### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-BNextendedexample.tex
Add files via upload

Bayesian network parameters
  Parameters of node Aval (multinomial distribution)
Conditional probability table:
, , Snow = high
     Earthquake
Aval         no       yes
  no  0.7500000 0.5000000
  yes 0.2500000 0.5000000
, , Snow = low
     Earthquake
Aval         no       yes
  no  0.6000000 0.6666667
  yes 0.4000000 0.3333333
  Parameters of node Bearing (multinomial distribution)
Conditional probability table:
       DTV
Bearing      high       low    medium
   bad  0.7000000 0.1666667 0.8571429
   good 0.3000000 0.8333333 0.1428571
  Parameters of node Bridge (multinomial distribution)
Conditional probability table:
, , Bearing = bad, Cracks = high, Flood = no, Joint = no, Rockfall = no
        Aval
Bridge   no yes
  high      1.0
  low       0.0
  medium    0.0
, , Bearing = good, Cracks = high, Flood = no, Joint = no, Rockfall = no
        Aval
Bridge    no yes
  high   0.5    
  low    0.0    
  medium 0.5    
, , Bearing = bad, Cracks = moderate, Flood = no, Joint = no, Rockfall = no
        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = moderate, Flood = no, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = small, Flood = no, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = small, Flood = no, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = high, Flood = yes, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = high, Flood = yes, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = moderate, Flood = yes, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high      1.0
  low       0.0
  medium    0.0

, , Bearing = good, Cracks = moderate, Flood = yes, Joint = no, Rockfall = no

        Aval
Bridge    no yes
  high   0.5    
  low    0.0    
  medium 0.5

, , Bearing = bad, Cracks = small, Flood = yes, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = small, Flood = yes, Joint = no, Rockfall = no

        Aval
Bridge   no yes
  high      1.0
  low       0.0
  medium    0.0

, , Bearing = bad, Cracks = high, Flood = no, Joint = yes, Rockfall = no

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = good, Cracks = high, Flood = no, Joint = yes, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = moderate, Flood = no, Joint = yes, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = moderate, Flood = no, Joint = yes, Rockfall = no

        Aval
Bridge   no yes
  high      1.0
  low       0.0
  medium    0.0

, , Bearing = bad, Cracks = small, Flood = no, Joint = yes, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = small, Flood = no, Joint = yes, Rockfall = no

        Aval
Bridge    no yes
  high   1.0    
  low    0.0    
  medium 0.0

, , Bearing = bad, Cracks = high, Flood = yes, Joint = yes, Rockfall = no

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = good, Cracks = high, Flood = yes, Joint = yes, Rockfall = no

        Aval
Bridge   no yes
  high      1.0
  low       0.0
  medium    0.0

, , Bearing = bad, Cracks = moderate, Flood = yes, Joint = yes, Rockfall = no

        Aval
Bridge    no yes
  high   1.0    
  low    0.0    
  medium 0.0

, , Bearing = good, Cracks = moderate, Flood = yes, Joint = yes, Rockfall = no

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = bad, Cracks = small, Flood = yes, Joint = yes, Rockfall = no

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = small, Flood = yes, Joint = yes, Rockfall = no

        Aval
Bridge    no yes
  high   1.0    
  low    0.0    
  medium 0.0

, , Bearing = bad, Cracks = high, Flood = no, Joint = no, Rockfall = yes

        Aval
Bridge    no yes
  high   1.0    
  low    0.0    
  medium 0.0

, , Bearing = good, Cracks = high, Flood = no, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = moderate, Flood = no, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high      1.0
  low       0.0
  medium    0.0

, , Bearing = good, Cracks = moderate, Flood = no, Joint = no, Rockfall = yes

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = bad, Cracks = small, Flood = no, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high      0.0
  low       0.0
  medium    1.0

, , Bearing = good, Cracks = small, Flood = no, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = high, Flood = yes, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high      0.0
  low       0.0
  medium    1.0

, , Bearing = good, Cracks = high, Flood = yes, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = moderate, Flood = yes, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = moderate, Flood = yes, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = small, Flood = yes, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = small, Flood = yes, Joint = no, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = high, Flood = no, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = good, Cracks = high, Flood = no, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high      0.0
  low       0.0
  medium    1.0

, , Bearing = bad, Cracks = moderate, Flood = no, Joint = yes, Rockfall = yes

        Aval
Bridge    no yes
  high   0.0    
  low    0.0    
  medium 1.0

, , Bearing = good, Cracks = moderate, Flood = no, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = small, Flood = no, Joint = yes, Rockfall = yes

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = good, Cracks = small, Flood = no, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high      0.0
  low       1.0
  medium    0.0

, , Bearing = bad, Cracks = high, Flood = yes, Joint = yes, Rockfall = yes

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = good, Cracks = high, Flood = yes, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = moderate, Flood = yes, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high      0.0
  low       0.0
  medium    1.0

, , Bearing = good, Cracks = moderate, Flood = yes, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

, , Bearing = bad, Cracks = small, Flood = yes, Joint = yes, Rockfall = yes

        Aval
Bridge    no yes
  high   0.0    
  low    1.0    
  medium 0.0

, , Bearing = good, Cracks = small, Flood = yes, Joint = yes, Rockfall = yes

        Aval
Bridge   no yes
  high         
  low          
  medium

  Parameters of node Cracks (multinomial distribution)

Conditional probability table:

          DTV
Cracks          high       low    medium
  high     0.4000000 0.2500000 0.4285714
  moderate 0.4000000 0.4166667 0.4285714
  small    0.2000000 0.3333333 0.1428571

  Parameters of node DTV (multinomial distribution)

Conditional probability table:
      high       low    medium 
0.3448276 0.4137931 0.2413793

  Parameters of node Earthquake (multinomial distribution)

Conditional probability table:
        no       yes 
0.6206897 0.3793103

  Parameters of node Flood (multinomial distribution)

Conditional probability table:

, , Snow = high

     Rain
Flood      high       low    medium
  no  0.6666667 0.5714286 0.3333333
  yes 0.3333333 0.4285714 0.6666667

, , Snow = low

     Rain
Flood      high       low    medium
  no  0.8000000 0.1666667 0.5000000
  yes 0.2000000 0.8333333 0.5000000

  Parameters of node Joint (multinomial distribution)

Conditional probability table:

     DTV
Joint      high       low    medium
  no  0.5000000 0.2500000 0.5714286
  yes 0.5000000 0.7500000 0.4285714

  Parameters of node Rain (multinomial distribution)

Conditional probability table:
      high       low    medium 
0.2758621 0.4482759 0.2758621

  Parameters of node Rockfall (multinomial distribution)

Conditional probability table:

        Earthquake
Rockfall        no       yes
     no  0.6666667 0.4545455
     yes 0.3333333 0.5454545

  Parameters of node Snow (multinomial distribution)

Conditional probability table:
      high       low 
0.5517241 0.4482759

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-BNsimplifiedexample.tex
Add files via upload

Bayesian network parameters
  Parameters of node Avalanche (multinomial distribution)
Conditional probability table:
         Earthquake
Avalanche        no       yes
      no  0.6666667 0.5454545
      yes 0.3333333 0.4545455
  Parameters of node Bridge (multinomial distribution)
Conditional probability table:
, , Rockfall = no
      Avalanche
Bridge        no       yes
   no  0.5000000 0.0000000
   yes 0.5000000 1.0000000
, , Rockfall = yes
      Avalanche
Bridge        no       yes
   no  0.8333333 0.8333333
   yes 0.1666667 0.1666667
  Parameters of node Earthquake (multinomial distribution)
Conditional probability table:
        no       yes 
0.6206897 0.3793103

  Parameters of node Rockfall (multinomial distribution)

Conditional probability table:

        Earthquake
Rockfall        no       yes
     no  0.6666667 0.4545455
     yes 0.3333333 0.5454545

  Bayesian network parameters

  Parameters of node Avalanche (multinomial distribution)

Conditional probability table:

         Earthquake
Avalanche        no       yes
      no  0.6304348 0.5312500
      yes 0.3695652 0.4687500

  Parameters of node Bridge (multinomial distribution)

Conditional probability table:

, , Rockfall = no

      Avalanche
Bridge        no       yes
   no  0.5000000 0.1666667
   yes 0.5000000 0.8333333

, , Rockfall = yes

      Avalanche
Bridge        no       yes
   no  0.7352941 0.7352941
   yes 0.2647059 0.2647059

  Parameters of node Earthquake (multinomial distribution)

Conditional probability table:
        no       yes 
0.5897436 0.4102564

  Parameters of node Rockfall (multinomial distribution)

Conditional probability table:

        Earthquake
Rockfall        no       yes
     no  0.6304348 0.4687500
     yes 0.3695652 0.5312500

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-avai-main/IMP-A-availability-network.R

@@ -1,0 +1,313 @@
+#This program is coded by Nam Lethanh for use in the class IMP-HS2014
+#Purpose: To calculate availability
+data <- read.csv("IMP-10-HS2014-A-availability-network.csv",header=TRUE,sep=",") #
+attach(data)
+TIME=20
+T=seq(0,TIME, by =1) #investigate time
+N=10
+cat("How data looks like?\n")
+print(data)
+
+#define function
+#reliability function for exponential distribution
+failureexp<-function(t,lambda){lambda*exp(-lambda*t)}
+survivalexp<-function(t,lambda){exp(-lambda*t)}
+#reliability function for weibull distribution
+
+failureweib<-function(t,theta,m){theta*m*(theta*t)^(m-1)*exp(-(theta*t)^m)}
+rateweib<-function(t,theta,m){theta*m*(theta*t)^(m-1)}
+survivalweib<-function(t,theta,m){exp(-(theta*t)^m)}
+#define structure function
+decompose<-function(x,y,z){
+  x*y+(1-x)*z
+}
+#calculating the structure function when link 1-3 is perfect and link 3-5 is perfect
+expression1<-function(xA1,xA4,x34,x45,xB5,xB2,x12,x23){
+(1-(1-xA1)*(1-xA4*(1-(1-x34)*(1-x45))))*(1-(1-xB5)*(1-xB2*(1-(1-x12)*(1-x23))))
+}
+
+#calculating the structure function when link 1-3 is perfect and link 3-5 is not perfect
+
+expression2<-function(xA1,xA4,x34,x45,xB5,xB2,x12,x23){
+xA1*x34*xB2*(1-(1-x12)*(1-x23))+xA1*x34*x45*xB5+xA4*x34*xB2*(1-(1-x12)*(1-x23))+xA4*x34*x45*xB5+xA1*xA4*x45*xB5*xB2*(1-(1-x12)*(1-x23))-xA1*xA4*x34*xB2*(1-(1-x12)*(1-x23))-xA1*x34*xB2*(1-(1-x12)*(1-x23))*x45*xB5-xA1*xA4*x34*x45*xB5-xA4*x34*xB2*(1-(1-x12)*(1-x23))*x45*xB5
+}
+
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-avai-main/IMP-A-availability.r

@@ -1,0 +1,91 @@
+#This program is coded by Nam Lethanh for use in the class IMP-HS2014
+#Purpose: To calculate availability
+data <- read.csv("IMP-A-availability.csv",header=TRUE,sep=",") #
+attach(data)
+T=50
+cat("How data looks like?\n")
+print(data)
+cat("range of observations \n")
+range<-max(fre)-min(fre)
+print(range)
+cat("range of observations - interval \n")
+
+rangetime<-c("9-12","13-16","17-20","21-24","25-28","29-32","33-36","37-40","41-44","45-48")
+
+fre<-c(2,11,24,27,20,15,16,7,2,2)
+plot.new()
+limy=c(0,1)
+barplot(fre, beside = TRUE, axes = FALSE, col = sample(colours(), 5), space = 0.5, names.arg=rangetime,ylim=c(0,30))
+axis(2, ylim = limy, col = "black", las = 1)
+mtext(expression(paste('Frequency')), side = 2, col = "black", line = 3)
+mtext(expression(paste('Range of time')), side = 1, col = "black", line = 3)
+box()
+
+library(graphics)
+par(new=TRUE)
+plot(fre,lwd=2,col="coral4",ylab="",xlab="",ylim=c(0,30),axes=FALSE,lty=1,type="b",pch=10,cex=0.2)
+legend("topright", inset = 0.05, col = "coral4", lty = 1, lwd = 2, legend = c("Fit spline"), bg = "aliceblue", cex = 0.8)
+
+#calculate the mean corrective intervention duration
+MCID<-sum((data$time)*data$fre)/sum(data$fre)
+cat("mean corrective intervention duration \n")
+print(MCID)
+#calculate the mean of sample
+VarX<-(sum(data$fre*(data$time-MCID)^2)/sum(data$fre))^(0.5)
+cat("value of standard deviation of sample \n")
+print(VarX)
+
+#calculate the geometric mean
+
+lognormalmean<-log(MCID)-0.5*(log(1+VarX^2/(MCID)^2))
+cat("value of mean of the lognormal distribution \n")
+print(lognormalmean)
+cat("value of the geometric mean \n")
+geomean<-exp(lognormalmean)
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-avai-main/IMP-E-availability.R

@@ -1,0 +1,122 @@
+#
+
+#This program is coded by Nam Lethanh for use in the class IMP
+#Purpose: To calculate reliability
+TIME=30
+T=seq(0,TIME, by =1) #investigate time
+N=6 #Number of item
+lambda<-matrix(double(1),nrow=1,ncol=N)
+lambda<-1/20 # value of hazard rate
+
+delta<-matrix(double(1),nrow=length(T),ncol=N-1)
+
+delta[1]<-1*7/365
+delta[2]<-2*7/365
+delta[3]<-5*7/365
+delta[4]<-4*7/365
+delta[5]<-3*7/365
+
+
+# define dimension of matrix
+fail<-matrix(double(1),nrow=length(T),ncol=N)
+survive<-matrix(double(1),nrow=length(T),ncol=N)
+availability<-matrix(double(1),nrow=length(T),ncol=N)
+
+
+#reliability function for exponential distribution
+failure<-function(t,lambda){lambda*exp(-lambda*t)}
+survival<-function(t,lambda){exp(-lambda*t)}
+#
+
+for (t in 1:length(T)){
+  for (n in 1:N){
+    if (n < N){
+      survive[t,n]<-survival(T[t],lambda)
+      availability[t,n]<-integrate(survival,0,t,lambda)$value/(integrate(survival,0,t,lambda)$value+1*delta[n]*(1-survive[t,n]))
+   #  availability[t,n]<-integrate(survival,0,t,lambda)$value/(integrate(survival,0,t,lambda)$value+delta[n]*integrate(failure,0,t,lambda)$value)
+    } else {
+      survive[t,n]=
+        survive[t,1]*survive[t,3]*survive[t,4]+
+        survive[t,1]*survive[t,3]*survive[t,5]+
+        survive[t,2]*survive[t,3]*survive[t,4]+
+        survive[t,2]*survive[t,3]*survive[t,5]+
+        survive[t,1]*survive[t,2]*survive[t,4]*survive[t,5]-
+        survive[t,1]*survive[t,2]*survive[t,3]*survive[t,4]-
+        survive[t,1]*survive[t,3]*survive[t,4]*survive[t,5]-
+        survive[t,1]*survive[t,2]*survive[t,3]*survive[t,5]-
+        survive[t,2]*survive[t,3]*survive[t,4]*survive[t,5]
+      
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-avai-main/IMP-E-tunnel.R

@@ -1,0 +1,61 @@
+#This program is coded by Nam Lethanh for use in the class IMP
+#Purpose: To calculate reliability, availability, and maintainability of tunnel
+
+data <- read.csv("IMP-E-tunnel.csv",header=TRUE,sep=",") #
+attach(data)
+TIME=20
+T=seq(0,TIME, by =1) #investigate time
+N=5 #Number of item
+M<-5 #number of configuration
+
+cat("how data look like? \n")
+print(data)
+theta<-matrix(double(1),nrow=M,ncol=N)
+
+MTBF<-matrix(double(1),nrow=M,ncol=1) #mean time between failure for each configuration
+Reli<-matrix(double(1),nrow=M,ncol=1) #Reliability for each configuration
+Avai<-matrix(double(1),nrow=M,ncol=1) #Availability for each configuration
+Cost<-matrix(double(1),nrow=M,ncol=1) #Cost for each configuration in 20 years
+Benefit_no_upfront<-matrix(double(1),nrow=M,ncol=1) #
+Benefit_with_upfront<-matrix(double(1),nrow=M,ncol=1) #
+
+for (i in 1:N){
+  for (j in 1:M){
+    theta[i,j]<-1/data[i,j+1]  
+  }
+}
+cat("Value of failure rate for each item i (column) in each configuration j (row) \n")
+print(theta)
+
+for (i in 1:N){
+  MTBF[i]<-1/sum(theta[i,])
+  Reli[i]<-exp(-1/MTBF[i]*1)
+  Avai[i]<-MTBF[i]/(MTBF[i]+data$MID[i]/12)
+  Cost[i]<-20*data$costcr[i]*data$MID[i]*(1-Reli[i])/(MTBF[i]+data$MID[i]/12)
+  
+}
+cat("Value of mean time between failure (MTBF) of each configuration \n")
+print(MTBF)
+
+cat("Reliability of each configuration (Reli) \n")
+print(Reli)
+
+cat("Availability of each configuration \n")
+print(Avai)
+cat("Cost of each configuration in 20 years \n")
+print(Cost)
+
+for (i in 1:M){
+  if (i==1){
+    Benefit_no_upfront[i]<-Cost[i]-Cost[i]
+    Benefit_with_upfront[i]<-Benefit_no_upfront[i]-data$costint[i]
+  } else {
+    Benefit_no_upfront[i]<-Cost[1]-Cost[i]
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-avai-main/tesr.R

@@ -1,0 +1,90 @@
+#This program is coded by Nam Lethanh for use in the class IMP-HS2014
+#Purpose: To calculate availability
+data <- read.csv("IMP-5-A-availability-network.csv",header=TRUE,sep=",") #
+attach(data)
+T=20
+N=10
+cat("How data looks like?\n")
+print(data)
+
+
+#define function
+#reliability function for exponential distribution
+failureexp<-function(t,lambda){lambda*exp(-lambda*t)}
+survivalexp<-function(t,lambda){exp(-lambda*t)}
+#reliability function for weibull distribution
+
+failureweib<-function(t,theta,m){theta*m*(theta*t)^(m-1)*exp(-(theta*t)^m)}
+rateweib<-function(t,theta,m){theta*m*(theta*t)^(m-1)}
+survivalweib<-function(t,theta,m){exp(-(theta*t)^m)}
+
+#define structure function
+decompose<-function(x,y,z){
+  x*y+(1-x)*z
+}
+#calculating the structure function when link 1-3 is perfect and link 3-5 is perfect
+expression1<-function(xA1,xA4,x34,x45,xB5,xB2,x12,x23){
+  (1-(1-xA1)*(1-xA4*(1-(1-x34)*(1-x45))))*(1-(1-xB5)*(1-xB2*(1-(1-x12)*(1-x23))))
+}
+
+#calculating the structure function when link 1-3 is perfect and link 3-5 is not perfect
+
+expression2<-function(xA1,xA4,x34,x45,xB5,xB2,x12,x23){
+  xA1*x34*xB2*(1-(1-x12)*(1-x23))+xA1*x34*x45*xB5+xA4*x34*xB2*(1-(1-x12)*(1-x23))+xA4*x34*x45*xB5+xA1*xA4*x45*xB5*xB2*(1-(1-x12)*(1-x23))-xA1*xA4*x34*xB2*(1-(1-x12)*(1-x23))-xA1*x34*xB2*(1-(1-x12)*(1-x23))*x45*xB5-xA1*xA4*x34*x45*xB5-xA4*x34*xB2*(1-(1-x12)*(1-x23))*x45*xB5
+}
+
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-BayesianNetwork.r

@@ -1,0 +1,23 @@
+# This program was coded by  Nam Lethanh (lethanh@ibi.baug.ethz.ch)
+#Bayesian network example
+library(deal) #this is dependency package
+data <- read.csv("BayesianNetwork-dataaccident.csv",header=TRUE)
+accident<-network(data)
+plot(accident)
+accident.nd <- nodes(accident)# the list of nodes
+accident.j <- jointprior(accident)
+accident.prior <- jointprior(accident)
+#accident.prior <- jointprior(accident,12)
+accident <- learn(accident,data,accident.prior)$nw
+nodes(accident)$Time$condprior
+nodes(accident)$Time$condposterior
+nodes(accident)$Time$loglik
+accident$score
+allaccident <- networkfamily(data,accident,accident.prior)
+allaccident <- nwfsort(allaccident$nw)
+print(allaccident)
+plot(allaccident)
+accident.s <- autosearch(accident,data,accident.prior)$nw
+plot(accident.s)
+#THE END
+cat("THE END")

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-bayesian-linear.r

@@ -1,0 +1,38 @@
+################################
+# example of Bayesian estimation for a linear model
+# Coded by Nam Lethanh (lethanh@ibi.baug.ethz.ch)
+################################
+library(LearnBayes) #requires R package to perform Bayesian estimation
+data=read.csv("accidentdata.csv", header=TRUE) #Data
+attach(data) 
+plot(noaccident,CS,xlab="Numbers of accident")
+plot(noaccident,dtv,xlab="Numbers of accident")
+plot(noaccident,slope,xlab="Numbers of accident")
+plot(noaccident,speed,xlab="Numbers of accident")
+##### Least-squares fit
+fit=lm(noaccident~CS+dtv+slope+speed,data=data,x=TRUE,y=TRUE)
+summary(fit)
+cat("RESULTS OF LEAST SQUARED FIT METHOD \n")
+print(summary(fit))
+##### Sampling from posterior
+N=10000 #sampling numbers
+theta.sample=blinreg(fit$y,fit$x,N) #Sampling from Posterior function
+hist(theta.sample$beta[,1],main="Intercept",
+     xlab=expression(beta[0]))
+hist(theta.sample$beta[,2],main="CS",
+     xlab=expression(beta[1]))
+hist(theta.sample$beta[,3],main="dtv",
+     xlab=expression(beta[2]))
+hist(theta.sample$beta[,4],main="slope",
+     xlab=expression(beta[3]))
+hist(theta.sample$beta[,5],main="speed",
+     xlab=expression(beta[4]))
+hist(theta.sample$sigma,main="error",
+     xlab=expression(epsilon))
+cat("RESULTS OF BAYESIAN ESTIMATION \n")
+cat("Confident interval \n")
+quantile1<-apply(theta.sample$beta,2,quantile,c(.05,.5,.95))
+quantile2<-quantile(theta.sample$sigma,c(.05,.5,.95))
+print(quantile1)
+print(quantile2)
+cat("THE END")

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-bayesian-weibull.r

@@ -1,0 +1,47 @@
+#This program was coded by Nam Lethanh for use in IMP class
+data=read.csv("pipelinedata.csv", header=TRUE) #Data
+attach(data) 
+library(LearnBayes) #R package for Bayesian statistics
+library(splines)
+library(survival) #R package for regression analysis with MLE approach
+#BEGIN Bayesian estimation approach on data
+#-----function to define posterior likelihood
+weibullregpost=function (theta, data)
+{
+  logf = function(t, c, x, sigma, mu, beta) {
+    z = (log(t) - mu - x %*% beta)/sigma
+    f = 1/sigma * exp(z - exp(z))
+    S = exp(-exp(z))
+    c * log(f) + (1 - c) * log(S) }
+  k = dim(data)[2]
+  p = k - 2
+  t = data[, 1]
+  c = data[, 2]
+  X = data[, 3:k]
+  sigma = exp(theta[1])
+  mu = theta[2]
+  beta = array(theta[3:k], c(p, 1))
+  return(sum(logf(t, c, X, sigma, mu, beta)))
+}
+#-----------------------------
+start=c(-.5,9,.5,-.05)
+d=cbind(time,status,waterlost,depth)
+fit=laplace(weibullregpost,start,d)
+proposal=list(var=fit$var,scale=1.5)
+bayesfit=rwmetrop(weibullregpost,proposal,fit$mode,10000,d) #This function rwmetrop is a standard function in LearnBayes R package describing the Metro-Polish Hasting Algorithm used to generate data based on MCMC.
+par(mfrow=c(2,2))
+sigma=exp(bayesfit$par[,1])
+mu=bayesfit$par[,2]
+beta1=bayesfit$par[,3]
+beta2=bayesfit$par[,4]
+hist(mu,xlab="Intercept", main="")
+hist(sigma,xlab="sigma",main="")
+#hist(mu,xlab="mu",main="")
+hist(beta1,xlab="Water lost",main="")
+hist(beta2,xlab="depth",main="")
+#END of Bayesian estimation approach for the data
+cat("RESULTS OF BAYESIAN ESTIMATION \n")
+cat("Confident interval \n")
+beta<-c(mu,beta1,beta2)
+cat("THE END")
+mean(beta1)

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-mle-weibull.r

@@ -1,0 +1,24 @@
+#This program is coded by Nam Lethanh @IBI, ETH Z?rich.
+# ESTIMATE THE MODEL'S PARAMETER and Draw the graph
+library(MASS)
+library(survival)
+data <- read.csv("pipelinedata.csv",header=TRUE,sep=",") 
+attach(data)
+sr.fit = survreg(Surv(time,status)~1+waterlost+depth,dist='weibull')
+summary(sr.fit)
+cat('model parameters \n')
+print(summary(sr.fit))
+scale=exp(sr.fit$icoef[1])
+shape=1/sr.fit$scale[1]
+YearMax=2000 # this is the time frame of investigation
+plot.new()
+#par(mar=c(5, 4, 4, 1) + 0.1)
+x=seq(0,YearMax,length=200)
+curve(pweibull(x, shape=shape,scale=scale,  lower.tail=FALSE), from=0, to=YearMax, col='red', lwd=2, ylab='', xlab='',bty='n',ylim=c(0,1),axes=FALSE)
+axis(2, ylim=c(0,1),col="black",las=1)  ## las=1 makes horizontal labels
+mtext(expression(paste("Reliability")),side=2,col="black",line=2.2) 
+axis(1,c(seq(0,YearMax,200)))
+mtext(expression(paste("Time (tus)")),side=1,col="black",line=2.2) 
+#draw the grid on the chart
+grid(10, 10, col = "lightgray", lty = "dotted",lwd = par("lwd"), equilogs = TRUE)
+box()

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/IMP-priordistribution.r

@@ -1,0 +1,20 @@
+# This program was coded by  Nam Lethanh (lethanh@ibi.baug.ethz.ch)
+#Bayesian statistic example - Accident risk analysis
+library(LearnBayes) #this is dependency package
+p = seq(0.05, 0.95, by = 0.1)
+prior = c(1, 5.2, 8, 7.2, 4.6, 2.1, 0.7, 0.1, 0, 0)
+prior = prior/sum(prior)
+plot(p, prior, type = "h", ylab="Prior Probability")
+data = c(11, 16)
+post = pdisc(p, prior, data)
+#post=round(cbind(p, prior, post),2)
+print(post)
+library(lattice)
+PRIOR=data.frame("prior",p,prior)
+POST=data.frame("posterior",p,post)
+names(PRIOR)=c("Type","P","Probability")
+names(POST)=c("Type","P","Probability")
+data=rbind(PRIOR,POST)
+print(xyplot(Probability~P|Type,data=data,layout=c(1,2), type="h",lwd=3,col="black"))
+#THE END
+cat("THE END")

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/bayesiannetwork-bridgereliabilitya.r

@@ -1,0 +1,52 @@
+#Created by Nam Lethanh for the IMP class
+library(bnlearn) #pls install the bnlearn package from CRAN
+
+#------------A simplified BN example-----------
+
+#creating an empty a network in abstract level
+
+dag1 <- empty.graph(nodes = c("Earthquake","Avalanche","Rockfall","Bridge"))
+#creating arcs for the network
+dag1 <- set.arc(dag1, from = "Earthquake", to = "Avalanche")
+dag1 <- set.arc(dag1, from = "Earthquake", to = "Rockfall")
+dag1 <- set.arc(dag1, from = "Avalanche", to = "Bridge")
+dag1 <- set.arc(dag1, from = "Rockfall", to = "Bridge")
+dag1 <- model2network("[Earthquake][Avalanche|Earthquake][Rockfall|Earthquake][Bridge|Avalanche:Rockfall]")
+plot(dag1) #plot the structure of the network
+graphviz.plot(dag1) ##plot the structure of the network but make it more appealling
+library(gRain)
+#Define the conditional probability table for the network CPT
+yn <- c("yes","no")
+c <- cptable(~Earthquake, values=c(0.01,0.99), levels=yn)
+s.c <- cptable(~Avalanche | Earthquake, values=c(0.7,0.3,0.1,0.9), levels=yn)
+r.c <- cptable(~Rockfall | Earthquake, values=c(0.4,0.6,0.05,0.95), levels=yn)
+w.src <- cptable(~Bridge | Avalanche:Rockfall,values=c(0.6,0.4,0.3,0.7,0.2,0.8,0.05,0.95),levels=yn)
+cpt.list <- compileCPT(list(c,s.c,r.c,w.src))
+print(cpt.list)
+#Create a graphical representation
+bnet <- grain(cpt.list)
+plot(bnet)
+# Compile network (details follow)
+bnet <- compile(bnet)
+# Query network to find marginal probabilities of diseases
+
+p<-querygrain(bnet, nodes=c("Avalanche","Rockfall","Bridge"))
+print(p)
+
+# Estimating CPT from data
+survey=read.csv("bayesiannetwork-bridgereliabilitya.csv", header=TRUE) #Data
+attach(survey) 
+#-------------Model from data file -------------------
+#define sets of variables for each nodes
+Earthquake.state <- c("yes", "no")
+Aval.state <- c("yes", "no")
+Rockfall.state <- c("yes", "no")
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/bayesiannetwork-bridgereliabilityb.r

@@ -1,0 +1,232 @@
+#Created by Nam Lethanh for the IMP class
+library(bnlearn) #pls install the bnlearn package from CRAN
+
+#------------A simplified BN example-----------
+
+#creating an empty a network in abstract level
+
+dag1 <- empty.graph(nodes = c("Earthquake","Avalanche","Rockfall","Bridge"))
+#creating arcs for the network
+dag1 <- set.arc(dag1, from = "Earthquake", to = "Avalanche")
+dag1 <- set.arc(dag1, from = "Earthquake", to = "Rockfall")
+dag1 <- set.arc(dag1, from = "Avalanche", to = "Bridge")
+dag1 <- set.arc(dag1, from = "Rockfall", to = "Bridge")
+dag1 <- model2network("[Earthquake][Avalanche|Earthquake][Rockfall|Earthquake][Bridge|Avalanche:Rockfall]")
+plot(dag1) #plot the structure of the network
+graphviz.plot(dag1) ##plot the structure of the network but make it more appealling
+library(gRain)
+#Define the conditional probability table for the network CPT
+yn <- c("yes","no")
+c <- cptable(~Earthquake, values=c(0.01,0.99), levels=yn)
+s.c <- cptable(~Avalanche | Earthquake, values=c(0.7,0.3,0.1,0.9), levels=yn)
+r.c <- cptable(~Rockfall | Earthquake, values=c(0.4,0.6,0.05,0.95), levels=yn)
+w.src <- cptable(~Bridge | Avalanche:Rockfall,values=c(0.6,0.4,0.3,0.7,0.2,0.8,0.05,0.95),levels=yn)
+cpt.list <- compileCPT(list(c,s.c,r.c,w.src))
+print(cpt.list)
+#Create a graphical representation
+bnet <- grain(cpt.list)
+plot(bnet)
+# Compile network (details follow)
+bnet <- compile(bnet)
+# Query network to find marginal probabilities of diseases
+
+p<-querygrain(bnet, nodes=c("Avalanche","Rockfall","Bridge"))
+print(p)
+
+
+
+
+
+
+#-------------Extended Example for more complex structure
+
+#creating an empty a network in detail
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/bayesiannetwork-failure.r

@@ -1,0 +1,29 @@
+library(gRain)
+#Define the conditional probability tables for the network CPT
+yn <- c("yes","no")
+a <- cptable(~asia, values=c(1,99), levels=yn)
+t.a <- cptable(~tub | asia, values=c(5,95,1,99), levels=yn)
+s <- cptable(~smoke, values=c(5,5), levels=yn)
+l.s <- cptable(~lung | smoke, values=c(1,9,1,99), levels=yn)
+b.s <- cptable(~bronc | smoke, values=c(6,4,3,7), levels=yn)
+e.lt <- cptable(~either | lung:tub,values=c(1,0,1,0,1,0,0,1),levels=yn)
+x.e <- cptable(~xray | either, values=c(98,2,5,95), levels=yn)
+d.be <- cptable(~dysp | bronc:either, values=c(9,1,7,3,8,2,1,9),levels=yn)
+cpt.list <- compileCPT(list(a, t.a, s, l.s, b.s, e.lt, x.e, d.be))
+print(cpt.list)
+#Create a graphical representation
+bnet <- grain(cpt.list)
+# Compile network (details follow)
+bnet <- compile(bnet)
+# Query network to find marginal probabilities of diseases
+querygrain(bnet, nodes=c("tub","lung","bronc"))
+
+
+
+
+
+
+
+
+
+

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/bayesiannetwork-travel.r

@@ -1,0 +1,45 @@
+library(bnlearn)
+#creating an empty a network
+dag <- empty.graph(nodes = c("A", "S", "E", "O", "R", "T"))
+#creating arcs for the network
+dag <- set.arc(dag, from = "A", to = "E")
+dag <- set.arc(dag, from = "S", to = "E")
+dag <- set.arc(dag, from = "E", to = "O")
+dag <- set.arc(dag, from = "E", to = "R")
+dag <- set.arc(dag, from = "O", to = "T")
+dag <- set.arc(dag, from = "R", to = "T")
+plot(dag)
+#define sets of variables for each nodes
+A.lv <- c("young", "adult", "old")
+S.lv <- c("M", "F")
+E.lv <- c("high", "uni")
+O.lv <- c("emp", "self")
+R.lv <- c("small", "big")
+T.lv <- c("car", "train", "other")
+#Assign conditional probability table for each node
+A.prob <- array(c(0.30, 0.50, 0.20), dim = 3,dimnames = list(A = A.lv))
+S.prob <- array(c(0.60, 0.40), dim = 2,dimnames = list(S = S.lv))
+O.prob <- array(c(0.96, 0.04, 0.92, 0.08), dim = c(2, 2),dimnames = list(O = O.lv, E = E.lv))
+R.prob <- array(c(0.25, 0.75, 0.20, 0.80), dim = c(2, 2), dimnames = list(R = R.lv, E = E.lv))
+E.prob <- array(c(0.75, 0.25, 0.72, 0.28, 0.88, 0.12, 0.64,
+                  0.36, 0.70, 0.30, 0.90, 0.10), dim = c(2, 3, 2),
+                dimnames = list(E = E.lv, A = A.lv, S = S.lv))
+T.prob <- array(c(0.48, 0.42, 0.10, 0.56, 0.36, 0.08, 0.58,
+                  0.24, 0.18, 0.70, 0.21, 0.09), dim = c(3, 2, 2),
+                dimnames = list(T = T.lv, O = O.lv, R = R.lv))
+#re-arrange the network so it looks nicer
+dag1 <- model2network("[A][S][E|A:S][O|E][R|E][T|O:R]")
+plot(dag3)
+cpt <- list(A = A.prob, S = S.prob, E = E.prob, O = O.prob,
+             R = R.prob, T = T.prob) #list all individual condition table into a network conditional table
+bn1 <- custom.fit(dag, cpt) #calculate the marginal probability distribution based on the defined network and conditional probability table
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-bayesian/bayesiannetwork-wetgrass.r

@@ -1,0 +1,29 @@
+library(gRain)
+#Define the conditional probability table for the network CPT
+trueorfault <- c("fault","true")
+c <- cptable(~cloudy, values=c(50,50), levels=trueorfault)
+s.c <- cptable(~sprinker | cloudy, values=c(50,50,90,10), levels=trueorfault)
+r.c <- cptable(~rain | cloudy, values=c(80,20,20,80), levels=trueorfault)
+w.src <- cptable(~wet | sprinker:rain,values=c(100,0,10,90,10,90,1,99),levels=trueorfault)
+cpt.list <- compileCPT(list(c,s.c,r.c,w.src))
+print(cpt.list)
+#Create a graphical representation
+bnet <- grain(cpt.list)
+plot(bnet)
+# Compile network (details follow)
+bnet <- compile(bnet)
+# Query network to find marginal probabilities of diseases
+
+p<-querygrain(bnet, nodes=c("sprinker","rain","wet"))
+print(p)
+
+
+
+
+
+
+
+
+
+
+

### 2019-05-05 InfrastructureMaintenance/Chapters/avai-main.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Availability and maintainability}
\label{avaimain} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Bryan T. Adey and Nam Lethanh}
\section{Maintainability of an item}
When managing infrastructure the maintainability of an item is often a useful
bit of information, i.e. how much effort is it to restore the item so that it
once again provides an adequate LOS. For example, if the speeds of trains need to
be reduced on a rail link due to deformations in a track, how much will it cost
to fix the rail link in a way that the trains can once again run at normal speed,
and how long will this take?

The maintainability of an item only has to do with the execution of
interventions on the item. The maintainability and reliability together give the
availability of an item.

\subsection{Time}
In order to estimate how long it will take to restore an item so that it
provides an adequate LOS, it is necessary to estimate the steps to be taken from
the instant that an inadequate LOS is not provided until it is restored. These
steps include the determination that a failure has occurred, the determination of
what exactly went wrong, the determination of how to restore the item, the
execution of the intervention, the testing that the item again can provide an
adequate LOS and then putting it back in operation so that it does provide an
adequate LOS. An illustration of this for the restoration of a part of an
electricity distribution network using Business Process Modelling Notation is
shown in Figure \ref{avaimain:1}.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{avaimain-1.eps}
\caption{Simplified process for the restoration of a part of an electricity
distribution network}\label{avaimain:1}
% \end{center}
\end{figure}
Once the process has been identified it is necessary to estimate the length of
time that each of the activities will take. This depends on many factors,
including the quantity and skill levels of persons required and available, the
quantity of required replacement parts required and available, and the amount of
time required to react to an indication that something is not working, sometimes
referred to as responsiveness.
\subsection{Indicators}
Indicators that are often used to express maintainability are the
mode\footnote{The duration or costs that occurs the most often},
mean\footnote{The average of the sum of all durations or costs divided by the
number of values} and maximum\footnote{The largest of all durations or costs
observed or expected. This is also often given as a percentile of the largest of
all durations or costs.} amount of duration or cost of interventions. Indicators
are usually developed for specific types of interventions, e.g. the mean duration
of a corrective intervention is 3 months, or the mean duration of a preventive
intervention is 1 week.
\section{Maintainability of an item comprised of sub-items}
The maintainability of an item can be estimated by taking into consideration the
structure of the sub-items in the item, and the maintainability of the sub-items.
This is advantageous if there is little data on the interventions that have been
executed on the item itself, e.g. a bridge, but abundant data on interventions
that have been executed on sub-items similar to those of which the item is
composed, e.g. a concrete abutment.

If it can be assumed that each intervention is only used to restore one sub-item
so that it provides an adequate LOS, the durations of the interventions can be
simply combined, e.g. if it takes 1 week to repair an abutment and 2 weeks to
repair a deck and both happen once every 10 years, the average time spent to
repair a bridge per year, i.e. its' maintainability, is 0.30 weeks ((1+2)/10).
If, however, 50\% of the time the abutment and deck are repaired at the same time
which takes 2 weeks, 50\% of the time they are repaired separately than the
average time spent to repair a bridge per year, i.e. its maintainability, is
0.175 weeks ((0.5(2)+0.25(1+2))/10).

The basic steps to combine sub-items, if it can be assumed that each
intervention is only used to restore one sub-item so that it provides an adequate
LOS, are:
\begin{itemize}
	\item divide all sub-items by type, so it can be assumed that each sub-item per
sub-item type behaves in the same way,
	\item determine the rate of occurrence of the type of intervention of interest, e.g.
corrective intervention, per sub-item type,
	\item estimate the rate of occurrence of the type of intervention, per sub-item type
by multiplying the rate of occurrence of the type of intervention by the number
of sub-items per sub-type,
	\item estimate the total duration of the type of intervention per sub-item type by
multiplying the rate of occurrence of the type of intervention by the mean
duration of the type of intervention per sub-item type
	\item estimate the rate of occurrence of an intervention on the item by summing the
rates of occurrence of the types of interventions per sub-item
	\item estimate the expected intervention duration, or maintainability, of the item by
summing the total duration of the type of intervention per sub-item.
\end{itemize}
\section{Availability of an item}
The availability of an item is based on both the reliability of an item and the
maintainability of an item. Simply, availability is:
\begin{eqnarray}
&& {A_i} = \frac{{E\left[ {t_i^{adequate}} \right]}}{{E\left[ {t_i^{adequate}}
\right] + E\left[ {t_i^{inadequate}} \right]}}
\label{avaimaineq:1}
%(1)
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$t_i^{adequate}$:] the duration of time item \textit{i} provides an adequate LOS,
\item[$t_i^{inadequate}$:] the duration of time item \textit{i} provides an inadequate LOS
\end{description}
%\end{flushright}
\end{adjustwidth}
When investigating the availability of an item the exact definition of
availability used is important. One should be explicit about
\begin{itemize}
	\item the types of interventions being considered, e.g. preventive and corrective
interventions or only corrective interventions
	\item the type of time being considered, e.g. is the time spent waiting for parts and
the time on-site to be considered, or only the time on-site.
\end{itemize}
Different measure of availability serve different purposes.
\subsection{Example}
A road has been in service for 10 years (3'652 days). Over those 10 years, 6
interventions have been executed. Some of those have been due to rock falls, some
due to scour and some due to chloride induced corrosion. The total amount of time
that an adequate LOS was not provided was 156 days, i.e. traffic flow was
restricted for 156 days in the last 3'652 days.
\subsubsection{Question A}
What was the availability of the road link?
\subsubsection{Answer A}
The total length of time when an adequate LOS was provided was 3'496 days (3'652
-- 156). The rate of occurrence of interventions was 0.001716 interventions/day
(6/3'496). The mean time between interventions is therefore 582.7 days
(1/0.001716). The availability is then
\begin{eqnarray}
&& \begin{array}{l}
 {A_i} = \frac{{E\left[ {t_i^a} \right]}}{{E\left[ {t_i^a} \right] + E\left[
{t_i^n} \right]}} = \frac{{582.7}}{{582.7 + 156/6}} = 0.9573
\end{array} \label{avaimaineq:2}
%(2)
\end{eqnarray}
\subsubsection{Question B}
Is this also the availability of the road next year?
\subsubsection{Answer B}
It depends on the types of failures and how they were repaired. In order to
state that the availability of the road next year is also 0.9573, it would be
necessary to be able to model the time between interventions as an exponential
distribution, meaning that rate of occurrence of failure can be modelled as
constant. This is essentially saying that no matter how many failure have
occurred in the past there will be on average the same number in the future. This
is in many cases not correct. For example, if there is an area in which a
landslide damages a road, it may be that there will never be another landslide in
this area. Or, if many interventions has to be executed due the corrosion of the
reinforcement in deck slabs and waterproofing layers on all than the number of
interventions per year in the future would change.
\section{Availability of an item comprised of sub-items}
In determining the availability of an item comprised of sub-items it is
important to explicitly take into consideration the structure of the item. The
availability of item \textit{i} comprised of sub-items connected in series is
given by:
\begin{eqnarray}
&& {A_i} = \prod\limits_j^J {\left( {\frac{{E\left[ {t_j^a} \right]}}{{E\left[
{t_j^a} \right] + E\left[ {t_j^n} \right]}}} \right)}
\label{avaimaineq:3}
%(3)
\end{eqnarray}
\subsection{Example}
You own a water distribution network that is connected as shown in Figure
\ref{avaimain:2}. The five pipes in the network have life expectancies of 20 years.
The life expectancies can be modelled using an exponential distribution. It takes
1 week to repair pipes 1 and 4, 2 weeks for pipes 2 and 5 and 3 weeks for pipe 3,
once it breaks.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{avaimain-2.eps}
\caption{Water distribution network}\label{avaimain:2}
% \end{center}
\end{figure}
\subsubsection{Question C}
What is the reliability of the network?
\subsubsection{Question D}
What is the availability of the network?
\subsubsection{Answer C}
Using the pivotal decomposition method, the structure function of the network is
\begin{eqnarray}
&& \begin{array}{l}
\phi (\vec x) = {x_1}{x_3}{x_4} + {x_1}{x_3}{x_5} + {x_2}{x_3}{x_4} +
{x_2}{x_3}{x_5} + {x_1}{x_2}{x_4}{x_5}\\
{\rm{          }} - {x_1}{x_2}{x_3}{x_4} - {x_1}{x_3}{x_4}{x_5} -
{x_1}{x_2}{x_3}{x_5} - {x_2}{x_3}{x_4}{x_5}
\end{array}
\label{avaimaineq:4}
%(4)
\end{eqnarray}
The steps to arrive at the above function has been described in section 3.3 of
the script of week 9.

In term of reliability, the structure function can be rewritten as
\begin{eqnarray}
&& \begin{array}{l}
{R_{network}}(t) = {R_1}(t) \cdot {R_3}(t) \cdot {R_4}(t) + {R_1}(t) \cdot
{R_3}(t) \cdot {R_5}(t) + {R_2}(t) \cdot {R_3}(t) \cdot {R_4}(t) + {R_2}(t) \cdot
{R_3}(t) \cdot {R_5}(t)\\
+ {R_1}(t) \cdot {R_2}(t) \cdot {R_4}(t) \cdot {R_5}(t){\rm{ }} - {R_1}(t)
\cdot {R_2}(t) \cdot {R_3}(t) \cdot {R_4}(t)\\
- {R_1}(t) \cdot {R_3}(t) \cdot {R_4}(t) \cdot {R_5}(t) - {R_1}(t) \cdot
{R_2}(t) \cdot {R_3}(t) \cdot {R_5}(t) - {R_2}(t) \cdot {R_3}(t) \cdot {R_4}(t)
\cdot {R_5}(t)
\end{array}
\label{avaimaineq:5}
%(5)
\end{eqnarray}
The reliability of any item \textit{i} at time \textit{t} is defined in \eqref{eqreliability:7}

As the expected time until an inadequate LOS is provided is about 20 years, the
``failure rate'' can be estimated as follows:

\[
 \int\limits_0^{ + \infty } {{R_i}(t)} dt = \int\limits_0^{ + \infty } {\exp
\left( { - {\theta _i} \cdot t} \right)dt}  = \frac{1}{{{\theta _i}}} = 20{\rm{
years}}
\]
\[
 {\theta _i} = \frac{1}{{20}} = 0.05
\]

Using equations \eqref{eqreliability:7} the reliability of each sub-item, and the item, in 30 years
can be estimated (Table \ref{tblavaimain:1}). For example, the probability that sub-item
1 will provide an adequate LOS until the start of year 14 is 0.522 and the
probability that the item will provide an adequate LOS until year 14 is 0.346.
This is illustrated in Figure \ref{avaimain:3}. Only one curve is used to represent
the reliability of a sub-item, since the reliability for all sub-items is the
same.
\begin{table}
\caption{Reliability}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Time} & \multicolumn{6}{c|}{Reliability} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{(years)} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{Item} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{0.951} & \multicolumn{1}{c|}{0.951} & \multicolumn{1}{c|}{0.951} & \multicolumn{1}{c|}{0.951} & \multicolumn{1}{c|}{0.951} & \multicolumn{1}{c|}{0.987} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{0.905} & \multicolumn{1}{c|}{0.905} & \multicolumn{1}{c|}{0.905} & \multicolumn{1}{c|}{0.905} & \multicolumn{1}{c|}{0.905} & \multicolumn{1}{c|}{0.952} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{0.861} & \multicolumn{1}{c|}{0.861} & \multicolumn{1}{c|}{0.861} & \multicolumn{1}{c|}{0.861} & \multicolumn{1}{c|}{0.861} & \multicolumn{1}{c|}{0.904} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{0.819} & \multicolumn{1}{c|}{0.819} & \multicolumn{1}{c|}{0.819} & \multicolumn{1}{c|}{0.819} & \multicolumn{1}{c|}{0.819} & \multicolumn{1}{c|}{0.847} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{0.779} & \multicolumn{1}{c|}{0.779} & \multicolumn{1}{c|}{0.779} & \multicolumn{1}{c|}{0.779} & \multicolumn{1}{c|}{0.779} & \multicolumn{1}{c|}{0.786} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{0.741} & \multicolumn{1}{c|}{0.741} & \multicolumn{1}{c|}{0.741} & \multicolumn{1}{c|}{0.741} & \multicolumn{1}{c|}{0.741} & \multicolumn{1}{c|}{0.723} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{0.705} & \multicolumn{1}{c|}{0.705} & \multicolumn{1}{c|}{0.705} & \multicolumn{1}{c|}{0.705} & \multicolumn{1}{c|}{0.705} & \multicolumn{1}{c|}{0.66} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{0.670} & \multicolumn{1}{c|}{0.670} & \multicolumn{1}{c|}{0.670} & \multicolumn{1}{c|}{0.670} & \multicolumn{1}{c|}{0.670} & \multicolumn{1}{c|}{0.599} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{0.638} & \multicolumn{1}{c|}{0.638} & \multicolumn{1}{c|}{0.638} & \multicolumn{1}{c|}{0.638} & \multicolumn{1}{c|}{0.638} & \multicolumn{1}{c|}{0.541} \\ 
\hline
\multicolumn{1}{|c|}{11} & \multicolumn{1}{c|}{0.607} & \multicolumn{1}{c|}{0.607} & \multicolumn{1}{c|}{0.607} & \multicolumn{1}{c|}{0.607} & \multicolumn{1}{c|}{0.607} & \multicolumn{1}{c|}{0.487} \\ 
\hline
\multicolumn{1}{|c|}{12} & \multicolumn{1}{c|}{0.577} & \multicolumn{1}{c|}{0.577} & \multicolumn{1}{c|}{0.577} & \multicolumn{1}{c|}{0.577} & \multicolumn{1}{c|}{0.577} & \multicolumn{1}{c|}{0.436} \\ 
\hline
\multicolumn{1}{|c|}{13} & \multicolumn{1}{c|}{0.549} & \multicolumn{1}{c|}{0.549} & \multicolumn{1}{c|}{0.549} & \multicolumn{1}{c|}{0.549} & \multicolumn{1}{c|}{0.549} & \multicolumn{1}{c|}{0.389} \\ 
\hline
\multicolumn{1}{|c|}{14} & \multicolumn{1}{c|}{0.522} & \multicolumn{1}{c|}{0.522} & \multicolumn{1}{c|}{0.522} & \multicolumn{1}{c|}{0.522} & \multicolumn{1}{c|}{0.522} & \multicolumn{1}{c|}{0.346} \\ 
\hline
\multicolumn{1}{|c|}{15} & \multicolumn{1}{c|}{0.497} & \multicolumn{1}{c|}{0.497} & \multicolumn{1}{c|}{0.497} & \multicolumn{1}{c|}{0.497} & \multicolumn{1}{c|}{0.497} & \multicolumn{1}{c|}{0.307} \\ 
\hline
\multicolumn{1}{|c|}{16} & \multicolumn{1}{c|}{0.472} & \multicolumn{1}{c|}{0.472} & \multicolumn{1}{c|}{0.472} & \multicolumn{1}{c|}{0.472} & \multicolumn{1}{c|}{0.472} & \multicolumn{1}{c|}{0.272} \\ 
\hline
\multicolumn{1}{|c|}{17} & \multicolumn{1}{c|}{0.449} & \multicolumn{1}{c|}{0.449} & \multicolumn{1}{c|}{0.449} & \multicolumn{1}{c|}{0.449} & \multicolumn{1}{c|}{0.449} & \multicolumn{1}{c|}{0.241} \\ 
\hline
\multicolumn{1}{|c|}{18} & \multicolumn{1}{c|}{0.427} & \multicolumn{1}{c|}{0.427} & \multicolumn{1}{c|}{0.427} & \multicolumn{1}{c|}{0.427} & \multicolumn{1}{c|}{0.427} & \multicolumn{1}{c|}{0.212} \\ 
\hline
\multicolumn{1}{|c|}{19} & \multicolumn{1}{c|}{0.407} & \multicolumn{1}{c|}{0.407} & \multicolumn{1}{c|}{0.407} & \multicolumn{1}{c|}{0.407} & \multicolumn{1}{c|}{0.407} & \multicolumn{1}{c|}{0.187} \\ 
\hline
\multicolumn{1}{|c|}{20} & \multicolumn{1}{c|}{0.387} & \multicolumn{1}{c|}{0.387} & \multicolumn{1}{c|}{0.387} & \multicolumn{1}{c|}{0.387} & \multicolumn{1}{c|}{0.387} & \multicolumn{1}{c|}{0.164} \\ 
\hline
\multicolumn{1}{|c|}{21} & \multicolumn{1}{c|}{0.368} & \multicolumn{1}{c|}{0.368} & \multicolumn{1}{c|}{0.368} & \multicolumn{1}{c|}{0.368} & \multicolumn{1}{c|}{0.368} & \multicolumn{1}{c|}{0.144} \\ 
\hline
\multicolumn{1}{|c|}{22} & \multicolumn{1}{c|}{0.350} & \multicolumn{1}{c|}{0.350} & \multicolumn{1}{c|}{0.350} & \multicolumn{1}{c|}{0.350} & \multicolumn{1}{c|}{0.350} & \multicolumn{1}{c|}{0.126} \\ 
\hline
\multicolumn{1}{|c|}{23} & \multicolumn{1}{c|}{0.333} & \multicolumn{1}{c|}{0.333} & \multicolumn{1}{c|}{0.333} & \multicolumn{1}{c|}{0.333} & \multicolumn{1}{c|}{0.333} & \multicolumn{1}{c|}{0.111} \\ 
\hline
\multicolumn{1}{|c|}{24} & \multicolumn{1}{c|}{0.317} & \multicolumn{1}{c|}{0.317} & \multicolumn{1}{c|}{0.317} & \multicolumn{1}{c|}{0.317} & \multicolumn{1}{c|}{0.317} & \multicolumn{1}{c|}{0.097} \\ 
\hline
\multicolumn{1}{|c|}{25} & \multicolumn{1}{c|}{0.301} & \multicolumn{1}{c|}{0.301} & \multicolumn{1}{c|}{0.301} & \multicolumn{1}{c|}{0.301} & \multicolumn{1}{c|}{0.301} & \multicolumn{1}{c|}{0.085} \\ 
\hline
\multicolumn{1}{|c|}{26} & \multicolumn{1}{c|}{0.287} & \multicolumn{1}{c|}{0.287} & \multicolumn{1}{c|}{0.287} & \multicolumn{1}{c|}{0.287} & \multicolumn{1}{c|}{0.287} & \multicolumn{1}{c|}{0.074} \\ 
\hline
\multicolumn{1}{|c|}{27} & \multicolumn{1}{c|}{0.273} & \multicolumn{1}{c|}{0.273} & \multicolumn{1}{c|}{0.273} & \multicolumn{1}{c|}{0.273} & \multicolumn{1}{c|}{0.273} & \multicolumn{1}{c|}{0.064} \\ 
\hline
\multicolumn{1}{|c|}{28} & \multicolumn{1}{c|}{0.259} & \multicolumn{1}{c|}{0.259} & \multicolumn{1}{c|}{0.259} & \multicolumn{1}{c|}{0.259} & \multicolumn{1}{c|}{0.259} & \multicolumn{1}{c|}{0.056} \\ 
\hline
\multicolumn{1}{|c|}{29} & \multicolumn{1}{c|}{0.247} & \multicolumn{1}{c|}{0.247} & \multicolumn{1}{c|}{0.247} & \multicolumn{1}{c|}{0.247} & \multicolumn{1}{c|}{0.247} & \multicolumn{1}{c|}{0.049} \\ 
\hline
\multicolumn{1}{|c|}{30} & \multicolumn{1}{c|}{0.235} & \multicolumn{1}{c|}{0.235} & \multicolumn{1}{c|}{0.235} & \multicolumn{1}{c|}{0.235} & \multicolumn{1}{c|}{0.235} & \multicolumn{1}{c|}{0.043} \\ 
\hline
\end{tabular}
\label{tblavaimain:1}
\end{table}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=415pt]{avaimain-3.eps}
\caption{Reliability and availability}\label{avaimain:3}
% \end{center}
\end{figure}
\subsubsection{Answer D}
The function form for calculating the availability is:
\begin{eqnarray}
&& \begin{array}{l}
{A_{network}}(t) = {A_1}(t) \cdot {A_3}(t) \cdot {A_4}(t) + {A_1}(t) \cdot
{A_3}(t) \cdot {A_5}(t) + {A_2}(t) \cdot {A_3}(t) \cdot {A_4}(t) + {A_2}(t) \cdot
{A_3}(t) \cdot {A_5}(t)\\
+ {A_1}(t) \cdot {A_2}(t) \cdot {A_4}(t) \cdot {A_5}(t){\rm{ }} - {A_1}(t)
\cdot {A_2}(t) \cdot {A_3}(t) \cdot {A_4}(t)\\
- {A_1}(t) \cdot {A_3}(t) \cdot {A_4}(t) \cdot {A_5}(t) - {A_1}(t) \cdot
{A_2}(t) \cdot {A_3}(t) \cdot {A_5}(t) - {A_2}(t) \cdot {A_3}(t) \cdot {A_4}(t)
\cdot {A_5}(t)
\end{array}
\label{avaimaineq:6}
%(9)
\end{eqnarray}
where ${A_i}(t)$is the availability of sub-item \textit{i}.

The availability of each sub-item is calculated with equation \eqref{avaimaineq:1}, in which, it
is necessary to calculate the mean time to repair or the mean time (duration) the
sub-item is in adequate level of service. This calculation involves the value of
reliability for each sub-item calculated in previous step.

\begin{eqnarray}
E(t_i^n) = {\Delta _i} \cdot {F_i}(t) = {\Delta _i} \cdot \left[ {1 - {R_i}(t)}
\right]
\label{avaimaineq:7}
%(10)
\end{eqnarray}

In equation \eqref{avaimaineq:7}, ${F_i}(t)$is the failure probability at time \textit{t}.

The availability of each sub-item in year \textit{t} will be
\begin{eqnarray}
A(t_i^{}) = \frac{{E\left[ {t_i^a} \right]}}{{E\left[ {t_i^a} \right] + E\left[
{t_i^n} \right]}} = \frac{{\int\limits_0^{tu} {{R_i}(t)dt} }}{{\int\limits_0^{tu}
{{R_i}(t)dt}  + {\Delta _i} \cdot \left[ {1 - {R_i}(tu)} \right]}}
\label{avaimaineq:8}
%(11)
\end{eqnarray}
The availability of each item and the network in 30 years.

Using equations \eqref{eqreliability:7} the availability of each sub-item, and the item, for periods
of time up to 30 years can be calculated (Table \ref{tblavaimain:2}). For example, the
availability of sub-item 1 for a 14 year time period is 0.999090421, and the
availability of the item for a 14 year time period is 0.999947505. This means
that on average over periods of time that are 14 years in length that one can
expect that sub-item 1 would provide an adequate LOS for 13.9872 years and an
inadequate LOS for 0.01273411 years (4.647949 days), and that the item would
provide an adequate LOS for 13.99927 years and an inadequate LOS for 0.00073
years (0.26645 days). The longer the period of time investigated the lower the
availability because the reliability of the sub-items decreases with age, when
there is a constant failure rate. This is illustrated in Figure \ref{avaimain:3}.

\begin{table}[h]
\caption{Availability}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Time} & \multicolumn{6}{c|}{Availability} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{(years)} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{Item} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{0.999509} & \multicolumn{1}{c|}{0.999018} & \multicolumn{1}{c|}{0.997549} & \multicolumn{1}{c|}{0.998038} & \multicolumn{1}{c|}{0.998528} & \multicolumn{1}{c|}{0.999985} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{0.999345} & \multicolumn{1}{c|}{0.998691} & \multicolumn{1}{c|}{0.996735} & \multicolumn{1}{c|}{0.997386} & \multicolumn{1}{c|}{0.998039} & \multicolumn{1}{c|}{0.999973} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{0.999264} & \multicolumn{1}{c|}{0.998528} & \multicolumn{1}{c|}{0.996329} & \multicolumn{1}{c|}{0.997061} & \multicolumn{1}{c|}{0.997794} & \multicolumn{1}{c|}{0.999966} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{0.999215} & \multicolumn{1}{c|}{0.998431} & \multicolumn{1}{c|}{0.996086} & \multicolumn{1}{c|}{0.996867} & \multicolumn{1}{c|}{0.997648} & \multicolumn{1}{c|}{0.999961} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{0.999182} & \multicolumn{1}{c|}{0.998366} & \multicolumn{1}{c|}{0.995925} & \multicolumn{1}{c|}{0.996737} & \multicolumn{1}{c|}{0.997551} & \multicolumn{1}{c|}{0.999958} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{0.999159} & \multicolumn{1}{c|}{0.99832} & \multicolumn{1}{c|}{0.995810} & \multicolumn{1}{c|}{0.996645} & \multicolumn{1}{c|}{0.997482} & \multicolumn{1}{c|}{0.999955} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{0.999142} & \multicolumn{1}{c|}{0.998285} & \multicolumn{1}{c|}{0.995724} & \multicolumn{1}{c|}{0.996576} & \multicolumn{1}{c|}{0.99743} & \multicolumn{1}{c|}{0.999953} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{0.999128} & \multicolumn{1}{c|}{0.998258} & \multicolumn{1}{c|}{0.995657} & \multicolumn{1}{c|}{0.996523} & \multicolumn{1}{c|}{0.99739} & \multicolumn{1}{c|}{0.999952} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{0.999118} & \multicolumn{1}{c|}{0.998237} & \multicolumn{1}{c|}{0.995604} & \multicolumn{1}{c|}{0.99648} & \multicolumn{1}{c|}{0.997358} & \multicolumn{1}{c|}{0.999951} \\ 
\hline
\multicolumn{1}{|c|}{11} & \multicolumn{1}{c|}{0.999109} & \multicolumn{1}{c|}{0.998219} & \multicolumn{1}{c|}{0.995561} & \multicolumn{1}{c|}{0.996445} & \multicolumn{1}{c|}{0.997332} & \multicolumn{1}{c|}{0.99995} \\ 
\hline
\multicolumn{1}{|c|}{12} & \multicolumn{1}{c|}{0.999102} & \multicolumn{1}{c|}{0.998205} & \multicolumn{1}{c|}{0.995525} & \multicolumn{1}{c|}{0.996416} & \multicolumn{1}{c|}{0.99731} & \multicolumn{1}{c|}{0.999949} \\ 
\hline
\multicolumn{1}{|c|}{13} & \multicolumn{1}{c|}{0.999096} & \multicolumn{1}{c|}{0.998193} & \multicolumn{1}{c|}{0.995494} & \multicolumn{1}{c|}{0.996392} & \multicolumn{1}{c|}{0.997292} & \multicolumn{1}{c|}{0.999948} \\ 
\hline
\multicolumn{1}{|c|}{14} & \multicolumn{1}{c|}{0.99909} & \multicolumn{1}{c|}{0.998182} & \multicolumn{1}{c|}{0.995469} & \multicolumn{1}{c|}{0.996372} & \multicolumn{1}{c|}{0.997276} & \multicolumn{1}{c|}{0.999948} \\ 
\hline
\multicolumn{1}{|c|}{15} & \multicolumn{1}{c|}{0.999086} & \multicolumn{1}{c|}{0.998174} & \multicolumn{1}{c|}{0.995446} & \multicolumn{1}{c|}{0.996354} & \multicolumn{1}{c|}{0.997263} & \multicolumn{1}{c|}{0.999947} \\ 
\hline
\multicolumn{1}{|c|}{16} & \multicolumn{1}{c|}{0.999082} & \multicolumn{1}{c|}{0.998166} & \multicolumn{1}{c|}{0.995427} & \multicolumn{1}{c|}{0.996338} & \multicolumn{1}{c|}{0.997251} & \multicolumn{1}{c|}{0.999947} \\ 
\hline
\multicolumn{1}{|c|}{17} & \multicolumn{1}{c|}{0.999079} & \multicolumn{1}{c|}{0.998159} & \multicolumn{1}{c|}{0.995410} & \multicolumn{1}{c|}{0.996325} & \multicolumn{1}{c|}{0.997241} & \multicolumn{1}{c|}{0.999946} \\ 
\hline
\multicolumn{1}{|c|}{18} & \multicolumn{1}{c|}{0.999076} & \multicolumn{1}{c|}{0.998153} & \multicolumn{1}{c|}{0.995395} & \multicolumn{1}{c|}{0.996313} & \multicolumn{1}{c|}{0.997232} & \multicolumn{1}{c|}{0.999946} \\ 
\hline
\multicolumn{1}{|c|}{19} & \multicolumn{1}{c|}{0.999073} & \multicolumn{1}{c|}{0.998148} & \multicolumn{1}{c|}{0.995382} & \multicolumn{1}{c|}{0.996302} & \multicolumn{1}{c|}{0.997224} & \multicolumn{1}{c|}{0.999945} \\ 
\hline
\multicolumn{1}{|c|}{20} & \multicolumn{1}{c|}{0.999071} & \multicolumn{1}{c|}{0.998143} & \multicolumn{1}{c|}{0.99537} & \multicolumn{1}{c|}{0.996293} & \multicolumn{1}{c|}{0.997217} & \multicolumn{1}{c|}{0.999945} \\ 
\hline
\multicolumn{1}{|c|}{21} & \multicolumn{1}{c|}{0.999068} & \multicolumn{1}{c|}{0.998139} & \multicolumn{1}{c|}{0.995359} & \multicolumn{1}{c|}{0.996284} & \multicolumn{1}{c|}{0.99721} & \multicolumn{1}{c|}{0.999945} \\ 
\hline
\multicolumn{1}{|c|}{22} & \multicolumn{1}{c|}{0.999066} & \multicolumn{1}{c|}{0.998135} & \multicolumn{1}{c|}{0.99535} & \multicolumn{1}{c|}{0.996276} & \multicolumn{1}{c|}{0.997205} & \multicolumn{1}{c|}{0.999945} \\ 
\hline
\multicolumn{1}{|c|}{23} & \multicolumn{1}{c|}{0.999065} & \multicolumn{1}{c|}{0.998131} & \multicolumn{1}{c|}{0.995341} & \multicolumn{1}{c|}{0.996269} & \multicolumn{1}{c|}{0.997199} & \multicolumn{1}{c|}{0.999945} \\ 
\hline
\multicolumn{1}{|c|}{24} & \multicolumn{1}{c|}{0.999063} & \multicolumn{1}{c|}{0.998128} & \multicolumn{1}{c|}{0.995333} & \multicolumn{1}{c|}{0.996263} & \multicolumn{1}{c|}{0.997195} & \multicolumn{1}{c|}{0.999944} \\ 
\hline
\multicolumn{1}{|c|}{25} & \multicolumn{1}{c|}{0.999062} & \multicolumn{1}{c|}{0.998125} & \multicolumn{1}{c|}{0.995326} & \multicolumn{1}{c|}{0.996257} & \multicolumn{1}{c|}{0.99719} & \multicolumn{1}{c|}{0.999944} \\ 
\hline
\multicolumn{1}{|c|}{26} & \multicolumn{1}{c|}{0.99906} & \multicolumn{1}{c|}{0.998123} & \multicolumn{1}{c|}{0.99532} & \multicolumn{1}{c|}{0.996252} & \multicolumn{1}{c|}{0.997186} & \multicolumn{1}{c|}{0.999944} \\ 
\hline
\multicolumn{1}{|c|}{27} & \multicolumn{1}{c|}{0.999059} & \multicolumn{1}{c|}{0.99812} & \multicolumn{1}{c|}{0.995314} & \multicolumn{1}{c|}{0.996247} & \multicolumn{1}{c|}{0.997183} & \multicolumn{1}{c|}{0.999944} \\ 
\hline
\multicolumn{1}{|c|}{28} & \multicolumn{1}{c|}{0.999058} & \multicolumn{1}{c|}{0.998118} & \multicolumn{1}{c|}{0.995308} & \multicolumn{1}{c|}{0.996243} & \multicolumn{1}{c|}{0.99718} & \multicolumn{1}{c|}{0.999944} \\ 
\hline
\multicolumn{1}{|c|}{29} & \multicolumn{1}{c|}{0.999057} & \multicolumn{1}{c|}{0.998116} & \multicolumn{1}{c|}{0.995303} & \multicolumn{1}{c|}{0.996239} & \multicolumn{1}{c|}{0.997176} & \multicolumn{1}{c|}{0.999944} \\ 
\hline
\multicolumn{1}{|c|}{30} & \multicolumn{1}{c|}{0.999056} & \multicolumn{1}{c|}{0.998114} & \multicolumn{1}{c|}{0.995298} & \multicolumn{1}{c|}{0.996235} & \multicolumn{1}{c|}{0.997174} & \multicolumn{1}{c|}{0.999943} \\ 
\hline
\end{tabular}
\label{tblavaimain:2}
\end{table}

The code to compute this example is given in the Appendix \ref{appavai-mai1}
\section{Trade-offs}
Indicators of performance are just that, indicators. One can minimize the amount
of time required to execute interventions on an item, i.e. maximize
maintainability, by constructing an expensive but easy to repair item. This would
most likely increase the availability of the item but may have little effect on
the reliability of the item.

One can maximize the reliability of an item by constructing an expensive but
difficult to repair item. This would most likely decrease the maintainability of
the item but may have little effect on the availability of the item.

Once can maximize the availability of an item by constructing an expensive item
that is fairly reliable and fairly easy to repair.

In each of these cases, however, the most important question for the
infrastructure manager is whether the improvements in reliability, availability
or maintainability are worth the money spent. For example, if a road has very
little traffic it may not be worthwhile to spend a large amount of money to
shorten the durations of interventions, to increase the reliability of the bridge
in the road, or to improve the availability of the road.
\subsection{Example}
You are the manager of a tunnel that is considering the construction of tunnel
with five possible configurations. Each configuration is comprised of five
components, but three of the components can be one of two different types. The
configurations to be considered, as well as the mean time between failures for
each component in each configuration, the initial cost of each configuration, and
the mean intervention duration are given in Table \ref{tblavaimain:3}.
\begin{table}[h]
\caption{Tunnel components}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|c|}{Components} & \multicolumn{1}{c|}{Option 1} & \multicolumn{1}{c|}{Option 2} & \multicolumn{1}{c|}{Option 3} & \multicolumn{1}{c|}{Option 4} & \multicolumn{1}{c|}{Option 5} \\ 
\hline
\multicolumn{1}{|c|}{No.} & Description & \multicolumn{5}{c|}{mean time between failures (years)} \\ 
\hline
\multicolumn{1}{|c|}{1} & a power supply system & \multicolumn{1}{c|}{245} & \multicolumn{1}{c|}{9'612} & \multicolumn{1}{c|}{245} & \multicolumn{1}{c|}{9'612} & \multicolumn{1}{c|}{9'612} \\ 
\hline
\multicolumn{1}{|c|}{2} & a ventilation system & \multicolumn{1}{c|}{4'819} & \multicolumn{1}{c|}{4'819} & \multicolumn{1}{c|}{36'955} & \multicolumn{1}{c|}{4'819} & \multicolumn{1}{c|}{36'955} \\ 
\hline
\multicolumn{1}{|c|}{3} & a fire services system & \multicolumn{1}{c|}{1.1x106} & \multicolumn{1}{c|}{1.1x106} & \multicolumn{1}{c|}{1.1x106} & \multicolumn{1}{c|}{1.1x106} & \multicolumn{1}{c|}{1.1x106} \\ 
\hline
\multicolumn{1}{|c|}{4} & a drainage system & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2'194} & \multicolumn{1}{c|}{2'194} \\ 
\hline
\multicolumn{1}{|c|}{5} & a central monitoring and control system & \multicolumn{1}{c|}{2.4x105} & \multicolumn{1}{c|}{2.4x105} & \multicolumn{1}{c|}{2.4x105} & \multicolumn{1}{c|}{2.4x105} & \multicolumn{1}{c|}{2.4x105} \\ 
\hline
\multicolumn{1}{|l}{} & Initial cost (mus) & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{12} \\ 
\hline
\multicolumn{1}{|l}{} & Mean intervention duration (tu) & \multicolumn{1}{c|}{0.25} & \multicolumn{1}{c|}{0.28} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{2} \\ 
\hline
\end{tabular}
\label{tblavaimain:3}
\end{table}
\subsubsection{Question}
Which configuration has the highest reliability? The highest maintainability?
The highest availability? And which configuration results in the lowest overall
net costs over a 20 \textit{tu} time period?
\subsubsection{Answer}
\paragraph{Reliability}
In the estimation of the reliability of the tunnel item one first has to
recognize that the 5 sub-items are conceptually connected in series, i.e. if one
sub-item fails the item fails. The next step is to calculate the failure rate of
each component, which if it is assumed that the failures of each component
follows can be modelled using an exponential distribution, is 1/the mean time
between failures. These are shown in Table \ref{tblavaimain:4}.
${\theta _1}$ ${\theta _2}$ ${\theta _3}$ ${\theta _4}$ ${\theta _5}$
The mean time between failures for the tunnel is then 1 over the sum of the
failure rates as shown in equation \eqref{avaimaineq:9}:
\begin{eqnarray}
&& \frac{1}{{\sum\limits_i^I {{\theta _i}} }}
\label{avaimaineq:9}
%(12)
\end{eqnarray}
Where i is the index of tunnel components.

And the reliability of the tunnel in one year is given by:
\begin{eqnarray}
&& R = \exp \left[ { - \left( {\sum\limits_i^I {{\theta _i}} } \right) \cdot 1}
\right]
\label{avaimaineq:10}
%(13)
\end{eqnarray}
These values for each of the options are shown in Table \ref{tblavaimain:4}. It can be
seen that option 5 is the most reliable, although not significantly more reliable
than option 4.
\begin{table}[h]
\caption{Failure rates of tunnel components, and mean time between failure and reliability of the tunnel}
\begin{tabular}{|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Option} & \multicolumn{1}{c|}{Item} & \multicolumn{5}{c|}{Components} & \multicolumn{1}{m{1.2cm}|}{MTBF of Tunnel} & \multicolumn{1}{c|}{Reliability} \\ 
\cline{3-7}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{m{1.2cm}|}{Power Supply System} & \multicolumn{1}{m{1.2cm}|}{Ventilation System} & \multicolumn{1}{m{1.2cm}|}{Fire Services System} & \multicolumn{1}{m{1.2cm}|}{Drainage System} & \multicolumn{1}{m{1.5cm}|}{Central monitoring and control system} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{MTBF (years)} & \multicolumn{1}{c|}{245} & \multicolumn{1}{c|}{4'819} & \multicolumn{1}{c|}{1.10E+06} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2.40E+05} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0.3663} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{$\theta_1$} & \multicolumn{1}{c|}{4.08E-03} & \multicolumn{1}{c|}{2.08E-04} & \multicolumn{1}{c|}{9.09E-07} & \multicolumn{1}{c|}{1.00E+00} & \multicolumn{1}{c|}{4.17E-06} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{MTBF (years)} & \multicolumn{1}{c|}{9'612} & \multicolumn{1}{c|}{4'819} & \multicolumn{1}{c|}{1.10E+06} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2.40E+05} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0.3678} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{$\theta_2$} & \multicolumn{1}{c|}{1.04E-04} & \multicolumn{1}{c|}{2.08E-04} & \multicolumn{1}{c|}{9.09E-07} & \multicolumn{1}{c|}{1.00E+00} & \multicolumn{1}{c|}{4.17E-06} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{MTBF (years)} & \multicolumn{1}{c|}{245} & \multicolumn{1}{c|}{36'955} & \multicolumn{1}{c|}{1.10E+06} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2.40E+05} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0.3664} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{$\theta_3$} & \multicolumn{1}{c|}{4.08E-03} & \multicolumn{1}{c|}{2.71E-05} & \multicolumn{1}{c|}{9.09E-07} & \multicolumn{1}{c|}{1.00E+00} & \multicolumn{1}{c|}{4.17E-06} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{MTBF (years)} & \multicolumn{1}{c|}{9'612} & \multicolumn{1}{c|}{4'819} & \multicolumn{1}{c|}{1.10E+06} & \multicolumn{1}{c|}{2'194} & \multicolumn{1}{c|}{2.40E+05} & \multicolumn{1}{c|}{1'295} & \multicolumn{1}{c|}{0.9992} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{$\theta_4$} & \multicolumn{1}{c|}{1.04E-04} & \multicolumn{1}{c|}{2.08E-04} & \multicolumn{1}{c|}{9.09E-07} & \multicolumn{1}{c|}{4.56E-04} & \multicolumn{1}{c|}{4.17E-06} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{MTBF (years)} & \multicolumn{1}{c|}{9'612} & \multicolumn{1}{c|}{36'955} & \multicolumn{1}{c|}{1.10E+06} & \multicolumn{1}{c|}{2'194} & \multicolumn{1}{c|}{2.40E+05} & \multicolumn{1}{c|}{1'689} & \multicolumn{1}{c|}{0.9994} \\ 
\cline{2-7}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{$\theta_5$} & \multicolumn{1}{c|}{1.04E-004} & \multicolumn{1}{c|}{2.71E-005} & \multicolumn{1}{c|}{9.09E-007} & \multicolumn{1}{c|}{4.56E-004} & \multicolumn{1}{c|}{4.17E-006} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\end{tabular}
\label{tblavaimain:4}
\end{table}
\paragraph{Maintainability}
In order to calculate the maintainability we can look simply at the average
length of time it will take to restore each configuration when it fails. It is
here clear that the configuration 1, which has an average down time of 0.25
months if it fails is the most maintainable. This can be seen in Table
\ref{tblavaimain:5}.
\begin{table}[h]
\caption{Initial cost vs reliability, maintainability, availability and net benefit of the five
configurations}
\begin{tabular}{|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Configuration} & \multicolumn{1}{c|}{Initial Cost} & \multicolumn{1}{c|}{Reliability} & \multicolumn{1}{c|}{Maintainability} & \multicolumn{1}{c|}{Availability} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{(mus)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{(months)} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.3663} & \multicolumn{1}{c|}{\cellcolor{blue!25} 0.25} & \multicolumn{1}{c|}{0.979506} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{0.3678} & \multicolumn{1}{c|}{0.28} & \multicolumn{1}{c|}{0.97759} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{0.3664} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.975512} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{0.9992} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{\cellcolor{blue!25}0.999903} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{\cellcolor{blue!25} 0.9994} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{0.999901} \\ 
\hline
\end{tabular}
\label{tblavaimain:5}
\end{table}
\paragraph{Availability}
The availability of each configuration is given by taking the mean time between
failures, i.e. the mean time the item works, and dividing it by the average time
it takes to restore the configuration if it fails plus the mean time it works,
i.e. the average amount of time from starting to providing an adequate LOS to
providing an adequate LOS again. One can also refer to this as the renewal
period. The configuration with the highest availability is configuration 4 (Table
\ref{tblavaimain:5}).

\paragraph{Costs}
The net costs are the initial costs plus the costs incurred over the 20 year
period due to the costs due to lost service when an inadequate LOS is provided,
which in this case are considered to include the cost of interventions when
executed. Assuming that these costs are directly and uniformly related to the
amount of time required to restore service, the annual maintenance costs for each
configuration are given by:
\begin{eqnarray}
&& {C_{am}} = \frac{{E\left[ {t_i^n} \right] \cdot {I_{cr}}/12}}{{E\left[ {t_i^a}
\right] + E\left[ {t_i^n} \right]}}
%(13)
\end{eqnarray}
Where \textit{I$_{cr}$} represents the cost of executing an intervention
multiplied by the probability of executing an intervention

The annual maintenance costs as well as the net benefit excluding upfront costs
and including upfront costs are shown for each configuration in Table
\ref{tblavaimain:6}. The net benefit of a configuration is the cost of configuration 1
minus the costs of the configuration being investigated. It can be seen that
configuration has the highest net benefit over the 20 \textit{tu} period.

\begin{eqnarray}
&& N{B^i} = NB_c^i + NB_m^i\,\,\,\,\,\,\,\,\,\, = \left( {C_c^r - C_c^i} \right) +
\left( {C_m^r - C_m^i} \right)
\label{avaimaineq:11}
%(15)
\end{eqnarray}
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$N{B^i}$:] is the net benefit of configuration i,
\item[$C_m^r$:] is the maintenance costs of the reference configuration
\item[$C_m^i$:] is the maintenance costs of configuration i
\end{description}
%\end{flushright}
\end{adjustwidth}

\begin{table}[h]
\caption{ Initial cost versus total costs of the five configurations}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Configuration} & \multicolumn{5}{c|}{Cost} & \multicolumn{1}{c|}{Net benefit} & \multicolumn{1}{c|}{Net benefit} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{initial} & \multicolumn{1}{c|}{ Cost} & \multicolumn{1}{c|}{Cost} & \multicolumn{1}{c|}{Cost per tu} & \multicolumn{1}{c|}{Cost for 20 tus} & \multicolumn{1}{c|}{excluding} & \multicolumn{1}{c|}{including} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{(mus)} & \multicolumn{1}{c|}{mus/month} & \multicolumn{1}{c|}{(mus)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{upfront costs} & \multicolumn{1}{c|}{upfront costs} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{$C_c$} & \multicolumn{1}{c|}{$I_{cr}$} & \multicolumn{1}{c|}{$E[t_i^n]\cdot I_{cr}$} & \multicolumn{1}{c|}{$C_{am}$} & \multicolumn{1}{c|}{$C_{am}\cdot 20$} & \multicolumn{1}{c|}{$NB_m^i$} & \multicolumn{1}{c|}{$NB^i$} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1.25} & \multicolumn{1}{c|}{$7.79x10^{-1}$} & \multicolumn{1}{c|}{15.5844} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1.38} & \multicolumn{1}{c|}{$8.50x10^{-1}$} & \multicolumn{1}{c|}{17.0023} & \multicolumn{1}{c|}{-1.418} & \multicolumn{1}{c|}{-3.418} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{$9.31x10^{-1}$} & \multicolumn{1}{c|}{18.6197} & \multicolumn{1}{c|}{-3.035} & \multicolumn{1}{c|}{-7.035} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{7.5} & \multicolumn{1}{c|}{$4.47x10^{-6}$} & \multicolumn{1}{c|}{0.0001} & \multicolumn{1}{c|}{15.584} & \multicolumn{1}{c|}{9.584} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{$3.50x10^{-6}$} & \multicolumn{1}{c|}{0.0001} & \multicolumn{1}{c|}{15.584} & \multicolumn{1}{c|}{3.584} \\ 
\hline
\end{tabular}
\label{tblavaimain:6}
\end{table}

\paragraph{Summary}
As can be seen in Table \ref{tblavaimain:7} that estimating the availability was the
only performance indicator that allowed the optimal configuration to be
determined. This will, however, not always be the case. This should be kept in
mind when selecting performance indicators to evaluate performance.
\begin{table}[h]
\caption{Initial cost vs reliability, maintainability, availability and net benefit of the five
configurations}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Configuration} & \multicolumn{1}{c|}{Initial Cost} & \multicolumn{1}{c|}{Reliability} & \multicolumn{1}{c|}{Maintainability} & \multicolumn{1}{c|}{Availability} & \multicolumn{1}{c|}{Net benefit} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{(mus)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{(months)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{including upfront costs} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.3663} & \multicolumn{1}{c|}{\cellcolor{blue!25} 0.25} & \multicolumn{1}{c|}{0.979506} & \multicolumn{1}{c|}{0.000} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{0.3678} & \multicolumn{1}{c|}{0.28} & \multicolumn{1}{c|}{0.97759} & \multicolumn{1}{c|}{-3.418} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{0.3664} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.975512} & \multicolumn{1}{c|}{-7.035} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{0.9992} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{\cellcolor{blue!25} 0.999903} & \multicolumn{1}{c|}{\cellcolor{blue!25} 9.584} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{\cellcolor{blue!25} 0.9994} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{0.999901} & \multicolumn{1}{c|}{3.584} \\ 
\hline
\end{tabular}
\label{tblavaimain:7}
\end{table}
The R code for this example is given in Appendix \ref{appavai-mai2}

\section{Assignments}
\subsection{Problem A}
An infrastructure manager is responsible for maintaining a number of
facilities over a period of time. The length of time required for past corrective
interventions are given in Table \ref{tblavaimain:8}

\begin{table}[h]
\caption{ Corrective maintenance task times}
\begin{tabular}{|c|c|}
\hline
Task time & Frequency \\ 
\hline
11 & 2 \\ 
\hline
13 & 3 \\ 
\hline
15 & 8 \\ 
\hline
17 & 12 \\ 
\hline
19 & 12 \\ 
\hline
21 & 14 \\ 
\hline
23 & 13 \\ 
\hline
25 & 10 \\ 
\hline
27 & 10 \\ 
\hline
29 & 8 \\ 
\hline
31 & 7 \\ 
\hline
33 & 6 \\ 
\hline
35 & 5 \\ 
\hline
36 & 5 \\ 
\hline
37 & 4 \\ 
\hline
39 & 3 \\ 
\hline
41 & 2 \\ 
\hline
47 & 2 \\ 
\hline
\end{tabular}
\label{tblavaimain:8}
\end{table}
\subsubsection{Question A1}
What is the range of observations?
\subsubsection{Answer A1}
The range of observations is the difference between the smallest and the
largest value in the data set. In this respect, the range of the above data set
is
\begin{eqnarray}
&& Range = \max (data) - \min (data)
\label{eqavaimain:1}
%(1)
\end{eqnarray}
\[
Range = 14 - 2 = 12
\]
\subsubsection{Question A2s}
Using a class interval of 4, determine the number of class intervals.
Plot the data and construct a curve. What is the most likely type of distribution
is indicated by the curve?
\subsubsection{Answer A2}
With a class interval of 4, one can construct a range of observations.
This is equivalent to the class interval of 4.

\begin{table}[h]
\caption{Range of observation}
\begin{tabular}{|l|l|}
\hline
\multicolumn{1}{|c|}{Range} & \multicolumn{1}{c|}{Frequency} \\ 
\hline
\multicolumn{1}{|c|}{9-12} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{13-16} & \multicolumn{1}{c|}{11} \\ 
\hline
\multicolumn{1}{|c|}{17-20} & \multicolumn{1}{c|}{24} \\ 
\hline
\multicolumn{1}{|c|}{21-24} & \multicolumn{1}{c|}{27} \\ 
\hline
\multicolumn{1}{|c|}{25-28} & \multicolumn{1}{c|}{20} \\ 
\hline
\multicolumn{1}{|c|}{29-32} & \multicolumn{1}{c|}{15} \\ 
\hline
\multicolumn{1}{|c|}{33-36} & \multicolumn{1}{c|}{16} \\ 
\hline
\multicolumn{1}{|c|}{37-40} & \multicolumn{1}{c|}{7} \\ 
\hline
\multicolumn{1}{|c|}{41-44} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{45-48} & \multicolumn{1}{c|}{2} \\ 
\hline
\end{tabular}
\label{tblavaimain:9}
\end{table}

From this table, a histogram of frequency can be constructed (Figure
\ref{figavaimain-a:1})

\begin{figure}[h]
% \begin{center}
\includegraphics[width=376pt]{avaimain-a-9.eps}
\caption{Histogram of data}\label{figavaimain-a:1}
% \end{center}
\end{figure}

By observing the shape of the histogram, one can conclude that the data
is distributed close to the density of the lognormal distribution function, which
has a skew tail on the right side of the density function.

The lognormal distribution function has following properties

Probability density function
\begin{eqnarray}
&& {f_X}(x,\mu ,\sigma ) = \frac{1}{{x \cdot \sigma \sqrt {2\pi } }}{e^{ -
\frac{{{{(\ln x - \mu )}^2}}}{{2{\sigma ^2}}}}}{\rm{,   }}x > 0
\label{eqavaimain:2}
%(2)
\end{eqnarray}
Cumulative distribution function
\begin{eqnarray}
&& {F_X}(x,\mu ,\sigma ) = \frac{1}{2} + \frac{1}{2}erf\left[ {\frac{{(\ln x - \mu
)}}{{\sqrt {2\sigma } }}} \right]
\label{eqavaimain:3}
%(3)
\end{eqnarray}
where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$x$:] is the random variable of the distribution,
\item[$\mu$:] is the mean of the variable's natural logarithm,
\item[$\sigma $:] is the standard deviation of the variable's natural logarithm,
\item[$erf $:] is the error function.
\end{description}
%\end{flushright}
\end{adjustwidth}
\subsubsection{Question A3}
What is the mean length of time for a corrective intervention,
\textit{M$_{ct}$}?
\subsubsection{Answer A3}
For estimating the mean corrective intervention duration
\begin{eqnarray}
&& {\overline M _{ct}} = \frac{{\sum {\left( {{f_{c{t_i}}}} \right)\left(
{{M_{c{t_i}}}} \right)} }}{{\sum {{f_{c{t_i}}}} }}
\label{eqavaimain:4}
%(4)
\end{eqnarray}
where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$f_{cti}$:] is the frequency of the \textit{i$^{th}$} corrective intervention in interventions per system operating time unit,
\item[$M_{cti}$:] is the elapsed time required for the \textit{i$^{th}$} corrective intervention.
\end{description}
%\end{flushright}
\end{adjustwidth}
The answer is 25.15 days
\begin{table}[h]
\caption{Mean time of intervention duration}
\begin{tabular}{|c|c|c|c|}
\hline
Time & Frequency & Mean time & Variance \\ 
\hline
(1) & (2) & (3)=(1)*(2) & $(4)=(2)*[(1)- \bar M_{ct}]^²$ \\ 
\hline
11 & 2 & 22 & 400.49 \\ 
\hline
13 & 3 & 39 & 442.93 \\ 
\hline
15 & 8 & 120 & 824.31 \\ 
\hline
17 & 12 & 204 & 797.23 \\ 
\hline
19 & 12 & 228 & 453.99 \\ 
\hline
21 & 14 & 294 & 241.21 \\ 
\hline
23 & 13 & 299 & 60.14 \\ 
\hline
25 & 10 & 250 & 0.23 \\ 
\hline
27 & 10 & 270 & 34.2 \\ 
\hline
29 & 8 & 232 & 118.53 \\ 
\hline
31 & 7 & 217 & 239.49 \\ 
\hline
33 & 6 & 198 & 369.66 \\ 
\hline
35 & 5 & 175 & 485.03 \\ 
\hline
36 & 5 & 180 & 588.53 \\ 
\hline
37 & 4 & 148 & 561.61 \\ 
\hline
39 & 3 & 117 & 575.4 \\ 
\hline
41 & 2 & 82 & 502.39 \\ 
\hline
47 & 2 & 94 & 954.78 \\ 
\hline
\textbf{Total} & \textbf{126} & \textbf{3'169} & \textbf{7'650.13} \\ 
\hline
\end{tabular}
\label{tblavaimain:10}
\end{table}
  \[
   {\overline M _{ct}} = \frac{{3'169}}{{126}} = 25.15
  \]
\subsubsection{Question A4}
What is the geometric mean of the repair times?
\subsubsection{Answer A4}
For calculating geometric mean, refer to general statistical materials
\begin{eqnarray}
&& {\bar \mu _{geo}} = {e^\mu }
\label{eqavaimai:5}
%(4)
\end{eqnarray}
Where $\mu{}$ is the mean that is calculated based on the mean ${\overline M _{ct}}$ in this case)
\begin{eqnarray}
&& \mu  = \ln (E\left[ X \right]) - \frac{1}{2}{\sigma ^2} \label{eqavaimai:6}
\end{eqnarray}
\begin{eqnarray}
&& {\sigma ^2} = \ln \left( {1 + \frac{{Var\left[ X \right]}}{{{{(E\left[ X \right])}^2}}}} \right)
\end{eqnarray}
from data, we have
\[
Var\left[ X \right] = \sqrt {\frac{{7'650.13}}{{126}}}  = 7.79
\]
\[
E\left[ X \right] = {\bar M_{ct}} = 25.15
\]
\[
\mu  = \ln (25.15) - \frac{1}{2}\left[ {\ln \left( {1 +
\frac{{{{7.79}^2}}}{{{{25.15}^2}}}} \right)} \right] = 3.179
\]
The geometric mean is then
\[
{\bar \mu _{geo}} = {e^{3.179}} = 24.02
\]
\subsubsection{Question A5}
What is the standard deviation of the sample data?
\subsubsection{Answer A5}
The standard deviation is 7.79, which has been calculated in previous
steps.
\subsubsection{Question A6}
Assume 90\% confidence level, what is the \textit{M$_{max}$} value?
\subsubsection{Answer A6}
For obtaining the the \textit{M$_{max}$} value with assumption of a 90\%
confidence level, refer to the formulation of the cumulative density function in
equation \eqref{eqavaimain:3} of the log-normal distribution and interpolate the data.

The Mmax is 27.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=396pt]{avaimain-a-10.eps}
\caption{Cumulative probability}
% \end{center}
\end{figure}
R code is given in Appendix \ref{appavai-mai3} for this assignment

\subsection{Problem B}
A transportation network connecting city A to city B is shown in Figure
\ref{figavaimain-a:2}. In order to go from city A to city B and vice versa, vehicles can
travel in both directions of every link. Links are either a road section, a
bridge or a tunnels. It is periodically affected by natural hazards such as
flood, avalanches, and rockfall. The probability of the links not providing an
adequate LOS due to the natural hazards can be modelled using the exponential or
the Weibull distributions as shown in Table \ref{tblavaimain:11}. The mean time required
to restore an adequate LOS following the occurrence of a natural hazard on each
of the links is also is included in Table \ref{tblavaimain:11}.

\begin{figure}[h]
\begin{center}
\includegraphics[width=392pt]{avaimain-a-1.eps}
\caption{A transportation network connecting city A to city B}\label{figavaimain-a:2}
\end{center}
\end{figure}

\begin{table}[h]
\caption{Characteristics of links in the network}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Link object} & \multicolumn{1}{c|}{Deterioration distribution} & \multicolumn{2}{c|}{Weibul} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{Mean time of intervention} \\ 
\cline{3-4}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$\alpha$} & \multicolumn{1}{c|}{$m$} & \multicolumn{1}{c|}{$\theta$} & \multicolumn{1}{c|}{(days)} \\ 
\hline
\multicolumn{1}{|c|}{Road A-1} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.024} & \multicolumn{1}{c|}{20} \\ 
\hline
\multicolumn{1}{|c|}{Road 2-3} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.023} & \multicolumn{1}{c|}{15} \\ 
\hline
\multicolumn{1}{|c|}{Road 3-4} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.03} & \multicolumn{1}{c|}{17} \\ 
\hline
\multicolumn{1}{|c|}{Road 3-5} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.06} & \multicolumn{1}{c|}{33} \\ 
\hline
\multicolumn{1}{|c|}{Road 4-5} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.04} & \multicolumn{1}{c|}{40} \\ 
\hline
\multicolumn{1}{|c|}{Bridge 1-2} & \multicolumn{1}{c|}{Weibull} & \multicolumn{1}{c|}{0.025} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{30} \\ 
\hline
\multicolumn{1}{|c|}{Bridge 1-3} & \multicolumn{1}{c|}{Weibull} & \multicolumn{1}{c|}{0.03} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{50} \\ 
\hline
\multicolumn{1}{|c|}{Bridge A-4} & \multicolumn{1}{c|}{Weibull} & \multicolumn{1}{c|}{0.025} & \multicolumn{1}{c|}{2.3} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{70} \\ 
\hline
\multicolumn{1}{|c|}{Tunnel B-2} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.04} & \multicolumn{1}{c|}{60} \\ 
\hline
\multicolumn{1}{|c|}{Tunnel B-5} & \multicolumn{1}{c|}{Exponential} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.03} & \multicolumn{1}{c|}{45} \\ 
\hline
\end{tabular}
\label{tblavaimain:11}
\end{table}

The impacts incurred when each link does not provide an adequate LOS,
and all others do, are given in Table \ref{tblavaimain:12}, in terms of monetary units
(\textit{mus}) incurred per day of inadequate service

\begin{table}[h]
\caption{Impacts}
\begin{tabular}{|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Link} & \multicolumn{4}{c|}{Impacts (mus)} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{Owner} & \multicolumn{1}{c|}{Users} & \multicolumn{1}{c|}{DAP} & \multicolumn{1}{c|}{IAP} \\ 
\hline
\multicolumn{1}{|c|}{Road A-1} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{3} \\ 
\hline
\multicolumn{1}{|c|}{Road 2-3} & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{Road 3-4} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{3} \\ 
\hline
\multicolumn{1}{|c|}{Road 3-5} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{Road 4-5} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{Bridge 1-2} & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{3} \\ 
\hline
\multicolumn{1}{|c|}{Bridge 1-3} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{Bridge A-4} & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{Tunnel B-2} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{Tunnel B-5} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{2} \\ 
\hline
\end{tabular}\\
Note: DAP and IAP stand for directly affected public and indirectly
affected public
\label{tblavaimain:12}
\end{table}
\subsubsection{Question B1}

Estimate the reliability and availability of each link and of the entire
network for each year in 20 years.
\subsubsection{Answer B1}
\underline{Step 1:} Estimate the reliability of each link

The network can be simplified with the circles representing the links
and the connecting point representing the cities and towns. The reliability of
each link is given in Table \ref{tblavaimain:13} and illustrated in Figure \ref{figavaimain-a:3}.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=386pt]{avaimain-a-2.eps}
\caption{Network simplification}
% \end{center}
\end{figure}

\begin{table}[h]
\caption{Reliability}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
Time & \multicolumn{10}{c|}{Links} & Networks \\ 
\cline{2-11}
(years) & \multicolumn{5}{c|}{Road sections} & \multicolumn{3}{c|}{Bridge} & \multicolumn{2}{c|}{Tunnels} &  \\ 
\cline{2-11}
 & A-1 & 2-3 & 3-4 & 3-5 & 4-5 & 1-2 & 1-3 & A-4 & B-2 & B-5 &  \\ 
\hline
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\ 
\hline
2 & 0.98 & 0.98 & 0.97 & 0.94 & 0.96 & 1 & 1 & 1 & 0.96 & 0.97 & 1 \\ 
\hline
3 & 0.95 & 0.96 & 0.94 & 0.89 & 0.92 & 0.99 & 1 & 1 & 0.92 & 0.94 & 0.99 \\ 
\hline
4 & 0.93 & 0.93 & 0.91 & 0.84 & 0.89 & 0.98 & 0.99 & 1 & 0.89 & 0.91 & 0.98 \\ 
\hline
5 & 0.91 & 0.91 & 0.89 & 0.79 & 0.85 & 0.97 & 0.99 & 1 & 0.85 & 0.89 & 0.97 \\ 
\hline
6 & 0.89 & 0.89 & 0.86 & 0.74 & 0.82 & 0.96 & 0.98 & 0.99 & 0.82 & 0.86 & 0.95 \\ 
\hline
7 & 0.87 & 0.87 & 0.84 & 0.7 & 0.79 & 0.94 & 0.97 & 0.99 & 0.79 & 0.84 & 0.92 \\ 
\hline
8 & 0.85 & 0.85 & 0.81 & 0.66 & 0.76 & 0.93 & 0.96 & 0.98 & 0.76 & 0.81 & 0.89 \\ 
\hline
9 & 0.83 & 0.83 & 0.79 & 0.62 & 0.73 & 0.91 & 0.94 & 0.98 & 0.73 & 0.79 & 0.86 \\ 
\hline
10 & 0.81 & 0.81 & 0.76 & 0.58 & 0.7 & 0.9 & 0.93 & 0.97 & 0.7 & 0.76 & 0.82 \\ 
\hline
11 & 0.79 & 0.79 & 0.74 & 0.55 & 0.67 & 0.88 & 0.91 & 0.96 & 0.67 & 0.74 & 0.78 \\ 
\hline
12 & 0.77 & 0.78 & 0.72 & 0.52 & 0.64 & 0.87 & 0.9 & 0.95 & 0.64 & 0.72 & 0.74 \\ 
\hline
13 & 0.75 & 0.76 & 0.7 & 0.49 & 0.62 & 0.85 & 0.88 & 0.94 & 0.62 & 0.7 & 0.69 \\ 
\hline
14 & 0.73 & 0.74 & 0.68 & 0.46 & 0.59 & 0.83 & 0.86 & 0.93 & 0.59 & 0.68 & 0.65 \\ 
\hline
15 & 0.71 & 0.72 & 0.66 & 0.43 & 0.57 & 0.81 & 0.84 & 0.91 & 0.57 & 0.66 & 0.6 \\ 
\hline
16 & 0.7 & 0.71 & 0.64 & 0.41 & 0.55 & 0.79 & 0.82 & 0.9 & 0.55 & 0.64 & 0.56 \\ 
\hline
17 & 0.68 & 0.69 & 0.62 & 0.38 & 0.53 & 0.78 & 0.79 & 0.89 & 0.53 & 0.62 & 0.52 \\ 
\hline
18 & 0.66 & 0.68 & 0.6 & 0.36 & 0.51 & 0.76 & 0.77 & 0.87 & 0.51 & 0.6 & 0.47 \\ 
\hline
19 & 0.65 & 0.66 & 0.58 & 0.34 & 0.49 & 0.74 & 0.75 & 0.85 & 0.49 & 0.58 & 0.43 \\ 
\hline
20 & 0.63 & 0.65 & 0.57 & 0.32 & 0.47 & 0.72 & 0.72 & 0.83 & 0.47 & 0.57 & 0.39 \\ 
\hline
\end{tabular}
\label{tblavaimain:13}
\end{table}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=429pt]{avaimain-a-11.eps}
\caption{Reliability of each link and network + network availability over 20
years}\label{figavaimain-a:3}
% \end{center}
\end{figure}

\underline{Step 2:} Estimate the reliability of the network

First, it is necessary to determine the structure function, which can be
calculated using the pivotal decomposition method.

First, link 1-3 is selected as the pivot. When link 1-3 is perfect, the
original network can be further simplified as shown in Figure \ref{figavaimain-a:3}, and when link
1-3 is not perfect, the original network diagram can be further simplified as
shown in Figure \ref{figavaimain-a:4}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=340pt]{avaimain-a-3.eps}
\caption{Network simplification - decomposition when when ${x_{1 - 3}} = 1$}\label{figavaimain-a:3}
% \end{center}
\end{figure}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=340pt]{avaimain-a-4.eps}
\caption{Network simplification - decomposition when when ${x_{1 - 3}} = 0$}\label{figavaimain-a:4}
% \end{center}
\end{figure}
The structure function of the network can then be expressed as:
\begin{eqnarray}
&& \phi (\vec x) = {x_{1 - 3}} \cdot r({1_{1 - 3}},\vec x) + (1 - {x_{1 - 3}})
\cdot r({0_{1 - 3}},\vec x)
\label{eqavaimain:5}
%(8)
\end{eqnarray}
However, even with the newly simplified network, still it is difficult
to estimate the value of the sub-structure function $r({1_{1 - 3}},\vec x)$ and
$r({0_{1 - 3}},\vec x)$directly. It is, therefore, necessary to further decompose
the simplified sub-networks.

With respect to the network $r({1_{1 - 3}},\vec x)$, the link 3-5 is
selected to be the pivot. When the link 3-5 is 100\% reliable, the network in
Figure 6 becomes the one shown in Figure \ref{figavaimain-a:5}, and when the link 3-5 is
0\% reliable, the network in Figure 6 becomes the one shown in Figure
\ref{figavaimain-a:6}.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=347pt]{avaimain-a-5.eps}
\caption{Network simplification - decomposition when $x_{3-5}=1$ and $x_{1-3}=1$}\label{figavaimain-a:5}
% \end{center}
\end{figure}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=374pt]{avaimain-a-6.eps}
\caption{Network simplification - decomposition when $x_{3-5}=0$ and $x_{1-3}=1$}\label{figavaimain-a:6}
% \end{center}
\end{figure}

The structure function $r({1_{1 - 3}},\vec x)$ is then:
\begin{eqnarray}
&& r({1_{1 - 3}},\vec x) = {x_{3 - 5}} \cdot r_{1 - 3}^1({1_{3 - 5}},\vec x) + (1 -
{x_{3 - 5}})r_{1 - 3}^1({0_{3 - 5}},\vec x)
\label{eqavaimain:6}
%(9)
\end{eqnarray}
where $r_{1 - 3}^1({1_{3 - 5}},\vec x)$ and $r_{1 - 3}^1({0_{3 -
5}},\vec x)$refer to the sub-structure functions after link 1-3 and link 3-5 have
been decomposed when link 1-3 is perfect.

The value of $r_{1 - 3}^1({1_{3 - 5}},\vec x)$ and $r_{1 - 3}^1({0_{3 -
5}},\vec x)$can now be directly computed as follows:
\begin{eqnarray}
&& \begin{array}{l}
r_{1 - 3}^1({1_{3 - 5}},\vec x) = \left[ {1 - (1 - {x_{A - 1}})(1 - {x_{A - 4}}
\cdot \left\{ {1 - (1 - {x_{3 - 4}}) \cdot (1 - {x_{4 - 5}})} \right\})} \right]
\cdot \\
\left[ {1 - (1 - {x_{B - 5}}) \cdot (1 - {x_{B - 2}} \cdot \left\{ {1 - (1 -
{x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right\}} \right]
\end{array}
\label{eqavaimain:7}
%(10)
\end{eqnarray}
\begin{eqnarray}
&& \begin{array}{l}
r_{1 - 3}^1({0_{3 - 5}},\vec x) = {x_{A - 1}}{x_{3 - 4}} \cdot {x_{B - 2}} \cdot
\left[ {1 - (1 - {x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right]\\
{\rm{                 }} + {x_{A - 1}}{x_{3 - 4}}{x_{4 - 5}}{x_{B - 5}}\\
{\rm{                 }} + {x_{A - 4}}{x_{3 - 4}}{x_{B - 2}} \cdot \left[ {1 -
(1 - {x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right]\\
{\rm{                 }} + {x_{A - 4}}{x_{3 - 4}}{x_{4 - 5}}{x_{B - 5}}\\
{\rm{                 }} + {x_{A - 1}}{x_{A - 4}}{x_{4 - 5}}{x_{B - 5}}{x_{B -
2}} \cdot \left[ {1 - (1 - {x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right]\\
{\rm{                 }} - {x_{A - 1}}{x_{A - 4}}{x_{3 - 4}}{x_{B - 2}} \cdot
\left[ {1 - (1 - {x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right]\\
{\rm{                 }} - {x_{A - 1}}{x_{3 - 4}}{x_{B - 2}} \cdot \left[ {1 -
(1 - {x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right]{x_{4 - 5}}{x_{B - 5}}\\
{\rm{                 }} - {x_{A - 1}}{x_{A - 4}}{x_{3 - 4}}{x_{4 - 5}}{x_{B -
5}}\\
{\rm{                 }} - {x_{A - 4}}{x_{3 - 4}}{x_{B - 2}} \cdot \left[ {1 -
(1 - {x_{1 - 2}}) \cdot (1 - {x_{2 - 3}})} \right]{x_{4 - 5}}{x_{B - 5}}
\end{array}
\label{eqavaimain:8}
%(11)
\end{eqnarray}
% Equation \eqref{eqavaimain:7} is described based on the results of structure function obtained in section 3.3 of the script of week 9.
With respect to the network $r({0_{1 - 3}},\vec x)$, the link 3-5 is
choosen to be the pivot. When link 3-5 is 100\% reliable, the network in Figure
\ref{figavaimain-a:4} becomes that shown in Figure \ref{figavaimain-a:7} and when the link 3-5 is 0\%
reliable, the network in Figure \ref{figavaimain-a:4} becomes that in Figure \ref{figavaimain-a:8}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=386pt]{avaimain-a-7.eps}
\caption{Network simplification - decomposition $x_{3-5}=1$ and $x_{1-3}=0$}\label{figavaimain-a:7}
% \end{center}
\end{figure}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=391pt]{avaimain-a-8.eps}
\caption{Network simplification - decomposition $x_{3-5}=0$ and $x_{1-3}=0$}\label{figavaimain-a:8}
% \end{center}
\end{figure}
The structure function $r({0_{1 - 3}},\vec x)$ is:
\begin{eqnarray}
&& r({0_{1 - 3}},\vec x) = {x_{3 - 5}} \cdot r_{1 - 3}^0({1_{3 - 5}},\vec x) + (1 -
{x_{3 - 5}}) \cdot r_{1 - 3}^0({0_{3 - 5}},\vec x)
\label{eqavaimain:9}
%(12)
\end{eqnarray}
where $r_{1 - 3}^0({1_{3 - 5}},\vec x)$ and $r_{1 - 3}^0({0_{3 -
5}},\vec x)$refer to the sub-structure function after link 1-3 and link 3-5 have
been decomposed when link 1-3 is 0\% reliable.

The values of $r_{1 - 3}^0({1_{3 - 5}},\vec x)$ and $r_{1 - 3}^0({0_{3 -
5}},\vec x)$ can now be directly computed

\begin{eqnarray}
&& \begin{array}{l}
r_{1 - 3}^0({1_{3 - 5}},\vec x) = {x_{A - 1}} \cdot {x_{1 - 2}} \cdot {x_{2 -
3}} \cdot {x_{B - 2}}\\
{\rm{                 }} + {x_{A - 1}} \cdot {x_{1 - 2}} \cdot {x_{2 - 3}} \cdot
{x_{B - 5}}\\
{\rm{                 }} + {x_{A - 4}} \cdot \left[ {1 - (1 - {x_{3 - 4}}) \cdot
(1 - {x_{4 - 5}})} \right] \cdot {x_{2 - 3}} \cdot {x_{B - 2}}\\
{\rm{                 }} + {x_{A - 4}} \cdot \left[ {1 - (1 - {x_{3 - 4}}) \cdot
(1 - {x_{4 - 5}})} \right] \cdot {x_{2 - 3}} \cdot {x_{B - 5}}\\
{\rm{                 }} + {x_{A - 1}} \cdot {x_{1 - 2}} \cdot {x_{A - 4}} \cdot
\left[ {1 - (1 - {x_{3 - 4}}) \cdot (1 - {x_{4 - 5}})} \right] \cdot {x_{B - 5}}
\cdot {x_{B - 2}}\\
{\rm{                 }} - {x_{A - 1}} \cdot {x_{1 - 2}} \cdot {x_{A - 4}} \cdot
\left[ {1 - (1 - {x_{3 - 4}}) \cdot (1 - {x_{4 - 5}})} \right] \cdot {x_{2 - 3}}
\cdot {x_{B - 2}}\\
{\rm{                 }} - {x_{A - 1}} \cdot {x_{1 - 2}} \cdot {x_{2 - 3}} \cdot
{x_{B - 2}} \cdot {x_{B - 5}}\\
{\rm{                 }} - {x_{A - 1}} \cdot {x_{1 - 2}} \cdot {x_{A - 4}} \cdot
\left[ {1 - (1 - {x_{3 - 4}}) \cdot (1 - {x_{4 - 5}})} \right] \cdot {x_{2 - 3}}
\cdot {x_{B - 5}}\\
{\rm{                 }} - {x_{A - 4}} \cdot \left[ {1 - (1 - {x_{3 - 4}}) \cdot
(1 - {x_{4 - 5}})} \right] \cdot {x_{2 - 3}} \cdot {x_{B - 5}} \cdot {x_{B - 2}}
\end{array}
\label{eqavaimain:10}
%(13)
\end{eqnarray}
The set of equation from equation \eqref{eqavaimain:5} to equation \eqref{eqavaimain:11} can be used
directly to calculate the reliability of the network given the reliability of the
individual link every year. The reliability of each link and the network for
periods of time ranging from 0 to 20 years are given in Table \ref{tblavaimain:13} and
illustrated in Figure \ref{figavaimain-a:3}.

R code for this example is given in Appendix B.

\underline{Step 3:} Estimate the availability of the network

Similar to reliability, the set of structure functions above can be used
to compute the availability of the network based on the reliability of individual
link. The availability of individual link is calculated based on following
equation
\begin{eqnarray}
&& A(t_i^{}) = \frac{{E\left[ {t_i^a} \right]}}{{E\left[ {t_i^a} \right] + E\left[
{t_i^n} \right]}} = \frac{{\int\limits_0^{tu} {{R_i}(t)dt} }}{{\int\limits_0^{tu}
{{R_i}(t)dt}  + {\Delta _i} \cdot \left[ {1 - {R_i}(tu)} \right]}}
\label{eqavaimain:11}
%(15)
\end{eqnarray}
where ${R_i}(t)$ is the reliability of link i at time t and ${\Delta
_i}$is the mean time of intervention for link \textit{i}.

The availabilities of each link and the network for periods of time from
0 to 20 years are given in Table \ref{tblavaimain:14}, and illustrated in Figure
\ref{figavaimain-a:3}.

\begin{table}[h]
\caption{Availability}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
Time & \multicolumn{10}{c|}{Links} & Networks \\ 
\cline{2-11}
(years) & \multicolumn{5}{c|}{Road sections} & \multicolumn{3}{c|}{Bridge} & \multicolumn{2}{c|}{Tunnels} &  \\ 
\cline{2-11}
 & A-1 & 2-3 & 3-4 & 3-5 & 4-5 & 1-2 & 1-3 & A-4 & B-2 & B-5 &  \\ 
\hline
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\ 
\hline
2 & 0.999335 & 0.999502 & 0.999431 & 0.998864 & 0.99865 & 0.999022 & 0.998376 & 0.99773 & 0.997976 & 0.998496 & 0.999995 \\ 
\hline
3 & 0.999114 & 0.999336 & 0.99924 & 0.998459 & 0.998187 & 0.998707 & 0.997859 & 0.997011 & 0.997282 & 0.99799 & 0.999992 \\ 
\hline
4 & 0.999003 & 0.999253 & 0.999142 & 0.998237 & 0.997945 & 0.998557 & 0.997615 & 0.996675 & 0.99692 & 0.997733 & 0.99999 \\ 
\hline
5 & 0.998937 & 0.999204 & 0.999083 & 0.998088 & 0.997791 & 0.998471 & 0.99748 & 0.996492 & 0.996691 & 0.997575 & 0.999988 \\ 
\hline
6 & 0.998892 & 0.999171 & 0.999042 & 0.997975 & 0.997682 & 0.998417 & 0.997397 & 0.996385 & 0.996527 & 0.997467 & 0.999987 \\ 
\hline
7 & 0.998861 & 0.999148 & 0.999012 & 0.997883 & 0.997599 & 0.998382 & 0.997344 & 0.996319 & 0.996402 & 0.997388 & 0.999986 \\ 
\hline
8 & 0.998837 & 0.999131 & 0.998988 & 0.997804 & 0.997531 & 0.998357 & 0.997308 & 0.996279 & 0.996301 & 0.997327 & 0.999986 \\ 
\hline
9 & 0.998819 & 0.999118 & 0.998969 & 0.997734 & 0.997473 & 0.998339 & 0.997284 & 0.996255 & 0.996215 & 0.997277 & 0.999985 \\ 
\hline
10 & 0.998804 & 0.999107 & 0.998954 & 0.99767 & 0.997424 & 0.998325 & 0.997266 & 0.996242 & 0.99614 & 0.997235 & 0.999985 \\ 
\hline
11 & 0.998792 & 0.999099 & 0.99894 & 0.99761 & 0.997379 & 0.998315 & 0.997254 & 0.996236 & 0.996074 & 0.9972 & 0.999984 \\ 
\hline
12 & 0.998782 & 0.999092 & 0.998929 & 0.997554 & 0.997339 & 0.998308 & 0.997244 & 0.996236 & 0.996014 & 0.997169 & 0.999984 \\ 
\hline
13 & 0.998774 & 0.999086 & 0.998918 & 0.9975 & 0.997302 & 0.998302 & 0.997236 & 0.996238 & 0.995958 & 0.997142 & 0.999984 \\ 
\hline
14 & 0.998766 & 0.999081 & 0.998909 & 0.997448 & 0.997268 & 0.998297 & 0.99723 & 0.996243 & 0.995907 & 0.997118 & 0.999983 \\ 
\hline
15 & 0.99876 & 0.999076 & 0.998901 & 0.997398 & 0.997235 & 0.998293 & 0.997225 & 0.99625 & 0.995858 & 0.997096 & 0.999983 \\ 
\hline
16 & 0.998755 & 0.999073 & 0.998893 & 0.997349 & 0.997204 & 0.998289 & 0.997219 & 0.996257 & 0.995812 & 0.997075 & 0.999983 \\ 
\hline
17 & 0.99875 & 0.99907 & 0.998886 & 0.997302 & 0.997175 & 0.998286 & 0.997214 & 0.996264 & 0.995768 & 0.997057 & 0.999983 \\ 
\hline
18 & 0.998746 & 0.999067 & 0.998879 & 0.997256 & 0.997147 & 0.998284 & 0.997208 & 0.996272 & 0.995726 & 0.997039 & 0.999983 \\ 
\hline
19 & 0.998742 & 0.999064 & 0.998873 & 0.99721 & 0.99712 & 0.998281 & 0.997202 & 0.996279 & 0.995686 & 0.997023 & 0.999982 \\ 
\hline
20 & 0.998738 & 0.999062 & 0.998867 & 0.997165 & 0.997093 & 0.998279 & 0.997196 & 0.996286 & 0.995647 & 0.997007 & 0.999982 \\ 
\hline
\end{tabular}
\label{tblavaimain:14}
\end{table}
\subsubsection{Question B2}
Calculate the total impacts due to links not being able to provide an
adequate LOS and the network over a 20 year period. (Assume that link failures
are mutually exclusive.)
\subsubsection{Answer B2}
The expected cost is estimated based on following equation 
\begin{eqnarray}
&& {C_{am}} = \frac{{E\left[ {t_i^n} \right] \cdot {I_{cr}}/12}}{{E\left[ {t_i^a}
\right] + E\left[ {t_i^n} \right]}}
\label{eqavaimain:12}
%(15)
\end{eqnarray}
The mean time between failures can be calculated with following
equation:
\begin{eqnarray}
&& E\left[ {t_i^a} \right] = {\Theta _i} = \int\limits_0^\infty  {{r_i}(t)dt} 
\label{eqavaimain:13}
%(15)
\end{eqnarray}
where ${r_i}(t)$ is the reliability of link \textit{i}.

Table \ref{tblavaimain:15} shows the summary of total cost for each link and the
total cost for the network

\begin{table}[h]
\caption{Cost}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
Time & \multicolumn{10}{c|}{Links} & Total \\ 
\cline{2-11}
(years) & \multicolumn{5}{c|}{Road sections} & \multicolumn{3}{c|}{Bridge} & \multicolumn{2}{c|}{Tunnels} & costs \\ 
\cline{2-11}
 & A-1 & 2-3 & 3-4 & 3-5 & 4-5 & 1-2 & 1-3 & A-4 & B-2 & B-5 &  \\ 
\hline
Owner & 18.27 & 18.56 & 18.39 & 44.18 & 42.18 & 29.96 & 57.86 & 43.03 & 94.78 & 38.87 & 406.08 \\ 
\hline
User & 21.93 & 13.26 & 25.74 & 66.27 & 42.18 & 25.68 & 36.16 & 43.03 & 63.19 & 29.15 & 366.59 \\ 
\hline
DAP & 10.96 & 10.61 & 14.71 & 44.18 & 31.64 & 17.12 & 28.93 & 18.44 & 63.19 & 38.87 & 278.64 \\ 
\hline
IAP & 10.96 & 5.3 & 11.03 & 22.09 & 10.55 & 12.84 & 14.46 & 6.15 & 31.59 & 19.43 & 144.42 \\ 
\hline
\textbf{Total} & 62.13 & 47.74 & 69.87 & 176.72 & 126.54 & 85.6 & 137.41 & 110.64 & 252.75 & 126.32 & 1195.73 \\ 
\hline
\end{tabular}
\label{tblavaimain:15}
\end{table}

\bibliographystyle{plainnat}
\bibliography{reference}
% \newpage
\pagebreak

\begin{subappendices}
% \label{appendix3}
\section{Example - Reliability \& Availability}\label{appavai-mai1}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-avai-main/IMP-E-availability.R}
\pagebreak
\section{Example - HongKong tunnel}\label{appavai-mai2}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-avai-main/IMP-E-tunnel.R}
\pagebreak
\section{Assignment - Availability-log-normal distribution}\label{appavai-mai3}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-avai-main/IMP-A-availability.r}
\pagebreak
\section{Assignment - availability, reliability, and impact estimation for the network}\label{appavai-mai3}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-avai-main/IMP-A-availability-network.R}
\end{subappendices}

### 2019-05-05 InfrastructureMaintenance/Chapters/bayesian.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Bayesian Computation and Bayesian Network Models}
\label{bayesianchapter} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Nam Lethanh}
\section{Introduction}

% Within the domain of infrastructure asset management (IAM), prediction models play an important role. The prediction models can be used to oversee the evolution of changes on various indicators and impacts affecting the decision on interventions. For instance, the prediction of evolution of discrete condition state representing the physical condition of bridge elements or road sections using mechanistic-empirical models or using Markov models; or the prediction of accident rate at each road section on an entire transportation network.
% 
% Prediction is closely dependent on data and data can be in various form and categories. For example, in order to predict the future condition of a bridge element, it is necessary to use data on time series data of cracking, chloride induced corrosion, fatigues, daily traffic volume, ambient temperature, etc.
% 
% In many case, mathematically, prediction models are of complex form that require highly computational efforts and development of algorithms. In addition, data under each categories can be linked as they are, in many cases, more or less statistically correlated. For example, accident rate at a road section is attributed to friction coefficient of the road surface, slope, speed limit, and also weather factors such as rain and snow. Also, in many cases, data is insufficient or missing due to measurement errors or limitation of resources.
% 
% One of the missions of research in this field is to come up with or apply the best method possible for prediction and analysis given requirements and constraints such as computational efforts, measurement errors, level of accuracy of prediction, missing data, and statistical correlated data.
Recently, Bayesian statistics has gained its ground in the research domain of the IAM and transportation engineering domain. It is considered as a powerful approach that can be used to solve many problems in the IAM. It is more advantageous than the traditional frequentist statistics \footnote{the terminologies ``frequentist statistics'' and ``Bayesian statistics'' are used commonly in various statistical textbooks} such as least squared estimation or regression analysis with Maximum likelihood estimation (MLE) approach.

A simple example on estimation of accident rate on a road link herewith to provide readers a basic difference between the traditional frequentist statistics and Bayesian statistics.

In the past 10 years, it is known that the probability of having an accident on a specific road section is 0.001 in a day. This value is considered as an expected mean probability, which is calculated by simply based on observed data. If someone asks you a question ``what will be the probability of having an accident on that road tomorrow?'', the correct answer is no longer with the value of 0.001, but it is different. The probability of having an accident tomorrow is a conditional probability that is calculated based on its dependency on observation of accident today, weather condition, number of daily traffic volume that are expected to be tomorrow. Frequentist statistics cannot provide answer to this question, but Bayesian statistics can because in the heart of Bayesian statistics, it is Bayes rule that allows to take into consideration of conditional probability and degree of belief based on various observed parameters such as accident today, weather condition, numbers of daily traffic volume, as well as opinions from experts \citep{Andrew2006}.

In briefly, Bayesian statistical approach can be used in replacement of frequentist statistical approaches for all of traditional statistical methods such as regression analysis, least squared estimation, and Maximum likelihood estimation approach. In addition, it can be used to solve complex and hierarchical models that methods used in traditional statistics cannot.

In the course of parameter estimation of statistical models, traditional statistics use analytical way to derive the solution. This analytical way faces difficulties when models are in complex hierarchical structures. In addition, traditional statistics is largely dependent on data. If data is insufficient or missing, the likelihood of estimation certainly involves greater bias and lower confidence \citep{Andrew2006,jeffgill}. Analytical method itself requires, for example, derivative approach for each model's parameters. For example, when using the MLE approach, it is necessary to obtain the first and second derivatives for each parameters. In mathematically term, it is necessary to derive the Jacobian matrices and Hessian matrices, then use Newton methods to obtain the values of model's parameters given data. However, in many cases, it is impossible to derive the Jacobian or Hessian matrices \footnote{Jacobian and Hessian matrices are matrices of the first and second derivatives of model's parameters}, leaving the problem unsolvable with the MLE approach. In this situation, the Bayesian approach can be used as it does not require the analytical solution, i.e. there is no need to obtain the Jacobian and Hessian matrices.

In this chapter, the basic elements of the Bayesian inferential approach are introduced through examples related to prediction models for the IAM.
%

\subsection{Bayes's theorem}\label{bayesiantheorem}
Although literature on Bayesian statistics has been documented in a countless number, we would prefer to briefly present the Bayes's theorem in this section for the convenience of the readers and to serve as connection to the examples presented in this chapter.

In Bayesian statistics, the probability $P(A)$ of an event $A$ is formulated as a degree of belief that $A$ will occur. This degree of belief is subsequently referred as a prior probability which can be assumed relying on a subjective experience (theoretical assumption) or based on data of experiments or observations (frequentistic, empirical). Bayesian inference combines prior information and current information to derive the probability $P(A)$. This approach is considered as a process of fitting a probabilistic model to a set of data and summarizing the result by a probability distribution on the parameters of the model and on unobserved quantities such as predictions for new observations \cite{Geman1984}.

In the Bayesian statistics, the posterior distribution of parameters is estimated by using the likelihood function, which is defined by the prior distribution of parameters and the observed data. Here, the likelihood function is represented by ${\cal L}(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})$. $\mbox{\boldmath$\beta$}$ and $\mbox{\boldmath$\xi$}$ denote the unknown parameter vector and the observed data, respectively. 
It is assumed that $\mbox{\boldmath$\beta$}$ is a random variable, and is subjected to the prior probability density function $p($\mbox{\boldmath$\beta$}$)$. Under these conditions and according to Baye's law, when the observed data $\mbox{\boldmath$\xi$}$ is given, the posterior probability density function $p($\mbox{\boldmath$\beta$}$|\mbox{\boldmath$\xi$})$ of the unknown parameters $\mbox{\boldmath$\beta$}$ is defined as:
 %
   \begin{eqnarray}
      && p(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})=
      \frac{{\cal L}(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})
      p(\mbox{\boldmath$\beta$})}
      {\int_{\Theta}{\cal L}(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})
      p(\mbox{\boldmath$\beta$})d\mbox{\boldmath$\beta$}}
      \label{bayes1}
   \end{eqnarray}
 %
where $\Theta$ represents the parameter space. At this time, $p($\mbox{\boldmath$\beta$}$|\mbox{\boldmath$\xi$})$ can be expressed as follows:
 %
   \begin{eqnarray}
      && p(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})
      \propto {\cal L}(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})
      p(\mbox{\boldmath$\beta$})
      \label{bayes2}
   \end{eqnarray}
 %
The symbol $\propto$ denotes 'be proportional to'. The denominator of the right-hand side of Eq. (\ref{bayes1}):
 %
   \begin{eqnarray}
      && m(\mbox{\boldmath$\xi$})
      =\int_{\Theta}{\cal L}(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})
      p(\mbox{\boldmath$\beta$})d\mbox{\boldmath$\beta$}
      \label{bayes3}
   \end{eqnarray}
 %
is called the normalization constant of $p(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})$, or the prior predictive distribution.

In short, the Bayes's theorem is summarized in following notation
\begin{center}
\textbf{\textit{posterior}} \hspace{2mm} $\propto$ \textbf{\textit{prior}} $\times$ \textbf{\textit{likelihood}}
\end{center}
In general, the procedures of the Bayesian estimation can be summarized in following steps:
\begin{itemize}
	\item Step 1: the prior probability distribution function $p(\mbox{\boldmath$\beta$})$ is specified, based on the prior information;
	\item Step 2: the likelihood function ${\cal L}(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})$ is defined based on the newly obtained data $\mbox{\boldmath$\xi$}$;
	\item Step 3: the prior probability density function is modified in accordance with the Bayes' theorem, and the posterior probability density function $p(\mbox{\boldmath$\beta$}|\mbox{\boldmath$\xi$})$ regarding the parameters $\mbox{\boldmath$\beta$}$ is updated.
\end{itemize}

% These procedure is regarded as 'Bayesian estimation rule' in our study. Theoretically, the Bayesian estimation method differs from the MLE method in that the probability distribution of the unknown parameters $\mbox{\boldmath$\beta$}$ is obtained as a posterior distribution. Moreover, in numerical solution with the MLE approach, to obtain the most likely value of model's parameters, it is required to derive the Jacobian (first derivative) and Hessian (second derivative) matrices for the objective function. Meanwhile, with Bayesian estimation method, Jacobian and Hessian matrices are not of requirement, and this feature is also considered as one of the advantage of Bayesian estimation method over the conventional MLE method.
\subsection{MCMC Method} \label{sec51}
In statistic with Bayesian inference, prior and posterior probability are employed with aim to estimate the values of model's parameters. However, in actual analysis, it is hard to define a prior probability distribution \citep{ibrahim01bo}. Methods to overcome the problems in the assumption of prior probability distribution often require numerical analysis with multi-dimensional integration, and thus remaining as a limitation in Bayesian estimation.

In recent years, an appealing solution to the problems in Bayesian estimation has been proposed, with the application of Markov Chain Monte Carlo (MCMC) simulation. The MCMC simulation technique does not require a high level of derivative and multi-dimensional integration of model's objective functions \citep{robert}. As a result, estimation results, in a great number of applied statistic research, have been improved through the combination of Bayesian estimation and MCMC simulation.

In MCMC simulation, Gibbs sampling and Metropolis Hastings (Metropolis-Hastings or MH) techniques have been extensively discussed \citep{robert}. Reference to research on image restoration is a good example of MCMC simulation \citep{Geman1984}. Of that study, the algorithm of Gibbs sampling was used to estimate the posterior distribution in Bayesian estimation. In MH law, the iterative parameter $\mbox{\boldmath$\beta$}$ is defined by repeatedly generating random numbers through the conditional probability density function.

% Regarding application of Bayesian estimation and MCMC method in infrastructure management, the authors of this paper has developed a hidden Markov model for elimination of selection bias \cite{Kobayashi2012a}. The use of Bayesian estimation and MCMC method has showed a great advantage over the conventional MLE approach in the case of having complete likelihood function with multiple integrations. Following sections detail our numerical solution to overcome the challanges in estimation of the COHA model's parameters.

\subsection{Example into Bayesian thinking}\label{bayesianexample1}
An infrastructure manager is interested in learning the accident risks on his/her highway network. He/she is given an information from a consultant that the numbers of accident per 100 $km$ of road section in a year should be less than 8 cases.

What proportion of all road sections in the entire network get at least 8 accidents in a year?

Here, it is thought that the entire road network is composed of hundreds of road sections. Let $p$ represents the proportion of this population which has a maximum level of 8 accidents per year. It is the interest of the infrastructure manager to understand the value of $p$, which is unknown.

As earlier discussed, in Bayesian inference, the belief of the manager on the proportion $p$ can be represented by a probability distribution. This distribution is referred as prior knowledge that the manager believe in about the value of $p$.

Certainly, the infrastructure manager might not have all data for the entire network, therefore, he/she decides to carry out a sample that focuses on only a number of road sections. However, before that, he/she does some initial investigation on this proportion by reading some scientific research papers. This will help him/her in formulating the prior distribution used to compute the value of $p$.

First based on the sample report, he/she comes to know that there are, on average, 6 accidents occurs in a year for that pool of sample. Second, he/she does reading some technical reports in other regions on 100 samples, it is concluded in that technical report that there are approximately 70\% of road sections have a numbers of accident per year falling in between 5 and 6 cases, 28\% of road sections have 7 to 8 accidents, and about 2\% of road section have more than 9 accident cases.

Based on these information, the manager believes that the numbers of accident per year for the entire network generally less than 8 cases per 100 $km$ of road is having a proportion smaller than 0.5. After some reflection, her/his best guess value of $p$ is 0.3. However, this is a very likely that the value of $p$ can fall arbitrarily between 0 and 0.5.

He/she decides to carry out an investigation with 27 samples (e.g. road sections). Among 27 samples, 11 records have at least 8 accidents.

Following assumptions are used:
\begin{itemize}
\item the prior density for $p$ is denoted by $g(p)$
\item the “good” case is when the number of accident is less than 8 cases and the “not good” is when the number of accident is greater or equal than 8 cases, then we can write down the likelihood function in equation \eqref{bayes4}                                                                                                                                                                                                                          \end{itemize}
\begin{eqnarray}
&& L(p) \propto p^n \cdot (1-p)^g\label{bayes4}
\end{eqnarray}
where $g$ and $n$ represent “good” and “not good”

The posterior density for $p$, with Bayesian inference, is obtained, up to a proportionality constant, by multiplying the prior density by the likelihood (equation \eqref{bayes1}).
\begin{eqnarray}
&& g(p|data) \propto g(p)\cdot L(p) \label{bayes5}
\end{eqnarray}
the word ``data'' in equation \eqref{bayes5} infers the notation ``$\mbox{\boldmath$\xi$}$'' in equation \eqref{bayes1}

To obtain the posterior value of $p$, one can use discrete prior distribution as explained below

A simple approach to represent the prior value of $p$ is to write it down as a list of reasonable proportion value and then assigns weights to these values. For example, following vector
\[
0.05 \hspace{0.5cm}0.15	\hspace{0.5cm}0.25\hspace{0.5cm}0.35\hspace{0.5cm}0.45\hspace{0.5cm}0.55\hspace{0.5cm}0.65\hspace{0.5cm}0.75\hspace{0.5cm}0.85\hspace{0.5cm}0.95
\]
represents possible value of $p$ discretely. Based on her/his belief, following weights to this vector are assigned

\begin{table}
%	\centering
% 	\caption{Evolution of erosion (mechanistic-empirical model)} \label{tbl:36}
\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Range of $p$} & \multicolumn{1}{m{1cm}|}{\centering 0.05} & \multicolumn{1}{m{1cm}|}{\centering0.15} & \multicolumn{1}{m{1cm}|}{\centering0.25} & \multicolumn{1}{m{1cm}|}{\centering0.35} & \multicolumn{1}{m{1cm}|}{\centering0.45} & \multicolumn{1}{m{1cm}|}{\centering0.55} & \multicolumn{1}{m{1cm}|}{\centering0.65} & \multicolumn{1}{m{1cm}|}{\centering0.75} & \multicolumn{1}{m{1cm}|}{\centering0.85} & \multicolumn{1}{m{1cm}|}{\centering0.95} \\ 
\hline
\multicolumn{1}{|c|}{Weight} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{5.2} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{7.2} & \multicolumn{1}{c|}{4.6} & \multicolumn{1}{c|}{2.1} & \multicolumn{1}{c|}{0.7} & \multicolumn{1}{c|}{0.1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\end{tabular}
\end{table}
Using this information, the discrete prior distribution can be shown graphically (Fig. \ref{figbayes1})
\begin{figure}[h]
% \begin{center}
\includegraphics[width=302pt]{figbayes1.eps}
\caption{Prior distribution for a proportion of $p$}\label{figbayes1}
% \end{center}
\end{figure}
In this example, it is known that there are 11 samples out of 27 samples have numbers of accidents equal or greater than 8, therefore the value of $n$=11 and $g$=16, the likelihood function in equation \eqref{bayes4} can be then written as
The posterior density for $p$, with Bayesian inference, is obtained, up to a proportionality constant, by multiplying the prior density by the likelihood (equation \eqref{bayes1}).
\begin{eqnarray}
&& L(p) \propto p^{11} \cdot (1-p)^{16}\label{bayes6}
\end{eqnarray}
Using R code for this problem (Appendix \ref{bayesian1}), following posterior probability of $p$ can be estimated
%
\begin{table}
%	\centering
	\caption{Prior and posterior of $p$} \label{tblbayes1}
\begin{tabular}{|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{No.} & \multicolumn{1}{m{1.5cm}|}{\centering $p$} & \multicolumn{1}{m{1.5cm}|}{\centering Prior} & \multicolumn{1}{m{1.5cm}|}{\centering Posterior} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{0.05} & \multicolumn{1}{c|}{0.03} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{0.15} & \multicolumn{1}{c|}{0.18} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{0.25} & \multicolumn{1}{c|}{0.28} & \multicolumn{1}{c|}{0.13} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{0.35} & \multicolumn{1}{c|}{0.25} & \multicolumn{1}{c|}{0.48} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{0.45} & \multicolumn{1}{c|}{0.16} & \multicolumn{1}{c|}{0.33} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{0.55} & \multicolumn{1}{c|}{0.07} & \multicolumn{1}{c|}{0.06} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{0.65} & \multicolumn{1}{c|}{0.02} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{0.75} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{0.85} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{0.95} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\end{tabular}
\end{table}
Fig. \ref{figbayes2} shows the distribution of the prior and posterior of $p$
\begin{figure}[h]
% \begin{center}
\includegraphics[width=402pt]{figbayes2.eps}
\caption{Prior and posterior distribution of $p$}\label{figbayes2}
% \end{center}
\end{figure}
The posterior value of $p$, as can be seen from the table and the figure, is about 95\% concentrating on the range from 0.25 to 0.45 . This infers that the probability that numbers of accidents per year for 100 $km$ road reaches equal or greater than 8 cases will likely to happen only in the range from 0.25 to 0.45.

% \cite{Albert2009}

% \section{Bayesian estimation for linear regression model}

\section{Bayesian inference for simple linear regression model} \label{bayline}
In section \ref{regressionmodel} of Chapter \ref{meempi}, a simple linear regression model is presented along with the squared root estimation approach. This chapter elaborates the Bayesian inference for solving this simple linear regression to show how it can be applied in a simplest model. A complete description of the approach is referred to \cite{Bolstad2007}.

In a simple regression model (e.g. Eq. \eqref{eq37}), we expect to model a relationship between two variables, $x$ and $y$. Both variables are observed data. $x$ is referred characteristic variables (sometimes also referred as predictor variable or independent variable). Example of $x$ in the context of data in the IAM are traffic volume, thickness of road sections, ambient temperature, etc. These data are observable. $y$ is often referred as outcome (e.g. condition state of road sections, percentage of cracks on a bridge deck), which is also observable. The mission is to model the relationship between $x$ and $y$ in a linear form given a data set of $n$ ordered pairs of sample ($x_i,y_i$) for $i=1,\cdots,n$.

In order to construct such a relationship, as shown in section \ref{regressionmodel} of Chapter \ref{meempi}, a derivative method is used. The method is referred as \textit{least squares} estimation.

Using this linear model, in fact, we can solve many problems in practices involving not only linear form but also nonlinear form of models. For example, in following equation describes a growth model in exponential form.
%
\begin{eqnarray}
&& y=e^{\alpha+\beta\cdot x}\label{bayes7}
\end{eqnarray}
%
This equation is used often in practice, also for the deterioration prediction with exponential form \citep{Lethanh2013b,Lethanh2015b}, for estimating impacts such as vehicle operating cost \citep{OpusCL1999} and  accident rate \citep{Kumares2007}.

The equation \eqref{bayes7} can be easily transformed into a linear form if we take logarithm for both sides of it.
\begin{eqnarray}
&& log(y)=\alpha+\beta\cdot x\label{bayes8}
\end{eqnarray}
Using the linear form in equation \eqref{bayes8}, again, we can use the least squares method to obtain the model's parameters $\alpha$ and $\beta$ given data.

% For convenience of readers, without having to refer to the regression equation in section \ref{regressionmodel} of Chapter \ref{meempi}, we rewrite that equation and name it with a new caption.
% \begin{eqnarray}
% && y=e^{\alpha+\beta\cdot x}\label{bayes9}
% \end{eqnarray}
In reference to the Bayes' theorem, for this regression model, it is necessary to define the likelihood and select 
the prior for the model's parameters $\alpha$ and $\beta$.
\subsection{The joint likelihood for $\alpha$ and $\beta$} \label{joinlik01}
For a sample $i^{th}$, the joint probability density function as a joint function of parameters $\alpha$ and $\beta$ for the paired observed data $(x_i,y_i)$ is
\begin{eqnarray}
&& likelihood_i(\alpha,\beta) \propto e^{-\frac{1}{2\sigma^2}\left[y_i-(\alpha+\beta(x_i-\bar x))\right]^2}\label{bayes10}
\end{eqnarray}
The likelihood of the entire data is the join probability and therefore is expressed as
\begin{eqnarray}
&& likelihood_{data}(\alpha,\beta) \propto \prod_{i=1}^{n} e^{-\frac{1}{2\sigma^2}\left[y_i-(\alpha+\beta(x_i-\bar x))\right]^2}\label{bayes11} \\
&& \hspace{3.5cm}  \propto -\frac{1}{2\sigma^2} \sum_{i=1}^{n}  \left[y_i-(\alpha+\beta(x_i-\bar x))\right]^2\label{bayes12}
\end{eqnarray}
The polynomial inside the bracket of equation \eqref{bayes12} equals to
\begin{eqnarray}
&& \sum_{i=1}^{n}  \left[y_i-\bar y + \bar y + (\alpha+\beta(x_i-\bar x))\right]^2\label{bayes13}
\end{eqnarray}
which equals to
\begin{eqnarray}
&& \sum_{i=1}^{n}(y_i-\bar y)^2 +2\sum_{i=1}^{n}(y_i-\bar y)(\bar y-(\alpha+\beta(x_i-\bar x)))+ \sum_{i=1}^{n} (\bar y -(\alpha+\beta(x_i-\bar x)))^2 \label{bayes14}
\end{eqnarray}
This is simplified as
\begin{eqnarray}
&& \epsilon_y -2\beta\epsilon_{xy}+\beta^2 \epsilon_x+n(\alpha-\bar y)^2 \label{bayes15}
\end{eqnarray}
where $\epsilon_y=\sum_{i=1}^{n}(y_i-\bar y)^2$, $\epsilon_{xy}=\sum_{i=1}^{n}(y_i-\bar y)(x_i-\bar x)$, and $\epsilon_x=\sum_{i=1}^{n}(x_i-\bar x)^2$.

Finally, the join likelihood can be described as
\begin{eqnarray}
&& likelihood_{data}(\alpha,\beta) \propto e^{-\frac{1}{2\sigma^2}\left[\epsilon_y -2\beta\epsilon_{xy}+\beta^2 \epsilon_x+n(\alpha-\bar y)^2\right]}\label{bayes16}\\
&& \hspace{30mm} \propto e^{-\frac{1}{2\sigma^2}\left[\epsilon_y -2\beta\epsilon_{xy}+\beta^2 \epsilon_x\right]}\times e^{-\frac{1}{2\sigma^2}\left[n(\alpha-\bar y)^2\right]} \label{bayes17}\\
&& \hspace{30mm} \propto e^{-\frac{1}{2\sigma^2/{\epsilon_x}}\left[\beta - \frac{\epsilon_{xy}}{\epsilon_x}\right]^2}\times e^{-\frac{1}{2\sigma^2/n}\left[(\alpha-\bar y)^2\right]} \label{bayes18}
\end{eqnarray}
In equation \eqref{bayes18}, it is obvious to recognize that the ratio $B=\frac{\epsilon_{xy}}{\epsilon_x}$ is the slop of the least squared line and $\bar y$ is the intercept of the vertical line $A=\bar y$. Evidently, the join likelihood of the two parameter has been factored .
\begin{eqnarray}
&& likelihood_{data}(\alpha,\beta) \propto likelihood_{data}(\alpha)\times likelihood_{data}(\beta) \label{bayes19}
\end{eqnarray}
where\\
\begin{eqnarray}
 && likelihood_{data}(\beta) = e^{-\frac{1}{2\sigma^2/{\epsilon_x}}\left[\beta - B\right]^2}  \label{bayes20}
\end{eqnarray}
and
\begin{eqnarray}
 && likelihood_{data}(\alpha) = e^{-\frac{1}{2\sigma^2/n}\left[(\alpha-A)^2\right]} \label{bayes21}
\end{eqnarray}
It can be seen from equation \eqref{bayes20} and \eqref{bayes21} of $\beta$ and $\alpha$ that the two likelihood functions are normal distributions $\mathcal{N}(B,\frac{\sigma^2}{\epsilon_x})$ and $\mathcal{N}(A,\frac{\sigma^2}{n})$, respectively.

\subsection{The joint prior for $\alpha$ and $\beta$}
The joint prior for $\alpha$ and $\beta$ is the multiplication of the two priors
\begin{eqnarray}
 && g(\alpha,\beta) = g(\alpha)\times g(\beta) \label{bayes22}
\end{eqnarray}
The selection for the distribution of $g(\alpha)$ and $g(\beta)$ can be normal distribution, which is a conjugate distribution to the likelihood.
\subsection{The join posterior for $\alpha$ and $\beta$}
From the definition of Bayes's theorem ($posterior \propto prior \times likelihood $), we can have the joint posterior as follows
\begin{eqnarray}
 && g(\alpha,\beta|data) \propto g(\alpha,\beta)\times likelihood_{data}(\alpha,\beta) \propto g(\alpha|data)\times g(\beta|data) \label{bayes23}
\end{eqnarray}
\subsection{Example - Accident data on road}
Table \ref{tblbayes2} shows the data on numbers of accident occurred on different road sections during the period of 2 year. In order to understand the functional relationship between the numbers of accident and the characteristic variables such as condition state (CS), daily traffic volume (DTV), slope, and speed of vehicle at the time of accident, investigator use a linear model (refer to equation \eqref{eq37}). In Figure \ref{figbayesx}, scatter plots are shown based on the data given in the table.

\begin{table}
	\centering
\caption{Accident data on a road link} \label{tblbayes2}
\begin{tabular}{|c|p{1.5cm}|c|c|c|c|}
\hline
Sections & \centering Numbers of accident & CS & DTV & Slope & Speed \\ 
\hline
1 & \centering 5 & 2 & 9'746 & 19 & 52 \\ 
\hline
2 & \centering 2 & 1 & 6'878 & 1 & 59 \\ 
\hline
3 & \centering 4 & 2 & 5'758 & 0 & 50 \\ 
\hline
4 & \centering 5 & 1 & 6'903 & 17 & 52 \\ 
\hline
5 & \centering 10 & 2 & 9'114 & 20 & 54 \\ 
\hline
6 & \centering 8 & 3 & 7'890 & 8 & 80 \\ 
\hline
7 & \centering 10 & 4 & 9'539 & 12 & 65 \\ 
\hline
8 & \centering 2 & 1 & 7'062 & 3 & 72 \\ 
\hline
9 & \centering 3 & 5 & 8'846 & 9 & 54 \\ 
\hline
10 & \centering 2 & 5 & 7'383 & 3 & 79 \\ 
\hline
11 & \centering 6 & 5 & 9'755 & 18 & 77 \\ 
\hline
12 & \centering 10 & 3 & 7'351 & 10 & 51 \\ 
\hline
13 & \centering 3 & 3 & 8'376 & 17 & 63 \\ 
\hline
14 & \centering 7 & 2 & 7'646 & 15 & 72 \\ 
\hline
15 & \centering 1 & 5 & 7'002 & 10 & 74 \\ 
\hline
16 & \centering 5 & 1 & 8'644 & 19 & 78 \\ 
\hline
17 & \centering 2 & 1 & 9'719 & 7 & 69 \\ 
\hline
18 & \centering 5 & 2 & 8'851 & 13 & 54 \\ 
\hline
19 & \centering 5 & 3 & 9'353 & 19 & 67 \\ 
\hline
20 & \centering 10 & 4 & 7'824 & 18 & 68 \\ 
\hline
21 & \centering 0 & 2 & 7'092 & 10 & 50 \\ 
\hline
22 & \centering 4 & 1 & 7'383 & 9 & 52 \\ 
\hline
23 & \centering 1 & 3 & 7'716 & 8 & 51 \\ 
\hline
24 & \centering 10 & 2 & 6'810 & 1 & 63 \\ 
\hline
25 & \centering 3 & 1 & 8'506 & 18 & 54 \\ 
\hline
26 & \centering 7 & 1 & 5'073 & 0 & 79 \\ 
\hline
27 & \centering 7 & 2 & 9'331 & 6 & 63 \\ 
\hline
28 & \centering 8 & 2 & 9'516 & 10 & 65 \\ 
\hline
29 & \centering 9 & 5 & 6'495 & 7 & 64 \\ 
\hline
30 & \centering 1 & 2 & 7'968 & 19 & 74 \\ 
\hline
\end{tabular}
\end{table}

\begin{figure}
 \begin{minipage}[h]{0.5\linewidth}
        \centering
        \includegraphics[scale=0.34]{figbayes3.eps}
				\subcaption{Numbers of accident vs CS}\label{figbayes3}
     \end{minipage}
\vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.34]{figbayes4.eps}
			\subcaption{Numbers of accident vs DTV}\label{figbayes4}
     \end{minipage}
\vspace{3.00mm} 
    \begin{minipage}[h]{0.5\linewidth}
        \centering
        \includegraphics[scale=0.34]{figbayes5.eps}
				\subcaption{Numbers of accident vs Slope}\label{figbayes5}
     \end{minipage}
\vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.34]{figbayes6.eps}
			\subcaption{Numbers of accident vs Speed}\label{figbayes6}
     \end{minipage}
\caption{Scatter plot of numbers of accident vs observed characteristic variables}
\label{figbayesx}
\end{figure}

\subsubsection{Question A}
Use the least squared method to obtain the model's parameter?
\subsubsection{Answer to question A}
The linear model shown in equation \eqref{eq37} can be simplified as follows
\begin{eqnarray}
 && No. of accident = 1\cdot \beta_0 + CS \cdot \beta_1 + DTV\cdot \beta_2 + Slope \cdot \beta_3 + Speed \cdot \beta_4 + \epsilon\label{bayes24}
\end{eqnarray}
Using the R code provided in the Appendix \ref{bayesian2}, results are obtained and shown in listing \ref{bayesls}.

\lstinputlisting[basicstyle=\ttfamily\scriptsize,float=h,frame=tb,caption=Results of least squared fit method,label=bayesls]{./Programs/IMP-bayesian/IMP-bayesian-linear-leastsquare.tex}

\subsubsection{Question B}
Use the Bayesian method to obtain the model's parameter? and give a discussion on the values of model's parameters in comparison with that values obtained by using the least squared method.
\subsubsection{Answer to question B}
Using the model presented in subsection \ref{bayline} with 10'000 numbers of iterations, values of models parameters can be obtained. As this process involves highly computational efforts, the R code shown in the Appendix \ref{bayesian2} is used.

\begin{figure}[ht]
 \begin{minipage}[h]{0.5\linewidth}
        \centering
        \includegraphics[scale=0.34]{figbayes7.eps}
				\subcaption{Intercept}\label{figbayes7}
     \end{minipage}
\vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.34]{figbayes8.eps}
			\subcaption{Condition state}\label{figbayes8}
     \end{minipage}
\vspace{3.00mm} 
    \begin{minipage}[h]{0.5\linewidth}
        \centering
        \includegraphics[scale=0.34]{figbayes9.eps}
				\subcaption{Daily traffic volume}\label{figbayes9}
     \end{minipage}
\vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.34]{figbayes10.eps}
			\subcaption{Slope}\label{figbayes10}
     \end{minipage}
     \vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.34]{figbayes11.eps}
			\subcaption{Speed}\label{figbayes11}
     \end{minipage}
     \vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.34]{figbayes12.eps}
			\subcaption{Error term}\label{figbayes12}
     \end{minipage}
\caption{Distribution of models parameters}
\label{figbayesxx}
\end{figure}

A summary of the mean values of model's parameters are shown in listing \ref{bayesconfident}. This listing also presents the values of quantiles at the boundaries 5\% and 95\%.

\lstinputlisting[basicstyle=\ttfamily\scriptsize,float=H,frame=tb,caption=Confident interval,label=bayesconfident]{./Programs/IMP-bayesian/IMP-bayesian-linear-results.tex}

As can be seen in Figure \ref{figbayesxx}, after 10'000 iterations of sampling, the distributions of model's parameters become normal distributed with their mean values almost equal to that values obtained by the least squared method in the previous section. This is exactly the highlight of difference between the traditional frequentist estimation technique and Bayesian technique for a simple example.

With respect to the statistical inferences on the values of model's parameters, it can be state that all characteristic variables are positively attributed to the occurrence of accident. However, contributions are different among them. It is likely that the condition state exposes to have the highest impact on the occurrence of accident as its value is 0.27, which is far more greater than that of other characteristic variables. The slope is the second important factor contributing to the occurrence of accident (value of 0.0255). Interesting, the DTV seems to have less impacts on the occurrence of accident for this road link. 
%
\section{Bayesian inference for a simple Weibull model} \label{bayweil}
In sub-section \ref{reliaweibl} of Chapter \ref{reliabilityc}, we learn that failure rate can be time-variant, and so does the reliability of an item. This type of time-variant failure and reliability models is commonly observed in reality. To address this time-variant characteristics of failure and reliability, a commonly used model is the Weibull proportional model.

In Weibull proportional model, the reliability function, density function, and failure rate, are shown in equations \eqref{eqreliability:10}, \eqref{eqreliability:11}, and \eqref{eqreliability:12}, respectively. For convenience of reading, they are rewritten as follows:

Reliability function
\[
 R(t) = {e^{ - {\alpha \cdot t^m}}}
\]
Probability density function (p.d.f)
\[
 f(t) = \alpha  \cdot m \cdot {t^{m - 1}} \cdot {e^{ - {\alpha \cdot {t}^m}}}
\]
Failure rate
\[
 \lambda (t) = \alpha  \cdot m \cdot {t^{m - 1}}
\]
where $\alpha $ and $m$ are scale and shape parameters, respectively. These are parameters that need to be estimated given data.
%
\subsection{Maximum likelihood estimation for the Weibull model} \label{mleweil}
In order to estimate the values of the models parameters $\alpha$ and $m$, the maximum likelihood estimation (MLE) can be used. The MLE is used extensively in frequentist statistic as a de-factor estimation method.

Let assume that the vector of parameter and data are $\theta$ and $\bar t$, respectively. The likelihood function has the following general form.
\begin{eqnarray}
 && L(t_1,t_2,\cdots,t_N,\theta) = \prod_{n=1}^N f(t_n,\theta) \label{mle01}
\end{eqnarray}
In this equation, $f$ represents the density function of a random variable $t$ and characterized by $\theta$. $n$ is the sample and $N$ is total numbers of samples.

The likelihood function \eqref{mle01} can be expressed in following form taking into consideration of data.
\begin{eqnarray}
 && L(t_1,t_2,\cdots,t_N,\theta) = \prod_{n=1}^N [f(t_n,\theta)]^{\delta_n}\cdot [1-F(t_n,\theta)]^{1-\delta_n} \label{mle02}
\end{eqnarray}
In equation \eqref{mle02}, $F(t_n,\theta)$ is the cumulative distribution function (c.d.f). Obviously, the second polynomial inside the product is the reliability function $(R=1-F)$ that we learn in Chapter \ref{reliabilityc}. $\delta_n$ is dummy variable taking its value of 1 when a sample (or item) has been observably failed and 0 otherwise.

Using the Weibull function for the p.d.f and the c.d.f, we can expressed equation \eqref{mle02} as follows
\begin{eqnarray}
 && L(\bar t,\theta) = \prod_{n=1}^N [\alpha\cdot m \cdot t_n^{m-1}\cdot exp(-\alpha\cdot t_n^m)]^{\delta_n}\cdot [exp(-\alpha\cdot t_n^m)]^{1-\delta_n} \label{mleweibull01}
\end{eqnarray}
The logarithmic form of the likelihood function will be
\begin{eqnarray}
 && ln[L(\bar t,\theta)] = \sum_{n=1}^N \delta_n \cdot \left[ ln(\alpha) + ln(m) + (m-1)ln(t_n)-\alpha\cdot t^m \right] + \sum_{n=1}^N (1-\delta_n)\cdot [-\alpha\cdot t^m]
   \label{mleweibull02}
\end{eqnarray}
In order to obtain models parameters, analytically, following derivative has to be solved for each parameter.
\begin{eqnarray}
 && \frac{\partial ln[L(\bar t,\theta)]}{\partial \theta_i} =0  \label{mleweibull03}
\end{eqnarray}
where the index $i$ represent the running index of models parameters. It is important to note that the scale parameter $\alpha$ can be further expressed as a function of characteristic variables in a linear form such as
\begin{eqnarray}
 && \alpha = \sum_{k=1}^K x_k\cdot \beta_k  \label{mleweibull04}
\end{eqnarray}
where, $k$ is the index of characteristic variable such as traffic volume, ambient temperature, etc and $\beta_k$ turns to be the models parameters associated with each characteristic variable $x_k$.
\subsection{Bayesian inference} \label{bayweil}
Using the Bayes law as shown in equation \eqref{bayes1}, we can express the posterior p.d.f of the model as follows
\begin{eqnarray}
 && p(\alpha,m|data)= \frac{L[\bar t|\alpha,m]\cdot p(\alpha)\cdot p(m)}{\int_0^{\infty}\int_0^{\infty}L[\bar t|\alpha,m]\cdot p(\alpha)\cdot p(m) d\alpha dm}   \label{mleweibull05}
\end{eqnarray}
where $p(\alpha)$ and $p(m)$ are prior distributions for $\alpha$ and $m$, respectively. As can be seen from this formula, the prior distributions for for $\alpha$ and $m$ infer that we can make use of knowledge that we might have known for these two parameters. This turns out to be extremely useful when dealing with small set of data. The choice of using which prior distribution to fit with $\alpha$ and $m$ can be varied depending on each modeler and knowledge gained from practice. This section leave out a detailed mathematical formulation of Bayesian inference for the Weibull model as it involves a great number of mathematic ingredients. For curious readers, who expect to fully capture the formulation of the model, research works of \cite{Dodson2006}, \cite{Nassar2005}, and \cite{Soliman2006} are recommended.
%
\subsection{Example: Underground pipeline deterioration}
A water distribution network of a mega urban city encompasses a large numbers of pipelines, which were laid underground. Pipelines were constructed in different time. Overtime, pipelines deteriorate, leakages or cracks might occur that lead to the lost of water in different part of the city. In some cases, pipelines were broken suddenly, resulting in a great amount of negative impacts to stakeholders (e.g. owners, users, directly affected public, and indirectly affected public). As pipelines were laid underground, it is not easy to perform frequent inspections. Moreover, inspection activities are also expensive. Most of data to date that is available for analysis is failure of pipelines over the entire city. The data is described in Table \ref{tblbayes3}.

\begin{table}
	\centering
	\caption{Data on pipeline failure $p$} \label{tblbayes3}
\begin{tabular}{|c|c|c|c|c|}
\hline
Segment & Time & Status & Water lost & Depth \\ 
 & (tus) &  & (normalized) & (cm) \\ 
\hline
1 & 156 & 1 & 0.5283 & 40 \\ 
\hline
2 & 1040 & 0 & 0.9542 & 45 \\ 
\hline
3 & 59 & 1 & 0.6773 & 47 \\ 
\hline
4 & 421 & 0 & 0.7279 & 41 \\ 
\hline
5 & 329 & 1 & 0.1481 & 42 \\ 
\hline
6 & 769 & 0 & 0.4236 & 47 \\ 
\hline
7 & 365 & 1 & 0.4951 & 47 \\ 
\hline
8 & 770 & 0 & 0.9857 & 42 \\ 
\hline
9 & 1227 & 0 & 0.1728 & 48 \\ 
\hline
10 & 268 & 1 & 0.0712 & 42 \\ 
\hline
11 & 475 & 1 & 0.5424 & 41 \\ 
\hline
12 & 1129 & 0 & 0.1673 & 43 \\ 
\hline
13 & 464 & 1 & 0.4240 & 50 \\ 
\hline
14 & 1206 & 0 & 0.9539 & 50 \\ 
\hline
15 & 638 & 1 & 0.2992 & 48 \\ 
\hline
16 & 563 & 1 & 0.0558 & 46 \\ 
\hline
17 & 1106 & 0 & 0.3183 & 42 \\ 
\hline
18 & 431 & 1 & 0.4762 & 44 \\ 
\hline
19 & 855 & 0 & 0.4101 & 42 \\ 
\hline
20 & 803 & 0 & 0.9394 & 45 \\ 
\hline
21 & 115 & 1 & 0.9573 & 43 \\ 
\hline
22 & 744 & 0 & 0.8402 & 46 \\ 
\hline
23 & 477 & 0 & 0.3166 & 47 \\ 
\hline
24 & 448 & 0 & 0.7257 & 46 \\ 
\hline
25 & 353 & 1 & 0.6178 & 41 \\ 
\hline
26 & 377 & 0 & 0.9832 & 45 \\ 
\hline
\end{tabular}
\end{table}

As can be seen in the table, there are 26 segments of pipelines in total. The column ``Time'' indicates the time in time units (tus) that each segment has been used since it was constructed. The column ``Status'' indicates the failure status of the pipe segment to date. The status receives value of 1 and 0. If the segment of pipe has failed, it receives value 1 and if not, the value is 0. The column ``Water lost'' indicates a normalized water lost recorded before the failure time for segments have failed and before inspection time for segments that have not failed. The last column indicates the depth in cm of pipe underneath the road surface.

\subsubsection{Question}
Use the MLE method and Bayesian method to estimate the models parameters assuming that the time to failure is dependent on water lost and depth. Interpret the meaning in the values of models parameters. 
\subsubsection{Answer}
Using the MLE method, following results are obtained. The results (listing \ref{mleweibull}) were obtained by using the R code in the Appendix \ref{bayesian4}

\lstinputlisting[basicstyle=\ttfamily\scriptsize,float=h,frame=tb,caption=Confident interval,label=mleweibull]{./Programs/IMP-bayesian/IMP-weibullMLE.tex}

From this result, the scale parameter and shape parameter of the distribution can be obtained \footnote{Scale and shape parameter of R function for Weibull model is described in \\https://stat.ethz.ch/R-manual/R-devel/library/survival/html/survreg.html}
\[
 \alpha = 2.7736+0.8820 \times mean(waterlost)+0.0873*mean(depth) 
 \]
\[
 \alpha = 2.7736+0.8820 \times 0.5466+0.0873*44.61538 = 7.1110381 
\]
\[
 m=1/Scale = 1/0.885=1.129402
\]
Using these parameters, the reliability function can be drawn 
\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{figbayes13.eps}
\caption{Reliability over time}\label{figbayes13}
\end{center}
\end{figure}

Using Bayesian method, the model's parameters can be summarized in distributions, with their means are the expected values of models parameters (Figure \ref{figbayesxxx}).

\begin{figure}[h]
\begin{center}
\includegraphics[width=450pt]{figbayesxxx.eps}
\caption{Distribution of model parameters}\label{figbayesxxx}
\end{center}
\end{figure}
\subsection{Conclusion}
Using Bayesian statistics, once can solve many models that cannot be solved with traditional frequentist statistics. These models allow us to have better prediction results given incomplete data set. Examples of using Bayesian statistics in the field of infrastructure asset management are not so many. Some of the recent research works in this direction are the works of \cite{Hong2006}, \cite{Gao2012}, \cite{Kobayashi2012}, \cite{Kobayashi2014}, and \cite{Lethanh2015b}.
\section{Bayesian network model}
Bayesian network modeling approach has recently gained its popularity in the fields of transportation engineering, structural analysis, and infrastructure asset management. It is a useful approach to quantify and measure risks related to transportation and infrastructure systems. Some examples of using Bayesian network models are with the works of \cite{Bayraktarli2011} for quantifying risks occurred in a mega city after an earthquake; the works of \cite{Bateni2007} for prediction of scour deterioration process of bridge piers; the works of \cite{Schubert2012}, \cite{Wang2013}, and \cite{Deublein2013} for predicting accident rates on a transportation network; and the works of \cite{Straub2010} for analysing structural reliability of an infrastructure object such as concrete beams and columns in a building or piers and decks of a bridge.

The principle concept to formulate a Bayesian network model is with the Bayes law that has been described in section \ref{bayesiantheorem} with equation \eqref{bayes1} and the graphical probabilistic method. There are many types of Bayesian network models, which really depend on the nature of observed data and assumptions of modelers.

This section aims to give a gentle introduction on Bayesian network modeling through a number of examples.
\subsection{General concept} \label{baynetconcept}
A Bayesian network (BN) is a graphical structure used to describe the interdependent of variables and parameters of a probabilistic model. The network with its notation $D=(V,E)$ is referred as a direct acyclic graph (DAG) where $V$ and $E$ representing vectors of finite sets of nodes and edges, respectively. The node $v\in V$ in a BN represent a random variable $X_v$, thus the entire node vector $V$ represents a set of random variables $\mathbf  X=X_1,X_2,\cdots,X_n$. The edge $e$ (sometimes refer to as arcs or links) connects pairs of nodes $X_i$ and $X_j$, i.e. $X_i \to X_j$, which represents the direct dependencies between variables.

In this context, the numbers of nodes are discrete and finite, thus the set of random variables is also discrete and finite. Each variable is characterized by a probability distribution. As nodes are undirectly or directly connected, they represent the dependencies. Thus, the probability of having one node (e.g. the event of a random variable to receive a value of x or y) is a conditional probability with its value dependent on other nodes. To represent the dependencies in a graphical model, two terminologies are used: ``parent nodes'' and ``child nodes''. Parent nodes are nodes being considered as sources that lead to the occurrence of child nodes. A node, can be either ``parent'' or ``child'' node. For example, in Figure \ref{baynet01}, node B is the parent node of node C but it is the child node of nodes A and D.

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.6]{baynet01.eps}
\caption{An example of a simplified BN model}\label{baynet01}
\end{center}
\end{figure}

To understand how a BN model is structured, let go through a simple example
\subsection{Introductory example - reliability of a bridge}
It is observed that over the last 15 years a bridge located in a mountainous regions had to close a numbers of times due to its condition that was unable to provide adequate level of services (LOS). The inadequate LOS is defined by manager as the levels of reliability and it is classified into three categories: ``low'' is when the reliability is less than 0.6; ``medium'' is when the reliability is between 0.6 and 0.8; and ``high'' is when the reliability is greater than 0.8. The reliability can be measured by performing stress tests.

Not providing adequate LOS might be due to latent process such as avalanche, rockfall, floods, and its manifest deterioration such as cracking of bridge decks, weakness of its bearing system, and bad quality of joints. The latent deteriorate process can be triggered by an earthquake, amount of snow accumulated in the starting zone of avalanches, and volume of run-off due to rains.

In each year, discrete values of these six variables were recorded.

\begin{itemize}
 \item \textbf{Avalanche}: the avalanche, recorded as ``yes'' if there was at least one occurrence of avalanches in that year and the avalanches came to contact with the bridge, and ``no'' if there was no avalanche occurred. 
 \item \textbf{Rockfall}: the rockfall, recorded as ``yes'' if there was at least one occurrence of rockfalls in that year and the rockfalls came to contact with the bridge, and ``no'' if there was no rockfall occurred. 
 \item \textbf{Floods}: the floods, recorded in three categories: ``low'', ``medium'', and ``high'' indicating levels of floods happened.
 \item \textbf{Cracks of bridge decks}: the cracks appeared on bridge decks indicates serious problems on the deterioration of the decks. The cracks were recorded in three categories: ``small'' is when the percentage of cracks is less than 5\% in the total amount of bridge decks areas; ``moderate'' is when the percentage of cracks is between 5\% and 10\% and ``high'' is when the percentage of cracks is higher than 10\%.
 \item \textbf{Weakness of the bearing system}: the bearing system is made of rubber pads. The condition of the system is recorded as ``good'' and ``bad''.
 \item \textbf{Quality of joints}: Quality of joints were recorded as ``good'' and ``bad'', which indicates the condition state of joint. It is good when the joint has no defect. It is bad when either top or bottom layers or both of the two layers are in bad condition.
 \item \textbf{Earthquake}: the earthquake is recorded as ``yes'' or ``no''.
 \item \textbf{Snow}: the snow is recorded as ``low'' and ``high'' depending on the amount of snow accumulated on the starting zone of avalanches
 \item \textbf{Rain}: the rain is recorded as ``low'', ``medium'', and ``high'' depending on the volume of run-off recorded at the weather station nearby the bridge.
 \item \textbf{Daily traffic volume (DTV)}: the daily traffic volume was recorded in numbers but it can also be categories into 3 discrete scales: ``low'', ``medium'', and ``high''.
\end{itemize}

The reliability of the bridge is dependent on both latent process and manifest process. Therefore, by study the structure of this system, it might be useful to calculate the reliability of bridge more accurate and eventually beneficial to come up with better management for the bridge.

\subsection{Graphical representation and values of nodes}
The way the variables were recorded in the inspection data, and more in general in two categories (latent and manifest) that each variable belongs to, infers the interdependency among them. Some of them have direct relationship such as the reliability of the bridge is directly related to a rockfall; and some of them have indirect relationship such as the reliability of the bridge is indirectly related to the volume of run-off.

Both these kinds of relationship can be graphically modeled using a direct graph as earlier mentioned in section \ref{baynetconcept}. In the graph, nodes ($V$) and edges ($E$) are used to systematically represent the relationship. 
In a very simple representation, a partial graphical representation of the network looks like the one shown in Figure \ref{baynet02}.

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.6]{baynet02.eps}
\caption{Partial representation of the Bayesian network for the bridge - Latent process}\label{baynet02}
\end{center}
\end{figure}

This figure infers that the reliability of the bridge is dependent on two factors: avalanche and rockfall. Those factors are dependent on natural hazard such as earthquake. Also in the figure, there are conditional probabilities table (cpt). Values shown in the tables indicate the probabilities of that variables (or event) receiving value ``yes'' or ``no''. For example, if there is a earthquake, there is 40\% chance that there will be rockfall. Certainly, the probability for the bridge not providing adequate LOS is conditional dependent on the probability of occurring avalanches or rockfall, which, in turn, dependent on the probability of earthquake occurrence.

With this rule of conditional probability in a chain, the joint probability of all nodes in this BN can be computed by using following formula.
%
\begin{eqnarray}
 && Pr(E,A,R,B)=Pr(E)\times Pr(A|E)\times Pr(R|E)\times Pr(B|E,A,R)   \label{baynetformula1}\\
 && Pr(E,A,R,B)=Pr(E)\times Pr(A|E)\times Pr(R|E)\times Pr(B|A,R)   \label{baynetformula2}
\end{eqnarray}
where, $E,A,R$, and $B$ stand for Earthquake, Avalanche, Rockfall, and Bridge.

Let say, we are now interested in calculating the likelihood that Avalanche will causes the bridge to be in condition that it will not provide an adequate LOS. Using the Bayes rule as shown in \eqref{bayes1}, we can formulate the posterior probability for the avalanche.
\begin{eqnarray}
 && Pr(A=yes|B=yes)= \frac{Pr(A=yes,B=yes)}{Pr(B=yes)} = \frac{\sum_{E,R}Pr(E,A=yes,R,B=yes)}{Pr(B=yes)} \label{baynetformula3}
\end{eqnarray}
The nominator $\sum_{E,R} Pr(A=yes,B=yes,E)$ infers that in order to calculate this probability, we have to consider the probability of earthquake occurrence and the probability of rockfall as well. Using equation \eqref{baynetformula2}, the nominator in equation \eqref{baynetformula3} becomes
\begin{eqnarray}
 && \sum_{E,R}Pr(E,A=yes,R,B=yes)= \nonumber \\ 
 && \hspace{1cm}= Pr(E=yes,A=yes,R=yes,B=yes)    \label{baynetformula4} \\
 && \hspace{1cm}+ Pr(E=no,A=yes,R=yes,B=yes)   \label{baynetformula5} \\
 && \hspace{1cm}+ Pr(E=yes,A=yes,R=no,B=yes)   \label{baynetformula6} \\
 && \hspace{1cm}+ Pr(E=no,A=yes,R=no,B=yes)   \label{baynetformula7} \\
 && \hspace{1cm}= Pr(E=yes)\times Pr(A=yes|E=yes)\times Pr(R=yes|E=yes)\times Pr(B=yes|A=yes,R=yes)   \label{baynetformula8} \\
 && \hspace{1cm}+ Pr(E=no)\times Pr(A=yes|E=no)\times Pr(R=yes|E=no)\times Pr(B=yes|A=yes,R=yes)   \label{baynetformula9} \\
 && \hspace{1cm}+ Pr(E=yes)\times Pr(A=yes|E=yes)\times Pr(R=no|E=yes)\times Pr(B=yes|A=yes,R=no)   \label{baynetformula10} \\
 && \hspace{1cm}+ Pr(E=no)\times Pr(A=yes|E=no)\times Pr(R=no|E=no)\times Pr(B=yes|A=yes,R=no)   \label{baynetformula11} 
\end{eqnarray}
%
\begin{eqnarray}
 && \sum_{E,R}Pr(E,A=yes,R,B=yes)= \nonumber \\ 
 && \hspace{1cm}= 0.01\times 0.70 \times 0.40 \times 0.60   \nonumber \\
 && \hspace{1cm}+ 0.99 \times 0.10 \times 0.05 \times 0.60   \nonumber \\
 && \hspace{1cm}+ 0.01 \times 0.70 \times 0.60 \times 0.20    \nonumber \\
 && \hspace{1cm}+ 0.99 \times 0.10 \times 0.95 \times 0.20   \nonumber \\
 && \hspace{1cm} = 0.0243 \nonumber 
\end{eqnarray}
Similarly using the \eqref{baynetformula1} for extending the formulation of the denominator of equation \eqref{baynetformula2}, its value is obtained
%
\begin{eqnarray}
 && Pr(B=yes)= \sum_{E,A,R,B} Pr(E,A,R,B=yes)=0.0804375
\end{eqnarray}
thus the likelihood of avalanche to be occurred if the bridge is not providing adequate LOS is
\begin{eqnarray}
 && Pr(A=yes|B=yes)= \frac{Pr(A=yes,B=yes)}{Pr(B=yes)} =  \frac{0.0243}{0.0804375}=0.302097 \nonumber
\end{eqnarray}
%
From this example, it is clear that the joint probability of variables in a BN model can be computed based on the rule of conditional probability. In general mathematical expression, that rule is described as follows
%
\begin{eqnarray}
 && Pr(V)=\prod_{v\in V} Pr(v|pa(v)) \label{baynetformula12} 
\end{eqnarray}
The notation $pa$ infers the parent nodes of node $v$ and $Pr(v|pa(v))$ is a function defined on $(v,pa(v)$. This function satisfies the condition that the sum of probability over a space equals to 1, i.e. $\sum_{v*}Pr(v=v*|pa(v))=1$. It can be seen that a set values of $Pr(v|pa(v))$ is actually the conditional probabilities table (the cpt in short). In this view, it can be said that a BN is a complex stochastic model built up by forming together simple components. In other words, it is referred as hierarchical Bayesian model.

The graph and result of this example can be easily drawn and calculated by hand. For convenient of readers, R code is provided in Appendix \ref{bayesian5}

The above example is a simplest assumption possible that we can think of. In the example, only latent deterioration process is considered. However, in practice, latent deterioration is not only the cause of deterioration. Manifest deterioration should also be considered in calculating the reliability of the bridge. In addition, both latent and manifest processes can be further expanded (like the event tree). A more realistic BN for the same example can be visualized in Figure \ref{baynet03}.
\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.6]{baynet03.eps}
\caption{Partial representation of the Bayesian network for the bridge - Latent process}\label{baynet03}
\end{center}
\end{figure}

Figure \ref{baynet03} shows that conceptually a BN model can be expanded to include many layers of variables possible. Also, values describing each variable can be also continuous and be assumed to follow a parametric distribution (e.g. Poisson, Weibull, Gamma) depending on the behaviors of data.

The calculation for this BN becomes more difficult than the earlier one as the size of the cpt exponentially growth with numbers of variables. The R code in Appendix \ref{bayesian6} is used to generate the BN shown in Figure \ref{baynet03}. It includes also a random generated cpt for each node of the BN model and perform calculation of marginal probability for each node. Noted that in this extended example, the reliability of the bridge is recorded in three categories ``low'', ``medium'', and ``high'' instead of of only two states in original example.
\subsection{Estimating the cpt from data}
In earlier section, the cpt is presented along with the graph of the BN models. Values of the properties in the cpt can be estimated from various source of information such as expert opinions, simulation, or data. Estimation the cpt from data is common in most cases. Typically, data is stored in various extension files (e.g. txt, tex, csv) or in database systems (e.g. MySQL, MsAccess, PostgreSQL). An example of how data looks like for the above example is listed in Table \ref{tblbayes4a} and Table \ref{tblbayes4b}.

\begin{table}
	\centering
	\caption{Simplified observed data} \label{tblbayes4a}
\begin{tabular}{|c|c|c|c|}
\hline
Earthquake & Avalanche & Rockfall & Bridge \\ 
\hline
yes & yes & no & yes \\ 
\hline
yes & no & no & yes \\ 
\hline
no & no & no & yes \\ 
\hline
no & no & no & no \\ 
\hline
no & no & no & no \\ 
\hline
no & no & yes & no \\ 
\hline
yes & no & no & no \\ 
\hline
no & no & no & no \\ 
\hline
no & yes & yes & no \\ 
\hline
yes & no & yes & no \\ 
\hline
no & yes & no & yes \\ 
\hline
no & no & no & no \\ 
\hline
no & yes & yes & no \\ 
\hline
no & yes & no & yes \\ 
\hline
no & no & no & yes \\ 
\hline
yes & yes & yes & no \\ 
\hline
no & yes & no & yes \\ 
\hline
no & no & yes & no \\ 
\hline
no & yes & yes & no \\ 
\hline
yes & no & yes & no \\ 
\hline
no & no & no & yes \\ 
\hline
no & no & yes & no \\ 
\hline
yes & yes & yes & yes \\ 
\hline
yes & no & no & yes \\ 
\hline
yes & yes & yes & no \\ 
\hline
no & no & no & no \\ 
\hline
no & no & no & yes \\ 
\hline
yes & no & yes & yes \\ 
\hline
yes & yes & no & yes \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{Extended observed data} \label{tblbayes4b}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|}
\hline
Earthquake & Snow & Rain & DTV & Aval & Rockfall & Flood & Cracks & Bearing & Joint & Bridge \\ 
\hline
yes & high & high & low & yes & no & no & moderate & good & yes & high \\ 
\hline
yes & low & high & high & no & no & yes & moderate & good & no & high \\ 
\hline
no & low & medium & high & no & no & no & high & good & no & high \\ 
\hline
no & low & low & low & no & no & no & high & good & no & medium \\ 
\hline
no & high & low & low & no & no & no & high & bad & yes & low \\ 
\hline
no & high & low & high & no & yes & yes & high & bad & yes & low \\ 
\hline
yes & high & medium & low & no & no & yes & moderate & good & yes & low \\ 
\hline
no & low & low & low & no & no & yes & moderate & good & no & medium \\ 
\hline
no & low & low & medium & yes & yes & yes & high & bad & no & medium \\ 
\hline
yes & low & high & medium & no & yes & no & moderate & bad & yes & medium \\ 
\hline
no & low & low & medium & yes & no & yes & moderate & bad & no & high \\ 
\hline
no & high & medium & low & no & no & yes & moderate & good & yes & low \\ 
\hline
no & low & high & high & yes & yes & no & small & bad & no & medium \\ 
\hline
no & high & medium & low & yes & no & yes & high & good & yes & high \\ 
\hline
no & low & low & high & no & no & yes & moderate & bad & yes & high \\ 
\hline
yes & high & low & low & yes & yes & no & small & good & yes & low \\ 
\hline
no & high & low & low & yes & no & yes & small & good & no & high \\ 
\hline
no & low & high & high & no & yes & no & moderate & good & no & low \\ 
\hline
no & low & low & low & yes & yes & yes & moderate & bad & yes & medium \\ 
\hline
yes & high & low & high & no & yes & yes & small & bad & yes & low \\ 
\hline
no & low & medium & low & no & no & yes & small & good & yes & high \\ 
\hline
no & high & medium & medium & no & yes & no & small & bad & yes & low \\ 
\hline
yes & high & low & medium & yes & yes & no & moderate & bad & no & high \\ 
\hline
yes & high & medium & high & no & no & yes & moderate & bad & yes & high \\ 
\hline
yes & high & medium & medium & yes & yes & no & high & good & yes & medium \\ 
\hline
no & high & high & high & no & no & yes & high & bad & yes & low \\ 
\hline
no & high & low & low & no & no & no & small & good & yes & high \\ 
\hline
yes & high & high & medium & no & yes & no & high & bad & no & high \\ 
\hline
yes & low & high & high & yes & no & no & high & bad & no & high \\ 
\hline
\end{tabular}
\end{table}

Table \ref{tblbayes4a} shows a simplified data that includes only 4 variables (Earthquake, Rockfall, Avalanche, Bridge) and for each variable, there is only two discrete value ``yes'' and ``no''. The conditional probability for each variable can be computed using the data shown in this table. Since values of variables are discrete, one way to do it is to use following equation

\begin{eqnarray}
 && Pr(B=yes|R=yes)=\frac{Pr(B=yes,R=yes)}{Pr(R=yes)} \label{baynetformula13} \\
 && \hspace{2cm} \frac{\text{Number of observations for which B=yes and R=yes}}{\text{Number of observations for which R=yes}}
\end{eqnarray}
This equation yields exact the conditional probability for this event to occur. This is corresponding to the traditional $frequentist$ and $maximum$ $likelihood$ estimates.

The R codes for computing the cpt based on these two tables are given in Appendix \ref{bayesian5} and \ref{bayesian6} and the estimated cpt for the data shown in Table \ref{tblbayes4a} and Table \ref{tblbayes4b} are given in Appendix \ref{Bndata1} and \ref{Bndata2}, respectively.
\subsection{Forming a BN model from data and the curse of dimensionality}
In the above example, it is given that the structure of the BN model is known. That is we know the relationship between nodes in the graphical model. In many practical situations, data is collected that includes many different types of variables and attributes that we do not precisely know their relationship. Fortunately, given the data, it is possible to learn from the behaviors of data in order to create a set of BN models that suit the data the most. Following example of data on accident risks illustrate the idea of how to form a BN model from data.

A city transportation department wants to understand the accident risks that occurs on the city
road network when three intervention strategies (IS)\footnote{ Intervention strategies can be the setup of traffic flows or the installation of traffic signals
} is pilot tested. These pilot tests can be
executed in different parts of the city network. The effectiveness of these tests are measured
by observing the outcome as the numbers of accidents occurs in two consecutive weeks. A
description of the data is given in following table \ref{tblbayes5}.

\begin{table}
	\centering
	\caption{Data on pipeline failure $p$} \label{tblbayes5}
\begin{tabular}{|c|c|c|c|}
\hline
Time & IS & week 1 & week 2 \\ 
\hline
day & 1 & 5 & 6 \\ 
\hline
day & 1 & 7 & 6 \\ 
\hline
day & 1 & 9 & 9 \\ 
\hline
day & 1 & 5 & 4 \\ 
\hline
day & 2 & 9 & 12 \\ 
\hline
day & 2 & 7 & 7 \\ 
\hline
day & 2 & 7 & 6 \\ 
\hline
day & 2 & 6 & 8 \\ 
\hline
day & 3 & 14 & 11 \\ 
\hline
day & 3 & 21 & 15 \\ 
\hline
day & 3 & 12 & 10 \\ 
\hline
day & 3 & 17 & 12 \\ 
\hline
night & 1 & 7 & 10 \\ 
\hline
night & 1 & 8 & 10 \\ 
\hline
night & 1 & 6 & 6 \\ 
\hline
night & 1 & 9 & 7 \\ 
\hline
night & 2 & 7 & 6 \\ 
\hline
night & 2 & 10 & 13 \\ 
\hline
night & 2 & 6 & 9 \\ 
\hline
night & 2 & 8 & 7 \\ 
\hline
night & 3 & 14 & 9 \\ 
\hline
night & 3 & 14 & 8 \\ 
\hline
night & 3 & 16 & 12 \\ 
\hline
night & 3 & 10 & 5 \\ 
\hline
\end{tabular}
\end{table}
The network has its primary property as node as graphically represented in Figure \ref{baynet04}. The cell
in the table represents each entry or node of the network. Each entry contains information
attached with each node (e.g. day or night, different numbers of accidents).

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.6]{baynet04.eps}
\caption{Graphical representation of the network for accidents}\label{baynet04}
\end{center}
\end{figure}
In this simple network, the input nodes are the time (day or night) and the intervention
strategies (IS), and the outcome are the numbers of accident occurs in each consecutive
weeks. The infrastructure manager wants to identify the probability that the accident occurs
due to time and intervention strategies. In addition, as the week 2 is after week 1, the node
representing week 1 can also become a parent node of week 2, i.e. the likelihood of accidents
occurs on week 2 is dependent also on the numbers of accidents that have had occurred in
week 1.

It is assumed that the joint probability for this problem is Gaussian distribution
(normal distribution). The Bayesian network allows to compute such joint probability for
every combination of nodes. For example, the joint probability of combination
$(Time\to IS \to W1 \to W2)$ or $(Time \to W1)$ or $(IS \to W1 \to W2)$.

Following graphs show the outcome of running a Bayesian network for this example (the R code
used to compute the example is given in Appendix \ref{bayesian7}).

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.45]{baynet05a.eps}
\caption{Possible BN networks (1)}\label{baynet05a}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.45]{baynet05b.eps}
\caption{Possible BN networks (2)}\label{baynet05b}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.45]{baynet05c.eps}
\caption{Possible BN networks (3)}\label{baynet05c}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.45]{baynet05d.eps}
\caption{Possible BN networks (4)}\label{baynet05d}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.45]{baynet05e.eps}
\caption{Possible BN networks (5)}\label{baynet05e}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.45]{baynet05f.eps}
\caption{Possible BN networks (6)}\label{baynet05f}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale =0.6]{baynet06.eps}
\caption{The optimal network structure}\label{baynet06}
\end{center}
\end{figure}

As can be seen from the figure Figure \ref{baynet05a} to Figure \ref{baynet05f}, there are many possible combination of nodes to form a network. The arrow in each of the graph represents the dependency. In total, with only 4 nodes, there are 144 networks being analyzed.

Table \ref{tblbayes6} lists a possible numbers of networks that can be formed given a certain numbers of nodes with their lower bound and upper bound values \citep{Heckerm1996}.

\begin{table}
	\centering
	\caption{Numbers of networks} \label{tblbayes6}
\begin{tabular}{|c|c|}
\hline
$\#Node$ & $\#networks$ \\ 
\hline
1 & 1 \\ 
\hline
2 & 2-3 \\ 
\hline
3 & 12-25 \\ 
\hline
4 & 144-543 \\ 
\hline
5 & 4'800-29'281 \\ 
\hline
6 & 320'000-3'781'503 \\ 
\hline
\end{tabular}
\end{table}

In these graphs, the statistic score $S(D)$ are displayed along with their rescaling score. The score is computed for every combination of nodes based on following equation
\begin{eqnarray}
 && S(D)=p(D,data)=p(data|D)\cdot p(D)
\end{eqnarray}

This score is used to assess the likelihood for that combination to be realized with certain probability. In this example, the network $IS \to W1 \to W2$ (Figure \ref{baynet06}) will be most likely to be realized with highest probability. It means that the numbers of accidents that will occurs in week 2 will have high correlation with the intervention strategies and the numbers of accidents in week 1.

The above example discusses a general Bayesian network frame, in which, data model can be formulated with different combination and interaction. As can be shown, there are many scenarios that are not necessary to be analyzed for the final result. For example, the scenario that the arrow is only connecting between “time” and “week 2”. This scenario is not likely to give a good answer as it is known from engineering background that the “IS” should be involved. By using engineering background, the numbers of combinations between nodes can be reduced significantly.

\bibliographystyle{plainnat}
\bibliography{reference}

\pagebreak

\begin{subappendices}
% % \label{appendix3}
\section{Example into Bayesian thinking}\label{bayesian1}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/IMP-priordistribution.r}
 \pagebreak
\section{Bayesian estimation - Linear model}\label{bayesian2}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/IMP-bayesian-linear.r}
  \pagebreak
  \section{Maximum likelihood estimation estimation - Weibull model}\label{bayesian4}
  \lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/IMP-mle-weibull.r}
 \pagebreak
\section{Bayesian estimation - Weibull model}\label{bayesian3}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/IMP-bayesian-weibull.r}
\pagebreak
\section{Bayesian network - bridge reliability (simplified)}\label{bayesian5}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/bayesiannetwork-bridgereliabilitya.r}
\pagebreak
\section{Bayesian network - bridge reliability (Extended)}\label{bayesian6}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/bayesiannetwork-bridgereliabilityb.r}
\pagebreak
\section{cpt of the BN simplified example}\label{Bndata1}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/IMP-BNsimplifiedexample.tex}
\pagebreak
\section{cpt of the BN extended example}\label{Bndata2}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/IMP-BNextendedexample.tex}
\pagebreak
\section{Bayesian network - Learning structure of a BN from data}\label{bayesian7}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-bayesian/bayesiannetwork-structurelearning.r}
\end{subappendices}
% 
% 
% 
%

### 2019-05-05 InfrastructureMaintenance/Chapters/chapter4.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Parameter based Intervention Strategies - Markov Model}
\label{markov} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head

\abstract{XXX.}
\section{Section Heading}

\section*{Problems}
\addcontentsline{toc}{section}{Problems}
%
% Use the following environment.
% Don't forget to label each problem;
% the label is needed for the solutions' environment
\begin{prob}
\label{prob1}
A given problem or Exercise is described here. The
problem is described here. The problem is described here.
\end{prob}

\begin{prob}
\label{prob2}
\textbf{Problem Heading}\\
(a) The first part of the problem is described here.\\
(b) The second part of the problem is described here.
\end{prob}

\input{referenc}

### 2019-05-05 InfrastructureMaintenance/Chapters/fault-event-tree.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Event Tree and Fault Tree Analysis}
\label{faulteventreechapter} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Bryan T. Adey and Nam Lethanh}
\section{Event tree analysis - Theory}
\subsection{Definition}
Event tree analysis (ETA) is a way to analyse and display different discrete
scenarios, their corresponding probability of occurrence and the resulting
consequences. ETA uses Boolean logic to analyse a chronological sequence of
events and estimate the probability of their occurrence. A key component of event
tree analysis is the event tree. An event tree is built from a starting event and
branches at each subsequent event based on the values of key parameters these key
parameters were intensity measures. When the event tree is complete, it is a
logical and visual representation of the set of scenarios that can occur, i.e.
multiple possible futures. An example is shown in Figure \ref{figeventfault:1}, the
initiating event is a 100-year precipitation, which may result in the occurrence
of a flood with water depths over 30 cm, under 30 cm, or not at all. In this
example, the flood can lead to different damage states related to an
infrastructure element, which can result in interruptions to traffic flow on the
road network. The consequences of the scenarios can then be quantified. Also the
heavy rain itself can cause an interruption.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{faulteventtree-1.eps}
\caption{Simplified event tree for a flood risk analysis}\label{figeventfault:1}
% \end{center}
\end{figure}
ETA is often referred to as having forward logic as it starts from an initiating
event and ends at a consequence. It is often used to analyze the risks associated
with systems, and the consequences of functioning or failed systems given that an
event(s) has (have) occurred. Simply the construction of an event tree can often
be useful in decision situations as it brings to light many scenarios that
decision makers otherwise would not have considered.

When conducting an ETA it is necessary to first determine the starting events
that are to be considered and then to think through each of the next possible
events to be considered. Each chain of events from initiating events to the end
event (or consequence) is referred to as a path, and may be thought of as a
scenario.

At each node in the event tree, the tree branches. There are often only two
branches that emanate from a node but it is possible that no split occurs or that
many splits occur. It is, however, of utmost importance to ensure that the split
represents mutually exclusive possibilities, i.e. if one possibility happens the
other cannot happen. The end events are often referred to as consequences.

The construction of an event tree requires knowledge about the system.
\subsection{Steps}

The basic steps to conduct an event tree analysis are:
\begin{enumerate}
\item Define static system: Determine what you would like to model and what is not to
be considered in the model
\item Define dynamic system: Determine what you would like to model over time. For
example, what hazards are of particular concern, e.g. flooding, what physical
things may happen, e.g. bridge collapse, what consequences might occur, e.g.
horrendous increases in travel times due to traffic jams
\item Identify specific initiating events: For example, heavy rain fall.
\item Identify specific intermediate events: For example, overtopping of a bridge
\item Construct the event tree
\item Calculate the probabilities of occurrence of each event: One possibility is
using fault tree analysis.
\item Calculate the probability of occurrence of each scenario or path:
\item Calculate the overall risk:
\item Evaluate the acceptability of the risk associated with each scenario and
overall:
\end{enumerate}
%
Due to the almost infinite number of ways to represent reality and the almost
infinite number of events that can occur, an appropriate system representation
needs to be developed. A good starting point is often the system representation
used to determine that there was a problem. It needs, however, to be verified
that this representation is adequate for the evaluation of the candidate
strategies. The necessary detail to be used depends on the specific question, the
strategies to be evaluated and the level of detail desired.
\subsection{Example in literature}
Please read the research work of \cite{Beim1997}, who constructed a event
tree analysis for a lock gate system, as part of the script.
\section{Event tree analysis - Example}
\subsection{Question A}
You are the road manager responsible for a particularly risky section of road.
It consists of a bridge that crosses a river and is just below a steep
embankment, upon which it is known that avalanches sometimes occur. This is
illustrated in Figure \ref{figfaultevent2x}.
%
\begin{figure}
 \begin{minipage}[h]{0.5\linewidth}
        \centering
        \includegraphics[scale=0.95]{faulteventtree-31.eps}
				\subcaption{Flood event}\label{faulteventtree-31}
     \end{minipage}
\vspace{3.00mm}
    \begin{minipage}[h]{0.5\linewidth}
       \centering
       \includegraphics[scale=0.95]{faulteventtree-32.eps}
			\subcaption{Avalanche event}\label{faulteventtree-32}
     \end{minipage}
\caption{Illustration of risks}
\label{figfaultevent2x}
\end{figure}
%
You suspect that there are substantial infrastructure related risks due to both
flooding and avalanches but would like to quantify the risks. Develop an event
tree to allow you to do this.
\subsection{Answer A}
One of the first things to do when developing an event tree to allow you to
estimate the infrastructure related risks due to flooding and avalanches is to
determine the what will cause the branching in the event tree\footnote{The answer
is an adapted version of that used in \cite{Adey2009}}. The event tree is a part of the system
representation\footnote{The system representation is a model of the relevant part
of reality used for the evaluation and consists of all possible realizations of
stochastic processes within the investigated time period.}. The event tree in
this case should have sufficiently good representations of the structures,
hazards, and consequences, as well as the interaction between them so that it can
be reasonably certain that there is an appropriate understanding of the system
and that the risks. The event tree should consist of all possible things that can
happen in a specified period of time, i.e. scenarios. In the event tree a
scenario is represented by a path in the event tree. In the event tree proposed
here the paths are divided into five components, 1) event occurrence, 2) effect
of event, 3) physical changes in infrastructure, 4) direct consequences, and 5)
indirect consequences. An example for a bridge exposed to flooding and avalanches
is shown in Figure \ref{figeventfault:3}, assuming that flood and avalanche events are
mutually exclusive.
\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-13.eps}
\caption{Event tree with mutually exclusive flood and avalanche
events.}\label{figeventfault:3}
\end{center}
\end{figure}
In this event tree the hazard consists of the event and the effects on the
bridge object that are considered to have consequences. The effects on the bridge
are represented by two levels of the key parameters, water depth \textit{$>$= a
}and water depth \textit{$<$ a}, where \textit{a }is the bridge clearance, and
snow depth \textit{$>$= b }and snow depth \textit{$<$ b}, where \textit{b }is the
depth of snow on the bridge.

The consequence portion of the event tree is composed of the non-monetarisable
and monetarisable direct and indirect consequences. For each level of the key
parameter, assumptions are made with respect to the non-monetarisable direct
consequences that can occur, e.g. the physical behavior of the structure when the
specified level of the key parameter(s) occurs. The effect on the structure
should be described with the fewest possible parameters. Assumptions are made
with respect to the monetarisable direct consequences that can occur for each
non-monetarisable direct consequence, e.g. the fatalities related to the collapse
of the bridge. Assumptions are made with respect to the monetarisable indirect
consequences that can occur for each monetarisable direct consequence, e.g. the
additional user traffic time due to a deviation. The monetarisable consequences
can be attributed to those that will carry the consequences if failure occurs and
can be grouped by consequences type.
In the event tree (Figure \ref{figeventfault:3}), each effect has two possible
non-monetarisable direct consequences (\textit{nmDC{\tiny 1}, nmDC{\tiny 2}}).
For example, when the water depth during a flood is \textit{$>$= a}, than the two
possible failure modes are the wash out of the abutments (\textit{nmDC{\tiny 1}})
and the wash out of the columns (\textit{nmDC{\tiny 2}}). The monetarisable
direct consequences deterministically follow the non monetarisable direct
consequences (\textit{nmDC{\tiny 1}}). The monetarisable indirect consequences
are branched into the monetarisable indirect consequences that would occur if the
abutments were washed out during winter with low traffic volumes
(\textit{mIC{\tiny 1}}) and those that would occur if the abutments were washed
out during summer with high traffic volumes (\textit{mIC{\tiny 2}})
\subsection{Question B}
Construct an event tree for a road segment and a bridge that may be affected by
a flood or an avalanche, but not at the same time.
\subsection{Answer B}
The branches in the event tree are comprised of a chain of possible events and
monetarisable direct and indirect consequences (Fig. \ref{figeventfault:3} and Fig. \ref{figeventfault:4}). A slightly more
detailed example of some of the avalanche scenarios is given in Table
\ref{tbleventfault:1}, where the effect on the physical object is listed simply as level.
It is easy to imagine how detailed this can become.
\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-14.eps}
\caption{Event tree with mutually exclusive flood event.}\label{figeventfault:3}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-15.eps}
\caption{Event tree with mutually exclusive avalanche event.}
\label{figeventfault:4}
\end{center}
\end{figure}

\begin{table}
	\centering
	\caption{Examples of scenarios} \label{tbleventfault:1}
\begin{tabular}{|c|c|c|c|}
\hline
Nr. & Effect & Non-monetarisable direct consequences & Monetarisable direct consequences \\ 
\hline
1 &  & Mild deformation of road surface & Level 1 \\ 
\cline{1-1}\cline{4-4}
2 &  &  & Level 2 \\ 
\cline{1-1}\cline{3-4}
3 &  & Severe deformation of road surface & Level 1 \\ 
\cline{1-1}\cline{4-4}
4 & On column level 1 &  & Level 2 \\ 
\cline{1-1}\cline{3-4}
5 &  & Collapse of object & Level 2 \\ 
\cline{1-1}\cline{4-4}
6 &  &  & Level 3 \\ 
\hline
- & - & - & - \\ 
- & - & - & - \\ 
- & - & - & - \\ 
\hline
31 &  & Mild displacement of deck & Level 1 \\ 
\cline{1-1}\cline{4-4}
32 &  &  & Level 2 \\ 
\cline{1-1}\cline{3-4}
33 &  & Severe displacement of deck & Level 1 \\ 
\cline{1-1}\cline{4-4}
34 & Horizontal loading of deck level 3 &  & Level 2 \\ 
\cline{1-1}\cline{3-4}
35 &  & Removal of superstructure & Level 2 \\ 
\cline{1-1}\cline{4-4}
36 &  &  & Level 3 \\ 
\hline
\end{tabular}
\end{table}
\subsection{Question C}
As a road manager you are considering the construction of avalanche protection
to reduce the infrastructure related risk.

You know that with respect to the probabilities of occurrence of events that
\begin{itemize}
\item if you do nothing,the probability of occurrence of an avalanche decreases from
0.5 in 2009 to 0.3 in 2029 due to ever decreasing snow fall.
\item If you construct avalanche protection, the probability of occurence of an
avalanche will be 0 for the next 10 years but will increase after that
\end{itemize}
You know that the conditional probabilities of the effects on the loading of the
column, the horizontal loading of the superstructure and the vertical loading of
the superstructure are constant and as shown in Table \ref{tbleventfault:2}. You know that
conditional probability of physical changes due to horizontal loading of the
columns and deck increases by 0.001 per year to take into account normal
deterioration (Table \ref{tbleventfault:3} and Table \ref{tbleventfault:4}) and due to vertical
loading of deck is 1 and constant (Table \ref{tbleventfault:5}). You also know that the due
to changes in traffic that the conditional probabilities of consequences are as
shown in Table \ref{tbleventfault:6}. You have also estimate the consequences of each
consequence level to be those as shown in Table \ref{tbleventfault:7}.

Is it worth it if only a 20 year horizon is analysed?  (Use a discount factor of
2\%. Assume that the probability of multiple failures of the same infrastructure
object are negligible. Show the yearly costs and benefits as well as the totals.
For simplicity, assume that there is only one level of monetarisable consequences
per level of physical damage.)

\begin{table}
	\centering
	\caption{Conditional probabilities of the effects} \label{tbleventfault:2}
\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Year} & \multicolumn{3}{c|}{Loading of the column} & \multicolumn{3}{m{2.5cm}|}{\centering Horizontal loading of the superstructure} & \multicolumn{3}{m{2.5cm}|}{\centering Vertical loading of the superstructure} \\ 
\cline{2-10}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{Level 1} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 3} & \multicolumn{1}{c|}{Level 1} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 3} & \multicolumn{1}{c|}{Level 1} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 3} \\ 
\hline
\multicolumn{1}{|c|}{2009} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2010} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2011} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2012} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2013} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c|}{-} \\ 
\multicolumn{1}{|c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c|}{-} \\ 
\hline
\multicolumn{1}{|c|}{2025} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2026} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2027} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2028} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\multicolumn{1}{|c|}{2029} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.3} & \multicolumn{1}{c|}{0.4} & \multicolumn{1}{c|}{0.3} \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{Conditional probabilities of the occurrence of the physical changes due to loading of the column} \label{tbleventfault:3}
\begin{tabular}{|c|c|c|c|}
\hline
Year & \multicolumn{3}{c|}{Effect} \\ 
\cline{2-4}
 & \multicolumn{3}{c|}{Loading of column} \\ 
\cline{2-4}
 & Mild deformation of road surface & Severe deformation of road surface & Collapse of object \\ 
\hline
2009 & 0.2 & 0.2 & 0.2 \\ 
\hline
2010 & 0.201 & 0.201 & 0.201 \\ 
\hline
2011 & 0.202 & 0.202 & 0.202 \\ 
\hline
2012 & 0.203 & 0.203 & 0.203 \\ 
\hline
- & - & - & - \\ 
- & - & - & - \\ 
\hline
2027 & 0.218 & 0.218 & 0.218 \\ 
\hline
2028 & 0.219 & 0.219 & 0.219 \\ 
\hline
2029 & 0.220 & 0.220 & 0.220 \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{Conditional probabilities of the occurrence of the physical changes due to horizontal loading of the deck} \label{tbleventfault:4}
\begin{tabular}{|c|c|c|c|}
\hline
Year & \multicolumn{3}{c|}{Effect} \\ 
\cline{2-4}
 & \multicolumn{3}{c|}{Horizontal loading of deck} \\ 
\cline{2-4}
 & Mild deformation of road surface & Severe deformation of road surface & Collapse of object \\ 
\hline
2009 & 0.2 & 0.2 & 0.2 \\ 
\hline
2010 & 0.201 & 0.201 & 0.201 \\ 
\hline
2011 & 0.202 & 0.202 & 0.202 \\ 
\hline
2012 & 0.203 & 0.203 & 0.203 \\ 
\hline
- & - & - & - \\ 
- & - & - & - \\ 
\hline
2027 & 0.218 & 0.218 & 0.218 \\ 
\hline
2028 & 0.219 & 0.219 & 0.219 \\ 
\hline
2029 & 0.220 & 0.220 & 0.220 \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{Conditional probabilities of the occurrence of the physical changes due to vertical loading of the deck} \label{tbleventfault:5}
\begin{tabular}{|c|c|c|c|}
\hline
Year & \multicolumn{3}{c|}{Effect} \\ 
\cline{2-4}
 & \multicolumn{3}{c|}{Vertical loading of deck} \\ 
\cline{2-4}
 & Slightly yielded reinforcement & Plasticized reinforcement & Collapse of object \\ 
\hline
2009 & 1 & 1 & 1 \\ 
\hline
2010 & 1 & 1 & 1 \\ 
\hline
2011 & 1 & 1 & 1 \\ 
\hline
2012 & 1 & 1 & 1 \\ 
\hline
- & - & - & - \\ 
- & - & - & - \\ 
\hline
2027 & 1 & 1 & 1 \\ 
\hline
2028 & 1 & 1 & 1 \\ 
\hline
2029 & 1 & 1 & 1 \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{Conditional probabilities of monetarisable (direct and indirect) consequences} \label{tbleventfault:6}
\begin{tabular}{|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Year} & \multicolumn{4}{m{4.5cm}|}{\centering Loading of column (L1, L2, L3) Horizonal loading of deck} & \multicolumn{2}{m{2cm}|}{\centering Vertical loading of deck (L1, L2)} & \multicolumn{2}{m{2cm}|}{\centering Vertical load of deck (L3)} \\ 
\cline{2-9}
\multicolumn{1}{|c|}{} & \multicolumn{8}{c|}{Failure mode} \\ 
\cline{2-9}
\multicolumn{1}{|c|}{} & \multicolumn{2}{m{2cm}|}{\centering Deformation of road surface} & \multicolumn{2}{m{2cm}|}{\centering Collapse of object} & \multicolumn{2}{m{2cm}|}{\centering Slightly yielded / plastized rebar} & \multicolumn{2}{m{2cm}|}{\centering Collapse of object} \\ 
\cline{2-9}
\multicolumn{1}{|c|}{} & \multicolumn{8}{c|}{Monetarisable direct and indirect consequences} \\ 
\cline{2-9}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{Level 1} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 1} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 1} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 2} & \multicolumn{1}{c|}{Level 3} \\ 
\hline
\multicolumn{1}{|c|}{2009} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.5} \\ 
\hline
\multicolumn{1}{|c|}{2010} & \multicolumn{1}{c|}{0.499} & \multicolumn{1}{c|}{0.501} & \multicolumn{1}{c|}{0.499} & \multicolumn{1}{c|}{0.501} & \multicolumn{1}{c|}{0.499} & \multicolumn{1}{c|}{0.501} & \multicolumn{1}{c|}{0.499} & \multicolumn{1}{c|}{0.501} \\ 
\hline
\multicolumn{1}{|c|}{2011} & \multicolumn{1}{c|}{0.498} & \multicolumn{1}{c|}{0.502} & \multicolumn{1}{c|}{0.498} & \multicolumn{1}{c|}{0.502} & \multicolumn{1}{c|}{0.498} & \multicolumn{1}{c|}{0.502} & \multicolumn{1}{c|}{0.498} & \multicolumn{1}{c|}{0.502} \\ 
\hline
\multicolumn{1}{|c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c|}{-} \\ 
\multicolumn{1}{|c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{-} & \multicolumn{1}{c|}{-} \\ 
\hline
\multicolumn{1}{|c|}{2027} & \multicolumn{1}{c|}{0.482} & \multicolumn{1}{c|}{0.518} & \multicolumn{1}{c|}{0.482} & \multicolumn{1}{c|}{0.518} & \multicolumn{1}{c|}{0.482} & \multicolumn{1}{c|}{0.518} & \multicolumn{1}{c|}{0.482} & \multicolumn{1}{c|}{0.518} \\ 
\hline
\multicolumn{1}{|c|}{2028} & \multicolumn{1}{c|}{0.481} & \multicolumn{1}{c|}{0.519} & \multicolumn{1}{c|}{0.481} & \multicolumn{1}{c|}{0.519} & \multicolumn{1}{c|}{0.481} & \multicolumn{1}{c|}{0.519} & \multicolumn{1}{c|}{0.481} & \multicolumn{1}{c|}{0.519} \\ 
\hline
\multicolumn{1}{|c|}{2029} & \multicolumn{1}{c|}{0.480} & \multicolumn{1}{c|}{0.520} & \multicolumn{1}{c|}{0.480} & \multicolumn{1}{c|}{0.520} & \multicolumn{1}{c|}{0.480} & \multicolumn{1}{c|}{0.520} & \multicolumn{1}{c|}{0.480} & \multicolumn{1}{c|}{0.520} \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
	\centering
	\caption{Consequences} \label{tbleventfault:7}
\begin{tabular}{|p{2cm}|p{2.5cm}|c|c|c|c|c|c|c|}
\hline
Effect & Physical change & Level 1 & \multicolumn{2}{c|}{Owner} & \multicolumn{2}{c|}{User} & \multicolumn{2}{c|}{Public} \\ 
\cline{4-9}
 &  &  & Inter-vention & Travel time & Nr. injuries & Nr. fatal-ities & Nr. injuries & Nr. fatal-ities \\ 
\cline{4-9}
 &  &  & 106 CHF & 106 CHF & 106 CHF & 106 CHF & 106 CHF & 106 CHF \\ 
\hline
On column level 1, 2 and 3 & None & None & 0 & 0 & 0 & 0 & 0 & 0 \\ 
\cline{2-9}
 & Light deformation  of road surface & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\ 
\cline{3-9}
 &  & 2 & 0.5 & 0.5 & 0.5 & 0.5 & 0.5 & 0.5 \\ 
\cline{2-9}
 & Mild deformation of road surface & 1 & 1.5 & 1.5 & 1.5 & 1.5 & 1.5 & 1.5 \\ 
\cline{3-9}
 &  & 2 & 2.5 & 2.5 & 2.5 & 2.5 & 2.5 & 2.5 \\ 
\cline{2-9}
 & Collapse of object & 2 & 5 & 5 & 5 & 5 & 5 & 5 \\ 
\cline{3-9}
 &  & 3 & 10 & 10 & 10 & 10 & 10 & 10 \\ 
\hline
 &  & - & - & - & - & - & - & - \\ 
\hline
Vertical loading of deck Stufe 3 & Collapse of object & 2 & 1 & 1 & 1 & 1 & 1 & 1 \\ 
\cline{3-9}
 &  & 3 & 2.5 & 2.5 & 2.5 & 2.5 & 2.5 & 2.5 \\ 
\hline
\end{tabular}
\end{table}
\subsection{Answer C}
The breakdown of the relative costs and benefits is given in Table \ref{tbleventfault:8}.
The yearly costs and benefits are given in Figure \ref{figeventfault:4}-Figure \ref{figeventfault:5}.
The avalanche protection should be built.
\begin{table}
	\centering
	\caption{ Costs, benefits and effectiveness} \label{tbleventfault:8}
\begin{tabular}{|c|p{1.2cm}|p{1.5cm}|c|p{1.5cm}|p{1.5cm}|c|c|}
\hline
Intervention & \multicolumn{3}{c|}{Relative costs ($10^6$ mu)} & \multicolumn{3}{c|}{Relative benefits (106 mu)} & Effectiveness (B-C) \\ 
\cline{2-8}
 & \centering Planned interventions & \centering Restoration following failure & Total & \centering Reduction of material damage risks & \centering Reduction of failure risks & Total &  \\ 
\hline
Do nothing & \centering 0 & \centering 0 & \centering 0 & \centering 0 & \centering 0 & \centering 0 & 0 \\ 
\hline
Avalanche protection & \centering 4.00 & \centering -1.00 & \centering 3.00 & \centering 2.00 & \centering 3.00 & \centering 5.00 & 2.00 \\ 
\hline
\end{tabular}
\end{table}

\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-16.eps}
\caption{Discounted costs of intervention per year for the do nothing
intervention}\label{figeventfault:4}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-17.eps}
\caption{Discounted risk per year for the do nothing intervention}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-18.eps}
\caption{Discounted costs of intervention per year for the avalanche protection
intervention}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-19.eps}
\caption{Discounted risk per year for the avalanche protection intervention}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=300pt]{faulteventtree-20.eps}
\caption{Discounted benefit (in terms of risk reduction) per year for the
avalanche protection intervention}\label{figeventfault:5}
\end{center}
\end{figure}
\section{Fault tree analysis -- Theory}
\subsection{Definition}
Fault tree analysis is often used when it is desired to focus on the particular
ways a system can fail. It is particularly well suited to determine the
probability of a specific event happening when this event can happen due to
different precursor events. Fault trees are often used in the evaluation of large
safety-critical systems \citep{Andrews2012}.

Fault tree analysis is often referred to as having backward logic as specific
consequences are first determined and then the causes of these consequences are
worked out, i.e. working backwards in time. Fault tree analysis is also sometimes
seen as being top down, i.e. how can the top event be determined from a
combination of lower level events.

In the development of a fault tree one normally starts by defining a top event
and then developing logical expressions of sub-events that may lead to the
occurrence of the top event. The last events in the tree are base events, or
initiating events. Once a fault tree is constructed and the probability of
occurrence of the base events are estimated the probability of occurrence of the
top event can be estimated. When constructing fault trees it is necessary to
ensure that that the events are binary events e.g. failure or no failure, that
they are statistically independent, and that the relationships between events and
causes can be represented by means of logical gates.

Fault trees can also be used to assess the infrastructure related risk due to
natural hazards. Their focus, however, is more on a specific type of consequence,
which may happen due to hazard of different types, e.g. a bridge may collapse due
to a flood or an earthquake.

A separate fault tree is required for every separate top event, or in other
words every failure mode.
\subsection{Symbols}
A fault tree makes use of events and logic gates. These are shown in Table
\ref{tbleventfault:9} and Table \ref{tbleventfault:10}.
\begin{table}
\centering
\caption{Fault tree events} \label{tbleventfault:9}
\begin{tabular}{|>{\centering\arraybackslash}m{2cm}|>{\raggedright\arraybackslash}m{3.5cm}|c|}
\hline
Event label & Event description & Event symbol \\ 
\hline
Basic event & failure or error in a subitem & \includegraphics[scale=0.5]{faulteventtree-2.eps} \\ 
\hline
House event & The event is known to be true or false & \includegraphics[scale=0.5]{faulteventtree-3.eps} \\ 
\hline
Undeveloped event & An event about which insufficient information is available, or which is of no consequence & \includegraphics[scale=0.5]{faulteventtree-4.eps} \\ 
\hline
Conditional event & conditions that restrict or affect logic gates & \includegraphics[scale=0.5]{faulteventtree-5.eps} \\ 
\hline
Intermediate event & can be used immediately above a primary event to provide more room to type the event description & \includegraphics[scale=0.5]{faulteventtree-6.eps} \\ 
\hline
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Fault tree logic gates} \label{tbleventfault:10}
\begin{tabular}{|>{\centering\arraybackslash}m{2cm}|>{\raggedright\arraybackslash}m{3.5cm}|>{\centering\arraybackslash}m{2cm}|}
\hline
Gate label & Gate description & Gate symbol \\ 
\hline
OR & the output occurs if any input occurs & \includegraphics[scale=0.5]{faulteventtree-7.eps} \\ 
\hline
AND & the output occurs only if all inputs occur (inputs are independent) & \includegraphics[scale=0.5]{faulteventtree-8.eps} \\ 
\hline
Exclusive OR & the output occurs if exactly one input occurs & \includegraphics[scale=0.5]{faulteventtree-9.eps} \\ 
\hline
Priority AND & the output occurs if the inputs occur in a specific sequence specified by a conditioning event & \includegraphics[scale=0.5]{faulteventtree-10.eps} \\ 
\hline
Inhibit & the output occurs if the input occurs under an enabling condition specified by a conditioning event & \includegraphics[scale=0.5]{faulteventtree-11.eps} \\ 
\hline
Voting & output event occurs if at least m of the input occur & \includegraphics[scale=0.5]{faulteventtree-12.eps} \\ 
\hline
\end{tabular}
\end{table}
%
\subsection{Steps}
FTA analysis involves five steps:
\begin{enumerate}
\item Define the undesired event, i.e. the top event.
\item Obtain an understanding of the system, i.e. determine all possible ways that
this undesired event may happen. This requires experts who understand the
functioning of the system well. If something is overlooked it will not be
included in the fault tree and this may result in less than optimal decision
being made to reduce the probability of occurrence of the undesired event.
\item Construct the fault tree:
\item Evaluate the fault tree: this requires the determination or estimation of the
probability of occurrence of all base events and then the estimation of the
probability of occurrence of the top event. As the determination of the
probability of occurrence of all base events may take considerable time and
effort it is worthwhile to keep in mind the exactness required so that a decision
maker can come to the optimal decision. This estimation of the probability of the
occurrence of the top event. This is explained in week 9 ``Performance
indicators: Reliability''.
\end{enumerate}
\subsection{Examples in literature}
FTA has been widely used in management of many infrastructure systems. Some of
examples are with gas system \cite{Andrews2012,Tian2012}; bridges \cite{Davis-McDaniel2013,Johnson1999,Sianipar1997,LeBeau2007}; tunneling \cite{You2012}, and urban transport \cite{Ma2014}. Students are requested to read these cited works as part of the handout.
\section{Fault tree analysis -- Example B}
\subsection{Question B1}
As the manager of the bridge shown in Figure \ref{figeventfault:6} and Figure \ref{figeventfault:7},
you are interested in having an overview of all the ways it might fail so you can
determine the most effective way for you to reduce your risks. Construct a
failure tree covering the following hazards 1) corrosion of the steel
reinforcement, 2) spalling of concrete, 3) corrosion of the steel beams, 4)
fatigue of the steel beams, 5) fire, 6) train impact, 7) over loading due to
vehicles, 8) earthquakes, 9) scour. Assume the top event to be bridge collapse.

\begin{figure}[h]
\begin{center}
\includegraphics[width=352pt]{faulteventtree-22.eps}
\caption{Getwing bridge}\label{figeventfault:6}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[scale=1.5]{faulteventtree-23.eps}
\caption{Design of Getwing bridge}\label{figeventfault:7}
\end{center}
\end{figure}

\subsection{Answer B1}

A possible fault tree for the case of bridge collapse can be constructed and
shown in following figures.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.6]{faulteventtree-24.eps}
\caption{First level of the fault tree}\label{figeventfault:8}
\end{center}
\end{figure}

In Figure \ref{figeventfault:8}, the first level of the fault tree is shown. The bridge
can collapse due to the collapse of either the superstructure (beams, girders) or
the substructures (abutment and piers). The failure of both superstructure and
the substructures are developed in the next figures.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.6]{faulteventtree-25.eps}
\caption{Second level of the fault tree (superstructure)}\label{figeventfault:9}
\end{center}
\end{figure}

Figure \ref{figeventfault:9} shows the cause of the failure of superstructure can be due to the
failure of the beams or girders (elements). Furthermore, the failure of each
element can be, either due to manifest deterioration processes or latent
deterioration processes.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.6]{faulteventtree-26.eps}
\caption{Second level of the fault tree (sub-structures)}\label{figeventfault:10}
\end{center}
\end{figure}

Similar to Figure \ref{figeventfault:9}, Figure \ref{figeventfault:10} shows the fault tree for the
substructure, which includes the piers, abutments and bearing system. For the
sake of simplicity, only three branches under sub-structure are shown. The full
diagram would include many more branches. For example, instead of considering the
failure of abutments as one event, the failure of each abutment may be considered
as a separate event, i.e. failure of abutment 1 and failure of abutment 2, and
therefore an illustration of this would include two more branches. Similarly, the
development of the fault tree for piers could include the failure of each pier as
a separate branch. The level of detail to be considered depends on the detail
required in the analysis and the amount of effort to be invested.

Figure \ref{figeventfault:11} shows the last step in development of the tree for the beam
that is affected by manifest processes, e.g. corrosion, cracking, and fatigue
under the normal traffic condition and no error in the design phase. There are a
set of mathematical models to capture these manifest deterioration process (see
script of week 4). It is noted that in this last step of tree development. The
round circle or rectangular is used. This step often involves with the estimation
of probability for the event to occur. For example, under the ``failure due to
corrosion of steels'', there are two direct event that can be consider: 1) the
first event is the concentration of chloride reaching a critical level of 0.48
(see script of week 4); 2) and other event could be the age of the bridge.
%
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.8]{faulteventtree-27.eps}
\caption{Third level of the fault tree (superstructure-beam- manifest
process)}\label{figeventfault:11}
\end{center}
\end{figure}
%
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=1]{faulteventtree-28.eps}
\caption{Third level of the fault tree (superstructure- latent process)}
\end{center}
\end{figure}
%
\begin{figure}[ht]
\begin{center}
\includegraphics[scale=1]{faulteventtree-29.eps}
\caption{Sub-structure (abutment)}
\end{center}
\end{figure}

The complete fault tree for this example is given in OpenFTA extension file.
Students are adviced to install an open source fault tree analysis software
called ``OpenFTA'' from following link \href{http://www.openfta.com/}{http://www.openfta.com/}.
\subsection{Question B2}
Expand the fault tree to investigate the failure due to scour of the foundation
of column 6.
\subsection{Answer B2}
This figure shows the extension of the fault tree, which includes the scouring
process affecting the abutments and the piers of the bridges. The scouring
process can be due to four process: 1) local scouring; 2) contraction; 3)
widening; and 4) lateral migration. To understand these terminology, students are
advised to read the work of \cite{Johnson1999}, which is provided as the class handout.

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=1]{faulteventtree-30.eps}
\caption{Fault tree for scouring of the pier}
\end{center}
\end{figure}
\section{Assignment}
\subsection{Problem A: Event tree analysis}
Water is one of the fundamental needs of millions of people living in a
megacity. Water of sufficient quality and quantity must be provided around the
clock. In order to fulfill this need, a city depends on its water distribution
infrastructure, which includes pipes made of different materials and laid at
different times. These pipes are affected by processes of different types and
deteriorate at different rates. The consequences related to pipe failure vary
significantly depending on the type of failure, e.g. a pipe break, or a leaking
pipe, as does the reaction time required to fix the pipe. For example, if a pipe
breaks, an inadequate LOS occurs immediately and a corrective intervention is to
be executed very quickly, and if it is noticed that there is progressive water
loss over time then a preventive intervention can be planned before there is an
inadequate LOS. Part of the water distribution network in mega-city Q is shown in
Figure \ref{fig24}. The pipe characteristics are given in Table \ref{tbl:213}.

\subsubsection{Question A1}
Construct an event tree for the pipe network if an earthquake occurs? The event
tree should include a sensible classification of events and consequences and it
should be realistic to determine the probability of occurrence of each scenario.
For the example, you should assume a probability of occurrence of each event in
your event tree, and then estimate the probability of occurrence of each
scenario. At whatever level you define your event tree be sure that it is
complete!
\subsubsection{Answer A1}
No specific answer is provided. Please submit your answer to obtain feedback. A
possible source event is the occurrence of an earthquake. Possible intermediary
events are the zones in which a pipe breaks occur if an earthquake occurs.
Possible network events are the loss of water and the loss of water pressure to
serviced areas.
\section{Problem B1 }
The manager of the pipeline network as shown in Figure \ref{fig24} expects to
understand how a large amount of water loss could occur. Water loss can be due to
many reasons, which includes the manifest and latent deterioration process.
\subsection{Question B1}
Construct a fault tree for the pipe network (Figure \ref{fig24}) if there is a
massive amount of water loss?
\subsection{Answer B1}
No specific answer is provided. Please submit your answer to obtain feedback.
Keep in mind though that water loss may happen in one or more zones, and that
these zones could be subdivided into specific pipes. Keep in mind that there are
many different ways for pipes to fail, e.g. holes in pipes can occur due to
differential soil movements, due to the penetration of roots into the pipes, the
corrosion of reinforcement, sabotage, mistakes on construction sites, etc.

\bibliographystyle{plainnat}
\bibliography{reference}
%

### 2019-05-05 InfrastructureMaintenance/Chapters/introduction.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Infrastructure Maintenance Processes: An Introduction}
\label{intro} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Bryan T. Adey and Nam Lethanh}

\section{General}
Infrastructure consists of the fixed physical objects used to ensure the functioning of a society, e.g. the bridges in a road network, the track in a rail network, the pipes in a water supply network or sewer network, the transformers in an energy distribution network, and the masts in telecommunication networks. Infrastructure management is the process used to ensure that existing infrastructure objects and networks provide adequate levels of service (LOS) for specified periods of time. Good infrastructure management requires taking into consideration the benefits and costs of infrastructure to all members of society, including economic, ecological and social benefits and costs, as well as balancing the need for prediction accuracy with analysis effort. The main tasks in the infrastructure management process are:

\begin{itemize}
	\item determine the required LOSs,
	\item assess the ability of the infrastructure to provide these LOSs, 
	\item determine the performance indicators to be used to determine that the required LOS are being provided, 
	\item develop and evaluate both development and maintenance strategies, which include general considerations of the interventions to be executed when certain situations exist.
	\item develop and evaluate work programs, which include the specific interventions to be executed taking into consideration the actual situation that exits. 
\end{itemize}

This process is true for all types of infrastructure. In order for it to be executed effectively in today’s world it also requires the use of a large amount of modern technology, including the use of advanced inspection techniques, computerized management systems and effective infrastructure management organizations. In this article an overview of the main tasks in the infrastructure management process is given.

\section{Levels of service}\label{1:los}
In order to manage infrastructure well, it is necessary to determine the LOS required of infrastructure and how it changes over time. The first part of this is the definition of how the future should be with new or modified infrastructure, e.g. the longest shortest drive between any two nodes in the road network is to be under x hours. The second part is the definition of how the future should be during the operation and maintenance of the existing, and the new and modified, infrastructure, e.g. the number of accidents should be less than x per year.

Although the initial definition of the LOS may be expressed in an incomplete fashion, as above, a complete definition in terms of impacts on humans should be made. This is required, in order to objectively evaluate how well infrastructure provides the required LOS when different intervention strategies and intervention programs are followed. The items to be included in the definition of the LOS should be included in an impact hierarchy that is complete, operational, quantifiable and non-redundant. An impact hierarchy needs to be complete to ensure that the value of one impact is not improved to the detriment of another. For example, the reduction of the cost of manual labor on a construction site which results in an inadvertent rise in the number, and therefore the costs, of accidents. An impact hierarchy needs to be operational so that it can be used. For example, there is little need to have an impact hierarchy which includes a breakdown of accidents by the age of the persons in the vehicles if it is considered to be too laborious to collect such information. An impact hierarchy needs to be quantifiable as it is difficult to compare strategies or intervention programs using qualitative measures. In this case it is better to make difficult to quantify impacts such as noise into something quantifiable and accept that there is a substantial variation in the possible values than to simply say that one intervention program would result in less noise than another. The two main possibilities to measure things in proportional units are utility or monetary units. The latter has the advantage that many impacts are already measure in terms of monetary units. An impact hierarchy needs to be non-redundant so that accurate results in analyses are obtained.

The LOS to be provided by infrastructure changes over time. It may change due changes in already defined types of impacts, e.g. last year travel time impacts were not to exceed x CHF and this year they are not to exceed y CHF. It may also change due to changes in not yet defined requirements, e.g. last year there were no limitation on the amount of CO2 emitted, and this year the amount of CO2 emitted is not to exceed x tons.

Changes in the LOS to be provided can be classified as gradual (manifest) or sudden (latent). A gradual change in the LOS is one that happens in a way that there is enough time to execute an intervention so as to ensure that the infrastructure continues to provide an adequate LOS. A sudden change in the LOS is one that happens in a way that there is not enough time to execute an intervention so as to ensure that the infrastructure continues to provide an adequate LOS. For example, a gradual change may be the increases in the amount of travel time due to increasingly congested roads, something that may happen at a rate of 2\% per year. A sudden change may be a new law that allows the axle loads to increase by 50\%. The classification of a change in LOS as gradual or sudden depends on the state of preparedness and the perception of the infrastructure management operator (IMO).

The evaluation of the LOS provided by infrastructure requires the use of models to measure the values of performance indicators associated with the LOS \citep{Talvitie1999}, e.g. analytical models to predict the impacts due to noise produced by vehicles on road networks \citep{Fyhri2010} or trains on rail network \citep{herrmann08}, neural networks to predict the impacts due to the execution of interventions on water distribution networks \citep{Bowden2006} or buildings \citep{Marzouk2013}, and Bayesian networks to predict the impacts due to accidents \citep{Wang2013}.

The task to identify the LOS to be provided involves consideration of what the IMO can offer or would like to offer, and consideration of what stakeholders of the infrastructure, e.g. users and persons who are directly affected by the infrastructure but who are not users and persons who are indirectly affected by the infrastructure. The latter can require considerable interaction between multiple parties and often includes dealing with stakeholder representatives such as politicians or lobbies, as well as through consideration of laws.

\section{Performance indicators}
The determination of whether or not a required LOS is provided by infrastructure if default strategies and intervention programs are being followed requires assessment. This is done through the estimation of performance indicators \citep{Talvitie1999}.

The most complete performance indicators are the values of the total impacts associated with the multiple possible futures that may happen and the probabilities of occurrence of these possible futures. Here the only relevant impacts are those related to the execution of interventions, the occurrence of accidents and the production of noise.

As the evaluation of total impacts is laborious, it is often advantageous to use partial performance indicators. Four common ones are reliability, availability, maintainability and safety. Reliability is the probability that infrastructure provides a specific LOS for a specified period of time. Availability is the amount of time that infrastructure provides a specific LOS over a specified period of time taking into consideration the length of time that the infrastructure is not providing the specific LOS due to the execution of interventions. Maintainability is a measure of the impacts associated with the execution of interventions or the time required to execute interventions over a specified period of time. Safety is determined through the assessment of the number of persons injured.

A significant part of the estimation of the values of performance indicators for infrastructure networks depends on the topology of the networks.  For example, the estimation of the reliability of traveling from A to B in a road network depends on whether A and B are connected in a series, parallel, or a more complex network, as well as how each of the objects in each of the network links performs \citep{Bell2000}. Similar considerations are applicable to the estimation of availability and maintainability. Reliability is often used when the impacts related to the infrastructure being non-operational are substantially higher than when it is operational, e.g. in the case of sudden deterioration of infrastructure due to floods. Availability is often used when the impacts related to the infrastructure being operational and non-operational are relatively similar, e.g. in the case of gradual deterioration of infrastructure due to chloride induced corrosion. Maintainability is often used when the impacts related to the execution of interventions on the infrastructure is of particular interest. Safety is often used when there is a perception that the number of accidents is too high.

The task of determining the performance indicators to be used involves consideration about how information is collected, i.e. monitoring, and making assumptions about what is important. This includes the generation and assessment of possible monitoring strategies, the estimation of the impacts associated with the performing inspections, the development of data management strategies, and consideration as to how the data will be used.

\section{Strategies}
Intervention strategies, whether they are development or maintenance strategies, include all interventions to be executed on an object taking into consideration many possible conditions of the object or many possible situations in which the object could be. For example, if there is cracking in the pavement part of a road section a crack filling intervention should be executed and if there is cracking and significant unevenness then a partial depth repair intervention should be executed. Intervention strategies are developed without consideration of externalities, such as the condition of other objects or other types of infrastructure.

A development strategy is one of how to modify existing infrastructure so that it provides a new LOS. It principally deals with possible changes in requirements. For example, if the number of trains traveling on a railway line is over 90\% of the design capacity then a new track will be built. A maintenance strategy is one of how to modify existing infrastructure so that it continues to provide the same LOS as initially intended. It principally deals with possible changes in the infrastructure. If more than 50\% of the material from track is worn away then the track will be replaced.

The generation of possible development and maintenance strategies requires in-depth infrastructure sector specific knowledge. The generation of such strategies requires good technical persons and organizations with structures that encourage both cooperation and innovation. The evaluation of the possible strategies and the determination of the optimal ones require a good foundation in decision theory and an ability to model how impacts will change over time if each of the strategies are followed. Some models used to evaluate strategies are common, such as block replacement models, age replacement models, and Markov models. Others are custom built to address specific questions, using standard operations research methods. In the evaluation of strategies both the impacts during execution of interventions (construction and maintenance) and between the executions of interventions must be considered.

The task to develop and evaluate both development and maintenance strategies involves consideration about the interventions that are possible in different situations, including those that affect the present and the future, e.g. the addition of a new concrete layer on a reinforced concrete element, and the ones that affect only the future, e.g. the addition of a waterproofing layer on a concrete bridge deck. It also involve the estimation of the impacts associated with the execution of the interventions, their effects on future LOSs, their duration, and the probability of them working as intended, as well as the consequences if they do not work as intended.

\section{Infrastructure}
In order to manage infrastructure well, it is necessary to determine the LOS provided by infrastructure and how it changes over time. Some, but normally not all, aspects of the LOS to be provided by infrastructure are often defined in building codes. For example, a highway design code may specify that all objects are to be designed to withstand a flood that returns on average every 100 years which indirectly limits the costs of interventions related to rebuilding the highway due to floods, but it may not specify the minimum conditions above which the infrastructure must be kept to ensure that the costs related to the noise produced by vehicles traveling on the highway remain above a specified limit.

The determination of the LOS provided by infrastructure requires understanding the objects that comprise the infrastructure network being managed, as well as how these objects, and therefore the network, perform.

Infrastructure loses its ability to provide LOS due to processes that can be classified as gradual and sudden. A gradual process is one that happens in a way that there is enough time to execute an intervention to ensure that the infrastructure continues to provide an adequate LOS. A sudden process is one that is happens in a way that there is not enough time to execute an intervention to ensure that the infrastructure continues to provide an adequate LOS. An example of a gradual process is chloride-induced corrosion of reinforced concrete in which an expansive rust product is created. An example of a sudden process is earthquake.

Many models can be used to predict how infrastructure performance changes over time. Popular models for gradual processes include mechanistic-empirical models (e.g. \cite{Newbery1988}), regression models (e.g. \cite{Madanat1995}), and Markov models (e.g. \cite{Lethanh2015}). Mechanistic-empirical models are used for example to predict the speed of chloride ingress into reinforced concrete bridge decks, the start of corrosion, and the speed with which the cross section area of the reinforcement will be reduced. Regression models are used to predict the number of potholes that will occur on road sections. Markov models are used to model the change in condition over time of bridge elements \citep{ASTRA2010}\footnote{ASTRA (the Swiss Federal Road  Office) uses KUBA, a bridge management system built on a Markov chain model, to predict deterioration of highway bridges and to determine optimal intervention strategies over a specified planning horizon}. Popular models for latent processes include event trees, Bayesian networks, and fault tree models. For example, event trees and fault tree models have been used to measure the reliability of dams and nuclear reactors \citep{Pate-Cornell1984}. Bayesian networks have been used to quantify the risks associated with road accidents \citep{Deublein2013,Kohler2007} or hazards occurred after an earthquake in a urban city \citep{Bayraktarli2011}.

The task to determine the LOS provided by the infrastructure involves surveying the condition and the performance of all infrastructure, which often includes discussing with the many persons taking care of the infrastructure and developing models of the connectivity of the objects, determining the LOS of the infrastructure is likely to provide, and the probability that it will do so. This requires the definition of the hazards that are likely to affect the infrastructure and having appropriate persons provide the assessments of the condition and performance of the infrastructure. This information then needs to be integrated in a way that provides a cohesive overarching picture.

\section{Intervention programs}
Intervention programs include the interventions to be executed on objects with consideration of externalities, such as the condition of other objects or other types of infrastructure. They include all interventions to be executed, regardless if they are due to development strategies or maintenance strategies. The consideration of externalities results in deviations from those that would result if optimal intervention strategies were followed without consideration of externalities. For example, if an object is not in a condition state in which the intervention strategy includes the execution of intervention, but part of the network is closed due to the execution of other interventions, than an intervention might also be executed on the object.

The development of optimal intervention programs is dependent on the location of the objects in networks and the timing of the interventions on the objects. The determination of optimal intervention programs requires making trade-offs and deviations from the theoretically optimal intervention strategies.

The generation of possible intervention programs requires good technical persons and organizations with structures that encourage both cooperation and innovation. One of the particular challenges in the generation of intervention programs that give adequate consideration to development and maintenance interventions, is that it requires understanding of

\begin{itemize}
\item how the infrastructure needs to be adapted or could be adapted to ensure that it provides adequate LOSs,
\item the impacts related to the execution of maintenance interventions and how they change over time,
\item how the impacts related to the execution of interventions are affected by the execution of other interventions simultaneously.
\end{itemize}

The generation of possible intervention programs increases in difficulty with the size of IMOs, as there is an increasing division of labor, i.e. different persons are responsible for the development strategies and maintenance strategies, and different persons are responsible for different types of infrastructure. This difficult can, however, be overcome with clear guidelines of how information is to be aggregated and how decisions are to be made as to which interventions are to be included in the intervention program.
Traditionally intervention programs have been developed through negotiations with these different persons. Increasingly, however, computers are being used to help determine the intervention programs to follow. Over time the algorithms used are increasingly more sophisticated.

The task to develop and evaluate intervention programs involves securing the financing to conduct the intervention programs, ensuring that all legal requirements are met and ensuring that there is enough staff available for conducting the intervention program.

\section{Projects}
Once intervention programs are finalized, planning can progress to a much more detailed level, i.e. the planning and execution of projects. This task involves tendering the interventions, reviewing the applications, selecting the best companies to execute the interventions, planning the interventions, communicating the activities with stakeholders, and all other normal project management related activities.
\section{Conclusions}
The management of infrastructure is an important process. Good infrastructure management can result in substantially reduced negative impacts to society when compared to bad infrastructure management. Good infrastructure management requires the exploitation of the latest advances in the field of infrastructure management.

\bibliographystyle{plainnat}
\bibliography{reference}

### 2019-05-05 InfrastructureMaintenance/Chapters/los.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter2.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Level of Service - Impact Hierarchy}
\label{2:los} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Bryan T. Adey and Nam Lethanh}
\section{General}
In order to determine what to do with infrastructure it is necessary to determine what is expected from the infrastructure, i.e. it is necessary to determine the required LOSs. This must be defined for a specific time period. It requires the determination of an impact hierarchy and the determination of the values of each impact indicator or impact value over the specified period of time \citep{Adey2012}.
\section{Impact hierarchy - Theory}
\subsection{General}
The best way to define LOS provided by infrastructure is to consider it to be identical to the amount that persons are affected by the functioning (or non-functioning) of infrastructure with respect to the required LOS. For example, on a road network, it may be expected that next year $x$ $mus$\footnote{$mus$ stands for monetary units} are to be spent on the execution of maintenance interventions and y mus worth of impacts are due to accidents. If more than $x$ $mus$ are to be spent on maintenance and more than $y$ $mus$ worth of impacts occur due to accidents than it can be considered that an inadequate LOS was provided. If less or equal, than it can be considered that an adequate LOS was provided.

If it is too difficult to quantify all impacts to be used to measure LOS in terms of directly comparable units (above they are given in $mus$) proxies can be used. For example, instead of stating that y mus worth of impacts due accidents are expected next year, the number of accidents may be stated.

The type of proxy used can be evaluated with respect to its usefulness in assessing the amount that persons are affected by the functioning (or non-functioning) of infrastructure with respect to the required LOS. For example, the number of accidents might be considered a relatively good proxy with respect to the amount persons are affected by accidents, whereas the number of potholes in the road, might not be. The number of accidents can be multiplied by the average impact of an accident to give the expected total impacts, whereas if the number of potholes is used to calculate the expected total impact, it must be multiplied with the expected number of accidents that occur when there are this number of and the average impact of an accident \footnote{the expected number of accidents when a road section has $x$ numbers of potholes is often estimated as probabilistic function that takes into account of some factors such as numbers of potholes, traffic volume, vehicle speed, etc.}. In general, the fewer the number of calculations that are required to determine the amount that persons are affected, the better.

In any case, the determination of the values of the proxies to be used should be approximated based on a rigorous assessment of the impacts related to these values, and a comparison of these values with the other impact types used to define LOS. If not, an optimal performance of the infrastructure will not be targeted.

It is important to define the LOS so that it covers all ways in which persons can be affected by the functioning (or non-functioning) of infrastructure. If this is not done, then there is a significant probability that managers will concentrate on some impacts and ignore others, which may also be important. For example, a manager might choose to reduce the probability of occurrence of an accident in a tunnel to a very small amount by imposing a regulation to reduce vehicle speed (e.g. from 100 km/hour to 80 km/hour) in the tunnel if only owner costs and accident costs are considered. If, however, owner costs, the costs of travel time and the costs of accidents are taken into consideration he may determine that the best strategy is to renew the road surface and the lighting system more frequently, even though it would result in higher costs for the owner.

To ensure that all this can be done it is useful to:

\begin{itemize}%[leftmargin=2pt-.8in]
\item determine the objective of setting the LOS
\item determine and classify the persons affected by the functioning (or non-functioning) of the infrastructure, i.e. stakeholders
\item determine and classify the ways in which each of the stakeholders are affected, i.e. the impact types. 
\item determine the indicators to be used to determine the value of the impacts incurred by stakeholders,
\item determine the value of a unit change in the indicators used in units that can be compared across all impact types.                                                                                                                    \end{itemize}

The grouping of stakeholders, their impacts at each level, and the indicators to be used to measure these impacts are referred to as an impact hierarchy. In the development of an impact hierarchy it is important that it is:

\begin{enumerate}
\item \textbf{complete} - all non-negligible ways in which persons are affected should be included. If not, there will be an unwanted focus on the items included in the hierarchy to the detriment of those that are not, which may lead overall to sub-optimal decisions.
\item \textbf{decomposable} - each impact should be broken down, or decomposed, to a level that is reasonable quantifiable. If the hierarchy is not decomposable to a reasonable good quantifiable level persons will be required to place numbers on impacts relatively subjectively.  
\item \textbf{operational} - the impacts should be defined in a way and a level which it is reasonable to quantify them in terms of effort. If a hierarchy is defined in too much detail the effort will be too large to conduct the required analysis or to collect the required information. If a hierarchy is defined in too little detail, there will be so much subjectivity in the assessment of the value of the impacts that the results will be relatively useless. 
\item \textbf{non-redundant, or orthogonal} - the impacts should be defined in a way that ensures impacts are not double counted. To help ensure orthogonality it is sometimes helpful to make use of the pillars of sustainability, i.e. economic, environmental, and social, in the definition of the lowest level impacts in the hierarchy. If the hierarchy is not orthogonal than the double counting of some, but not all, of the impacts will result in overall sub-optimal decisions.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 \end{enumerate}
%
\subsection{Objective}
The stakeholders, impacts types and proxies to be considered depend on the objective of determining the LOS. For example, if a road manager wants to determine the work program that will maximize the net positive impact that society as a whole obtains from the infrastructure, only changes that are relevant to society as a whole are considered, and therefore positive impacts on one stakeholder that are negative impacts on another are not taken into consideration. For example, the income that is received from charging a toll on a public road would not be considered as it is a positive impact on the owner but a negative impact of equal magnitude on the user.
%
\subsection{Stakeholder}
There are many ways to classify these persons. It is, however, advantageous to group them keeping in mind how models will be built to assess how they are affected over time. For example, a person affected by noise who is driving a car may not be affected in the same way as a person who lives next to road on which cars pass.

Being a stakeholder is time dependent. For example, when a person is driving a vehicle on a road he is a user at that point in time, but when he is off of the road and in his house far from the road he is part of the indirectly affected public. Although he is the same person he is affected by the road when he is using it in one way, and when he is not using it in another way.

Depending on the objective it can be helpful to classify stakeholders as either first level, or second level or third level stakeholders. The first level stakeholders are those whose net positive impacts should be maximized. The second level stakeholders are those whose impacts are the outcome of the maximization of the net impacts of the first level stakeholders, and should be monitored.  The third level of stakeholders are those whose impacts are of no concern.
%
\subsection{Impact}
Each stakeholder can be affected by the functioning (or non-functioning) of the infrastructure in different ways. For example, someone driving on a road can be affected by being in an accident or through the amount of time they spend traveling.

The impacts are attributed to the stakeholder who is most directly affected. For example, the money spent to hire persons to execute an intervention on a bridge should be attributed to the owner of the bridge rather than the tax payers.

The impact are to be grouped in a hierarchy and given to a level of detail in a way in which it can be reasonably expected that they can be quantified.

The impact hierarchy should include all non-negligible impacts to the “principal” stakeholders of public road infrastructure, i.e. those whose positive impacts should be maximized.

\subsection{Indicator}
In order to estimate the impacts it is necessary to determine impact indicators, which are considered to be representative of each impact type. It is only by estimating the values of these indicators over time, and attributing monetary values to each unit change in the indicators, that the impact on the stakeholders can be correctly evaluated. If the values of these indicators are used directly to set a LOS they are referred to as proxies.

\subsection{Conclusions}
Determination of the impact hierarchy is the precursor to setting the required LOS. It is important that sufficient time be spent to define it well, and to ensure that decision makers understand it. It is only with this understanding that the statement of the required LOS will make sense, which will set the stage for good decision making in the future.

Ideally the determination of the required LOS would be closely linked to the theoretically optimal LOS, i.e. the stakeholders as a whole receive the maximum net benefit from the infrastructure. As there are, however, many different stakeholders affected by the functioning (or non-functioning) of infrastructure, with many different opinions of the values to be associated with the unit changes in different types of impacts, the fixing of the required LOS is something that is associated with considerable negotiation between all involved stakeholders \citep{Brauers204}.

\section{Impact hierarchy - an example}
\subsection{Question A}

Develop an impact hierarchy that will allow a public road manager to determine the work program that will maximize the net positive impact that society as a whole obtains from the infrastructure over the next 20 years. Be clear as to the objective of developing the impact hierarchy. Explain the main components of the hierarchy in general. Give the complete hierarchy. Explain how you would attempt to quantify each impact type at the lowest level.

\subsection{Answer A}

\label{GrindEQpgref4d79f6d52}\subsubsection{Objective}

This impact hierarchy is developed for public roads assuming that a road manager wants to determine the work program that will maximize the net positive impact that society as a whole obtains from the infrastructure Therefore, only changes that are relevant to society as a whole are considered, i.e. positive impacts on one stakeholder that are negative impacts on another are not taken into consideration. For example, the income that is received from charging a toll on public road is not considered as it is a positive impact on the owner but a negative impact of equal magnitude on the user.

\subsubsection{Components}

The impact hierarchy is to consist of stakeholders, impact types and impact indicators.

\paragraph{\underline{Stakeholders}}

A stakeholder is considered as an individual, group, or organization, which is affected by changes to public roads. Being a stakeholder is time dependent, i.e. when a person is driving a vehicle on a road he is a user at that point in time. When he is off of the road and in his house far from the road he is part of the indirectly affected public. It is considered that all stakeholders can be grouped as either first level or second level stakeholders. The first level stakeholders are those whose net positive impacts should be maximized. The second level stakeholders are those whose impacts are the outcome of the maximization of the net impacts of the first level stakeholders, and should be monitored.

The four first level stakeholder groups are the owner, the user, the directly affected public (DAP), and the indirectly affected public (IAP). It is assumed that all impacts to be maximized can be attributed to one of these four stakeholder groups. The impacts are attributed to the stakeholder who is most directly affected. The definitions of each stakeholder group are given in later parts of this section.

The proposed impact hierarchy is to be used to take into consideration all non-negligible impacts to the ``principal'' stakeholders of public road infrastructure, i.e. those whose positive impacts should be maximized. It is considered that all people at a point in time can be divided into one of four stakeholder groups, the owner, the user, the directly affected public, and the indirectly affected public. The impact types in the hierarchy are given to a level of detail in which they could be reasonably expected to be quantified. To help ensure orthogonality the pillar of sustainability, i.e. economic, environmental, social, is stated for each of the lowest level impacts. An example of each impact type is given.

\begin{table}[htbp]
\caption{Stakeholders}
\begin{tabular}{|p{90pt}|p{170pt}|p{170pt}|}
\hline
Stakeholders & Definition & Examples \\ \hline
Owner & the persons who are responsible for decisions with respect to physically modifying the infrastructure & a federal road authority  \\ \hline
Users & the persons who are using the roads & a driver and passengers of a vehicle on a road. \\ \hline
Directly affected public & the persons who are in the vicinity of the road but are not using it & persons in a house next to the road that hear vehicles driving on the road. \\ \hline
Indirectly affected public & the persons who are not in the vicinity of the road but are affected by its use & persons in a house far away from the road that do not hear vehicles driving on the road, but are affected by a changing climate due to the emissions produced by vehicles driving on the road. \\ \hline
\end{tabular}
\label{tbl:21}
\end{table}

Second level stakeholder groups, such as contractors, financial institutions, and operators are not considered further than they are considered when considered as one of the above listed stakeholders. The impacts on these stakeholder groups are only the outcome of our efforts to maximize the positive net impact of the four first level stakeholders in the first level.

\paragraph{\underline{Impact types}}

The impacts on each stakeholder are grouped as impact types. The impact types are subdivided at increasingly fine levels until the impact of each type can be reasonably and objectively quantified and modelled. To help to ensure orthogonality in the impact hierarchy, each impact type, on the lowest defined level, is explained and classified as contributing to one of the pillars of sustainability (economic, societal, environmental). An example is given for each to help clarify its meaning.

\paragraph{\underline{Impact indicators}}

Impact indicators, which are considered to be representative of each impact type are given for each impact type. By estimating the values of these indicators over time, and attributing monetary values to each unit change in the indicators, it is possible to evaluate the impact of the stakeholders. Examples of models to be used to estimate how their values change over time per impact type are given in section 5. The impact hierarchy is given in Table \ref{tbl:22} to Table 5 in the following sections.

\subsubsection{Impact hierarchy}

\paragraph{\underline{Owner}}

The impacts attributed to the owner are grouped as intervention costs (Table \ref{tbl:22}), i.e. the impact on the owner of executing interventions. The impact indicators are the amount of labour, the amount of equipment, and the amount of material to be used to execute interventions, $I_{labor}^o$\textit{,,}$I_{material}^o$ e.g. intervention \textit{w} required \textit{x} man-hours, \textit{y} generator-hours and \textit{z} kilograms of material\footnote{Throughout the document, I represents ``impact'' and it is used to as mathematical notation in functions where appropriate.}.

The monetary value placed on the

\begin{itemize}
	\item labor used represents the economic impact of persons performing tasks, i.e. the value from society's perspective of the person doing the intervention. 
	\item material used represents the economic impact of people ensuring that materials are available for use, i.e. the value from society's perspective of persons preparing the materials for use. 
	\item equipment used represents the economic impact of people ensuring that equipment is available for use, i.e. the value from society's perspective of persons preparing the equipment for use.
\end{itemize}

The estimation of the value of an intervention is often done using one of two approaches:

\begin{itemize}
	\item a disaggregate approach where expenditures for each item or activity are estimated and summed. When this approach is used the work break down structure of the intervention project it is often used.
	\item an aggregate approach where sum of all expenditures is estimated directly. An aggregate approach often includes regression analysis and historical information. 
\end{itemize}

\begin{table}[htbp]
\caption{Owner impact types}
\begin{tabular}{|l|p{100pt}|p{70pt}|p{170pt}|}
\hline
\multicolumn{2}{|l|}{Level 1} & \multicolumn{2}{l|}{Level 2} \\ 
\hline
Label & Description & Label & Description \\ 
\hline
Intervention & the impact of executing interventions & Labor* & the economic impact of people performing tasks \\ 
\cline{3-4}
 &  & Material* & the economic impact of people ensuring that materials are available for use \\ 
\cline{3-4}
 &  & Equipment* & the economic impact of people ensuring that equipment is available for use \\ 
\hline
\end{tabular}\\
\label{tbl:22}
\footnotesize * These could be further subdivided based on the type of activity performed, e.g. administration, planning, etc...
\end{table}

\paragraph{\underline{User}}

During the interventions, users experience inconveniences such as traffic jams, bumping condition of temporarily roads (detour roads). The unfavorable conditions of traveling via the road sections and links subjected to intervention will eventually lead to higher possibility of accidents, various types of physical exhaustion and illness as comfort levels decrease, loss of travel time, and increases in fuel consumption and frequency of the maintenance of vehicles.

In between interventions, deterioration processes result in a worsening condition of any infrastructure object. The increasingly poor condition of infrastructure also results in a change in how stakeholders are affected e.g. increases in the number of accidents, travel time, the costs of operating and maintaining vehicles.

The impacts attributed to the user are grouped as: safety, operation efficiency, operation quality, and environment preservation (Table \ref{tbl:23}).

\begin{table}[htbp]
\caption{User impact types}
\begin{tabular}{|p{70pt}|p{100pt}|p{70pt}|p{150pt}|}
\hline
\multicolumn{2}{|l|}{Level 1} & \multicolumn{2}{l|}{Level 2} \\ 
\hline
Label & Description & Label & Description \\ 
\hline
Safety & the impact on the user due to the user being involved in an accident & property damage & the economic impact of repairing the vehicle \\ 
\cline{3-4}
 &  & injury & the societal impact due to the injury \\ 
\cline{3-4}
 &  & death & the societal impact due to death \\ 
\hline
Operation efficiency & the impact of travel in terms of time required & work & the economic impact of wasting work time traveling \\ 
\cline{3-4}
 &  & leisure & the economic impact of wasting leisure time traveling \\ 
\cline{2-4}
 & the impact of travel in terms of vehicle costs & operation & the economic impact of people ensuring that fuel and oil is available for use \\ 
\cline{3-4}
 &  & maintenance & the economic impact of people repairing vehicles and ensuring that materials, e.g. tires and brake pads, are available for use \\ 
\hline
Operation quality & the impact of traveling on the user & physical & the societal impact of obtaining for example, bruises from an extremely bumpy ride \\ 
\cline{3-4}
 &  & psychological & the societal impact of having for example, anxiety due to a perceived increase in the probability of being involved in an accident, or of seeing things while traveling. \\ 
\hline
Environment preservation (reduction of noise) & the societal impact due to the user coming in contract with sound emissions &  &  \\ 
\hline
\end{tabular}\label{tbl:23}
\end{table}

\begin{enumerate}
 \item Safety (Accidents)

 Accidents result in damage to the property of involved parties. The owners will often repair the damaged objects so as to provide adequate service to users after accidents (this would be attributed to the owner). The users, however, will be required to repair their vehicles, and will also be affected by any injury and of course, death, that may befall them. The safety impact type attributed to the user is subdivided into property damage, injury and death impact types.
 \item Property damage

  The property damage impact type represents the economic impact of repairing the vehicle, i.e. of providing the user with a functioning mode of transport similar to the one being used before the accident, e.g. the costs of the labor, materials and equipment required to replace the bumper on a vehicle that has been in an accident. The impact indicators for property damage are the amount of labor and materials used to repair vehicles damaged in an accident, expressed as the property damage costs given by $I_{acc - pd}^u$. The value of this impact type can be approximated using the receipts from past repairs.
\item Injury and death

The injury impact type and the death impact type attributed to the user represent the social impact due to injuries and deaths, respectively. They represent the change in interactions between persons that will occur because the user is injured or dead. It is not to be confused with the injury impact type and the death impact type attributed to the directly affected public (refer to section \ref{2dap}) or to the indirectly affected public (refer to section \ref{2iap}). The impact indicators are the number of injuries and deaths incurred in a specified time interval, given by$I_{acc - injury}^u$, $I_{acc - death}^u$, respectively

The value of these impact types can be estimated by using the user's willingness to pay to avoid injury or death. Care must; however, be taken to avoid double counting with the injury impact type and the death impact type attributed to the indirectly affected public.

\item Operation efficiency

The operation efficiency impact type represents the impact on the traveling of users and on the maintenance and operation of the vehicles.

\item Travel time

The amount of time traveling on the road is determined by various factors such as: speed, road condition (drivers feel comfortable on a smooth road, and therefore drives faster than on a bumpy road), and the DTV (the daily traffic volume, especially in relation to the road capacity), and road geometry. Furthermore, in case of an intervention, a detour might be needed.

The economic impact of wasting work and leisure time traveling may be thought of as the loss of productivity of the users due to time spent traveling. The impact indicators are the amounts of work and leisure time wasted while traveling and are given by $I_{tt - work}^u$, $I_{tt - leisure}^u$

The value of travel time can be determined using willingness to pay surveys.

\item Vehicle operation and maintenance

The vehicle operation and maintenance impact types represent the economic impact of people ensuring that fuel and oil is available for use, and the economic impact of people repairing vehicles and ensuring that materials, e.g. tires and brake pads, are available for use, respectively. The impact indicators are$I_{v - oper}^u$$I_{v - main}^u$.

The value of the vehicle operation and maintenance impact types can be approximated using the receipts from fuel and vehicle service receipts.

\item Operation quality

The operation quality impact type is subdivided into the physical and psychological impact types. The physical impact type represents the social impact of obtaining for example, bruises from an extremely bumpy ride. It represents the change in interactions between people that will occur because the physical change in the user due to the bumpy ride. The psychological impact type represents the impact of having for example, anxiety due to a perceived increase in the probability of being involved in an accident, or of seeing things while traveling, e.g. aesthetics. It is believed that any economic impacts relevant to society due to the physical and psychological impacts of traveling, such as the loss of productivity, are negligible, in developed countries.

The impact indicators are the amounts of physical and psychological impacts of traveling, given by $I_{physical}^u$, $I_{psychological}^u$.

The value of degrees of bumpiness could be determined through willingness to pay investigations.

\item Environment preservation (noise)

The environment preservation impact type represents the social impact due to the user coming in contact with sound emissions. It is meant to capture the changes that occur in the interactions between people due to sound emissions, e.g. the inability to communicate between driver and passenger while driving.

The impact indicator is the amount of sound emissions to which the directly affected public is exposed $I_{noise}^u$.

The value of an amount of sound emissions can be determined through willingness to pay investigations.

\end{enumerate}

\paragraph{\underline{Directly affected public (DAP)}} \label{2dap}

The impacts attributed to the directly affected public are grouped as: safety, operation quality, and environment preservation (Table \ref{tbl:24}), similar to the impact types of the user. The reason they are handled separately is that the directly affected public is affected in fundamentally different ways than the user.

\begin{table}
\caption{Directly affected impact types}
\begin{tabular}{|p{70pt}|p{100pt}|p{70pt}|p{150pt}|}
\hline
\multicolumn{2}{|l|}{Level 1} & \multicolumn{2}{l|}{Level 2} \\ 
\hline
Label & Description & Label & Description \\ 
\hline
"Safety & the impact on the directly affected public due being involved in an accident & property damage & the \underline{economic} impact of repairing the vehicle \\ 
\cline{3-4}
(Accidents) &  & injury & the \underline{societal} impact due to the injury \\ 
\cline{3-4}
 &  & death & the \underline{societal} impact due to death \\ 
\hline
Operation quality & the impact of traveling on the directly affected public & physical & the \underline{societal} impact of experiencing vibrations in a building next to a road \\ 
\cline{3-4}
 &  & psychological & the \underline{societal} impact of having for example, anxiety due to a perceived increase in the probability of being involved in an accident, or of seeing things while being next to a road. \\ 
\hline
Environment preservation (reduction of noise) & the \underline{societal} impact due to the directly affected public coming in contract with sound emissions &  &  \\ 
\hline
\end{tabular}
\label{tbl:24}
\end{table}

\begin{enumerate}
 \item Safety (Accidents)

The safety impact type is subdivided into property damage, injury and death impact types.

\item Property damage

The property damage impact type represents the economic impact of repairing damaged property, to the condition it was prior to the occurrence of the accident, e.g. the costs of the labor, and materials required to repair a retaining wall that has been damaged in an accident. The impact indicator is the property damage cost, $I_{acc - pd}^{dap}$.

The value of this impact type can be approximated using the receipts from past repairs.

\item Injury and death

The injury impact type and the death impact type attributed to the directly affected public represent the societal impact due to injuries and deaths, respectively. They represent the change in interactions between persons that will occur because someone other than the user is injured or dead. It is not to be confused with the injury impact type and the death impact type attributed to the indirectly affected public. The impact indicators are the number of injuries and deaths incurred in a specified time interval, given by$I_{acc - injury}^{dap}$, $I_{acc - death}^{dap}$, respectively

The value of these impact types can be estimated by using willingness to pay of the directly affected public to avoid injury or death. Care must; however, be taken to avoid double counting with the injury impact type and the death impact type attributed to the indirectly affected public.

\item Operation quality (Comfort)

The operation quality impact type is subdivided into the physical and psychological impact types. The physical impact type represents the societal impact of obtaining for example, discomfort through vibrations that occur due to road use. It represents the change in interactions between persons that will occur because the physical change in the directly affected public.

The psychological impact type represents the social impact of having for example, anxiety due to a perceived increase in the probability of being involved in an accident, or of seeing the infrastructure, e.g. aesthetics. It is believed that any economic impacts to be attributed to the directly affected public due to the physical and psychological impacts of others traveling, such as the loss of productivity, are negligible.

The impact indicators are the amounts of physical and psychological impacts of traveling, given by $I_{physical}^{dap}$, $I_{psychological}^{dap}$.

The value of degrees of bumpiness can be determined through willingness to pay investigations.

\item Environment preservation (Noise)

The environment preservation impact type represents the social impact due to the directly affected public coming in contact with sound emissions. It is meant to capture the changes that occur in the interactions between people due to sound emissions, e.g. the necessity to change where people meet due to excess noise.

The impact indicator is the amount of sound emissions to which the directly affected public is exposed $I_{noise}^{dap}$.

The value of an amount of sound emissions can be determined through willingness to pay investigations.
\end{enumerate}
%
\paragraph{\underline{Indirectly affected public (IAP)}} \label{2iap}
%
The indirectly affected public are those that are affected by roads through other mediums, e.g. a person who is affected by an increase in the temperature of the earth due to the CO$_{2 }$emitted during the execution of an intervention on a road. The impacts attributed to the indirectly affected public are grouped as: safety, socio-economic activity, environment preservation, and environment consumption.

Table \ref{tbl:25} and Table \ref{tbl:26} list the most common impact types and indicators. It is noted that several impact types, such as gas and particle emission also directly affect the users and directly affected public group. However, it is assumed that these impacts are minimal.

\begin{table}
\caption{Indirectly affected public impact types (safety, socio-economic activity)}
\begin{tabular}{|p{70pt}|p{70pt}|p{70pt}|p{70pt}|p{70pt}|p{70pt}|}
\hline
\multicolumn{1}{|l}{Level 1} &  & \multicolumn{1}{l}{Level 2} &  & \multicolumn{1}{l}{Level 3} &  \\ 
\hline
Label & Description & Label & Description & Label & Description \\ 
\hline
Safety (Accidents) & The impact on the indirectly affected public of accidents occurring on roads & injuries & the economic impact due to an injury &  &  \\ 
\cline{3-6}
 &  & deaths & the economic impact due to a death &  &  \\ 
\hline
Socio-economic activity & The impact on the ...due to changes in socio-economic development & Persons & the impact of not  being able to transport people & Productivity & the economic impact due to not being able to transport people, e.g. a crop cannot be harvested as workers cannot be transported to the field \\ 
\cline{5-6}
 &  &  &  & Health & the societal impact due to injuries and deaths of not being able to get proper medical care \\ 
\cline{3-6}
 &  & Goods & the impact of not being able to move goods & Productivity & the economic impact due to not being able to deliver goods, e.g. because of not being able to work as planned \\ 
\cline{5-6}
 &  &  &  & Health & the societal impact due to not being able to deliver goods, e.g. due to deaths because of lack of food or medical supplies \\ 
\cline{3-6}
 &  & Employment & social and economic impact of interventions in terms of employing people &  &  \\ 
\hline
\end{tabular}
\label{tbl:25}
\end{table}

\begin{table}
\caption{Indirectly affected public impact types (environment preservation, environment consumption)}
\begin{tabular}{|p{70pt}|p{70pt}|p{70pt}|p{70pt}|p{70pt}|p{70pt}|}
%\begin{tabular}{|l|l|l|l|ll|}
\hline
\multicolumn{1}{|l}{Level 1} &  & \multicolumn{1}{l}{Level 2} &  & Level 3 &  \\ 
\hline
Label & Description & Label & Description & \multicolumn{1}{l|}{Label} & Description \\ 
\hline
"Environment preservation & The impact on ... due to the environment being impacted by particle emissions & CO2 & the impact due to the emissions  & \multicolumn{1}{l|}{Production} & the environmental impact of emissions emitted during the production of materials \\ 
\cline{5-6}
(Reduction of gas and particles emissions) &  &  &  & \multicolumn{1}{l|}{Material transport} & the environmental impact of emissions emitted during the transport of materials to and from the construction site  \\ 
\cline{5-6}
 &  &  &  & \multicolumn{1}{l|}{Person transport} & the environmental impact of emissions emitted during travel \\ 
\cline{5-6}
 &  &  &  & \multicolumn{1}{l|}{Health} & the societal impact due to emissions (human health) \\ 
\cline{3-6}
 &  & PM10 & Same as for CO2 &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & nitrogen &  &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & carbon monoxide &  &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & aldehydes &  &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & nitrogen dioxide &  &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & sulphur dioxide &  &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & polycyclic aromatic hydrocarbons &  &  &  \\ 
\cline{3-3}\cline{5-6}
 &  & dust &  &  &  \\ 
\hline
Environment consumption & The impact on ... due to the slower depletion of finite amounts of non renewable  resources & Energy & the environmental impact due to the consumption of energy not related to emissions, e.g. depletion of finite amounts of non-renewable energy sources &  &  \\ 
\cline{3-6}
 &  & Materials & the environmental impact of consuming materials, not related to emissions  &  &  \\ 
\cline{3-6}
 &  & Land & The land impact type represents the environmental impact due to the consumption of land not related to emissions &  &  \\ 
\cline{3-6}
 &  & Culture & The culture impact type represents the societal impact of changing things important to our identity (of which heritage is part) &  &  \\ 
\hline
\end{tabular}
\label{tbl:26}
\end{table}

\begin{enumerate}
 \item Safety (Accidents)
The injury impact type and the death impact type attributed to the indirectly affect public represents the economic impact due to injuries and deaths, respectively. They represent the loss in productivity due to injuries and deaths, respectively. It includes changes to human activity, such as a doctor's time in an emergency room and the time required ensuring that an insurance company conducts the required financial transactions. The impact indicator is the amount of work time lost, when compared to the reference case where the accident had not occurred given by $I_{safety\_time}^{iap}$.

The value of each amount can be determined by estimating the loss of productivity of the person involved in the accident.

\item Socio-economic activity

The socio-economic activity impact type represents the contribution of the road to socio-economic development. It is composed of persons, goods and employment impact types.

\item Persons

The person impact type is further divided into a productiveness impact type and a health impact type.

\begin{itemize}
\item Productivity

The productiveness impact type represents the economic impact due to not being able to travel, e.g. a farmer cannot harvest his entire crop because he needs to spend a significantly larger portion of his time getting his goods to market. The impact indicator is the amount of lost work, given by $I_{trans - p - wt}^{iap}$, and expressed in units of time.

The value of each amount can be determined by conducting through simulations of the performance of the region.

\item Health

The health impact type represents the societal impact due to injuries and deaths that occur due to not being able to obtain standard medical care, due to a shortage of available persons. The impact indicators are the number of injuries and deaths incurred in a specified time interval, given by$I_{trans - p - injury}^{iap}$, $I_{trans - p - death}^{iap}$, respectively. The value of each amount can be determined by conducting through simulations of the region.
\end{itemize}

\item Goods

The goods impact type is also further divided into a productiveness impact type and a health impact type.

\begin{itemize}
\item Productivity

The productiveness impact type represents the economic impact due to not being able to deliver goods, e.g. a farmer cannot plant his crop on time since fertilizer could not be delivered as planned. The impact indicator is the amount of lost work, given by $I_{trans - g - wt}^{iap}$.and expressed in units of time.

\item Health

The health impact type represents the societal impact due to injuries and deaths due to goods such as food or medical supplies not being delivered as planned, e.g. the change in society that occurs due to the death of someone in a hospital that would not have died if medical supplies had been delivered as planned. The impact indicators are the number of injuries and deaths incurred in a specified time interval, given by$I_{trans - g - injury}^{iap}$, $I_{trans - g - death}^{iap}$, respectively.
\end{itemize}

\item Employment

The employment impact type represents the societal impact of executing interventions in terms of employing people that is not captured by the impact type attributed to the owner due to the execution of interventions or the user due to the maintenance of vehicles used for traveling. It includes economic development. The impact indicator is the amount of work provided given by $I_{work}^{idap}$.

The value can be estimated as using economic impact assessment models such as those proposed in \cite{cubrc2001}; \cite{Davis2001}; \cite{Kumares2007}, using predictions of business output, value added, employment level, wages and salaries, and wealth are made.

\item Environment preservation (emissions)

The environment preservation (emissions) impact type of the indirectly affected public represents the environmental and societal impact due emissions emitted during the production and transport of materials and persons.

It is subdivided by particle emitted, e.g. CO$_{2}$, PM10, NO, CO, aldehydes, NO$_{2}$, SO$_{2}$, polycyclic aromatic hydrocarbons and dust. The impact indicators are the amounts of each that are emitted, given by $I_{C{O_2}}^{iap}$., $I_{P{M_{10}}}^{iap}$, $I_{CO}^{iap}$, $I_{aldehydes}^{iap}$, $I_{N{O_2}}^{iap}$, $I_{S{O_2}}^{iap}$, $I_{pah}^{iap}$,  $I_{dust}^{iap}$ .

Each of these are further subdivided into

\begin{itemize}
	\item[-] the production impact type, which represents the \underline{environmental} impact of emissions emitted during the production of materials
	\item[-] the material transport impact type, which represents the \underline{environmental} impact of emissions emitted during the transport of materials to and from the construction site
	\item[-] the person transport impact type, which represents the \underline{environmental} impact of emissions emitted during travel
	\item[-] the health transport impact type, which \underline{societal} impact due to emissions (human health). It is meant to capture the changes that occur in the interactions between people due the changes in the people, e.g. due to sickness.  
\end{itemize}

The value of each amount can be determined by analyzing historical records or by conducting empirical studies using emission measurement tools and instruments.

\item Environment consumption

The environment consumption impact type represents the depletion of finite amounts of non-renewable resources. It is subdivided into the energy impact type, the material impact type, the land impact type and the culture impact type.

\begin{itemize}
\item Energy

The energy impact type represents the environmental impact due to the consumption of energy not related to emissions, e.g. depletion of finite amounts of non-renewable energy sources. The impact indicator is a measure of the significance of a unit depletion of a finite resource, given by $I_{energy}^{iap}$.

\item Material

The material impact type represents the environmental impact of consuming materials, not related to emissions, e.g. the consumption of wood has an impact on woodland areas. The impact indicator is the amount of consumed material, e.g. kg of wood cut, given by $I_{materials}^{iap}$.

\item Land

The land impact type represents the environmental impact due to the consumption of land not related to emissions, e.g. increased environmental damages due to floods. The impact indicator is the m$^{2}$ of land converted from a natural state to a built state, given by $I_{land}^{iap}$.

\item Culture

The culture impact type represents the societal impact of changing things important to our identity (of which heritage is part).The impact indicator is a measure of the significance of an object to our identity, given by $I_{culture}^{iap}$. The value can be estimated through willingness to pay investigations. 
\end{itemize}

\end{enumerate}

\subsection{Question B}

In reading such an impact hierarchy some people will imagine things that may seemingly not be in the list of impacts. To help clarify this, explain why your impact hierarchy does not include an impact type ``access'' or an impact type ``natural hazard risk''.

\subsection{Answer B}

\subsubsection{Access}

The improvement of access is often stated as an impact \citep{Kumares2007}. Improved access, however, is a proxy for the many things that actually happen when access is improved. For example, if access to a highway is improved a user traveling from point A, to point B, who normally would not take the highway, would perhaps have a shorter travel time, reduced vehicle costs, and reduced accident costs. It is proposed that if it is desired to approximate the value of improved access that the changes in the relevant impact types for each stakeholder are calculated.

A more numeric example is as follows (Table \ref{tbl:27}). If there are two interventions being considered for a highway (intervention A with a medium level of access and intervention B with a high level of access), it is possible that intervention B will result in a shorter distance for users to travel to arrive at their destinations in comparison with intervention A. This in turn results in a reduction in:

\begin{itemize}
	\item[-] accident costs of 2 mus, 
	\item[-] comfort costs of 5 \textit{mus},
	\item[-] travel time costs of 4 \textit{mus}, and
	\item[-] vehicle costs of 2 \textit{mus}.
\end{itemize}

\begin{table}
\caption{Access: Example}
\begin{tabular}{|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Interventions} & \multicolumn{1}{c|}{Accident} & \multicolumn{1}{c|}{Comfort} & \multicolumn{1}{c|}{Travel Time} & \multicolumn{1}{c|}{Vehicle} \\ 
\hline
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{8} \\ 
\hline
\multicolumn{1}{|c|}{B} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{6} \\ 
\hline
\end{tabular}
\label{tbl:27}
\end{table}

This means that intervention B is 13 \textit{mus} better than intervention A, or in other words, the improvement in access has a value of 13 \textit{mus}.

\subsubsection{Natural hazard risks}

The reduction of natural hazard risks is often stated as an impact. The reduction of natural hazard risk is, however, a proxy for the many things that may actually happen when the risk of natural hazards is reduced. It is proposed that if it is desired to approximate the value of reduced natural hazard risk that the changes in the relevant impact types (probability of occurrence and values) for each stakeholder be estimated.

For example, if there are two interventions (Table \ref{tbl:28}) being considered for the modification of a section of highway, one of them (A) may make it so that it is much quicker to restore service following the interruption of traffic flow than the other (B). Based on exactly how the interventions are planned and executed this might result in increased risk to the user as follows:

\begin{itemize}
	\item[-] accident costs of 1.5 \textit{mus} (0.1 x 15 \textit{mus}), 
	\item[-] comfort costs of 2.4 \textit{mus} (0.1 x 24 \textit{mus}),
	\item[-] travel time costs of 11.3 \textit{mus} (0.1 x 113 \textit{mus}), and
	\item[-] vehicle costs of 5.7 \textit{mus} (0.1 x 57 \textit{mus}).
\end{itemize}
\begin{table}
\caption{Natural hazard risk: example}
\begin{tabular}{|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Interventions} & \multicolumn{1}{c|}{Accident} & \multicolumn{1}{c|}{Comfort} & \multicolumn{1}{c|}{Travel Time} & \multicolumn{1}{c|}{Vehicle} \\ 
\hline
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{8} \\ 
\hline
\multicolumn{1}{|c|}{B} & \multicolumn{1}{c|}{11.5} & \multicolumn{1}{c|}{17.4} & \multicolumn{1}{c|}{25.3} & \multicolumn{1}{c|}{13.7} \\ 
\hline
\end{tabular}
\label{tbl:28}
\end{table}
\section{Changes in LOSs - Theory}

\subsection{General}

It is important to note that neither the required nor the provided LOSs are constant over time. Provided LOSs change due to deterioration of the infrastructure. Required LOSs change due to changing needs for the infrastructure. This is illustrated in Figure \ref{fig21}.

\begin{figure}[h]
\includegraphics[width=391pt]{fig21.eps}
\caption{Illustration of changes in required and provided LOS}\label{fig21}
\end{figure}

An example to help the interpretation of Figure 1 is as follows. At t = 0, a road link consisting of a bridge and a road section starts to provide a LOS that is higher than the initially agreed upon required LOS (e.g. the travel time costs required per vehicle to go from A to B should be below x, and earthquake risks should be below y). To ensure this, the road is to have a roughness above 20 mm/m, and the bridge is designed to withstand an earthquake of magnitude 4 with a very low probability of damage. Over time both objects deteriorate and the provided LOS decreases, i.e. due to the worsening road condition vehicles on average take more time to go from A to B, and the earthquake risks increases because the probability of the bridge withstanding an earthquake of magnitude 4 with no damage decrease.

As the infrastructure manager is aware that the provided LOS is decreasing over time, he executes two interventions a few years apart to restore the provided LOS to the initially agreed required level. The first time due planning problems the provided LOS falls below the initially agreed required LOS. The second time this does not happen.

Following the second intervention, the government introduces a new law, which requires that all bridges be able to resist an earthquake of a magnitude of 8 with no interruption to traffic flow, i.e. the earthquake risks are to be reduced. The infrastructure manager then executes an intervention to bring the provided LOS up to the initially required LOS for the road section and the bridge up to, and beyond, the new required LOS perfectly coordinated with the wishes of the government.

As deterioration of the road section and bridge continue, the infrastructure manager executes another intervention to restore the provided LOS to the new required LOS, but again due to planning problems later than he should have.

Immediately after the execution of the intervention, the public votes to have the speeds on the road increased from 100 to 120 kph, and there is a very short time frame given for the infrastructure manager to respond. It is also decided that the required LOS should be fixed not only with the traffic seen today but taking into consideration the growth in the future. 
The infrastructure manager responds with an intervention to expand the road at the same time improving the condition of both the road and the bridge and establishes a plan to execute interventions at appropriate times in the future so adequate LOSs are provided. 
%

\subsection{Ability to provide LOSs (deterioration)}

Deterioration is the process of becoming worse. The deterioration of infrastructure is the process with which infrastructure loses its ability to provide a specific LOS.

From an infrastructure managers perspective, deterioration processes can be classified as either manifest or latent \cite{Lethanh2015}. Manifest processes are processes that are observable under the inspection strategies being followed, so that there is enough time for a manager to execute an intervention so as ensure that the infrastructure provides an adequate LOS.

Latent processes are processes that are unobservable under the inspection strategies being followed, so that there is enough time for a manager to execute an intervention so as ensure that the infrastructure provides an adequate LOS.

Although it does depend on the monitoring strategies in place, typical examples of manifest processes are pavement cracking, the leaching of mortar between masonry blocks and chloride induced corrosion, where the corrosion is expansive. Typical examples of latent processes are those that result in floods, earthquakes, and the overloading of infrastructure. Other examples are given in Table \ref{tbl:29}. An example of a process that is not as clear is the fatigue of metal.

\begin{table}
\caption{Examples of manifest deterioration processes}
\begin{tabular}{|l|p{120pt}|p{200pt}|}
\hline
Process & Subprocess & Comments \\ 
\hline
Concrete surface wear & Abrasion & moving objects in contact with concrete which can wear away and often desired roughness \\ 
\hline
Aggregate deterioration & Alkali-silika reaction & a reaction between the hydroxyl ions in the alkaline cement pore solution in the concrete and reactive forms of silica in the aggregate form a gel is produced, which increases in volume by taking up water and so exerts an expansive pressure \\ 
\cline{2-3}
 & Freezing and thawing & under the humid conditions pores may be water-filled and under freezing conditions, water in these pores may freeze and thaw, which causes immense hydraulic pressure, which cracks concrete \\ 
\hline
Cement paste deterioration & Deterioration due to sulphates  & exposure to industrial and agricultural chemicals can result in the break-down of the cement paste, the glue that holds concrete together \\ 
\hline
Reinforcement corrosion & Chloride induced corrosion of steel & Corrosion of steel can cause the production of a rust product that is significantly larger than the normal steel cracking the surrounding concrete and destroying the bond between concrete and steel. There may also be section loss of the steel reinforcement. \\ 
\cline{2-3}
 & Carbonation induced corrosion of steel & Carbonation, or neutralization, is a chemical reaction between carbon dioxide in the air with calcium hydroxide and hydrated calcium silicate in the concrete.  \\ 
 &  & The carbon dioxide in the air reacts with the alkali in the cement and makes the pore water more acidic, thus lowering the pH.  \\ 
\hline
\end{tabular}
\label{tbl:29}
\end{table}

In order to understand how the provided LOS changes over time it is necessary to understand the deterioration processes themselves, how they affect the materials in which they are occurring, as well as how the changes being caused in the materials affect the functionality of the elements composed of these materials, the objects composed of these elements, and the networks composed of these objects.

\subsubsection{Depth of process analysis}

There is, of course, considerable depth with which deterioration processes can be analyzed. A few examples are given in Table \ref{tbl:29}. Even the lowest levels shown there, however, can be analyzed in more detail. For example, to analyze chloride induced corrosion of steel one can investigate, or model, the processes shown in Table \ref{tbl:210}. Much more information can be found in \cite{Elsener2013} for the interested reader.

{\raggedright

\begin{table}[h]
\caption{Examples of the subprocesses within the chloride induced corrosion of steel process for a reinforced concrete bridge}

\vspace{3pt} \noindent
\begin{tabular}{|p{99pt}|p{122pt}|p{189pt}|}
\hline
\parbox{99pt}{\raggedright 
Subprocesses
} & \parbox{122pt}{\raggedright 
Sub-subprocesses
} & \parbox{189pt}{\raggedright 
Description
} \\
\hline
\parbox{99pt}{\raggedright 
spreading processes
} & \parbox{122pt}{\raggedright } & \parbox{189pt}{\raggedright 
in which chlorides are spread on a concrete bridge
} \\
\hline
\parbox{99pt}{\raggedright 
transport processes
} & \parbox{122pt}{\raggedright } & \parbox{189pt}{\raggedright 
in which the chlorides enter the bridge
} \\
\hline
\parbox{99pt}{\raggedright } & \parbox{122pt}{\raggedright 
capillary absorption
} & \parbox{189pt}{\raggedright 
in which substances in an aqueous solution get into the concrete near the surface
} \\
\hline
\parbox{99pt}{\raggedright } & \parbox{122pt}{\raggedright 
diffusion
} & \parbox{189pt}{\raggedright 
in which chloride ions due to different concentration of substances or different pressures within the concrete move within the concrete
} \\
\hline
\parbox{99pt}{\raggedright } & \parbox{122pt}{\raggedright 
ion migration in electric fields
} & \parbox{189pt}{\raggedright 
in which chloride ions move within the concrete
} \\
\hline
\parbox{99pt}{\raggedright 
depassivation process
} & \parbox{122pt}{\raggedright } & \parbox{189pt}{\raggedright 
in which the steel in the concrete is depassivated which in turn allows corrosion to start if moisture and oxygen are present
} \\
\hline
\parbox{99pt}{\raggedright 
corrosion process
} & \parbox{122pt}{\raggedright } & \parbox{189pt}{\raggedright 
In which RUST is formed
} \\
\hline
\parbox{99pt}{\raggedright } & \parbox{122pt}{\raggedright 
cathodic (oxidation) process
} & \parbox{189pt}{\raggedright 
in which oxidizing iron produces iron(II) ions and electrons , and electrons combine with water and oxygen to form hydroxide ions, $2{{\rm{H}}_{\rm{2}}}{\rm{O + }}{{\rm{O}}_{\rm{2}}}{\rm{ + }}4{{\rm{e}}^ - } \to 4{\rm{O}}{{\rm{H}}^ - }$
} \\
\hline
\parbox{99pt}{\raggedright } & \parbox{122pt}{\raggedright 
anodic (reduction) process
} & \parbox{189pt}{\raggedright 
hydroxide ions react with iron(II) ions to form Iron(II) hydroxide $2{\rm{F}}{{\rm{e}}^{{\rm{2 + }}}}{\rm{ + 4O}}{{\rm{H}}^ - } \to 2{\rm{Fe}}{({\rm{OH)}}_2}$, and Iron(II) hydroxide reacts with oxygen to form ${\rm{4Fe + 3}}{{\rm{O}}_{\rm{2}}} \to {\rm{F}}{{\rm{e}}_{\rm{2}}}{{\rm{O}}_3}$rust
} \\
\hline
\end{tabular}
\vspace{2pt}
\label{tbl:210}\end{table}

}
\subsubsection{Relationship deterioration process - LOS}

A simple example of how deterioration processes change the provided LOS is illustrated in Table \ref{tbl:211} and Figure \ref{fig22} for the chloride induced corrosion of reinforced concrete bridge. It can be seen that a change in the phase of a deterioration process does not necessarily mean that there is a change in the LOS. The relationship depends on the how the processes affect the materials, elements, objects and networks. As phases of deterioration are often expressed as condition states and LOSs are sometimes expressed as performance states, the relationship may also be considered as a condition state - performance state relationship. An example is given in Table \ref{tbl:212} for a steel element.

\begin{table}
\caption{Illustration of relationship between a deterioration process and provided LOSs}
\begin{tabular}{|l|p{110pt}|p{190pt}|l|l|}
\hline
\multicolumn{2}{|l|}{Deterioration process phases} & Visual condition indications & \multicolumn{1}{c|}{\hspace{2mm}Time \hspace{2mm}} & \multicolumn{1}{c|}{Provided LOSs} \\ 
\hline
\multicolumn{1}{|c|}{\hspace{2mm}  1 \hspace{2mm} } & Penetration of water and de-icing salts into surface concrete & None & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{2} & Breakdown of depassivation layer of top reinforcement & None & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{3} & Rusting of top reinforcement & None & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{4} & Cracking and spalling of surface concrete & Small pockmarks on the surface, often accompanied with a rust stain, and usually at a point corresponding to the location of a top reinforcing bar & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{5} & Shifting of the compressive zone in the concrete deck & Spalling in multiple locations and some of these locations had significant amounts of exposed reinforcement & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1} \\ 
\multicolumn{1}{|c|}{} &  & Uniformly spaced cracks on the underside of the slab, beneath spalled areas & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{6} & Over-stressing of the bottom reinforcement & No additional visual indications & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{7} & Cracking of the bottom concrete & Spalling in numerous widely dispersed locations over relatively large areas, more intense under the truck lanes. & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{8} & Punching failures & "Spalling in numerous widely dispersed locations over relatively large areas, extensive cracking was observed & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{2} \\ 
\multicolumn{1}{|c|}{} &  & two punching failures of the deck slab occurred" & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{9} & Collapse & - & \multicolumn{1}{c|}{?} & \multicolumn{1}{c|}{3} \\ 
\hline
\end{tabular}
\label{tbl:211}
\end{table}

\begin{figure}[h]
\includegraphics[width=376pt]{fig22.eps}
\caption{Illustration of relationship between deterioration processes and the provided LOS}\label{fig22}
\end{figure}

\subsection{Required LOSs}

Changes in required LOSs are the result of changing human needs. They often result in existing infrastructure providing an inadequate LOS. For example, if a bridge was designed to have two lanes but a decision was made to expand the road to four lanes to accommodate changing human needs, then the bridge would no longer provide an adequate LOS.

From an infrastructure managers perspective, processes that result in changes in required LOSs, as deterioration processes, can be classified as either manifest or latent.

With respect to changes in required LOSs most processes are considered manifest if development strategies are well developed and triggers for the execution of interventions are determined. For example, the need to expand a road due to it reaching its carrying capacity in terms of number of vehicles in peak periods can be considered as a manifest process if the number of vehicles per hour are recorded every day and a development strategy has been determined that states that once the road is at 80\% its maximum capacity the road will be widened.

Most processes are considered as latent if development strategies are not well developed and no triggers for the execution of interventions are determined. For example, if the infrastructure manager is focused simply on maintaining the physical condition of the road it may come to him as a surprise when suddenly politicians request that a certain road be widened.

Some example of processes that can result in changes in the required LOS in buildings can be are shown in Table \ref{tbl:212}.

\begin{table}
\caption{Example changes for buildings and related processes}
\begin{tabular}{|l|l|ll}
\cline{1-2}
Change & Related process &  &  \\ 
\cline{1-2}
Floor space layout & Demand for apartments suitable for growing elderly population &  &  \\ 
\cline{1-2}
New windows & Demand for energy efficient houses and rises.  &  &  \\ 
\cline{1-2}
New heating system & Demand for less maintenance on the heating system &  &  \\ 
\cline{1-2}
New control system & Demand for better control over the quality of the indoor air condition. &  &  \\ 
\cline{1-2}
\end{tabular}
\label{tbl:212}
\end{table}

Changes to the required LOS can happen due the wishes of the owners of infrastructure or be imposed by external persons such as government. The government can also provide incentives for the owner to change the required LOS. Other stakeholders can exert influence on the owner to change the required LOS, e.g. tenants in an apartment building can threaten not to pay the rent unless an air conditioning system is installed. The development of new technologies can also instigate changes in the required LOS as new developments may make things possible that were not before, e.g. new heating systems.

\subsection{Methodology}

The five basic steps to modelling changes in the required LOS are (adapted from \cite{Neufville2011}):

\begin{itemize}
	\item \underline{Step 1}:  Identify key performance drivers, i.e. determine the things that if changed will result in a change in the required LOS, e.g. if the number of vehicles per day pass x, then the road will need to be widen.
	\item \underline{Step 2}: Analyse historical trends, i.e. obtain as much historical data as possible and model how it changed over time with the plan to use this model to predict the future values, e.g. obtain the evolution of vehicles on the road over the last 20 years.
	\item \underline{Step 3}: Identify trend breakers, i.e. dig deeper and try to determine if there is something that can be used to help you predict if this trend will continue or might change suddenly, e.g. a new rail link on which passenger trains will circulate is about to be completed where there was none before.
	\item \underline{Step 4}: Establish forecast accuracy, i.e. if enough data exists put yourself mentally in the past and see how could you have predicted the future for which you have data, e.g. using only traffic data from 1980 to 1999 how could would you be at predicting the traffic between 2000 and 2014. 
	\item \underline{Step 5}: Build a dynamic model, i.e. build a model that allows a range of predictions to be made using a sensible distribution. 
\end{itemize}

An example for a hospital is provided in  the work of \cite{Neufville2011} .

More detail on model development will be provided later.

\section{Change in LOS-Example}

\subsection{Question C}

Using the owner and user impacts of the impact hierarchy developed in section 3, develop simple models to be used to estimate how the provided LOS of a road will change over 20 years to the relatively slow deterioration of the pavement surface.

\subsection{Answer C}

\subsubsection{Background}

In order to develop simple models to quantify the values of the various impact types it is useful to think of the object under investigation as being in one of two states, i.e. when no intervention is being executed and when an intervention is being executed. In general, these can be represented by the functions $f_{}^k(t,x)$ and $g_{}^k(d,x)$, respectively (here, \textit{k} represents intervention to be executed, e.g. renewal of the road section; \textit{x} represents the performance indicator, \textit{t} and \textit{d} represent in between interventions and during intervention, respectively). It is assumed that they are expressed in \textit{mus}. Both can be given by functions that vary over time. For example, two following generic exponential functions which can be used are:

\begin{eqnarray}
      && {f^k}(t,x) = {a^{k,x}} + {b^{k,x}} \cdot \exp ({\beta ^{k,x}} \cdot t) \label{fx}\\
      && {g^k}(d,x) = {\bar a^{k,x}} + {\bar b^{k,x}} \cdot \exp ({\bar \beta ^{k,x}} \cdot {d^k}) \label{gx}
\end{eqnarray}

Where  or ${f^k}(t,x)$ and $g_{}^k(d,x)$ represent the impacts (costs) that are incurred as a function of time (\textit{t} or \textit{d}) and parameters $a$, $b$, and $\beta $.

If it is assumed that the change in condition of the road over time is known, this equation links the changes any impact on any stakeholder to the change in condition of the road over time.

An example of the evolution of vehicle operation cost (VoC) along with the evolution of roughness for 1 m length of road is shown in Figure \ref{fig23}. In the figure, Eq. (1) has parameter values \textit{a}=500, \textit{b}=800, and $\beta  = 0.095$. These values are often estimated using empirical models. For example, in order to estimate the evolution of the VoC in between two consecutive interventions (from ${t_1}$ to ${t_2}$), \cite{Ouyang2004} used the following function.

\begin{eqnarray}
      && \int\limits_{{t_1}}^{{t_2}} {{f^k}\left[ {s(t)} \right]} {e^{ - rt}}dt = c\int\limits_{{t_1}}^{{t_2}} {s(t){e^{ - rt}}} dt \label{stx}
\end{eqnarray}

In Eq. \eqref{stx}, $s(t)$ represents the roughness indicators; c is constant associated with cost; $r$ is discount factor. The functional form of the roughness can be estimated with a simple exponential function as follows:

\begin{eqnarray}
      && s(t) = s({t_0}) \cdot \exp \left[ {\alpha (t - {t_0})} \right] \hspace{2mm} (for \hspace{2mm} t \ge {t_0}) \label{stx1}
\end{eqnarray}

Where $s({t_0})$ is initial roughness indicator (the initial LOS) and $\alpha $is a deterioration parameter.

When the time \textit{t} is discrete, \cite{Ouyang2004} suggests to use following function

\begin{eqnarray}
      && s(t + 1) = \left[ {s(t) + \upsilon } \right] \cdot \exp (\alpha ) \label{stx2}
\end{eqnarray}

With $\upsilon $ is the pre-defined model's parameter.

The cost curve (green curve in Figure \ref{fig23}) represents the increase of vehicle operation cost over time. The discount rate used is 2\%. Evolution of roughness is estimated from using Eq. \eqref{stx2} with $\upsilon  = 2$ and $\alpha  = 0.0153$. Value of parameter $c$in Eq. \eqref{stx} is 2.1.

\begin{figure}[h]
\begin{center}
\includegraphics[width=402pt]{fig23.eps}
\caption{Evolution of VoC and roughness indicator}\label{fig23}
\end{center}
\end{figure}

Being generic, Eqs. \eqref{fx} and \eqref{gx} can be used to model the evolution of any impact that varies as a function of the condition of a road section, and allows the heterogeneity of each road section, e.g. the different deterioration rates, to be taken into consideration. For example, if it is desired to model the relationship between impacts on the user and the condition of a road section, the values of a, b, and \textit{$\beta{}$} (e.g. for user) could be selected as non-negative values, ensuring that the function \textit{f }will result in the values of the impact on the users that increase exponentially with time. This is similar to that of the user cost estimated using Eq. \eqref{stx}. Although the form of functions \textit{f} and \textit{g} are flexible, here, the exponential form is often used. 
%%%
\subsubsection{Owner}
%%%
\begin{eqnarray}
      && f_{}^k(t,x_i^o) = \sum\limits_{i = 1}^3 {I_i^o(t) \cdot c_i^o(t)} \label{fx1}\\
      && g_{}^k(d,x_i^o) = \sum\limits_{j = 1}^3 {I_i^o(d) \cdot c_i^o(d)} \label{gx1}
\end{eqnarray}
Where:\\
%\begin{flushright}
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$i$:] is the index of impact types: labor, equipment, and materials
\item[$I_i^o$:] is the owner impact indicator for impact type i. In this case it is the number of person-hours, numbers of equipment use hours, and quantity of materials used for intervention
\item[$c_i^o(t)$ and $c_i^o(d)$:] are owner unit costs, which vary as a function of time $t$ and $d$
\end{description}
%\end{flushright}
\end{adjustwidth}
\subsubsection{Users}
\begin{eqnarray}
      && f_{}^k(t,x_i^u) = \sum\limits_{i = 1}^3 {I_i^u(t) \cdot c_i^u(t)} \label{fx2}\\
      && g_{}^k(d,x_i^u) = \sum\limits_{i = 1}^3 {I_i^u(d) \cdot c_i^u(d)} \label{gx2}
\end{eqnarray}
Where:
%\begin{flushright}
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$i$:] is the index of impact types
\item[$I_i^u$:] is the user impact indicator for impact type $i$ 
\item[$c_i^u(t)$ and $c_i^u(d)$:] are user unit costs, which vary as a function of time $t$ and $d$.
\end{description}
%\end{flushright}
\end{adjustwidth}
%%%
\paragraph{\underline{Safety}}
The values of impact indicators ($I_i^u(t)$and $I_i^u(d)$) are generally estimated through regression analysis using empirical models. In empirical models, values of impact indicators associated with accidents can be estimated by using the ``accident rate'' multiplied by the cost of the accidents if they occur. Both depend on multiple factors such as the condition of the road, daily traffic volume, the physical condition of the driver of the vehicle and environmental factors such as, the occurrence of poor weather that can effect accident rate. For example, a popular way to estimate the value of impact indicators is by use of following equation \citep{Lindenmann2008}.

\begin{eqnarray}
      && I_{property}^u = t \cdot DTV \cdot s \cdot \theta  \cdot \omega  \cdot \upsilon \label{safety1}
\end{eqnarray}
Where:
%\begin{flushright}
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$t$:] is the number of days,
\item[$DTV$:] is the daily traffic volume,
\item[$s$:] is the length of the object,
\item[$\theta$:] is the coefficient depending on the deterioration level of the civil infrastructure,
\item[$\omega$:] is the correction factor for accident rate, and
\item[$\upsilon$:] is the approximate numbers of vehicles involved in an accident.
\end{description}
%\end{flushright}
\end{adjustwidth}
According to \cite{tarko2000}, three ways commonly used to estimate the accident rate are 
\begin{enumerate}
	\item crash rate method, which recommends a rate of crash for a specific type of road, in specific location, and with specified level of traffic volume; 
	\item crash equation method, which relies on a functional form of traffic volume and engineering factors; 
	\item crash reduction factor method, which is an extension method of crash equation method, where the crash reduction factor is given by: 
\end{enumerate}

\begin{eqnarray}
      && f(X) = k \cdot Y \cdot {Q^\gamma } \cdot {e^{\beta x}} \label{safety2}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$f(X)$:] is the crash reduction function,
\item[$k$:] is a constant,
\item[$Y, Q$:] are the exposure variables representing the temporal span of data and indicate the section length and traffic volume, respectively, 
\item[$\beta$:] is the an unknown parameter associated with the variable X.
\item[$X$:] is the variable vector of engineering factors, including the performance condition of the road.
\end{description}
%\end{flushright}
\end{adjustwidth}
After obtaining the accident rate related to the execution of an intervention for each intervention type on an each object (i.e. between and during interventions), the value of impact indicators $I_i^u(t)$ and can be calculated.

For example, after the execution of an intervention, an asphalt road section of 20 km is renewed to be in condition state 1. The daily traffic volume is 600 vehicles/day. The accident rates during and after intervention are estimated to be 0.03\% and 0.005\% out of the total traffic volume in a day. The duration of the intervention is 30 days. It is assumed that the average number of vehicles involved in an accident are two. 
%
The expected total numbers of vehicles involved in accidents in a year after intervention will be:

$I^u_{property}(t=365) \hspace{1mm}days \hspace{1mm} =365\times600\times0.005\%\times2 \approx 22 \hspace{1mm}$ vehicles 
%

The expected total numbers of vehicles involved in accidents during intervention will be:

$I^u_{property}(d=30) \hspace{1mm}days \hspace{1mm} =30\times600\times0.03\%\times2 \approx 11 \hspace{1mm}$ vehicles

The average unit cost $c_{{\rm{property}}}^u(t)$per vehicle can be approximated from historical data \cite{Lindberg1999}.

\paragraph{\underline{Operation efficiency}}

The values of the travel time impact indicator can be approximated as a function of multiple factors including vehicle speed, amount of congestion, and curviness of the road.

The values of the vehicle operation and maintenance impact indicators can be approximated as a function of multiple factors including the total number of vehicle type \textit{j}. (type $j$ means type of vehicles such as car, truck, bus, etc), 
\begin{eqnarray}
      && f^{k_{n_l}}_{n_l}(t,x^u_i)=I^u_{i,k_{n_l}} = \sum_{i=1}^2\sum_{j=1}^J C^u_{ij}(t) \cdot VH_j (t) \label{operation1}\\
      && g^{k_{n_l}}_{n_l}(d,x^u_i)=I^u_{i,k_{n_l}} = \sum_{i=1}^2\sum_{j=1}^J C^u_{ij}(d) \cdot VH_j (d) \label{operation2}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$I_i^u$:] is the impact indicator for the index \textit{i}
\item[$i$:] is the index for vehicle maintenance cost and vehicle operation cost.
\item[$VH$:] is total number of vehicle type \textit{j}. (type \textit{\underline{j}} means type of vehicles such as car, truck, bus, etc). The value of \textit{VH} can be obtained from examining historical record on traffic volume and annual growth of traffic volume.
\end{description}
%\end{flushright}
\end{adjustwidth}
%
The value of $C_{ij}^u(t)$can be estimated by empirical study or regression analysis based on recorded numbers of bill paid for operation and maintenance respectively. To date, several models have been developed to relate such cost to deterioration or performance of the road. For example,\cite{OpusCL1999} studied the relationship between vehicle operation cost and international roughness index. The value of \textit{VH} can be obtained from examining historical record on traffic volume and annual growth of traffic volume.
%
\paragraph{\underline{Operational Quality}}
Impact indicators associated with and  can be measured by using qualitative scale (e.g, scale from 1 to 5, with 1 is the best and 5 is the worst) or by carrying out an empirical study on the loss in effective working time if users travel on a target road link. The values of the differences between these states in these scales can be determined through willing to pay investigations.

Following proposed function can be used for estimating the values of impact indicators $BI^u_{physical, k_{n_l}}$ and $BI^u_{psychological, k_{n_l}}$.
\begin{eqnarray}
      && BI^u_{i,k_{n_l}} = t\cdot U\cdot \mu_i \label{operation3}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$t$:] is the number of days,
\item[$U$:] is expected number of users per day, which can be approximated by means of daily traffic volume,
\item[$\mu$:] is the mean value of amount of physical and psychological impacts. It can be obtained by carrying out surveys on a population sample of the users.
\end{description}
%\end{flushright}
\end{adjustwidth}

For example, a survey using qualitative method on 300 users, who travel on a certain road link of 20 km, reveals a mean value $\mu_{physical}=3.4$ (from a scale of 5) per 1 km. The daily users have a factor of 2.5 times the daily traffic volume, which is 600 vehicles/day. Thus, in 30 days, the values of $I^u_{physical}=3.4$ will be determined as:

$I^u_{physical} = 30 \times (600 \times 2.5) \times 3.4 = 153'000 $ units

If each unit of physical impact is relatively cost 0.05 $mus$, then in total 30 days, the cost incurred to users due to physical impact will be $0.05(mus/unit)\times 153'000 (units)=7'650$  (mus).

\paragraph{Environment preservation (reduction of noise)}

Impact indicators associated with $I_{noise}^u$ can be approximated by use of following equation.

\begin{eqnarray}
      && I^u_{noise} = t\cdot \overline{dBA} \cdot U \label{operation4}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$t$:] are the number of days in between interventions and during intervention,
\item[$\overline{dBA}$:] is the expected increase unit of noise (in $dBA$).
\item[$U$:] is the expected number of users within a specific period.
\end{description}
%\end{flushright}
\end{adjustwidth}

For example, an intervention is scheduled to last 30 days. The expected numbers of users could be estimated as 300 persons. If during intervention, the noise due to construction activities increases by 2 $dBA$ compared to the normal day without intervention, and one additional dBA has a value of 100 mus. Using Eq. \eqref{operation4}, the values of impact indicator will be

$I^u_{noise}=30 \times 2 \time 300 = 18'000 dBA-person-days$

and thus, the values of function $g^d(d,x^u_{noise})$ will be

$g^d(d,x^u_{noise}) = 18'000 \times 100 = 1'800'000 \hspace {1mm} mus$
%
\section{Assignments}
\subsection{Problem A}
In class an impact hierarchy was given to be used to determine optimal intervention strategies and work programs for a road network. Discuss the suitability of this impacts hierarchy if it was to be used to determine optimal intervention strategies and work programs for 
\begin{adjustwidth}{1cm}{}
\begin{itemize}
\item[a)] a private road network, and 
\item[b)] a public rail network.                         
\end{itemize}
\end{adjustwidth}
%
\subsection{Answer A-a}
No specific answer is provided. Please submit your answer to obtain feedback. 
\subsection{Answer A-b}
No specific answer is provided. Please submit your answer to obtain feedback. 
%
\subsection{Problem B}
Water is one of the fundamental needs of millions of people living in a megacity. Water of sufficient quality and quantity must be provided around the clock. In order to fulfill this need, a city depends on its water distribution infrastructure, which includes pipes made of different materials and laid at different times. These pipes are affected by processes of different types and deteriorate at different rates. The consequences related to pipe failure vary significantly depending on the type of failure, e.g. a pipe break, or a leaking pipe, as does the reaction time required to fix the pipe. For example, if a pipe breaks, a corrective intervention must be executed immediately, and if it is noticed that there is progressive water loss over time then a preventive intervention can be planned before there is an inadequate level of service.  Part of the water distribution network in mega-city Q is shown in Figure \ref{fig24}. The pipe characteristics are given in Table \ref{tbl:213}.

\begin{figure}[h]
\includegraphics[width=200pt]{fig24.eps}
\caption{Simplified network of water supply pipes}\label{fig24}
\end{figure}

\begin{table}
\caption{Pipe and their attributes}
\begin{tabular}{|l|l|l|l|l|p{80pt}|}
\hline
\multicolumn{1}{|c|}{Pipe} & Material & \multicolumn{1}{c|}{Demand (m3/day)} & \multicolumn{1}{c|}{Length (m)} & \multicolumn{1}{c|}{Year of construction} & Usage \\ 
\hline
\multicolumn{1}{|c|}{M} & Concrete & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{5'000} & \multicolumn{1}{c|}{1998} & Distribution only \\ 
\hline
\multicolumn{1}{|c|}{A.1} & PVC & \multicolumn{1}{c|}{250'000} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & Distribution only \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.2} & PVC & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & Distribution only \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.3} & PVC & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & Distribution and connection to buildings \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.4} & PVC & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & Distribution only \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.5} & PVC & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{1998} & Distribution and connection to buildings \\ 
\hline
\multicolumn{1}{|c|}{B} & Cast iron type 1 & \multicolumn{1}{c|}{200'000} & \multicolumn{1}{c|}{2'500} & \multicolumn{1}{c|}{1993} & Distribution and connection to buildings \\ 
\hline
\multicolumn{1}{|c|}{C} & Cast iron type 2 & \multicolumn{1}{c|}{100'000} & \multicolumn{1}{c|}{1'600} & \multicolumn{1}{c|}{1993} & Distribution and connection to buildings \\ 
\hline
\multicolumn{1}{|c|}{D} & Cast iron type 3 & \multicolumn{1}{c|}{350'000} & \multicolumn{1}{c|}{4'000} & \multicolumn{1}{c|}{1983} & Distribution and connection to buildings \\ 
\hline
\end{tabular}
\label{tbl:213}
\end{table}
%
\subsection{Question B}
Make an impact hierarchy that you would use to measure the performance of the network. Be complete at the highest level, i.e. the stakeholders, and be detailed for one of the stakeholders besides the owner.
\subsection{Answer B}
No specific answer is provided. Please submit your answer to obtain feedback. 
\subsection{Problem C}
\subsection{Question C}
Using the impact hierarchy for public roads in the script explain three processes that would result in a change in a required level of service defined using accident costs, travel time costs and maintenance costs. 
\subsection{Answer C}
No specific answer is provided. Please submit your answer to obtain feedback. 
%
\bibliographystyle{plainnat}
\bibliography{reference}

### 2019-05-05 InfrastructureMaintenance/Chapters/mecha-empi.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Mechanistic-empirical models and regression models }
\label{meempi} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Bryan T. Adey and Nam Lethanh}
\section{General}
\subsection{Manifest and latent processes}
Models are required to predict both the ability to provide a specified level of service (LOS) and the required LOS. The processes that result in changes to either of these can be classified as gradual (manifest) or sudden (latent), depending on how the processes are monitored, the speed with which the infrastructure manager can react and the defined LOS. The classification of a change in LOS as gradual or sudden depends on the state of preparedness and the perception of the infrastructure management owner (IMO).

A gradual change in the LOS provided is one that happens in a way that there is enough time to execute an intervention so as to ensure that the infrastructure continues to provide an adequate LOS. For example, a gradual change may be the increases in the amount of travel time due to increasingly congested roads, something that may happen at a rate of 2\% per year.

A sudden change in the LOS provided is one that happens in a way that there is not enough time to execute an intervention so as to ensure that the infrastructure continues to provide an adequate LOS. A sudden change may be a new law that allows the axle loads to increase by 50\%.

Typical models to be used to predict either the provided or the required LOS due to gradual processes are mechanistic-empirical models, regression models, Markov models, neural network models, and Bayesian networks.

Typical models to be used to predict with either the provided or the required LOS due to sudden processes are event trees and fault trees.
%
\subsection{Continuous vs. discrete states}
%
Condition of an infrastructure object changes over time. The range of condition in which an object can change before it is referred to as something else is referred to as a ``state''. An adjective or a noun in front of the word ``state'' refers to the factor being used to classify the object. For example, a ``condition state'' is defined based on the physical characteristics of an object, and a ``performance state'' is defined based on the ability of the object to perform, or provide a level of service.

As states are man-made descriptions of the object there are almost an infinite number of possible ways that states can be defined. How states are defined will affect the illustration of the movement of the object over time through the states. This can be considered in the simplification of the mathematical modeling of the behavior of infrastructure. 
%
\subsection{Deterministic vs. probabilistic models}
Models of the behavior of infrastructure over time can be done using models that are classified as deterministic or probabilistic. Both have advantages and disadvantages.

The use of deterministic models gives the impression that the values of the performance indicators is known with certainty at every time \textit{t}. They give very precise, but perhaps not very accurate predictions.

The use of probabilistic models gives the probability of having each value of the performance indicators at every time \textit{t}. They give less precise predictions than deterministic models but may be more accurate. 
\subsection{Classification of models}
The models that are typically used to make prediction of the future states of infrastructure objects can be classified in different ways. Some examples are given in Table \ref{tbl:31}. The mechanistic-empirical and regression models will be discussed in more detail in the next sections.
\begin{table}
\caption{Types of models for manifest processes}
\begin{tabular}{|l|p{300pt}|}
\hline
Type of model & Definition \\ 
\hline
Analytical & has a closed form solution, i.e. the solution to the equations used to describe changes in a system can be expressed as a mathematical analytic function. \\ 
\hline
Numerical & use some sort of numerical time-stepping procedure to obtain the models behaviour over time \\ 
\hline
Mechanistic & is based on an understanding of the behavior of the elements in a the system, often developed using the laws of physics \\ 
\hline
Empirical & is based on direct observation, measurement and extensive data records, often without understanding the physical processes at work \\ 
\hline
Extrapolation & can be used to estimate a value of a variable outside a known range from values within a known range \\ 
\hline
Interpolation & can be used to estimate a value of a variable within a known range from other values within this range \\ 
\hline
Regression & is developed using regression analysis, which is a statistical approach to predicting the change in the value of a dependent variable based on the changes in the values of an independent variables  \\ 
\hline
Markovian & is stochastic and where the conditional probability distribution of future states of the process (conditional on both past and present values) depends only upon the present state, not on the sequence of events that preceded it \\ 
\hline
\end{tabular}
\label{tbl:31}
\end{table}
\section{Mechanistic-empirical models}
\subsection{Theory}
A mechanistic model is one that is based on an understanding of the behavior of the elements in the system, often developed using the laws of physics. An empirical model is one that is based on direct observation, measurement and extensive data records, often without understanding the physical processes at work. A mechanistic - empirical model is a combination of both.

These types of models are often used to predict the occurrence of very specific conditions, e.g. the amount of chlorides at a specific depth in a concrete beam. Coupled with the probabilistic distributions to represent the values of the variables and parameters in the models, they can be used to estimate the probable specific conditions.

They can also be used to predict a LOS, e.g. the amount of intervention costs in a specified unit of time. Or more specifically, the probability of occurrence of each LOS within each investigated time interval can be estimated using a Monte Carlo simulation or a random sampling algorithm, given the probabilistic distributions of each parameters, and the mechanistic-empirical model.
%
\subsection{Example}
You are an infrastructure manager that would like to predict the evolution of a reinforced concrete bridge slab whose main deterioration process is chloride-induced corrosion of the steel reinforcement over time (More information related to such processes can be found in \cite{Elsener2013} and \cite{Bertolini2013} so that you can plan an intervention before the bridge slab provides an inadequate LOS. You know that the chloride-induced corrosion is composed of two deterioration phases: the first phase is called initiation phase, which covers the entire time from construction to the time the first visual crack appears. In this phase the chloride starts to penetrate into concrete cover, reach the reinforcement with a sufficient concentration to start corrosion and continue until there is enough corrosion to cause cracks to be visible. The second phase of deterioration then starts and includes the continued widening of the cracks in the concrete due to the continued corrosion of the reinforcement. The entire deterioration process can be graphically illustrated in Figure \ref{fig:31}.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig31.eps}
\caption{Deterioration process of reinforced concrete due to chloride-induced corrosion}\label{fig:31}
% \end{center}
\end{figure}
According to \cite{DuraCrete2000}, chloride penetration into reinforced concrete is a complicated process, which involves inter alia ion diffusion and convection \citep{Bertolini2013}. This complex transport mechanism can be represented by Fick's second law of diffusion.
\begin{eqnarray}
      && \frac{{\partial {C_{cl}}}}{{\partial t}} = {D_{cl}}\frac{{{\partial ^2}{C_{cl}}}}{{\partial x_{cl}^2}} \label{eq31}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$C_{cl}$:] is the chloride ion concentration at the depth,
\item[$x_{cl}$:] from the surface of concrete after it exposes to chlorides till time \textit{t} (here small notation cl denotes the abbreviation for chloride).  
\item[$D_{cl}$:] is the chloride diffusion coefficient. 
\end{description}
%\end{flushright}
\end{adjustwidth}
The solution for partial differential equation of Eq. \eqref{eq31} gives the following explicit form to calculate the chloride concentration as a function of depth ${x_{cl}}$and time \textit{t}.
\begin{eqnarray}
      && {C_{cl}}({x_{cl}},t) = {C_s}\left( {1 - erf\left( {\frac{{{x_{cl}}}}{{2\sqrt {{D_{cl}}t} }}} \right)} \right) \label{eq32}
\end{eqnarray}
where  $erf\left(  \right)$denotes the error function\footnote{Definition of error function at http://en.wikipedia.org/wiki/Error\_function}.

The duration for the diffusion process of chloride ion to initiate corrosion, that is time \textit{t} in Eq. \eqref{eq32}, can be estimated by setting the value of ${C_{cl}}$ to be equal to the chloride concentration. In other words, by setting the value of ${C_{cl}}$for each discrete condition state \textit{i}, the time to arrive at that CS can be obtained by solving Eq. \eqref{eq32} with respect to time \textit{t} and a certain depth of concrete cover from the re-bar.

For a bridge slab, value of variables ${C_{cl}}$,${C_s}$, and ${D_{cl}}$in the equation are can be considered as random, each one is associated with its own statistical distribution. Thus, estimation of time to arrive at certain value of chloride concentration is probabilistic in nature.

After the value of chloride concentration reaches a certain limit, corrosion starts on the reinforcing bars and after it reaches another higher limit the cracking process starts. The crack initiation phase is referred to here as the propagation phase and the width of the cracks over time can be determined by using following equations:
\begin{eqnarray}
      && w(t) = {w_0} + \beta  \cdot \left( {P(t) - {P_0}} \right) \label{eq33}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$w(t)$:] width (mm) of crack over time,
\item[$\beta $:] the parameter that controls the propagation
\item[${w_0}$:] width of crack when it is visible (\textasciitilde{}0.05 mm)
\item[${P_0}$:] the amount of loss of re-bar diameter (mm) when crack width is visible 
\item [$P(t)$:] the amount of loss of re-bar diameter (mm) at \textit{t}
\end{description}
%\end{flushright}
\end{adjustwidth}

The reinforcement loss function can be represented as:
%
\begin{eqnarray}
      && P(t) = \int\limits_0^t {{V_{corr}} \cdot \alpha  \cdot wet \cdot tdt}  \label{eq34}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[${V_{corr}}$:] corrosion rate coefficient (mm/year)
\item[$wet$:] wet period in a year (equal to the ratio between total numbers of rainy day and 365 days)
\item[$\alpha $:] pitting factor that takes non-uniform corrosion of the re-bars into consideration
\end{description}
%\end{flushright}
\end{adjustwidth}

Based on field tests and experiments, statistical values of models coefficients and parameters are given in Table \ref{tbl:32}
% 
\begin{table}
\caption{Values of models parameters and coefficients}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Parameters/ Coefficients} & \multicolumn{1}{c|}{Units} & Mean ($\mu$) & Standard deviation ($\sigma$) & Distribution & Reference equations \\ 
\hline
\multicolumn{1}{|c|}{$C_s$} & \multicolumn{1}{c|}{$wt.-\%Cl-/binder$} & \multicolumn{1}{c|}{2.565} & \multicolumn{1}{c|}{0.405} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq32}} \\ 
\hline
\multicolumn{1}{|c|}{$x_{cl}$} & \multicolumn{1}{c|}{mm} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{Eq. \eqref{eq32}} \\ 
\hline
\multicolumn{1}{|c|}{$D_{cl}$} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{100} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq32}} \\ 
\hline
\multicolumn{1}{|c|}{$w_0$} & \multicolumn{1}{c|}{mm} & \multicolumn{1}{c|}{0.05} & \multicolumn{1}{c|}{0.005} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq33}} \\ 
\hline
\multicolumn{1}{|c|}{$P_0$} & \multicolumn{1}{c|}{mm} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq33}} \\ 
\hline
\multicolumn{1}{|c|}{$\beta$} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{0.001} & \multicolumn{1}{c|}{0.001} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq33}} \\ 
\hline
\multicolumn{1}{|c|}{$V_{corr}$} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{0.012} & \multicolumn{1}{c|}{0.006} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq34}} \\ 
\hline
\multicolumn{1}{|c|}{$wet$} & \multicolumn{1}{c|}{year} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0.25} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq34}} \\ 
\hline
\multicolumn{1}{|c|}{$\alpha$} & \multicolumn{1}{c|}{NA} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{4.04} & \multicolumn{1}{c|}{Normal} & \multicolumn{1}{c|}{Eq. \eqref{eq34}} \\ 
\hline
\end{tabular}
\label{tbl:32}
\end{table}

\subsubsection{Question}
If you know that the chloride concentration ${C_{crit}}$ which starts the propagation phase is ${C_{cl}} = {C_{crit}} = 0.48$, and you consider that once a crack width is 0.5 mm, that an intervention is required to ensure that an adequate LOS is provided, when do you need to plan an intervention to ensure that an adequate LOS is provided?

Hint: Construct a probabilistic deterioration model with discrete condition states for the entire chloride induced corrosion process of the bridge slab.
\subsubsection{Answer}
\paragraph{Condition states}
Since the deterioration process of chloride induced corrosion is a continuous process with two distinct phases, it is useful to define ranges of condition states for each phase in order to construct the probabilistic deterioration model with discrete condition states. This is illustrated in Figure \ref{fig:32}, when 5 condition states are used. Two are attributed to the initiation phase and two to the propagation phase.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig32.eps}
\caption{Deterioration process with discrete condition states}\label{fig:32}
% \end{center}
\end{figure}
In Figure \ref{fig:32}, five discrete condition states are used to illustrate the process. Their definitions are explained in Table \ref{tbl:33}.

\begin{table}
%	\centering
	\caption{Definition of condition states} \label{tbl:33}
		\begin{tabular}{|l|l|p{3.5cm}|p{5.5cm}|l|}
\hline
\multicolumn{1}{|c|}{Phase} & \multicolumn{1}{c|}{CS} & \multicolumn{1}{c|}{Description} & \multicolumn{1}{c|}{Indicator} & \multicolumn{1}{c|}{Criteria} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{1} & New/partial new & Amount of chlorides in & \multicolumn{1}{c|}{$0 < C_{cl}\le 0.24$} \\ 
\cline{2-3}\cline{5-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{2} & Concrete contaminated & the concrete at reinforcing bar level $C_{cl}$ & \multicolumn{1}{c|}{$0.24 < C_{cl}\le 0.48$} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{3} & Corrosion has initiated, no visible cracking has occurred & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$C_{cl}> 0.48, w\le 0.25$} \\ 
\cline{2-3}\cline{5-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{4} & Visible cracking has occurred & Width of crack ($w$) & \multicolumn{1}{c|}{$0.25 < w \le 0.5$} \\ 
\cline{2-3}\cline{5-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{5} & Visible cracking has occurred and cover has spalled off & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$ w > 0.5$} \\ 
\hline
\end{tabular}
\end{table}

\paragraph{Deterioration: Chloride concentration}

Using Eq. \eqref{eq32}, the concentration of chlorides at any time \textit{t} can be deterministically estimated given the mean values of the parameters of the model. These are, however, probabilistic in nature, thus, an explicit and straightforward arithmetic calculation is not always desirable. At any time in the future, for example, the value of ${C_{cl}}$ can be represented by using the mean and standard deviation. This concept is illustrated in Figure \ref{fig:33} for a normal distribution, where the mean value $\mu$ and standard deviation $\sigma$ are 0.3 and 0.08, respectively

\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig33.eps}
\caption{Probability of condition state CS 1-initiation phase.}\label{fig:33}
% \end{center}
\end{figure}

In order to determine when an intervention is required to ensure that an adequate LOS is provided, the probability of being in each condition state over time needs to be calculated.

Let assume that it is time \textit{t} now, and the probability of the bridge slab being in CS 1 will be equivalent to the probability that its continuous value falls between 0 and 0.24. The equation used to estimate the red area is.

In order to determine when an intervention is required to ensure that an adequate LOS is provided, the probability of being in each condition state over time needs to be calculated.

Let assume that it is time t now, and the probability of the bridge slab being in $CS1$ will be equivalent to the probability that its continuous value falls between 0 and 0.24. The equation used to estimate the red area is.
\begin{eqnarray}
      && P_1 =P_1^{cl} = Prob[CS=1]=Prob[0 \leq C_{cl} \leq 0.24] = \int_0^{0.24}f(x) \cdot dx = 0.227 \label{stateprobcs1}
\end{eqnarray}
where $f(x)$ is density function of the normal distribution. 
\begin{eqnarray}
      && f(x)= \frac{1}{\sigma \sqrt{2\pi}}\cdot e^{-\frac{(x-\mu)^2}{2\sigma^2}}\label{densitynormaldistribu}
\end{eqnarray}
Similarly, the probability that the bridge element is in $CS2$ is given by:
\begin{eqnarray}
      && P_2 = P_2^{cl} = Prob[CS=2]=Prob[0.24 \leq C_{cl} \leq 0.48] \nonumber \\
      && = \int_{0.24}^{0.48}f(x) \cdot dx = 0.761 \label{stateprobcs2}
\end{eqnarray}
The probability that the bridge slab is in $CS3$ is in a condition state greater than CS2 is then given by:
\begin{eqnarray}
      && P_{>2} = P_{>2}^{cl} = 1-Prob[CS=2]-Prob[CS=2]=1-0.227-0.761=0.012\label{stateprobcs345}
\end{eqnarray}
%
\begin{figure}[h!]
% \centering
\includegraphics[width=300pt]{fig34.eps} 
\caption{Probability of condition state CS-2-initiation phase.} 
\label{fig:34}
\end{figure}
%
\paragraph{Deterioration: Corrosion propagation}

If it is assumed that the crack width can be modeled using a normal distribution with $\left[ {\mu  = 0.22,\sigma  = 0.03} \right]$ at time \textit{t} (the same time that is used to compute the state probability of the initiation phase), then the probability of being in condition state 3 is given by:

\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig35.eps}
\caption{Probability of condition state CS 3-propagation phase.}
% \end{center}
\end{figure}

\begin{eqnarray}
      && P_3 = P_3^{cr} = Prob[CS=3]=Prob[w\leq0.25]\cdot Prob[C_{cl}>0.48] \nonumber \\
      && = 0.841 \cdot 0.012 = 0.010092\label{eq35}
\end{eqnarray}

Eq. \eqref{eq35} shows a joint probability of the event that chloride concentration level has reached more than its critical level 0.48 and corrosion has progressed to a point where at least one crack of 0.05 mm has opened up. The width of crack is, however, less than 0.25 mm.

Similarly, following state transition probabilities are obtained

\begin{eqnarray}
      && P_4 = P_4^{cr} = Prob[CS=4]=Prob[0.25 <w\leq0.5]\cdot Prob[C_{cl}>0.48] \nonumber \\
      && = 0.159 \cdot 0.012 = 0.001908\label{stateprobcs4} \\
      && P_5 = P_5^{cr} = Prob[CS=5]=Prob[w\geq 0.5]\cdot Prob[C_{cl}>0.48] = 0 \cdot 0.012 = 0 \label{stateprobcs5}
\end{eqnarray}

The above steps of calculation are used for a specific case when the condition state is defined in Table \ref{tbl:33}. For a general case that mathematically describes the relationship and for sake of programing, following generic formulation is used.

%
\begin{eqnarray}
&& {P_k} = \left\{ {\begin{array}{*{10}{c}}
{P_i^{cl} = \int\limits_a^b {f(x)dx} \hspace{0.2cm}i = (1,...,I){\rm{  }}}\\
{P_I^{cl} \cdot P_j^{cr} \hspace{1cm}   where \hspace{1cm} P_j^{cr} = \int\limits_c^d {g(x)dx} \hspace{1cm}j = (1,...,J)}
\end{array}{\rm{    }}} \right. \label{eq36}
\end{eqnarray}

Where:%k = (1,...,\left[ {I + J - 1} \right])
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$P_i^{cl}$:] is state probability of CS $i$ of the initiation phase with total number CS of $I$
\item[$P_j^{cl}$:] is state probability of CS $j$ of the propagation phase with total number CS of $J$
\item[$f(x)$ and $g(x)$:] are density function of the normal distribution for the initiation phase and propagation phase, respectively.
\end{description}
%\end{flushright}
\end{adjustwidth}
\paragraph{State distribution as proportional data}
Once the mechanistic-empirical equation, and the probabilistic distributions of each parameter to be used are known than the probable value of both chloride concentration and crack width can be generated by running many simulations, e.g. $20’000$. Once enough points are generated at the point of time to be investigated, either the probabilistic distribution that best fits these points can be determined and from this distribution the probability of the bridge slab being in a specific condition state can be estimated, or the probability of the bridge being in each condition state can be estimated directly from the number of data points, e.g. $10’000/20’000$ are in the specified range.

In previous steps of calculation, the example is shown only for a specific time $t$ given the mean $\mu$ and the standard deviation $\sigma$ for each process. However, these values have to be estimated at each time step given the probabilistic distribution of models parameters. Results of estimation are the state probabilities over a pre-defined number of years. This pool of state probabilities is proportional data and state probabilities of the bridge in previous year is not related to the following year.

\paragraph{Deterioration prediction}

Once the mechanistic-empirical equation, and the probabilistic distributions of each parameter to be used are known than the probable value of both chloride concentration and crack width can be generated by running many simulations, e.g. 20'000. Once enough points are generated at the point of time to be investigated, either the probabilistic distribution that best fits these points can be determined and from this distribution the probability of the bridge slab being in a specific condition state can be estimated, or the probability of the bridge being in each condition state can be estimated directly from the number of data points, e.g. 10'000 / 20'000 are in the specified range.

In previous steps of calculation, the example is shown only for a specific time \textit{t} given the mean $\mu$ and the standard deviation $\sigma$ for each process. However, these values have to be estimated at each time step given the probabilistic distribution of models parameters (Table \ref{tbl:32}).

Results of the estimation are the state probability over a pre-defined number of years. In this example, calculation is done for 60 years. At each year, the probabilistic value of chloride concentration ${C_{cl}}$ and the width of crack $X$ are re-computed using a pool of 20'000 samples that were randomly generated using Monte Carlo simulation. Then, the state probability that the bridge slab arrives at each discrete CS was computed for every year. These are shown in graphically in Figure \ref{fig:36}.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=324pt]{fig36.eps}
\caption{Condition state distribution over 60 years}\label{fig:36}
% \end{center}
\end{figure}

\paragraph{Results}

As can be seen from Figure \ref{fig:36}, in the first 7 years after the construction of the bridge slab, the chloride concentration is not expected to reach a value of 0.25. In other words, the bridge slab is expected to remain in CS 1. Then, the probability of the bridge slab being in CS 1 decreases, in the mean-time, of course, the probability that the bridge slab is in CS 2 increases. After 10 years, there is almost a 50\% chance that the bridge slab will be in CS 2.

Between the 7$^{th}$ and 10$^{th}$ years, there is a non-zero probability that the critical chloride concentration has been reached (0.48), i.e. the propagation phase has started. This explains the appearance of CS 3, CS 4 and CS 5 in the chart.

After 20 years, there is a high probability that the critical chloride concentration has been reached and the propagation phase has started. There is also approximately a 40\% chance that the surface of the slab will have at least one crack with a width between 0.00 mm and 0.25 mm. More or less the same chance that the crack width will be between 0.25 mm and 0.5 mm. There is, however, also a 20\% chance that the crack width will be greater than 0.5 mm, and that an intervention will have to be executed to ensure that an adequate LOS is provided.

So, since I know that the chloride concentration ${C_{crit}}$ that starts the propagation phase is ${C_{cl}} = {C_{crit}} = 0.48$, and if I consider that once a crack width is 0.5 mm that an intervention is required to ensure that an adequate LOS is provided, I will plan an intervention in year 15, when the probability of being in CS5 is still, according to my calculations is under 0.01 (Table \ref{tbl:34}). 
\begin{table}
%	\centering
	\caption{Evolution of state probability over 30 years} \label{tbl:34}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|m{1.5cm}|}{\centering year} & \multicolumn{5}{c|}{Condition states} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{m{1.5cm}|}{\centering 1} & \multicolumn{1}{m{1.5cm}|}{\centering 2} & \multicolumn{1}{m{1.5cm}|}{\centering 3} & \multicolumn{1}{m{1.5cm}|}{\centering 4} & \multicolumn{1}{m{1.5cm}|}{\centering 5} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{0.9174} & \multicolumn{1}{c|}{0.0826} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{0.5520} & \multicolumn{1}{c|}{0.3907} & \multicolumn{1}{c|}{0.0574} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{0.2977} & \multicolumn{1}{c|}{0.3571} & \multicolumn{1}{c|}{0.3452} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{0.1718} & \multicolumn{1}{c|}{0.2123} & \multicolumn{1}{c|}{0.6149} & \multicolumn{1}{c|}{0.001} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{11} & \multicolumn{1}{c|}{0.1065} & \multicolumn{1}{c|}{0.1181} & \multicolumn{1}{c|}{0.7640} & \multicolumn{1}{c|}{0.0114} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{12} & \multicolumn{1}{c|}{0.0693} & \multicolumn{1}{c|}{0.0672} & \multicolumn{1}{c|}{0.8153} & \multicolumn{1}{c|}{0.0482} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{13} & \multicolumn{1}{c|}{0.0466} & \multicolumn{1}{c|}{0.0396} & \multicolumn{1}{c|}{0.7965} & \multicolumn{1}{c|}{0.1173} & \multicolumn{1}{c|}{0.0001} \\ 
\hline
\multicolumn{1}{|c|}{14} & \multicolumn{1}{c|}{0.0319} & \multicolumn{1}{c|}{0.0242} & \multicolumn{1}{c|}{0.7360} & \multicolumn{1}{c|}{0.2069} & \multicolumn{1}{c|}{0.0011} \\ 
\hline
\multicolumn{1}{|c|}{15} & \multicolumn{1}{c|}{0.0221} & \multicolumn{1}{c|}{0.0151} & \multicolumn{1}{c|}{0.6584} & \multicolumn{1}{c|}{0.2977} & \multicolumn{1}{c|}{0.0068} \\ 
\hline
\multicolumn{1}{|c|}{16} & \multicolumn{1}{c|}{0.0154} & \multicolumn{1}{c|}{0.0097} & \multicolumn{1}{c|}{0.5795} & \multicolumn{1}{c|}{0.3720} & \multicolumn{1}{c|}{0.0234} \\ 
\hline
\multicolumn{1}{|c|}{17} & \multicolumn{1}{c|}{0.0107} & \multicolumn{1}{c|}{0.0063} & \multicolumn{1}{c|}{0.5073} & \multicolumn{1}{c|}{0.4195} & \multicolumn{1}{c|}{0.0561} \\ 
\hline
\multicolumn{1}{|c|}{18} & \multicolumn{1}{c|}{0.0075} & \multicolumn{1}{c|}{0.0041} & \multicolumn{1}{c|}{0.4447} & \multicolumn{1}{c|}{0.4387} & \multicolumn{1}{c|}{0.1050} \\ 
\hline
\multicolumn{1}{|c|}{19} & \multicolumn{1}{c|}{0.0052} & \multicolumn{1}{c|}{0.0027} & \multicolumn{1}{c|}{0.3918} & \multicolumn{1}{c|}{0.4344} & \multicolumn{1}{c|}{0.1659} \\ 
\hline
\multicolumn{1}{|c|}{20} & \multicolumn{1}{c|}{0.0036} & \multicolumn{1}{c|}{0.0018} & \multicolumn{1}{c|}{0.3477} & \multicolumn{1}{c|}{0.4139} & \multicolumn{1}{c|}{0.2330} \\ 
\hline
\multicolumn{1}{|c|}{21} & \multicolumn{1}{c|}{0.0025} & \multicolumn{1}{c|}{0.0012} & \multicolumn{1}{c|}{0.3110} & \multicolumn{1}{c|}{0.3842} & \multicolumn{1}{c|}{0.3011} \\ 
\hline
\multicolumn{1}{|c|}{22} & \multicolumn{1}{c|}{0.0017} & \multicolumn{1}{c|}{0.0008} & \multicolumn{1}{c|}{0.2805} & \multicolumn{1}{c|}{0.3506} & \multicolumn{1}{c|}{0.3664} \\ 
\hline
\multicolumn{1}{|c|}{23} & \multicolumn{1}{c|}{0.0012} & \multicolumn{1}{c|}{0.0005} & \multicolumn{1}{c|}{0.2551} & \multicolumn{1}{c|}{0.3164} & \multicolumn{1}{c|}{0.4268} \\ 
\hline
\multicolumn{1}{|c|}{24} & \multicolumn{1}{c|}{0.0008} & \multicolumn{1}{c|}{0.0004} & \multicolumn{1}{c|}{0.2337} & \multicolumn{1}{c|}{0.2839} & \multicolumn{1}{c|}{0.4812} \\ 
\hline
\multicolumn{1}{|c|}{25} & \multicolumn{1}{c|}{0.0006} & \multicolumn{1}{c|}{0.0002} & \multicolumn{1}{c|}{0.2157} & \multicolumn{1}{c|}{0.2539} & \multicolumn{1}{c|}{0.5296} \\ 
\hline
\multicolumn{1}{|c|}{26} & \multicolumn{1}{c|}{0.0004} & \multicolumn{1}{c|}{0.0002} & \multicolumn{1}{c|}{0.2005} & \multicolumn{1}{c|}{0.2269} & \multicolumn{1}{c|}{0.5721} \\ 
\hline
\multicolumn{1}{|c|}{27} & \multicolumn{1}{c|}{0.0003} & \multicolumn{1}{c|}{0.0001} & \multicolumn{1}{c|}{0.1874} & \multicolumn{1}{c|}{0.2030} & \multicolumn{1}{c|}{0.6093} \\ 
\hline
\multicolumn{1}{|c|}{28} & \multicolumn{1}{c|}{0.0002} & \multicolumn{1}{c|}{0.0001} & \multicolumn{1}{c|}{0.1762} & \multicolumn{1}{c|}{0.1819} & \multicolumn{1}{c|}{0.6417} \\ 
\hline
\multicolumn{1}{|c|}{29} & \multicolumn{1}{c|}{0.0001} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.1664} & \multicolumn{1}{c|}{0.1634} & \multicolumn{1}{c|}{0.6700} \\ 
\hline
\multicolumn{1}{|c|}{30} & \multicolumn{1}{c|}{0.0001} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.1580} & \multicolumn{1}{c|}{0.1472} & \multicolumn{1}{c|}{0.6947} \\ 
\hline
\end{tabular}
\end{table}
As I know, however, that CS5 does not necessarily mean that an inadequate LOS is provided, I can also treat it as a warning that an inadequate LOS is coming if I don't do anything, but that I have a bit of time before this will happen. In this case, I may plan the intervention to be in year 20, when the probability of being in CS5 is 0.233. If I decide to wait until there is 50\% of being in CS5 than I will plan the intervention for either year 24 or 25. Whether or not this is a good decision depends on the probability of an inadequate LOS occurring once the slab reaches CS5 and the consequences if the adequate LOS happens, i.e. the risk associated with being in CS5.

The steps of numerical calculation can be tracked by reading the R script for this example provided at the Appendix \ref{mepi1}.
%
\section{Regression models} \label{regressionmodel}
\subsection{Theory}
Regression models represent empirical relationships between multiple parameters and are developed using regression analysis. They are used to predict the change in the value of a dependent variable based on the changes in the values of independent variables. Each variable is described in terms of its mean and variance. The simplest form is a linear regression model as shown in equation \ref{eq37}. Other forms are possible.

\begin{eqnarray}
 && {Y_i} = \alpha  + \beta  \cdot {X_i} + {\varepsilon _i} \label{eq37}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[${Y_i}$:] is the dependent variable or output (e.g. state probability of \textit{CS i}) that is observable
\item[${X_i}$:] is the characteristic or explanatory variable that is also observable (e.g. time, traffic volume, ambient temperature).
\item[$\alpha ,\beta $:] are parameters that need to be estimated.
\item[$\varepsilon $:] is the prediction error that can be assumed to follow a certain distribution (e.g. normal or log-normal distributions)
\end{description}
%\end{flushright}
\end{adjustwidth}

The mean or expected value of $Y_i$ for each value of input $X_i$ is
\begin{eqnarray}
 && E{Y_i} = \hat{\alpha}  + \hat{\beta}  \cdot {X_i} + {\varepsilon _i} \label{eq38}
\end{eqnarray}
The overhead mark ${\rm{\hat{\hspace{1mm}}}}$ indicates the expected values.

The purpose of performing a regression analysis is to predict the expected values of the regression parameters $\hat \alpha ,\hat \beta $ that minimize following objective function.
\begin{eqnarray}
 && s = {\sum\limits_{i = 1}^N {\left[ {{Y_i} - {{\hat Y}_i}} \right]} ^2} = {\sum\limits_{i = 1}^N {\left[ {{Y_i} - \hat \alpha  - \hat \beta  \cdot {X_i}} \right]} ^2} \label{eq39}
\end{eqnarray}

In words, the objective of Eq. \eqref{eq39} is to minimize the sum of the prediction errors. Here \textit{N} is the total of data points.

In order to obtain the values of the regression parameters, one way of analytically solving Eq. \eqref{eq39} with respect to two variables $\hat \alpha ,\hat \beta $ (hereafter refer as a vector set $\hat \theta $). This can be done by taking the first derivative of Eq. \eqref{eq39} and set it to be equal to 0, and then solve for $\hat \theta $.
\begin{eqnarray}
 && \frac{{\partial s}}{{\partial \hat \theta }} = 0 \label{eq310}
\end{eqnarray}
For the value of $\hat \alpha $
\begin{manyeqns}
 && \frac{{\partial s}}{{\partial \hat \alpha }} = 0 \label{eq311} \\
 && \Leftrightarrow \frac{{{{\bar Y}^2} - 2 \cdot \bar Y(\hat \alpha  + \hat \beta  \cdot \bar X) + {{(\hat \alpha  + \hat \beta  \cdot \bar X)}^2}}}{{\partial \hat \alpha }} = 0\\
 && \Leftrightarrow  - 2 \cdot \bar Y + 2 \cdot \hat \alpha  + 2 \cdot \hat \beta  \cdot \bar X = 0\\
 && \Leftrightarrow \hat \alpha  = \bar Y - \hat \beta  \cdot \bar X
\end{manyeqns}
%
Here, the bar ${\rm{\bar{\hspace{1mm}}  }}$ indicates the value of $\frac{{\sum\limits_{i = 1}^N {{X_i}} }}{N}$ and $\frac{{\sum\limits_{i = 1}^N {{Y_i}} }}{N}$, which are the mean values.

Similarly, for the value of $\hat \beta $
\begin{manyeqns}
  && \frac{{\partial s}}{{\partial \hat \beta }} = 0\\
 && \Leftrightarrow \frac{{{{\bar Y}^2} - 2 \cdot \bar Y(\hat \alpha  + \hat \beta  \cdot \bar X) + {{(\hat \alpha  + \hat \beta  \cdot \bar X)}^2}}}{{\partial \hat \beta }} = 0\\
 && \Leftrightarrow \beta  \cdot \sum\limits_{i = 1}^N {X_i^2}  - \sum\limits_{i = 1}^N {{X_i} \cdot {Y_i}}  + \alpha \sum\limits_{i = 1}^N {{X_i}}  = 0\\
 && \Leftrightarrow \frac{{\sum\limits_{i = 1}^N {{X_i} \cdot {Y_i}}  - n \cdot \bar X \cdot \bar Y}}{{\sum\limits_{i = 1}^N {X_i^2 - n \cdot \bar X} }} = \frac{{\sum\limits_{i = 1}^N {({X_i} - \bar X) \cdot ({Y_i} - \bar Y)} }}{{\sum\limits_{i = 1}^N {{{({X_i} - \bar X)}^2}} }}
\end{manyeqns}
Following figure graphically represents the regression equation and the calculation.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig37.eps}
\caption{Deriving linear regression coefficients}
% \end{center}
\end{figure}

The goodness of fit of the regression line can be measured by the coefficient ${R^2}$, which measures the proportion of total variation about the mean $\bar Y$, which is explained by regression:
\begin{eqnarray}
 && {R^2} = \frac{{\sum\limits_{i = 1}^N {{{({{\hat Y}_i} - \bar Y)}^2}} }}{{\sum\limits_{i = 1}^N {{{({Y_i} - \bar Y)}^2}} }} = \frac{{{\beta ^2} \cdot \sum\limits_{i = 1}^N {{{({X_i} - \bar X)}^2}} }}{{\sum\limits_{i = 1}^N {{{({{\hat Y}_i} - \bar Y)}^2}} }} \label{eq321}
\end{eqnarray}
The prediction errors can be followed an independent type of distribution. In many practical situation, normal distribution is often used. In case of normal distribution $N(\mu ,\sigma )$, its mean and standard deviation are defined as:
\begin{eqnarray}
 && {\mu _\varepsilon } = {Y_i} - {\hat Y_i} \label{eq322}\\
 && {\sigma _\varepsilon } = \sqrt {\frac{{\sum\limits_{i = 1}^N {{{({Y_i} - {Y_i})}^2}} }}{{N - k - 1}}}
\end{eqnarray}
Where \textit{k} is number of independent variable, which is equal to 1 in this example (as there is only one input X).

The above example is for the case of simple linear regression with only one independent variable. In many practical situations, linear regression models can be built for multiple variables, which often takes following form.
\begin{eqnarray}
 && Y = \sum\limits_{i = 0}^N {\beta _{}^k}  \cdot X_i^k \label{eq3223}
\end{eqnarray}
Where $k = 1,...,K$ is a vector set of numbers of independent variables. 
\subsection{Example}
Assuming that you have many reinforced concrete slabs, as the one discussed in section 2.2 and have measured chloride concentrations and crack widths over the years, you may have the plots shown in Figure \ref{fig:38} and Figure \ref{fig:39}\footnote{Data used for this example is selected from the pool of randomly generated total data used in the previous example, taking into consideration of the mean and standard deviation at each time t.}. (This data will be distributed in a .csv file in addition to this text).
\begin{figure}[h]
%\begin{center}
\includegraphics[width=300pt]{fig38.eps}
\caption{Measured chloride concentrations in similar deck slabs over time}\label{fig:38}
%\end{center}
\end{figure}
\begin{figure}[h]
%\begin{center}
\includegraphics[width=300pt]{fig39.eps}
\caption{Measured crack widths in similar deck slabs over time}\label{fig:39}
%\end{center}
\end{figure}
\subsubsection{Question}
If you know that the chloride concentration ${C_{crit}}$ which starts the propagation phase is ${C_{cl}} = {C_{crit}} = 0.48$, as in the previous example, and you consider that the once a crack width of 0.5 mm is reached, that an intervention is required to ensure that an adequate LOS is provided, when do you need to plan an intervention to ensure that an adequate LOS is provided? (Use linear regression analysis with the values of chloride concentration and crack width provided, estimate the functional form of deterioration).
\subsubsection{Answer}
As deterioration of the bridge slab is seen as happening in two different phases, separate models are developed using regression analysis for both phases; one for the initiation phase with chloride concentration and other for the propagation phase with the development of crack.
\paragraph{Condition states}
The condition states are defined exactly as in Table \ref{tbl:33}.
\paragraph{Deterioration: Chloride concentration} \label{chrloride}
The form of regression equation for chloride concentration, when it is assumed to be linear is:
\begin{eqnarray}
 && Y_i^{cl} = {\alpha _{cl}} + {\beta _{cl}} \cdot X_i^{cl} + \varepsilon _i^{cl} \label{eq3224}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$Y_i^{cl}$:] is the value of chloride concentration at year \textit{i}.
\item[$X_i^{cl}$:] is the year \textit{i}.
\item[${\alpha _{cl}},{\beta _{cl}}$:] are regression parameters that need to be estimated.
\item[$\varepsilon _i^{cl}$:] is the prediction errors following a normal distribution
\end{description}
%\end{flushright}
\end{adjustwidth}
The expected values of regression parameters can be obtained using the R code in the Appendix \ref{mepi3} on data created using the Appendix \ref{mepi2}. The results are in listing \ref{mechaempichlorine}.

\lstinputlisting[basicstyle=\ttfamily\scriptsize,float=h,frame=tb,caption=Results-Chloride concentration,label=mechaempichlorine]{./Programs/IMP-mecha-empi/IMP-chloridecon.tex}

The value of the regression parameters are: ${\alpha _{cl}} =  - 3.40275$ and ${\beta _{cl}} = 0.47975$. The value of ${R^2}$ is greater than 0.8959, which is nearly 10\% far from 1. This means that the fitness of the linear model to this data set infers a small variation. This can also be seen from Figure \ref{fig:310}, which shows fitted line of chloride concentration at the surface of the reinforcement as time goes by.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=304pt]{fig310.eps}
\caption{Development of chloride concentration}\label{fig:310}
% \end{center}
\end{figure}
The equation to be used for estimating the chloride concentration over time is then

\begin{eqnarray}
 && Y_i^{cl} =  - 3.40275 + 0.47975 \cdot X_i^{cl} \label{eq3225}
\end{eqnarray}
Noted that the value of chloride concentration cannot be negative, therefore, it is necessary to consider a dummy variable to control the results of calculation
\begin{eqnarray}
 && Y_i^{cl} = ( - 3.40275 + 0.47975 \cdot X_i^{cl}) \cdot \delta _i^{cl} \label{eq3226}\\
 && where{\rm{\hspace{2mm} }}\delta _i^{cl} = \left\{ {\begin{array}{*{20}{c}}
{0{\rm{\hspace{2mm}   when \hspace{2mm} X}}_i^{cl} < 7}\\
{1{\rm{\hspace{2mm}   when \hspace{2mm}  X}}_i^{cl} \ge 7{\rm{ }}}
\end{array}} \right.
\end{eqnarray}
Where 7 is the integer value closest to the ratio 3.40275/0.4975. An integer value is used here as the years are used as discrete units of time.

This result is obtained by running the R code provided in the Appendix \ref{mepi3} on data created using the Appendix \ref{mepi2}.
%
\paragraph{Deterioration: Crack propagation}
To estimate the development of crack width using a regression model, a step similar to the one in section \ref{chrloride} is used. Following results (listing \ref{mechaempicrack}) are obtained (using the R code provided in the Appendix \ref{mepi3}).

\lstinputlisting[basicstyle=\ttfamily\scriptsize,float=h,frame=tb,caption=Results-Crack,label=mechaempicrack]{./Programs/IMP-mecha-empi/IMP-crackpropa.tex}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=305pt]{fig311.eps}
\caption{Development of crack width}
\label{fig:311}
% \end{center}
\end{figure}
The equation to be used to predict crack width is then:
\begin{eqnarray}
 && Y_i^{cr} = ( - 0.500755 + 0.053150 \cdot X_i^{cr}) \cdot \delta _i^{cr} \label{eq3227}\\
 && where{\rm{\hspace{2mm} }}\delta _i^{cr} = \left\{ {\begin{array}{*{20}{c}}
{0{\rm{ \hspace{2mm}  when \hspace{2mm} X}}_i^{cr} < 9}\\
{1{\rm{ \hspace{2mm}  when \hspace{2mm}  X}}_i^{cr} \ge 9{\rm{ }}}
\end{array}} \right.
\end{eqnarray}
Where 9 is the integer value closest to the ratio 0.500755/0.053150. An integer value is used here as the years are used as discrete units of time.
%
\paragraph{Deterioration prediction}

Using the regression equations given above, the predicted deterioration is as shown in Figure \ref{fig:312}. The curve consists of three pieces. The first piece represents the period of time from new construction until the chloride concentration at the reinforcement level is thought to be equivalent to that determined by the regression equation. The second piece represents the period of time where the chloride concentration at the reinforcement can be represented by the regression equation to the point where corrosion has started to an extent to cause at least one crack wider than 0.5 mm, and the third piece shows the propagation phase, i.e. the extension of cracking once at least one crack is wider than 0.5 mm. This is represented by the second regression equation.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=305pt]{fig312.eps}
\caption{Development of crack width}
 \label{fig:312}
% \end{center}
\end{figure}
\paragraph{Results}
As can be seen from the deterioration curve estimated by using regression model, by the time of 17 years, the chloride concentration certainly reach to a level that higher than 0.48. It means the bridge slab will enter CS 3 and the second phase of deterioration will start, which initiates cracking. Since the time to the worst condition state (CS 5) from the time propagation phase starts will be about 10 years, meaning that the total duration of deterioration from CS1 to CS 5 is about 27 years.

So, since I know that the chloride concentration ${C_{crit}}$ that starts the propagation phase is ${C_{cl}} = {C_{crit}} = 0.48$, and if I consider that once a crack width is 0.5 mm that an intervention is required to ensure that an adequate LOS is provided, I will plan an intervention before year 27. 
\section{Comparison mechanistic-empirical models vs. regression models}
Using both models, and planning an intervention when there is a 50\% chance that the bridge slab will be in CS5 are the same (27 years).

Using the mechanistic model, in the way we did we could also determine the probability of the bridge slab being in CS5 in each time interval through-out the investigated time period. This made it possible to set other criteria on when to intervene, e.g. the probability of being in CS5 is negligible, or less than 20\%. This was not possible using the regression model the way we did. If it was desired to take into consideration risk, using the regression analysis one would need to arbitrarily pick an earlier point in time to plan the intervention, e.g. in year 20 and not in year 25.

The use of the mechanistic-empirical model requires a lot of information on the specific values of the parameters to be used in the model, estimates of the uncertainty of the model, and the model. It is suitable in situations where there is not a lot of data on the performance of the deck slabs, and it is feasible to estimate the distributions to be used to represent the values of the parameters, develop the model, and estimate model uncertainty.

The use of the regression model requires a lot of information on the performance of similar deck slabs. It is suitable in situations where this information exists.
\section{Assignments}
\subsection{Problem A - Erosion of concrete cover of waste water tanks}
In a waste water treatment plant, reinforced concrete tanks are used to store the waste water before going through the cleaning process. Waste water is stored in the tanks and various chemical substances are poured into the tanks for the purposes of disinfection. The cement deteriorates due to this exposure.

The infrastructure manager is interested in predicting the condition of the tanks so that the maintenance interventions can be better planned over the next 30 years. She realizes that these can be estimated using the results of past inspections performed at 2 years intervals (Table \ref{tbl:35}). The first inspection time was executed 1 year after the begin use of the tanks. It is a general rule of thumb that the erosion should not be more than 12 mm.
%
\begin{table}
%	\centering
	\caption{Inspection data} \label{tbl:35}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Tank} & \multicolumn{5}{c|}{Depth of erosion (m)} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{inspection 1} & \multicolumn{1}{c|}{inspection 2} & \multicolumn{1}{c|}{inspection 3} & \multicolumn{1}{c|}{inspection 4} & \multicolumn{1}{c|}{inspection 5} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{0.2} & \multicolumn{1}{c|}{1.1} & \multicolumn{1}{c|}{2.4} & \multicolumn{1}{c|}{3.8} & \multicolumn{1}{c|}{4.9} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{0.9} & \multicolumn{1}{c|}{1.7} & \multicolumn{1}{c|}{3.5} & \multicolumn{1}{c|}{5.2} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{2.7} & \multicolumn{1}{c|}{3.2} & \multicolumn{1}{c|}{4.5} & \multicolumn{1}{c|}{5.5} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{3.1} & \multicolumn{1}{c|}{3.6} & \multicolumn{1}{c|}{5.5} & \multicolumn{1}{c|}{6.3} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{1.1} & \multicolumn{1}{c|}{1.8} & \multicolumn{1}{c|}{1.9} & \multicolumn{1}{c|}{3.1} & \multicolumn{1}{c|}{4.9} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{0.7} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{3.8} & \multicolumn{1}{c|}{5.1} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{0.8} & \multicolumn{1}{c|}{0.9} & \multicolumn{1}{c|}{1.8} & \multicolumn{1}{c|}{4.2} & \multicolumn{1}{c|}{5.4} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{1.9} & \multicolumn{1}{c|}{2.6} & \multicolumn{1}{c|}{3.9} & \multicolumn{1}{c|}{5} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{0.9} & \multicolumn{1}{c|}{1.7} & \multicolumn{1}{c|}{3.6} & \multicolumn{1}{c|}{4.3} & \multicolumn{1}{c|}{7.1} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{0.8} & \multicolumn{1}{c|}{0.9} & \multicolumn{1}{c|}{1.6} & \multicolumn{1}{c|}{4.1} & \multicolumn{1}{c|}{6.7} \\ 
\hline
\multicolumn{1}{|c|}{11} & \multicolumn{1}{c|}{0.5} & \multicolumn{1}{c|}{1.8} & \multicolumn{1}{c|}{2.3} & \multicolumn{1}{c|}{3.7} & \multicolumn{1}{c|}{5.8} \\ 
\hline
\multicolumn{1}{|c|}{12} & \multicolumn{1}{c|}{0.6} & \multicolumn{1}{c|}{1.2} & \multicolumn{1}{c|}{2.2} & \multicolumn{1}{c|}{3.6} & \multicolumn{1}{c|}{6} \\ 
\hline
\end{tabular}
\end{table}

She also realizes that this can be done using by constructing and testing tank walls in the laboratory. After having done this her engineers have come up with the following mechanistic deterioration model.
\begin{eqnarray}
 && f(t + 1) = f(t) \cdot {e^\beta }\label{eq3228}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$f(t+1)$:] is the depth of corrosion.
\item[$\beta$:] is the deterioration coefficient taking its value of 0.1.
\end{description}
%\end{flushright}
\end{adjustwidth}
It is assumed that $f(0) = 0$ and $f(1) = 1$\footnote{This assumption is used to prevent the value of function $f$ to be equal to 0 at year t=0}.

\subsection{Question A1}
What is the condition of each object in 30 years if the mechanistic-empirical model is used?
\subsection{Question A2}
What is the condition of each object in 30 years if predictions are made using a regression analysis?
\subsection{Answer A1}
Using Eq. \eqref{eq3228}, the evolution of erosion over 30 years is shown in Table \ref{tbl:36} and Figure \ref{fig:313}.

\begin{table}
%	\centering
	\caption{Evolution of erosion (mechanistic-empirical model)} \label{tbl:36}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Time} & \multicolumn{1}{c|}{Erosion} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Erosion} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Erosion} \\ 
\multicolumn{1}{|c|}{(years)} & \multicolumn{1}{c|}{(mm)} & \multicolumn{1}{c|}{(years)} & \multicolumn{1}{c|}{(mm)} & \multicolumn{1}{c|}{(years)} & \multicolumn{1}{c|}{(mm)} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{2.7183} & \multicolumn{1}{c|}{21} & \multicolumn{1}{c|}{7.3891} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{1.1052} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{3.0042} & \multicolumn{1}{c|}{22} & \multicolumn{1}{c|}{8.1662} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{1.2214} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{3.3201} & \multicolumn{1}{c|}{23} & \multicolumn{1}{c|}{9.0250} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{1.3499} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{3.6693} & \multicolumn{1}{c|}{24} & \multicolumn{1}{c|}{9.9742} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{1.4918} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{4.0552} & \multicolumn{1}{c|}{25} & \multicolumn{1}{c|}{11.0232} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{1.6487} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{4.4817} & \multicolumn{1}{c|}{26} & \multicolumn{1}{c|}{12.1825} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{1.8221} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{4.9530} & \multicolumn{1}{c|}{27} & \multicolumn{1}{c|}{13.4637} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{2.0138} & \multicolumn{1}{c|}{18} & \multicolumn{1}{c|}{5.4739} & \multicolumn{1}{c|}{28} & \multicolumn{1}{c|}{14.8797} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{2.2255} & \multicolumn{1}{c|}{19} & \multicolumn{1}{c|}{6.0496} & \multicolumn{1}{c|}{29} & \multicolumn{1}{c|}{16.4446} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{2.4596} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{6.6859} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{18.1741} \\ 
\hline
\end{tabular}
\end{table}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig313.eps}
\caption{Evolution of erosion – mechanistic-empirical model}
 \label{fig:313}
% \end{center}
\end{figure}
%
By the year 26, if the infrastructure manager has not already executed an intervention, the depth of erosion will reach 12 mm, which is beyond the critical level. In order to prevent the re-bars of tanks from the onset of corrosion, an intervention (e.g. renewal of cover layer of the tanks walls) should be executed before year 26. 
\subsection{Answer A2}
Assuming that the form of the regression is linear, the general regression equation is:
\begin{eqnarray}
 && Y_i^{ero} = {\alpha _{ero}} + {\beta _{ero}} \cdot X_i^{ero} + \varepsilon _i^{ero} \label{eq3229}
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$Y_i^{ero}$:] is the value of erosion at year \textit{i}.
\item[$X_i^{ero}$:] is the year \textit{i}.
\item[${\alpha _{ero}},{\beta _{ero}}$:] are regression parameters that need to be estimated.
\item[$\varepsilon _i^{ero}$:] is the prediction errors following a normal distribution
\end{description}
%\end{flushright}
\end{adjustwidth}
Using the linear regression analysis (refer to section \ref{regressionmodel}), and in the Appendix \ref{mepi4}, following results are obtained (listing \ref{mechaempiwaste}).

\lstinputlisting[basicstyle=\ttfamily\scriptsize,float=h,frame=tb,caption=Results,label=mechaempiwaste]{./Programs/IMP-mecha-empi/IMP-A-waste-water-tanks.tex}

\begin{eqnarray}
 && Y_i^{ero} =  - 0.03630 + 0.59171 \cdot X_i^{ero} \label{eq3230}
\end{eqnarray}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=302pt]{fig314.eps}
\caption{Evolution of erosion - linear regression model}\label{fig:314}
% \end{center}
\end{figure}

The fitted line plotted from Eq. \eqref{eq3230} is shown in Figure \ref{fig:314}. When the level of erosion becomes 12 mm, it corresponds to 20 years. If the infrastructure manager relies solely on this analysis she will plan the intervention to be executed before year 20.
\subsection{Problem B - Bridge wear out}
An infrastructure manager wants to make a work program for a bridge, i.e. when an intervention needs to be executed and what type of intervention. He normally models deterioration using the following mechanistic-empirical model:
\begin{eqnarray}
 && I = \exp (\lambda  \cdot t) - 1 \label{eq3231}
\end{eqnarray}
Where is $\lambda $is the deterioration coefficient, which is given by:
\begin{eqnarray}
 && \lambda  = \frac{{\ln (2)}}{{Tc}} \label{eq3232}
\end{eqnarray}
In Eq. \eqref{eq3231}, $Tc$ represents the average life time of the structure elements of the bridge. It is modeled using a normal distribution with mean and standard deviation equal to 60 years and 7 years, respectively. A detail description of the model can be referred to the work of \cite{Brodsky2006}. The manager divides condition into five discrete states given in Table \ref{tbl:37}. The wear out of the bridge has its value from 0 to 1 as defined in Eq. \eqref{eq3232}.
\begin{table}
%	\centering
	\caption{Definition of condition states} \label{tbl:37}
\begin{tabular}{|l|l|l|}
\hline
\multicolumn{1}{|c|}{Condition states} & \multicolumn{1}{c|}{Description} & \multicolumn{1}{c|}{Corresponding values of I} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{New/Partially new} & \multicolumn{1}{c|}{$I<0.2$} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{Good} & \multicolumn{1}{c|}{$0.2\le I <0.4$} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{Average} & \multicolumn{1}{c|}{$0.4\le I <0.6$} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{Bad} & \multicolumn{1}{c|}{$0.6\le I <0.8$} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{Very bad} & \multicolumn{1}{c|}{$I\ge0.8$}\\
\hline
\end{tabular}
\end{table}
\subsection{Question}
What is the probabilistic distribution of discrete condition states over the next 50 years for the bridge? What does this mean?

\subsection{Answer}
As the value of $Tc$ is probabilistic in nature, the value of I is also probabilistic. A Monte Carlo simulation with 20'000 iterations is used to estimate the probability of \textit{I} having values that correspond with each of the condition states over time. The results are shown in Figure \ref{fig:315}. The R code in the Appendix \ref{mepi5} is used to estimate the probability of the bridge being in each condition state over 50 years. 
\begin{figure}[h]
% \begin{center}
\includegraphics[width=300pt]{fig315.eps}
\caption{Condition state distribution over 50 years}\label{fig:315}
% \end{center}
\end{figure}
As can be seen from the figure, in the first 10 years after the construction of the bridge, the bridge will remains in CS1. Then, the probability that the bridge is in CS1 decreases and the probability that it will be in CS2 increases. After 30 years, there is a 50\% probability that the bridge will be in CS2.

After 30 years, there is a small probability that the bridge will be in CS3, and after 40 years, a 50\% probability that the bridge will be in CS 4. The speed of degradation increases rapidly after 30 years. By year 50, the probability that the bridge is in CS 5 is more than 50\%.

Based on this information the infrastructure manager might consider executing an intervention before year 40 to avoid an inadequate LOS from occurring. The steps of numerical calculation can be tracked by reading the R script for this example provided at the the Appendix \ref{mepi5}.

%
\bibliographystyle{plainnat}
\bibliography{reference}
\pagebreak

\begin{subappendices}
% \label{appendix3}
% \label{appendix3}
\section{R code for the mechanistic-empirical model-concrete bridge slab}\label{mepi1}
% \textbf{R code for the mechanistic-empirical model-concrete bridge slab}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-mecha-empi/IMP-Mechanistic-Emprical-model.r}
\pagebreak
\section{R code for generating data}\label{mepi2}
% \textbf{R code for generating data}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-mecha-empi/IMP-data-generation.r}
\pagebreak
\section{R code for regression analysis} \label{mepi3}
% \textbf{R code for regression analysis}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-mecha-empi/IMP-regression.r}
\pagebreak
\section{R code for estimation of erosion evolution over 30 years} \label{mepi4}
% \textbf{R code for estimation of erosion evolution over 30 years}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-mecha-empi/IMP-A-waste-water-tanks.r}
\pagebreak
\section{R code for chloride concentration} \label{mepi5}
% \textbf{R code for chloride concentration}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-mecha-empi/IMP-A-bridge-wear-out.r}
\end{subappendices}
% \pagebreak
% \section*{Appendix 3.4} \label{mepi4}
% \textbf{R code for regression analysis}
% \lstinputlisting{./Programs/IMP-chapter3-R/IMP-4-regression.r}

### 2019-05-05 InfrastructureMaintenance/Chapters/part1.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%%part.tex%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% sample part title
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{partbacktext}
\part{Introduction}
\noindent Use the template \emph{part.tex} together with the Springer document class SVMono (monograph-type books) or SVMult (edited books) to style your part title page and, if desired, a short introductory text (maximum one page) on its verso page in the Springer layout.

\end{partbacktext}

### 2019-05-05 InfrastructureMaintenance/Chapters/reliability.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%% chapter.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample chapter
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%
%\motto{Use the template \emph{chapter.tex} to style the various elements of your chapter content.}
\chapter{Level of services indicators - reliability }
\label{reliabilityc} % Always give a unique label
% use \chaptermark{}
% to alter or adjust the chapter heading in the running head
\chapterauthor{Bryan T. Adey and Nam Lethanh}
\section{Introduction}
In order to define an adequate LOS, to determine if an adequate LOS is provided,
and to determine optimal intervention strategies and work programs, it would be
best to focus on the maximizing the net benefit from the road infrastructure, as
defined in an all-inclusive impact hierarchy over a specific time period. Due to
the significant effort required to determine the values of some impact types, and
collecting these values over time, proxies are often used. Four commonly used
proxies are the reliability, availability, maintainability and safety of the
infrastructure. Of these, reliability is discussed in the remaining part of this
document. Availability and maintainability are discussed in the subsequent
document. Safety refers to the likelihood of persons being injured or killed due
to the infrastructure or use of the infrastructure. It is not discussed here
further. The definitions of the three performance indicators discussed are:

\begin{itemize}
	\item Reliability is the probability that an item to provide an adequate LOS.
	\item Availability is the proportion of time an item provides an adequate LOS.
	\item Maintainability is the ease with which an item can be maintained
\end{itemize}

The use of the word item in all definitions is to allow it to be replaced by any
term that refers to physical infrastructure that is composed of smaller parts.
For example, a network is comprised of objects, objects are comprised of
elements, and elements are comprised of segments. In such a case, the word item
in each of the three definitions could be replaced by either the word
``network'', the word ``object'', the word ``element'' or the word ``segment''.
For example, reliability is the ability of a network to provide an adequate level
of service, or reliability is the ability of an object to provide an adequate
level of service.
%
\section{Reliability of an item}
\subsection{General}
When managing infrastructure the reliability of an item is often a useful bit of
information, i.e. how likely is it that an item is going to provide an adequate
LOS for a specified period of time. If one speaks of the collapse of a bridge as
being the first level of inadequate service, one often refers to structural
reliability.

If the adequate LOS and the specified time period are defined, then one is
interested in knowing the reliability of the item. If the adequate LOS is defined
but not the specified time period then one is interested in knowing the amount of
time required until an item provides an inadequate LOS. This is, often referred
to as the service life of the item

It is important to realize that there is a tight connection between reliability,
the definition of adequate LOS, and the service life of an item. For example, two
identical items could have the same reliability, if one an adequate LOS is
defined to be high and the service life short, and the other one had an adequate
LOS defined to be low and the service life long.
\subsection{Reliability of an item over a specified time period}
The reliability of an item in the time interval from 0 to \textit{t} is
equivalent to the probability that the item provides an adequate LOS for a longer
time than \textit{t}, which can be expressed as:
\begin{eqnarray}
&& {R_i}\left( t \right) = P\left( {{\tau _i} > t} \right)
\label{eqreliability:1}
%1
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$\tau$:] is the amount of time that an item provides an adequate LOS,
\item[$i$:] is the item identifier
\end{description}
%\end{flushright}
\end{adjustwidth}
When item \textit{i} provides an adequate LOS for $\tau $ \textit{tus,} which is
a continuous positive random variable with c.d.f, ${\tau _i}$ represents the
amounts of time that the items provide adequate LOSs.

Mathematically,
\begin{eqnarray}
&& \begin{array}{l}
{X_i}\left( t \right) = 1\;\hspace{2mm}if \hspace{2mm}{\tau _i} > t\\
{X_i}\left( t \right) = 0\;\hspace{2mm}if\hspace{2mm}{\tau _i} \le t
\end{array}
\label{eqreliability:2}
%(2)
\end{eqnarray}
When multiple items are being investigated \textit{\textbf{X}(t) =
(X$_{1}$(t),\ldots{}X$_{n}$(t)) }is the item state vector at time \textit{t.}

In the estimation of reliability of items over multiple time periods \textit{t},
it is important to realize that it is the same items that survive each successive
interval. In other words, it is assumed that the items provide an adequate LOS at
t = 0, that there are no interventions executed on the items over the time period
investigated, and that the probability of an item providing an inadequate LOS at
\textit{t} = 2 depends on it providing an adequate level of service until
\textit{t} = 2, i.e. it is conditional on the item surviving until \textit{t} =
2.
\subsection{Conditional probability of failure and failure rate}
The probability of failure in the time interval \textit{(t, t+$\Delta{}$],}
given that the item provides an adequate LOS at \textit{t}, is the conditional
probability of failure, as shown in Equation \eqref{eqreliability:3}.
\begin{eqnarray}
&& P\left( t \le \tau \le t+\Delta \right|\tau > t) = \frac{P\left( \text{failure appears in } (t,t+\Delta) \right)}{P(\tau >t)}
\label{eqreliability:3}
%(2)
\end{eqnarray}
If the \textit{$\tau{}$} is a positive random variable with d.f. \textit{f(t)}
and c.d.f \textit{F(t)}, then for small delta the conditional failure probability
in \textit{(t, t+$\Delta{}$]} given the item is operational at \textit{t} is, for
small delta, approximately equal to the failure rate \textit{f(t) / (1-F(t)),
i.e.}
\begin{eqnarray}
&& P\left( t \le \tau \le t+\Delta \right|\tau > t) = \frac{f(t)\cdot \Delta}{1-F(t)}
\label{eqreliability:4}
%(2)
\end{eqnarray}
\begin{eqnarray}
&& {\rm{Prob}}(t \le \tau  \le t + \Delta t|\tau  \ge t) = \lambda (t)\Delta t =
\frac{{f(t)\Delta t}}{{\tilde F(t)}}
\label{eqreliability:5}
%(4)
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[$f(t)$:] is the probability that an item will not provide an adequate LOS at any instance from 0 to $t$,
\item[$1-F(t)$:] is the probability that an item provides an adequate LOS from 0 to $t$. This is often referred to as the survival function. 
\end{description}
%\end{flushright}
\end{adjustwidth}
The failure rate can then be expressed as:
\begin{eqnarray}
&& \lambda (t) = \frac{{f(t)}}{{\tilde F(t)}}
%(4)
\end{eqnarray}
Note that \textit{$\lambda{}$(t)} is defined only for \textit{t} such that
\textit{F(t) $<$ 1}.
%
\subsection{Failure rates}
As can be seen from equation \eqref{eqreliability:5}, the failure rate is a time dependent, i.e. it
is not necessarily constant over time. For example, some engineering systems have
failure rates that exhibit bathtub curve characteristics (Figure \ref{reliability:1}).
It is common to divide these failure rates over time for these systems in three
phases. In the first phase is there is a monotonic decrease in the failure rate,
in the second phase the failure rate is more or less constant, and in the third
phase the failure rate monotonically increases.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=326pt]{reliability-1.eps}
\caption{Bathtub curve}\label{reliability:1}
% \end{center}
\end{figure}
When the failure can be considered to be constant, it can be approximated by
assuming that the probability of failure can be modeled using an exponential
distribution function. In this case,
\begin{eqnarray}
&& \lambda (t) = \theta 
%(5)
\end{eqnarray}
where the probability density function and the cumulative distribution function
are:
\begin{eqnarray}
&& f(t) = \theta \exp ( - \theta  \cdot t) \\ 
&& F(t) = 1- \exp ( - \theta  \cdot t)
\end{eqnarray}
\subsection{Failure rate and rate of occurrence of failures}
Failure rate should not be confused with the rate of occurrence of failures.
Failure rate is defined only for a random variable describing the lifetime of a
non-renewable item. The rate of occurrence of failures is the derivative of the
mean number of failures \textit{[0,t]} with respect to \textit{t} of a point
process N\textit{(t), t $>$= 0,} describes failures of a renewable item, i.e.
\begin{eqnarray}
&& v(t)=\frac{E\left( N(t) \right)}{dt} \label{eqreliability:6}
\end{eqnarray}
\subsection{Reliability of an item with a time invariant failure rate}
When an item has a constant failure rate, the reliability of the item, which is
sometimes referred to as the survival probability of the item, is given as:
\begin{eqnarray}
&& R(t) = \tilde F(t) = 1 - F(t) = \exp ( - \theta  \cdot t)
\label{eqreliability:7}
%(9)
\end{eqnarray}
Example of how the reliability of the item changes over time for three different
constant failure rates is given in Figure \ref{reliability-2} and the corresponding
probability density functions are given in Figure \ref{reliability-3}.
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=386pt]{reliability-2.eps}
\caption{Reliability}\label{reliability-2}
% \end{center}
\end{figure}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=394pt]{reliability-3.eps}
\caption{Probability density function}\label{reliability-3}
% \end{center}
\end{figure}
Interestingly the reliability of items with constant failure rates, regardless
of their values, is approximately 37\% at the mean life of the item. This can be
seen easily mathematically, since the length of time that an item provides an
adequate LOS if it has a constant failure rate can be calculated as:
\begin{eqnarray}
&& \Theta  = \int\limits_0^\infty  {\tilde F(t)dt}  = \int\limits_0^\infty  {\exp (
- \theta  \cdot t)dt}  = \frac{1}{\theta } \label{eqreliability:8}
%(10)
\end{eqnarray}
where $\theta $ is the failure rate and its reliability is then given by:
\begin{eqnarray}
&& R(t) = R(\frac{1}{\theta }) = \exp ( - \theta  \cdot \frac{1}{\theta }) = \exp (
- 1) = 0.3678794 \approx 0.37
\label{eqreliability:9}
%(11)
\end{eqnarray}

No matter what the value of the failure rate, the reliability at the mean
service life is 0.37, i.e. if the failure rate is 0.5, and, therefore, the mean
service life is 2 \textit{tus}, then the probability that the item provides an
adequate LOS for longer than 2 \textit{tus} is 0.37. This is illustrated
graphically in Figure \ref{reliability-4}, for items with failure rates of 0.5, 1, 1.5,
where the mean service lives are 2, 1 and 0.67.
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=346pt]{reliability-4.eps}
\caption{Probable service life and reliability}\label{reliability-4}
% \end{center}
\end{figure}
The codes for generating these graphs are given in Appendix \ref{appen31}.
\subsection{Estimating the remaining amount of time an item will provide an
adequate LOS}
In many cases when it is know that an item has provided an adequate LOS until
\textit{t} it is desired to know how much longer it might continue to provide an
adequate LOS. For example, if it known that an item provides an adequate LOS for
37 years and it has now been in service 37 years, what is the expected amount of
time that it will continue to provide an adequate LOS. How to calculate this was
explained in class.
\subsection{Reliability of an item with time-variant failure rate} \label{reliaweibl}
When an item has a time variant failure rate, the probability of failure cannot
be modeled using an exponential distribution. If it is modeled with the Weibull
function, a common selection, the reliability of the item can be estimated as:
\begin{eqnarray}
&& R(t) = {e^{ - \alpha{{( t)}^m}}}
\label{eqreliability:10}
%(12)
\end{eqnarray}
where $\alpha $ and m are scale and shape parameters, respectively.

In this case the probability density function is:
\begin{eqnarray}
&& f(t) = \alpha  \cdot m \cdot {t^{m - 1}} \cdot {e^{ - \alpha{{(t)}^m}}}
\label{eqreliability:11}
%(13)
\end{eqnarray}
And the failure rate
\begin{eqnarray}
&& \lambda (t) = \alpha  \cdot m \cdot {t^{m - 1}}
\label{eqreliability:12}
%(14)
\end{eqnarray}
Here, it can be see that if the value of $m=1$, this equation is identical to the
one used to estimate the constant failure rate.
\section{Estimating reliability when LOS data available}
The reliability of an item can be estimated in different ways. The way to be
used depends on the availability of information and the composition or structure
of the item (e.g. the item can be composed of many sub-items). If data is
available it can be used as described above to estimate the failure rate of an
item and then estimate the reliability of an item. An example is shown for a case
where the failure is time invariant and a case where the failure is time variant.
\subsection{Example 1: Reliability of an item with time invariant failure rate}
A railway tests 10 switches, over 4'180 days of use under the same loads. It was
observed that:
\begin{itemize}
	\item Switch 1 failed in 75 days
	\item Switch 2 failed in 125 days
	\item Switch 3 failed in 130 days
	\item Switch 4 failed in 325 days
	\item Switch 5 failed in 525 days
\end{itemize}
The failure rate is estimated as
\begin{eqnarray}
\theta  = \frac{5}{{4'180}} = 0.001196
\nonumber
%(15)
\end{eqnarray}
Assuming that this rate is constant, then the reliability of a switch of this
type is estimated as:
\begin{eqnarray}
R(t) = \exp ( - \theta  \cdot t) = \exp ( - 0.001196 \cdot t)
\nonumber
%(16)
\end{eqnarray}
Which shows, for example, that the reliability of a switch, i.e. the probability
that a switch will provide an adequate LOS for 1000 days, is 0.302.
\subsection{Example 2: Reliability of an item with time-variant failure rate}
The lights in street lamps frequently burn out. Regular inspections are done to
see if the lights are function or not. The data shown in Table \ref{tbl-reliability:1} has
been collected. In the table, the column ``time'' is the units of time that the
items have been in service. The column ``status'' refers to the state of the
lighting, 1 is ``burnt out'' and 0 is ``working''.
\begin{table}[h]
\caption{Failure data of the lighting system}
\vspace{3pt} \noindent
\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} & \multicolumn{1}{c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} & \multicolumn{1}{c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} & \multicolumn{1}{c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} & \multicolumn{1}{c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{9} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{28} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{161} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{21} & \multicolumn{1}{c|}{33} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{7} & \multicolumn{1}{c|}{31} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{22} & \multicolumn{1}{c|}{43} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{34} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{18} & \multicolumn{1}{c|}{23} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{23} & \multicolumn{1}{c|}{45} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{18} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{9} & \multicolumn{1}{c|}{45} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{19} & \multicolumn{1}{c|}{27} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{23} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{48} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\end{tabular}\vspace{2pt}
\label{tbl-reliability:1}\end{table}
Using the Weibull distribution as described in equation \eqref{eqreliability:11}, the value of
$\theta $ and m are 0.027 and 1.097, respectively. These values are estimated
using the Maximum likelihood estimation approach. The R code is given in Appendix
\ref{appen32}.

With these values, the failure rate is given by:
\begin{eqnarray}
&& \lambda (t) = 0.027 \times 1.097 \cdot {t^{1.097 - 1}}
\nonumber
%(17)
\end{eqnarray}
And the probability density function and the reliability function are therefore:
\begin{eqnarray}
&& f(t) = 0.027 \times 1.097 \cdot {t^{1.097 - 1}} \cdot {e^{ - {{(0.027 \cdot
t)}^{1.097}}}}
\nonumber
%(18)
\end{eqnarray}
\begin{eqnarray}
&& R(t) = {e^{ - {{(0.027 \cdot t)}^{1.097}}}}
\nonumber
%(19)
\end{eqnarray}
The evolution of reliability and failure rate over time are shown in Figure \ref{reliability-5}\pagebreak{}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=436pt]{reliability-5.eps}
\caption{Reliability and failure rate - Weibull distribution
function}\label{reliability-5}
% \end{center}
\end{figure}
\section{Estimating reliability when LOS data not available}
There are many cases where the LOS data is not available, e.g. the failure of a
bridge. One way to estimate the reliability of the item in this case is to model
the uncertainty in the variables of the models used and compare the expected
resistance to the expected loads.
%
\subsection{Based on models and uncertainty of variables}
%
In this case a model of the performance of the infrastructure object is selected
and laboratory tests or simulations are done to determine probabilistic
distributions to represent the uncertainties in key parameters in the models. For
example, if one was to estimate the reliability of a bridge with respect to
landslides, one almost certainly does not have enough data to say with any
confidence the reliability of a bridge in the future, so one may embark upon
determining the probability that the bridge would continue to provide an adequate
LOS if it came in contact with specific volumes of rock. This is done by modeling
the performance of the bridge if it came in contact with rock falls of different
intensities, taking into consideration, for example, the strength of the
concrete, the strength of the reinforcement, the location of the reinforcement,
and how the elements in the bridge work together as a system. Such information is
often included in fragility curves. An example is given in  Figure \ref{reliability-6} and more can be found in  \cite{Schultz2010} and \cite{Lethanh2015}
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=418pt]{reliability-6.eps}
\caption{Example bridge deck fragility curves}\label{reliability-6}
% \end{center}
\end{figure}
The fragility curves shown in  Figure \ref{reliability-6} represent the probability (vertical axis) of
bridge deck failure, taking into consideration the initial condition of the
bridge deck, i.e. the manifest deterioration state of the bridge deck (\textit{i
}= 1, 2, and 3) prior to hazard occurrence given the volume of rock fall that
comes in contact with the deck. As can be seen from the figure, if 15 m$^{3}$ of
rock come in contact with a bridge deck in state 1, there is a 0.089 probability
(point A) that the bridge deck would fail. It can also be seen that the higher
the initial state the higher the probability of bridge deck failure. e.g. when
the bridge deck is initially in state 2 and state 3 the probability that the
bridge deck would fail is 0.372 and 0.716, respectively).

The probability of each failure state, given the intensity of the rock fall,
which are indicated with different curves in Figure \ref{reliability-6}, are given by
\begin{eqnarray}
&& \Omega_{s,i}^l=prob\left[CS>l|s,i\right] \label{eqreliability:13}
%(18)
\end{eqnarray}
where the probability of exceeding condition state is conditional on the intensity of the load effect event and the
condition state of the object, immediately prior to occurrence of the load effect event.

Of course, to estimate the reliability of the bridge, in general, the probability of occurrence of the load effect event itself has to be estimated, i.e.
\begin{eqnarray}
&& H_s(t)=Prob\left[\text{intensity} > s(0,t) \right] \label{eqreliability:14}
%(18)
\end{eqnarray}
\subsection{Based on reliability of sub-items}
When data does not exist on the LOS of the item itself, but it does on the
sub-items of which it is composed, the sub-items can be combined in ways to allow
the estimation of the reliability of the item. Four basic ways to combine things
are as sub-items in series, sub-items in parallel, parallel sub-items connected
in series, and series sub-items connected in parallel. These are explained in the
following four sub-sections.
\subsubsection{Sub-items in series}
If the sub-items are in series (e.g. Figure \ref{reliability-17}) and the LOS provided
by the item is equal to the LOS provided by the worst sub-item, then the LOS
provided by the item is given by:
\begin{eqnarray}
&& \Phi(\overrightarrow{x})=\prod_{i=1}^n x_i= \min\limits_{1 \leq i \leq n} x_i \label{eqreliability:15}
%(18)
\end{eqnarray}
where \textit{x$_{i}$} is the state of sub-item \textit{i}, \textit{i} =
\textit{1,\ldots{}n}, and it is represented as a binary variable, where 1
indicates the sub-item provides an adequate LOS and the 0 indicates that it does
not.

The reliability that the item will provide an adequate LOS is, therefore, given by:
\begin{eqnarray}
&& \vec R(t) = \prod\limits_{n = 1}^N {{R_n}(t)}
\label{eqreliability:17}
%(22)
\end{eqnarray}

\begin{figure}[h]
% \begin{center}
\includegraphics[width=410pt]{reliability-17.eps}
\caption{Sub-items in series}\label{reliability-17}
% \end{center}
\end{figure}
for example, a bridge can be considered to provide an adequate LOS if and only
if the abutment 1 and 2, and the deck provide adequate LOSs, and the probabilities
of each providing an inadequate LOS in one year are 0.01, 0.02 and 0.03,
respectively, and their failure rates are constant, then the reliabilities of the
bridge over 5 years are:

\begin{eqnarray}
&& \begin{array}{l}
\vec R(t = 5) = \prod\limits_{n = 1}^N {{R_n}(5)}  = {R_{abutment1}}(5) \cdot
{R_{deck}}(5) \cdot {R_{abutment2}}(5)\\
\hspace{20mm}= {e^{ - 0.02 \times 5}} \cdot {e^{ - 0.01 \times 5}} \cdot {e^{ - 0.03 \times
5}} = {e^{ - (0.02 + 0.01 + 0.03) \cdot 5}} \approx 0.74
\end{array}
%\label{eqreliability:18}
%(23)
\end{eqnarray}
\subsubsection{Sub-items in parallels}
If the sub-items are in parallel (e.g. Figure \ref{reliability-19}), and the LOS provided
by the item is equal to the LOS provided by the best sub-item, then the LOS
provided by the item is given by:
\begin{eqnarray}
&& \Phi(\overrightarrow{x})=1-\prod_{i=1}^n (1-x_i)= \max\limits_{1 \leq i \leq n} x_i \label{eqreliability:18}
%(18)
\end{eqnarray}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=399pt]{reliability-19.eps}
\caption{Sub-items in parallel}\label{reliability-19}
% \end{center}
\end{figure}
The reliability that the item will provide an adequate LOS is, therefore, given by:
\begin{eqnarray}
&& \vec R(t) = 1 - \prod\limits_{i = 1}^N {\left( {1 - {R_n}(t)} \right)}
\label{eqreliability:19}
%(25)
\end{eqnarray}
\subsubsection{Parallel sub-items connected in series}
In some cases items are comprised of sub-items that can be thought of as being
groups of sub-items connected in parallel where the groups are then connected in
series (e.g. Figure \ref{reliability-21}). In this case the LOS provided by the item is
connected to the LOS provided by the sub-items as follows:
\begin{eqnarray}
&& \Phi (\vec x) = \left[ {1 - (1 - {x_1}) \cdot (1 - {x_2})} \right] \cdot \left[ {1 - (1
- {x_3}) \cdot (1 - {x_4})} \right]
\label{eqreliability:201}
%(27)
\end{eqnarray}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{reliability-21.eps}
\caption{Parallel sub-items connected in series}\label{reliability-21}
% \end{center}
\end{figure}
The reliability that the item will provide an adequate LOS is, therefore, given by:
\begin{eqnarray}
&& \vec R = \left[ {1 - (1 - {R_1}) \cdot (1 - {R_2})} \right] \cdot \left[ {1 - (1
- {R_3}) \cdot (1 - {R_4})} \right]
\label{eqreliability:20}
%(27)
\end{eqnarray}
\subsubsection{Series sub-items connected in parallel}
In some cases items are comprised of sub-items that can be thought of as being
groups of sub-items connected in series where the groups are then connected in
parallel (e.g. Figure \ref{reliability-23}). In this case the LOS provided by the item is
connected to the LOS provided by the sub-items as follows: 
%
%(29)
\begin{eqnarray}
&& \Phi (\vec x) = 1 - (1 - {x_1} \cdot {x_2}) \cdot (1 - {x_3} \cdot {x_4} \cdot {x_5})
\label{eqreliability:211}
\end{eqnarray}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{reliability-23.eps}
\caption{Series sub-items connected in parallel}\label{reliability-23}
% \end{center}
\end{figure}
The reliability that the item will provide an adequate LOS is, therefore, given by:
\begin{eqnarray}
&& \vec R = 1 - (1 - {R_1} \cdot {R_2}) \cdot (1 - {R_3} \cdot {R_4} \cdot {R_5})
\label{eqreliability:21}
\end{eqnarray}
\subsubsection{Complex configurations}
As not all items can be thought of as neat combinations of series and parallel
sub-items, it is useful to have a systematic way to determine the structure
function for the item. Three ways are pivotal decomposition, minimal paths and
minimal cuts.
\paragraph{Pivotal decomposition}
Pivotal decomposition is a method to determine the reliability of an item by
replacing unreliable sub-items with either 100\% reliable sub-items and/or 0\%
reliable (i.e. failed) sub-items until a configuration is created for which the
reliability is easily computable. The reliability of the item is then calculated
as:
\begin{eqnarray}
&& R\left( {\overrightarrow x } \right) = {R_i} \cdot R\left(
{{1_i},\overrightarrow x } \right) + \left( {1 - {R_i}} \right) \cdot R\left(
{{0_i},\overrightarrow x } \right)
\end{eqnarray}
Where \textit{ }$\left( {{\alpha _i},\overrightarrow x }
\right)$ represents the vector of sub-items  with $\overrightarrow x $ the sub-item
\textit{i} replaced by \textit{$\alpha{}$$_{i}$}. , which is either totally
reliable and has the value of 1 or failed and has the value of 0.
\paragraph{Minimal path and minimal cuts}
In the case of minimal cuts, it is necessary to consider
\begin{itemize}
	\item the vector of sub-item states a \textit{cut vector} if \textit{$\varphi{}$}(x) =
0,
	\item the specific sub-items in the cut vector that are required to not provide an
adequate LOS to mean that the item does not provide an adequate LOS, form the
\textit{cut set}, and
	\item a cut set in which there is no possibility to improve the LOS provided by a
sub-item without improving the LOS provided by the item is a \textit{minimal cut
set}.
\end{itemize}
In the case of minimal paths, it is necessary to consider
\begin{itemize}
	\item the vector of sub-item states a \textit{path vector} if \textit{$\varphi{}$}(x)
= 1
	\item the specific sub-items in the path vector that are required to provide an
adequate LIS to mean that the item provides an adequate LOS, form the
\textit{path set}, and
	\item a path set in which there is no possibility to reduce the LOS provided by a
sub-item without reducing the LOS provided by the item is a \textit{minimal path
set}.
\end{itemize}
\paragraph{Example}
Question

Define the structure function to be used to estimate the reliability of the item
composed of 5 sub-items as shown in Figure \ref{reliability-24}.
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{reliability-24.eps}
\caption{Item with 5 sub-items in a complex structure}\label{reliability-24}
% \end{center}
\end{figure}

Answer

Using the pivotal decomposition, the item structure function can be determined
by substituting sub-item 3 with a sub-item with a reliability of 1 or 0, yielding
either equation \eqref{eqreliability:22} or \eqref{eqreliability:23},respectively. These are illustrated in Figure \ref{reliability-25} and Figure \ref{reliability-26}, respectively
\begin{eqnarray}
R({1_3},x) = \left[ {1 - (1 - {x_1})(1 - {x_2})} \right] \cdot \left[ {1 - (1 -
{x_4})(1 - {x_5})} \right]
\label{eqreliability:22}
%(30)
\end{eqnarray}
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{reliability-25.eps}
\caption{Sub-item 3 provides an adequate LOS}\label{reliability-25}
% \end{center}
\end{figure}
%
\begin{eqnarray}
&& R({0_3},\vec x) = \left[ {1 - (1 - {x_1} \cdot {x_4})} \cdot (1 - {x_2} \cdot {x_5}) \right]
\label{eqreliability:23}
%(31)
\end{eqnarray}
\begin{figure}[h]
% \begin{center}
\includegraphics[width=450pt]{reliability-26.eps}
\caption{Sub-item 3 does not provide an adequate LOS}\label{reliability-26}
% \end{center}
\end{figure}
This means that the structure function of the item to be used to estimate the
reliability of the item is:
\begin{eqnarray}
&& R(\vec x) = {R_3} \cdot R({1_3},\vec x) + (1 - {R_3}) \cdot R({0_3},\vec x)
\label{eqreliability:24}
%(32)
\end{eqnarray}
\begin{eqnarray}
&& \begin{array}{l}
R(\vec x) = {R_1}{R_3}{R_4} + {R_1}{R_3}{R_5} + {R_2}{R_3}{R_4} +
{R_2}{R_3}{R_5} + {R_1}{R_2}{R_4}{R_5}\\
{\rm{          }} - {R_1}{R_2}{R_3}{R_4} - {R_1}{R_3}{R_4}{R_5} -
{R_1}{R_2}{R_3}{R_5} - {R_2}{R_3}{R_4}{R_5}
\end{array}
\label{eqreliability:25}
%(33)
\end{eqnarray}

Question

Show that the minimal path method will give the same answer as pivotal
decomposition.

Answer

The ``minimal path sets'' include
\begin{eqnarray}
&& \begin{array}{l}
{P_1} = (1,4)\\
{P_2} = (1,3,5)\\
{P_3} = (2,5)\\
{P_1} = (2,3,4)
\end{array} \nonumber
\end{eqnarray}
%
\begin{eqnarray}
&& \begin{array}{l}
\phi (\vec x) = 1 - \prod\limits_{s = 1}^S {\left( {1 - \prod\limits_{n \in
{P_s}} {{x_n}} } \right)} \\
{\rm{         = }}1 - (1 - {x_1}{x_4})(1 - {x_1}{x_3}{x_5})(1 - {x_2}{x_5})(1 -
{x_2}{x_3}{x_4})\\
{\rm{         = }}{x_1}{x_3}{x_4} + {x_1}{x_3}{x_5} + {x_2}{x_3}{x_4} +
{x_2}{x_3}{x_5} + {x_1}{x_2}{x_4}{x_5}\\
{\rm{          }} - {x_1}{x_2}{x_3}{x_4} - {x_1}{x_3}{x_4}{x_5} -
{x_1}{x_2}{x_3}{x_5} - {x_2}{x_3}{x_4}{x_5}
\end{array}
\label{eqreliability:26}
%(34)
\end{eqnarray}
\section{Conclusion}
Once adequate LOSs are defined, one is often interested in estimating the
likelihood that they will be provided. Without thinking of the specific
consequences that might occur, reliability can be used as a performance
indicator. Keep in mind though that the consequences related to the occurrence of
inadequate services are almost never constant, and, therefore, focusing
exclusively on reliability as a performance indicator will not allow the
prioriotization of activities to maximize net benefits related to infrastructure.
An example is included at the end of the next chapter.

When using reliability as a performance indicator for an item, there are
multiple ways to estimate the reliability. When using reliability as a
performance indicator for a sub-item one should not forget that the LOS it
provides depends on how it is integrated into the item as a whole and the
reliability of the other sub-items in the item. There are different ways to
measure this that vary in terms of accuracy and effort.
%
\section{Assignments}
\subsection{Problem A}
\subsubsection{Question A1}
Determine the reliability of the road network shown in Figure \ref{reliability-a-1}, when
the reliability of each object is as given in Table \ref{tbl-reliability-a:1}.
\begin{figure}[h]
% \begin{center}
\includegraphics[width=406pt]{reliability-a-1.eps}
\caption{A network of 7 objects}\label{reliability-a-1}
% \end{center}
\end{figure}
\begin{table}[h]
\caption{Object reliabilities}
% \vspace{3pt} \noindent
\begin{tabular}{|l|l|}
\hline
\multicolumn{1}{|c|}{Object} & \multicolumn{1}{c|}{Reliability} \\ 
\hline
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{0.95} \\ 
\hline
\multicolumn{1}{|c|}{B} & \multicolumn{1}{c|}{0.97} \\ 
\hline
\multicolumn{1}{|c|}{C} & \multicolumn{1}{c|}{0.92} \\ 
\hline
\multicolumn{1}{|c|}{D} & \multicolumn{1}{c|}{0.94} \\ 
\hline
\multicolumn{1}{|c|}{E} & \multicolumn{1}{c|}{0.90} \\ 
\hline
\multicolumn{1}{|c|}{F} & \multicolumn{1}{c|}{0.88} \\ 
\hline
\multicolumn{1}{|c|}{G} & \multicolumn{1}{c|}{0.98} \\ 
\hline
\end{tabular}
% \vspace{2pt}
\label{tbl-reliability-a:1}\end{table}
\subsubsection{Answer A1}
Seeing that the network can be broken down into series and parallel networks and
remembering that the structure function for series, and parallel objects are as
given in Equations \eqref{eqreliability:17} and \eqref{eqreliability:19}, respectively.
\begin{eqnarray}
&& {\vec R_C} = 1 - \left[ {1 - ({R_A} \cdot {R_B} \cdot {R_{CD}} \cdot {R_{EF}})}
\right] \cdot \left[ {1 - {R_G}} \right]
\label{eq-reliability-a3}
%(3)
\end{eqnarray}
where ${R_{CD}},{R_{EF}}$ represents the reliability of the parallel components
composed of item C, D and E, F, respectively. ${\vec R_{Network}}$ represents
reliability of the network configuration.
%
\begin{eqnarray}
&& \begin{array}{l}
{{\vec R}_{Network}} = 1 - \left[ {1 - ({R_A} \cdot {R_B} \cdot {R_{CD}} \cdot
{R_{EF}})} \right] \cdot \left[ {1 - {R_G}} \right]\\
{\rm{      =  }}1 - \left[ {1 - ({R_A} \cdot {R_B} \cdot \left\{ {(1 - (1 -
{R_C}) \cdot (1 - {R_D})} \right\} \cdot \left\{ {(1 - (1 - {R_E}) \cdot (1 -
{R_F})} \right\})} \right] \cdot \left[ {1 - {R_G}} \right]
\end{array}
\label{eq-reliability-a4}
%(4)
\end{eqnarray}
Using the value of reliability for each item shown in Figure \ref{fig:1},
following result is obtained
\begin{eqnarray}
&& {\vec R_{Network}} = 0.998121 \nonumber
%(3)
\end{eqnarray}
%
\subsection{Problem B}
Imagine that the infrastructure organization offers bid to external contractors
for construction and installation of its infrastructure network. The networks
consist of 17 objects (Table \ref{tbl-reliability-a:2}) and, minimum reliability of the network
reliability must be 70\%. The A, B and C bid 57'000 CHF, 39'000 CHF, and 42'000
CHF, respectively and proposed the configuration shown in Figure \ref{reliability-a-2},
Figure \ref{reliability-a-3}, and Figure \ref{reliability-a-3}.
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=408pt]{reliability-a-2.eps}
\caption{Configuration A}\label{reliability-a-2}
% \end{center}
\end{figure}
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=371pt]{reliability-a-3.eps}
\caption{Configuration B}\label{reliability-a-3}
% \end{center}
\end{figure}
%
\begin{figure}[h]
% \begin{center}
\includegraphics[width=415pt]{reliability-a-4.eps}
\caption{Configuration B}\label{reliability-a-4}
% \end{center}
\end{figure}
%
\begin{table}[h]
\caption{Object reliabilities}
\vspace{3pt} \noindent
\begin{tabular}{|p{28pt}|p{63pt}|p{28pt}|p{59pt}|p{34pt}|p{55pt}|}
\hline
\parbox{28pt}{\centering 
Object
} & \parbox{63pt}{\centering 
Reliability
} & \parbox{28pt}{\centering 
Object
} & \parbox{59pt}{\centering 
Reliability
} & \parbox{34pt}{\centering 
Object
} & \parbox{55pt}{\centering 
Reliability
} \\
\hline
\parbox{28pt}{\centering 
A
} & \parbox{63pt}{\centering 
0.840
} & \parbox{28pt}{\centering 
G
} & \parbox{59pt}{\centering 
0.870
} & \parbox{34pt}{\centering 
M
} & \parbox{55pt}{\centering 
0.830
} \\
\hline
\parbox{28pt}{\centering 
B
} & \parbox{63pt}{\centering 
0.860
} & \parbox{28pt}{\centering 
H
} & \parbox{59pt}{\centering 
0.880
} & \parbox{34pt}{\centering 
N
} & \parbox{55pt}{\centering 
0.850
} \\
\hline
\parbox{28pt}{\centering 
C
} & \parbox{63pt}{\centering 
0.890
} & \parbox{28pt}{\centering 
I
} & \parbox{59pt}{\centering 
0.890
} & \parbox{34pt}{\centering 
O
} & \parbox{55pt}{\centering 
0.840
} \\
\hline
\parbox{28pt}{\centering 
D
} & \parbox{63pt}{\centering 
0.860
} & \parbox{28pt}{\centering 
J
} & \parbox{59pt}{\centering 
0.860
} & \parbox{34pt}{\centering 
P
} & \parbox{55pt}{\centering 
0.890
} \\
\hline
\parbox{28pt}{\centering 
E
} & \parbox{63pt}{\centering 
0.870
} & \parbox{28pt}{\centering 
K
} & \parbox{59pt}{\centering 
0.850
} & \parbox{34pt}{\centering 
Q
} & \parbox{55pt}{\centering 
0.890
} \\
\hline
\parbox{28pt}{\centering 
F
} & \parbox{63pt}{\centering 
0.820
} & \parbox{28pt}{\centering 
L
} & \parbox{59pt}{\centering 
0.860
} & \parbox{34pt}{\centering } & \parbox{55pt}{\centering } \\
\hline
\end{tabular}
\vspace{2pt}
\label{tbl-reliability-a:2}\end{table}
\subsubsection{Question B1}
Given this information, determine the system reliability for each of the three
configurations.
\subsubsection{Answer B1}
Configurations A, B, and C are composed of both items connected in series and
parallel. Equations \eqref{eqreliability:17} and \eqref{eqreliability:19} are used.

One possible formula to be used to determine the reliability of configuration A,
B and C are:

\underline{Configuration A}

\begin{itemize}
	\item B and C are in the same link (Eq. \eqref{eqreliability:17})
	\item D and E are in the same link (Eq. \eqref{eqreliability:17})
	\item \{B,C\} and \{D, E\} are in parallel (Eq. \eqref{eqreliability:19})
	\item G and H are in parallel (Eq. \eqref{eqreliability:19})
	\item Set 1 = [\{(B,C),(D,E)\}, F, (G,H), and I] are in the same link  (Eq. \eqref{eqreliability:17})
	\item K and L are in parallel (Eq. \eqref{eqreliability:19})
	\item O, P, and Q are in parallel (Eq. \eqref{eqreliability:19})
	\item Set 2 = [J, (K,L), M, N, and (O,P,Q)] are in the same link (Eq. \eqref{eqreliability:17})
	\item Set 1 and 2 are in parallel (Eq. \eqref{eqreliability:19})
\end{itemize}
\begin{eqnarray}
&& {\vec R_{configuration{\rm{ }}A}} = {R_A} \cdot {R_{B \to Q}}
\label{eq-reliability-a5}
%(5)
\end{eqnarray}
where  ${R_{B \to Q}}$ represents the the reliability of the sub-system
%
\begin{eqnarray}
&& {R_{B \to Q}} = 1 - (1 - {R_{B \to I}}) \cdot (1 - {R_{J \to Q}})
\label{eq-reliability-a6}
%(6)
\end{eqnarray}
%
\begin{eqnarray}
&& \begin{array}{l}
{R_{B \to I}} = {R_{B \to E}} \cdot {R_F} \cdot {R_{G \to H}} \cdot {R_I}\\
{\rm{          = }}\left[ {1 - (1 - {R_B} \cdot {R_C}) \cdot (1 - {R_D} \cdot
{R_E})} \right] \cdot {R_F} \cdot \left[ {1 - (1 - {R_G}) \cdot (1 - {R_H})}
\right] \cdot {R_I}
\end{array}
\label{eq-reliability-a7}
%(7)
\end{eqnarray}
%%
\begin{eqnarray}
&& \begin{array}{l}
{R_{J \to Q}} = {R_J} \cdot {R_{K \to L}} \cdot {R_M} \cdot {R_N} \cdot {R_{O
\to Q}}\\
{\rm{          = }}{R_J} \cdot \left[ {1 - (1 - {R_K}) \cdot (1 - {R_L})}
\right] \cdot {R_M} \cdot {R_N} \cdot \left[ {1 - (1 - {R_O}) \cdot (1 - {R_P})
\cdot (1 - {R_Q})} \right]
\end{array}
\label{eq-reliability-a8}
%(8)
\end{eqnarray}
Using the reliability values of items in Table \ref{tbl-reliability-a:2}, the reliability of
the configuration A is
\begin{eqnarray}
&& {\vec R_{configuration{\rm{ }}A}} = 0.729 \nonumber
\end{eqnarray}

\underline{Configuration B}

\begin{eqnarray}
{\vec R_{configuration{\rm{ }}B}} = {R_A} \cdot {R_B} \cdot {R_{C \to O}}
\label{eq-reliability-a9}
%(9)
\end{eqnarray}
%
\begin{eqnarray}
&& {R_{C \to O}} = 1 - (1 - {R_{C \to J}} \cdot {R_{K \to M}}) \cdot (1 - {R_{N \to
O}})
\label{eq-reliability-a10}\\
&& {R_{C \to J}} = 1 - (1 - {R_C} \cdot {R_F} \cdot {R_G})(1 - {R_E} \cdot {R_{H
\to J}})
\label{eq-reliability-a11}\\
&&{R_{H \to J}} = 1 - (1 - {R_H}) \cdot (1 - {R_I}) \cdot (1 - {R_J})
\label{eq-reliability-a12}\\
&& {R_{K \to M}} = 1 - (1 - {R_K}) \cdot (1 - {R_L}) \cdot (1 - {R_M})
\label{eq-reliability-a13}\\
%(13)
&& {R_{N \to O}} = {R_N} \cdot {R_O}
\label{eq-reliability-a14}
%(14)
\end{eqnarray}
%
%
Using the reliability values of items in Table \ref{tbl-reliability-a:2}, the reliability of
the configuration B is
%
\begin{eqnarray}
&& {\vec R_{configuration{\rm{ }}B}} = 0.714 \nonumber
\end{eqnarray}

\underline{Configuration C}
\begin{eqnarray}
&& {\vec R_{configuration{\rm{ C}}}} = {R_A} \cdot {R_{B \to K}} \cdot {R_L} \cdot
{R_{M \to N}}
\label{eq-reliability-a15}
%(15)
\end{eqnarray}
\begin{eqnarray}
&& {R_{B \to K}} = 1 - (1 - {R_{B \to F}} \cdot {R_{H \to J}}) \cdot (1 - {R_{G \to
K}})
\label{eq-reliability-a16}\\
%(16)
&& {R_{B \to F}} = 1 - (1 - {R_B} \cdot {R_C}) \cdot (1 - {R_D} \cdot {R_E}) \cdot
(1 - {R_F})
\label{eq-reliability-a17}\\
&& {R_{H \to J}} = 1 - (1 - {R_H}) \cdot (1 - {R_I}) \cdot (1 - {R_J})
\label{eq-reliability-a18}\\
%(18)
&& {R_{G \to K}} = {R_G} \cdot {R_K}
\label{eq-reliability-a19}
%(19)
\end{eqnarray}
Using the reliability values of items in Table \ref{tbl-reliability-a:2}, the reliability of
the configuration C is
\begin{eqnarray}
&& {\vec R_{configuration{\rm{ }}C}} = 0.702\nonumber
\end{eqnarray}

\underline{Result}

The reliability of configurations A, B and C are 0.729, 0.714, and 0.702
respectively. Based on reliability alone configuration A would be chosen.
%
\subsubsection{Question B2}
How high would the failure costs have to be before you would not choose the
network with the lowest upfront costs? (Use a one year time period and assume
that the network would not failure more than once).
%
\subsubsection{Answer B2}
If failure costs nothing than network configuration results in the lowest cost
(Table \ref{tbl-reliability-a:3}).

\begin{table}[h]
\caption{Total cost of networks}
\begin{tabular}{|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Failure cost} & \multicolumn{1}{c|}{Network} & \multicolumn{1}{c|}{Reliability} & \multicolumn{1}{c|}{Construction cost} & \multicolumn{1}{c|}{Total cost in one year} \\ 
\multicolumn{1}{|c|}{($10^3$ x CHF)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{($10^3$ x CHF)"} & \multicolumn{1}{c|}{($10^3$ x CHF)} \\ 
\hline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A} & \multicolumn{1}{c|}{0.729} & \multicolumn{1}{c|}{57} & \multicolumn{1}{c|}{57} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{0} & \multicolumn{1}{c|}{B} & \multicolumn{1}{c|}{0.714} & \multicolumn{1}{c|}{39} & \multicolumn{1}{c|}{39} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{C} & \multicolumn{1}{c|}{0.702} & \multicolumn{1}{c|}{42} & \multicolumn{1}{c|}{42} \\ 
\hline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A} & \multicolumn{1}{c|}{0.729} & \multicolumn{1}{c|}{57} & \multicolumn{1}{c|}{301} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{900} & \multicolumn{1}{c|}{B} & \multicolumn{1}{c|}{0.714} & \multicolumn{1}{c|}{39} & \multicolumn{1}{c|}{296} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{C} & \multicolumn{1}{c|}{0.702} & \multicolumn{1}{c|}{42} & \multicolumn{1}{c|}{310} \\ 
\hline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A} & \multicolumn{1}{c|}{0.729} & \multicolumn{1}{c|}{57} & \multicolumn{1}{c|}{328} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{1000} & \multicolumn{1}{c|}{B} & \multicolumn{1}{c|}{0.714} & \multicolumn{1}{c|}{39} & \multicolumn{1}{c|}{325} \\ 
\cline{2-5}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{C} & \multicolumn{1}{c|}{0.702} & \multicolumn{1}{c|}{42} & \multicolumn{1}{c|}{340} \\ 
\hline
\end{tabular}
\label{tbl-reliability-a:3}
\end{table}
\subsection{Problem C}
You have a series network with 5 infrastructure objects, two pavement sections
and three bridges. The probability of failure in each of the next 5 years of each
bridge is 0.1 (before and after repair) and the probability of failure for each
road section is 0.15 (before and after repair). Each bridge failure will take 60
days to repair and will cost 250'000 mus during which time the road will be
unusable in which users will incur 300'000 mus.  Each road section failure will
take 20 days to repair and will cost 75'000 mus during which time the road will
be unusable in which users will incur 100'000 mus. Failures are considered
mutually exclusive and that only one failure can occur per year.
\subsubsection{Question C1}
Calculate the reliability of the road over the 5 year-period.
\subsubsection{Answer C1}
The reliability is defined \eqref{eqreliability:7}. The hazard rate $\theta$ can be estimated.
%
\begin{eqnarray}
&& ln[R(t)] = \ln [\exp ( - \theta  \cdot t)] \Leftrightarrow \ln [R(t)] =  -
\theta .t
\label{eq-reliability-a21} \\
&& \to \theta  =  - \frac{{\ln [R(t)]}}{t}
\label{eq-reliability-a22}
%(21)
\end{eqnarray}
The reliability in one year of the network $R(t = 1)$is estimated as
\[
R(t = 1) = (1-0.1)^3 \cdot  (1-0.15)^2 = 0.5267025
\]
The join hazard rate will be 
\[
\to \theta  =  - \frac{{\ln [0.5267025]}}{1} = 0.6411194
%(24)
\]
The reliability in 5 year of the network will be
\[
R(t = 5) = \exp ( - 0.6411194 \times 5) = 0.0405347
%(25)
\]
\subsubsection{Question C2}
Calculate the maintainability of the road (in terms of time and money).
\subsubsection{Answer C2}
The formulation to estimate the maintainability is
\begin{eqnarray}
&& \bar M = \frac{{\sum\limits_{n = 1}^N {{p_n} \cdot N \cdot \mu _n^M}
}}{{\sum\limits_{n = 1}^N {{p_n} \cdot N} }}
\label{eq-reliability-a26}
%(26)
\end{eqnarray}
Where:
\begin{adjustwidth}{1cm}{}
\begin{description}
\item[${p_n}$:] is the failure probability of item type $n$,
\item[$N$:] is the number of item type $n$ 
\item[$\mu _n^M$:] is the mean time of impacts occur during the maintenance
\end{description}
%\end{flushright}
\end{adjustwidth}
Thus, in term of time, we have
\[
{\bar M^{time}} = \frac{{0.1 \times 3 \times 60}}{{0.1 \times 3 + 0.15 \times
2}} + \frac{{0.15 \times 2 \times 20}}{{0.1 \times 3 + 0.15 \times 2}} = 40{\rm{
days}}
\]
and in term of money, we have
\[
{\bar M^{money}} = \frac{{0.1 \times 3 \times (250'000 + 300'000)}}{{0.15 \times
2 + 0.1 \times 3}} + \frac{{0.15 \times 2 \times (75'000 + 100'000)}}{{0.15
\times 2 + 0.1 \times 3}} = 362'500{\rm{ mus}}
\]
\subsubsection{Question C3}
What is the impact on reliability and maintainability of repairing one bridge
perfectly, i.e failure will not occur?
\subsubsection{Answer C3}
When one bridge is perfectly repaired, its reliability is 1 and thus the network
can be considered to have only two bridges instead of 3. The same way of
calculation for a new configuration of the network with only 2 bridges and 2 road
sections is used to compute the reliability and maintainability.

The reliability in one year of the network $R(t = 1)$ is estimated as
\[
R(t = 1) = (1-0.1)^2 \times (1-0.15)^2 = 0.585225
\]
\[
\to \theta  =  - \frac{{\ln [0.585225]}}{1} = 0.5357589
\]
The reliability in 5 year of the network will be
\[
R(t = 5) = \exp ( - 0.5357589 \times 5) = 0.06864586
%(25)
\]
\[
{\bar M^{time}} = \frac{{0.15 \times 2 \times 60}}{{0.15 \times 2 + 0.1 \times
2}} + \frac{{0.1 \times 2 \times 20}}{{0.15 \times 2 + 0.1 \times 2}} = 36{\rm{
days}}
\]
and in terms of money, we have
\[
{\bar M^{money}} = \frac{{0.15 \times 2 \times (250'000 + 300'000)}}{{0.15
\times 2 + 0.1 \times 2}} + \frac{{0.1 \times 2 \times (75'000 + 100'000)}}{{0.15
\times 2 + 0.1 \times 2}} = 325'000{\rm{ mus}}
\]
\subsubsection{Question C4}
How much should we be willing to pay to execute such an intervention?
\subsubsection{Answer C4}
The amount of money that we are willing to pay depending on the expected impact
due to failure that can be generated under the two above options. Following table
describes the calculation of the amount of money that we are willing to pay. The
option 1 and 2 are corresponding to the case one bridge is not perfectly repaired
and another case when a bridge is perfectly repaired, respectively. The
reliability and the expect cost to repair are calculated from previous steps.
From the reliability, the failure probability can be straightforward estimated
and it is used to compute the possible cost (expected cost) if failure happens in
the future. As can be seen from the amount of possible cost under option 1 and 2,
option 2 yields less amount of cost than option 1, and the willingness to pay is
the difference of the twos, which is about 45.115 mus.

\begin{table}[h]
\caption{Comparison of two options}

\vspace{3pt} \noindent
\begin{tabular}{|p{38pt}|p{108pt}|p{51pt}|p{60pt}|}
\hline
\parbox{38pt}{\centering 
{\footnotesize Option}
} & \parbox{108pt}{\centering 
{\footnotesize Reliability}
} & \parbox{51pt}{\centering 
{\footnotesize Cost (mus)}
} & \parbox{60pt}{\centering 
{\footnotesize Failure probability}
} \\
\hline
\parbox{38pt}{\centering 
{\footnotesize 1}
} & \parbox{108pt}{\centering 
{\footnotesize 0.0405347}
} & \parbox{51pt}{\centering 
{\footnotesize 362'500}
} & \parbox{60pt}{\centering 
{\footnotesize 0.9594653}
} \\
\hline
\parbox{38pt}{\centering 
{\footnotesize 2}
} & \parbox{108pt}{\centering 
{\footnotesize 0.06864586}
} & \parbox{51pt}{\centering 
{\footnotesize 325'000}
} & \parbox{60pt}{\centering 
{\footnotesize 0.93135414}
}  \\
\hline
\parbox{38pt}{\centering } & \multicolumn{3}{|c|}{\parbox{219pt}{\centering 
{\footnotesize Willingness to pay  = 45.115 mus}
}} \\
\hline
\end{tabular}
\vspace{2pt}
\end{table}
\subsubsection{Question C5}
Discuss the assumptions mutual exclusiveness and that only one failure can occur
per year.
\subsubsection{Answer C5}
The mutual exclusiveness assumed in this example means that only one failure can
occur per year. This assumption can be used but it also bears a limitation that
it might not perfectly reflect the increasing failure rate over time as the
bridges and road sections are aging year by year. In addition, as the bridges and
road sections are independent, they can actually failure to provide adequate
level of services simultaneously in the same year. This possibility should be
taken into consideration of calculating the reliability and maintainability in
reality.
\subsection{Problem D}
A network is connected and represented as shown in the following Figure
\ref{reliability-a-5}. All 6 objects are identical. In one month, on average it is recorded
that 1 out of 6 objects fails and is renewed.

\begin{figure}[h]
% \begin{center}
\includegraphics[width=354pt]{reliability-a-5.eps}
\caption{Network}\label{reliability-a-5}
% \end{center}
\end{figure}
\subsubsection{Question D1}
Calculate the reliability of the network in 5 months, i.e. that it is possible
to send one item from the start node (S) to the end note (E)? Note: you can round
up the values in 2 digits.
\subsubsection{Answer D1}
One month failure probability for each object is 1/6. The reliability for the
network in one month is defined as
\begin{eqnarray}
&& \vec R_{network}^1 = R_1^1 \cdot R_{2 \to 5}^1 \cdot R_6^1
\label{eq-reliability-a30} \\
&& R_{2 \to 5}^1 = 1 - (1 - R_2^1) \cdot \left[ {\left\{ {1 - (1 - R_3^1) \cdot (1
- R_4^1)} \right\}} \right] \cdot (1 - R_5^1)
\label{eq-reliability-a31} \\
&& \vec R_{network}^1 \approx 0.694 \nonumber
%(30)
\end{eqnarray}
The reliability in 5 months of the network will be estimated based on the
failure rate of individual item
\[
\to \theta  =  - \frac{{\ln [1 - 1/6]}}{1} \approx 0.182
%(32)
\]
The reliability of each item in 5 months is
\[
R_{item}^5 = \exp ( - 0.183 \times 5) \approx 0.402
\]
Using equations \eqref{eq-reliability-a30} and \eqref{eq-reliability-a31} for a period of 5 months, we have
\begin{eqnarray}
&& \vec R_{network}^5 = R_1^5 \cdot R_{2 \to 5}^5 \cdot R_6^5
\label{eq-reliability-a34} \\
&& R_{2 \to 5}^5 = 1 - (1 - R_2^5) \cdot \left[ {\left\{ {1 - (1 - R_3^5) \cdot (1
- R_4^5)} \right\}} \right] \cdot (1 - R_5^5)
\label{eq-reliability-a35}\\
&& \vec R_{network}^5 \approx 0.141 \nonumber
%(34)
\end{eqnarray}
\subsubsection{Question D2}
If the network is not working, a loss of 30 \textit{mus} is expected. In order
to improve the overall reliability of the network over the next 5 months, manager
proposes 2 options

\textbf{Option 1:} Replace link 1 and link 6 with links that are 5 \% more
reliable than the existing ones.

\textbf{Option 2:} Replace links 2, 3, 4, and 5 with the links that are 8\% more
reliable than the existing ones.

The differences in replacement costs are negligible and thus not considered.

Which option shall be chosen?
\subsubsection{Answer D2}
\textbf{Option 1:} Replace link 1 and link 6 with links that are 5 \% more
reliable than the existing ones.

Under this option, new reliability of link 1 and link 6 become
\[
R_1^1 = R_6^1 = (1 - \frac{1}{6}) + (1 - \frac{1}{6}) \cdot \frac{5}{{100}} =
0.875
\]

Using the new value of reliability for item 1 and 6 and the values of
reliability for item 2, 3, 4, and 5 ($R_2^1 = R_3^1 = R_4^1 = R_5^1 = (1 -
\frac{1}{6})$) for the equations (30) and (31), we have

\[
 \vec R_{network}^{1,option1} \approx 0.765
\]

\textbf{Option 2:} Replace links 2, 3, 4, and 5 with the links that are 8\% more
reliable than the existing ones.

Under this option, new reliability of link 1 and link 6 become

\[
 R_2^1 = R_3^1 = R_4^1 = R_5^1 = (1 - \frac{1}{6}) + (1 - \frac{1}{6}) \cdot
\frac{8}{{100}} = 0.9
\]

Using the new value of reliability for item 2, 3, 4 and 5 and the old values of
reliability for item 1, and 6 ($R_1^1 = R_6^1 = (1 - \frac{1}{6})$) for the
equations (30) and (31), we have

\[
 \vec R_{network}^{1,option1} \approx 0.694
\]

Similarly, for 5 months, we have

\[
 R_1^5 = R_6^5 \approx 0.402 + 0.402 \cdot \frac{5}{{100}} = 0.4221
\]

Using the new value of reliability for item 1 and 6 and the values of
reliability for item 2, 3, 4, and 5 ($R_2^5 = R_3^5 = R_4^5 = R_5^5 = 0.402$) for
the equations (30) and (31), we have

\[
 \vec R_{network}^{5,option1} \approx 0.155
\]

\textbf{Option 2:} Replace links 2, 3, 4, and 5 with the links that are 8\% more
reliable than the existing ones.

Under this option, new reliability of link 1 and link 6 become

\[
 R_2^2 = R_3^3 = R_4^4 = R_5^5 = 0.402 + 0.402 \cdot \frac{8}{{100}} = 0.434
\]

Using the new value of reliability for item 2, 3, 4 and 5 and the old values of
reliability for item 1, and 6 ($R_1^5 = R_6^5 = 0.402$) for the equations (30)
and (31), we have

\[
 \vec R_{network}^{5,option2} \approx 0.145
\]

Following table compares the expected impact under the two options
\begin{table}
\caption{Comparison of two options}
\label{tbl-reliability-a:4}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Time} & \multicolumn{1}{c|}{Options} & \multicolumn{1}{c|}{Reliability} & \multicolumn{1}{c|}{Failure} & \multicolumn{1}{c|}{Unit impact} & \multicolumn{1}{c|}{Total impact} \\ 
\multicolumn{1}{|c|}{(months)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{probability} & \multicolumn{1}{c|}{(mus)} & \multicolumn{1}{c|}{(mus)} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0.765} & \multicolumn{1}{c|}{0.235} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{7.05} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{0.694} & \multicolumn{1}{c|}{0.306} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{9.18} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0.155} & \multicolumn{1}{c|}{0.845} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{25.35} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{0.145} & \multicolumn{1}{c|}{0.855} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{25.65} \\ 
\hline
\end{tabular}
\end{table}
As can be seen from the table, under both 1 month and 5 months the expected
total impacts incurred by option 1 is smaller than that of option 2, thus, option
1 is preferable than option 2.
% \begin{enumerate}
\subsection{Problem E}
% \end{enumerate}

You are evaluating two offers for replacement of an existing bridge on a
highway. You have determined that the link to which the existing bridge belongs
has a reliability of 0.9 over the time period of interest, if it is assumed that
the bridge is perfectly reliable. The consequences of not being able to use the
link due to failure of the bridge are 100 million CHF, principally due to lost
travel time and economic impact.

\begin{itemize}
	\item Offer A, for 15 million CHF, is a wide bridge that can accommodate two lanes of
traffic in both directions and has reliability over the time period of interest
of 0.95.
	\item Offer B, for 19 million CHF, is for two narrow bridges that can accommodate two
lanes of traffic each, and each have a reliability of 0.95 over the same time
period.
\end{itemize}
% \begin{enumerate}
\subsubsection{Question E1}
% \end{enumerate}

If it is assumed that the only differences between the two offers in terms of
costs and benefits are the cost of construction and the cost of not being able to
use the link if failure occurs,

which offer should be chosen?

Back up your argument with exact numbers. In words explain how your decision
would be affected if you discovered that you over-estimated the reliability of
the rest of the link.

% \begin{enumerate}
\subsubsection{Answer E1}
% \end{enumerate}

Option B without the link has a higher reliability than the option A without the
link. Therefore if the reliability of the link is lower than assumed, the failure
cost of both options will be increasing accordingly. Therefore, option B will
remain as the best option. The calculations can be seen in Table \ref{tbl-reliability-a:4a}.
\begin{table}
\caption{Calculations}
\label{tbl-reliability-a:4a}
{\tabcolsep2pt
\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|P{90}{2.2cm}|}{Offers} & \multicolumn{1}{P{90}{2.2cm}|}{Cost} & \multicolumn{1}{P{90}{2.2cm}|}{Object reliability} & \multicolumn{1}{P{90}{2.2cm}|}{Link reliability w/o the object} & \multicolumn{1}{P{90}{2.0cm}|}{Link reliability w object} & \multicolumn{1}{P{90}{2.2cm}|}{Link failure probability} & \multicolumn{1}{P{90}{2.2cm}|}{Costs due to not being able to use the link if failure occurs} & \multicolumn{1}{P{90}{2.2cm}|}{Expected costs due to not being able to use the link} & \multicolumn{1}{P{90}{1.6cm}|}{Differential cost} & \multicolumn{1}{P{90}{2.2cm}|}{Differential benefit} & \multicolumn{1}{P{90}{2.2cm}|}{ Net benefit} \\ 
\hline
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{0.95} & \multicolumn{1}{c|}{0.9} & \multicolumn{1}{c|}{0.855} & \multicolumn{1}{c|}{0.145} & \multicolumn{1}{c|}{100} & \multicolumn{1}{c|}{14.500} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{B} & \multicolumn{1}{c|}{19} & \multicolumn{1}{c|}{0.95} & \multicolumn{1}{c|}{0.9} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{100} & \multicolumn{1}{c|}{10.225} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{4.275} & \multicolumn{1}{c|}{0.275} \\ 
\cline{3-3}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{0.95} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\end{tabular}
}
\end{table}
\subsection{Problem F}
Water is one of the fundamental needs of millions of people living in a
megacity. Water of sufficient quality and quantity must be provided around the
clock. In order to fulfill this need, a city depends on its water distribution
infrastructure, which includes pipes made of different materials and laid at
different times. These pipes are affected by processes of different types and
deteriorate at different rates. The consequences related to pipe failure vary
significantly depending on the type of failure, e.g. a pipe break, or a leaking
pipe, as does the reaction time required to fix the pipe. For example, if a pipe
breaks, a corrective intervention must be executed immediately, and if it is
noticed that there is progressive water loss over time then a preventive
intervention can be planned before there is an inadequate level of service.

Part of the water distribution network in mega-city Q is shown in Figure
\ref{fig24}. The pipe characteristics are given in Table \ref{tbl-reliability-a:5}.
% \begin{figure}[h]
% % \begin{center}
% \includegraphics[width=292pt]{reliability-a-6.eps}
% \caption{Simplified network of water supply pipes}\label{reliability-a-6}
% % \end{center}
% \end{figure}

\begin{table}
\caption{Pipe and their attributes}
\label{tbl-reliability-a:5}
{\tabcolsep2pt
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Pipe} & \multicolumn{1}{c|}{Material} & \multicolumn{1}{c|}{Demand} & \multicolumn{1}{c|}{Length} & \multicolumn{1}{c|}{Year of construction} & \multicolumn{1}{c|}{Usage} & \multicolumn{1}{c|}{Distribution of PS in year 2014} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{($m^3/day$} & \multicolumn{1}{c|}{($m$)} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{M} & \multicolumn{1}{c|}{Concrete} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{5'000} & \multicolumn{1}{c|}{1998} & \multicolumn{1}{c|}{Distribution only} & \multicolumn{1}{c|}{[0.9; 0.1; 0; 0; 0]} \\ 
\hline
\multicolumn{1}{|c|}{A.1} & \multicolumn{1}{c|}{PVC} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & \multicolumn{1}{c|}{Distribution only} & \multicolumn{1}{c|}{} \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.2} & \multicolumn{1}{c|}{PVC} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & \multicolumn{1}{c|}{Distribution only} & \multicolumn{1}{c|}{} \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.3} & \multicolumn{1}{c|}{PVC} & \multicolumn{1}{c|}{250'000} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & \multicolumn{1}{c|}{Distribution and connection to buildings} & \multicolumn{1}{c|}{[0.85; 0.1; 0.05; 0; 0]} \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.4} & \multicolumn{1}{c|}{PVC} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & \multicolumn{1}{c|}{Distribution only} & \multicolumn{1}{c|}{} \\ 
\cline{1-2}\cline{4-6}
\multicolumn{1}{|c|}{A.5} & \multicolumn{1}{c|}{PVC} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{600} & \multicolumn{1}{c|}{2003} & \multicolumn{1}{c|}{Distribution and connection to buildings} & \multicolumn{1}{c|}{} \\ 
\hline
\multicolumn{1}{|c|}{B} & \multicolumn{1}{c|}{Cast iron type 1} & \multicolumn{1}{c|}{200'000} & \multicolumn{1}{c|}{2'500} & \multicolumn{1}{c|}{1993} & \multicolumn{1}{c|}{Distribution and connection to buildings} & \multicolumn{1}{c|}{[0.8; 0.2; 0; 0; 0]} \\ 
\hline
\multicolumn{1}{|c|}{C} & \multicolumn{1}{c|}{Cast iron type 2} & \multicolumn{1}{c|}{100'000} & \multicolumn{1}{c|}{1'600} & \multicolumn{1}{c|}{1993} & \multicolumn{1}{c|}{Distribution and connection to buildings} & \multicolumn{1}{c|}{[0.83; 0.1; 0.07; 0; 0]} \\ 
\hline
\multicolumn{1}{|c|}{D} & \multicolumn{1}{c|}{Cast iron type 3} & \multicolumn{1}{c|}{350'000} & \multicolumn{1}{c|}{4'000} & \multicolumn{1}{c|}{1983} & \multicolumn{1}{c|}{Distribution and connection to buildings} & \multicolumn{1}{c|}{[0.8; 0.1; 0.1; 0; 0]} \\ 
\hline
\end{tabular}
}
\end{table}

When newly built the pipes to not leak but over time they do. The amount of
water lost due to these leaks increases over time. In order to model the increase
in leakage over time the pipes the performance states in Table \ref{tbl-reliability-a:6} are
used. The one year transition probabilities for pipe M and A are given in Table
\ref{tbl-reliability-a:7}.

\begin{table}%[h]
\caption{Performance states used to model water leakage over time}

\vspace{3pt} \noindent
\begin{tabular}{|p{95pt}|p{111pt}|}
\hline
\parbox{95pt}{\centering 
Performance

state
} & \parbox{111pt}{\centering 
Water loss
\\
(\%)
} \\
\hline
\parbox{95pt}{\centering 
1
} & \parbox{111pt}{\centering 
W $<$ 5\%
} \\
\hline
\parbox{95pt}{\centering 
2
} & \parbox{111pt}{\centering 
5\% $<$= W $<$ 10\%
} \\
\hline
\parbox{95pt}{\centering 
3
} & \parbox{111pt}{\centering 
10\% $<$= W $<$ 15\%
} \\
\hline
\parbox{95pt}{\centering 
4
} & \parbox{111pt}{\centering 
15\% $<$= W $<$ 20\%
} \\
\hline
\parbox{95pt}{\centering 
5
} & \parbox{111pt}{\centering 
20\% $<$= W
} \\
\hline
\end{tabular}
\vspace{2pt}
\label{tbl-reliability-a:6}\end{table}

\begin{table}[h]
\caption{One year transition probabilities for pipe M and A}
\label{tbl-reliability-a:7}
{\tabcolsep2pt
\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Pipe} & \multicolumn{5}{c|}{M} & \multicolumn{5}{c|}{A} \\ 
\cline{2-11}
\multicolumn{1}{|c|}{PS} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{5} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{0.93} & \multicolumn{1}{c|}{0.07} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.91} & \multicolumn{1}{c|}{0.09} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.87} & \multicolumn{1}{c|}{0.13} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.88} & \multicolumn{1}{c|}{0.12} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.85} & \multicolumn{1}{c|}{0.15} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.85} & \multicolumn{1}{c|}{0.15} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.8} & \multicolumn{1}{c|}{0.2} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0.82} & \multicolumn{1}{c|}{0.18} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{1} \\ 
\hline
\end{tabular}
}
\end{table}

In addition to the pipes leaking they sometimes break. The processes that lead
to these breaks are not the same as the ones that cause leakage. The reliability
of the pipes affected by these processes are modeled with the Weibull function
 with the values of the parameters given in Table \ref{tbl-reliability-a:8}.

\begin{table}%[h]
\caption{Deterioration parameters and year of construction and intervention information}
\label{tbl-reliability-a:8}
{\tabcolsep2pt
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Zone} & \multicolumn{1}{c|}{Section} & \multicolumn{1}{c|}{Scale parameter} & \multicolumn{1}{c|}{Shape parameter} & \multicolumn{1}{c|}{Duration of corrective intervention} & \multicolumn{1}{c|}{Cost of corrective intervention} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$\alpha$} & \multicolumn{1}{c|}{$m$} & \multicolumn{1}{c|}{(days)} & \multicolumn{1}{c|}{(mus)} \\ 
\hline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.1} & \multicolumn{1}{c|}{0.005} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{5} & \multicolumn{1}{c|}{8'000} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.2} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.4} & \multicolumn{1}{c|}{2} & \multicolumn{1}{c|}{12'000} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{A.3} & \multicolumn{1}{c|}{0.003} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{8} & \multicolumn{1}{c|}{10'000} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.4} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{6} & \multicolumn{1}{c|}{20'000} \\ 
\cline{2-6}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.5} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{4} & \multicolumn{1}{c|}{15'000} \\ 
\hline
\multicolumn{1}{|c|}{B} & \multicolumn{1}{c|}{B} & \multicolumn{1}{c|}{0.003} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{9} & \multicolumn{1}{c|}{45'000} \\ 
\hline
\multicolumn{1}{|c|}{C} & \multicolumn{1}{c|}{C} & \multicolumn{1}{c|}{0.001} & \multicolumn{1}{c|}{1.7} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{60'000} \\ 
\hline
\multicolumn{1}{|c|}{D} & \multicolumn{1}{c|}{D} & \multicolumn{1}{c|}{0.002} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{9} & \multicolumn{1}{c|}{50'000} \\ 
\hline
\multicolumn{1}{|c|}{M} & \multicolumn{1}{c|}{M} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.2} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{70'000} \\ 
\hline
\end{tabular}
}
\end{table}

The types of intervention considered possible for the pipes, as well as their
unit costs, are given in Table \ref{tbl-reliability-a:9}. If a pipe breaks, it will be replaced
for a cost of 500'000 CHF/100 m. The types of inspection that can be included in
monitoring strategies are given in Table \ref{tbl-reliability-a:10}.

\begin{table}[h]
\caption{Interventions}
\begin{tabular}{|l|l|l|}
\hline
\multicolumn{1}{|c|}{Performance} & \multicolumn{1}{c|}{Intervention} & \multicolumn{1}{c|}{Unit cost} \\ 
\multicolumn{1}{|c|}{state} & \multicolumn{1}{c|}{type} & \multicolumn{1}{c|}{$10^3$ mus/100m} \\ 
\hline
\multicolumn{1}{|c|}{1,2,3,4} & \multicolumn{1}{c|}{Do nothing} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{Rehabilitation} & \multicolumn{1}{c|}{25} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{Rehabilitation} & \multicolumn{1}{c|}{30} \\ 
\hline
\multicolumn{1}{|c|}{3,4,5} & \multicolumn{1}{c|}{Replacement} & \multicolumn{1}{c|}{100} \\ 
\hline
\multicolumn{1}{|c|}{1,2,3,4,5} & \multicolumn{1}{c|}{Replacement} & \multicolumn{1}{c|}{500} \\ 
\hline
\end{tabular}
\label{tbl-reliability-a:9}
\end{table}

\begin{table}[h]
\caption{Inspection}
\begin{tabular}{|l|l|}
\hline
\multicolumn{1}{|c|}{Intervention} & \multicolumn{1}{c|}{Unit cost} \\ 
\multicolumn{1}{|c|}{type} & \multicolumn{1}{c|}{$10^3$ mus/100m} \\ 
\hline
\multicolumn{1}{|c|}{Ultra-sonic} & \multicolumn{1}{c|}{3} \\ 
\hline
\multicolumn{1}{|c|}{Acoustic emission} & \multicolumn{1}{c|}{2} \\ 
\hline
\multicolumn{1}{|c|}{Wireless sensor} & \multicolumn{1}{c|}{4} \\ 
\hline
\multicolumn{1}{|c|}{Camera} & \multicolumn{1}{c|}{5} \\ 
\hline
\end{tabular}
\label{tbl-reliability-a:10}
\end{table}
\subsubsection{Question F1}
% \newpage

Estimate the \textit{failure rate} for performance states 1-4 of pipe M

\subsubsection{Answer F1}
%

The failure rate for each performance state is estimated from the reliability of
that state. The reliability of each state equals to the survival probability of
that state. The value of the reliability of each state is therefore equivalent to
the transition probability that the pipe will remains in that state. Values of
transition probability in Table \ref{tbl-reliability-a:7} will be used for computing the
failure rate of each performance state.

Using equation \eqref{eq-reliability-a22}, following results are shown.

\[
 {\theta _1} =  - \frac{{\ln [{R_1}(t = 1)]}}{{t = 1}} =  - \frac{{\ln
[0.93]}}{{t = 1}} \approx 0.073
\]
\[
 {\theta _2} =  - \frac{{\ln [{R_2}(t = 1)]}}{{t = 1}} =  - \frac{{\ln
[0.87]}}{{t = 1}} \approx 0.139
\]
\[
 {\theta _3} =  - \frac{{\ln [{R_3}(t = 1)]}}{{t = 1}} =  - \frac{{\ln
[0.85]}}{{t = 1}} \approx 0.162
\]
\[
 {\theta _4} =  - \frac{{\ln [{R_4}(t = 1)]}}{{t = 1}} =  - \frac{{\ln
[0.8]}}{{t = 1}} \approx 0.223
\]
\subsubsection{Question F2}
Estimate the expected amount of time that pipe M remains in each performance
state
\subsubsection{Answer F2}
The amount of time that pipe M remains in each performance state can be directly
estimated from the reliability function, that is
\begin{eqnarray}
&& {\Theta _i} = \int\limits_0^\infty  {{{\tilde F}_i}(t)dt}  =
\int\limits_0^\infty  {\exp ( - {\theta _i} \cdot t)dt}  = \frac{1}{{{\theta
_i}}}
\label{eq-reliability-a38}
%(38)
\end{eqnarray}
In the case of exponential distribution, the amount of time is the invert value
of the failure rate

\[
 \Theta _1 = \frac{1}{{{\theta _1}}} = \frac{1}{{0.073}} = 13.69 {\rm{ 
years}}
\]
\[
 {\Theta _2} = \frac{1}{{{\theta _2}}} = \frac{1}{{0.139}} = 7.19{\rm{  years}}
\]
\[
 {\Theta _3} = \frac{1}{{{\theta _3}}} = \frac{1}{{0.162}} = 6.17{\rm{  years}}
\]
\[
 {\Theta _4} = \frac{1}{{{\theta _4}}} = \frac{1}{{0.223}} = 4.48{\rm{  years}}
\]

\subsubsection{Question F3}
What is the probability that there will be an adequate level of service in Zone
A in year 2015 (PS4, PS5 = inadequate level of service)? The suspected amount of
water loss in 2014 is given with the following probabilities (PS1 = 0.85, PS2 =
0.10, PS3 = 0.05, PS4 = 0.00; PS5 = 0.00).

\subsubsection{Answer F3}
Using the following equation (in Chapter describing Markov model)

\begin{eqnarray}
&& \overrightarrow {{x^{t + 1}}}  = \left[ {\begin{array}{*{20}{c}}
{{x_1}}&{{x_2}}&{{x_3}}&{{x_4}}&{{x_5}}
\end{array}} \right] \cdot \left[ {\begin{array}{*{20}{c}}
{{p_{11}}}&{{p_{12}}}&{{p_{13}}}&{{p_{14}}}&{{p_{15}}}\\
0&{{p_{22}}}&{{p_{23}}}&{{p_{24}}}&{{p_{25}}}\\
0&0&{{p_{33}}}&{{p_{34}}}&{{p_{35}}}\\
0&0&0&{{p_{44}}}&{{p_{45}}}\\
0&0&0&0&{{p_{55}}}
\end{array}} \right]
\label{eq-reliability-a39}
%(39)
\end{eqnarray}
to estimate the state probability in year 2015
%
\begin{eqnarray}
\begin{array}{l}
\overrightarrow {{x^{t + 1}}}  = \left[ {\begin{array}{*{20}{c}}
{0.85}&{0.1}&{0.05}&0&0
\end{array}} \right] \cdot \left[ {\begin{array}{*{20}{c}}
{0.91}&{0.09}&0&0&0\\
0&{0.88}&{0.12}&0&0\\
0&0&{0.85}&{0.15}&0\\
0&0&0&{0.82}&{0.18}\\
0&0&0&0&1
\end{array}} \right]\\
\;\;\;\;\;\\
\;\;\;\;\; = \left[ {\begin{array}{*{20}{c}}
{0.7735}&{0.1645}&{0.0545}&{0.0075}&0
\end{array}} \right]
\end{array}
\label{eq-reliability-a40}
%(40)
\end{eqnarray}
The probability that pipe in zone A will be in an adequate level of service is
the sum of the state probability from CS 1 to CS 3, that is
\[
\Pr {o^{2015}}\left[ {{\text{Zone A in adequate level of service}}} \right] =
\sum\limits_{i = 1}^3 {x_i^{2015} = 0.7735 + 0.1645 + 0.0545 = 0.9925}
\]
\subsubsection{Question F4}
Calculate the probability of there being at least one pipe break in zone A in
year 2020 (5 points).
\subsubsection{Answer F4}

\underline{Step 1:} Calculate the reliability for each pipe A1, A2, A3, A4, A5 in Zone A. The failure probability of each pipe is independent from each other.

Example is done for pipe A1
\[
{R^{A1}}(t = 22) = {e^{ - {{(0.005 \cdot 22)}^{1.3}}}} = 0.945
\]
the failure probability is
\[
{F^{A1}}(t = 22) = 1 - {R^{A1}}(t = 22) = 1 - 0.945 = 0.055
\]
\underline{Step 2:} Since the break of pipe is not dependent, it is not joint
probability and thus, it is simple the average failure probability, which equals
to 0.025.

The calculations are shown in Table \ref{tbl-reliability-a:11}.
\begin{table}[h]
\caption{Calculations (2020)}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Zone} & \multicolumn{1}{c|}{Section} & \multicolumn{1}{c|}{Scale parameter} & \multicolumn{1}{c|}{Shape parameter} & \multicolumn{1}{c|}{Elapsed time till 2013} & \multicolumn{1}{c|}{Elapsed time till 2020} & \multicolumn{1}{c|}{Reliability} & \multicolumn{1}{c|}{Failure probability} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$\alpha$} & \multicolumn{1}{c|}{$m$} & \multicolumn{1}{c|}{(years)} & \multicolumn{1}{c|}{(years)} & \multicolumn{1}{c|}{($R(t)$)} & \multicolumn{1}{c|}{($F(t)$)} \\ 
\hline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.1} & \multicolumn{1}{c|}{0.005} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{22} & \multicolumn{1}{c|}{0.945} & \multicolumn{1}{c|}{0.055} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.2} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.4} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{0.977} & \multicolumn{1}{c|}{0.023} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{A.3} & \multicolumn{1}{c|}{0.003} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{0.988} & \multicolumn{1}{c|}{0.012} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.4} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{0.982} & \multicolumn{1}{c|}{0.018} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.5} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{0.982} & \multicolumn{1}{c|}{0.018} \\ 
\hline
\end{tabular}
\label{tbl-reliability-a:11}
\end{table}

\subsubsection{Question F5}
When is it expected that the annual probability of pipe A1 breaking will exceed
0.3 given that it did not break earlier?

\subsubsection{Answer F5}
Value of time t can be estimated from from equation \eqref{eqreliability:10}
\[
\begin{array}{l}
R(t) = {e^{ - {{(\alpha  \cdot t)}^m}}} \Leftrightarrow \ln \left[ {R(t)}
\right] = \ln \left[ {{e^{ - {{(\alpha  \cdot t)}^m}}}} \right] =  - {(\alpha 
\cdot t)^m}\\
\to {t^m} =  - \frac{{\ln \left[ {R(t)} \right]}}{{{\alpha ^m}}} =  -
\frac{{\ln \left[ {0.945} \right]}}{{{{0.005}^{1.3}}}} = 55.45 \approx 56{\rm{
years}}
\end{array}
\]
\subsubsection{Question F6}
Calculate the reliability of being able to provide an adequate level of service
to zone A in year 2014 (Keep in mind that water can flow in both directions
through the pipes).
\subsubsection{Answer F6}
\underline{Step 1:} Calculate the reliability of each pipe in Zone A using given
reliability function. This is exactly the same calculation as shown in subsection
6.8, except the elapsed time till 2014 for each pipe is different. Table
\ref{tbl-reliability-a:12} shows the results of estimation for reliability and failure in year
2014.

\begin{table}[h]
\caption{Calculations (2014)}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Zone} & \multicolumn{1}{c|}{Section} & \multicolumn{1}{c|}{Scale parameter} & \multicolumn{1}{c|}{Shape parameter} & \multicolumn{1}{c|}{Elapsed time till 2013} & \multicolumn{1}{c|}{Elapsed time till 2020} & \multicolumn{1}{c|}{Reliability} & \multicolumn{1}{c|}{Failure probability} \\ 
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & \multicolumn{1}{c|}{$\alpha$} & \multicolumn{1}{c|}{$m$} & \multicolumn{1}{c|}{(years)} & \multicolumn{1}{c|}{(years)} & \multicolumn{1}{c|}{($R(t)$)} & \multicolumn{1}{c|}{($F(t)$)} \\ 
\hline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.1} & \multicolumn{1}{c|}{0.005} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{0.963} & \multicolumn{1}{c|}{0.037} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.2} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.4} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{0.987} & \multicolumn{1}{c|}{0.013} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{A} & \multicolumn{1}{c|}{A.3} & \multicolumn{1}{c|}{0.003} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{0.994} & \multicolumn{1}{c|}{0.006} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.4} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.5} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{0.991} & \multicolumn{1}{c|}{0.009} \\ 
\cline{2-8}
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{A.5} & \multicolumn{1}{c|}{0.004} & \multicolumn{1}{c|}{1.3} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{0.991} & \multicolumn{1}{c|}{0.009} \\ 
\hline
\end{tabular}
\label{tbl-reliability-a:12}
\end{table}
\underline{Step 2:} if for example pipe break at A.2 or A.4, the system is still
functioning because the water can flow into both direction. And even A.3 break,
it can still provide adequate level of service. In that situation, the system
turns to be like a series network with independent reliability.

3 scenarios can be seen

\begin{itemize}
	\item Scenario 1: If A.1 breaks, it affects the entire zone A
	\item Scenario 2: If A.2 breaks, it does not, because water can flow into A4
	\item Scenario 3: If A.4 breaks, it does not, because water can flow in to A2
\end{itemize}
Following is the calculation for reliability of each scenario
\[
 {R^{scenario1}}(2014) = {R^{A1}}(2014) = 0.963
\]
\[
 {R^{scenario2}}(2014) = {R^{A1}}(2014) \cdot {R^{A3}}(2014) \cdot
{R^{A4}}(2014) \cdot {R^{A5}}(2014) = 0.940
\]
\[
 {R^{scenario3}}(2014) = {R^{A1}}(2014) \cdot {R^{A2}}(2014) \cdot
{R^{A3}}(2014) \cdot {R^{A5}}(2014) = 0.937
\]
As each scenario can occurs independently from other scenarios, the average
reliability is then evaluated as the average reliability of the three scenario
\[
\begin{array}{l}
{R^{{\rm{Adequate level of service}}}}(2014) = \frac{{{R^{{\rm{scenario
1}}}}(2014) + {R^{{\rm{scenario 2}}}}(2014) + {R^{{\rm{scenario
3}}}}(2014)}}{3}\\
{\rm{                                         =  }}\frac{{0.963 + 0.940 +
0.937}}{3} = 0.947
\end{array}
\]
\subsection{Problem G}
The details of the number of fan units used in road tunnels retired at different
ages are given in Table \ref{tbl-reliability-a:13}.
\begin{table}[h]
\caption{Record of fan units failed}
\begin{tabular}{|l|l|l|l|l|l|l|l|l|}
\hline
\multicolumn{1}{|c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} & \multicolumn{1}{c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} & \multicolumn{1}{c|}{Item} & \multicolumn{1}{c|}{time} & \multicolumn{1}{c|}{status} \\ 
\hline
\multicolumn{1}{|c|}{1} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{25} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{2} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{26} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{3} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{27} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{4} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{28} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{5} & \multicolumn{1}{c|}{18} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{29} & \multicolumn{1}{c|}{19} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{6} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{18} & \multicolumn{1}{c|}{18} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{30} & \multicolumn{1}{c|}{11} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{7} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{19} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{31} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{8} & \multicolumn{1}{c|}{19} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{17} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{32} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{9} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{21} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{33} & \multicolumn{1}{c|}{10} & \multicolumn{1}{c|}{0} \\ 
\hline
\multicolumn{1}{|c|}{10} & \multicolumn{1}{c|}{13} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{22} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{34} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{11} & \multicolumn{1}{c|}{20} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{23} & \multicolumn{1}{c|}{12} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{35} & \multicolumn{1}{c|}{14} & \multicolumn{1}{c|}{1} \\ 
\hline
\multicolumn{1}{|c|}{12} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{0} & \multicolumn{1}{c|}{24} & \multicolumn{1}{c|}{15} & \multicolumn{1}{c|}{1} & \multicolumn{1}{c|}{36} & \multicolumn{1}{c|}{16} & \multicolumn{1}{c|}{0} \\ 
\hline
\end{tabular}
\label{tbl-reliability-a:13}
\end{table}
Value in the column ``status'' has binary value 1 and 0. Value 1 and 0 represent
the ``fail'' or ``not fail'' of the tunnel fans at the inspection time.
\subsubsection{Question G1}
What is the average expected life of the units that have survived until year 4?
\subsubsection{Answer G1}
The observed data infers a case that the failure rate is not a constant value.
The failure rate of the fan is a time-varying variable. The failure rate can be
followed a Weibull function (Equation \eqref{eqreliability:12}).

The average life of the units that have survived until a certain year will be
defined as
\begin{eqnarray}
&& \Phi \left[ {\tau  \ge t} \right] = \left[ {\int\limits_t^\infty  {R(\tau )d\tau
} } \right] = \left[ {\int\limits_t^\infty  {{e^{ - {{(\alpha  \cdot \tau
)}^m}}}d\tau } } \right]
\label{eq-reliability-a44}
%(44)
\end{eqnarray}
In this example, t=4 years. The unknown parameters are value of $\alpha $ and m.
Values of these parameters can be estimated with the Maximum likelihood
estimation approach (MLE) (see the R code in the Appendix).

Using the MLE, value of $\alpha $ and m are 0.0542 and 5.8582, respectively. The
average life of the units that have survived after 4 years will be
\[
\Phi \left[ {\tau  \ge 4} \right] = \left[ {\int\limits_4^\infty  {{e^{ -
{{(0.0542 \cdot \tau )}^{5.8584}}}}d\tau } } \right] \approx 13{\rm{ years}}
\]
\bibliographystyle{plainnat}
\bibliography{reference}

\pagebreak

\begin{subappendices}
% \label{appendix3}
\section{Reliability of an item with time invariant failure rate}\label{appen31}
% \textbf{Reliability of an item with time invariant failure rate}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-reliability/IMP-example1-time-invarying.r}
\pagebreak
\section{Reliability of an item with time-variant failure rate}\label{appen32}
% \textbf{Reliability of an item with time-variant failure rate}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-reliability/IMP-example2-time-varying.r}
\pagebreak
\section{Maximum likelihood estimation for Weibull model}\label{appen33}
% \textbf{Reliability of an item with time-variant failure rate}
\lstinputlisting[basicstyle=\ttfamily\scriptsize]{./Programs/IMP-reliability/IMP-A-time-varying-fan.r}
\end{subappendices}

### 2019-05-05 InfrastructureMaintenance/General/acronym.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%%%acronym.tex%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% sample list of acronyms
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%

\extrachap{Acronyms}

Use the template \emph{acronym.tex} together with the Springer document class SVMono (monograph-type books) or SVMult (edited books) to style your list(s) of abbreviations or symbols in the Springer layout.

Lists of abbreviations\index{acronyms, list of}, symbols\index{symbols, list of} and the like are easily formatted with the help of the Springer-enhanced \verb|description| environment.

### 2019-05-05 InfrastructureMaintenance/General/foreword.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%%%foreword.tex%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\foreword
This document has been compiled to provide students taking the above mentioned course in understanding the examples given in class and assignments. All questions presented have been prepared by the authors based to varying degrees on different sources. The sources are given for each example.

This document is being issued so that students taking the course may better prepare for the exam. As it has not yet undergone the complete rigorous control processes of IBI this document is to be considered a work in progress and is hence labeled as a DRAFT.

\vspace{\baselineskip}
\begin{flushright}\noindent
IBI-ETH, September 2015\hfill {\it Professor: Bryan T. Adey}\\
\hfill {\it Dr: Nam Lethanh}\\
\end{flushright}

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-reliability/IMP-A-time-varying.R
Add files via upload

@@ -1,0 +1,36 @@
+#This program is coded by Nam Lethanh for use in the class IMP-HS2014
+#Purpose: To calculate reliability from Cox Model
+
+density<-function(t,theta,m){theta*m*(theta*t)^(m-1)*exp(-(theta*t)^m)}
+rate<-function(t,theta,m){theta*m*(theta*t)^(m-1)}
+reliability<-function(t,theta,m){exp(-(theta*t)^m)}
+
+theta<-c(0.005,0.004,0.003,0.004,0.004)
+m<-c(1.3,1.4,1.5,1.5,1.5)
+elapsedtime20<-c(22,17,17,17,17)
+elapsedtime14<-c(16,11,11,11,11)
+
+No =5 
+R2020<-matrix(double(1),nrow=No,ncol=1)
+for (i in 1:No){
+  R2020[i]<-reliability(elapsedtime20[i],theta[i],m[i])
+}
+cat("Reliability of each pipe in Zone A in year 2020")
+print(R2020)
+
+mean(1-R2020)
+
+
+R2014<-matrix(double(1),nrow=No,ncol=1)
+for (i in 1:No){
+  R2014[i]<-reliability(elapsedtime14[i],theta[i],m[i])
+}
+cat("Reliability of each pipe in Zone A in year 2014")
+print(R2014)
+
+
+
+
+cat("THE END")
+
+

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-reliability/IMP-example1-time-invarying.r

@@ -1,0 +1,93 @@
+#This program is coded by Nam Lethanh for use in the class IMP
+#Purpose: To calculate reliability
+TIME=4
+T=seq(0,TIME, by =0.5) #investigate time
+N=3 #Number of item
+lambda<-matrix(double(1),nrow=1,ncol=N)
+lambda<-c(0.5,1,1.5) # value of hazard rate
+# define dimension of matrix
+fail<-matrix(double(1),nrow=length(T),ncol=N)
+survive<-matrix(double(1),nrow=length(T),ncol=N)
+#reliability function for exponential distribution
+failure<-function(t,lambda){lambda*exp(-lambda*t)}
+survival<-function(t,lambda){exp(-lambda*t)}
+for (i in 1:N){
+for (t in (1:length(T))){
+  fail[t,i]<-failure(T[t],lambda[i])
+  survive[t,i]<-survival(T[t],lambda[i])
+}
+}
+cat('PROBABILITY DENSITY \n')
+print(cbind(T,fail))
+cat('RELIABILITY \n')
+print(cbind(T,survive))
+cat('EXPECTED LIFE TIME \n')
+theta=lambda
+print(1/theta)
+#Plotting for reliability
+plot.new()
+par(mar=c(5,4,4,6)+0.3)
+limy=c(0,1)
+limx=c(0,4)
+plot(T,survive[,1],lwd=2,col="red",ylab="",xlab="",xlim=limx,ylim=limy,axes=FALSE,lty=1,type="b",pch=4)
+axis(2,ylim=limy,col="black",las=1)
+axis(1,c(seq(0,max(limx),by=0.5)),c(seq(0,max(limx),by=0.5)))
+mtext(expression(paste('Reliability')),side=2,col="black",line=3)
+mtext(expression(paste('Units of time')),side=1,col="black",line=3)
+grid(10, 10, col = "lightgray", lty = "dotted",lwd = par("lwd"), equilogs = TRUE)
+box()
+par(new=TRUE)
+plot(T,survive[,2],lwd=2,col="blue",ylab="",xlab="",xlim=limx,ylim=limy,axes=FALSE,lty=1,type="b",pch=2)
+par(new=TRUE)
+plot(T,survive[,3],lwd=2,col="cyan4",ylab="",xlab="",xlim=limx,ylim=limy,axes=FALSE,lty=1,type="b",pch=1)
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-reliability/IMP-example2-time-varying.R

@@ -1,0 +1,53 @@
+#This program is coded by Nam Lethanh for use in the class IMP-HS2014
+#Purpose: To calculate reliability from Cox Model
+data=read.csv("example2.csv",header=TRUE)
+attach(data)
+library(survival)
+likelihood=survreg(Surv(time,status)~1,data=data,dist="weibull")
+print(likelihood)
+
+density<-function(t,theta,m){theta*m*(theta*t)^(m-1)*exp(-(theta*t)^m)}
+rate<-function(t,theta,m){theta*m*(theta*t)^(m-1)}
+reliability<-function(t,theta,m){exp(-(theta*t)^m)}
+TIME=100
+T=seq(0,TIME, by =1)
+den<-matrix(double(1),nrow=T,ncol=1)
+hazard<-matrix(double(1),nrow=T,ncol=1)
+reli<-matrix(double(1),nrow=T,ncol=1)
+#theta=0.912
+#m=3.624
+
+theta=1/exp(3.624)
+m=1/0.912
+
+for (t in (1:length(T))){
+  den[t]<-density(T[t],theta,m)
+  hazard[t]<-rate(T[t],theta,m)
+  reli[t]<-reliability(T[t],theta,m)
+}
+results=cbind(den,hazard,reli)
+print(results)
+
+plot.new()
+par(mar=c(5,4,4,6)+0.3)
+limy=c(0,1)
+limx=c(0,100)
+plot(T,reli,lwd=2,col="red",ylab="",xlab="",xlim=limx,ylim=limy,axes=FALSE,lty=1,type="b",pch=4)
+axis(2,ylim=limy,col="black",las=1)
+axis(1,c(seq(0,100,by=10)),c(seq(0,100,by=10)))
+mtext(expression(paste('Reliability')),side=2,col="black",line=3)
+mtext(expression(paste('Units of time')),side=1,col="black",line=3)
+grid(10, 10, col = "lightgray", lty = "dotted",lwd = par("lwd"), equilogs = TRUE)
+box()
+par(new=TRUE)
+limy=c(0,0.05)
+limx=c(0,100)
+plot(T,hazard,lwd=2,col="blue",ylab="",xlab="",xlim=limx,ylim=limy,axes=FALSE,lty=1,type="b",pch=2)
+axis(4,ylim=limy,col="black",las=1)
+mtext(expression(paste('Failure rate')),side=4,col="black",line=3)
+colors=c("red","blue")
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-reliability/IMP-fragility.R

@@ -1,0 +1,36 @@
+# Evolution
+data=read.csv("fragility.csv", header=TRUE)
+attach(data)
+#object 1
+scalefactor=1
+trucy=1
+colors=c("red", "darkorchid","green", "violet", "darkorange3","brown3")
+
+plot(data$CS1/scalefactor,pch=16,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="red",type="o",lwd=4)
+axis(2, ylim=c(0,trucy),col="darkblue",las=1)
+mtext(expression(paste("Probability of failure")),side=2,line=2.2, adj = 0.5 )
+axis(1,pretty(range(data$rockfall),10))
+mtext(expression(paste("Rock fall volume" (m^3))),side=1,col="black",line=2.2)  
+box()
+
+par(new=TRUE)
+plot(data$CS2/scalefactor,pch=15,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="darkorchid",type="o",lwd=4)
+par(new=TRUE)
+plot(data$CS3/scalefactor,pch=14,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="green",type="o",lwd=4)
+grid(10, 10, col = "lightgray", lty = "dotted",
+     lwd = par("lwd"), equilogs = TRUE)
+
+legend("topleft",inset=.02,legend=c("i=1","i=2","i=3"),text.col=colors,pch=c(16,15,14),col=colors,horiz=F,bty != "n",bg="white")
+
+
+text(15, 18.5,expression(paste("I3-GCS1 (14.73x",10^6, "mu-year 15)")),pos=1,cex = 0.8, srt = 0)
+text(15, data$CS1[15],"A (0.089)",pos=4,cex = 0.8, srt = 0)
+text(15, data$CS2[15],"B (0.372)",pos=4,cex = 0.8, srt = 0)
+text(15, data$CS3[15],"C (0.716)",pos=4,cex = 0.8, srt = 0)
+points(15,data$CS1[15],pch=23,bg="black",lwd=3)
+points(15,data$CS2[15],pch=23,bg="black",lwd=3)
+points(15,data$CS3[15],pch=23,bg="black",lwd=3)
+segments(15, 0, 15, data$CS3[15], col= 'cadetblue',lty=2,lwd=0.5)
[…]

### 2019-05-05 InfrastructureMaintenance/Programs/IMP-reliability/IMP-reliability.R

@@ -1,0 +1,20 @@
+#
+R=c(1-1/6,1-1/6,1-1/6,1-1/6,1-1/6,1-1/6)
+R_network_1=R[1]*R[6]*(1-(1-R[2])*(1-R[5])*(1-(1-(1-R[3])*(1-R[4]))))
+print(R_network_1)
+hazardrate=-log(R_network_1)
+cat("hazard rate for system")
+print(hazardrate)
+cat("Reliability for 5 month")
+R=exp(-hazardrate*5)
+print(R)
+cat("hazard rate for one item")
+hazardrate=-log(1-1/6)
+print(hazardrate)
+cat("Reliability for 1 item in 5 month")
+R=exp(-hazardrate*5)
+print(R)
+cat("Reliability for system in 5 month")
+R=c(R,R,R,R,R,R)
+R_network_5=R[1]*R[6]*(1-(1-R[2])*(1-R[5])*(1-(1-(1-R[3])*(1-R[4]))))
+print(R_network_5)

### 2023-03-02 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.   
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

\documentclass[utf8]{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass[utf8]{frontiersHLTH} % for Health articles
%\documentclass[utf8]{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\usepackage[utf8]{vietnam}
\usepackage{hyperref}
%\linenumbers

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{22-03-2023} %use et al only if is more than 1 author
\def\Authors{Lê Thanh Nam\,$^{1,*, 2, 3, 4}$}
%\def\Authors{Lê Thanh Nam\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^{1,2}$}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Tiến Sĩ, Kĩ Sư Xây Dựng: Viện Đô Thị Thông Minh và Quản Lý (ISCM), Trường Công Nghệ và Thiết Kế, Đại Học UEH, tp. Hồ Chí Minh -  \href{https://www.iscm.ueh.edu.vn/}{www.iscm.ueh.edu.vn} \\
$^{2}$Giám Đốc Kĩ Thuật, Chuyên Gia Soạn Hợp Đồng (FIDIC): Công Ty Tư Vấn ARCADIS, phụ trách thị trường Đông Nam Á và Ấn Độ -  \href{www.arcadis.com}{www.arcadis.com}\\
$^{3}$Giám Đốc: Công ty TNHH ASQ Việt Nam - \href{https://www.asq.vn}{www.asq.vn}\\
$^{4}$Cố Vấn Kĩ Thuật: Công ty EMAPTA - \href{www.emapta.com}{www.emapta.com}}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Di Động: +84-9-8378-0100 (Việt Nam) / +91-98-1044-4723 (Ấn Độ) Email: namlt@ueh.edu.vn}

%\def\corrEmail{Email: namlt@ueh.edu.vn}

\begin{document}
\onecolumn
\firstpage{1}

\title[Dữ Liệu Lớn/Quản Lý Hạ Tầng - Điện Lực]{Dữ Liệu Lớn và Quản Lý Hạ Tầng Trong Ngành Điện Lực}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}
Ngành điện lực là một trong những ngành xương sống cho sự phát triển kinh tế của quốc gia. Trong những năm gần đây, các công ty điện lực đã chứng kiến và trải nghiệm nhiều thử thách và sự đổi mới trong cả kĩ thuật và quản lý, và phần nào đã có những bước tiến đáng kể trong công tác chuyển đổi số (digitization) cùng với khái niệm về hệ thống truyền tải điện thông minh(Smart Grids). Các hệ thống điện thường luôn hoạt động trong tình trạng áp lực, chủ yếu là do sự gia tăng hàng năm trong nhu cầu sử dụng điện, sự thiếu hụt về nguồn cung về nhiên liệu/năng lượng để sản xuất ra điện, và các ràng buộc về môi trường áp đặt lên hệ thống sản xuất điện và sự mở rộng các đường dây tải điện.

Ở Việt Nam, tổng công ty điện lực Việt Nam (EVN) cùng các bộ ngành liên quan đã soạn thảo và đề xuất "Qui Hoạch Điện VIII" hướng tới sự phát triển ổn định và bền vững chiến lược đến năm 2050. Một trong những nội dung quan trọng của Qui Điện VIII có chú trọng đến việc phát triển và nâng cao khả năng ứng dụng các công nghệ tiên tiến và áp dụng dữ liệu lớn cũng như trí thông minh nhân tạo vào công tác xây dựng, điều hành và quản lý các hệ thống điện.

Bài viết này chú trọng đến vài trò của công tác phân tích dữ liệu lớn và vai trò của công tác quản lý cơ sở hạ tầng trong các lĩnh vực quan trọng của nền kinh tế, đặc biệt chú trọng tới ngành điện. Nội dung bài viết đề cập đến việc sử dụng các nguồn cung cấp dữ liệu lớn mà rất khó có thể truy cập vào các hệ thống cơ sở dữ liệu tiêu chuẩn, việc quản lý và quan trắc trong ngành điện. Đảm bảo sự ổn định của hệ thống thông qua việc sử dụng hiệu quả các nguồn dữ liệu lớn là một trong những mục tiêu quan trọng của các công ty điện lực.

\tiny
 \keyFont{ \section{Từ Khóa:} Dữ Liệu Lớn, Quản Lý Tài Sản, Cơ Sở Hạ Tầng, Độ Tin Cậy, Trí Thông Minh Nhân Tạo} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

\section{Giới Thiệu}
%For Original Research Articles \citep{conference}, Clinical Trial Articles \citep{article}, and Technology Reports \citep{patent}, the introduction should be succinct, with no subheadings \citep{book}. For Case Reports the Introduction should include symptoms at presentation \citep{chapter}, physical exams and lab results \citep{dataset}.

\section{Dữ Liệu Lớn trong Nhiều Lĩnh Vực của Nền Kinh Tế}

% For Original Research articles, please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.

\section{Dữ Liệu Lớn Trong Ngành Điện Lực}

\section*{Lời Cám Ơn}
This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.

%\section*{Figure captions}

%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
%%% If using *.tif files convert them to .jpg or .png
%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%

%\begin{figure}[h!]
%\begin{center}
%\includegraphics[width=10cm]{logo1}% This is a *.eps file
%\end{center}
%\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
%\end{figure}

\end{document}

### 2023-03-17 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update

@@
 
-\title[Dữ Liệu Lớn/Quản Lý Hạ Tầng - Điện Lực]{Dữ Liệu Lớn và Quản Lý Hạ Tầng Trong Ngành Điện Lực}
+\title[Ứng Dụng Dữ Liệu Lớn/Quản Lý Hạ Tầng - Điện Lực]{Ứng Dụng Dữ Liệu Lớn và Quản Lý Tài Sản Hạ Tầng Trong Ngành Điện Lực}
 
@@
 
-\section*{Lời Cám Ơn}
-This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.
+\section{Quản Lý Tài Sản Hạ Tầng}
+
+\section{}
+
+%\section*{Lời Cám Ơn}
+%This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.
 

### 2023-03-19 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
commit

@@
 \begin{abstract}
+%Bài viết này tổng hợp 
 Ngành điện lực là một trong những ngành xương sống cho sự phát triển kinh tế của quốc gia.
@@
 
-Bài viết này chú trọng đến vài trò của công tác phân tích dữ liệu lớn và vai trò của công tác quản lý cơ sở hạ tầng trong các lĩnh vực quan trọng của nền kinh tế, đặc biệt chú trọng tới ngành điện.
+Bài viết này cung cấp cho người đọc một cái nhìn tổng quan về các ứng dụng phân tích dữ liệu lớn, công tác quản lý cơ sở hạ tầng trong ngành điện lực, và các vấn đề liên quan đến việc triển khai.
+Chú trọng của bài viết sẽ là các ứng dụng mới đã mang lại giá trị tích cực trong ngành, một số bài học rút ra từ việc triển khai ứng dụng tại một số tổ chức, và một số ý tưởng và chủ đề cần khám phá.
+Tổng quan các nghiên cứu khoa học trong ngành khoa học dữ liệu trong ngành điện sẽ được đề cập nhằm tạo ra được cái nhìn tổng quát cho việc ứng dụng và tiếp tục phát triển của ngành trong tương lai.
+Cuối cùng, bài viết nêu ra các cơ hội và thách thức cũng như các mục tiêu và chiến lược để đạt được các kết quả có ý nghĩa.
+
+%Bài viết này chú trọng đến vài trò của công tác phân tích dữ liệu lớn và vai trò của công tác quản lý cơ sở hạ tầng trong các lĩnh vực quan trọng của nền kinh tế, đặc biệt chú trọng tới ngành điện.
 Nội dung bài viết đề cập đến việc sử dụng các nguồn cung cấp dữ liệu lớn mà rất khó có thể truy cập vào các hệ thống cơ sở dữ liệu tiêu chuẩn, việc quản lý và quan trắc trong ngành điện.
@@
 
+Có rất nhiều định nghĩa và cách hiểu khác nhau về phân tích dữ liệu lớn trong các ứng dụng ngành điện, đặc biệt là công tác sản xuất, truyền tải, và phân phối.
+Điều này chủ yếu là vì lý do có khá nhiều cách tiếp cận trong ngành khoa học dữ liệu và mục đích rộng liên quan (ví dụ như việc sử dụng các công cụ toán xác xác thống kê, trí thông minh nhân tạo (AI), học máy).
+Tính phức tạp và đa dạng của phân tích dữ liệu lớn trong ngành còn đến từ nhiều cách thức ứng dụng trong các lĩnh vực như quản lý tài sản (Asset Management), điều hành, điều khiển hệ thống, an toàn và an ninh, các quyết định lập kế hoạch và thị trường.
+Trong đó, có một vấn đề nền tảng làm cho công tác phân tích dữ liệu lớn trong ngành điện lực đặc biệt khó khăn là liên quan đến dung lượng của dữ liệu.
+Dung lượng của dữ liệu trong ngành thường ở cấp độ Terabyte, lớn hơn rất nhiều cấp độ Petabyte là cấp độ đã được coi là lớn trong các lĩnh vực khác.
+Cuối cùng, khái niệm "phân tích dữ liệu" còn là khá mơ hồ và đôi khi bị hiểu sai bởi vì hầu hết các ứng dựng cơ bản hay ứng dụng truyền thống trong ngành điện là dựa trên việc xử lý các dữ liệu được đo đạc (measurement data).
+Công việc xử lý truyền thống này đôi khi không được coi là phân tích dữ liệu lớn trong ngành khoa học dữ liệu.
+Ví dụ như, các phương pháp truyền thống liên quan đến tính toán ước lượng trạng thái (state estimation) hay vị trí sự cố (outage location) thường không được coi là phân tích dữ liệu lớn nếu công tác phân tích này chỉ đơn thuần dựa vào các phương trình/công thức toán được phát triển trên nền tảng vật lý.
+Trong khi đó, ở một khía cạnh khác, việc phán đoán sự cố dựa trên các mô hình nền tảng dữ liệu sử dụng các dữ liệu về thời tiết và mất điện có độ phân giải cao thì được coi là phân tích dữ liệu lớn.
+
+Bài viết này không chú trọng đến việc đưa ra một định nghĩa rành mạch về phân tích dữ liệu lớn trong ngành điện mà chỉ chú trọng đến việc tổng quan hóa và nêu ra một số nghiên cứu và ứng dụng gần nhất sử dụng các phương pháp tiên tiến trong ngành khoa học dữ liệu trong ngành điện, đặc biệt chú trọng đến công tác quản lý tài sản cơ sở hạ tầng ngành điện, quản lý sự cố, và tính tích hợp với năng lượng tái tạo.
+
+Có thể nói, một trong những bài báo nghiên cứu đầu tiên liên quan đến việc ứng dụng dữ liệu lớn trong ngành điện là bài báo của tác giả .
+Tuy nhiên, nghiên cứu này chỉ dừng lại ở dạng mô hình và lý thuyết mà chưa đưa ra ứng dụng cụ thể vào thực tế.
+Việc ứng dụng vào thực tế mới chỉ được tiến hành trong vài năm gần đây.
+
+Với mục đích hướng tới việc truyền tải kiến thức và đào tạo, ở bài viết này, một số khái niệm cơ bản cấu thành nên công tác phân tích dữ liệu lớn trong ngành điện và các khía cạnh đặc thù của ngành điện sẽ được đề cập.
+
+\section{Tầm Quan Trọng và Tính Khả Thi của Phân Tích Dữ Liệu Lớn trong Ngành Điện}
+\subsection{Sự Tác Động của Phân Tích Dữ Liệu Lớn}
+
+Sự thay đổi mạnh mẽ trong ngành điện dường như là chưa có tiền lệ, bao gồm sự chuyển dịch trong phân bổ các dạng năng lượng (Energy Mix), yêu cầu và nhu cầu ngành càng cao của khách hàng, sự xuất hiện các kĩ thuật tiên tiến và thiết bị mới, và các mô hình kinh doanh mới.
+Chính những thay đổi này dẫn đến tính phức tạp (Complexity) và tính không chắc chắn (Uncertainty), cũng như mang tới các thách thức và cơ hội.
+
+\subsection{Các Nguồn Dữ Liệu Mới}
+
+\subsection{Các Cân Nhắc Quan Trọng Khi Tạo Ra Các Gói Dữ Liệu}
+
+\section{Các Thách Thức của Công Việc Phân Tích Dữ Liệu Lớn}
+\subsection{Các Nền Tảng của Khoa Học Dữ Liệu}
+
+\subsection{Các Khía Cạnh Kĩ Thuật}
+
+\subsection{Khung Đưa ra Quyết Định}
+
+\section{Ứng Dụng}
+\subsection{Quản Lý tài Sản Cơ Sở Hạ Tầng và Mất Điện}
+
+\subsubsection{Phân Loại Tài Sản}
+
+\subsubsection{Các Tác Động của Thời Tiết và Việc Mất Điện}
+
+\subsubsection{Cơ Bản về Quản Lý Mất Điện}
+
+\subsubsection{Ước Đoán Sự Mất Điện và Mạng Lưới Truyền Tải}
+
 \section{Dữ Liệu Lớn trong Nhiều Lĩnh Vực của Nền Kinh Tế}
-
 % For Original Research articles, please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.

### 2023-03-19 2023-EVN-Data-Analytics-for-CEO/paper/frontiers.tex
update the literature review sections

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.   
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

\documentclass[utf8]{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass[utf8]{frontiersHLTH} % for Health articles
%\documentclass[utf8]{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

\linenumbers

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
\def\Authors{First Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^{1,2}$}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X \\
$^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Corresponding Author}

\def\corrEmail{email@uni.edu}

\begin{document}
\onecolumn
\firstpage{1}

\title[Running Title]{Article Title}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle
\begin{abstract}

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\section{}
For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.

As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership. References should not be cited in the abstract. Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.

\tiny
 \keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

\section{Introduction}

For Original Research Articles \citep{conference}, Clinical Trial Articles \citep{article}, and Technology Reports \citep{patent}, the introduction should be succinct, with no subheadings \citep{book}. For Case Reports the Introduction should include symptoms at presentation \citep{chapter}, physical exams and lab results \citep{dataset}.

\section{Article types}

For requirements for a specific article type please refer to the Article Types on any Frontiers journal page. Please also refer to  \href{http://home.frontiersin.org/about/author-guidelines#Sections}{Author Guidelines} for further information on how to organize your manuscript in the required sections or their equivalents for your field

% For Original Research articles, please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.

\section{Manuscript Formatting}

\subsection{Heading Levels}

%There are 5 heading levels

\subsection{Level 2}
\subsubsection{Level 3}
\paragraph{Level 4}
\subparagraph{Level 5}

\subsection{Equations}
Equations should be inserted in editable format from the equation editor.

\begin{equation}
\sum x+ y =Z\label{eq:01}
\end{equation}

\subsection{Figures}
Frontiers requires figures to be submitted individually, in the same order as they are referred to in the manuscript. Figures will then be automatically embedded at the bottom of the submitted manuscript. Kindly ensure that each table and figure is mentioned in the text and in numerical order. Figures must be of sufficient resolution for publication \href{http://home.frontiersin.org/about/author-guidelines#ResolutionRequirements}{see here for examples and minimum requirements}. Figures which are not according to the guidelines will cause substantial delay during the production process. Please see \href{http://home.frontiersin.org/about/author-guidelines#GeneralStyleGuidelinesforFigures}{here} for full figure guidelines. Cite figures with subfigures as figure \ref{fig:2}B.

\subsubsection{Permission to Reuse and Copyright}
Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet). It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.
%%Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet). It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.

\subsection{Tables}
Tables should be inserted at the end of the manuscript. Please build your table directly in LaTeX.Tables provided as jpeg/tiff files will not be accepted. Please note that very large tables (covering several pages) cannot be included in the final PDF for reasons of space. These tables will be published as \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} on the online article page at the time of acceptance. The author will be notified during the typesetting of the final article if this is the case.

\section{Nomenclature}

\subsection{Resource Identification Initiative}
To take part in the Resource Identification Initiative, please use the corresponding catalog number and RRID in your current manuscript. For more information about the project and for steps on how to search for an RRID, please click \href{http://www.frontiersin.org/files/pdf/letter_to_author.pdf}{here}.

\subsection{Life Science Identifiers}
Life Science Identifiers (LSIDs) for ZOOBANK registered names or nomenclatural acts should be listed in the manuscript before the keywords. For more information on LSIDs please see \href{http://www.frontiersin.org/about/AuthorGuidelines#InclusionofZoologicalNomenclature}{Inclusion of Zoological Nomenclature} section of the guidelines.

\section{Additional Requirements}

For additional requirements for specific article types and further information please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines#AdditionalRequirements}{Author Guidelines}.

\section*{Conflict of Interest Statement}
%All financial, commercial or other relationships that might be perceived by the academic community as representing a potential conflict of interest must be disclosed. If no such relationship exists, authors will be asked to confirm the following statement:

The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
%\section*{Author Contributions}

The Author Contributions section is mandatory for all articles, including articles by sole authors. If an appropriate statement is not provided on submission, a standard one will be inserted during the production process. The Author Contributions statement must describe the contributions of individual authors referred to by their initials and, in doing so, all authors agree to be accountable for the content of the work. Please see  \href{http://home.frontiersin.org/about/author-guidelines#AuthorandContributors}{here} for full authorship criteria.

\section*{Funding}
Details of all funding sources should be provided, including grant numbers if applicable. Please ensure to add all necessary funding information, as after publication this is no longer possible.

\section*{Acknowledgments}
This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.

\section*{Supplemental Data}
 \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} should be uploaded separately on submission, if there are Supplementary Figures, please include the caption in the same file as the figure. LaTeX Supplementary Material templates can be found in the Frontiers LaTeX folder.

\section*{Data Availability Statement}
The datasets [GENERATED/ANALYZED] for this study can be found in the [NAME OF REPOSITORY] [LINK].
% Please see the availability of data guidelines for more information, at https://www.frontiersin.org/about/author-guidelines#AvailabilityofData

\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
%\bibliographystyle{frontiersinHLTH&FPHY} % for Health, Physics and Mathematics articles
\bibliography{test}

%%% Make sure to upload the bib file along with the tex file and PDF
%%% Please see the test.bib file for some examples of references

\section*{Figure captions}

%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
%%% If using *.tif files convert them to .jpg or .png
%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{logo1}% This is a *.eps file
\end{center}
\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=15cm]{logos}
\end{center}
\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
\end{figure}

%%% If you are submitting a figure with subfigures please combine these into one image file with part labels integrated.
%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
%%% Frontiers will add the figures at the end of the provisional pdf automatically
%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.

\end{document}

### 2023-03-19 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update the literature review sections

@@
 
-Có thể nói, một trong những bài báo nghiên cứu đầu tiên liên quan đến việc ứng dụng dữ liệu lớn trong ngành điện là bài báo của tác giả .
+Có thể nói, một trong những bài báo nghiên cứu đầu tiên liên quan đến việc ứng dụng dữ liệu lớn trong ngành điện là bài báo của tác giả \cite{Kezunovic2013}.
 Tuy nhiên, nghiên cứu này chỉ dừng lại ở dạng mô hình và lý thuyết mà chưa đưa ra ứng dụng cụ thể vào thực tế.
@@
 
-Sự thay đổi mạnh mẽ trong ngành điện dường như là chưa có tiền lệ, bao gồm sự chuyển dịch trong phân bổ các dạng năng lượng (Energy Mix), yêu cầu và nhu cầu ngành càng cao của khách hàng, sự xuất hiện các kĩ thuật tiên tiến và thiết bị mới, và các mô hình kinh doanh mới.
-Chính những thay đổi này dẫn đến tính phức tạp (Complexity) và tính không chắc chắn (Uncertainty), cũng như mang tới các thách thức và cơ hội.
+Sự thay đổi mạnh mẽ trong ngành điện dường như là chưa có tiền lệ, bao gồm sự chuyển dịch trong phân bổ tỷ lệ các dạng năng lượng (Energy Mix), yêu cầu và nhu cầu ngành càng cao của khách hàng, sự xuất hiện các kĩ thuật tiên tiến và thiết bị mới, và các mô hình kinh doanh mới.
+Chính những thay đổi này dẫn đến tính phức tạp (Complexity) và tính không chắc chắn (Uncertainty), cũng như mang tới các thách thức và cơ hội mới.
+Ở trung tâm của việc giải quyết các thách thức chính là việc đưa ra các quyết định tốt hơn và tối ưu hơn trong giai đoạn vận hành và lập kế hoạch, bao gồm cả công tác đầu tư dài hạn và phát triển chính sách.
+
+Như chúng ta đã biết, lưới điện thông minh ngày này đã và đang tiếp tục được trang bị các thiết bị có khả năng cảm biến (sensing) và thu thập dữ liệu với độ phân giải cao hơn nhiều so với trước đây.
+Dữ liệu mới kèm theo các phương pháp phân tích mới có thể hỗ trợ các mục tiêu trong quản lý hệ thống điện như việc đạt được độ bền cao hơn, hiệu quả kinh tế tốt hơn và giảm khí thải.
+
+Phân tích dữ liệu vì thế trở thành một trong những yếu tố quan trọng giúp phần đảm bảo và nâng cao các công nghệ hiện có nhằm đưa ra các quyết định tối ưu hơn.
+Có thể nói, công việc phân tích dữ liệu đã và đang trở thành khả năng cơ bản và lợi thế chiến lượng cho các tổ chức muốn đổi mới và luôn hướng tới việc cung cấp dịch vụ chất lượng tốt hơn cho khách hàng.
+Các ứng dụng của phân tích dữ liệu cho ngành điện là rất nhiều và có thể được xác định trong nhiều hoạt động của ngành.
+Do đó, phân tích dữ liệu lớn đã, đang và sẽ là một bước đổi mới quan trọng trong đối trong các tổ chức cơ quan ngành điện lực.
 
 \subsection{Các Nguồn Dữ Liệu Mới}
+Việc áp dụng các giải pháp phân tích dữ liệu lớn trong các hệ thống phân phối và truyền tải điện khác nhau thường được tập trung vào việc khai thác các nguồn dữ liệu hỗn hợp (heterogeneous), là các khối dữ liệu có chất lượng riêng biệt, độ phân giải không gian/và thời gian khách nhau, và có cách trình bày thông tin khác nhau.
+Sẽ là khả thi để nâng cao ứng dụng nguồn dữ liệu này thông qua các phương pháp như: (a) kết hợp các nguồn dữ liệu mới và nguồn dữ liệu truyền thống, ví dụ thông qua việc sử dụng lý thuyết hợp nhất dữ liệu (data fusion) \citep{SimoesCosta2013}; (b) trích xuất và kết hợp thông tin từ các dữ liệu có định dạng khác nhau (ví dụ như dữ liệu ở hạng ảnh và ở dạng văn bản) thông qua các mô hình đã chiều (multimodal learning) \citep{Srivastava2012} hoặc qua một mạng lưới thông tin hỗ hợp (heterogeneous) \citep{Sun2016}; và (c) kết hợp dữ liệu từ các nguồn dữ liệu phân tán về mặt địa lý, ví dụ như việc sử dụng các phương pháp vector autoregressive (VAR) truyền thống \citep{Cavalcante2017} hay phương pháp học sâu (deep learning) \citep{Zhu2020}.
 
-\subsection{Các Cân Nhắc Quan Trọng Khi Tạo Ra Các Gói Dữ Liệu}
+Các nguồn dữ liệu mới đến từ các lĩnh vực khác nhau:
 
-\section{Các Thách Thức của Công Việc Phân Tích Dữ Liệu Lớn}
-\subsection{Các Nền Tảng của Khoa Học Dữ Liệu}
+\begin{itemize}
+	\item \textbf{Hạ tầng mạng lưới điện}: Các cơ quan quản lý và điều hành hệ thống đang cải thiện và nâng cao khả năng quan sát của hệ thống mạng bằng cách lắp đặt các đơn vị đo pha (PMU) có thể cung cấp dữ liệu với tốc độ báo cáo cao (ví dụ, 30 đơn vị đo lường trên mỗi giây của độ lớn điện áp/dòng điện, pha và tần số) và các thiết bị đầu cuối điểm từ xa (Remote Terminal Units - RTU) tại các trạm biến áp và các công tơ điện thông minh lắp đặt cho khách hàng.
+Các cảm biến được dùng cho việc giám sát từ xa các trạm biến áp cũng đang được thử nghiệm để theo dõi tình trạng làm việc của thiết bị và cải thiện chất lượng dịch vụ \citep{Leitao2015}.
 
-\subsection{Các Khía Cạnh Kĩ Thuật}
+	\item \textbf{Các nhà máy điện tái tạo}: Công nghiệp năng lượng tái tạo đang lắp đặt và vận hành các cảm biến giám sát tại cấp độ cánh quạt điện gió và tấm năng lượng mặt trời, tạo ra một lượng lớn dữ liệu (ví dụ, một cánh quạt điện gió có thể có hơn 100 cảm biến bên trong cánh quạt, thu thập hơn 10.000 điểm dữ liệu mỗi giây) có thể được sử dụng cho việc bảo trì dự đoán (và giảm chi phí Vận hành và Bảo trì); dữ liệu từ một lưới dự báo thời tiết số, các cảm biến phân tán địa lý (ví dụ, cánh quạt điện gió, các cảm biến đo độ chiếu sáng), máy ảnh ghi lại bầu trời và hình ảnh vệ tinh có thể được kết hợp để cải thiện khả năng dự báo công suất (và thời tiết) trong nhiều khung thời gian khác nhau \citep{Sweeney2020}.
+Trong việc dự báo điện năng tái tạo, quy mô của các nghiên cứu cũng đã tăng từ một địa điểm đến hơn 100 địa điểm \citep{Messner2019}.
+
+	\item \textbf{Người tiêu dùng và mạng xã hội}: trong khi đang ở giai đoạn đầu của triển khai, sự lan rộng của các thiết bị liên kết Internet trong các căn hộ và tòa nhà thông minh đang tạo điều kiện cho các dịch vụ về năng lượng và phi năng lượng dựa trên dữ liệu \citep{Ahmed2016}, ảnh hưởng của chúng phụ thuộc vào việc giải quyết các thách thức như bảo vệ/quyền riêng tư dữ liệu và tương tác với người tiêu dùng.
+Hơn nữa, sự gia tăng của mạng xã hội đã cho phép các công ty điện lực hiểu và tương tác với khách hàng tốt hơn bao giờ hết \citep{MorenoMunoz2016}.
+Các nhà nghiên cứu cũng đã thử kết hợp dữ liệu Twitter vào việc phát hiện cục bộ mất điện \citep{Sun2016}.
+
+	\item \textbf{Thị trường điện}: Trong vài năm qua, đặc biệt tại châu Âu, tính minh bạch của thị trường điện đã được cải thiện đáng kể, và sau khi xuất bản Quy định (EU) số 5 43/2013 [22], lượng dữ liệu công khai có sẵn đang tăng lên [23], bao gồm cả truy cập vào các đề xuất riêng lẻ từ các công ty cung cấp điện trên thị trường (thông thường có sẵn với độ trễ vài tháng).
+Cùng xu hướng đó đang diễn ra ở Mỹ, với các nền tảng như bộ sưu tập dữ liệu Form EIA-930 cung cấp nguồn dữ liệu toàn diện và tập trung cho các dữ liệu hoạt động hàng giờ của lưới điện cao áp tại 48 tiểu bang.
+Dữ liệu mở này có thể được sử dụng cho các mục đích khác nhau: cải thiện kỹ năng dự báo giá điện bằng cách kết hợp phân tích giá điện từ các khu vực khác nhau [24] hoặc đánh giá ảnh hưởng quy mô lớn của việc phát điện từ nguồn năng lượng tái tạo được cấp bởi lưới điện liên kết với các quốc gia láng giềng [25].
+
+	\item \textbf{Lĩnh vực môi trường và thời tiết}: Dữ liệu thời tiết đóng vai trò quan trọng trong việc dự đoán điều kiện hoạt động, bao gồm cả lỗi hỏng hóc thiết bị và đường dây.
+Dữ liệu từ các trạm thời tiết trên mặt đất [26], vệ tinh [27] và tài nguyên radar [28] có sẵn từ cơ sở dữ liệu của chính phủ.
+Mạng lưới cảm biến chuyên dụng, chẳng hạn như mạng lưới phát hiện sét quốc gia ở Mỹ [29], cũng là nguồn cung cấp dữ liệu thời tiết khá hữu ích.
+Nhiều dịch vụ dự báo thời tiết cũng có sẵn để cung cấp các tính năng được tính trước của các tập dữ liệu thời tiết [30].
+Ngoài ra, dữ liệu về thực vật, đất đai, di cư động vật và các điều kiện xung quanh khác cũng có thể có sẵn từ nhiều nguồn khác nhau [31].
+Cách sử dụng dữ liệu chính xác cao bằng cách sử dụng các cơ sở dữ liệu chuyên dụng như hệ thống (LIDAR - Light Detection and Ranging) hoặc khảo sát bằng thiết bị bay drone cũng được nghiên cứu và đưa vào thực tiễn [32].
+Những dữ liệu này thường không được thu thập trong phạm vi chức năng của ngành điện nhưng chúng lại là một nguồn dữ liệu quan trọng cho ngành này.
+Chẳng hạn, trong dự báo tải, các nghiên cứu đã chuyển từ việc sử dụng nhiệt độ được thu thập tại một trạm đến nhiều biến thời tiết và nhiều trạm thời tiết [33-35].
+Trong dự báo điện mặt trời, dữ liệu hình ảnh bầu trời được sử dụng nhiều để phát hiện đám mây [36].
+\end{itemize}
+
+Các nhà nghiên cứu và chuyên gia hiện nay tập trung vào việc khai thác dữ liệu hiện có và khám phá các nguồn dữ liệu mới nổi và dữ liệu với quy mô lớn để theo đuổi các cải tiến trong kế hoạch và vận hành lưới điện.
+Có nhiều khía cạnh quan trọng khác của phân tích dữ liệu lớn, chẳng hạn như xây dựng thuật toán có thể tận dụng môi trường tính toán hiệu suất cao và mở rộng kích thước mô hình để nắm bắt các đặc trưng chi tiết trong dữ liệu [37].
+Một ví dụ khác là sử dụng dữ liệu tải và thời tiết hàng giờ để dự báo tải dài hạn, thông thường dựa trên dữ liệu hàng tháng [38,39].
+Một bài đánh giá gần đây về phân tích dữ liệu lấy từ các công tơ thông minh liệt kê 10 bộ dữ liệu công khai về nhu cầu điện [40].
+Hình \ref{fig:1} là một ví dụ mô tả các thuộc tính cơ bản của dữ liệu lớn trong ngành điện.
+
+\begin{figure}[h!]
+\centering
+	\begin{center}
+		\includegraphics[width=10cm]{logo1}% This is a *.eps file
+	\end{center}
+	\caption{ Các thuộc tính cơ bản của dữ liệu lớn trong ngành điện}\label{fig:1}
+\end{figure}
+
+\subsection{Các Cân Nhắc Quan Trọng Khi Tạo Ra Các Tập Dữ Liệu}
+
+Cách tạo các tập dữ liệu như thế nào là khá quan trọng để sử dụng và ứng dụng các công cụ phân tích dữ liệu lớn.
+Một số cân nhắc sau cần được quan tâm khi thiết lập, tạo lập, và thu thập các gói dữ liệu:
+
+\begin{itemize}
+\item Tính đồng bộ và mối tương quan theo không gian và thời gian
+\item Tính mở rộng
+\item Dữ liệu trống
+\item Sự đa dạng của dữ liệu xấu
+\item Các loại dữ liệu có độ tin cậy thấp/có tính biến thiên cao
+\end{itemize}
+
+Cách mà mỗi yếu tố này phản ánh lên các ứng dụng dữ liệu lớn trong lưới điện nằm ngoài phạm vi của bài viết này, nhưng chắc chắn đáng khám phá khi các tập dữ liệu mới được thêm vào và kết hợp với nhau trong quá trình xử lý và phân tích.
+
+\section{Thách Thức của Công Việc Phân Tích Dữ Liệu Lớn}
+\subsection{Nền Tảng của Khoa Học Dữ Liệu}
+Mục tiêu của khoa học dữ liệu là trích xuất giá trị từ dữ liệu.
+Các bước của vòng đời quản lý dữ liệu bao gồm thu thập dữ liệu; tiền xử lý (khám phá, lấy mẫu, giảm chiều/thu gọn các đặc trưng và thuộc tính, tạo đặc trưng và thuộc tính, biến đổi, làm sạch và tích hợp); xử lý phân tích (mô hình hóa, thường bao gồm nhiều khối mô hình xây dựng liên kết với nhau); giải thích và báo cáo kết quả [61].
+Những kỹ năng chính cần thiết trong lĩnh vực này thường được xem là đa ngành ở giao điểm của khoa học máy tính, toán học, thống kê và các lãnh vực khoa học kĩ thuật và xã hội khác.
+Trên mặt kỹ thuật, các thách thức chính thường liên quan đến dữ liệu lớn, trí thông minh nhân tạo và các phương pháp học máy, trong khi quá trình khoa học dữ liệu áp dụng cũng có thể yêu cầu các kỹ năng khoa học xã hội, giao tiếp và kinh doanh.
+
+Một cái nhìn toàn diện về khoa học dữ liệu nhấn mạnh rằng khoa học dữ liệu là "hơn là sự kết hợp của thống kê và khoa học máy tính" vì "nó yêu cầu đào tạo làm sao để tổng hợp và kết nối giữa các kĩ thuật thống kê và sức mạnh của máy tính trong một khung làm việc lớn hơn, giải quyết từng vấn đề một cũng như giải quyết các câu hỏi cụ thể trong từng lĩnh vực".
+Các nhà nghiên cứu cũng có cùng quan điểm cho rằng ngành khoa học dữ liệu đòi hỏi: (1) hiểu về bối cảnh của dữ liệu, (2) đánh giá các trách nhiệm liên quan đến việc sử dụng dữ liệu riêng tư và công khai; và (3) giao tiếp rõ ràng về những gì có thể và không thể được suy ra từ một bộ dữ liệu.
+
+Các thành phần cốt lõi của khoa học dữ liệu là các phương pháp dựa trên học máy (Machine Learning - ML) để tìm kiếm các khuôn mẫu (pattern) trong dữ liệu và dựa vào đó để cung cấp thông tin về hiện tượng được mô tả bởi dữ liệu và các dự đoán liên quan đến các sự kiện trong tương lai.
+Trong học máy , mục tiêu là học/tạo ra một hàm (ánh xạ) mô tả và chứa đựng các dự liệu đầu vào (thường gọi là các biến giải thích - explanatory variable) với tham số đầu ra quan sát được (gọi là biến phản ứng).
+Một biểu diễn đơn giản hóa của thực tế được tạo ra cho mục đích này, gọi là mô hình, được sử dụng để ước tính phản ứng chưa biết cho các trường hợp mới dựa trên các biến giải thích quan sát được quan tâm, và quá trình này được gọi là suy luận hoặc đơn giản hơn là dự đoán.
+
+Các mục tiêu của học máy thường được nhóm thành các nhiệm vụ mô tả và nhiệm vụ dự đoán.
+Các nhiệm vụ mô tả nhằm khám phá các mô hình có thể diễn giải được mô tả dữ liệu trong quá khứ, và các nhiệm vụ dự đoán là những nhiệm vụ mà mục tiêu là xác định các mô hình được quan sát trong dữ liệu huấn luyện để ước tính các dự đoán về các rủi ro và các kết quả khác trong tương lai.
+Các nhiệm vụ mô tả thường là không giám sát, có nghĩa là chỉ các biến giải thích được xem xét trong phân tích.
+Các mục tiêu mô tả phổ biến bao gồm phân cụm dữ liệu [64,65], khám phá mối liên hệ [66], và phát hiện sự khác biệt so với hành vi bình thường, bao gồm phân tích giá trị cực đại, phát hiện giá trị ngoại lệ và xác định các mô hình mới nổi [67].
+Các nhiệm vụ dự đoán là giám sát, có nghĩa là chúng không chỉ yêu cầu các biến giải thích mà còn giá trị của biến phụ thuộc đang được dự đoán.
+Các ví dụ thực tế bao gồm đánh giá rủi ro [68] và chẩn đoán bệnh [69].
+
+Có các phương pháp bán giám sát [70] và tự đào tạo [71], trong đó dữ liệu huấn luyện bao gồm một số dữ liệu có nhãn và nhiều dữ liệu không có nhãn.
+
+Trong phân loại, các biến phản hồi được dự đoán là một lớp (ví dụ, một trong vài loại nhãn dữ liệu), hoặc trong trường hợp hồi quy, đó là một giá trị liên tục.
+Một trong những phương pháp thường được sử dụng cho phân loại là cây quyết định (decision tree).
+Thuật toán Hunt [67], một trong những phương pháp cây quyết định sớm nhất, đề xuất quy trình tổng quát của việc phân chia dữ liệu dựa trên giá trị của một thuộc tính duy nhất và tiếp tục đệ quy trên các tập con nếu lớp không đủ thuần khiết ở các tập con.
+Nhiều phương pháp đã được đề xuất để đo độ không thuần khiết của tập dữ liệu và xác định phân nhánh tiếp theo (ví dụ: entropy trong CART [72] hoặc chỉ số Gini trong ID3 [73] và C4.5 [74]), cũng như để tỉa cây, từ đó cải thiện khả năng tổng quát của mô hình.
+Cây quyết định dễ hiểu, rất rẻ để xây dựng và rất nhanh trong việc phân loại các trường hợp không rõ ràng.
+Chúng cũng khá bền với nhiễu và có thể xử lý các thuộc tính trùng lặp hoặc không liên quan, nhưng không tính đến sự tương tác giữa các thuộc tính.
+Một trong những hạn chế của cây quyết định là chúng yêu cầu tỉa, nếu không chúng sẽ trở nên quá lớn dẫn đến các vấn đề quá khớp.
+Giới hạn này được giải quyết thành công bằng Random Forests, được xây dựng như một tập hợp các cây quyết định không tương quan [75,76].
+Tập hợp này được lấy cảm hứng từ phương pháp Bagging, một phương pháp tổng hợp dựa trên lấy mẫu bootstrap, được phát triển để giảm phương sai mà không làm tăng sai số [77].
+Trong Random Forests, ý tưởng này được mở rộng hơn nữa bằng cách giới hạn mỗi nút chỉ xem xét một tập con ngẫu nhiên nhỏ của các thuộc tính.
+Giải pháp kết quả được chứng minh là chính xác hơn so với thuật toán AdaBoost.
+
+Một kỹ thuật thay thế phổ biến có thể xử lý các tương tác giữa các biến giải thích là phân loại một trường hợp mới bằng cách tính khoảng cách đến k láng giềng gần nhất trong tập huấn luyện và dự đoán lớp dựa trên đa số hoặc đa số có trọng số của các láng giềng được xác định kết quả của chúng.
+Đây là một phương pháp học lười vì mô hình không được xây dựng rõ ràng và thời gian suy luận cần thiết để phân loại một trường hợp mới khá lớn.
+Nó cũng yêu cầu so sánh mỗi điểm dữ liệu mới với mỗi điểm dữ liệu trong tập huấn luyện.
+Ngoài ra, kỹ thuật này không dễ sử dụng khi nhiều giá trị thuộc tính bị thiếu, vì trong những trường hợp như vậy, phương pháp dựa trên khoảng cách để xác định láng giềng gần nhất có thể không đáng tin cậy.
+Một số trong số những giới hạn này có thể được vượt qua bằng cách sử dụng đồ thị gần nhất, trong đó các nút được kết nối nếu các điều kiện hình học nhất định được đáp ứng.
+Trong một công thức như vậy, các thuật toán đồ thị hiệu quả khác nhau (ví dụ: cây khung nhỏ nhất và tam giác hóa) có thể được sử dụng để xác định láng giềng gần nhất và tương quan hơn [80].
+
+Một phương pháp phân loại thay thế toán học nghiêm ngặt hơn là ước tính xác suất hậu nghiệm của lớp mục tiêu bằng cách sử dụng định lý Bayes.
+Một phương pháp đơn giản nhưng thanh lịch và mạnh mẽ, được gọi là Naive Bayes, giả định rằng các giá trị thuộc tính độc lập có điều kiện với nhau, cho trước nhãn lớp y.
+Trong trường hợp như vậy, xác suất có điều kiện của lớp của tất cả các thuộc tính có thể được phân tách thành tích của xác suất có điều kiện của mỗi thuộc tính.
+Phương pháp này chống lại nhiễu, giá trị bị thiếu và các thuộc tính không liên quan.
+Tuy nhiên, sự độc lập có điều kiện giữa các biến giải thích là một giả định mạnh mẽ không hợp lệ trong nhiều ứng dụng.
+Đối với các kịch bản như vậy, một lớp mô hình đồ thị xác suất được gọi là Mạng tin cậy Bayesian đã được phát triển bằng cách mô hình hóa các phụ thuộc có điều kiện thông qua các đồ thị vô hướng có hướng.
+Sự suy luận chính xác trên các đồ thị như vậy là NP-khó [81], do đó, các ứng dụng của chúng bị giới hạn cho số lượng thuộc tính nhỏ hơn hoặc cho các loại cấu trúc đồ thị đặc biệt.
+
+Một bộ phân loại xác suất hiệu quả khác là hồi quy logistic [82].
+Chúng tôi đã thành công trong việc sử dụng phương pháp này để dự đoán xác suất mất điện liên quan đến thời tiết [83].
+Giả sử một vấn đề hai lớp (phản hồi y = 0 hoặc 1), phương pháp này tránh việc ước tính trực tiếp xác suất có điều kiện của một trường hợp, mà thay vào đó ước tính tỷ lệ giữa xác suất lớp hậu nghiệm P(y=1|x)/P(y=0|x).
+Một lợi thế lớn của hồi quy logistic so với k láng giềng gần nhất là nó có thể áp dụng cho các vấn đề có số chiều cao, vì phương pháp không phụ thuộc vào việc đo đạc sự tương đồng giữa các điểm dữ liệu.
+Lợi ích khác của phương pháp này là các tham số trọng lượng tương ứng với các thuộc tính riêng lẻ và do đó cung cấp tính giải thích khá dễ dàng.
+Tuy nhiên, sự hiện diện của một số lượng lớn các thuộc tính không liên quan là một thách thức đối với hồi quy logistic, và phương pháp này không áp dụng cho việc phân loại các trường hợp với giá trị bị thiếu, điều này có thể là một hạn chế nghiêm trọng trong thực tế.
+
+Mô hình hồi quy logistic có thể được xem như một trường hợp của mô hình tuyến tính tổng quát.
+Các mô hình mạnh mẽ khác trong danh mục này là Máy vector hỗ trợ (SVM) và Mạng neural đa tầng (MNN).
+Trong SVM, vấn đề tối ưu hóa được định hình như tìm kiếm ranh giới lớn nhất phân tách siêu mặt phẳng cho một khu vực lớn tồn tại ở mỗi bên của ranh giới quyết định [84].
+Điều này được định hình như một vấn đề lập trình phi tuyến tính bị ràng buộc được biểu thị dưới dạng một hàm của các hệ số của siêu mặt phẳng phân tách, được giải quyết bằng phương pháp bội số Lagrange.
+Đối với phân loại phi tuyến tính, dữ liệu được chuyển đổi một cách ngầm định thành không gian có số chiều cao, nơi vấn đề có thể được phân tách tuyến tính.
+Điều này được đạt được bằng cách sử dụng kernel trick, để giảm bớt vấn đề hiệu quả thành tình huống phân loại tuyến tính.
+Sử dụng các kernel được chọn cẩn thận (Gaussian, đa thức hoặc sigmoid) cho phép xấp xỉ các ranh giới quyết định tùy ý.
+Các lợi ích chính của phương pháp SVM là nó chống lại nhiễu và giảm thiểu quá khớp trong khi tìm kiếm giá trị nhỏ nhất toàn cục của hàm mục tiêu.
+Tuy nhiên, chi phí tính toán của SVM là cao, và vẫn là một thách thức để sử dụng mô hình này khi các biến mô tả bị thiếu một phần trong dữ liệu quan sát.
+
+Mạng nơ-ron tiến thẳng đa tầng (FNN) cũng được sử dụng thành công cho phân loại trong nhiều ứng dụng thách thức khác nhau [85].
+Ví dụ, chúng tôi đã thành công trong việc huấn luyện FNN để phân biệt giữa dòng vào từ biến áp và dòng lỗi [86].
+Mô hình này có ít nhất một lớp đơn vị ẩn, mỗi lớp tính toán một hàm phi tuyến mượt và khả vi của tổng đầu vào có trọng số (ví dụ, hàm sigmoid).
+Trong mô hình này, vấn đề cập nhật các tham số khi phát hiện lỗi ở đầu ra thường được giải quyết bằng cách truyền ngược lỗi từ đầu ra về các lớp trước đó.
+Trong quá trình này, lỗi của một nút trong lớp ẩn được ước tính dưới dạng một hàm của các ước tính lỗi và trọng số trong các nút trong lớp trước đó, và giá trị này được sử dụng để cập nhật trọng số của nút ẩn này bằng cách tính toán độ dốc lỗi đối với các trọng số trong nút [87].
+FNN có thể xấp xỉ các hàm tùy ý, và do đó mạnh mẽ hơn SVM.
+Tuy nhiên, khi thiết kế một mạng, quá khớp phải được giải quyết cẩn thận.
+Ngoài ra, nhiễu trong dữ liệu có thể gây ra vấn đề huấn luyện, vì mô hình có thể hội tụ đến một cực tiểu cục bộ, và quá trình huấn luyện có thể yêu cầu một thời gian dài, giới hạn các ứng dụng thực tế.
+Vấn đề khác với FNN cổ điển là học các mạng sâu rất khó, do tác động kết hợp của việc bão hòa hàm kích hoạt sigmoid khi truyền ngược các lỗi nhỏ, dẫn đến sự hội tụ rất chậm.
+Tiến bộ lớn đã được đạt trong việc giải quyết vấn đề này, được gọi là vấn đề gradient biến mất, trong những năm gần đây.
+Điều này, cùng với tiến bộ trong cơ sở hạ tầng tính toán phân tán dựa trên GPU và sự có sẵn của các tập dữ liệu rất lớn, đã cho phép phát triển các mạng nơ-ron sâu hiệu quả, vượt trội hơn rất nhiều so với tất cả các phương pháp thách thức khác nhau, bao gồm thị giác máy tính, xử lý ngôn ngữ tự nhiên, phát âm.
+
+Trong các hệ thống điện, dữ liệu thường được quan sát trên không gian và thời gian, và do đó, các phương pháp hồi quy cấu trúc dựa trên đồ thị tiên tiến hơn được sử dụng để khai thác các phụ thuộc cấu trúc.
+Ví dụ, chúng tôi đã sử dụng học có cấu trúc trong các trường ngẫu nhiên có điều kiện Gaussian để đánh giá rủi ro của sự phá vỡ cách điện cho một sự tiếp xúc cụ thể và các mối đe dọa thời tiết liên quan trong một mạng điện [91].
+Nghiên cứu mới nhất về học sâu [88,92,93] cho thấy rằng một loạt các ứng dụng, bao gồm hồi quy cấu trúc, có thể được hưởng lợi từ việc học các biểu diễn tiềm ẩn cho dữ liệu đầu vào.
+Trong nghiên cứu của chúng tôi [94], việc học biểu diễn cho các trạm hệ thống điện, dựa trên sự gần nhau về không gian của chúng, đã rất có lợi cho việc dự đoán các lỗi cung cấp điện và ước tính xác suất lỗi cung cấp điện.
+Trong một phương pháp như vậy, các nút của đồ thị được nhúng trong không gian chiều thấp hơn, nơi các phương pháp học máy tiêu chuẩn có thể được áp dụng dễ dàng hơn.
+Các thuật toán nhúng nhằm giữ cấu trúc đồ thị và đơn giản hóa các mô hình học bằng cách di chuyển khỏi các biểu diễn đồ thị.
+Một lợi thế của việc sử dụng các phương pháp này là chúng có thể tiết lộ các phụ thuộc không gian phức tạp hơn có chứa một số tương tác xa hơn ngoài tác động của khu vực lân cận.
+
+Quá trình nhúng nút đại diện cho đồ thị ban đầu trong một không gian đặc trưng mới, mô tả tốt nhất các mối quan hệ không gian của các nút trong đồ thị ban đầu.
+Đặc điểm này của quá trình nhúng nút nhằm bắt giữ các mối quan hệ cần thiết của cấu trúc đồ thị ban đầu trong khi đơn giản hóa biểu diễn thành một danh sách các vector đặc trưng chiều thấp hơn.
+
+Có nhiều thuật toán để thu được nhúng như vậy; Hai thuật toán thường được sử dụng là DeepWalk [95] và Node2Vec [96].
+Cả hai thuật toán đều dựa trên thông tin cộng đồng được thu được bằng các bước đi ngẫu nhiên, được sử dụng để học các biểu diễn không gian tiềm ẩn.
+Ngoài ra, DeepWalk có thể thực hiện khám phá cục bộ một cách hiệu quả và có thể chứa đựng các thay đổi nhỏ trong cấu trúc đồ thị mà không cần tính toán toàn cầu lại.
+Node2Vec là một khung công cụ thuật toán hóa mà tổng quát hóa quá trình DeepWalk để cung cấp một khái niệm linh hoạt về khu vực lân cận của một nút, cho phép học các biểu diễn phong phú hơn bằng cách khám phá hiệu quả các khu vực lân cận đa dạng.
+Giải pháp này đã được sử dụng thành công để phát triển một phương pháp mới cho dự báo bức xạ mặt trời, dựa trên các nhúng không gian và thời gian sử dụng mô hình Node2Vec cho dữ liệu đồ thị [97].
+Phương pháp này đơn giản hóa các mô hình học bằng cách di chuyển khỏi các đồ thị phức tạp.
+Mô hình được phát triển cho các dự báo từ 3 đến 12 giờ.
+Mô hình đã dự đoán bức xạ mặt trời với độ chính xác rất cao vào mùa hè, khi có nhiều ngày trời quang đãng.
+Trong những tháng mùa đông, độ chính xác có một chút giảm, nhưng vẫn rất cao và vẫn mạnh mẽ ngay cả khi dữ liệu quan sát bị thiếu về mặt không gian và thời gian.
+
+\subsection{Khía Cạnh Kĩ Thuật}
+Trong khi phân tích dữ liệu lớn dựa trên nền tảng khoa học dữ liệu mạnh mẽ, cũng có một số khía cạnh quan trọng khác để các phương pháp đó được sử dụng trong thực tế.
+Khi tương tác với các chuyên gia và những người trong ngành cố gắng đạt được những lợi ích rõ ràng từ việc sử dụng khoa học dữ liệu, người ta thường nghe thấy rằng phần thực tế của khoa học dữ liệu có thể chỉ chiếm 10\% công việc, trong khi 90\% liên quan đến thiết lập quy trình làm việc, quản lý dữ liệu (và lưu trữ) cũng như các khía cạnh tính toán.
+Do đó, việc quan sát một số khía cạnh liên quan đến kỹ thuật trong phân tích dữ liệu lớn là rất quan trọng.
+Chúng đã được xác định là các thách thức chính cho sự thành công của phân tích dữ liệu lớn [98,99].
+
+Ở trung tâm của khái niệm phân tích dữ liệu lớn là ý tưởng cơ bản rằng dữ liệu cần được xử lý là "lớn".
+Để định nghĩa đúng dữ liệu lớn và các tính năng cần thiết của nó, người đọc được giới thiệu đến [100].
+Các ví dụ điển hình liên quan đến việc thu thập dữ liệu PMU, cũng như dữ liệu có độ phân giải cao ở mức tài sản, (ví dụ, từ các tuabin gió, bộ biến tần điện từ (PV), đồng hồ thông minh, v.v.).
+Tốc độ thu thập dữ liệu này ở các thang thời gian từ giây đến phút, và đồng thời ở nhiều vị trí địa lý khác nhau, trong khi cũng bao gồm nhiều loại biến số khác nhau.
+Nói chung, các dữ liệu như vậy trong lưới điện bao gồm các đo lường điểm, hình ảnh và có thể là văn bản.
+Một số khía cạnh của dữ liệu lớn này cho các hệ thống điện, từ thách thức đến ứng dụng, đã được Arghandeh và Zhou đề cập gần đây [9].
+
+Trong hầu hết các tài liệu khoa học mô tả các ứng dụng điện và ngoài ra, được cho rằng dữ liệu có sẵn và chất lượng tốt.
+Tuy nhiên, bất kể khó khăn đến đâu, đảm bảo giao tiếp dữ liệu, đảm bảo tính toàn vẹn của dữ liệu, cũng như đảm bảo chất lượng dữ liệu, là các bước cần thiết trước khi thiết kế và triển khai một giải pháp dựa trên dữ liệu [101].
+Ngược lại, những khía cạnh liên quan đến tính sẵn có và chất lượng dữ liệu đã được xem xét trong một thời gian khá lâu bởi cộng đồng khoa học máy tính, cho đó đề xuất một số phương pháp [102].
+Sau đó, việc làm sạch dữ liệu và sửa đổi các bộ dữ liệu thường được liên quan đến, mặc dù người ta nên nhận thức rằng những hành động này thực sự có thể ảnh hưởng đến thông tin ban đầu trong các bộ dữ liệu, ví dụ như trong các thuộc tính thống kê của nó [103].
+Đối với các ứng dụng trong các thị trường điện liên quan đến năng lượng tái tạo, một vấn đề cổ điển ví dụ là điền vào các khoảng trống trong chuỗi thời gian, tức là có thể có các khoảng thời gian mà dữ liệu đơn giản không có sẵn, do các lỗi trong việc đăng nhập, lưu trữ hoặc truyền dữ liệu.
+Để điền vào những khoảng trống này, đã đề xuất nhiều phương pháp, tận dụng dữ liệu xung quanh khoảng thời gian đó, dữ liệu với phụ thuộc không gian thời gian (đặc biệt liên quan đến việc tạo ra năng lượng tái tạo dựa trên thời tiết), tính sẵn có của dữ liệu ở các mức tổng hợp khác nhau, cũng như mối quan hệ vật lý giữa các biến số quan tâm (ví dụ, trong một vấn đề luồng điện tối ưu).
+Đối với một ví dụ gần đây liên quan đến dữ liệu tải điện, xem thảo luận trong [104].
+
+Ngoài những khía cạnh liên quan đến dữ liệu, các phương pháp dựa trên dữ liệu được sử dụng trên các tập dữ liệu lớn đòi hỏi năng lực tính toán đáng kể để giải quyết các vấn đề mô phỏng và tối ưu hóa liên quan.
+Để tập trung dữ liệu, những vấn đề này có thể được giải quyết thông qua tính toán hiệu năng cao (HPC), ngày càng trở nên phổ biến trong các ứng dụng lưới điện và thị trường điện [105].
+Những ví dụ đáng chú ý bao gồm luồng công suất tối ưu [106] và vấn đề chuyển mạch truyền tải [107].
+Việc giải quyết các vấn đề quy mô lớn đó với HPC sẽ liên quan đến một số kỹ thuật phân rã, cũng trở nên phổ biến trong các vấn đề vận hành, thị trường và kế hoạch hệ thống điện [108].
+Tuy nhiên, cũng có thể có một số ứng dụng mà điều này không phải là khả thi hoặc không muốn tập trung dữ liệu.
+Trong trường hợp đó, các phương pháp tương tự có thể được sử dụng để giải quyết các vấn đề này theo cách phi tập trung, tuy nhiên, có thể phải chịu gánh nặng giao tiếp tăng do tính chất lặp lại của các phương pháp tối ưu hóa phân tán.
+Những ví dụ đáng chú ý bao gồm các vấn đề luồng công suất tối ưu [109] và học phân tán cho dự báo năng lượng tái tạo [110].
+Có lẽ trong tương lai, các thiết lập liên quan sẽ không được tập trung hoàn toàn hoặc phi tập trung hoàn toàn và sẽ phụ thuộc vào tính toán đám mây, fog và edge computing [111].
+
+Phân tích dữ liệu lớn cần được đặt trong khung cảnh tổng thể của giải quyết vấn đề.
+Trong thực tế, đó chỉ là một công cụ bổ sung để hỗ trợ hoạt động và ra quyết định.
+Do đó, trước khi đầu tư vào các thiết lập dữ liệu lớn cụ thể và các công cụ phân tích, vấn đề và phương pháp giải quyết vấn đề liên quan cần được xác định rõ ràng.
+Ví dụ, nếu các dự báo được sử dụng để hỗ trợ ra quyết định, loại dự báo (ví dụ, xác định hoặc xác suất) và các sản phẩm dự báo (độ phân giải, chuẩn hóa, vv.) nên được quyết định dựa trên vấn đề quyết định cụ thể.
+Ngoài ra, khi chuyển từ dữ liệu sang phân tích, thường có nhu cầu cho một lớp trích xuất đúng loại và mức thông tin từ dữ liệu thô.
+Điều này có thể được thực hiện dựa trên bộ lọc và làm mịn, kỹ thuật kỹ thuật đặc trưng, vv.
+Một ví dụ điển hình sẽ là phát hiện sự kiện dựa trên các luồng dữ liệu, sau đó được sử dụng làm đầu vào cho một số vấn đề phân tích khác.
+
+Các yêu cầu bổ sung có thể mang lại một cấp độ phức tạp khác cho phân tích dữ liệu lớn.
+Yêu cầu quan trọng liên quan đến dữ liệu chính nó (và các luồng dữ liệu liên quan) là cách xử lý an ninh mạng và quyền riêng tư trong lưới điện.
+Hiện nay, an ninh mạng đại diện cho một thành phần quan trọng của hệ thống năng lượng phân phối trong tương lai, mà trên đó phân tích dữ liệu lớn có thể được thực hiện [112].
+Do đó, thiết lập cho phân tích dữ liệu lớn, cũng như các công cụ được sử dụng, cần phải mạnh mẽ để có thể chịu đựng được việc loại bỏ các dữ liệu quan trọng hoặc làm giả dữ liệu.
+Ngoài ra, quyền riêng tư dữ liệu đang trở thành mối quan tâm ngày càng tăng vì nếu dữ liệu đang được thu thập được chia sẻ, có thể suy ra thông tin về các tài sản hoặc người tiêu dùng cụ thể, điều đó không hề được định sẵn.
+Mối quan tâm về quyền riêng tư đặc biệt hiện nay khi các đồng hồ đo thông minh được triển khai rộng rãi [113], cho phép một số người có thể có thông tin về thói quen tiêu dùng của người tiêu dùng để tiến hành các hoạt động tiếp thị và tội phạm.
+Yêu cầu chính khác liên quan đến nhu cầu có các mô hình và kết quả có thể giải thích được.
+Điều này đã kích hoạt một lĩnh vực nghiên cứu mới liên quan đến việc học máy có thể giải thích và học máy dựa trên vật lý.
 
 \subsection{Khung Đưa ra Quyết Định}
+Việc sử dụng phân tích dữ liệu lớn tất yếu dẫn đến việc nâng cao khả năng ra quyết định.
+Do đó, trong phương pháp phân tích dữ liệu được đề xuất, chú ý đặc biệt được dành cho việc trực quan hóa kết quả.
+
+Một cách tiếp cận được sử dụng để dự báo các sự cố mất điện là phát triển bản đồ rủi ro như các bản đồ được hiển thị trong Hình 2.
+Chúng đại diện cho nguy cơ thời tiết, độ dễ tổn thương và rủi ro tính toán dưới dạng tích của hai yếu tố này.
+
+Một khung công việc đánh giá rủi ro được định hình bởi Văn phòng Cứu trợ Khẩn cấp của Liên Hợp Quốc (UNDRO), đã được khám phá gần đây bởi Văn phòng Liên Hợp Quốc về Giảm thiểu Rủi ro Thiên tai (UNISDR) trong báo cáo liên quan đến họ [114], sau đó được Cơ quan Quản lý Khẩn cấp Liên bang (FEMA) của Hoa Kỳ đưa ra, vì trọng tâm chính là ảnh hưởng liên quan đến khí hậu đến cơ sở hạ tầng, xã hội và môi trường.
+Điều này giới thiệu Trạng thái Rủi ro (SoR) như sau:
+
+trong đó P(T) là Nguy cơ hoặc xác suất của một mức độ mối đe dọa cụ thể (T); u(C) là Tổn thất (xã hội, kinh tế hoặc môi trường) liên quan đến mức độ Hậu quả (C), liên quan đến mức độ mối đe dọa (T); và P(C|T) là Tính dễ tổn thương hoặc xác suất của việc trải qua mức độ hậu quả (C) được xác định dựa trên mức độ mối đe dọa (T).
+Đơn vị rủi ro được biểu thị trong đơn vị của các tổn thất.
+
+Trong bối cảnh này, quyết định liên quan đến quá trình tối ưu hóa, trong đó hàm mục tiêu và ràng buộc nhằm mục đích hành động giảm thiểu được xác định.
+Một khung quyết định rộng hơn cho các ví dụ về dự báo mất điện đã được hiển thị trong Hình 3.
 
@@
 
-\section{}
-
 %\section*{Lời Cám Ơn}
@@
 
-%\section*{Figure captions}
+\section*{Ghi Chú}
+Bài viết này được viết dựa trên việc tổng hợp kiến thức từ các bài báo được trích dẫn liệt kê trong phần Tài Liệu.
+Tác giả đã sử dụng tối ưu công cụ \href{https://chat.openai.com/chat}{ChatGPT} được phát triển bởi công ty \href{www.openai.com}{OpenAI} để làm công tác biên dịch và dựa vào gợi ý của ChatGPT, tác giả tiếp tục chỉnh sửa nội dung và câu chữ cho phù hợp hơn với ngôn ngữ Tiếng Việt.
+Ngoài việc sử dụng  \href{https://chat.openai.com/chat}{ChatGPT}, tác giả cũng đối chiếu phần dịch nội dung với công cụ AI được tích hợp trên phần mềm Bing và Edge của Microsoft.
 
@@
 
+\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
+%\bibliographystyle{frontiersinHLTH&FPHY} % for Health, Physics and Mathematics articles
+\bibliography{../ref/UEH-EVN-Data}
+
 \end{document}

### 2023-03-21 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update paper

@@
 	\begin{center}
-		\includegraphics[width=10cm]{logo1}% This is a *.eps file
+		\includegraphics[width=\textwidth]{big-data-properties-VN}% This is a *.eps file
 	\end{center}
-	\caption{ Các thuộc tính cơ bản của dữ liệu lớn trong ngành điện}\label{fig:1}
+	\caption{Các thuộc tính cơ bản của dữ liệu lớn trong ngành điện}\label{fig:1}
 \end{figure}
@@
 
-Một cách tiếp cận được sử dụng để dự báo các sự cố mất điện là phát triển bản đồ rủi ro như các bản đồ được hiển thị trong Hình 2.
+Một cách tiếp cận được sử dụng để dự báo các sự cố mất điện là phát triển bản đồ rủi ro như các bản đồ được hiển thị trong Hình \ref{fig:2}.
 Chúng đại diện cho nguy cơ thời tiết, độ dễ tổn thương và rủi ro tính toán dưới dạng tích của hai yếu tố này.
+
+\begin{figure}[h!]
+	\centering
+	\begin{center}
+		\includegraphics[width=\textwidth]{ruido}% This is a *.eps file
+	\end{center}
+	\caption{Bản đồ thể hiện nguy cơ, sự tổn thương, và rủi do của mạng lưới}\label{fig:2}
+\end{figure}
 
@@
 Trong bối cảnh này, quyết định liên quan đến quá trình tối ưu hóa, trong đó hàm mục tiêu và ràng buộc nhằm mục đích hành động giảm thiểu được xác định.
-Một khung quyết định rộng hơn cho các ví dụ về dự báo mất điện đã được hiển thị trong Hình 3.
+Một khung quyết định rộng hơn cho các ví dụ về dự báo mất điện đã được hiển thị trong Hình \ref{fig:3}.
+
+\begin{figure}[h!]
+	\centering
+	\begin{center}
+		\includegraphics[width=\textwidth]{decisionmaking-risk}% This is a *.eps file
+	\end{center}
+	\caption{Khung ra quyết định cho việc đánh giá rủi do}\label{fig:3}
+\end{figure}
 
 \section{Ứng Dụng}
-\subsection{Quản Lý tài Sản Cơ Sở Hạ Tầng và Mất Điện}
+\subsection{Quản Lý tài Sản Cơ Sở Hạ Tầng và Sự Cố Mất Điện}
+Các thị trường cạnh tranh về điện, sự tư nhân hóa và các yêu cầu kỹ thuật hoặc quy định bắt buộc các công ty điện phải tối ưu hóa các thực tiễn quản lý vận hành, gián đoạn và tài sản và phát triển các kế hoạch quyết định cần thiết kỹ thuật kinh tế.
+Trong thực tế hoạt động của công ty điện hôm nay, quản lý tài sản và gián đoạn được xử lý bởi các nhóm khác nhau và được coi là các nỗ lực lập kế hoạch dài hạn và ngắn hạn, tương ứng.
+Chúng tôi đã giữ cuộc thảo luận về các vấn đề có vẻ không liên quan này cùng với nhau để nhấn mạnh sự tương quan chặt chẽ giữa chúng về việc sử dụng dữ liệu, vì quản lý gián đoạn có thể sử dụng cùng dữ liệu với quản lý tài sản và ngược lại vì tình trạng cơ sở của tài sản điều khiển cả hai ứng dụng:
 
+%\begin{enumerate}
 \subsubsection{Phân Loại Tài Sản}
 
+“Quản lý tài sản trong các hệ thống điện có thể được phân loại rộng rãi thành bốn loại chính dựa trên các thời gian có thể, tức là, thời gian thực, ngắn hạn, trung hạn và dài hạn [115]”
+
+Quản lý tài sản thời gian thực chủ yếu bao gồm các nguyên tắc bền vững của hệ thống điện chính và xử lý các sự cố gián đoạn không đáng kể của thiết bị hệ thống điện và sự cố gián đoạn lưới điện.
+Bằng cách cải thiện nhận thức tình hình, các nhà điều hành lưới điện có thể giám sát và kiểm soát hệ thống một cách hiệu quả.
+Quản lý tài sản ngắn hạn cố gắng tối đa hóa tỷ lệ lợi nhuận liên quan đến đầu tư tài sản.
+Giá trị chủ yếu phụ thuộc vào giá thị trường không chắc chắn thông qua các thực hiện thị trường khác nhau.
+Đánh giá rủi ro thị trường là một yếu tố quan trọng và phân phối doanh thu / lợi nhuận được đạt thông qua phân tích tính lợi nhuận.
+Lập lịch bảo trì tối ưu thuộc quản lý tài sản trung hạn.
+Nó hướng dẫn kế hoạch bảo trì đến các mục tiêu mong muốn của toàn hệ thống một cách đầy đủ.
+
+Các nỗ lực được tập trung vào tối ưu hóa phân bổ tài nguyên tài chính hạn chế ở những nơi và thời điểm cần thiết để quản lý gián đoạn tối ưu mà không đánh đổi tính đáng tin cậy của hệ thống.
+Sử dụng rộng rãi các công nghệ cảm biến thông minh và giám sát để đánh giá sức khỏe và độ tin cậy của thiết bị hệ thống theo thời gian và tối ưu hóa các kế hoạch bảo trì tương ứng [116].
+Đầu tư vào kế hoạch mở rộng hệ thống điện, cũng như triển khai rộng rãi các nguồn điện phân tán, thuộc phạm vi của quản lý tài sản dài hạn, trong đó các nhà đầu tư, các đối thủ cạnh tranh và các bên liên quan khác được mời tham gia vào các kế hoạch kinh tế trong tương lai.
+
+%\end{enumerate}
+
 \subsubsection{Các Tác Động của Thời Tiết và Việc Mất Điện}
+Tác động của thời tiết đối với gián đoạn trong hệ thống điện có thể được phân loại thành trực tiếp và gián tiếp [117]:
 
+\begin{itemize}
+\item Tác động trực tiếp đối với tài sản của tiện ích: Loại tác động này bao gồm tất cả các tình huống khi điều kiện thời tiết nghiêm trọng trực tiếp gây ra sự cố của thành phần.
+Ví dụ như sét đánh vào tài sản của tiện ích, tác động của gió khiến cho cây hoặc cành cây chạm vào đường dây, v.v.
+Những loại gián đoạn này được đánh dấu là gián đoạn do thời tiết gây ra.
+
+\item Tác động gián tiếp đối với tài sản của tiện ích: Loại tác động này xảy ra khi thời tiết tạo ra một tình huống trong mạng lưới mà gián tiếp gây ra sự cố của thành phần.
+Các ví dụ bao gồm: điều kiện thời tiết nóng làm tăng nhu cầu, gây quá tải cho đường dây dẫn đến giảm độ cao của đường dây, tăng nguy cơ sự cố do tiếp xúc với cây, tài sản tiếp xúc với tác động của thời tiết trong dài hạn dẫn đến suy giảm thành phần, v.v.
+Những loại gián đoạn này được đánh dấu là lỗi thiết bị.
+
+\end{itemize}
 \subsubsection{Cơ Bản về Quản Lý Mất Điện}
 
+Khả năng theo dõi đồng thời nhiều mối đe dọa thời tiết khi chúng phát triển và đánh giá các tác động kết quả đa dạng đến tài sản ngành điện, cơ sở hạ tầng và các đường đời mà chúng hỗ trợ là rất quan trọng trong ngành điện.
+Lưới điện được phân bổ trên các vùng rộng lớn với việc phát điện thường đặt ở những vùng xa xôi.
+Tiêu thụ lớn ở các khu đô thị có nghĩa là lưới truyền tải phải đưa điện từ các trang trại phát điện ở xa đến các trung tâm tiêu thụ và hệ thống phân phối phải cung cấp các đường dẫn tiện ích cho từng khách hàng cá nhân.
+Để đạt được điều này, lưới phải trải qua các trạng thái hoạt động khác nhau (Hình 4).
+Các trạng thái thị trường điện tương ứng được hiển thị trong Bảng 3.
+Bằng cách kết hợp quản lý tài sản và gián đoạn, ta xử lý các tác động một cách hiệu quả nhất.
+
 \subsubsection{Ước Đoán Sự Mất Điện và Mạng Lưới Truyền Tải}
+Kiến thức từ dữ liệu lịch sử có thể được sử dụng để dự đoán những cúp điện truyền tải liên quan đến thời tiết 1-3 giờ trước.
+Nhúng không gian được thêm vào bộ dữ liệu đầu vào [94] để bắt các phụ thuộc không gian giữa các nút và sự kiện.
+Xem xét ví dụ với dữ liệu cúp điện lịch sử được thu thập từ Cơ quan Điện lực Bonneville (BPA) [120].
+Chương trình Hệ thống Quan sát Bề Mặt Tự Động (ASOS) [26] được sử dụng để thu thập các đo lường thời tiết lịch sử cho các tham số sau: Hướng gió [độ], Tốc độ gió [nút], Cơn gió mạnh [nút], Nhiệt độ [F], Điểm sương [F], Độ ẩm tương đối [\%], Áp suất [mb], Lượng mưa/giờ [inch] và Mã thời tiết hiện tại.
+Cơ sở Dữ liệu Dự báo Kỹ thuật số Quốc gia (NDFD) [121] được sử dụng để trích xuất dữ liệu dự báo thời tiết lịch sử được sử dụng để kiểm tra hệ thống ánh xạ xác suất cúp điện thời gian thực được sử dụng trong nghiên cứu phối hợp cách điện [41].
 
-\section{Dữ Liệu Lớn trong Nhiều Lĩnh Vực của Nền Kinh Tế}
-% For Original Research articles, please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.
+Việc đặt chỗ lý tưởng của các thiết bị bảo vệ sét dòng dây nhằm giảm thiểu tổng rủi ro của các sự cố và độ nhiễu liên quan đến sét khi vẫn duy trì trong giới hạn ngân sách yêu cầu [41].
+Mạng lưới và tác động của nó được mô hình hóa bằng một đồ thị trọng số đa chế độ sử dụng dữ liệu từ nhiều nguồn khác nhau.
+Mô hình rủi ro được phát triển xem xét tác động tích lũy của các sự cố sét trước đó để tạo ra một ước tính chính xác hơn về độ cứng của bộ cách điện và dự đoán hiệu suất của bộ cách điện cho tình huống gây ra quá áp do sét trong tương lai, sử dụng Gaussian Conditional Random Fields (GCRF) [122].
+Phương pháp quản lý thực vật dựa trên dữ liệu dự đoán trong phân phối được giới thiệu trong [123].
+Một mô hình cho mối tương quan không gian thời gian của một loạt dữ liệu được phát triển, cho phép phân tích thời gian thực về tác động của thực vật lên dây cấp điện dựa trên bản đồ rủi ro dự đoán.
+Thuật toán dự đoán dựa trên GCRF regression predictor.
+Thuật toán tối ưu được sử dụng để tìm lịch cắt tỉa cây động hiệu quả chi phí nhất để giảm thiểu tổng rủi ro của mạng lưới cho mỗi quý.
 
-\section{Dữ Liệu Lớn Trong Ngành Điện Lực}
+\subsubsection{Đánh giá tình trạng hoạt động và sự xuống cấp của trạm biến áp}
 
-\section{Quản Lý Tài Sản Hạ Tầng}
+Các phương pháp truyền thống để đánh giá sức khỏe của bộ biến áp được phát triển bằng cách sử dụng kiến thức lĩnh vực về các quá trình vật lý và hóa học xảy ra bên trong bồn dầu của bộ biến áp, sau đó được xác nhận bằng các nghiên cứu kinh nghiệm.
+Một số ví dụ là tam giác Duval [124], tỷ số khí IEC [125] hoặc phân tích khí chính [126].
+Việc thu thập dữ liệu ngày càng tăng (ví dụ: phân tích khí và dầu định kỳ, các cảm biến thu thập dữ liệu thời gian thực, v.v.) bởi các công ty điện đã thúc đẩy sự phát triển các phương pháp phân tích dữ liệu dựa trên học có giám sát để phân loại tình trạng bộ biến áp và loại lỗi.
+Một số ví dụ là mạng nơ-ron nhân tạo đa tầng [127], máy vector hỗ trợ [128] và mạng Bayesian [129].
+Những sự cố của bộ biến áp hạ áp (22,9KV-220V) được sử dụng trong các lĩnh vực phân phối ở Hàn Quốc được nghiên cứu trong [130].
+
+Các thuật toán học có giám sát đối mặt với các thách thức sau: (i) đa số các thuật toán chỉ cung cấp khả năng giải thích thấp cho người ra quyết định, điều này là yêu cầu cơ bản trong vấn đề này; (ii) thiếu dữ liệu lỗi có chất lượng cao; và (iii) dữ liệu được gán nhãn về tình trạng biến áp trong hầu hết các trường hợp không có sẵn hoặc được xác định bởi con người (tức là phân loại chủ quan của tình trạng).
+Sử dụng học không giám sát là một giải pháp thay thế và hấp dẫn, nhưng trong tài liệu vẫn còn hạn chế.
+Phương pháp không giám sát đầu tiên là chỉ số sức khỏe được mô tả trong tài liệu tham khảo [131] tóm tắt sức khỏe tổng thể của tài sản bằng cách kết hợp kết quả quan sát vận hành, kiểm tra trường và kiểm tra phòng thí nghiệm thành một chỉ số duy nhất.
+Tuy nhiên, các hạn chế chính của chỉ số này là: (i) định nghĩa kinh nghiệm của trọng số cho mỗi tiêu chí, và (ii) thiếu thông tin về loại lỗi.
+Các phương pháp thay thế khác là phân cụm [132] và học bán giám sát với Low Dimensional Scaling [133].
+Hơn nữa, nghiên cứu về học sâu có thể đóng góp vào việc bổ sung dữ liệu [134] cho dữ liệu biến áp và cung cấp các kỹ thuật mới cho học không giám sát như mạng Siamese [135].
+Cần nhấn mạnh rằng chúng ta nên mong đợi hiệu suất thấp hơn từ các kỹ thuật không giám sát so với học có giám sát, nhưng tiềm năng ứng dụng lớn hơn trong tập dữ liệu thực tế.
+
+\subsubsection{Những dự đoán về thiệt hại cơ sở hạ tầng thảm khốc gây ra các tình trạng mất điện}
+
+Phân tích dữ liệu lớn có thể được sử dụng để dự đoán các sự cố về tài sản thảm khốc do các sự kiện thời tiết khắc nghiệt như bão, vòng xoáy, lốc xoáy và sóng thần [136,137].
+Những nghiên cứu như vậy chủ yếu liên quan đến dự đoán thiệt hại cơ sở hạ tầng, bao gồm số cột bị đổ, trạm biến áp bị phá hủy và các thiệt hại khác đòi hỏi phải tái thiết kế toàn bộ cơ sở hạ tầng lưới điện.
+Một đánh giá ngầm cũng liên quan đến các tình trạng mất điện vì cần phải tái thiết kế trước khi khôi phục dịch vụ
+
+\subsection{Phân tích dữ liệu thu thập từ các công tơ thông minh}
+
+Ứng dụng này có nhiều trường hợp sử dụng khác nhau:
+
+\begin{itemize}
+\item Các ứng dụng của phân tích một đồng hồ thông minh duy nhất
+\item Các ứng dụng của các nhóm đồng hồ thông minh
+\item Các đồng hồ thông minh được kết nối với các mô hình lưới
+\end{itemize}
+
+Sự xâm nhập của các đồng hồ thông minh vào Mỹ đã vượt quá 50\%, và ở một số quốc gia châu Âu vượt quá 50\%, cung cấp cơ hội quan trọng cho phân tích dữ liệu để nâng cao quản lý khách hàng và hoạt động và kế hoạch lưới.
+Các đồng hồ thông minh cung cấp các đọc số về năng lượng, công suất và điện áp ở độ tinh granularities thường là một giờ hoặc 15 phút.
+Các công ty điện, nhà cung cấp dịch vụ năng lượng, khách hàng và nhà nghiên cứu đã xác định nhiều trường hợp sử dụng cho phân tích đồng hồ thông minh như dự báo, phân loại và phân loại tải khách hàng, ước tính tải và mô hình lưới nâng cao [40].
+Khi kết hợp với các nguồn dữ liệu khác và các hệ thống tiện ích, phân tích đồng hồ thông minh có thể mở rộng thêm lợi ích của nó cho toàn doanh nghiệp hoạt động.
+Tóm tắt của ứng dụng đồng hồ thông minh được minh họa trong Hình 5.
+
+Dưới đây, chúng tôi cung cấp một tóm tắt các ứng dụng nổi bật của thông minh phân tích dữ liệu đồng hồ với các tham chiếu tương ứng.
+Một số ứng dụng sẽ được mở rộng trong các phần tiếp theo dưới các tiêu đề riêng biệt.
+
+\subsubsection{Dự báo tải}
+Ngành công nghiệp điện đã sử dụng dự báo tải cho hoạt động và kế hoạch lưới và quản lý khách hàng.
+Dữ liệu đo đạc thông minh đã thu hút sự chú ý của các nhà nghiên cứu trong thập kỷ qua cho cả dự báo tải điểm và xác suất [138,139].
+Một xử lý toàn diện hơn về chủ đề này được đề cập trong phần con.
+
+\subsubsection{Hồ sơ tiêu thụ điện của khách hàng}
+
+Load profiling đề cập đến phân loại các đọc lịch sử về nhu cầu của khách hàng thành các nhóm dựa trên hành vi của họ.
+Các kỹ thuật phân cụm như k-means, phân cụm phân cấp và bản đồ tự tổ chức đã được sử dụng cho việc phân loại tải [140-142].
+Các mô hình biến thời gian kết hợp với phân cụm đã được sử dụng để phát triển mô hình tải điện phức tạp bằng cách sử dụng dữ liệu Cơ sở hạ tầng Đo lường tiên tiến (AMI) [143]
+
+\subsubsection{Phân tích DER}
+Một lượng đáng kể các nguồn năng lượng phân tán (DER) đang được kết nối với lưới điện, bao gồm điện mặt trời PV, lưu trữ năng lượng và xe điện.
+Những nguồn tài nguyên này trong lưới tạo ra những thách thức mới cho các nhà cung cấp tiện ích bao gồm biến động điện áp [144], vi phạm giới hạn nhiệt, dòng ngược và ảnh hưởng đến tuổi thọ dự kiến của cơ sở hạ tầng như máy biến áp và bộ điều chỉnh điện áp.
+Việc có thông tin chính xác liên quan đến Nguồn năng lượng phân tán (DER) tại mạch phân phối và đằng sau đồng hồ điện (BTM) là rất quan trọng đối với các tiện ích.
+Các nhà nghiên cứu đã chứng minh khả năng phát hiện điện mặt trời trên mái nhà [145] và xe điện sạc tại đầu tiêu dùng [146] bằng cách sử dụng phân tích không xâm nhập trên dữ liệu đồng hồ thông minh.
+Các mạng nơ-ron sâu đã được sử dụng để phát hiện kích thước, độ nghiêng và thông số hướng của các cài đặt điện mặt trời dựa trên dữ liệu AMI [147]
+
+\subsubsection{Các ứng dụng với mạng lưới điện}
+Dữ liệu đồng hồ thông minh có thể được sử dụng kết hợp với dữ liệu nguồn cấp phân phối để có được các mô hình tinh vi cho kế hoạch phân phối hoặc để có được cái nhìn sâu sắc vào các vấn đề mô hình hóa cụ thể.
+Ví dụ, dữ liệu đồng hồ thông minh đã được sử dụng để phát hiện bất thường [148] như thay đổi đáng kể về nhu cầu hoặc điện áp.
+Dữ liệu đồng hồ thông minh bị gián đoạn đã được sử dụng cho phân tích và phục hồi gián đoạn tối ưu [149] và thiết kế tái cấu hình nguồn cấp.
+Phân tích đồng hồ thông minh cũng đã được sử dụng để sửa chữa kết nối biến áp [149], xác định pha [150], xác định độ phức tạp [151] và ước lượng tham số [152]
+
+\subsubsection{Dự báo tải}
+Trong một cuộc khảo sát phân tích tiện ích được tiến hành vào năm 2017 với 136 tiện ích từ 24 quốc gia, 52\% số người trả lời coi dự báo năng lượng là một ứng dụng ưu tiên cao nhất, tỷ lệ cao nhất trong tất cả các ứng dụng trong cuộc khảo sát [153].
+Cải thiện dự báo có thể dẫn đến các quyết định vận hành và kế hoạch tốt hơn và do đó dẫn đến tiết kiệm tiền tệ hoặc nâng cao độ tin cậy của hệ thống.
+Tùy thuộc vào các yếu tố như kích thước của công ty và mức giảm lỗi, việc giảm lỗi dự báo có thể dẫn đến tiết kiệm hàng triệu đô la mỗi năm [154].
+Trong thời đại dữ liệu lớn, sự tăng trưởng của dữ liệu, sự tiến bộ của các công nghệ tính toán và các đột phá trong phân tích tiên tiến tiếp tục kích thích cải tiến các kỹ thuật và phương pháp dự báo năng lượng.
+Nhiều trong số những phát triển gần đây này đã được công nhận là các giải thưởng trong các Cuộc thi Dự báo Năng lượng Toàn cầu (GEFCom) [42-44].
+
+Các tiện ích đã thực hành dự báo tải trong hơn một thế kỷ [155].
+Theo sự phát triển của lưới điện và ngành công nghiệp điện, dự báo tải dài hạn, hoặc dự báo tải không gian, đặc biệt là đã được coi là một thành phần quan trọng trong kế hoạch hệ thống điện vào cuối thế kỷ 20 đến đầu thế kỷ 21 [156-158], khi dữ liệu tải được thu thập từ các biến áp phân phối và ở độ phân giải thấp, chẳng hạn như hàng tháng hoặc hàng năm.
+Tuy nhiên, nhu cầu vận hành phức tạp ngày càng đòi hỏi dự báo tải ngắn hạn chính xác [159].
+Các kỹ thuật trí tuệ nhân tạo, chẳng hạn như mạng nơ-ron nhân tạo, đã chiếm đa số các tài liệu vào những năm 1990 [160].
+Mặc dù nhiều mô hình được phát triển cho Luồng Tải Ngắn Hạn (STLF), nhưng hầu hết đều không có giá trị thực tiễn.
+Một thành công đáng chú ý phát triển từ giới học thuật vào những năm 1990 là một mô hình mạng nơ-ron, sau đó được thương mại hóa và vẫn được sử dụng trong ngành công nghiệp ngày nay [161].
+Một khám phá nghiên cứu học thuật gần đây khác đã được thương mại hóa và triển khai trên toàn cầu trong những năm 2010 là một khung mô hình dựa trên hồi quy [39,162].
+
+Khi các công nghệ tự động hóa phân phối và lưới điện thông minh cung cấp dữ liệu độ phân giải không gian thời gian cao cho các nhà dự báo tải, nghiên cứu về dự báo tải cũng phát triển mạnh.
+Ví dụ, các nhà cung cấp điện bán lẻ bắt đầu sử dụng dữ liệu hàng giờ cho dự báo tải dài hạn.
+Một số nhà khai thác thị trường và tiện ích cũng dựa vào dữ liệu hàng giờ hoặc hàng phút để dự báo tải ở mức điện áp cao [38,39,164].
+Những dữ liệu tải có độ phân giải cao này cho phép các nhà dự báo xây dựng mô hình với hàng trăm thông số có thể bắt được nhiều đặc trưng quan trọng trong tải [37].
+Chúng cũng cho phép các nhà dự báo tải cải thiện dự báo tổng thể bằng cách tận dụng thông tin tải cấp đồng hồ.
+[138].
+
+Trong khi khai thác việc sử dụng dữ liệu tải có độ phân giải cao, các nhà nghiên cứu và chuyên gia cũng đầu tư một số nỗ lực vào các cách sáng tạo để khai thác dữ liệu thời tiết.
+Truyền thống, chỉ có một số lượng nhỏ các trạm thời tiết được sử dụng để dự báo tải trên một khu vực cụ thể.
+Tại GEFCom2012, chuỗi nhiệt độ từ 11 trạm thời tiết đã được công bố cho cộng đồng nghiên cứu [42].
+Một phương pháp lựa chọn trạm thời tiết sau đó được đề xuất và cho thấy giá trị trong việc cải thiện dự báo tải tiên tiến nhất.
+Phương pháp này sau đó đã được sử dụng bởi một đội chiến thắng tại GEFCom2014 [165].
+Gần đây, nó đã được cải thiện thêm bởi các nhà nghiên cứu khác [166,167].
+Vì dữ liệu thời tiết có thể được sử dụng cho nhiều mô hình dự báo tải, các phương pháp lựa chọn trạm thời tiết này có lợi cho cả dự báo tải dài hạn và ngắn hạn.
+
+Trong khi xu hướng nghiên cứu dự báo tải chính là tận dụng dữ liệu lớn, một xu hướng khác là nhúng thông tin toàn diện về tương lai vào dự báo tải.
+Dự báo tải xác suất, cung cấp dự báo theo phân vị, khoảng xác suất hoặc hàm mật độ, là chủ đề nóng trong thập kỷ qua [168].
+Những dự báo tải xác suất này có thể được tạo ra thông qua mô phỏng phần dư [169,170], tạo ra các kịch bản đầu vào [171] hoặc kết hợp dự báo điểm [172].
+Hồi quy phân vị cũng đã được áp dụng để tạo ra dự báo tải xác suất [139,172,173].
+Một số khía cạnh phương pháp của dự báo cũng đã được nghiên cứu đặc biệt cho dự báo tải xác suất, chẳng hạn như lựa chọn đặc trưng [174,175] và kết hợp dự báo [176].
+
+Ngoài việc cải thiện dự báo tải ở mức điện áp cao, các nhà nghiên cứu và chuyên gia cũng đã dành nhiều nỗ lực cho dự báo tải ở mức điện áp trung bình và thấp.
+Một số dự báo tại mức điểm cung cấp [42, 44], trong khi những người khác là ở các đồng hồ đo riêng lẻ [40, 139, 177].
+Chủ đề phổ biến khác là dự báo tải cho tòa nhà, từ các tòa nhà học thuật [178, 179] đến các tòa nhà dân cư [180, 181].
+Trong khi thời tiết vẫn đóng vai trò chính trong nhiều mô hình dự báo tải ở mức tòa nhà, tác động của thời tiết đối với tải công nghiệp là khá nhỏ.
+Dự báo tải cho nhà máy và các cơ sở công nghiệp là một chủ đề mới nổi trong loại này [182-184], và bao gồm dự báo tải công suất phản kháng.
+
+\subsection{Dự báo và phân tích năng lượng tái tạo}
+Việc triển khai năng lượng tái tạo đã tiếp tục được duy trì với tốc độ ổn định trong thập kỷ qua.
+Do tính không đồng nhất của việc tạo điện từ các nguồn năng lượng tái tạo, cũng như do sự khó lường hạn chế, sự chú trọng đã được đặt vào việc phát triển các phương pháp tích hợp những nguồn tái tạo đó.
+Ngoài các khía cạnh liên quan đến mã lưới điện và một số phương pháp quản lý năng lượng phân tích mới được đề cập trong các phần khác, một phần lớn nỗ lực đã tập trung vào cách tận dụng tối đa các dữ liệu có sẵn để cải thiện kiến thức về việc tạo điện từ nguồn năng lượng tái tạo, thường là với mục đích dự báo.
+Một trong những bài báo đầu tiên đề xuất mô hình động để dự đoán tốc độ gió và sản lượng điện tương ứng là [185].
+Mặc dù tập trung vào các mô hình đơn giản và một bố cục ảo, bản thảo đó đã đặt nền tảng cho một số phát triển tiếp theo.
+Rõ ràng, vào thời điểm đó, các khía cạnh dữ liệu lớn chưa được thảo luận, và số chiều của các mô hình liên quan cũng nhỏ.
+
+Hôm nay cho các trang trại gió, đặc biệt là các trang trại ngoài khơi, việc thu thập dữ liệu ở mức độ động cơ gió được thực hiện với độ phân giải một giây.
+Tương tự, đối với các nhà máy năng lượng mặt trời, dữ liệu có thể được thu thập ở mức độ biến tần và với độ phân giải cùng cấp độ giây.
+Những dữ liệu ở mức độ rất tinh vi này được sử dụng để cải thiện phân tích và dự báo tại mức độ trang trại gió (hoặc nhà máy năng lượng mặt trời) [186].
+Ngoài ra, vì khả năng sinh năng lượng từ các nguồn tái tạo bắt đầu trở nên đa dạng và phân tán địa lý một cách dày đặc, ta cũng có thể sử dụng tất cả dữ liệu được thu thập tại các trang trại để cải thiện dự báo [14,18,187].
+
+Ngoài ra, còn có rất nhiều nguồn dữ liệu khác có liên quan, chủ yếu là liên quan đến quan sát và dự báo khí tượng, mô tả các quy trình phức tạp và mang lại khối lượng dữ liệu rất lớn.
+Về phía quan sát khí tượng:
+
+\begin{itemize}
+\item Sky imagers có tiềm năng lớn cho mô hình hóa và dự báo năng lượng mặt trời với độ phân giải cao vì chúng theo dõi các đám mây di chuyển và tác động của chúng đến các tấm pin mặt trời [188].
+Chúng có thể cung cấp hình ảnh của bầu trời trên các nhà máy điện mặt trời mỗi 30 giây;
+
+\item Radar thời tiết cũng đã chứng minh sự quan tâm của họ trong đánh giá và mô hình các chế độ động để ứng dụng vào dự báo năng lượng gió độ phân giải cao [189].
+Tùy thuộc vào công nghệ, hình ảnh radar có thể có sẵn giữa mỗi phút và mỗi 10 phút, với bán kính hình ảnh từ 60 đến 250 km;
+
+\item LIDAR ngày càng được coi là rất liên quan đến các đo lường gió ở phía trước của các tuabin gió và tích hợp vào các phương pháp dự báo [190], hoặc nói chung là các quan sát tiềm năng mới của các hồ sơ gió được sử dụng trong dự báo thời tiết và năng lượng tái tạo [191].
+LIDAR cung cấp các đo lường gió cho hình nón mà chúng quét (theo chiều dọc hoặc chiều ngang, tùy thuộc vào cách chúng được thiết lập) mỗi vài giây.
+\end{itemize}
+
+Ngoài ra, có thể đề cập đến hình ảnh vệ tinh, có tiềm năng quan trọng cho năng lượng gió, mặt trời và sóng.
+Tuy nhiên, tần suất cập nhật thấp hơn khiến chúng ít quan trọng hơn ở thời điểm hiện tại.
+Thông tin từ các loại thiết bị này được gọi là thông tin được cảm biến từ xa.
+
+Một thách thức đầu tiên và phức tạp khi xử lý thông tin cảm biến từ xa là giảm chiều.
+Điều này có thể được thực hiện: (i) dựa trên các kỹ thuật xử lý tín hiệu và thống kê, ví dụ như Phân tích thành phần độc lập-ICA, cho các trường chuyển động trong hình ảnh radar thời tiết; (ii) bằng cách trích xuất các đặc trưng vật lý như mây trong hình ảnh bầu trời [188] hoặc hệ thống mưa và đặc điểm của chúng trong hình ảnh radar [192]; hoặc cuối cùng (iii) thông qua các mô hình chức năng như các hồ sơ gió cho các đo đạc dọc theo chiều dọc của LiDAR.
+
+Ngoài các quan sát khí tượng, dữ liệu đầu vào chiều cao mới có thể có dạng dự báo thời tiết.
+Trước hết, thông tin liên quan trong dự báo thời tiết không chỉ có thể dành cho điểm gần nhất với một điểm quan tâm mà còn có thể được cung cấp trên toàn khu vực của điểm đó [59].
+Thứ hai, để diễn đạt dự báo theo cách xác suất, các dự báo thời tiết tập hợp bao gồm một tập hợp các quỹ đạo thay thế và có khả năng như nhau (thông thường, giữa 10 và 100) cho các biến thời tiết liên quan.
+Chúng có thể được hiểu là các thực hiện mẫu từ các dự báo xác suất đa biến để cung cấp cho các phương pháp dự báo năng lượng tái tạo.
+Chúng có sẵn trên các khu vực lớn.
+ví dụ, toàn bộ châu Âu, cung cấp thông tin về phụ thuộc không gian-thời gian đa biến trong việc tạo điện năng tái tạo [193,194].
+Cuối cùng, dự báo thời tiết độ phân giải rất cao (độ phân giải không gian trong khoảng vài mươi đến vài trăm mét và độ phân giải thời gian trong khoảng vài giây đến vài phút) đang mang lại cơ hội mới cho dự báo năng lượng tái tạo, như trong ví dụ gần đây trong [186] và các ứng dụng tại các trang trại gió ngoài khơi.
+
+Hiện nay, sự sẵn có của số lượng lớn dữ liệu này yêu cầu những thay đổi cơ bản trong phân tích và dự báo năng lượng tái tạo, không chỉ về phương pháp mà còn về các mô hình kinh doanh.
+Chúng ta mong đợi sẽ có nhiều công trình đổi mới xuất hiện trong tương lai, đề xuất các phương pháp dựa trên phương trình vi phân ngẫu nhiên, học sâu, học phân tán và liên kết, cũng như thị trường dữ liệu.
+
+\subsection{Hiệu quả và tối ưu hóa năng lượng}
+Các công nghệ lưới thông minh cung cấp tiềm năng lớn để tăng cường hiệu quả sử dụng năng lượng trong các lĩnh vực khác nhau.
+Tuy nhiên, hiện tại, các hoạt động tăng cường hiệu quả năng lượng chủ yếu bị hạn chế trong việc thực thi chứng nhận ISO 50001 như sau: (i) lắp đặt thiết bị bổ sung (đồng hồ đo, cảm biến, v.v.) để đo lường tiêu thụ năng lượng; (ii) lắp đặt phần cứng mới và thay thế thiết bị; và (iii) trực quan hóa dữ liệu trong giao diện người dùng liền mạch, tìm các mô hình bất thường; và xác định các quy trình tiêu thụ năng lượng cao.
+Thực tiễn tiêu chuẩn này cung cấp việc giám sát và nhận thức về tiêu thụ năng lượng cho các quyết định của con người, nhưng nó không cho phép phân tích theo chỉ đạo và kiểm soát quy trình tự động.
+
+Các tác phẩm khác nhau trong văn học khai thác các kỹ thuật tối ưu hóa năng lượng dựa trên mô hình, chẳng hạn như lập trình số nguyên cho việc giảm tải đỉnh trong các nhà máy thép [195] và lập trình tuyến tính số hỗn hợp cho các thiết bị gia dụng nhiệt [196].
+Các phương pháp dựa trên mô hình có nhược điểm là yêu cầu một mô hình (toán học) về quá trình vật lý và các ràng buộc, có thể phức tạp trong một số trường hợp.
+Chúng không cho phép cải tiến liên tục các chính sách kiểm soát.
+
+Sự xuất hiện của công nghệ internet-of-things cung cấp điều kiện kỹ thuật cho mô hình hóa động lực dữ liệu trong tối ưu hóa năng lượng.
+Một số ví dụ là việc sử dụng học tăng cường dạng batch (fitted Q-iteration) để điều khiển một nhóm máy sưởi nước điện gia đình cho dịch vụ đáp ứng nhu cầu [197]; fitted Q-iteration kết hợp với auto-encoders để tối ưu hóa năng lượng trong máy sưởi nước điện [198]; học tăng cường sâu cho tối ưu hóa năng lượng dự đoán các trạm bơm nước thải [199]; và mô hình cây cho tiêu thụ năng lượng của tòa nhà để lên lịch hệ thống sưởi ấm tối ưu [200].
+Phương pháp này không yêu cầu việc mô hình hoá đầy đủ các phương trình quá trình vì sự hiểu biết của nó được thực hiện theo thời gian thực thông qua dữ liệu.
+Tuy nhiên, sẵn có dữ liệu và thời gian (ví dụ: số lần tương tác với hệ thống vật lý) cần thiết để "huấn luyện" các phương pháp tối ưu dữ liệu vẫn là thách thức thực tế cho sự áp dụng trong ngành công nghiệp.
+
+Ngoài tối ưu hóa năng lượng dữ liệu, các dịch vụ năng lượng dựa trên dữ liệu khác có thể được cung cấp bởi các công ty dịch vụ năng lượng (ESCO), nhà tập đoàn và nhà bán lẻ nhằm tối đa hóa nhận thức của người sử dụng cuối đối với các hành động tăng hiệu quả sử dụng năng lượng và rút ra giá trị kinh doanh từ dữ liệu.
+
+Một đóng góp quan trọng để tăng sự nhận thức của người sử dụng cuối là giám sát tải không xâm phạm từ dữ liệu của các công tơ thông minh, có thể được áp dụng để phát hiện và ước tính các cài đặt điện PV dân dụng [145], tiêu thụ bơm nhiệt [201] và các hồ sơ tải cá nhân từ các đường cong tải nguồn [202].
+Phân đoạn khách hàng với các hồ sơ tải có thể được áp dụng bởi các nhà tập đoàn và nhà bán lẻ để xác định khách hàng mục tiêu, thiết kế một loại giá động hoặc thời gian thực tùy chỉnh [203], mô hình hóa hành vi động của các tải/công tắc có thể kiểm soát được [204], hoặc thông báo cho người tiêu dùng nếu họ đang đối mặt với một sự thay đổi không bình thường của hồ sơ tải.
+Các kỹ thuật tiêu chuẩn là phân cụm chuỗi thời gian lô, nhưng phân cụm trực tuyến là một yêu cầu cơ bản do tính động của hành vi tiêu thụ [205].
+
+Dữ liệu từ đồng hồ đo thông minh kết hợp với các biến môi trường bên ngoài (ví dụ: nhiệt độ ngoài trời) có thể hỗ trợ cho các nhà bán lẻ và nhà tổng hợp ước tính khả năng phản ứng yêu cầu của khách hàng.
+Trong [206], một vấn đề bao lưu ngẫu nhiên được đặt ra để lựa chọn khách hàng trong các chương trình phản ứng yêu cầu sử dụng dữ liệu tiêu thụ, cũng như để ước tính xác suất đạt được mục tiêu giảm tải.
+Các phương pháp suy luận nhân quả giữa các chương trình tính tiền Nguồn Phân tán (DR) và tiêu thụ được sử dụng trong [207] để ước tính độ đàn hồi của người tiêu dùng và xác định liệu các chương trình tính tiền động có ảnh hưởng đến sơ đồ tiêu thụ thông thường của người tiêu dùng hay không; mục tiêu tương tự cũng được đạt được với phương pháp dựa trên tương quan [208].
+Hơn nữa, học trực tuyến có thể được sử dụng để điều chỉnh tín hiệu giá động để đạt được hành vi sử dụng mong muốn, ví dụ, bằng cách đặt ra một vấn đề tối ưu lồi trực tuyến [209] hoặc thông qua một mô hình tiện ích tham số hóa [210].
+
+Dữ liệu được thu thập từ đánh giá và chứng nhận hiệu quả năng lượng rất có giá trị cho các bên liên quan và dịch vụ khác nhau [211].
+Ví dụ, Cơ quan Năng lượng (DOE) cơ sở dữ liệu hiệu suất tòa nhà có thể được sử dụng cho các mục đích khác nhau [212]: so sánh hiệu suất tiết kiệm năng lượng của các loại tòa nhà khác nhau và địa điểm địa lý; ước tính tiềm năng tiết kiệm năng lượng liên quan đến các hành động cải tiến cụ thể; và đánh giá tác động của các công nghệ tiết kiệm năng lượng ở mức độ danh mục.
+Trên thực tế, các kỹ thuật dựa trên dữ liệu có thể được sử dụng để đánh giá trước các biện pháp cải tiến năng lượng về mặt tiết kiệm năng lượng và "giảm rủi ro" đầu tư vào công nghệ tiết kiệm năng lượng.
+Ví dụ, hồi quy đa tuyến có thể sản xuất một ước tính xác suất về lợi nhuận liên quan đến các biện pháp cải tiến khác nhau, xem xét các đặc điểm và hệ thống của tòa nhà [213].
+
+\subsection{Phân tích dữ liệu synchrophasor và sự kiện}
+Để đánh giá tầm quan trọng của dữ liệu Đơn vị Đo Pha (PMU), ta phải trở về các báo cáo về nguyên nhân của sự cố mất điện lớn tại đông bắc Hoa Kỳ vào ngày 13 tháng 8 năm 2003 [214].
+Ban Nghiên cứu Sự cố Hệ thống Điện quốc tế Hoa Kỳ-Canada đã kết luận rằng một trong những lý do chính dẫn đến sự cố mất điện là thiếu nhận thức về tình hình thực tế.
+Điều này, lần lượt, được cho là do tầm nhìn hạn chế của nhà điều hành về các sự kiện và động lực liên quan đến hệ thống điện được cung cấp bởi Hệ thống Quản lý Năng lượng (EMS) thông qua các thiết bị đo trường như hệ thống Giám sát và Thu thập Dữ liệu (SCADA) và các thiết bị ghi lại tại trạm biến áp như máy ghi sự cố kỹ thuật số, máy bộ định hướng kỹ thuật số và máy ghi sự kiện trình tự.
+Thiếu sót chính là tốc độ báo cáo đo đạc (Đơn vị Trung tâm Tầm xa SCADA), thiếu dấu thời gian khác biệt (cơ sở dữ liệu SCADA) và không thể cập nhật mô hình hệ thống điện kịp thời để phản ánh các sự kiện chuyển mạch liên tiếp (bộ ước lượng trạng thái).
+Ngoài ra, các thiết bị ghi sự cố không được đồng bộ thời gian và đánh dấu thời gian đầy đủ.
+Để khắc phục những thiếu sót đó, các PMU và hạ tầng đo đạc phụ thuộc vào synchrophasor được xác định là cơ sở đo lường phù hợp.
+Với sự tăng cường nguồn kinh phí từ Đạo luật Phục hồi và Tái đầu tư Hoa Kỳ năm 2009, nhiều PMU đã được lắp đặt trên khắp Hoa Kỳ và hiện đang có hơn 2.500 đơn vị.
+Một số lượng PMU ngay cả lớn hơn đã được lắp đặt tại Trung Quốc và đang tiến hành kế hoạch lắp đặt một số lượng lớn tại Ấn Độ.
+
+Một trong những điểm mạnh chính của PMU là cung cấp các phép đo đồng bộ thời gian của phasor điện áp và dòng điện thời gian thực với tốc độ báo cáo cao hơn đáng kể so với các Đơn vị Điều khiển Từ xa SCADA (RTU).
+Có thể hiểu được cách thức thực hiện này thông qua Hình 6, trong đó hiển thị kiến trúc tổng quát của PMU.
+Các đặc tính chính của PMU là xung một lần mỗi giây (1PPS) và mã thời gian được cung cấp bởi bộ đếm thời gian toàn cầu GPS.
+Tín hiệu đồng hồ chính xác với tốc độ 1PPS cung cấp cho mạch lấy mẫu và giữ (S/H) ở tất cả các PMU, cho phép đồng bộ lấy mẫu của các dạng sóng đầu vào trên không chỉ các đầu vào của một PMU mà là đầu vào của tất cả các PMU được lắp đặt trong một hệ thống điện cụ thể.
+Mã thời gian đại diện cho thời gian tuyệt đối cung cấp khả năng đánh dấu thời gian tham chiếu cho tất cả các giá trị phasor được tính toán.
+Các giá trị phasor tính toán như vậy được báo cáo với tốc độ mỗi giây là 30, 60, 120 hoặc thậm chí cao hơn, cung cấp hai lợi thế chính: a) cung cấp dữ liệu chất lượng cao với tốc độ dữ liệu cao, và b) đảm bảo tính toán độ chênh lệch pha chính xác cao giữa bất kỳ vị trí nào trong hệ thống và phasor tham chiếu.
+
+Điều này cung cấp một cơ hội độc đáo cho phân tích sau sự kiện, đặc biệt là trong trường hợp các sự kiện phức tạp (ví dụ: lỗi lan truyền do sự cố thiết bị dẫn đến mất điện).
+Ngoài ra, việc bắt và phân tích các dao động trong hệ thống điện quan trọng vì nó báo hiệu về khả năng sụp đổ của hệ thống [216].
+PMU đóng vai trò quan trọng trong việc ghi nhận các sự cố [217] với sự phát triển gần đây của các công nghệ xác định và phân loại sự cố [217-220].
+Trong những năm gần đây, khả năng của hệ thống điện đã được mở rộng để bao gồm các nguồn năng lượng tái tạo, nhu cầu năng lượng tăng, tích hợp xe điện, vv.
+Tất cả các công nghệ này đều đặt ra những thách thức mới cho hoạt động hệ thống.
+Dữ liệu PMU cung cấp một nguồn thông tin có giá trị có thể giúp đáp ứng các thách thức và tăng cường sự phù hợp của lưới kết hợp.
+
+Do đó, dữ liệu được thu thập từ PMU đóng một vai trò quan trọng trong các ứng dụng như giám sát, điều khiển, bảo vệ, ước lượng trạng thái, đánh giá sự ổn định và phát hiện sự cố [221].
+Tuy nhiên, phân tích sau sự kiện vẫn được thực hiện bằng tay trong nhiều trường hợp với khả năng dự đoán hạn chế hoặc không có.
+Các kỹ thuật được đề xuất trong các giải pháp phần mềm được nhắc đến nhằm mục đích phân tích tự động có khả năng không chỉ phân loại các sự kiện quá khứ mà còn dự đoán những trường hợp sự cố trong tương lai sử dụng các bản ghi dữ liệu trực tiếp.
+
+Phần lớn các phương pháp dự đoán cho phân tích dữ liệu PMU trong văn bản tập trung vào việc cho phép nhận thức tình huống có ý nghĩa hơn so với những gì được bao phủ bởi EMS SCADA, từ đó đảm bảo sự ổn định động của hệ thống [222-229].
+Các phương pháp dự đoán có thể mở rộng khả năng phân biệt các loại sự kiện của các nhà điều hành, từ hoạt động bình thường đến hoạt động trong điều kiện cực đoan bằng cách bắt giữ các đặc trưng dạng sóng PMU và tự động xác định các sự kiện trong quá trình phân loại.
+Điều này cho phép ứng dụng của dữ liệu lớn, trí tuệ nhân tạo và học máy giúp đoán trước nhiều loại cảnh báo và sự cố khẩn cấp trong hệ thống điện, ngoài việc theo dõi hoạt động bình thường và cực đại ổn định động lực.
+Kiến thức ban đầu có thể được sử dụng như hướng dẫn cho nghiên cứu tương lai trong lĩnh vực này và có thể mở cánh cửa cho khám phá sâu hơn của các thuật toán dự đoán dựa trên PMU cho các ứng dụng khác, chẳng hạn như quản lý và giám sát tài sản hoặc dự đoán và quản lý sự cố.
+
+Một thách thức mà tất cả các đơn vị liên quan đến phát triển các ứng dụng trực tuyến và ngoại tuyến đều đang đối mặt là sự gia tăng của một lượng dữ liệu rất lớn đến từ PMUs.
+Để cải thiện tính tin cậy, bảo mật và hiệu quả của hệ thống, cần phát triển các công cụ tự động có khả năng phân tích các sự kiện quá khứ và thông báo quyết định trong thời gian thực.
+Điều này có nghĩa là các phương pháp đề xuất cần phân biệt giữa các kỹ thuật phù hợp để truy cập và xử lý một lượng lớn dữ liệu lịch sử, trong đó thời gian xử lý không quan trọng bằng các kỹ thuật nhằm vào xử lý dữ liệu luồng trong thời gian thực, trong đó tính hiệu quả tính toán là rất quan trọng.
+
+Khi tài nguyên tính toán tiến bộ, việc học và trích xuất các mẫu hữu ích từ dữ liệu lớn tạo ra cơ hội mới.
+Học sâu [88,92,93] cho phép tìm thấy các mẫu trừu tượng hơn từ dữ liệu phức tạp lớn, thậm chí là không đồng nhất [230].
+Một mô hình sâu học một cách đa tầng.
+Mỗi khi dữ liệu mới hoặc tập các đặc trưng được truyền qua một tầng, một cấp độ trừu tượng dữ liệu mới được giới thiệu.
+Do đó, thông qua nhiều tầng liên tiếp, các mô hình sâu học được học các biểu diễn giàu, cấp cao của dữ liệu thô cấp thấp như hình ảnh, âm thanh và văn bản.
+Nói cách khác, các mô hình sâu cung cấp một khung việc toàn diện cho việc trích xuất đặc trưng tự động, phức tạp ở một cấp độ trừu tượng cao.
+Các mô hình sâu không trích xuất các biểu diễn được xác định trước - ngược lại, chúng có xu hướng tìm ra các mẫu không thay đổi bằng cách loại bỏ sự biến động trong dữ liệu [231].
+Giả thuyết cuối cùng trong vấn đề này cho rằng càng nhiều dữ liệu, càng trích xuất được nhiều kiến thức [232], và do đó khả năng tổng quát hóa càng lớn hơn có thể đạt được.
+Các biểu diễn được học là rất gọn nhẹ, giúp giảm thiểu tính toán và từ đó làm cho việc học tiếp theo rất hiệu quả.
+
+Kiến trúc tổng quan của quá trình phân tích dữ liệu PMU tự động được trình bày trong Hình 6.
+
+\subsection{Các ứng dụng cho lưới điện}
+Sự đầu tư vào các công nghệ Lưới điện Thông minh (ví dụ như đồng hồ thông minh, PMU và các thiết bị điện tử thông minh [IED]) cho lưới điện phân phối và truyền tải đang cải thiện khả năng giám sát và điều khiển của chúng [233].
+Đồng thời, các khung pháp lý và thị trường mới đang được thử nghiệm (ở cấp độ thử nghiệm) và triển khai ở các quốc gia khác nhau để hỗ trợ tích hợp các nguồn điện phân tán linh hoạt.
+
+\subsubsection{Tính quan sát}
+Mặc dù có tất cả những tiến bộ công nghệ, khả năng quan sát lưới điện áp thấp (LV) vẫn là một trở ngại chính để khai thác tối đa tiềm năng từ những công nghệ này và tích hợp các nguồn điện phân tán linh hoạt vào thị trường điện và quản lý hệ thống điện.
+Điều này có thể được chia thành hai thách thức chính: (i) đặc trưng của tôpô và các thông số lưới điện; và (ii) giám sát thời gian thực.
+
+Thông tin từ các đồng hồ thông minh được lắp đặt tại khách hàng áp thấp và trạm phân phối cấp hai có thể được sử dụng để xây dựng tôpô lưới sử dụng nhiều phương pháp khác nhau, chẳng hạn như: mô hình đồ thị xác suất và hồi quy tuyến tính LASSO [234, 235]; và luồng điện kết hợp với thông tin chung [151].
+Trong một số trường hợp, thông tin về tôpô lưới có sẵn, nhưng các thông số điện có sai số lớn hoặc không biết (tức là sử dụng các thông số từ các danh mục dây cáp), và do đó, cần thực hiện một ước lượng chắc chắn (và dựa trên dữ liệu) của các thông số lưới [236, 237].
+Vấn đề này cũng yêu cầu phát triển các phương pháp để lập kế hoạch vị trí cảm biến và giảm thiểu chi phí giám sát [238].
+Một nguồn không chắc chắn khác là giai đoạn kết nối của mỗi khách hàng.
+Gom nhóm các phép đo độ lớn điện áp 15 phút từ các đồng hồ thông minh có thể xác định các nhóm khách hàng được kết nối với cùng một giai đoạn và giảm chi phí nhân công cho việc xác định giai đoạn trên thực địa [239].
+Dữ liệu của đồng hồ thông minh cũng có thể được sử dụng để ước tính công suất định mức của các tấm pin năng lượng mặt trời cài đặt sau đồng hồ [240] và phát hiện xe điện đang sạc [241].
+
+Về mặt giám sát thời gian thực, thông tin về hồ sơ điện áp là rất quan trọng để tạo ra các cảnh báo cho các nhân viên vận hành tại các trung tâm phân phối điện áp thấp.
+Tuy nhiên, ngay cả với các giao thức truyền thông như Power Line Communication (PLC) PRIME hoặc General Packet Radio Service (GRPS), việc thu thập thông tin 15 phút một lần từ hàng trăm khách hàng là không khả thi hoặc chi phí truyền thông rất cao.
+Các chức năng ước tính trạng thái dựa trên dữ liệu giúp ước tính độ lớn điện áp tại mỗi nút điện áp thấp trong thời gian thực chỉ bằng cách sử dụng thông tin từ một tập hợp con các đồng hồ thông minh (tức là 10-20\%) có khả năng truyền thông thời gian thực [242].
+Dự báo tải và sản xuất có thể được sử dụng như đầu vào trong thuật toán ước tính trạng thái tiêu chuẩn và tạo ra một ước tính xác suất của điện áp nút [243].
+
+Các mức điện áp khác cũng được hưởng lợi từ các chức năng ước tính trạng thái dựa trên dữ liệu để giám sát thời gian thực.
+Một số ví dụ bao gồm: một chức năng ước tính trạng thái dựa trên dữ liệu mạnh mẽ được đề xuất trong [244] để khai thác dữ liệu lịch sử được thu thập bởi các đồng hồ đo phân phối và PMU, bằng cách tìm các đặc trưng tương đồng trong một khung làm việc học có giám sát với hồi quy kernel ridge; một bộ ước tính trạng thái động có khả năng dự báo được mô tả trong [245]; kết hợp dữ liệu SCADA từ các trạm phân phối chính (thông thường có sẵn trong cơ sở dữ liệu trung tâm với độ trễ vài giây/phút) và trạm phân phối phụ (thông thường có sẵn trong cơ sở dữ liệu trung tâm với độ trễ một vài giờ) có thể suy ra điều kiện vận hành gần thời gian thực của lưới điện áp trung bình (MV) [246,247]; và một khối ghép của các bộ tự mã hóa cạnh tranh địa phương ước tính trạng thái của các công tắc lưới dựa trên một tập hợp các đo lường điện cục bộ [248].
+Hơn nữa, các chức năng truyền thống, chẳng hạn như phát hiện cục bộ, cũng có thể được hưởng lợi từ các chức năng phân tích dữ liệu tích hợp thông tin văn bản, thời gian và không gian từ các phương tiện truyền thông xã hội [21] và dữ liệu PMU [218].
+
+\subsubsection{Hộ trợ ra quyết định và điều khiển}
+Các chức năng dựa trên dữ liệu cũng đang được tích hợp vào các hệ thống quản lý năng lượng truyền thống (EMS) và quản lý phân phối (DMS) để bổ sung hoặc thay thế các chức năng phân tích lưới điện cổ điển như lưu lượng điện, cấu hình mạng, lưu lượng điện tối ưu, vv.
+Trong một số trường hợp, một mô hình học máy được sử dụng như là một proxy cho các chức năng truyền thống, ví dụ như ước tính lưu lượng điện của ma trận Jacobian với dữ liệu PMU [249] và tính toán cam kết đơn vị [250].
+Một điểm bất lợi tiềm tàng của phương pháp này là cần thu thập hoặc mô phỏng một bộ dữ liệu lớn để vừa với các mô hình.
+Tuy nhiên, sau khi được vừa, nó có thể được tích hợp vào các ứng dụng khác nhau và cung cấp các ước tính nhanh chóng.
+
+Trong những trường hợp khác, học máy được sử dụng để học từ dữ liệu lịch sử về các hành động điều khiển của các nhà điều hành và hỗ trợ ra quyết định [251] hoặc để khám phá các giải pháp mới (và tốt hơn) bằng cách sử dụng các hệ thống chuyên gia [252].
+Một lợi thế quan trọng của hai phương pháp này, được phát triển cho Chủ sở hữu Hệ thống Truyền tải (TSOs), là tính giải thích cao đối với những người ra quyết định và khả năng khai thác kiến thức chuyên gia (tức là quyết định quá khứ và các heuristics điều khiển).
+Tuy nhiên, cả hai đều thuộc mô hình học bằng mô phỏng, tức là học có giám sát được áp dụng cho các quyết định từ chuyên gia, và không khám phá (hoặc tìm kiếm) các giải pháp mới.
+Một lựa chọn khác là học tăng cường, cung cấp một sự đánh đổi giữa khai thác và khám phá và trong một số trường hợp, như trường hợp được mô tả trong [253], có thể kết hợp với học bằng mô phỏng.
+
+Một số trường hợp sử dụng cho phương pháp điều khiển dữ liệu này là: xếp hạng nhanh các trường hợp dự phòng cấp cao hơn (theo nguy cơ của chúng) để ưu tiên mô phỏng hệ thống điện năng [254]; và phân tích nhân quả trên dữ liệu đo lường để thực hiện các chiến lược tấn công node tối ưu [255].
+
+Cuối cùng, việc tích hợp các hệ thống năng lượng tái tạo và sự áp dụng rộng rãi của IEC 61850 đang tăng thêm khối lượng dữ liệu cảnh báo và số lượng cảnh báo đòi hỏi sự chú ý và hành động điều khiển trong phòng điều khiển [256].
+Các phương pháp truyền thống dựa trên hệ thống chuyên gia dựa trên quy tắc [257], nhưng thiếu khả năng cung cấp những thông tin bổ ích về các thông tin chứa trong chuỗi cảnh báo và giảm tải trí nhớ của nhân viên điều khiển.
+Trong [258], một kỹ thuật phân loại thô không giám sát được đề xuất để giảm khối lượng dữ liệu và thông báo nhận được tại các trạm biến áp trong các tình huống khẩn cấp và cải thiện quyết định, nhưng không cung cấp đề xuất về hành động điều khiển.
+Bước quyết định cuối cùng này (tức là xác định một chuỗi các thao tác thủ công bởi người điều khiển) thiếu trong hầu hết các phương pháp xử lý dữ liệu từ các hệ thống bảo vệ và là một yếu tố cốt lõi để tạo ra tác động thực sự từ các chức năng phân tích dữ liệu, chẳng hạn như giảm thời gian để đưa ra quyết định đầu tiên và chỉ số thời gian gián đoạn trung bình của hệ thống (SAIDI).
+
+\section{Cơ hội và thách thức trong tương lai}
+
+Cuộc khảo sát cho thấy có nhiều ứng dụng đang ở các giai đoạn thực hiện thực tế khác nhau.
+
+Trong khi số lượng tài liệu tham khảo về chủ đề tổng quan về dữ liệu lớn đã rất lớn, dự kiến công việc sẽ tiến triển với tốc độ chưa từng có, và số lượng tài liệu tham khảo sẽ tiếp tục tăng trong tương lai gần.
+
+Tuy nhiên, trong một số cuộc khảo sát đã công bố, không rõ ứng dụng cuối cùng của một số kỹ thuật được khảo sát là gì, và lợi ích và động lực kinh doanh liên quan là gì.
+Chúng tôi nhận thấy lợi ích của nghiên cứu tương lai là tập trung vào các ứng dụng cuối cùng trong khi đồng thời giải quyết các vấn đề cơ bản để thúc đẩy những cơ hội sau đây cuối cùng:
+
+\begin{itemize}
+\item Dự đoán các sự kiện trước thời gian và cho phép triển khai các chiến lược giảm thiểu rủi ro và tính toán rủi ro với dự báo không chắc chắn.
+
+\item Tối ưu hóa dữ liệu lịch sử cho lợi ích của chủ sở hữu và người sử dụng.
+Kết hợp mô hình vật lý và dữ liệu để cải thiện phân tích nguyên nhân gốc rễ.
+Ngăn chặn các sự cố và hạn chế vận hành hệ thống điện hiện tại gây tổn thất hàng tỷ đô la.
+
+\item Thông báo cho người dùng về những hạn chế cung cấp điện sắp tới và sử dụng các nguồn lực phân tán hiệu quả hơn để giảm thiểu sự cố và những tình huống khẩn cấp khác.
+
+\item Thúc đẩy sự đổi mới bằng cách sử dụng kinh nghiệm đa ngành từ các lĩnh vực khác nhau nhưng có mối liên hệ mạnh về thuộc tính dữ liệu và yêu cầu phân tích.
+
+\item Sử dụng dữ liệu từ nhiều nguồn phổ biến khác nhau như máy tính bảng, điện thoại thông minh và các thiết bị điện tử cá nhân khác, khiến chủ sở hữu và các bên liên quan của doanh nghiệp phân tích dữ liệu trở thành các nhân vật tham gia.
+
+\item Thay thế một số nhiệm vụ được thực hiện bởi các nhà điều hành và chuyên gia ngày nay bằng các thuật toán và quy trình tự động hóa.
+\end{itemize}
+
+\section{Kết luận}
+
+Bản khảo sát này đã cố gắng đạt được một số mục tiêu sau:
+\begin{itemize}
+\item Nó đã giải quyết được sự đa dạng và chiều sâu của các ứng dụng thực tiễn về phân tích dữ liệu lớn trong lưới điện.
+\item Nó đã đánh giá lại các vấn đề chính trong việc triển khai phân tích dữ liệu lớn trong các lĩnh vực lưới được chọn.
+\item Nó cung cấp nhiều ví dụ về các xu hướng mới nhất trong khung quyết định và phân tích dự đoán.
+\item Nó liệt kê một số tài liệu tham khảo có thể được sử dụng bởi các nhà nghiên cứu để nghiên cứu của họ tiếp tục phát triển.
+\item Nó đưa ra một số hướng đi của nghiên cứu tương lai và phác thảo các thách thức và cơ hội.
+
+\end{itemize}
+
+Bài báo này không phải là một tài liệu toàn diện về tất cả các tác phẩm hoặc xu hướng đã được xuất bản.
+Nếu quan tâm đến chủ đề, người đọc của cuộc khảo sát này nên khám phá các khía cạnh bổ sung không được đề cập ở đây.
 

### 2023-03-21 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update citations of the paper

@@
 
-Như chúng ta đã biết, lưới điện thông minh ngày này đã và đang tiếp tục được trang bị các thiết bị có khả năng cảm biến (sensing) và thu thập dữ liệu với độ phân giải cao hơn nhiều so với trước đây.
-Dữ liệu mới kèm theo các phương pháp phân tích mới có thể hỗ trợ các mục tiêu trong quản lý hệ thống điện như việc đạt được độ bền cao hơn, hiệu quả kinh tế tốt hơn và giảm khí thải.
+Như chúng ta đã biết, lưới điện thông minh ngày nay đã và đang tiếp tục được trang bị các thiết bị tiên tiến có khả năng cảm biến (sensing) và thu thập dữ liệu với độ phân giải cao hơn nhiều so với trước đây.
+Dữ liệu mới kèm theo các phương pháp phân tích mới có thể hỗ trợ các mục tiêu trong quản lý hệ thống điện như việc đạt được độ bền thiết bị cao hơn, hiệu quả kinh tế tốt hơn và giảm khí thải.
 
-Phân tích dữ liệu vì thế trở thành một trong những yếu tố quan trọng giúp phần đảm bảo và nâng cao các công nghệ hiện có nhằm đưa ra các quyết định tối ưu hơn.
-Có thể nói, công việc phân tích dữ liệu đã và đang trở thành khả năng cơ bản và lợi thế chiến lượng cho các tổ chức muốn đổi mới và luôn hướng tới việc cung cấp dịch vụ chất lượng tốt hơn cho khách hàng.
+Phân tích dữ liệu vì thế trở thành một phần quan trọng giúp doanh nghiệp nâng cao khả năng ứng dụng các công nghệ hiện có nhằm đưa ra các quyết định tối ưu hơn.
 Các ứng dụng của phân tích dữ liệu cho ngành điện là rất nhiều và có thể được xác định trong nhiều hoạt động của ngành.
@@
 
-\subsection{Các Nguồn Dữ Liệu Mới}
-Việc áp dụng các giải pháp phân tích dữ liệu lớn trong các hệ thống phân phối và truyền tải điện khác nhau thường được tập trung vào việc khai thác các nguồn dữ liệu hỗn hợp (heterogeneous), là các khối dữ liệu có chất lượng riêng biệt, độ phân giải không gian/và thời gian khách nhau, và có cách trình bày thông tin khác nhau.
-Sẽ là khả thi để nâng cao ứng dụng nguồn dữ liệu này thông qua các phương pháp như: (a) kết hợp các nguồn dữ liệu mới và nguồn dữ liệu truyền thống, ví dụ thông qua việc sử dụng lý thuyết hợp nhất dữ liệu (data fusion) \citep{SimoesCosta2013}; (b) trích xuất và kết hợp thông tin từ các dữ liệu có định dạng khác nhau (ví dụ như dữ liệu ở hạng ảnh và ở dạng văn bản) thông qua các mô hình đã chiều (multimodal learning) \citep{Srivastava2012} hoặc qua một mạng lưới thông tin hỗ hợp (heterogeneous) \citep{Sun2016}; và (c) kết hợp dữ liệu từ các nguồn dữ liệu phân tán về mặt địa lý, ví dụ như việc sử dụng các phương pháp vector autoregressive (VAR) truyền thống \citep{Cavalcante2017} hay phương pháp học sâu (deep learning) \citep{Zhu2020}.
+\subsection{Nguồn Dữ Liệu}
+Việc áp dụng các giải pháp phân tích dữ liệu lớn trong các hệ thống phân phối và truyền tải điện khác nhau thường được tập trung vào việc khai thác các nguồn dữ liệu không có tính đồng nhất (heterogeneous), là các khối dữ liệu có chất lượng riêng biệt, độ phân giải không gian/và thời gian khách nhau, và có cách trình bày thông tin khác nhau.
+Sẽ là khả thi để nâng cao ứng dụng nguồn dữ liệu này thông qua các phương pháp như: (a) kết hợp các nguồn dữ liệu mới và nguồn dữ liệu truyền thống, ví dụ thông qua việc sử dụng lý thuyết hợp nhất dữ liệu (data fusion) \citep{SimoesCosta2013}; (b) trích xuất và kết hợp thông tin từ các dữ liệu có định dạng khác nhau (ví dụ như dữ liệu ở hạng ảnh và ở dạng văn bản) thông qua các mô hình đa chiều (multimodal learning) \citep{Srivastava2012} hoặc qua một mạng lưới thông tin có tính không đồng nhất (heterogeneous) \citep{Sun2016}; và (c) kết hợp dữ liệu từ các nguồn dữ liệu phân tán về mặt địa lý, ví dụ như việc sử dụng các phương pháp Vector Autoregressive (VAR) truyền thống \citep{Cavalcante2017} hay phương pháp học sâu (Deep Learning) \citep{Zhu2020}.
 
-Các nguồn dữ liệu mới đến từ các lĩnh vực khác nhau:
+Nguồn dữ liệu đến từ các lĩnh vực khác nhau:
 
 \begin{itemize}
-	\item \textbf{Hạ tầng mạng lưới điện}: Các cơ quan quản lý và điều hành hệ thống đang cải thiện và nâng cao khả năng quan sát của hệ thống mạng bằng cách lắp đặt các đơn vị đo pha (PMU) có thể cung cấp dữ liệu với tốc độ báo cáo cao (ví dụ, 30 đơn vị đo lường trên mỗi giây của độ lớn điện áp/dòng điện, pha và tần số) và các thiết bị đầu cuối điểm từ xa (Remote Terminal Units - RTU) tại các trạm biến áp và các công tơ điện thông minh lắp đặt cho khách hàng.
+	\item \textbf{Hạ tầng mạng lưới điện}: Các doanh nghiệp quản lý và điều hành hệ thống đang cải thiện và nâng cao khả năng quan sát của hệ thống mạng bằng cách lắp đặt các đơn vị đo pha (PMU) có thể cung cấp dữ liệu với tốc độ báo cáo cao (ví dụ, 30 đơn vị đo lường trên mỗi giây của độ lớn điện áp/dòng điện, pha và tần số) và các thiết bị đầu cuối điểm từ xa (Remote Terminal Units - RTU) tại các trạm biến áp và các công tơ điện thông minh lắp đặt cho khách hàng.
 Các cảm biến được dùng cho việc giám sát từ xa các trạm biến áp cũng đang được thử nghiệm để theo dõi tình trạng làm việc của thiết bị và cải thiện chất lượng dịch vụ \citep{Leitao2015}.
 
-	\item \textbf{Các nhà máy điện tái tạo}: Công nghiệp năng lượng tái tạo đang lắp đặt và vận hành các cảm biến giám sát tại cấp độ cánh quạt điện gió và tấm năng lượng mặt trời, tạo ra một lượng lớn dữ liệu (ví dụ, một cánh quạt điện gió có thể có hơn 100 cảm biến bên trong cánh quạt, thu thập hơn 10.000 điểm dữ liệu mỗi giây) có thể được sử dụng cho việc bảo trì dự đoán (và giảm chi phí Vận hành và Bảo trì); dữ liệu từ một lưới dự báo thời tiết số, các cảm biến phân tán địa lý (ví dụ, cánh quạt điện gió, các cảm biến đo độ chiếu sáng), máy ảnh ghi lại bầu trời và hình ảnh vệ tinh có thể được kết hợp để cải thiện khả năng dự báo công suất (và thời tiết) trong nhiều khung thời gian khác nhau \citep{Sweeney2020}.
-Trong việc dự báo điện năng tái tạo, quy mô của các nghiên cứu cũng đã tăng từ một địa điểm đến hơn 100 địa điểm \citep{Messner2019}.
+	\item \textbf{Nhà máy điện tái tạo}: Các nhà máy sản xuất điện năng lượng tái tạo đang tiến hành lắp đặt và vận hành các cảm biến giám sát trong turbin và trong các cánh quạt (có những trường hợp lên đến hàng trăm cảm biến), các cảm biến này thu nhập hơn 10.000 điểm dữ liệu mỗi giây).
+Các dữ liệu này có thể được dùng cho công tác dự đoán sự cố và bảo trì thiết bị (giảm chi phí vận hành và bảo trì; dữ liệu từ mạng lưới dự báo thời tiết số, các cảm biến phân tán địa lý (ví dụ, cánh quạt điện gió, các cảm biến đo độ chiếu sáng), máy ảnh ghi lại bầu trời và hình ảnh vệ tinh có thể được kết hợp để cải thiện khả năng dự báo công suất (và thời tiết) trong nhiều khung thời gian khác nhau \citep{Sweeney2020}.
+Trong việc dự báo điện năng tái tạo, quy mô kết nối và sử dụng dữ liệu cũng đã tăng từ một địa điểm đến hàng trăm địa điểm \citep{Messner2019}.
 
 	\item \textbf{Người tiêu dùng và mạng xã hội}: trong khi đang ở giai đoạn đầu của triển khai, sự lan rộng của các thiết bị liên kết Internet trong các căn hộ và tòa nhà thông minh đang tạo điều kiện cho các dịch vụ về năng lượng và phi năng lượng dựa trên dữ liệu \citep{Ahmed2016}, ảnh hưởng của chúng phụ thuộc vào việc giải quyết các thách thức như bảo vệ/quyền riêng tư dữ liệu và tương tác với người tiêu dùng.
-Hơn nữa, sự gia tăng của mạng xã hội đã cho phép các công ty điện lực hiểu và tương tác với khách hàng tốt hơn bao giờ hết \citep{MorenoMunoz2016}.
+Hơn nữa, sự gia tăng của mạng xã hội đã cho phép các công ty điện lực hiểu và tương tác với khách hàng tốt hơn trước rất nhiều\citep{MorenoMunoz2016}.
 Các nhà nghiên cứu cũng đã thử kết hợp dữ liệu Twitter vào việc phát hiện cục bộ mất điện \citep{Sun2016}.
 
-	\item \textbf{Thị trường điện}: Trong vài năm qua, đặc biệt tại châu Âu, tính minh bạch của thị trường điện đã được cải thiện đáng kể, và sau khi xuất bản Quy định (EU) số 5 43/2013 [22], lượng dữ liệu công khai có sẵn đang tăng lên [23], bao gồm cả truy cập vào các đề xuất riêng lẻ từ các công ty cung cấp điện trên thị trường (thông thường có sẵn với độ trễ vài tháng).
+	\item \textbf{Thị trường điện}: Trong vài năm qua, đặc biệt tại châu Âu, tính minh bạch của thị trường điện đã được cải thiện đáng kể, và sau khi xuất bản Quy định (EU) số 543/2013 \citep{EU2013}, lượng dữ liệu công khai có sẵn đang tăng lên \citep{Hirth2018}, bao gồm cả truy cập vào các đề xuất riêng lẻ từ các công ty cung cấp điện trên thị trường (thông thường có sẵn với độ trễ vài tháng).
 Cùng xu hướng đó đang diễn ra ở Mỹ, với các nền tảng như bộ sưu tập dữ liệu Form EIA-930 cung cấp nguồn dữ liệu toàn diện và tập trung cho các dữ liệu hoạt động hàng giờ của lưới điện cao áp tại 48 tiểu bang.
-Dữ liệu mở này có thể được sử dụng cho các mục đích khác nhau: cải thiện kỹ năng dự báo giá điện bằng cách kết hợp phân tích giá điện từ các khu vực khác nhau [24] hoặc đánh giá ảnh hưởng quy mô lớn của việc phát điện từ nguồn năng lượng tái tạo được cấp bởi lưới điện liên kết với các quốc gia láng giềng [25].
+Dữ liệu mở này có thể được sử dụng cho các mục đích khác nhau: cải thiện kỹ năng dự báo giá điện bằng cách kết hợp phân tích giá điện từ các khu vực khác nhau \citep{Ziel2018} hoặc đánh giá ảnh hưởng quy mô lớn của việc phát điện từ nguồn năng lượng tái tạo được cấp bởi lưới điện liên kết với các quốc gia láng giềng \citep{Zugno2013}.
 
-	\item \textbf{Lĩnh vực môi trường và thời tiết}: Dữ liệu thời tiết đóng vai trò quan trọng trong việc dự đoán điều kiện hoạt động, bao gồm cả lỗi hỏng hóc thiết bị và đường dây.
-Dữ liệu từ các trạm thời tiết trên mặt đất [26], vệ tinh [27] và tài nguyên radar [28] có sẵn từ cơ sở dữ liệu của chính phủ.
-Mạng lưới cảm biến chuyên dụng, chẳng hạn như mạng lưới phát hiện sét quốc gia ở Mỹ [29], cũng là nguồn cung cấp dữ liệu thời tiết khá hữu ích.
-Nhiều dịch vụ dự báo thời tiết cũng có sẵn để cung cấp các tính năng được tính trước của các tập dữ liệu thời tiết [30].
-Ngoài ra, dữ liệu về thực vật, đất đai, di cư động vật và các điều kiện xung quanh khác cũng có thể có sẵn từ nhiều nguồn khác nhau [31].
-Cách sử dụng dữ liệu chính xác cao bằng cách sử dụng các cơ sở dữ liệu chuyên dụng như hệ thống (LIDAR - Light Detection and Ranging) hoặc khảo sát bằng thiết bị bay drone cũng được nghiên cứu và đưa vào thực tiễn [32].
+	\item \textbf{Môi trường và thời tiết}: Dữ liệu thời tiết đóng vai trò quan trọng trong việc dự đoán điều kiện hoạt động, bao gồm cả lỗi hỏng hóc thiết bị và đường dây.
+Dữ liệu từ các trạm thời tiết trên mặt đất, vệ tinh và tài nguyên radar có sẵn từ cơ sở dữ liệu của chính phủ.
+Mạng lưới cảm biến chuyên dụng, chẳng hạn như mạng lưới phát hiện sét quốc gia ở Mỹ, cũng là nguồn cung cấp dữ liệu thời tiết khá hữu ích.
+Nhiều dịch vụ dự báo thời tiết cũng có sẵn để cung cấp các tính năng được tính trước của các tập dữ liệu thời tiết.
+Ngoài ra, dữ liệu về rừng và nông nghiệp, đất đai, di cư động vật và các điều kiện xung quanh khác cũng có thể có sẵn từ nhiều nguồn khác nhau.
+Cách sử dụng dữ liệu chính xác cao bằng cách sử dụng các cơ sở dữ liệu chuyên dụng như hệ thống (LIDAR - Light Detection and Ranging) hoặc khảo sát bằng thiết bị bay drone cũng được nghiên cứu và đưa vào thực tiễn.
 Những dữ liệu này thường không được thu thập trong phạm vi chức năng của ngành điện nhưng chúng lại là một nguồn dữ liệu quan trọng cho ngành này.
-Chẳng hạn, trong dự báo tải, các nghiên cứu đã chuyển từ việc sử dụng nhiệt độ được thu thập tại một trạm đến nhiều biến thời tiết và nhiều trạm thời tiết [33-35].
-Trong dự báo điện mặt trời, dữ liệu hình ảnh bầu trời được sử dụng nhiều để phát hiện đám mây [36].
+Chẳng hạn, trong dự báo tải, các nghiên cứu đã chuyển từ việc sử dụng nhiệt độ được thu thập tại một trạm đến nhiều biến thời tiết và nhiều trạm thời tiết \citep{Hong2015}.
+Trong dự báo điện mặt trời, dữ liệu hình ảnh bầu trời được sử dụng nhiều để phát hiện đám mây \citep{Kleissl2013}.
 \end{itemize}
 
-Các nhà nghiên cứu và chuyên gia hiện nay tập trung vào việc khai thác dữ liệu hiện có và khám phá các nguồn dữ liệu mới nổi và dữ liệu với quy mô lớn để theo đuổi các cải tiến trong kế hoạch và vận hành lưới điện.
-Có nhiều khía cạnh quan trọng khác của phân tích dữ liệu lớn, chẳng hạn như xây dựng thuật toán có thể tận dụng môi trường tính toán hiệu suất cao và mở rộng kích thước mô hình để nắm bắt các đặc trưng chi tiết trong dữ liệu [37].
-Một ví dụ khác là sử dụng dữ liệu tải và thời tiết hàng giờ để dự báo tải dài hạn, thông thường dựa trên dữ liệu hàng tháng [38,39].
-Một bài đánh giá gần đây về phân tích dữ liệu lấy từ các công tơ thông minh liệt kê 10 bộ dữ liệu công khai về nhu cầu điện [40].
+Các nhà nghiên cứu và chuyên gia hiện nay tập trung vào việc khai thác dữ liệu hiện có và khám phá các nguồn dữ liệu mới và dữ liệu với quy mô lớn để theo đuổi các cải tiến trong lập kế hoạch và vận hành lưới điện.
+Có nhiều khía cạnh quan trọng khác của phân tích dữ liệu lớn, chẳng hạn như xây dựng thuật toán có thể tận dụng môi trường tính toán hiệu suất cao và mở rộng kích thước mô hình để nắm bắt các đặc trưng chi tiết trong dữ liệu \citep{Wang2016}.
+Một ví dụ khác là sử dụng dữ liệu tải và thời tiết hàng giờ để dự báo tải dài hạn, thông thường dựa trên dữ liệu hàng tháng \citep{Hyndman2009,Hong2014}.
+Một bài đánh giá gần đây về phân tích dữ liệu lấy từ các công tơ thông minh liệt kê 10 bộ dữ liệu công khai về nhu cầu điện \citep{Wang2018}.
 Hình \ref{fig:1} là một ví dụ mô tả các thuộc tính cơ bản của dữ liệu lớn trong ngành điện.
@@
 
-\subsection{Các Cân Nhắc Quan Trọng Khi Tạo Ra Các Tập Dữ Liệu}
+\subsection{Một Số Lưu Ý Quan Trọng Khi Tạo Ra Các Tập Dữ Liệu}
 
@@
 Mục tiêu của khoa học dữ liệu là trích xuất giá trị từ dữ liệu.
-Các bước của vòng đời quản lý dữ liệu bao gồm thu thập dữ liệu; tiền xử lý (khám phá, lấy mẫu, giảm chiều/thu gọn các đặc trưng và thuộc tính, tạo đặc trưng và thuộc tính, biến đổi, làm sạch và tích hợp); xử lý phân tích (mô hình hóa, thường bao gồm nhiều khối mô hình xây dựng liên kết với nhau); giải thích và báo cáo kết quả [61].
-Những kỹ năng chính cần thiết trong lĩnh vực này thường được xem là đa ngành ở giao điểm của khoa học máy tính, toán học, thống kê và các lãnh vực khoa học kĩ thuật và xã hội khác.
-Trên mặt kỹ thuật, các thách thức chính thường liên quan đến dữ liệu lớn, trí thông minh nhân tạo và các phương pháp học máy, trong khi quá trình khoa học dữ liệu áp dụng cũng có thể yêu cầu các kỹ năng khoa học xã hội, giao tiếp và kinh doanh.
+Các bước của vòng đời quản lý dữ liệu bao gồm thu thập dữ liệu; tiền xử lý (khám phá, lấy mẫu, giảm chiều/thu gọn các đặc trưng và thuộc tính, tạo đặc trưng và thuộc tính, biến đổi, làm sạch và tích hợp); xử lý phân tích (mô hình hóa, thường bao gồm nhiều khối mô hình xây dựng liên kết với nhau); giải thích và báo cáo kết quả \citep{Aggarwal2015}.
+Những kỹ năng chính cần thiết trong lĩnh vực này thường được xem là đa ngành ở giao điểm của khoa học máy tính, toán học, thống kê và các lãnh vực khoa học kĩ thuật và xã hội khác (Hình \ref{fig:khoahocdulieu01}).
+Trên mặt kỹ thuật, các thách thức chính thường liên quan đến dữ liệu lớn, trí thông minh nhân tạo và các phương pháp học máy, trong khi quá trình khoa học dữ liệu áp dụng cũng có thể yêu cầu các kỹ năng khoa học xã hội, giao tiếp và kinh doanh (Hình \ref{fig:khoahocdulieu02}).
 
-Một cái nhìn toàn diện về khoa học dữ liệu nhấn mạnh rằng khoa học dữ liệu là "hơn là sự kết hợp của thống kê và khoa học máy tính" vì "nó yêu cầu đào tạo làm sao để tổng hợp và kết nối giữa các kĩ thuật thống kê và sức mạnh của máy tính trong một khung làm việc lớn hơn, giải quyết từng vấn đề một cũng như giải quyết các câu hỏi cụ thể trong từng lĩnh vực".
+\begin{figure}[h!]
+	\centering
+	\begin{center}
+		\includegraphics[width=8cm]{khoahocdulieu01}% This is a *.eps file
+	\end{center}
+	\caption{Giao thoa các mảng kiến thức - khoa học dữ liệu}\label{fig:khoahocdulieu01}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\begin{center}
+		\includegraphics[width=\textwidth]{khoahocdulieu02}% This is a *.eps file
+	\end{center}
+	\caption{Các kĩ năng giao thoa - khoa học dữ liệu}\label{fig:khoahocdulieu02}
+\end{figure}
+
+Một cái nhìn toàn diện về khoa học dữ liệu nhấn mạnh rằng "khoa học dữ liệu hơn là sự kết hợp của thống kê và khoa học máy tính" vì "nó yêu cầu đào tạo làm sao để tổng hợp và kết nối giữa các kĩ thuật thống kê và sức mạnh của máy tính trong một khung làm việc lớn hơn, giải quyết từng vấn đề một cũng như giải quyết các câu hỏi cụ thể trong từng lĩnh vực".
 Các nhà nghiên cứu cũng có cùng quan điểm cho rằng ngành khoa học dữ liệu đòi hỏi: (1) hiểu về bối cảnh của dữ liệu, (2) đánh giá các trách nhiệm liên quan đến việc sử dụng dữ liệu riêng tư và công khai; và (3) giao tiếp rõ ràng về những gì có thể và không thể được suy ra từ một bộ dữ liệu.
@@
 Các thành phần cốt lõi của khoa học dữ liệu là các phương pháp dựa trên học máy (Machine Learning - ML) để tìm kiếm các khuôn mẫu (pattern) trong dữ liệu và dựa vào đó để cung cấp thông tin về hiện tượng được mô tả bởi dữ liệu và các dự đoán liên quan đến các sự kiện trong tương lai.
-Trong học máy , mục tiêu là học/tạo ra một hàm (ánh xạ) mô tả và chứa đựng các dự liệu đầu vào (thường gọi là các biến giải thích - explanatory variable) với tham số đầu ra quan sát được (gọi là biến phản ứng).
+Trong học máy, mục tiêu là học/tạo ra một hàm (ánh xạ) mô tả và chứa đựng các dự liệu đầu vào (thường gọi là các biến giải thích - explanatory variable) với tham số đầu ra quan sát được (gọi là biến phản ứng - response variable).
 Một biểu diễn đơn giản hóa của thực tế được tạo ra cho mục đích này, gọi là mô hình, được sử dụng để ước tính phản ứng chưa biết cho các trường hợp mới dựa trên các biến giải thích quan sát được quan tâm, và quá trình này được gọi là suy luận hoặc đơn giản hơn là dự đoán.
@@
 Các nhiệm vụ mô tả thường là không giám sát, có nghĩa là chỉ các biến giải thích được xem xét trong phân tích.
-Các mục tiêu mô tả phổ biến bao gồm phân cụm dữ liệu [64,65], khám phá mối liên hệ [66], và phát hiện sự khác biệt so với hành vi bình thường, bao gồm phân tích giá trị cực đại, phát hiện giá trị ngoại lệ và xác định các mô hình mới nổi [67].
+Các mục tiêu mô tả phổ biến bao gồm phân cụm dữ liệu \citep{Albarakati2019, Gligorijevic2016}, khám phá mối liên hệ \citep{Gligorijevic2016a}, và phát hiện sự khác biệt so với hành vi bình thường, bao gồm phân tích giá trị cực đại, phát hiện giá trị ngoại vi và xác định các mô hình mới \citep{Tan2016}.
 Các nhiệm vụ dự đoán là giám sát, có nghĩa là chúng không chỉ yêu cầu các biến giải thích mà còn giá trị của biến phụ thuộc đang được dự đoán.
-Các ví dụ thực tế bao gồm đánh giá rủi ro [68] và chẩn đoán bệnh [69].
+Các ví dụ thực tế bao gồm đánh giá rủi ro và chẩn đoán bệnh \citep{Ghalwash2013}.
 
-Có các phương pháp bán giám sát [70] và tự đào tạo [71], trong đó dữ liệu huấn luyện bao gồm một số dữ liệu có nhãn và nhiều dữ liệu không có nhãn.
+Có các phương pháp bán giám sát và tự đào tạo, trong đó dữ liệu huấn luyện bao gồm một số dữ liệu có nhãn và nhiều dữ liệu không có nhãn \citep{Triguero2015}.
 
@@
 Một trong những phương pháp thường được sử dụng cho phân loại là cây quyết định (decision tree).
-Thuật toán Hunt [67], một trong những phương pháp cây quyết định sớm nhất, đề xuất quy trình tổng quát của việc phân chia dữ liệu dựa trên giá trị của một thuộc tính duy nhất và tiếp tục đệ quy trên các tập con nếu lớp không đủ thuần khiết ở các tập con.
-Nhiều phương pháp đã được đề xuất để đo độ không thuần khiết của tập dữ liệu và xác định phân nhánh tiếp theo (ví dụ: entropy trong CART [72] hoặc chỉ số Gini trong ID3 [73] và C4.5 [74]), cũng như để tỉa cây, từ đó cải thiện khả năng tổng quát của mô hình.
-Cây quyết định dễ hiểu, rất rẻ để xây dựng và rất nhanh trong việc phân loại các trường hợp không rõ ràng.
+Thuật toán Hunt \citep{Tan2016}, một trong những phương pháp cây quyết định sớm nhất, đề xuất quy trình tổng quát của việc phân chia dữ liệu dựa trên giá trị của một thuộc tính duy nhất và tiếp tục đệ quy trên các tập con nếu lớp không đủ thuần khiết ở các tập con.
+Nhiều phương pháp đã được đề xuất để đo độ không thuần khiết của tập dữ liệu và xác định phân nhánh tiếp theo (ví dụ: entropy trong CART \citep{Breiman2017} hoặc chỉ số Gini trong ID3 \citep{Quinlan1986} và C4.5 \citep{Quinlan2014}), cũng như để tỉa cây, từ đó cải thiện khả năng tổng quát của mô hình.
+Cây quyết định dễ hiểu và không tốn kém nguồn lực để xây dựng và rất nhanh trong việc phân loại các trường hợp không rõ ràng.
 Chúng cũng khá bền với nhiễu và có thể xử lý các thuộc tính trùng lặp hoặc không liên quan, nhưng không tính đến sự tương tác giữa các thuộc tính.
-Một trong những hạn chế của cây quyết định là chúng yêu cầu tỉa, nếu không chúng sẽ trở nên quá lớn dẫn đến các vấn đề quá khớp.
-Giới hạn này được giải quyết thành công bằng Random Forests, được xây dựng như một tập hợp các cây quyết định không tương quan [75,76].
-Tập hợp này được lấy cảm hứng từ phương pháp Bagging, một phương pháp tổng hợp dựa trên lấy mẫu bootstrap, được phát triển để giảm phương sai mà không làm tăng sai số [77].
+Một trong những hạn chế của cây quyết định là chúng yêu cầu tỉa, nếu không chúng sẽ trở nên quá lớn dẫn đến các vấn đề kĩ thuật như quá khớp (overfit).
+Giới hạn này được giải quyết thành công bằng Random Forests, được xây dựng như một tập hợp các cây quyết định không tương quan \citep{Breiman2001}.
+Tập hợp này được lấy cảm hứng từ phương pháp Bagging, một phương pháp tổng hợp dựa trên lấy mẫu bootstrap, được phát triển để giảm phương sai mà không làm tăng sai số \citep{Breiman1996}.
 Trong Random Forests, ý tưởng này được mở rộng hơn nữa bằng cách giới hạn mỗi nút chỉ xem xét một tập con ngẫu nhiên nhỏ của các thuộc tính.
@@
 Một kỹ thuật thay thế phổ biến có thể xử lý các tương tác giữa các biến giải thích là phân loại một trường hợp mới bằng cách tính khoảng cách đến k láng giềng gần nhất trong tập huấn luyện và dự đoán lớp dựa trên đa số hoặc đa số có trọng số của các láng giềng được xác định kết quả của chúng.
-Đây là một phương pháp học lười vì mô hình không được xây dựng rõ ràng và thời gian suy luận cần thiết để phân loại một trường hợp mới khá lớn.
+Đây là một phương pháp học ít tốn kém vì mô hình không được xây dựng rõ ràng và thời gian suy luận cần thiết để phân loại một trường hợp mới khá lớn.
 Nó cũng yêu cầu so sánh mỗi điểm dữ liệu mới với mỗi điểm dữ liệu trong tập huấn luyện.
 Ngoài ra, kỹ thuật này không dễ sử dụng khi nhiều giá trị thuộc tính bị thiếu, vì trong những trường hợp như vậy, phương pháp dựa trên khoảng cách để xác định láng giềng gần nhất có thể không đáng tin cậy.
-Một số trong số những giới hạn này có thể được vượt qua bằng cách sử dụng đồ thị gần nhất, trong đó các nút được kết nối nếu các điều kiện hình học nhất định được đáp ứng.
-Trong một công thức như vậy, các thuật toán đồ thị hiệu quả khác nhau (ví dụ: cây khung nhỏ nhất và tam giác hóa) có thể được sử dụng để xác định láng giềng gần nhất và tương quan hơn [80].
+Một số trong số những giới hạn này có thể được vượt qua bằng cách sử dụng nhánh mạng gần nhất (proximity graphs), trong đó các nút mạng được kết nối nếu các điều kiện hình học nhất định được đáp ứng.
+Trong một công thức như vậy, các thuật toán đồ thị hiệu quả khác nhau (ví dụ: cây khung nhỏ nhất và tam giác hóa) có thể được sử dụng để xác định láng giềng gần nhất và tương quan hơn \citep{Toussaint2005}.
 
 Một phương pháp phân loại thay thế toán học nghiêm ngặt hơn là ước tính xác suất hậu nghiệm của lớp mục tiêu bằng cách sử dụng định lý Bayes.
-Một phương pháp đơn giản nhưng thanh lịch và mạnh mẽ, được gọi là Naive Bayes, giả định rằng các giá trị thuộc tính độc lập có điều kiện với nhau, cho trước nhãn lớp y.
+Một phương pháp khá đơn giản nhưng trơn tru và mạnh mẽ, được gọi là phương pháp Naive Bayes.
+Phương pháp này giả định rằng các giá trị thuộc tính độc lập có điều kiện với nhau, cho trước nhãn lớp $y$.
 Trong trường hợp như vậy, xác suất có điều kiện của lớp của tất cả các thuộc tính có thể được phân tách thành tích của xác suất có điều kiện của mỗi thuộc tính.
-Phương pháp này chống lại nhiễu, giá trị bị thiếu và các thuộc tính không liên quan.
+Phương pháp này ngăn được nhiễu, giá trị bị thiếu và các thuộc tính không liên quan.
 Tuy nhiên, sự độc lập có điều kiện giữa các biến giải thích là một giả định mạnh mẽ không hợp lệ trong nhiều ứng dụng.
-Đối với các kịch bản như vậy, một lớp mô hình đồ thị xác suất được gọi là Mạng tin cậy Bayesian đã được phát triển bằng cách mô hình hóa các phụ thuộc có điều kiện thông qua các đồ thị vô hướng có hướng.
-Sự suy luận chính xác trên các đồ thị như vậy là NP-khó [81], do đó, các ứng dụng của chúng bị giới hạn cho số lượng thuộc tính nhỏ hơn hoặc cho các loại cấu trúc đồ thị đặc biệt.
+Đối với các kịch bản như vậy, một lớp mô hình mạng lưới xác suất được gọi là Mạng tin cậy Bayesian đã được phát triển bằng cách mô hình hóa các phụ thuộc có điều kiện thông qua các mạng vô hướng và có hướng.
+Sự suy luận chính xác trên các mạng như vậy gọ là NP-hard \citep{Cooper1990}, do đó, các ứng dụng của chúng bị giới hạn cho số lượng thuộc tính nhỏ hơn hoặc cho các loại cấu trúc mạng lưới đặc biệt.
 

### 2023-03-21 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update

@@
 
-Có các phương pháp bán giám sát và tự đào tạo, trong đó dữ liệu huấn luyện bao gồm một số dữ liệu có nhãn và nhiều dữ liệu không có nhãn \citep{Triguero2015}.
+Có các phương pháp bán giám sát và tự đào tạo, trong đó dữ liệu huấn luyện bao gồm một số dữ liệu có nhãn và nhiều dữ liệu không có nhãn.
 
@@
 Một số trong số những giới hạn này có thể được vượt qua bằng cách sử dụng nhánh mạng gần nhất (proximity graphs), trong đó các nút mạng được kết nối nếu các điều kiện hình học nhất định được đáp ứng.
-Trong một công thức như vậy, các thuật toán đồ thị hiệu quả khác nhau (ví dụ: cây khung nhỏ nhất và tam giác hóa) có thể được sử dụng để xác định láng giềng gần nhất và tương quan hơn \citep{Toussaint2005}.
+Trong một công thức như vậy, các thuật toán đồ thị hiệu quả khác nhau (ví dụ: cây khung nhỏ nhất và tam giác hóa) có thể được sử dụng để xác định láng giềng gần nhất và tương quan hơn.
 
@@
 
-Một bộ phân loại xác suất hiệu quả khác là hồi quy logistic [82].
-Chúng tôi đã thành công trong việc sử dụng phương pháp này để dự đoán xác suất mất điện liên quan đến thời tiết [83].
-Giả sử một vấn đề hai lớp (phản hồi y = 0 hoặc 1), phương pháp này tránh việc ước tính trực tiếp xác suất có điều kiện của một trường hợp, mà thay vào đó ước tính tỷ lệ giữa xác suất lớp hậu nghiệm P(y=1|x)/P(y=0|x).
-Một lợi thế lớn của hồi quy logistic so với k láng giềng gần nhất là nó có thể áp dụng cho các vấn đề có số chiều cao, vì phương pháp không phụ thuộc vào việc đo đạc sự tương đồng giữa các điểm dữ liệu.
-Lợi ích khác của phương pháp này là các tham số trọng lượng tương ứng với các thuộc tính riêng lẻ và do đó cung cấp tính giải thích khá dễ dàng.
-Tuy nhiên, sự hiện diện của một số lượng lớn các thuộc tính không liên quan là một thách thức đối với hồi quy logistic, và phương pháp này không áp dụng cho việc phân loại các trường hợp với giá trị bị thiếu, điều này có thể là một hạn chế nghiêm trọng trong thực tế.
+Một phương pháp phân loại theo xác xuất cũng khá hiệu quả là phương pháp hồi quy logistic.
+Khá nhiều ứng dụng sử dụng phương pháp này để dự đoán xác xuất mất điện liên quan đến thời tiết đã được triển khai.
+Điểm mạnh của phương pháp hồi quy logistic so với phương pháp láng giếng gần (k-nearest neighbour) là nó có thể giải quyết các vấn đề có nhiều chiều, bởi vì phương pháp này không dựa vào việc đo lường sự giống nhau giữa các điểm dữ liệu.
+Một lợi ích khác của phương pháp này là các tham số trọng lượng tương ứng với các thuộc tính riêng lẻ và do đó có tính giải thích khá dễ dàng.
+Tuy nhiên, sự hiện diện của một số lượng lớn các thuộc tính không liên quan là một thách thức đối với hồi quy logistic, và phương pháp này không áp dụng cho việc phân loại các trường hợp với giá trị rỗng, điều này có thể là một hạn chế lớn khi áp dụng trong thực tế với gói dữ liệu tồn tại số lượng lớn giá trị rỗng.
 
 Mô hình hồi quy logistic có thể được xem như một trường hợp của mô hình tuyến tính tổng quát.
-Các mô hình mạnh mẽ khác trong danh mục này là Máy vector hỗ trợ (SVM) và Mạng neural đa tầng (MNN).
-Trong SVM, vấn đề tối ưu hóa được định hình như tìm kiếm ranh giới lớn nhất phân tách siêu mặt phẳng cho một khu vực lớn tồn tại ở mỗi bên của ranh giới quyết định [84].
+Các mô hình mạnh mẽ khác trong danh mục này bao gồm Support Vector Machines (SVM) và Mạng neural đa tầng (Multilayer Neural Network - MNN).
+Trong phương pháp SVM, vấn đề tối ưu hóa được định hình như tìm kiếm ranh giới lớn nhất phân tách siêu mặt phẳng cho một khu vực lớn tồn tại ở mỗi bên của ranh giới quyết định \citep{Cortes1995}.
 Điều này được định hình như một vấn đề lập trình phi tuyến tính bị ràng buộc được biểu thị dưới dạng một hàm của các hệ số của siêu mặt phẳng phân tách, được giải quyết bằng phương pháp bội số Lagrange.
-Đối với phân loại phi tuyến tính, dữ liệu được chuyển đổi một cách ngầm định thành không gian có số chiều cao, nơi vấn đề có thể được phân tách tuyến tính.
-Điều này được đạt được bằng cách sử dụng kernel trick, để giảm bớt vấn đề hiệu quả thành tình huống phân loại tuyến tính.
-Sử dụng các kernel được chọn cẩn thận (Gaussian, đa thức hoặc sigmoid) cho phép xấp xỉ các ranh giới quyết định tùy ý.
-Các lợi ích chính của phương pháp SVM là nó chống lại nhiễu và giảm thiểu quá khớp trong khi tìm kiếm giá trị nhỏ nhất toàn cục của hàm mục tiêu.
+Đối với phân loại phi tuyến tính, dữ liệu được chuyển đổi một cách ngầm định thành không gian đa chiều, nơi vấn đề có thể được phân tách theo dạng tuyến tính.
+Điều này được đạt được bằng cách sử dụng thủ thuật Kernel, để giảm bớt vấn đề thành tình huống phân loại tuyến tính.
+Sử dụng các Kernel được chọn cẩn thận (Gaussian, đa thức, hoặc sigmoid) cho phép xấp xỉ các ranh giới quyết định tùy ý.
+Các lợi ích chính của phương pháp SVM là nó chống lại nhiễu và giảm thiểu overfitting trong khi tìm kiếm giá trị nhỏ nhất toàn cục của hàm mục tiêu.
 Tuy nhiên, chi phí tính toán của SVM là cao, và vẫn là một thách thức để sử dụng mô hình này khi các biến mô tả bị thiếu một phần trong dữ liệu quan sát.
 
-Mạng nơ-ron tiến thẳng đa tầng (FNN) cũng được sử dụng thành công cho phân loại trong nhiều ứng dụng thách thức khác nhau [85].
-Ví dụ, chúng tôi đã thành công trong việc huấn luyện FNN để phân biệt giữa dòng vào từ biến áp và dòng lỗi [86].
+Mạng nơ-ron tiến thẳng đa tầng (Multiplayer Feed-Forward Neural Network - FNN) cũng được sử dụng thành công cho phân loại trong nhiều ứng dụng khác nhau \citep{Zhang2000}.
+Ví dụ như việc áp dụng thành công trong việc huấn luyện FNN để phân biệt giữa dòng vào từ biến áp và dòng lỗi \citep{Perez1994}.
 Mô hình này có ít nhất một lớp đơn vị ẩn, mỗi lớp tính toán một hàm phi tuyến mượt và khả vi của tổng đầu vào có trọng số (ví dụ, hàm sigmoid).
 Trong mô hình này, vấn đề cập nhật các tham số khi phát hiện lỗi ở đầu ra thường được giải quyết bằng cách truyền ngược lỗi từ đầu ra về các lớp trước đó.
-Trong quá trình này, lỗi của một nút trong lớp ẩn được ước tính dưới dạng một hàm của các ước tính lỗi và trọng số trong các nút trong lớp trước đó, và giá trị này được sử dụng để cập nhật trọng số của nút ẩn này bằng cách tính toán độ dốc lỗi đối với các trọng số trong nút [87].
+Trong quá trình này, lỗi của một nút trong lớp ẩn được ước tính dưới dạng một hàm của các ước tính lỗi và trọng số trong các nút trong lớp trước đó, và giá trị này được sử dụng để cập nhật trọng số của nút ẩn này bằng cách tính toán độ dốc lỗi đối với các trọng số trong nút \citep{Werbos1994}.
 FNN có thể xấp xỉ các hàm tùy ý, và do đó mạnh mẽ hơn SVM.
-Tuy nhiên, khi thiết kế một mạng, quá khớp phải được giải quyết cẩn thận.
-Ngoài ra, nhiễu trong dữ liệu có thể gây ra vấn đề huấn luyện, vì mô hình có thể hội tụ đến một cực tiểu cục bộ, và quá trình huấn luyện có thể yêu cầu một thời gian dài, giới hạn các ứng dụng thực tế.
-Vấn đề khác với FNN cổ điển là học các mạng sâu rất khó, do tác động kết hợp của việc bão hòa hàm kích hoạt sigmoid khi truyền ngược các lỗi nhỏ, dẫn đến sự hội tụ rất chậm.
-Tiến bộ lớn đã được đạt trong việc giải quyết vấn đề này, được gọi là vấn đề gradient biến mất, trong những năm gần đây.
-Điều này, cùng với tiến bộ trong cơ sở hạ tầng tính toán phân tán dựa trên GPU và sự có sẵn của các tập dữ liệu rất lớn, đã cho phép phát triển các mạng nơ-ron sâu hiệu quả, vượt trội hơn rất nhiều so với tất cả các phương pháp thách thức khác nhau, bao gồm thị giác máy tính, xử lý ngôn ngữ tự nhiên, phát âm.
+Tuy nhiên, khi thiết kế một mạng, overfitting phải được giải quyết cẩn thận.
+Ngoài ra, nhiễu trong dữ liệu có thể gây ra vấn đề huấn luyện, vì mô hình có thể hội tụ đến một cực tiểu cục bộ, và quá trình huấn luyện có thể yêu cầu một thời gian dài, do đó giới hạn tới việc triển khai cho các ứng dụng thực tế.
+Vấn đề khác với phương pháp FNN truyền thống là học các mạng sâu rất khó, do tác động kết hợp của việc bão hòa hàm kích hoạt sigmoid khi truyền ngược các lỗi nhỏ, dẫn đến sự hội tụ rất chậm.
+Tiến bộ lớn đã được đạt trong việc giải quyết vấn đề này, được gọi là vấn đề gradient biến mất (Vanishing Gradient Problems), trong những năm gần đây.
+Điều này, cùng với tiến bộ trong cơ sở hạ tầng tính toán phân tán dựa trên GPU và sự có sẵn của các tập dữ liệu rất lớn, đã cho phép phát triển các mạng nơ-ron sâu hiệu quả, vượt trội hơn rất nhiều so với tất cả các phương pháp truyền thống, bao gồm thị giác máy tính (Computer Vision), xử lý ngôn ngữ tự nhiên, phát âm.
+
+Rất nhiều kiến trúc học sâu được đề xuất để xử lý các thuộc tính dữ liệu khác nhau.
+Một số giải pháp đã được thiết lập thường được áp dụng cho nhiều loại dữ liệu bao gồm mạng nơ-ron tích chập cho dữ liệu dựa trên lưới (ví dụ: hình ảnh) [89] và mạng nơ-ron hồi quy cho chuỗi và dữ liệu thời gian [90].
 

### 2023-03-21 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
update

@@
 Điều này, cùng với tiến bộ trong cơ sở hạ tầng tính toán phân tán dựa trên GPU và sự có sẵn của các tập dữ liệu rất lớn, đã cho phép phát triển các mạng nơ-ron sâu hiệu quả, vượt trội hơn rất nhiều so với tất cả các phương pháp truyền thống, bao gồm thị giác máy tính (Computer Vision), xử lý ngôn ngữ tự nhiên, phát âm.
+trong những năm gần đây, nhiều kiến trúc học sâu đã được đề xuất và phát triển để xử lý dữ liệu với nhiều thuộc tính khác nhau, tiêu biểu trong số đó là phương pháp mạng nơ-ron tích chập (Convutional Neural Network) áp dụng để xử lý dữ liệu toàn mạng lưới điện (ví dụ như các dữ liệu về ảnh) và mạng nơ-ron hồi quy cho chuỗi dữ liệu thời gian.
 
-Rất nhiều kiến trúc học sâu được đề xuất để xử lý các thuộc tính dữ liệu khác nhau.
 Một số giải pháp đã được thiết lập thường được áp dụng cho nhiều loại dữ liệu bao gồm mạng nơ-ron tích chập cho dữ liệu dựa trên lưới (ví dụ: hình ảnh) [89] và mạng nơ-ron hồi quy cho chuỗi và dữ liệu thời gian [90].

### 2023-03-21 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
final update on paper

--- 2023-03-21 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
@@
 \usepackage{hyperref}
+\usepackage{placeins}
 %\linenumbers
@@
 \end{itemize}
-
 Các nhà nghiên cứu và chuyên gia hiện nay tập trung vào việc khai thác dữ liệu hiện có và khám phá các nguồn dữ liệu mới và dữ liệu với quy mô lớn để theo đuổi các cải tiến trong lập kế hoạch và vận hành lưới điện.
@@
 
-\begin{figure}[h!]
+\FloatBarrier
+\begin{figure}[h]
 \centering
@@
 \end{figure}
+\FloatBarrier
 
@@
 
-\begin{figure}[h!]
+\FloatBarrier
+\begin{figure}[h]
 	\centering
@@
 \end{figure}
+\FloatBarrier
 
-\begin{figure}[h!]
+\FloatBarrier
+\begin{figure}[h]
 	\centering
@@
 Điều này, cùng với tiến bộ trong cơ sở hạ tầng tính toán phân tán dựa trên GPU và sự có sẵn của các tập dữ liệu rất lớn, đã cho phép phát triển các mạng nơ-ron sâu hiệu quả, vượt trội hơn rất nhiều so với tất cả các phương pháp truyền thống, bao gồm thị giác máy tính (Computer Vision), xử lý ngôn ngữ tự nhiên, phát âm.
+trong những năm gần đây, nhiều kiến trúc học sâu đã được đề xuất và phát triển để xử lý dữ liệu với nhiều thuộc tính khác nhau, tiêu biểu trong số đó là phương pháp mạng nơ-ron tích chập (Convolutional Neural Network) áp dụng để xử lý dữ liệu toàn mạng lưới điện (ví dụ như các dữ liệu về ảnh) và mạng nơ-ron hồi quy cho chuỗi dữ liệu thời gian \citep{Werbos1994, LeCun2015}.
 
-Rất nhiều kiến trúc học sâu được đề xuất để xử lý các thuộc tính dữ liệu khác nhau.
-Một số giải pháp đã được thiết lập thường được áp dụng cho nhiều loại dữ liệu bao gồm mạng nơ-ron tích chập cho dữ liệu dựa trên lưới (ví dụ: hình ảnh) [89] và mạng nơ-ron hồi quy cho chuỗi và dữ liệu thời gian [90].
+%Đối với các hệ thống điện, dữ liệu thường được quan sát và thu thập cả về mặt không gian và thời gian, do đó, các phương pháp hồi quy tiên tiến dựa trên cấu trúc  mạng lưới thường được áp dụng hơn để khai thác các sự phụ thuộc về mặt trúc (Ví dụ như sử dụng mô hình học máy được tích hợp phương pháp Bayes với trường ngẫu nhiên để đánh giá rủi do liên quan đến sự hủy hoại lớp cách điện khi thiết bị được đặt vào môi trường có nguy cơ liên quan đến thời tiết trong một mạng lưới).
+Các nghiên cứu về học sâu đều đưa ra kết luận rằng các ứng dụng, bao gồm hồi quy cấu trúc, có thể được hưởng lợi từ việc học các biểu diễn tiềm ẩn cho dữ liệu đầu vào.
+Trong một số nghiên cứu, việc học biểu diễn cho các trạm hệ thống điện, dựa trên sự gần nhau về không gian của chúng, đã chứng minh là rất có lợi cho việc dự đoán các lỗi liên quan đến cung cấp điện và ước tính xác suất lỗi đó.
 
-Trong các hệ thống điện, dữ liệu thường được quan sát trên không gian và thời gian, và do đó, các phương pháp hồi quy cấu trúc dựa trên đồ thị tiên tiến hơn được sử dụng để khai thác các phụ thuộc cấu trúc.
-Ví dụ, chúng tôi đã sử dụng học có cấu trúc trong các trường ngẫu nhiên có điều kiện Gaussian để đánh giá rủi ro của sự phá vỡ cách điện cho một sự tiếp xúc cụ thể và các mối đe dọa thời tiết liên quan trong một mạng điện [91].
-Nghiên cứu mới nhất về học sâu [88,92,93] cho thấy rằng một loạt các ứng dụng, bao gồm hồi quy cấu trúc, có thể được hưởng lợi từ việc học các biểu diễn tiềm ẩn cho dữ liệu đầu vào.
-Trong nghiên cứu của chúng tôi [94], việc học biểu diễn cho các trạm hệ thống điện, dựa trên sự gần nhau về không gian của chúng, đã rất có lợi cho việc dự đoán các lỗi cung cấp điện và ước tính xác suất lỗi cung cấp điện.
-Trong một phương pháp như vậy, các nút của đồ thị được nhúng trong không gian chiều thấp hơn, nơi các phương pháp học máy tiêu chuẩn có thể được áp dụng dễ dàng hơn.
+%Trong một phương pháp như vậy, các nút của đồ thị được nhúng trong không gian chiều thấp hơn, nơi các phương pháp học máy tiêu chuẩn có thể được áp dụng dễ dàng hơn.
 Các thuật toán nhúng nhằm giữ cấu trúc đồ thị và đơn giản hóa các mô hình học bằng cách di chuyển khỏi các biểu diễn đồ thị.
@@
 
-Quá trình nhúng nút đại diện cho đồ thị ban đầu trong một không gian đặc trưng mới, mô tả tốt nhất các mối quan hệ không gian của các nút trong đồ thị ban đầu.
+%Quá trình nhúng nút đại diện cho đồ thị ban đầu trong một không gian đặc trưng mới, mô tả tốt nhất các mối quan hệ không gian của các nút trong đồ thị ban đầu.
 Đặc điểm này của quá trình nhúng nút nhằm bắt giữ các mối quan hệ cần thiết của cấu trúc đồ thị ban đầu trong khi đơn giản hóa biểu diễn thành một danh sách các vector đặc trưng chiều thấp hơn.
 
-Có nhiều thuật toán để thu được nhúng như vậy; Hai thuật toán thường được sử dụng là DeepWalk [95] và Node2Vec [96].
+%Có nhiều thuật toán để thu được nhúng như vậy; Hai thuật toán thường được sử dụng là DeepWalk [95] và Node2Vec [96].
 Cả hai thuật toán đều dựa trên thông tin cộng đồng được thu được bằng các bước đi ngẫu nhiên, được sử dụng để học các biểu diễn không gian tiềm ẩn.
@@
 \subsection{Khía Cạnh Kĩ Thuật}
-Trong khi phân tích dữ liệu lớn dựa trên nền tảng khoa học dữ liệu mạnh mẽ, cũng có một số khía cạnh quan trọng khác để các phương pháp đó được sử dụng trong thực tế.
-Khi tương tác với các chuyên gia và những người trong ngành cố gắng đạt được những lợi ích rõ ràng từ việc sử dụng khoa học dữ liệu, người ta thường nghe thấy rằng phần thực tế của khoa học dữ liệu có thể chỉ chiếm 10\% công việc, trong khi 90\% liên quan đến thiết lập quy trình làm việc, quản lý dữ liệu (và lưu trữ) cũng như các khía cạnh tính toán.
-Do đó, việc quan sát một số khía cạnh liên quan đến kỹ thuật trong phân tích dữ liệu lớn là rất quan trọng.
-Chúng đã được xác định là các thách thức chính cho sự thành công của phân tích dữ liệu lớn [98,99].
+%Trong khi phân tích dữ liệu lớn dựa trên nền tảng khoa học dữ liệu mạnh mẽ, cũng có một số khía cạnh quan trọng khác để các phương pháp đó được sử dụng trong thực tế. 
+Có một thực tế đã được nhiều chuyên gia và nhà nghiên cứu tổng kết là, thực tế làm các công việc phân tích và mô hình liên quan đến xử lý dữ liệu lớn thường chỉ chiếm khoảng 10\% tổng số công việc (số giờ, tài nguyên) liên quan đến cả quá trình xử lý dữ liệu.
+Có đến 90\% công việc lại trực tiếp liên quan đến thiết lập quy trình làm việc, quản lý dữ liệu (và lưu trữ) cũng như các khía cạnh tính toán.
+Do đó, việc quan sát và tập trung nguồn lực vào một số khía cạnh liên quan đến kỹ thuật trong phân tích dữ liệu lớn là rất quan trọng.
+Chúng đã được xác định là các thách thức chính cho sự thành công của phân tích dữ liệu lớn \citep{Wickham2023}.
 
-Ở trung tâm của khái niệm phân tích dữ liệu lớn là ý tưởng cơ bản rằng dữ liệu cần được xử lý là "lớn".
-Để định nghĩa đúng dữ liệu lớn và các tính năng cần thiết của nó, người đọc được giới thiệu đến [100].
-Các ví dụ điển hình liên quan đến việc thu thập dữ liệu PMU, cũng như dữ liệu có độ phân giải cao ở mức tài sản, (ví dụ, từ các tuabin gió, bộ biến tần điện từ (PV), đồng hồ thông minh, v.v.).
+Ở trung tâm của khái niệm phân tích dữ liệu lớn cơ bản là việc chúng ta cho rằng dữ liệu cần được xử lý là "lớn".
+Để định nghĩa đúng dữ liệu lớn và các tính năng cần thiết của nó, người đọc được giới thiệu đến \citep{DeMauro2016}.
+Các ví dụ điển hình liên quan đến việc thu thập dữ liệu PMU, cũng như dữ liệu có độ phân giải cao ở mức tài sản, (ví dụ, từ các tuabin gió, bộ biến tần điện từ (PV), công tơ thông minh, v.v.).
 Tốc độ thu thập dữ liệu này ở các thang thời gian từ giây đến phút, và đồng thời ở nhiều vị trí địa lý khác nhau, trong khi cũng bao gồm nhiều loại biến số khác nhau.
-Nói chung, các dữ liệu như vậy trong lưới điện bao gồm các đo lường điểm, hình ảnh và có thể là văn bản.
-Một số khía cạnh của dữ liệu lớn này cho các hệ thống điện, từ thách thức đến ứng dụng, đã được Arghandeh và Zhou đề cập gần đây [9].
+Nói chung, các dữ liệu như vậy trong lưới điện bao gồm các dữ liệu đo lường điểm, hình ảnh và có thể là văn bản.
+Một số khía cạnh của dữ liệu lớn này cho các hệ thống điện, từ thách thức đến ứng dụng, đã được \cite{Arghandeh2017} đề cập gần đây .
 
-Trong hầu hết các tài liệu khoa học mô tả các ứng dụng điện và ngoài ra, được cho rằng dữ liệu có sẵn và chất lượng tốt.
-Tuy nhiên, bất kể khó khăn đến đâu, đảm bảo giao tiếp dữ liệu, đảm bảo tính toàn vẹn của dữ liệu, cũng như đảm bảo chất lượng dữ liệu, là các bước cần thiết trước khi thiết kế và triển khai một giải pháp dựa trên dữ liệu [101].
-Ngược lại, những khía cạnh liên quan đến tính sẵn có và chất lượng dữ liệu đã được xem xét trong một thời gian khá lâu bởi cộng đồng khoa học máy tính, cho đó đề xuất một số phương pháp [102].
-Sau đó, việc làm sạch dữ liệu và sửa đổi các bộ dữ liệu thường được liên quan đến, mặc dù người ta nên nhận thức rằng những hành động này thực sự có thể ảnh hưởng đến thông tin ban đầu trong các bộ dữ liệu, ví dụ như trong các thuộc tính thống kê của nó [103].
-Đối với các ứng dụng trong các thị trường điện liên quan đến năng lượng tái tạo, một vấn đề cổ điển ví dụ là điền vào các khoảng trống trong chuỗi thời gian, tức là có thể có các khoảng thời gian mà dữ liệu đơn giản không có sẵn, do các lỗi trong việc đăng nhập, lưu trữ hoặc truyền dữ liệu.
-Để điền vào những khoảng trống này, đã đề xuất nhiều phương pháp, tận dụng dữ liệu xung quanh khoảng thời gian đó, dữ liệu với phụ thuộc không gian thời gian (đặc biệt liên quan đến việc tạo ra năng lượng tái tạo dựa trên thời tiết), tính sẵn có của dữ liệu ở các mức tổng hợp khác nhau, cũng như mối quan hệ vật lý giữa các biến số quan tâm (ví dụ, trong một vấn đề luồng điện tối ưu).
-Đối với một ví dụ gần đây liên quan đến dữ liệu tải điện, xem thảo luận trong [104].
+Việc đảm bảo tính của giao tiếp dữ liệu, đảm bảo tính toàn vẹn của dữ liệu, cũng như đảm bảo chất lượng dữ liệu, là các bước cần thiết và quan trọng trước khi tiến hành thiết kế và triển khai một giải pháp dựa trên dữ liệu \citep{Cai2015}.
+Ngược lại, những khía cạnh liên quan đến tính sẵn có và chất lượng dữ liệu đã được xem xét trong một thời gian khá lâu bởi cộng đồng khoa học máy tính \citep{Batini2009}.
+Tiếp theo đó là việc làm sạch dữ liệu và sửa đổi các bộ dữ liệu có liên quan, mặc dù người ta nên nhận thức rằng những hành động này thực sự có thể ảnh hưởng đến thông tin ban đầu trong các bộ dữ liệu.
+Đối với các ứng dụng trong các thị trường điện liên quan đến năng lượng tái tạo, một vấn đề khá truyền thống liên quan đến ví dụ là điền vào các khoảng trống trong chuỗi thời gian, tức là có thể có các khoảng thời gian mà dữ liệu không có sẵn, do các lỗi trong việc đăng nhập, lưu trữ hoặc truyền dữ liệu.
+Để điền vào những khoảng trống này, đã có nhiều nhiều phương pháp được đưa ra như việc  tận dụng dữ liệu xung quanh khoảng thời gian đó, hay dữ liệu với phụ thuộc không gian thời gian (đặc biệt liên quan đến việc tạo ra năng lượng tái tạo dựa trên thời tiết), tính sẵn có của dữ liệu ở các mức tổng hợp khác nhau, cũng như mối quan hệ vật lý giữa các biến số quan tâm.
 
 Ngoài những khía cạnh liên quan đến dữ liệu, các phương pháp dựa trên dữ liệu được sử dụng trên các tập dữ liệu lớn đòi hỏi năng lực tính toán đáng kể để giải quyết các vấn đề mô phỏng và tối ưu hóa liên quan.
-Để tập trung dữ liệu, những vấn đề này có thể được giải quyết thông qua tính toán hiệu năng cao (HPC), ngày càng trở nên phổ biến trong các ứng dụng lưới điện và thị trường điện [105].
-Những ví dụ đáng chú ý bao gồm luồng công suất tối ưu [106] và vấn đề chuyển mạch truyền tải [107].
-Việc giải quyết các vấn đề quy mô lớn đó với HPC sẽ liên quan đến một số kỹ thuật phân rã, cũng trở nên phổ biến trong các vấn đề vận hành, thị trường và kế hoạch hệ thống điện [108].
+Chẳng hạn như việc sử dụng phương pháp tính toán hiệu năng cao  (High Performance Computing - HPC) cho việc tập trung hóa xử lý.
+Phương pháp HPC ngày càng trở nên phổ biến trong các ứng dụng phân tích lưới điện và thị trường điện \citep{Khaitan2012}.
+%Việc giải quyết các vấn đề quy mô lớn với HPC sẽ liên quan đến một số kỹ thuật phân rã, cũng trở nên phổ biến trong các vấn đề vận hành, thị trường và kế hoạch hệ thống điện [108].
 Tuy nhiên, cũng có thể có một số ứng dụng mà điều này không phải là khả thi hoặc không muốn tập trung dữ liệu.
@@
 
-Phân tích dữ liệu lớn cần được đặt trong khung cảnh tổng thể của giải quyết vấn đề.
+Phân tích dữ liệu lớn cần được đặt trong khung quản lý tổng thể để giải quyết một hay nhiều vấn đề cụ thể.
 Trong thực tế, đó chỉ là một công cụ bổ sung để hỗ trợ hoạt động và ra quyết định.
@@
 Ví dụ, nếu các dự báo được sử dụng để hỗ trợ ra quyết định, loại dự báo (ví dụ, xác định hoặc xác suất) và các sản phẩm dự báo (độ phân giải, chuẩn hóa, vv.) nên được quyết định dựa trên vấn đề quyết định cụ thể.
-Ngoài ra, khi chuyển từ dữ liệu sang phân tích, thường có nhu cầu cho một lớp trích xuất đúng loại và mức thông tin từ dữ liệu thô.
-Điều này có thể được thực hiện dựa trên bộ lọc và làm mịn, kỹ thuật kỹ thuật đặc trưng, vv.
+Ngoài ra, khi chuyển từ dữ liệu sang dạng thứ cấp để phục vụ phân tích, thường nên tạo ra một lớp trích xuất/chú giải đúng loại và mức thông tin từ dữ liệu thô để về sau dễ liên hệ.
+%Điều này có thể được thực hiện dựa trên các phương pháp liên quan đến lọc và làm mịn, kỹ thuật kỹ thuật đặc trưng, vv.
 Một ví dụ điển hình sẽ là phát hiện sự kiện dựa trên các luồng dữ liệu, sau đó được sử dụng làm đầu vào cho một số vấn đề phân tích khác.
@@
 Yêu cầu quan trọng liên quan đến dữ liệu chính nó (và các luồng dữ liệu liên quan) là cách xử lý an ninh mạng và quyền riêng tư trong lưới điện.
-Hiện nay, an ninh mạng đại diện cho một thành phần quan trọng của hệ thống năng lượng phân phối trong tương lai, mà trên đó phân tích dữ liệu lớn có thể được thực hiện [112].
-Do đó, thiết lập cho phân tích dữ liệu lớn, cũng như các công cụ được sử dụng, cần phải mạnh mẽ để có thể chịu đựng được việc loại bỏ các dữ liệu quan trọng hoặc làm giả dữ liệu.
-Ngoài ra, quyền riêng tư dữ liệu đang trở thành mối quan tâm ngày càng tăng vì nếu dữ liệu đang được thu thập được chia sẻ, có thể suy ra thông tin về các tài sản hoặc người tiêu dùng cụ thể, điều đó không hề được định sẵn.
-Mối quan tâm về quyền riêng tư đặc biệt hiện nay khi các đồng hồ đo thông minh được triển khai rộng rãi [113], cho phép một số người có thể có thông tin về thói quen tiêu dùng của người tiêu dùng để tiến hành các hoạt động tiếp thị và tội phạm.
-Yêu cầu chính khác liên quan đến nhu cầu có các mô hình và kết quả có thể giải thích được.
+Hiện nay, an ninh mạng đại diện cho một thành phần quan trọng của hệ thống năng lượng phân phối trong tương lai, mà trên đó phân tích dữ liệu lớn có thể được thực hiện \citep{Li2017}.
+Do đó, công tác thiết lập cho việc phân tích dữ liệu lớn, cũng như các công cụ được sử dụng, cần phải mạnh mẽ để có thể chịu đựng được việc loại bỏ các dữ liệu quan trọng hoặc làm giả dữ liệu.
+Ngoài ra, quyền riêng tư dữ liệu đang trở thành mối quan tâm ngày càng tăng vì nếu dữ liệu đang được thu thập được chia sẻ, có thể suy ra thông tin các nhân về các tài sản hoặc người tiêu dùng cụ thể.
+Mối quan tâm về quyền riêng tư đặc biệt hiện nay khi các công tơ thông minh được triển khai rộng rãi, cho phép một số người có thể có thông tin về thói quen tiêu dùng của người khách hàng để tiến hành các hoạt động tiếp thị và tội phạm.
+%Yêu cầu chính khác liên quan đến nhu cầu có các mô hình và kết quả có thể giải thích được.
 Điều này đã kích hoạt một lĩnh vực nghiên cứu mới liên quan đến việc học máy có thể giải thích và học máy dựa trên vật lý.
@@
 \subsection{Khung Đưa ra Quyết Định}
-Việc sử dụng phân tích dữ liệu lớn tất yếu dẫn đến việc nâng cao khả năng ra quyết định.
-Do đó, trong phương pháp phân tích dữ liệu được đề xuất, chú ý đặc biệt được dành cho việc trực quan hóa kết quả.
+Việc sử dụng phân tích dữ liệu lớn tất yếu sẽ dẫn đến việc nâng cao khả năng ra quyết định.
+Do đó, việc trực quan hóa trong toàn bộ quá trình xử lý nên được đặc biệt quan tâm vì các nhà quản lý và điều hành thường đưa ra các quyết định dựa trên các con số và bảng biểu hay đồ thị được trực quan hóa.
 
-Một cách tiếp cận được sử dụng để dự báo các sự cố mất điện là phát triển bản đồ rủi ro như các bản đồ được hiển thị trong Hình \ref{fig:2}.
-Chúng đại diện cho nguy cơ thời tiết, độ dễ tổn thương và rủi ro tính toán dưới dạng tích của hai yếu tố này.
+Trong trong những vấn đề giành được sự quan tấm lớn hiện nay, không những chỉ trong ngành điện, mà còn trong công tác quản lý hạ tầng nói chung đó là việc quản lý rủi do.
+Trong công tác quản lý rủi do, ngoài việc đưa ra các mức rủi do mang tính chất định tính, các nhà quản lý cũng cần phải yêu cầu các kĩ sư và nhà phân tích đưa ra các con số định lượng và trực quan hóa các nguy cơ gây ra rủi do, mức độ tổn thương mạng lưới và rủi do cho từng bộ phân và khu vực trên toàn mạng lưới (Hình \ref{fig:2}).
+Như ví dụ ở Hình \ref{fig:2}, một bản đồ rủi do được thiết lập dựa trên việc tích hợp của hai bản đồ nền tảng là bản đồ nguy cơ và bản đồ tổn thương.
 
-\begin{figure}[h!]
+\FloatBarrier
+\begin{figure}[h]
 	\centering
 	\begin{center}
-		\includegraphics[width=\textwidth]{ruido}% This is a *.eps file
+		\includegraphics[width=15cm]{ruido}% This is a *.eps file
 	\end{center}
@@
 \end{figure}
+\FloatBarrier
 
-Một khung công việc đánh giá rủi ro được định hình bởi Văn phòng Cứu trợ Khẩn cấp của Liên Hợp Quốc (UNDRO), đã được khám phá gần đây bởi Văn phòng Liên Hợp Quốc về Giảm thiểu Rủi ro Thiên tai (UNISDR) trong báo cáo liên quan đến họ [114], sau đó được Cơ quan Quản lý Khẩn cấp Liên bang (FEMA) của Hoa Kỳ đưa ra, vì trọng tâm chính là ảnh hưởng liên quan đến khí hậu đến cơ sở hạ tầng, xã hội và môi trường.
+Như chúng ta đã thấy, việc đánh giá rủi do là khá phức tạp, đặc biệt là trên phương diện mạng lưới.
+Do đó, để tính toán được một cách có hệ thống và đạt được một mức độ chính xác nhất định, chúng ta cần phải xây dựng một khung công việc (framewwork) hay một dòng chảy công việc (work flow) nền tảng.
+Các khung công việc hay dòng chảy công việc này cần được thiết lập cho từng cấp độ khác nhau.
+Ở phạm vi quốc tế, ví dụ chúng ta có thể thấy đó là khung công việc cho đánh giá rủi do được đưa ra bởi Văn phòng Cứu trợ Khẩn cấp của Liên Hợp Quốc (UNDRO), hay Văn phòng Liên Hợp Quốc về Giảm thiểu Rủi ro Thiên tai (UNISDR).
+Ở mức độ quốc gia, chúng ta có thể lấy ví dụ về các tiêu chuẩn được đưa ra bởi Cơ quan Quản lý Khẩn cấp Liên bang (FEMA) của Hoa Kỳ.%., vì trọng tâm chính là ảnh hưởng liên quan đến khí hậu đến cơ sở hạ tầng, xã hội và môi trường.
 Điều này giới thiệu Trạng thái Rủi ro (SoR) như sau:
 
-trong đó P(T) là Nguy cơ hoặc xác suất của một mức độ mối đe dọa cụ thể (T); u(C) là Tổn thất (xã hội, kinh tế hoặc môi trường) liên quan đến mức độ Hậu quả (C), liên quan đến mức độ mối đe dọa (T); và P(C|T) là Tính dễ tổn thương hoặc xác suất của việc trải qua mức độ hậu quả (C) được xác định dựa trên mức độ mối đe dọa (T).
+Rủi được thể hiện qua công thức tính toán đơn giản sau.
+
+\begin{equation}\label{math01}
+	\textbf{R}isk = \textbf{H}azard \times \textbf{V}ulnerability \times \textbf{C}onsequences
+\end{equation}
+
+Nguy cơ (H) và sự tổn thương (V) có thể được xác định bằng xác xuất.
+Nguy cơ có thể được tính toán dựa trên xác xuất của một sự kiện nào đó (chẳng hạn như động đất, lũ lụt, mưa lớn) xảy ra với mức độ hay độ mạnh khác nhau (thường gọi là Intensity).
+Tổn thương (V) liên quan trực tiếp tới yếu tố bản thân của thiết bị tài sản, như xự xuống cấp hay các thông số thiết kế và độ mạnh của nguy cơ.
+Ở nhiều góc độ khác, tổn thương còn liên quan đến các yêu tố quản lý và sự sẵn sàng của doanh nghiệp trước các sự cố và hiện tượng bất thường.
+
+Thiệt hại (C) cần phải được tính toán cụ thể và thường được đo bởi đơn vị tiền tệ.
+Ở đây, thiệt hại nên được hiểu theo nghĩa rộng, tức là không chỉ dừng lại ở việc đo lường thiệt hại cho doanh nghiệp, mà còn phải đo lường thiệt hai cho cộng đồng và cho môi trường.
+Để tính toán và lượng hóa được thiệt hại cần phải xây dựng một hệ thống hay cấu trúc thông xuốt để loại trừ cách tính trùng lặp (thường được hiểu là thiết lập một Impact Hierachy).
+Một số phương pháp tiêu biểu có thể được áp dụng để tính toán H, V, và C là phương pháp Fault-Tree Analysis (FTA), hay còn gọi là phân tích lỗi theo dạng cây và phương pháp Event Tree Analysis (ETA) hay còn gọi là phân tích sự kiện theo dạng cây.
+
+%trong đó P(T) là Nguy cơ hoặc xác suất của một mức độ mối đe dọa cụ thể (T); u(C) là Tổn thất (xã hội, kinh tế hoặc môi trường) liên quan đến mức độ Hậu quả (C), liên quan đến mức độ mối đe dọa (T); và P(C|T) là Tính dễ tổn thương hoặc xác suất của việc trải qua mức độ hậu quả (C) được xác định dựa trên mức độ mối đe dọa (T).
 Đơn vị rủi ro được biểu thị trong đơn vị của các tổn thất.
 
-Trong bối cảnh này, quyết định liên quan đến quá trình tối ưu hóa, trong đó hàm mục tiêu và ràng buộc nhằm mục đích hành động giảm thiểu được xác định.
+Việc tính toán rủi do là quan trọng, nhưng cũng chỉ là một bước nếu đặt trong một mô hình toán tối ưu (Optimization Model) khi mà các nhà quản lý cần quan tâm đến nguồn lực nội tại cũng như các khó khăn phải đối mặt (Constraints).
+Do đó, sau khi tính toán được rủi do, cần xây dựng mô hình tối ưu với một hàm mục tiêu cụ thể và các giàng buộc đi kèm.
+Hình \ref{fig:3} mô tả khái quan một khung làm việc cho ví dụ về dự báo mất điện mạng lưới gây ra bởi thời tiết.
+
+%Trong bối cảnh này, quyết định liên quan đến quá trình tối ưu hóa, trong đó hàm mục tiêu và ràng buộc nhằm mục đích hành động giảm thiểu được xác định.
 Một khung quyết định rộng hơn cho các ví dụ về dự báo mất điện đã được hiển thị trong Hình \ref{fig:3}.
-
-\begin{figure}[h!]
+\FloatBarrier
+\begin{figure}[h]
 	\centering
 	\begin{center}
-		\includegraphics[width=\textwidth]{decisionmaking-risk}% This is a *.eps file
+		\includegraphics[width=15cm]{decisionmaking-risk}% This is a *.eps file
 	\end{center}
@@
 \subsection{Quản Lý tài Sản Cơ Sở Hạ Tầng và Sự Cố Mất Điện}
-Các thị trường cạnh tranh về điện, sự tư nhân hóa và các yêu cầu kỹ thuật hoặc quy định bắt buộc các công ty điện phải tối ưu hóa các thực tiễn quản lý vận hành, gián đoạn và tài sản và phát triển các kế hoạch quyết định cần thiết kỹ thuật kinh tế.
-Trong thực tế hoạt động của công ty điện hôm nay, quản lý tài sản và gián đoạn được xử lý bởi các nhóm khác nhau và được coi là các nỗ lực lập kế hoạch dài hạn và ngắn hạn, tương ứng.
-Chúng tôi đã giữ cuộc thảo luận về các vấn đề có vẻ không liên quan này cùng với nhau để nhấn mạnh sự tương quan chặt chẽ giữa chúng về việc sử dụng dữ liệu, vì quản lý gián đoạn có thể sử dụng cùng dữ liệu với quản lý tài sản và ngược lại vì tình trạng cơ sở của tài sản điều khiển cả hai ứng dụng:
+Tại các thị trường có sự cạnh tranh về điện, tư nhân hóa và các yêu cầu kỹ thuật hoặc quy định bắt buộc các công ty điện phải tối ưu hóa công tác quản lý và vận hành.
+Trong thực tế hoạt động của công ty điện, quản lý tài sản và quản lý sự cố mất điện được xử lý bởi các nhóm khác nhau và thường được đề cập trong việc lập kế hoạch từ ngắn hạn đến dài hạn.
+Việc quản lý tài sản và quản lý sự cố mất điện có mối tương quan với nhau và chính vì thế việc sử dụng dữ liệu của cả hai nhóm này trên cùng một hệ thống và nền tảng phân tích sẽ là một trong những ưu tiên nên được đầu tư và quan tâm thích đáng.
+
+Các công ty và doanh nghiệp điện lực thường chú trọng nhiều hơn đến việc quản lý sự cố mất điện chứ chưa có đầu tư đích đáng vào quản lý tài sản.
+Ở các nước phát triển, có một thực tế là việc quản lý tài sản nói chung và ngành điện nói riêng cũng chỉ được quan tâm trong vòng 15 năm trở lại đây.
+Đã có khá nhiều phần mềm quản lý tài sản, gọi chung là CMMS (Computerized Maintenance and Management Systems) đã được phát triển và đưa vào sử dụng.
+Nhưng ở các nước đang phát triển, trong đó có Việt Nam, quản lý tài sản là một khái niệm tương đối mới và chưa có một trường đại học nào đào tạo về ngành này.
+
+Về cơ bản, theo tác giả, quản lý tài sản có thể được định nghĩa là "\textbf{Tối ưu hóa việc phân bổ nguồn tài nguyên có hạn cho công tác làm mới hay duy tu vả bảo dưỡng các thiết bị và hệ thống sẵn có.
+Việc tối ưu hóa phải được tính đến lợi ích và ảnh hưởng của doanh nghiệp cũng như của các bên liên quan và môi trường}".
+Hình \ref{fig:asset} mô tả các qui trình và công đoạn cần thiết cấu thành lên một hệ thống quản lý tài sản hoàn chỉnh.
+
+\FloatBarrier
+\begin{figure}[h]
+	\centering
+	\begin{center}
+		\includegraphics[width=14cm]{assetmanagement}% This is a *.eps file
+	\end{center}
+	\caption{Qui trình và công đoạn trong quản lý tài sản hạ tầng}\label{fig:asset}
+\end{figure}
+\FloatBarrier
+
+%Chúng tôi đã giữ cuộc thảo luận về các vấn đề có vẻ không liên quan này cùng với nhau để nhấn mạnh sự tương quan chặt chẽ giữa chúng về việc sử dụng dữ liệu, vì quản lý gián đoạn có thể sử dụng cùng dữ liệu với quản lý tài sản và ngược lại vì tình trạng cơ sở của tài sản điều khiển cả hai ứng dụng:
 
@@
 
-“Quản lý tài sản trong các hệ thống điện có thể được phân loại rộng rãi thành bốn loại chính dựa trên các thời gian có thể, tức là, thời gian thực, ngắn hạn, trung hạn và dài hạn [115]”
+Quản lý tài sản trong các hệ thống điện có thể được phân loại rộng rãi thành bốn loại chính dựa trên các thời gian thời gian thực, ngắn hạn, trung hạn và dài hạn.
 
@@
 Quản lý tài sản ngắn hạn cố gắng tối đa hóa tỷ lệ lợi nhuận liên quan đến đầu tư tài sản.
-Giá trị chủ yếu phụ thuộc vào giá thị trường không chắc chắn thông qua các thực hiện thị trường khác nhau.
-Đánh giá rủi ro thị trường là một yếu tố quan trọng và phân phối doanh thu / lợi nhuận được đạt thông qua phân tích tính lợi nhuận.
+Giá trị chủ yếu phụ thuộc vào giá thị trường không chắc chắn thông qua việc điều chỉnh và tác động vào thị trường.
+Đánh giá rủi ro thị trường là một yếu tố quan trọng và phân phối doanh thu/lợi nhuận được đạt thông qua phân tích tính lợi nhuận.
 Lập lịch bảo trì tối ưu thuộc quản lý tài sản trung hạn.
@@
 Các nỗ lực được tập trung vào tối ưu hóa phân bổ tài nguyên tài chính hạn chế ở những nơi và thời điểm cần thiết để quản lý gián đoạn tối ưu mà không đánh đổi tính đáng tin cậy của hệ thống.
-Sử dụng rộng rãi các công nghệ cảm biến thông minh và giám sát để đánh giá sức khỏe và độ tin cậy của thiết bị hệ thống theo thời gian và tối ưu hóa các kế hoạch bảo trì tương ứng [116].
+Sử dụng rộng rãi các công nghệ cảm biến thông minh và giám sát để đánh giá tình trạng làm việc và độ tin cậy của thiết bị và hệ thống theo thời gian và tối ưu hóa các kế hoạch bảo trì tương ứng.
 Đầu tư vào kế hoạch mở rộng hệ thống điện, cũng như triển khai rộng rãi các nguồn điện phân tán, thuộc phạm vi của quản lý tài sản dài hạn, trong đó các nhà đầu tư, các đối thủ cạnh tranh và các bên liên quan khác được mời tham gia vào các kế hoạch kinh tế trong tương lai.
@@
 
-\subsubsection{Các Tác Động của Thời Tiết và Việc Mất Điện}
-Tác động của thời tiết đối với gián đoạn trong hệ thống điện có thể được phân loại thành trực tiếp và gián tiếp [117]:
+\subsubsection{Tác Động của Thời Tiết và Sự Cố Mất Điện}
+Tác động của thời tiết đối dẫn đến sự cố mất điện và gián đoạn trong hệ thống điện có thể được phân loại thành trực tiếp và gián tiếp:
 
 \begin{itemize}
-\item Tác động trực tiếp đối với tài sản của tiện ích: Loại tác động này bao gồm tất cả các tình huống khi điều kiện thời tiết nghiêm trọng trực tiếp gây ra sự cố của thành phần.
-Ví dụ như sét đánh vào tài sản của tiện ích, tác động của gió khiến cho cây hoặc cành cây chạm vào đường dây, v.v.
-Những loại gián đoạn này được đánh dấu là gián đoạn do thời tiết gây ra.
+\item Tác động trực tiếp đối với tài sản hạ tầng điện: Loại tác động này bao gồm tất cả các tình huống khi điều kiện thời tiết nghiêm trọng trực tiếp gây ra sự cố với tài sản.
+Ví dụ như sét đánh vào thiết bị, tác động của gió khiến cho cây hoặc cành cây chạm vào đường dây, v.v.
+Những sự cố này được đánh dấu là sự cố do thời tiết gây ra.
 
-\item Tác động gián tiếp đối với tài sản của tiện ích: Loại tác động này xảy ra khi thời tiết tạo ra một tình huống trong mạng lưới mà gián tiếp gây ra sự cố của thành phần.
-Các ví dụ bao gồm: điều kiện thời tiết nóng làm tăng nhu cầu, gây quá tải cho đường dây dẫn đến giảm độ cao của đường dây, tăng nguy cơ sự cố do tiếp xúc với cây, tài sản tiếp xúc với tác động của thời tiết trong dài hạn dẫn đến suy giảm thành phần, v.v.
+\item Tác động gián tiếp đối với tài sản hạ tầng điện: Loại tác động này xảy ra khi thời tiết tạo ra một tình huống trong mạng lưới mà gián tiếp gây ra sự cố của thiết bị.
+Các ví dụ bao gồm: điều kiện thời tiết nóng làm tăng nhu cầu, gây quá tải cho đường dây dẫn đến giảm độ cao của đường dây, tăng nguy cơ sự cố do tiếp xúc với cây, thiết bị tiếp xúc với tác động của thời tiết trong dài hạn dẫn đến xuống cấp và giảm khả năng cung ứng dịch vụ, v.v.
 Những loại gián đoạn này được đánh dấu là lỗi thiết bị.
@@
 \end{itemize}
-\subsubsection{Cơ Bản về Quản Lý Mất Điện}
 
-Khả năng theo dõi đồng thời nhiều mối đe dọa thời tiết khi chúng phát triển và đánh giá các tác động kết quả đa dạng đến tài sản ngành điện, cơ sở hạ tầng và các đường đời mà chúng hỗ trợ là rất quan trọng trong ngành điện.
-Lưới điện được phân bổ trên các vùng rộng lớn với việc phát điện thường đặt ở những vùng xa xôi.
-Tiêu thụ lớn ở các khu đô thị có nghĩa là lưới truyền tải phải đưa điện từ các trang trại phát điện ở xa đến các trung tâm tiêu thụ và hệ thống phân phối phải cung cấp các đường dẫn tiện ích cho từng khách hàng cá nhân.
-Để đạt được điều này, lưới phải trải qua các trạng thái hoạt động khác nhau (Hình 4).
-Các trạng thái thị trường điện tương ứng được hiển thị trong Bảng 3.
-Bằng cách kết hợp quản lý tài sản và gián đoạn, ta xử lý các tác động một cách hiệu quả nhất.
+Ngoài hai khái niệm trên, thực tế là chúng ta có thể chia sự xuống cấp của thiết bị theo hai xu hướng: (1) Xuống cấp theo thời gian và có thể quan sát được (Manifest Deterioration) và (2) xuống cấp bất ngờ hay đột ngột (Latent/Sudden) do các nguyên nhân như thiên tai.
 
-\subsubsection{Ước Đoán Sự Mất Điện và Mạng Lưới Truyền Tải}
-Kiến thức từ dữ liệu lịch sử có thể được sử dụng để dự đoán những cúp điện truyền tải liên quan đến thời tiết 1-3 giờ trước.
-Nhúng không gian được thêm vào bộ dữ liệu đầu vào [94] để bắt các phụ thuộc không gian giữa các nút và sự kiện.
-Xem xét ví dụ với dữ liệu cúp điện lịch sử được thu thập từ Cơ quan Điện lực Bonneville (BPA) [120].
-Chương trình Hệ thống Quan sát Bề Mặt Tự Động (ASOS) [26] được sử dụng để thu thập các đo lường thời tiết lịch sử cho các tham số sau: Hướng gió [độ], Tốc độ gió [nút], Cơn gió mạnh [nút], Nhiệt độ [F], Điểm sương [F], Độ ẩm tương đối [\%], Áp suất [mb], Lượng mưa/giờ [inch] và Mã thời tiết hiện tại.
-Cơ sở Dữ liệu Dự báo Kỹ thuật số Quốc gia (NDFD) [121] được sử dụng để trích xuất dữ liệu dự báo thời tiết lịch sử được sử dụng để kiểm tra hệ thống ánh xạ xác suất cúp điện thời gian thực được sử dụng trong nghiên cứu phối hợp cách điện [41].
+\subsubsection{Cơ Bản về Quản Lý Sự Cố Mất Điện}
 
-Việc đặt chỗ lý tưởng của các thiết bị bảo vệ sét dòng dây nhằm giảm thiểu tổng rủi ro của các sự cố và độ nhiễu liên quan đến sét khi vẫn duy trì trong giới hạn ngân sách yêu cầu [41].
-Mạng lưới và tác động của nó được mô hình hóa bằng một đồ thị trọng số đa chế độ sử dụng dữ liệu từ nhiều nguồn khác nhau.
-Mô hình rủi ro được phát triển xem xét tác động tích lũy của các sự cố sét trước đó để tạo ra một ước tính chính xác hơn về độ cứng của bộ cách điện và dự đoán hiệu suất của bộ cách điện cho tình huống gây ra quá áp do sét trong tương lai, sử dụng Gaussian Conditional Random Fields (GCRF) [122].
-Phương pháp quản lý thực vật dựa trên dữ liệu dự đoán trong phân phối được giới thiệu trong [123].
-Một mô hình cho mối tương quan không gian thời gian của một loạt dữ liệu được phát triển, cho phép phân tích thời gian thực về tác động của thực vật lên dây cấp điện dựa trên bản đồ rủi ro dự đoán.
+Khả năng theo dõi đồng thời nhiều mối đe dọa do thời tiết gây ra và đánh giá các tác động và thiệt hại đến tài sản ngành điện hay các sở hạ tầng (hạ tầng giao thông) liên quan đến việc hỗ trợ cho ngành điện là rất quan trọng.
+Đơn giản bời vì, lưới điện được phân bổ trên các vùng rộng lớn với các nhà máy phát điện thường đặt ở những vùng xa xôi.
+Tiêu thụ lớn ở các khu đô thị có nghĩa là lưới truyền tải phải đưa điện từ các nhà máy phát điện ở xa đến các trung tâm tiêu thụ và hệ thống phân phối phải cung cấp các đường dẫn tiện ích cho từng khách hàng.
+Để đạt được điều này, lưới phải trải qua các trạng thái hoạt động khác nhau (Hình \ref{fig:grid}).
+Bằng cách kết hợp quản lý tài sản và quản lý sự cố mất điện, chúng ta có thể xử lý và giảm thiểu các tác động gây thiệt hại một cách hiệu quả nhất.
+
+\FloatBarrier
+\begin{figure}[h]
+	\centering
+	\begin{center}
+		\includegraphics[width=6cm]{electricgridstates}% This is a *.eps file
+	\end{center}
+	\caption{Các trạng thái của lưới điện}\label{fig:grid}
+\end{figure}
+\FloatBarrier
+
+\subsubsection{Tiên Đoán Mất Điện trên Mạng Lưới Truyền Tải}
+Kiến thức từ dữ liệu thu thập trong quá khứ có thể được sử dụng để dự đoán những sự cố mất điện trên lưới điện liên quan đến thời tiết trong vòng 1-3 giờ trước.
+Các biến liên quan đến không gian được thêm vào bộ dữ liệu để tính toán được mối liên hệ tương hỗ giữa các sự kiện và các nút trên mạng lưới.
+Chương trình Hệ thống Quan sát Bề Mặt Tự Động (ASOS) được sử dụng để thu thập các đo lường thời tiết lịch sử cho các tham số sau: Hướng gió [độ], Tốc độ gió [km/h], Gió giật [m/s], Nhiệt độ [C], Điểm sương [F], Độ ẩm tương đối [\%], Áp suất [mb], Lượng mưa/giờ [mm].
+Cơ sở Dữ liệu Dự báo Kỹ thuật số Quốc gia (NDFD) của Mỹ được sử dụng để trích xuất dữ liệu dự báo thời tiết trong quá khứ và được sử dụng để kiểm tra khả năng theo thời gian thực tính phản xạ của hệ thống tương ứng với các xác suất mất điện khác nhau theo thời gian.
+
+Việc bố trí tối ưu các thiết bị bảo vệ sét cho dây điện và cột điện nhằm giảm thiểu tổng rủi ro của các sự cố và độ nhiễu liên quan đến sét khi vẫn duy trì trong giới hạn ngân sách yêu cầu.
+Mạng lưới và tác động của nó được mô hình hóa bằng một mạng có trọng số đa chế độ sử dụng dữ liệu từ nhiều nguồn khác nhau.
+Mô hình rủi ro được phát triển (sử dụng Gaussian Conditional Random Fields (GCRF) \citep{Radosavljevic2013}) có xem xét đến tác động tích lũy của các sự cố sét trước đó để tạo ra một ước tính chính xác hơn về độ nhiễu của bộ cách điện và dự đoán hiệu suất của bộ cách điện khi có tình huống xảy ra quá áp do sét đánh trong tương lai.
+%cho phép phân tích thời gian thực về tác động của thực vật lên dây cấp điện dựa trên bản đồ rủi ro dự đoán.
 Thuật toán dự đoán dựa trên GCRF regression predictor.
@@
 
-Các phương pháp truyền thống để đánh giá sức khỏe của bộ biến áp được phát triển bằng cách sử dụng kiến thức lĩnh vực về các quá trình vật lý và hóa học xảy ra bên trong bồn dầu của bộ biến áp, sau đó được xác nhận bằng các nghiên cứu kinh nghiệm.
-Một số ví dụ là tam giác Duval [124], tỷ số khí IEC [125] hoặc phân tích khí chính [126].
-Việc thu thập dữ liệu ngày càng tăng (ví dụ: phân tích khí và dầu định kỳ, các cảm biến thu thập dữ liệu thời gian thực, v.v.) bởi các công ty điện đã thúc đẩy sự phát triển các phương pháp phân tích dữ liệu dựa trên học có giám sát để phân loại tình trạng bộ biến áp và loại lỗi.
-Một số ví dụ là mạng nơ-ron nhân tạo đa tầng [127], máy vector hỗ trợ [128] và mạng Bayesian [129].
-Những sự cố của bộ biến áp hạ áp (22,9KV-220V) được sử dụng trong các lĩnh vực phân phối ở Hàn Quốc được nghiên cứu trong [130].
+Các phương pháp truyền thống để đánh giá tình trạng hoạt động (sức khỏe) của trạm biến áp được phát triển bằng cách sử dụng kiến thức lĩnh vực về các quá trình vật lý và hóa học xảy ra bên trong bồn dầu của bộ biến áp, sau đó được xác nhận bằng các nghiên cứu kinh nghiệm.
+Một số ví dụ là việc sử dụng phương pháp tam giác Duval, tỷ số khí IEC hoặc phân tích khí chính.
+Việc thu thập dữ liệu ngày càng tăng (ví dụ: phân tích khí và dầu định kỳ, các cảm biến thu thập dữ liệu thời gian thực, v.v.) bởi các công ty điện đã thúc đẩy sự phát triển các phương pháp phân tích dữ liệu dựa trên học máy có giám sát để phân loại tình trạng bộ biến áp và loại ra các lỗi.
+Một số ví dụ là việc sử dụng mạng nơ-ron nhân tạo đa tầng, Support Vector Machine và mạng Bayesian .
+%Những sự cố của bộ biến áp hạ áp (22,9KV-220V) được sử dụng trong các lĩnh vực phân phối ở Hàn Quốc được nghiên cứu trong [130].
 
-Các thuật toán học có giám sát đối mặt với các thách thức sau: (i) đa số các thuật toán chỉ cung cấp khả năng giải thích thấp cho người ra quyết định, điều này là yêu cầu cơ bản trong vấn đề này; (ii) thiếu dữ liệu lỗi có chất lượng cao; và (iii) dữ liệu được gán nhãn về tình trạng biến áp trong hầu hết các trường hợp không có sẵn hoặc được xác định bởi con người (tức là phân loại chủ quan của tình trạng).
-Sử dụng học không giám sát là một giải pháp thay thế và hấp dẫn, nhưng trong tài liệu vẫn còn hạn chế.
-Phương pháp không giám sát đầu tiên là chỉ số sức khỏe được mô tả trong tài liệu tham khảo [131] tóm tắt sức khỏe tổng thể của tài sản bằng cách kết hợp kết quả quan sát vận hành, kiểm tra trường và kiểm tra phòng thí nghiệm thành một chỉ số duy nhất.
+Các thuật toán học máy có giám sát đối mặt với các thách thức sau: (i) đa số các thuật toán chỉ cung cấp khả năng giải thích thấp cho người ra quyết định; (ii) thiếu dữ liệu về sự cố có chất lượng cao; và (iii) dữ liệu được gán nhãn về tình trạng biến áp trong hầu hết các trường hợp không có sẵn hoặc được xác định bởi con người (tức là phân loại chủ quan của tình trạng).
+Sử dụng học máy không giám sát là một giải pháp thay thế và hấp dẫn, nhưng trong tài liệu vẫn còn hạn chế.
+Phương pháp không giám sát đầu tiên là chỉ số sức khỏe được mô tả trong tài liệu tham khảo \citep{Jahromi2009} tóm tắt sức khỏe tổng thể của tài sản bằng cách kết hợp kết quả quan sát vận hành, kiểm tra trường và kiểm tra phòng thí nghiệm thành một chỉ số duy nhất.
 Tuy nhiên, các hạn chế chính của chỉ số này là: (i) định nghĩa kinh nghiệm của trọng số cho mỗi tiêu chí, và (ii) thiếu thông tin về loại lỗi.
-Các phương pháp thay thế khác là phân cụm [132] và học bán giám sát với Low Dimensional Scaling [133].
-Hơn nữa, nghiên cứu về học sâu có thể đóng góp vào việc bổ sung dữ liệu [134] cho dữ liệu biến áp và cung cấp các kỹ thuật mới cho học không giám sát như mạng Siamese [135].
-Cần nhấn mạnh rằng chúng ta nên mong đợi hiệu suất thấp hơn từ các kỹ thuật không giám sát so với học có giám sát, nhưng tiềm năng ứng dụng lớn hơn trong tập dữ liệu thực tế.
+Các phương pháp thay thế khác là phân cụm \citep{Islam2016} và học bán giám sát với Low Dimensional Scaling \citep{Mirowski2012}.% Hơn nữa, nghiên cứu về học sâu có thể đóng góp vào việc bổ sung dữ liệu \citep{Antoniou2018} cho dữ liệu biến áp và cung cấp các kỹ thuật mới cho học không giám sát như mạng Siamese \citep{Koch2015}.
+%Cần nhấn mạnh rằng chúng ta nên mong đợi hiệu suất thấp hơn từ các kỹ thuật không giám sát so với học có giám sát, nhưng tiềm năng ứng dụng lớn hơn trong tập dữ liệu thực tế.
 
-\subsubsection{Những dự đoán về thiệt hại cơ sở hạ tầng thảm khốc gây ra các tình trạng mất điện}
+\subsubsection{Dự đoán về thiệt hại cơ sở hạ tầng thảm khốc gây ra tình trạng mất điện}
 
-Phân tích dữ liệu lớn có thể được sử dụng để dự đoán các sự cố về tài sản thảm khốc do các sự kiện thời tiết khắc nghiệt như bão, vòng xoáy, lốc xoáy và sóng thần [136,137].
-Những nghiên cứu như vậy chủ yếu liên quan đến dự đoán thiệt hại cơ sở hạ tầng, bao gồm số cột bị đổ, trạm biến áp bị phá hủy và các thiệt hại khác đòi hỏi phải tái thiết kế toàn bộ cơ sở hạ tầng lưới điện.
-Một đánh giá ngầm cũng liên quan đến các tình trạng mất điện vì cần phải tái thiết kế trước khi khôi phục dịch vụ
+%Phân tích dữ liệu lớn có thể được sử dụng để dự đoán các sự cố thảm khốc do các sự kiện thời tiết khắc nghiệt như bão, lốc xoáy, lũ lụt và sóng thần [136,137].
+Hiện tại, nghiên cứu trong mảng này chủ yếu liên quan đến dự đoán thiệt hại cơ sở hạ tầng, bao gồm số cột bị đổ, trạm biến áp bị phá hủy và các thiệt hại khác đòi hỏi phải tái thiết toàn bộ hay một phần của cơ sở hạ tầng lưới điện.% Một đánh giá ngầm cũng liên quan đến các tình trạng mất điện vì cần phải tái thiết kế trước khi khôi phục dịch vụ
+
+Một trong những vấn đề mà ít có nghiên cứu nào đề cập đó là vấn đề liên quan đến phục hồi (restoration) khi có thiên tai xảy ra, các thiên tai như mưa lớn lũ lụt gây ra sự cố mất điện trên diện rộng và để phục hồi việc cấp điện nhiều khi cần phải huy động nguồn lực không những trong ngành điện mà còn các ngành và cơ quan chính quyền như quân đội và người dân.
+Hơn nữa, công tác phục hồi thường diễn ra không phải trong một vài ngày mà có thể kéo dài và tốn nhiều nguồn lực hơn, cần sự điều phối cao và chặt chẽ hơn.
+Trong những tình huống như vậy, chúng ta có thể sử dụng các công cụ toán tối ưu để hỗ trợ.
+Các công cụ toán này được xây dựng với biến đầu vào cả bên trong và bên ngoài ngành điện, cả nguy cơ, tổn thương, và tác động của nguy cơ gây ra, cũng như nguồn lực trong một khoảng thời gian cụ thể.
+Hình \ref{fig:restoration} biểu diễn mối quan hệ tương hỗ lẫn nhau trong công tác phục hồi lưới điện khi có sự cố xảy ra do tác động của thiên tai.
+Phân tích dữ liệu lớn có thể được sử dụng để dự đoán các sự cố thảm khốc do các sự kiện thời tiết khắc nghiệt như bão, lốc xoáy, lũ lụt và sóng thần.
+
+\FloatBarrier
+\begin{figure}[h]
+	\centering
+	\begin{center}
+		\includegraphics[width=14cm]{restoration}% This is a *.eps file
+	\end{center}
+	\caption{Phục hồi mạng lưới điện khi có sự cố gây ra bởi thiên tai}\label{fig:restoration}
+\end{figure}
+\FloatBarrier
 
@@
 
-Ứng dụng này có nhiều trường hợp sử dụng khác nhau:
-
+Ứng dụng phân tích dữ liệu lớn cho dữ liệu thu thập từ các công tơ thông minh có thể là
 \begin{itemize}
-\item Các ứng dụng của phân tích một đồng hồ thông minh duy nhất
-\item Các ứng dụng của các nhóm đồng hồ thông minh
-\item Các đồng hồ thông minh được kết nối với các mô hình lưới
+\item Các ứng dụng về phân tích dự trên một công tơ thông minh
+\item Các ứng dụng về phân tích dự trên nhiều công tơ thông minh
+\item Các công tơ thông minh được kết nối với các mô hình mạng lưới
 \end{itemize}
 
-Sự xâm nhập của các đồng hồ thông minh vào Mỹ đã vượt quá 50\%, và ở một số quốc gia châu Âu vượt quá 50\%, cung cấp cơ hội quan trọng cho phân tích dữ liệu để nâng cao quản lý khách hàng và hoạt động và kế hoạch lưới.
-Các đồng hồ thông minh cung cấp các đọc số về năng lượng, công suất và điện áp ở độ tinh granularities thường là một giờ hoặc 15 phút.
-Các công ty điện, nhà cung cấp dịch vụ năng lượng, khách hàng và nhà nghiên cứu đã xác định nhiều trường hợp sử dụng cho phân tích đồng hồ thông minh như dự báo, phân loại và phân loại tải khách hàng, ước tính tải và mô hình lưới nâng cao [40].
-Khi kết hợp với các nguồn dữ liệu khác và các hệ thống tiện ích, phân tích đồng hồ thông minh có thể mở rộng thêm lợi ích của nó cho toàn doanh nghiệp hoạt động.
-Tóm tắt của ứng dụng đồng hồ thông minh được minh họa trong Hình 5.
+Số lượng đồng hồ thông minh được lắp đặt ở Mỹ và Châu Âu đã vượt quá 50\% trên tổng số lượng công tơ điện tính tới thời điểm năm 2020.
+Điều này hiển nhiên cung cấp cơ hội lớn cho việc phân tích dữ liệu để nâng cao quản lý khách hàng và hoạt động và kế hoạch quản lý lưới điện.
+Các công tơ thông minh cung cấp các chỉ số về năng lượng, công suất và điện áp ở mức độ phân giải cao, thường là một giờ hoặc 15 phút.
+Số liệu lấy từ công tơ điện thông minh được sử dụng cho phân tích như dự báo, phân loại tải cho từng đối tượng khách hàng, ước tính tải.
+Khi kết hợp với các nguồn dữ liệu khác và các hệ thống tiện ích, phân tích dựa trên dữ liệu từ công tơ thông minh có thể được mở rộng và tạo thêm lợi ích cho hoạt động của doanh nghiệp.
+%Tóm tắt của ứng dụng đồng hồ thông minh được minh họa trong Hình 5.
 
-Dưới đây, chúng tôi cung cấp một tóm tắt các ứng dụng nổi bật của thông minh phân tích dữ liệu đồng hồ với các tham chiếu tương ứng.
+%Dưới đây, chúng tôi cung cấp một tóm tắt các ứng dụng nổi bật của thông minh phân tích dữ liệu đồng hồ với các tham chiếu tương ứng.
 Một số ứng dụng sẽ được mở rộng trong các phần tiếp theo dưới các tiêu đề riêng biệt.
 
-\subsubsection{Dự báo tải}
-Ngành công nghiệp điện đã sử dụng dự báo tải cho hoạt động và kế hoạch lưới và quản lý khách hàng.
+%\subsubsection{Dự báo tải}
+%Ngành công nghiệp điện đã sử dụng dự báo tải cho hoạt động và kế hoạch lưới và quản lý khách hàng.
 Dữ liệu đo đạc thông minh đã thu hút sự chú ý của các nhà nghiên cứu trong thập kỷ qua cho cả dự báo tải điểm và xác suất [138,139].
@@
 
-\subsubsection{Hồ sơ tiêu thụ điện của khách hàng}
+%\subsubsection{Hồ sơ tiêu thụ điện của khách hàng}
 
-Load profiling đề cập đến phân loại các đọc lịch sử về nhu cầu của khách hàng thành các nhóm dựa trên hành vi của họ.
+%Load profiling đề cập đến phân loại các đọc lịch sử về nhu cầu của khách hàng thành các nhóm dựa trên hành vi của họ.
 Các kỹ thuật phân cụm như k-means, phân cụm phân cấp và bản đồ tự tổ chức đã được sử dụng cho việc phân loại tải [140-142].
@@
 
-\subsubsection{Phân tích DER}
-Một lượng đáng kể các nguồn năng lượng phân tán (DER) đang được kết nối với lưới điện, bao gồm điện mặt trời PV, lưu trữ năng lượng và xe điện.
+%\subsubsection{Phân tích DER}
+%Một lượng đáng kể các nguồn năng lượng phân tán (DER) đang được kết nối với lưới điện, bao gồm điện mặt trời PV, lưu trữ năng lượng và xe điện.
 Những nguồn tài nguyên này trong lưới tạo ra những thách thức mới cho các nhà cung cấp tiện ích bao gồm biến động điện áp [144], vi phạm giới hạn nhiệt, dòng ngược và ảnh hưởng đến tuổi thọ dự kiến của cơ sở hạ tầng như máy biến áp và bộ điều chỉnh điện áp.
@@
 
-\subsubsection{Các ứng dụng với mạng lưới điện}
-Dữ liệu đồng hồ thông minh có thể được sử dụng kết hợp với dữ liệu nguồn cấp phân phối để có được các mô hình tinh vi cho kế hoạch phân phối hoặc để có được cái nhìn sâu sắc vào các vấn đề mô hình hóa cụ thể.
+%\subsubsection{Các ứng dụng với mạng lưới điện}
+%Dữ liệu đồng hồ thông minh có thể được sử dụng kết hợp với dữ liệu nguồn cấp phân phối để có được các mô hình tinh vi cho kế hoạch phân phối hoặc để có được cái nhìn sâu sắc vào các vấn đề mô hình hóa cụ thể.
 Ví dụ, dữ liệu đồng hồ thông minh đã được sử dụng để phát hiện bất thường [148] như thay đổi đáng kể về nhu cầu hoặc điện áp.
@@
 
-\subsubsection{Dự báo tải}
-Trong một cuộc khảo sát phân tích tiện ích được tiến hành vào năm 2017 với 136 tiện ích từ 24 quốc gia, 52\% số người trả lời coi dự báo năng lượng là một ứng dụng ưu tiên cao nhất, tỷ lệ cao nhất trong tất cả các ứng dụng trong cuộc khảo sát [153].
+%\subsubsection{Dự báo tải}
+%Trong một cuộc khảo sát phân tích tiện ích được tiến hành vào năm 2017 với 136 tiện ích từ 24 quốc gia, 52\% số người trả lời coi dự báo năng lượng là một ứng dụng ưu tiên cao nhất, tỷ lệ cao nhất trong tất cả các ứng dụng trong cuộc khảo sát [153].
 Cải thiện dự báo có thể dẫn đến các quyết định vận hành và kế hoạch tốt hơn và do đó dẫn đến tiết kiệm tiền tệ hoặc nâng cao độ tin cậy của hệ thống.
@@
 
-Các tiện ích đã thực hành dự báo tải trong hơn một thế kỷ [155].
+%Các tiện ích đã thực hành dự báo tải trong hơn một thế kỷ [155].
 Theo sự phát triển của lưới điện và ngành công nghiệp điện, dự báo tải dài hạn, hoặc dự báo tải không gian, đặc biệt là đã được coi là một thành phần quan trọng trong kế hoạch hệ thống điện vào cuối thế kỷ 20 đến đầu thế kỷ 21 [156-158], khi dữ liệu tải được thu thập từ các biến áp phân phối và ở độ phân giải thấp, chẳng hạn như hàng tháng hoặc hàng năm.
@@
 
-Khi các công nghệ tự động hóa phân phối và lưới điện thông minh cung cấp dữ liệu độ phân giải không gian thời gian cao cho các nhà dự báo tải, nghiên cứu về dự báo tải cũng phát triển mạnh.
+%Khi các công nghệ tự động hóa phân phối và lưới điện thông minh cung cấp dữ liệu độ phân giải không gian thời gian cao cho các nhà dự báo tải, nghiên cứu về dự báo tải cũng phát triển mạnh.
 Ví dụ, các nhà cung cấp điện bán lẻ bắt đầu sử dụng dữ liệu hàng giờ cho dự báo tải dài hạn.
@@
 
-Trong khi khai thác việc sử dụng dữ liệu tải có độ phân giải cao, các nhà nghiên cứu và chuyên gia cũng đầu tư một số nỗ lực vào các cách sáng tạo để khai thác dữ liệu thời tiết.
+%Trong khi khai thác việc sử dụng dữ liệu tải có độ phân giải cao, các nhà nghiên cứu và chuyên gia cũng đầu tư một số nỗ lực vào các cách sáng tạo để khai thác dữ liệu thời tiết.
 Truyền thống, chỉ có một số lượng nhỏ các trạm thời tiết được sử dụng để dự báo tải trên một khu vực cụ thể.
@@
 
-Trong khi xu hướng nghiên cứu dự báo tải chính là tận dụng dữ liệu lớn, một xu hướng khác là nhúng thông tin toàn diện về tương lai vào dự báo tải.
+%Trong khi xu hướng nghiên cứu dự báo tải chính là tận dụng dữ liệu lớn, một xu hướng khác là nhúng thông tin toàn diện về tương lai vào dự báo tải.
 Dự báo tải xác suất, cung cấp dự báo theo phân vị, khoảng xác suất hoặc hàm mật độ, là chủ đề nóng trong thập kỷ qua [168].
@@
 
-Ngoài việc cải thiện dự báo tải ở mức điện áp cao, các nhà nghiên cứu và chuyên gia cũng đã dành nhiều nỗ lực cho dự báo tải ở mức điện áp trung bình và thấp.
+%Ngoài việc cải thiện dự báo tải ở mức điện áp cao, các nhà nghiên cứu và chuyên gia cũng đã dành nhiều nỗ lực cho dự báo tải ở mức điện áp trung bình và thấp.
 Một số dự báo tại mức điểm cung cấp [42, 44], trong khi những người khác là ở các đồng hồ đo riêng lẻ [40, 139, 177].
@@
 \subsection{Dự báo và phân tích năng lượng tái tạo}
-Việc triển khai năng lượng tái tạo đã tiếp tục được duy trì với tốc độ ổn định trong thập kỷ qua.
+%Việc triển khai năng lượng tái tạo đã tiếp tục được duy trì với tốc độ ổn định trong thập kỷ qua.
 Do tính không đồng nhất của việc tạo điện từ các nguồn năng lượng tái tạo, cũng như do sự khó lường hạn chế, sự chú trọng đã được đặt vào việc phát triển các phương pháp tích hợp những nguồn tái tạo đó.
@@
 
-Hôm nay cho các trang trại gió, đặc biệt là các trang trại ngoài khơi, việc thu thập dữ liệu ở mức độ động cơ gió được thực hiện với độ phân giải một giây.
-Tương tự, đối với các nhà máy năng lượng mặt trời, dữ liệu có thể được thu thập ở mức độ biến tần và với độ phân giải cùng cấp độ giây.
-Những dữ liệu ở mức độ rất tinh vi này được sử dụng để cải thiện phân tích và dự báo tại mức độ trang trại gió (hoặc nhà máy năng lượng mặt trời) [186].
-Ngoài ra, vì khả năng sinh năng lượng từ các nguồn tái tạo bắt đầu trở nên đa dạng và phân tán địa lý một cách dày đặc, ta cũng có thể sử dụng tất cả dữ liệu được thu thập tại các trang trại để cải thiện dự báo [14,18,187].
+Ngày nay tại các trang trại điện gió, đặc biệt là các trang trại điện gió ngoài khơi, việc thu thập dữ liệu ở mức độ động cơ gió được thực hiện với độ phân giải một giây.
+Tương tự, đối với các nhà máy năng lượng mặt trời, dữ liệu có thể được thu thập ở mức độ biến tần và với độ phân giải cùng cấp độ giây \citep{Gilbert2020}.
+Những dữ liệu ở mức độ rất tinh vi này được sử dụng để cải thiện phân tích và dự báo.
+Ngoài ra, vì khả năng sinh năng lượng từ các nguồn tái tạo bắt đầu trở nên đa dạng và phân tán địa lý một cách dày đặc, ta cũng có thể sử dụng tất cả dữ liệu được thu thập tại các trang trại để cải thiện dự báo.
 
 Ngoài ra, còn có rất nhiều nguồn dữ liệu khác có liên quan, chủ yếu là liên quan đến quan sát và dự báo khí tượng, mô tả các quy trình phức tạp và mang lại khối lượng dữ liệu rất lớn.
-Về phía quan sát khí tượng:
+%Về phía cạnh quan sát khí tượng:
 
 \begin{itemize}
-\item Sky imagers có tiềm năng lớn cho mô hình hóa và dự báo năng lượng mặt trời với độ phân giải cao vì chúng theo dõi các đám mây di chuyển và tác động của chúng đến các tấm pin mặt trời [188].
+\item Sky imagers có tiềm năng lớn cho mô hình hóa và dự báo năng lượng mặt trời với độ phân giải cao vì chúng theo dõi các đám mây di chuyển và tác động của chúng đến các tấm pin mặt trời \citep{Chow2011}.
 Chúng có thể cung cấp hình ảnh của bầu trời trên các nhà máy điện mặt trời mỗi 30 giây;
 
-\item Radar thời tiết cũng đã chứng minh sự quan tâm của họ trong đánh giá và mô hình các chế độ động để ứng dụng vào dự báo năng lượng gió độ phân giải cao [189].
+\item Radar thời tiết cũng đã chứng minh tầm ảnh hưởng trong đánh giá và mô hình hóa các chế độ làm việc của các trang trại điện gió.
 Tùy thuộc vào công nghệ, hình ảnh radar có thể có sẵn giữa mỗi phút và mỗi 10 phút, với bán kính hình ảnh từ 60 đến 250 km;
 
-\item LIDAR ngày càng được coi là rất liên quan đến các đo lường gió ở phía trước của các tuabin gió và tích hợp vào các phương pháp dự báo [190], hoặc nói chung là các quan sát tiềm năng mới của các hồ sơ gió được sử dụng trong dự báo thời tiết và năng lượng tái tạo [191].
+\item Dữ liệu từ LIDAR ngày càng được coi là rất liên quan đến các đo lường gió ở phía trước của các tuabin gió và chúng được tích hợp vào các phương pháp dự báo, hoặc nói chung là các quan sát tiềm năng mới về gió được sử dụng trong dự báo thời tiết và năng lượng tái tạo.
 LIDAR cung cấp các đo lường gió cho hình nón mà chúng quét (theo chiều dọc hoặc chiều ngang, tùy thuộc vào cách chúng được thiết lập) mỗi vài giây.
@@
 
-Một thách thức đầu tiên và phức tạp khi xử lý thông tin cảm biến từ xa là giảm chiều.
+%Một thách thức đầu tiên và phức tạp khi xử lý thông tin cảm biến từ xa là giảm chiều.
 Điều này có thể được thực hiện: (i) dựa trên các kỹ thuật xử lý tín hiệu và thống kê, ví dụ như Phân tích thành phần độc lập-ICA, cho các trường chuyển động trong hình ảnh radar thời tiết; (ii) bằng cách trích xuất các đặc trưng vật lý như mây trong hình ảnh bầu trời [188] hoặc hệ thống mưa và đặc điểm của chúng trong hình ảnh radar [192]; hoặc cuối cùng (iii) thông qua các mô hình chức năng như các hồ sơ gió cho các đo đạc dọc theo chiều dọc của LiDAR.
 
-Ngoài các quan sát khí tượng, dữ liệu đầu vào chiều cao mới có thể có dạng dự báo thời tiết.
+%Ngoài các quan sát khí tượng, dữ liệu đầu vào chiều cao mới có thể có dạng dự báo thời tiết.
 Trước hết, thông tin liên quan trong dự báo thời tiết không chỉ có thể dành cho điểm gần nhất với một điểm quan tâm mà còn có thể được cung cấp trên toàn khu vực của điểm đó [59].
@@
 Hiện nay, sự sẵn có của số lượng lớn dữ liệu này yêu cầu những thay đổi cơ bản trong phân tích và dự báo năng lượng tái tạo, không chỉ về phương pháp mà còn về các mô hình kinh doanh.
-Chúng ta mong đợi sẽ có nhiều công trình đổi mới xuất hiện trong tương lai, đề xuất các phương pháp dựa trên phương trình vi phân ngẫu nhiên, học sâu, học phân tán và liên kết, cũng như thị trường dữ liệu.
+Chúng ta mong đợi sẽ có nhiều công trình đổi mới xuất hiện trong tương lai, đề xuất các phương pháp dựa trên phương trình vi phân ngẫu nhiên, học sâu, học phân tán và liên kết, cũng như dữ liệu về thị trường.
 
 \subsection{Hiệu quả và tối ưu hóa năng lượng}
-Các công nghệ lưới thông minh cung cấp tiềm năng lớn để tăng cường hiệu quả sử dụng năng lượng trong các lĩnh vực khác nhau.
-Tuy nhiên, hiện tại, các hoạt động tăng cường hiệu quả năng lượng chủ yếu bị hạn chế trong việc thực thi chứng nhận ISO 50001 như sau: (i) lắp đặt thiết bị bổ sung (đồng hồ đo, cảm biến, v.v.) để đo lường tiêu thụ năng lượng; (ii) lắp đặt phần cứng mới và thay thế thiết bị; và (iii) trực quan hóa dữ liệu trong giao diện người dùng liền mạch, tìm các mô hình bất thường; và xác định các quy trình tiêu thụ năng lượng cao.
+Các công nghệ về lưới điện thông minh cung cấp tiềm năng lớn để tăng cường hiệu quả sử dụng năng lượng trong các lĩnh vực khác nhau.
+Tuy nhiên, hiện tại, các hoạt động tăng cường hiệu quả năng lượng chủ yếu bị hạn chế trong việc thực thi tiêu chuẩn ISO 50001 như: (i) lắp đặt thiết bị bổ sung (đồng hồ đo, cảm biến, v.v.) để đo lường tiêu thụ năng lượng; (ii) lắp đặt phần cứng mới và thay thế thiết bị; và (iii) trực quan hóa dữ liệu, tìm các mô hình bất thường; và xác định các quy trình tiêu thụ năng lượng cao.
 Thực tiễn tiêu chuẩn này cung cấp việc giám sát và nhận thức về tiêu thụ năng lượng cho các quyết định của con người, nhưng nó không cho phép phân tích theo chỉ đạo và kiểm soát quy trình tự động.
 
-Các tác phẩm khác nhau trong văn học khai thác các kỹ thuật tối ưu hóa năng lượng dựa trên mô hình, chẳng hạn như lập trình số nguyên cho việc giảm tải đỉnh trong các nhà máy thép [195] và lập trình tuyến tính số hỗn hợp cho các thiết bị gia dụng nhiệt [196].
+%Các tác phẩm khác nhau trong văn học khai thác các kỹ thuật tối ưu hóa năng lượng dựa trên mô hình, chẳng hạn như lập trình số nguyên cho việc giảm tải đỉnh trong các nhà máy thép [195] và lập trình tuyến tính số hỗn hợp cho các thiết bị gia dụng nhiệt [196].
 Các phương pháp dựa trên mô hình có nhược điểm là yêu cầu một mô hình (toán học) về quá trình vật lý và các ràng buộc, có thể phức tạp trong một số trường hợp.
@@
 Sự xuất hiện của công nghệ internet-of-things cung cấp điều kiện kỹ thuật cho mô hình hóa động lực dữ liệu trong tối ưu hóa năng lượng.
-Một số ví dụ là việc sử dụng học tăng cường dạng batch (fitted Q-iteration) để điều khiển một nhóm máy sưởi nước điện gia đình cho dịch vụ đáp ứng nhu cầu [197]; fitted Q-iteration kết hợp với auto-encoders để tối ưu hóa năng lượng trong máy sưởi nước điện [198]; học tăng cường sâu cho tối ưu hóa năng lượng dự đoán các trạm bơm nước thải [199]; và mô hình cây cho tiêu thụ năng lượng của tòa nhà để lên lịch hệ thống sưởi ấm tối ưu [200].
+Một số ví dụ là việc sử dụng học tăng cường dạng batch (fitted Q-iteration) để điều khiển một nhóm máy sưởi nước điện gia đình cho dịch vụ đáp ứng nhu cầu \citep{Ruelens2014}; fitted Q-iteration kết hợp với auto-encoders để tối ưu hóa năng lượng trong máy sưởi nước điện \citep{Ruelens2018}; học tăng cường sâu cho tối ưu hóa năng lượng dự đoán các trạm bơm nước thải \citep{Filipe2019}; và mô hình cây cho tiêu thụ năng lượng của tòa nhà để lên lịch hệ thống sưởi ấm tối ưu \citep{Kouzelis2015}.
 Phương pháp này không yêu cầu việc mô hình hoá đầy đủ các phương trình quá trình vì sự hiểu biết của nó được thực hiện theo thời gian thực thông qua dữ liệu.
@@
 
-Ngoài tối ưu hóa năng lượng dữ liệu, các dịch vụ năng lượng dựa trên dữ liệu khác có thể được cung cấp bởi các công ty dịch vụ năng lượng (ESCO), nhà tập đoàn và nhà bán lẻ nhằm tối đa hóa nhận thức của người sử dụng cuối đối với các hành động tăng hiệu quả sử dụng năng lượng và rút ra giá trị kinh doanh từ dữ liệu.
+%Ngoài tối ưu hóa năng lượng dữ liệu, các dịch vụ năng lượng dựa trên dữ liệu khác có thể được cung cấp bởi các công ty dịch vụ năng lượng (ESCO), nhà tập đoàn và nhà bán lẻ nhằm tối đa hóa nhận thức của người sử dụng cuối đối với các hành động tăng hiệu quả sử dụng năng lượng và rút ra giá trị kinh doanh từ dữ liệu.
 
-Một đóng góp quan trọng để tăng sự nhận thức của người sử dụng cuối là giám sát tải không xâm phạm từ dữ liệu của các công tơ thông minh, có thể được áp dụng để phát hiện và ước tính các cài đặt điện PV dân dụng [145], tiêu thụ bơm nhiệt [201] và các hồ sơ tải cá nhân từ các đường cong tải nguồn [202].
+%Một đóng góp quan trọng để tăng sự nhận thức của người sử dụng cuối là giám sát tải không xâm phạm từ dữ liệu của các công tơ thông minh, có thể được áp dụng để phát hiện và ước tính các cài đặt điện PV dân dụng [145], tiêu thụ bơm nhiệt [201] và các hồ sơ tải cá nhân từ các đường cong tải nguồn [202].
 Phân đoạn khách hàng với các hồ sơ tải có thể được áp dụng bởi các nhà tập đoàn và nhà bán lẻ để xác định khách hàng mục tiêu, thiết kế một loại giá động hoặc thời gian thực tùy chỉnh [203], mô hình hóa hành vi động của các tải/công tắc có thể kiểm soát được [204], hoặc thông báo cho người tiêu dùng nếu họ đang đối mặt với một sự thay đổi không bình thường của hồ sơ tải.
@@
 
-Dữ liệu từ đồng hồ đo thông minh kết hợp với các biến môi trường bên ngoài (ví dụ: nhiệt độ ngoài trời) có thể hỗ trợ cho các nhà bán lẻ và nhà tổng hợp ước tính khả năng phản ứng yêu cầu của khách hàng.
+%Dữ liệu từ đồng hồ đo thông minh kết hợp với các biến môi trường bên ngoài (ví dụ: nhiệt độ ngoài trời) có thể hỗ trợ cho các nhà bán lẻ và nhà tổng hợp ước tính khả năng phản ứng yêu cầu của khách hàng.
 Trong [206], một vấn đề bao lưu ngẫu nhiên được đặt ra để lựa chọn khách hàng trong các chương trình phản ứng yêu cầu sử dụng dữ liệu tiêu thụ, cũng như để ước tính xác suất đạt được mục tiêu giảm tải.
@@
 
-Dữ liệu được thu thập từ đánh giá và chứng nhận hiệu quả năng lượng rất có giá trị cho các bên liên quan và dịch vụ khác nhau [211].
+%Dữ liệu được thu thập từ đánh giá và chứng nhận hiệu quả năng lượng rất có giá trị cho các bên liên quan và dịch vụ khác nhau [211].
 Ví dụ, Cơ quan Năng lượng (DOE) cơ sở dữ liệu hiệu suất tòa nhà có thể được sử dụng cho các mục đích khác nhau [212]: so sánh hiệu suất tiết kiệm năng lượng của các loại tòa nhà khác nhau và địa điểm địa lý; ước tính tiềm năng tiết kiệm năng lượng liên quan đến các hành động cải tiến cụ thể; và đánh giá tác động của các công nghệ tiết kiệm năng lượng ở mức độ danh mục.
@@
 
-\subsection{Phân tích dữ liệu synchrophasor và sự kiện}
-Để đánh giá tầm quan trọng của dữ liệu Đơn vị Đo Pha (PMU), ta phải trở về các báo cáo về nguyên nhân của sự cố mất điện lớn tại đông bắc Hoa Kỳ vào ngày 13 tháng 8 năm 2003 [214].
+%\subsection{Phân tích dữ liệu synchrophasor và sự kiện}
+%Để đánh giá tầm quan trọng của dữ liệu Đơn vị Đo Pha (PMU), ta phải trở về các báo cáo về nguyên nhân của sự cố mất điện lớn tại đông bắc Hoa Kỳ vào ngày 13 tháng 8 năm 2003 [214].
 Ban Nghiên cứu Sự cố Hệ thống Điện quốc tế Hoa Kỳ-Canada đã kết luận rằng một trong những lý do chính dẫn đến sự cố mất điện là thiếu nhận thức về tình hình thực tế.
@@
 
-Một trong những điểm mạnh chính của PMU là cung cấp các phép đo đồng bộ thời gian của phasor điện áp và dòng điện thời gian thực với tốc độ báo cáo cao hơn đáng kể so với các Đơn vị Điều khiển Từ xa SCADA (RTU).
+%Một trong những điểm mạnh chính của PMU là cung cấp các phép đo đồng bộ thời gian của phasor điện áp và dòng điện thời gian thực với tốc độ báo cáo cao hơn đáng kể so với các Đơn vị Điều khiển Từ xa SCADA (RTU).
 Có thể hiểu được cách thức thực hiện này thông qua Hình 6, trong đó hiển thị kiến trúc tổng quát của PMU.
@@
 
-Điều này cung cấp một cơ hội độc đáo cho phân tích sau sự kiện, đặc biệt là trong trường hợp các sự kiện phức tạp (ví dụ: lỗi lan truyền do sự cố thiết bị dẫn đến mất điện).
+%Điều này cung cấp một cơ hội độc đáo cho phân tích sau sự kiện, đặc biệt là trong trường hợp các sự kiện phức tạp (ví dụ: lỗi lan truyền do sự cố thiết bị dẫn đến mất điện).
 Ngoài ra, việc bắt và phân tích các dao động trong hệ thống điện quan trọng vì nó báo hiệu về khả năng sụp đổ của hệ thống [216].
@@
 
-Do đó, dữ liệu được thu thập từ PMU đóng một vai trò quan trọng trong các ứng dụng như giám sát, điều khiển, bảo vệ, ước lượng trạng thái, đánh giá sự ổn định và phát hiện sự cố [221].
+%Do đó, dữ liệu được thu thập từ PMU đóng một vai trò quan trọng trong các ứng dụng như giám sát, điều khiển, bảo vệ, ước lượng trạng thái, đánh giá sự ổn định và phát hiện sự cố [221].
 Tuy nhiên, phân tích sau sự kiện vẫn được thực hiện bằng tay trong nhiều trường hợp với khả năng dự đoán hạn chế hoặc không có.
@@
 
-Phần lớn các phương pháp dự đoán cho phân tích dữ liệu PMU trong văn bản tập trung vào việc cho phép nhận thức tình huống có ý nghĩa hơn so với những gì được bao phủ bởi EMS SCADA, từ đó đảm bảo sự ổn định động của hệ thống [222-229].
+%Phần lớn các phương pháp dự đoán cho phân tích dữ liệu PMU trong văn bản tập trung vào việc cho phép nhận thức tình huống có ý nghĩa hơn so với những gì được bao phủ bởi EMS SCADA, từ đó đảm bảo sự ổn định động của hệ thống [222-229].
 Các phương pháp dự đoán có thể mở rộng khả năng phân biệt các loại sự kiện của các nhà điều hành, từ hoạt động bình thường đến hoạt động trong điều kiện cực đoan bằng cách bắt giữ các đặc trưng dạng sóng PMU và tự động xác định các sự kiện trong quá trình phân loại.
@@
 
-Một thách thức mà tất cả các đơn vị liên quan đến phát triển các ứng dụng trực tuyến và ngoại tuyến đều đang đối mặt là sự gia tăng của một lượng dữ liệu rất lớn đến từ PMUs.
+%Một thách thức mà tất cả các đơn vị liên quan đến phát triển các ứng dụng trực tuyến và ngoại tuyến đều đang đối mặt là sự gia tăng của một lượng dữ liệu rất lớn đến từ PMUs.
 Để cải thiện tính tin cậy, bảo mật và hiệu quả của hệ thống, cần phát triển các công cụ tự động có khả năng phân tích các sự kiện quá khứ và thông báo quyết định trong thời gian thực.
@@
 
-Khi tài nguyên tính toán tiến bộ, việc học và trích xuất các mẫu hữu ích từ dữ liệu lớn tạo ra cơ hội mới.
+%Khi tài nguyên tính toán tiến bộ, việc học và trích xuất các mẫu hữu ích từ dữ liệu lớn tạo ra cơ hội mới.
 Học sâu [88,92,93] cho phép tìm thấy các mẫu trừu tượng hơn từ dữ liệu phức tạp lớn, thậm chí là không đồng nhất [230].
@@
 
-Kiến trúc tổng quan của quá trình phân tích dữ liệu PMU tự động được trình bày trong Hình 6.
+%Kiến trúc tổng quan của quá trình phân tích dữ liệu PMU tự động được trình bày trong Hình 6.
 
-\subsection{Các ứng dụng cho lưới điện}
-Sự đầu tư vào các công nghệ Lưới điện Thông minh (ví dụ như đồng hồ thông minh, PMU và các thiết bị điện tử thông minh [IED]) cho lưới điện phân phối và truyền tải đang cải thiện khả năng giám sát và điều khiển của chúng [233].
+%\subsection{Các ứng dụng cho lưới điện}
+%Sự đầu tư vào các công nghệ Lưới điện Thông minh (ví dụ như đồng hồ thông minh, PMU và các thiết bị điện tử thông minh [IED]) cho lưới điện phân phối và truyền tải đang cải thiện khả năng giám sát và điều khiển của chúng \citep{Simoes2012}.
 Đồng thời, các khung pháp lý và thị trường mới đang được thử nghiệm (ở cấp độ thử nghiệm) và triển khai ở các quốc gia khác nhau để hỗ trợ tích hợp các nguồn điện phân tán linh hoạt.
 
-\subsubsection{Tính quan sát}
-Mặc dù có tất cả những tiến bộ công nghệ, khả năng quan sát lưới điện áp thấp (LV) vẫn là một trở ngại chính để khai thác tối đa tiềm năng từ những công nghệ này và tích hợp các nguồn điện phân tán linh hoạt vào thị trường điện và quản lý hệ thống điện.
-Điều này có thể được chia thành hai thách thức chính: (i) đặc trưng của tôpô và các thông số lưới điện; và (ii) giám sát thời gian thực.
+%\subsubsection{Tính quan sát}
+%Mặc dù có tất cả những tiến bộ công nghệ, khả năng quan sát lưới điện áp thấp (LV) vẫn là một trở ngại chính để khai thác tối đa tiềm năng từ những công nghệ này và tích hợp các nguồn điện phân tán linh hoạt vào thị trường điện và quản lý hệ thống điện.
+Điều này có thể được chia thành hai thách thức chính: (i) đặc trưng của địa hình (topo) và các thông số lưới điện; và (ii) giám sát thời gian thực.
 
-Thông tin từ các đồng hồ thông minh được lắp đặt tại khách hàng áp thấp và trạm phân phối cấp hai có thể được sử dụng để xây dựng tôpô lưới sử dụng nhiều phương pháp khác nhau, chẳng hạn như: mô hình đồ thị xác suất và hồi quy tuyến tính LASSO [234, 235]; và luồng điện kết hợp với thông tin chung [151].
+%Thông tin từ các đồng hồ thông minh được lắp đặt tại khách hàng áp thấp và trạm phân phối cấp hai có thể được sử dụng để xây dựng topo lưới sử dụng nhiều phương pháp khác nhau, chẳng hạn như: mô hình đồ thị xác suất và hồi quy tuyến tính LASSO [234, 235]; và luồng điện kết hợp với thông tin chung [151].
 Trong một số trường hợp, thông tin về tôpô lưới có sẵn, nhưng các thông số điện có sai số lớn hoặc không biết (tức là sử dụng các thông số từ các danh mục dây cáp), và do đó, cần thực hiện một ước lượng chắc chắn (và dựa trên dữ liệu) của các thông số lưới [236, 237].
@@
 
-Về mặt giám sát thời gian thực, thông tin về hồ sơ điện áp là rất quan trọng để tạo ra các cảnh báo cho các nhân viên vận hành tại các trung tâm phân phối điện áp thấp.
+%Về mặt giám sát thời gian thực, thông tin về hồ sơ điện áp là rất quan trọng để tạo ra các cảnh báo cho các nhân viên vận hành tại các trung tâm phân phối điện áp thấp.
 Tuy nhiên, ngay cả với các giao thức truyền thông như Power Line Communication (PLC) PRIME hoặc General Packet Radio Service (GRPS), việc thu thập thông tin 15 phút một lần từ hàng trăm khách hàng là không khả thi hoặc chi phí truyền thông rất cao.
@@
 
-Các mức điện áp khác cũng được hưởng lợi từ các chức năng ước tính trạng thái dựa trên dữ liệu để giám sát thời gian thực.
+%Các mức điện áp khác cũng được hưởng lợi từ các chức năng ước tính trạng thái dựa trên dữ liệu để giám sát thời gian thực.
 Một số ví dụ bao gồm: một chức năng ước tính trạng thái dựa trên dữ liệu mạnh mẽ được đề xuất trong [244] để khai thác dữ liệu lịch sử được thu thập bởi các đồng hồ đo phân phối và PMU, bằng cách tìm các đặc trưng tương đồng trong một khung làm việc học có giám sát với hồi quy kernel ridge; một bộ ước tính trạng thái động có khả năng dự báo được mô tả trong [245]; kết hợp dữ liệu SCADA từ các trạm phân phối chính (thông thường có sẵn trong cơ sở dữ liệu trung tâm với độ trễ vài giây/phút) và trạm phân phối phụ (thông thường có sẵn trong cơ sở dữ liệu trung tâm với độ trễ một vài giờ) có thể suy ra điều kiện vận hành gần thời gian thực của lưới điện áp trung bình (MV) [246,247]; và một khối ghép của các bộ tự mã hóa cạnh tranh địa phương ước tính trạng thái của các công tắc lưới dựa trên một tập hợp các đo lường điện cục bộ [248].
@@
 
-\subsubsection{Hộ trợ ra quyết định và điều khiển}
-Các chức năng dựa trên dữ liệu cũng đang được tích hợp vào các hệ thống quản lý năng lượng truyền thống (EMS) và quản lý phân phối (DMS) để bổ sung hoặc thay thế các chức năng phân tích lưới điện cổ điển như lưu lượng điện, cấu hình mạng, lưu lượng điện tối ưu, vv.
+%\subsubsection{Hộ trợ ra quyết định và điều khiển}
+%Các chức năng dựa trên dữ liệu cũng đang được tích hợp vào các hệ thống quản lý năng lượng truyền thống (EMS) và quản lý phân phối (DMS) để bổ sung hoặc thay thế các chức năng phân tích lưới điện truyền thống như lưu lượng điện, cấu hình mạng, lưu lượng điện tối ưu, vv.
 Trong một số trường hợp, một mô hình học máy được sử dụng như là một proxy cho các chức năng truyền thống, ví dụ như ước tính lưu lượng điện của ma trận Jacobian với dữ liệu PMU [249] và tính toán cam kết đơn vị [250].
@@
 
-Trong những trường hợp khác, học máy được sử dụng để học từ dữ liệu lịch sử về các hành động điều khiển của các nhà điều hành và hỗ trợ ra quyết định [251] hoặc để khám phá các giải pháp mới (và tốt hơn) bằng cách sử dụng các hệ thống chuyên gia [252].
+%Trong những trường hợp khác, học máy được sử dụng để học từ dữ liệu lịch sử về các hành động điều khiển của các nhà điều hành và hỗ trợ ra quyết định [251] hoặc để khám phá các giải pháp mới (và tốt hơn) bằng cách sử dụng các hệ thống chuyên gia [252].
 Một lợi thế quan trọng của hai phương pháp này, được phát triển cho Chủ sở hữu Hệ thống Truyền tải (TSOs), là tính giải thích cao đối với những người ra quyết định và khả năng khai thác kiến thức chuyên gia (tức là quyết định quá khứ và các heuristics điều khiển).
@@
 
-Một số trường hợp sử dụng cho phương pháp điều khiển dữ liệu này là: xếp hạng nhanh các trường hợp dự phòng cấp cao hơn (theo nguy cơ của chúng) để ưu tiên mô phỏng hệ thống điện năng [254]; và phân tích nhân quả trên dữ liệu đo lường để thực hiện các chiến lược tấn công node tối ưu [255].
+%Một số trường hợp sử dụng cho phương pháp điều khiển dữ liệu này là: xếp hạng nhanh các trường hợp dự phòng cấp cao hơn (theo nguy cơ của chúng) để ưu tiên mô phỏng hệ thống điện năng [254]; và phân tích nhân quả trên dữ liệu đo lường để thực hiện các chiến lược tấn công node tối ưu [255].
 
-Cuối cùng, việc tích hợp các hệ thống năng lượng tái tạo và sự áp dụng rộng rãi của IEC 61850 đang tăng thêm khối lượng dữ liệu cảnh báo và số lượng cảnh báo đòi hỏi sự chú ý và hành động điều khiển trong phòng điều khiển [256].
+%Cuối cùng, việc tích hợp các hệ thống năng lượng tái tạo và sự áp dụng rộng rãi của IEC 61850 đang tăng thêm khối lượng dữ liệu cảnh báo và số lượng cảnh báo đòi hỏi sự chú ý và hành động điều khiển trong phòng điều khiển [256].
 Các phương pháp truyền thống dựa trên hệ thống chuyên gia dựa trên quy tắc [257], nhưng thiếu khả năng cung cấp những thông tin bổ ích về các thông tin chứa trong chuỗi cảnh báo và giảm tải trí nhớ của nhân viên điều khiển.
@@
 
-Cuộc khảo sát cho thấy có nhiều ứng dụng đang ở các giai đoạn thực hiện thực tế khác nhau.
+%Cuộc khảo sát cho thấy có nhiều ứng dụng đang ở các giai đoạn thực hiện thực tế khác nhau.
 
-Trong khi số lượng tài liệu tham khảo về chủ đề tổng quan về dữ liệu lớn đã rất lớn, dự kiến công việc sẽ tiến triển với tốc độ chưa từng có, và số lượng tài liệu tham khảo sẽ tiếp tục tăng trong tương lai gần.
+%Trong khi số lượng tài liệu tham khảo về chủ đề tổng quan về dữ liệu lớn đã rất lớn, dự kiến công việc sẽ tiến triển với tốc độ chưa từng có, và số lượng tài liệu tham khảo sẽ tiếp tục tăng trong tương lai gần.
 
-Tuy nhiên, trong một số cuộc khảo sát đã công bố, không rõ ứng dụng cuối cùng của một số kỹ thuật được khảo sát là gì, và lợi ích và động lực kinh doanh liên quan là gì.
+%Tuy nhiên, trong một số cuộc khảo sát đã công bố, không rõ ứng dụng cuối cùng của một số kỹ thuật được khảo sát là gì, và lợi ích và động lực kinh doanh liên quan là gì.
 Chúng tôi nhận thấy lợi ích của nghiên cứu tương lai là tập trung vào các ứng dụng cuối cùng trong khi đồng thời giải quyết các vấn đề cơ bản để thúc đẩy những cơ hội sau đây cuối cùng:
+Các thách thức và cơ hội liên quan đến công tác phân tích dữ liệu lớn trong tương lai có thể được tóm tắt theo liệt kê dưới đây:
+\begin{itemize}
+\item Dự đoán các sự kiện trước thời gian và cho phép triển khai các chiến lược giảm thiểu rủi ro và tính toán rủi ro.
 
-\begin{itemize}
-\item Dự đoán các sự kiện trước thời gian và cho phép triển khai các chiến lược giảm thiểu rủi ro và tính toán rủi ro với dự báo không chắc chắn.
-
-\item Tối ưu hóa dữ liệu lịch sử cho lợi ích của chủ sở hữu và người sử dụng.
-Kết hợp mô hình vật lý và dữ liệu để cải thiện phân tích nguyên nhân gốc rễ.
-Ngăn chặn các sự cố và hạn chế vận hành hệ thống điện hiện tại gây tổn thất hàng tỷ đô la.
+\item Tối ưu hóa sử dụng dữ liệu đã thu thập cho lợi ích của doanh nghiệp và người sử dụng.
+Kết hợp mô hình vật lý và dữ liệu để cải thiện việc phân tích nguyên nhân gốc rễ của các vấn đề.
+Ngăn chặn các sự cố và hạn chế trong vận hành hệ thống điện hiện tại liên quan đến tác động và hậu quả kinh tế lớn.
 
@@
 
-\item Sử dụng dữ liệu từ nhiều nguồn phổ biến khác nhau như máy tính bảng, điện thoại thông minh và các thiết bị điện tử cá nhân khác, khiến chủ sở hữu và các bên liên quan của doanh nghiệp phân tích dữ liệu trở thành các nhân vật tham gia.
+\item Sử dụng dữ liệu từ nhiều nguồn phổ biến khác nhau như máy tính bảng, điện thoại thông minh và các thiết bị điện tử cá nhân khác, khiến doanh nghiệp và các bên liên quan của doanh nghiệp phân tích dữ liệu trở thành các chủ thể cùng tham gia.
 
 \item Thay thế một số nhiệm vụ được thực hiện bởi các nhà điều hành và chuyên gia ngày nay bằng các thuật toán và quy trình tự động hóa.
+
+\item Xây dựng được các mô phỏng không gian và thời gian liên quan đến sự xuống cấp tài sản thiết bị và mạng lưới, đồng thời tích hợp định lượng nguy cơ rủi do, tổn thương, và rủi do vào mô phỏng để đưa ra các quyết định quản lý tối ưu, đặc biệt là trong các hoàn cảnh khó khăn và tình trạng khẩn cấp.
+
+\item Xây dựng được các mô hình tích hợp công tác làm mới hay sửa chữa cho hạ tầng điện lực kết hợp với hạ tầng các ngành khác nhau \citep{Kielhauser2017}.
 \end{itemize}
 
-\section{Kết luận}
+%\section{Kết luận}
 
-Bản khảo sát này đã cố gắng đạt được một số mục tiêu sau:
-\begin{itemize}
-\item Nó đã giải quyết được sự đa dạng và chiều sâu của các ứng dụng thực tiễn về phân tích dữ liệu lớn trong lưới điện.
-\item Nó đã đánh giá lại các vấn đề chính trong việc triển khai phân tích dữ liệu lớn trong các lĩnh vực lưới được chọn.
-\item Nó cung cấp nhiều ví dụ về các xu hướng mới nhất trong khung quyết định và phân tích dự đoán.
-\item Nó liệt kê một số tài liệu tham khảo có thể được sử dụng bởi các nhà nghiên cứu để nghiên cứu của họ tiếp tục phát triển.
-\item Nó đưa ra một số hướng đi của nghiên cứu tương lai và phác thảo các thách thức và cơ hội.
+%Trong bài viết này, tác giả đã cố gắng tổng hợp và tổng kết lại một số điểm và xu hướng chính liên quan đến việc phân tích dữ liệu lớn trong ngành điện.
+Đặc biệt, bài viết có liệt kê ra danh sách các tài liệu tham khảo liên quan để người đọc dễ dàng trong công tác tìm hiểm thêm nếu có nhu cầu.
+Bài viết này còn nhiều thiếu sót
+%
+%\begin{itemize}
+%\item Nó đã giải quyết được sự đa dạng và chiều sâu của các ứng dụng thực tiễn về phân tích dữ liệu lớn trong lưới điện.
+%\item Nó đã đánh giá lại các vấn đề chính trong việc triển khai phân tích dữ liệu lớn trong các lĩnh vực lưới được chọn.
+%\item Nó cung cấp nhiều ví dụ về các xu hướng mới nhất trong khung quyết định và phân tích dự đoán.
+%\item Nó liệt kê một số tài liệu tham khảo có thể được sử dụng bởi các nhà nghiên cứu để nghiên cứu của họ tiếp tục phát triển.
+%\item Nó đưa ra một số hướng đi của nghiên cứu tương lai và phác thảo các thách thức và cơ hội.
+%
+%\end{itemize}
 
-\end{itemize}
-
-Bài báo này không phải là một tài liệu toàn diện về tất cả các tác phẩm hoặc xu hướng đã được xuất bản.
+%không phải là một tài liệu toàn diện về tất cả các tác phẩm hoặc xu hướng đã được xuất bản.
 Nếu quan tâm đến chủ đề, người đọc của cuộc khảo sát này nên khám phá các khía cạnh bổ sung không được đề cập ở đây.
@@
 
-\section*{Ghi Chú}
-Bài viết này được viết dựa trên việc tổng hợp kiến thức từ các bài báo được trích dẫn liệt kê trong phần Tài Liệu.
-Tác giả đã sử dụng tối ưu công cụ \href{https://chat.openai.com/chat}{ChatGPT} được phát triển bởi công ty \href{www.openai.com}{OpenAI} để làm công tác biên dịch và dựa vào gợi ý của ChatGPT, tác giả tiếp tục chỉnh sửa nội dung và câu chữ cho phù hợp hơn với ngôn ngữ Tiếng Việt.
-Ngoài việc sử dụng  \href{https://chat.openai.com/chat}{ChatGPT}, tác giả cũng đối chiếu phần dịch nội dung với công cụ AI được tích hợp trên phần mềm Bing và Edge của Microsoft.
+\section*{Ghi Chú Quan Trọng}
+\textit{Bài viết này được viết dựa trên việc tổng hợp kiến thức từ các bài báo được trích dẫn liệt kê trong phần Tài Liệu, đặc biệt là bài viết của các tác giả \cite{Kezunovic2020,Angadi2020}.
+Tác giả đã sử dụng tối ưu công cụ \href{https://chat.openai.com/chat}{ChatGPT} được phát triển bởi công ty \href{www.openai.com}{OpenAI} để làm công tác biên soạn cho phù hợp với người đọc là các cán bộ và nhà quản lý của các Công Ty Điện Lực EVN ở Thành Phố Hồ Chí Minh, dựa vào gợi ý của ChatGPT, tác giả tiếp tục chỉnh sửa nội dung và câu chữ cho phù hợp hơn với ngôn ngữ Tiếng Việt.
+Ngoài việc sử dụng  \href{https://chat.openai.com/chat}{ChatGPT}, tác giả cũng đối chiếu phần biên soạn nội dung với công cụ AI được tích hợp trên phần mềm Bing và Edge của Microsoft.}
 

### 2023-03-22 2023-EVN-Data-Analytics-for-CEO/paper/paper.tex
udpate

@@
 
-Ở Việt Nam, tổng công ty điện lực Việt Nam (EVN) cùng các bộ ngành liên quan đã soạn thảo và đề xuất "Qui Hoạch Điện VIII" hướng tới sự phát triển ổn định và bền vững chiến lược đến năm 2050.
-Một trong những nội dung quan trọng của Qui Điện VIII có chú trọng đến việc phát triển và nâng cao khả năng ứng dụng các công nghệ tiên tiến và áp dụng dữ liệu lớn cũng như trí thông minh nhân tạo vào công tác xây dựng, điều hành và quản lý các hệ thống điện.
+Ở Việt Nam, Tập Đoàn Điện Lực Việt Nam (EVN) cùng các bộ ngành liên quan đã soạn thảo và đề xuất "Qui Hoạch Điện 8 - PDP8" hướng tới sự phát triển ổn định và bền vững chiến lược đến năm 2050.
+Một trong những nội dung quan trọng của PDP8 có chú trọng đến việc phát triển và nâng cao khả năng ứng dụng các công nghệ tiên tiến và áp dụng dữ liệu lớn cũng như trí thông minh nhân tạo vào công tác xây dựng, điều hành và quản lý các hệ thống điện.
 
@@
 \tiny
- \keyFont{ \section{Từ Khóa:} Dữ Liệu Lớn, Quản Lý Tài Sản, Cơ Sở Hạ Tầng, Độ Tin Cậy, Trí Thông Minh Nhân Tạo} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+ \keyFont{ \section{Từ Khóa:} Dữ Liệu Lớn, Quản Lý Tài Sản, Cơ Sở Hạ Tầng, Độ Tin Cậy, Trí Thông Minh Nhân Tạo, Rủi Do, Tổn Thương, Nguy Cơ} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
 \end{abstract}

### 2023-04-21 2023-EVN-Data-Analytics-for-CEO/Vi-Du/TDD/tdd.Rmd
update

---
title: "Technical Due Diligence"
author: "Nam Le"
date: "February 7, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rationale

Technical Due Diligence (TDD) is an engineering management process for evaluation and assessment on operation and condition of assets of engineering systems (e.g. buildings, industrial plants, factories). TDD can be considered as a simple application of the Asset Management Framework.

TDD is often executed by an engineering firm for Clients who aim to

* Buy/Acquire the property
* Execute an intervention strategy/program for the targeted engineering systems at a set of points in time

TDD is usually done in a quick fashion by a team of a few senior engineers and specialists who understand the targeted engineering systems. The team will normally visit the site/office to perform visual inspection and desktop study on historical records (e.g. as-building drawings, O&M manual, corrective and intervention intervention logbooks). In some case, the work might also involve physical testings for a certain level of audit.

## Methodology
### Condition State

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read_excel("tdd.xlsx",sheet="CS")
library(knitr)
kable(data, caption = "Condition State definition")
```

Aside from CS, it might be of usefulness to use a range of RISK

1 - <span style="color: green;">Green</span>

2 - <span style="color: #FF5733;">Amber</span>

3 - <span style="color: red;">Red</span>

These codes are sometimes attrative to managers but not for engineers :) as it is too much of high level of abstraction and not really accurate and very difficult to quantify.

### Intervention Strategy

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read_excel("tdd.xlsx",sheet="IS")
library(knitr)
kable(data, caption = "Intervention Strategy")
```

### The Model

The model is a simple one below

$$\Omega = \sum_{t=0}^{T} \sum_{n=1}^{N} \pi_{i,j,t}^{n} C_{i,j,t}^{n}  Q_{t}^{n}   \delta_{i,j,t}^{n}  e^{-\rho  t} \label{omega}$$

the objective is to minimize this objective function by allocating possible intervention types along time axis.

In the equation, $C_{i,j,t}^{n}$ is the unit cost to execute an intervention $j$ on asset $n$ when the asset is at CS $i$ at time $t$. $\pi$, $\rho$ are probability and discount factor, respectively. $Q$ is the quantity.

It is important to note that $\rho$, the discount factor, is often not used in business context, though it is taught in the school :). I work with many investment capital firms and they simply want to set $\rho = 0$ so they are at a conservative position to make/negociate the deal.

## Example
This section describes a fictive example of TDD with inputs and outputs all generated using R code.

### Asset Hierarachy

Assets should belong to Area, Zone, Disciplines, and Facilities. In addition, Assets should also belong to Tenant, User, Owner, and Operator. They are stakeholders involved in owning, managing, and using the assets.

An example of the hierarachy is

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read_excel("tdd.xlsx",sheet="Levels")
library(knitr)
options(knitr.kable.NA = '')
kable(data, caption = "Asset Hierarachy")
```

### The data
Now time to generate random data for fun. It is easier to use excel worksheet to generate random data than using R. Another reason to use excel in this case is that it is most convenient for members of a team to share using Cloud services such as SharePoint or Dropbox. They can basically work simultanenously on one file. Another benefits of using excel file for recording this type of data is to present to Clients and get them agree and appreciate your work. Moreover, it is also easy to plug the excel file to any Business Analytics software such as Power BI and Tableau.

```{r echo=TRUE, message=FALSE}
# this is the code saved in tdd.R file
library(readxl) 
library(dplyr)
library(DT)
library(ggplot2)
library(gridExtra)
library(grid)
library(png)
library(downloader)
library(grDevices)
library(cowplot)
library(reshape)
library(lubridate)
library(tidyverse)
library(janitor)
Capex=read_excel("tdd.xlsx",sheet="Capex")
epsilon=1000
#--------------------------------------------------------
#--------------------------------------------------------
#--------------------------------------------------------

cs1 = aggregate(list(CS = Capex$States), list(Disciplines = factor(Capex$Disciplines)), mean,na.rm = TRUE) %>%
 drop_na(CS)

plotcs1=ggplot(cs1, aes(x=reorder(Disciplines,-CS),y = CS)) +
  ylim(0, 5.2)+
  geom_bar(stat = "identity",fill = "#FF6666")+
  labs(title = "", x = "Disciplines",y = "States")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  geom_text(aes(Disciplines, CS+0.1, label = format(CS,digits=3), fill = NULL), data = cs1,cex=3)

plotcs1
#ggsave("picture/plotcs1.png",width = 8, height = 6)

risk1 = aggregate(list(Risk = Capex$Risk), list(Disciplines = factor(Capex$Disciplines)), mean,na.rm = TRUE) %>%
  drop_na(Risk)
plotrisk1=ggplot(risk1, aes(x=reorder(Disciplines,-Risk),y = Risk)) +
  geom_bar(stat = "identity",fill = "#0000FF")+
  ylim(0, 3.2)+
  labs(title = "", x = "Disciplines",y = "Risk")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  geom_text(aes(Disciplines, Risk+0.1, label = format(Risk,digits=3), fill = NULL), data = risk1,cex=3)
plotrisk1
#ggsave("picture/plotrisk1.png",width = 8, height = 6) This is optional

#--------------------------------------------------------
#--------------------------------------------------------
#--------------------------------------------------------

#Estimate the total values of CAPEX for each year with percentage of distribution

#Transform table by year

df<-Capex%>%
  select(InterYear,Disciplines,NPVCapex)%>%
  group_by(Disciplines)
df1<-melt(df,id=c("InterYear","Disciplines"))
df2<-cast(df1,Disciplines~InterYear,sum)
df3<-df2%>%
  mutate(immediate=df2[,2],shortterm=rowSums(df2[,c(3:4)]),mediumterm=rowSums(df2[,c(5:7)]),longterm=rowSums(df2[,c(8:12)]),CAPEX=rowSums(df2[,c(2:12)]))%>%
  mutate(freq = 100*CAPEX / sum(CAPEX))%>%
  arrange(desc(CAPEX))%>%
   adorn_totals()

x01 <- Capex %>%
  group_by(InterYear) %>%
  summarize(total = sum(NPVCapex/epsilon)) %>%
  arrange(desc(total))
x01=mutate(x01,weight_pct=100*total/sum(total))

#plot the graph for yearly CApex distribution per level 4
plot01 <- ggplot(Capex)+
  geom_bar(aes(x = InterYear, y = NPVCapex/epsilon,fill=Disciplines),
           stat='identity')+
  labs(title = "", x = "Year",y = "USD (1000)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+ 
  geom_text(aes(InterYear, total+3, label = round(total,2), fill = NULL), data = x01,cex=3,angle=90)
plot01
#ggsave("picture/plot01.png",width = 10, height = 6)
#Estimate the total values of CAPEX for each generic with percentage of distribution
x11 <- Capex %>%
  group_by(Zone) %>%
  summarize(total = sum(NPVCapex/epsilon)) %>%
  arrange(desc(total))
x11=mutate(x11,weight_pct=100*total/sum(total))

#plot the graph for yearly CApex distribution per level 1
plot11<- ggplot(Capex)+
  geom_bar(aes(x = reorder(Zone, -NPVCapex/epsilon, sum), y = NPVCapex/epsilon,fill=Disciplines),
           stat='identity')+
  labs(title = "", x = "Year",y = "USD (1000)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  geom_text(aes(Zone, total+3, label = round(total,2), fill = NULL), data = x11,cex=3,angle=0)

plot11
#ggsave("picture/plot11.png",width = 10, height = 6)

#Estimate the total values of CAPEX for each generic with percentage of distribution
x21 <- Capex %>%
  group_by(Facilities) %>%
  summarize(total = sum(NPVCapex/epsilon)) %>%
  arrange(desc(total))
x21=mutate(x21,weight_pct=100*total/sum(total))

#plot the graph for yearly CApex distribution per level 1
plot21<- ggplot(Capex)+
  geom_bar(aes(x = reorder(Facilities, -NPVCapex/epsilon, sum), y = NPVCapex/epsilon,fill=Disciplines),
           stat='identity')+
  labs(title = "", x = "Year",y = "USD (1000)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  geom_text(aes(Facilities, total+3, label = round(total,2), fill = NULL), data = x21,cex=3,angle=90)

plot21
#ggsave("picture/plot21.png",width = 10, height = 6)

#Estimate the total values of CAPEX for each generic with percentage of distribution
x31 <- Capex %>%
  group_by(Facilities) %>%
  summarize(total = sum(NPVCapex/epsilon)) %>%
  arrange(desc(total))
x31=mutate(x31,weight_pct=100*total/sum(total))

#plot the graph for yearly CApex distribution per level 1
plot31<- ggplot(Capex)+
  geom_bar(aes(x = reorder(Facilities, -NPVCapex/epsilon, sum), y = NPVCapex/epsilon,fill=Tenant),
           stat='identity')+
  labs(title = "", x = "Year",y = "USD (1000)")+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  geom_text(aes(Facilities, total+3, label = round(total,2), fill = NULL), data = x31,cex=3,angle=90)

plot31
#ggsave("picture/plot31.png",width = 10, height = 6)

```

```{r echo=TRUE, message=FALSE}
source("tdd.R")
library(knitr)
library(kableExtra)
datatable(Capex,filter = 'top') %>%
  formatRound(columns=c('Quantity','UnitCost','NPVCapex'),digits=0)%>%
  formatStyle('NPVCapex',  color = 'red', backgroundColor = 'orange', fontWeight = 'bold')
cat("Total Value =",format(sum(Capex$NPVCapex),digits=5, big.mark=",", small.mark = ".",small.interval=3),"$")
```

### Results

We can summary the results by use of pivot tables and graphs using ggplot2 packages.

#### Condition States and Risks

```{r dodge-st, fig.show = "hold", out.width = "50%",echo=TRUE, message=FALSE}
source("tdd.R")
plotcs1
plotrisk1
```

#### Cost Distribution

```{r echo=TRUE, message=FALSE}
source("tdd.R")
library(knitr)
library(kableExtra)
datatable(df3,filter = 'top',options = list(pageLength = 25)) %>%
  formatRound(columns=c(2:17),digits=0)%>%
  formatRound(columns=c(18),digits=2)%>%
  formatStyle('CAPEX',  color = 'red', backgroundColor = 'orange', fontWeight = 'bold')%>%      formatStyle('Disciplines', target = 'row',  backgroundColor = styleEqual('Total', 'yellow'))
```

```{r fig.width=14, fig.height=8, echo=TRUE, message=FALSE}
source("tdd.R")
plot01
```

```{r fig.width=14, fig.height=8, echo=TRUE, message=FALSE}
source("tdd.R")
plot11
```

```{r fig.width=14, fig.height=8, echo=TRUE, message=FALSE}
source("tdd.R")
plot21
```

```{r fig.width=14, fig.height=8, echo=TRUE, message=FALSE}
source("tdd.R")
plot31
```

## Conclusion
This process can be automated for any TDD job.

# Thank you for reading the article ^_+

### 2023-04-21 2023-EVN-Data-Analytics-for-CEO/Vi-Du/EVM/main.R

@@ -1,0 +1,87 @@
+#Coded by Nam Le (namlt@prontonmail.ch)
+library(scales)
+library(reshape2)
+library(ggplot2)
+library(plotrix)
+
+
+
+#----------DATA-----------------
+data <- read.csv(file="Scurve.csv", header=TRUE, sep=",")
+k=data.frame(data)
+
+# -----------------------------This needs to be modified whenever you plot
+
+plot.new()
+par(mar=c(4, 4, 4, 4), bg="ivory")
+week= length(data[!is.na(data[,3]),3]) #the lastest week that EV value is available. Or you can simply put the number of week in instead of using length(data[!is.na(data[,3]),3]). Here 3 is the column.
+scalefactor=100
+trucy=100
+ticktime=40   #Set the Project Week at which you want to limit the X-axis
+varrange=c(-20,50)
+
+#Posion of texbox (PV, EV, and VAR in the graph)
+htextbox=c(week-5,week)
+vtextbox=20
+
+# --------------------------------
+
+
+# ----------------------------
+#Plot the baseline
+
+plot(data$BS01[1:ticktime],pch=4,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="blue",type="o",lwd=1,lty=1, main =paste("S-curve of week", week))
+axis(2, ylim=c(0,trucy),col="darkblue",las=1)
+mtext(expression(paste("Cummulative percentage (%)")),side=2,line=2.2, adj = 0.5 )
+axis(1,pretty(range(data$Week),30))
+mtext(expression(paste("Project Week")),side=1,col="black",line=2.2)  
+box()
+
+#plot the EV curve
+par(new=TRUE)
+plot(data$EV[1:ticktime],pch=18,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="red",type="o",lwd=1,lty=2)
+
+#adding stick at the actual week
+abline(v=week, col="darkviolet",lty=3)
+
+par(new=TRUE)
+#library(gplots)
+library(astro) #this package is for astronomy but it has some fantastic function for plotting. Here I use the textbox function
[…]

### 2023-04-21 2023-EVN-Data-Analytics-for-CEO/Vi-Du/TDD/tdd.R

@@ -1,0 +1,142 @@
+library(readxl) 
+library(dplyr)
+library(DT)
+library(ggplot2)
+library(gridExtra)
+library(grid)
+library(png)
+library(downloader)
+library(grDevices)
+library(cowplot)
+library(reshape)
+library(lubridate)
+library(tidyverse)
+library(janitor)
+Capex=read_excel("tdd.xlsx",sheet="Capex")
+epsilon=1000
+#--------------------------------------------------------
+#--------------------------------------------------------
+#--------------------------------------------------------
+
+cs1 = aggregate(list(CS = Capex$States), list(Disciplines = factor(Capex$Disciplines)), mean,na.rm = TRUE) %>%
+ drop_na(CS) 
+
+plotcs1=ggplot(cs1, aes(x=reorder(Disciplines,-CS),y = CS)) +
+  ylim(0, 5.2)+
+  geom_bar(stat = "identity",fill = "#FF6666")+
+  labs(title = "", x = "Disciplines",y = "States")+
+  theme(axis.text.x=element_text(angle=45, hjust=1))+
+  geom_text(aes(Disciplines, CS+0.1, label = format(CS,digits=3), fill = NULL), data = cs1,cex=3)
+
+
+plotcs1
+#ggsave("picture/plotcs1.png",width = 8, height = 6)
+
+risk1 = aggregate(list(Risk = Capex$Risk), list(Disciplines = factor(Capex$Disciplines)), mean,na.rm = TRUE) %>%
+  drop_na(Risk)
+plotrisk1=ggplot(risk1, aes(x=reorder(Disciplines,-Risk),y = Risk)) +
+  geom_bar(stat = "identity",fill = "#0000FF")+
+  ylim(0, 3.2)+
+  labs(title = "", x = "Disciplines",y = "Risk")+
+  theme(axis.text.x=element_text(angle=45, hjust=1))+
+  geom_text(aes(Disciplines, Risk+0.1, label = format(Risk,digits=3), fill = NULL), data = risk1,cex=3)
+plotrisk1
+#ggsave("picture/plotrisk1.png",width = 8, height = 6) This is optional
+
+#--------------------------------------------------------
+#--------------------------------------------------------
+#--------------------------------------------------------
+
+
+#Estimate the total values of CAPEX for each year with percentage of distribution
+
[…]

### 2023-04-21 2023-EVN-Data-Analytics-for-CEO/Vi-Du/aWeiBullLcc/app.R

@@ -1,0 +1,2 @@
+# Run the app ----
+shinyApp(ui = ui, server = server)

### 2023-04-21 2023-EVN-Data-Analytics-for-CEO/Vi-Du/aWeiBullLcc/server.R

@@ -1,0 +1,158 @@
+# Define server logic ----
+server <- function(input, output) {
+  
+  #define the object
+  
+#  d<-reactive({
+  
+  #dist <- reactive(sample(input$range[1]:input$range[2], input$obs, replace=FALSE))
+  
+  d<-reactive(    
+    #if (input$generateBt > 0 ) 
+    #isolate(
+      floor(runif(input$obs, min = input$range[1], max = input$range[2]))
+    )
+  #)
+  
+  l=reactive(length(d()))
+  
+  
+  #-------Weibull function
+  median_percentile_ranks<-reactive(getPPP(d()))
+  #print(median_percentile_ranks)#[,2]
+  MLEfit<-reactive(mlefit(mleframe(d())))
+  MLE_Unbiased<-reactive(c(MLEfit()[1],MLEfit()[2]*hrbu(l())))
+  
+  eta<-reactive(MLE_Unbiased()[[1]])
+  beta<-reactive(MLE_Unbiased()[[2]])
+  
+  da1 <- reactive(data.frame(
+    serial=as.character(c(1:l())),
+    time=c(d()),
+    event=rep(1,l())))
+  
+  da2 <- reactive(wblr(da1(), col="red"))
+
+  da3 <- reactive(wblr.fit(da2(), col="darkgreen"))
+  
+  da4 <- reactive(wblr.conf(da3(), col="blue"))
+  
+  #----------------
+  
+  
+
+  
+  
+  
+  output$result <- renderPrint({ 
+  d()
+  })
+  
+  
+  
+  # Generate a summary of the dataset ----
+  output$summary <- renderPrint({
+    dataset <- d()
+    summary(dataset)
+  })
+  
+### display histograph
+  
+  output$distPlot <- renderPlot({
+    dataset <- d()
+    barplot(dataset,col="turquoise3",ylim=c(0,10000))
+  })
+ 
+ 
+ 
+### Display the estimation for alpha and eta
+
+  output$weibullplot <- renderPlot({
+    plot(da4())
+  })
+  
+  
+## Print value of Alpha and eta
+  
+  output$weibullvalue <- renderPrint({ 
+    c(da4()$fit[[1]]$beta,da4()$fit[[1]]$eta)
+  })
+  
+## Reliability curve
+  
+  output$weibullreli <- renderPlot({
[…]

### 2023-04-21 2023-EVN-Data-Analytics-for-CEO/Vi-Du/aWeiBullLcc/ui.R

@@ -1,0 +1,111 @@
+library(shiny)
+library(ggplot2)
+library(WeibullR)
+
+survi=function(t,eta,beta){exp(-(1/eta*t)^beta)} #this is survival function
+
+breakpoint=10
+
+
+# Define UI ----
+ui <- fluidPage(
+  titlePanel("Reliability and Optimal Preventive Intervention - A Weibull case"),
+  
+  helpText("Author: ", a("Nam Le (PhD)",  href="https://namkyodai.github.io"), "- Email address: ", a("namlt@protonmail.com",  href="namlt@protonmail.com")),
+  
+  helpText("This app can be extended for reliability estimation and determination of optimal intervention strategy in various real life cases (e.g. determination of preventive maintenance program of water and wastewater ultilities, facilities of power stations and industrial plants). "),
+  
+  
+  helpText("References:"),  
+  
+  helpText(a("Naoto Kaio, Shunji Osaki, 'Extended block replacement models', AFCET 1984",  href="https://doi.org/10.1051/ro/1984180100591")),  
+  
+ # helpText(a("Nam Le, 'Stochastic Optimization Methods for Infrastructure Management with Incomplete Monitoring', PhD Dissertation, Kyoto University, (2009)")),  
+  
+# helpText("Nam et al. Reports of the Plant Audit for various Pump Stations and Reservoirs - Maynilad Water Services Inc. (2019) "), 
+ 
+  sidebarLayout(
+    sidebarPanel(
+    
+      helpText("Inputs to estimate Reliability"),
+      
+      # Input: Specify the number of observations to view ----
+      numericInput("obs", "Number of observations to view:", 10, width="40%"),
+    
+      
+      
+      sliderInput("range", 
+                  label = "Range of duration between failure",
+                  min = 0, max = 10000, value = c(3000,6000), step=200),#,
+  
+    #  actionButton("generateBt", "Generate Numbers")
+      
+  #  helpText("Weibull distribution"),
+    
+ #   withMathJax(),
+
+    helpText("Inputs for Determination of Optimal Time Windows of Preventive Intervention."),
+    
+    sliderInput("cipiratio", "Ratio CI/PI:",  
+                min = 1, max = 20, value = 5, step=0.5),
+ 
+ sliderInput("interest", "Interest (Discount for NPV):",  
+             min = 0, max = 20, value = 8.5, step=0.5)
+        ),
+  
+
+
+
+    
+    
+    
+    
+    
+    
+#This is the main panel
+    mainPanel(
+      
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Bao-Hiem/R/tong-hop-du-lieu-file-goc.r
update

@@ -27,6 +27,16 @@ rm004=read_excel("../raw/4-RM-Phuong.xlsx",sheet="Sale",skip = 2)
 #nối các dữ liệu có cùng định dạng với nhau thành 1 dữ liệu tổng hợp, đồng thời biến dữ liệu đó thành dataframe để có thể sử dụng dplyr và ggplot2.
 df <-data.frame(rbind(rm001,rm002,rm003,rm004))
 
+
+
+
+
+
+
+
+
+
+
 #Kiểm tra các trường dữ liệu để xem có phải sửa gì không.
 glimpse(df)
 
@@ -47,8 +57,6 @@ df3<-df2 %>%
 
 df4<-df2 %>%
   bind_rows(summarise_all(., ~if(is.numeric(.)) sum(.) else "Tổng Cột"))
-
-
 
 
 p1<-df%>%

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Covid-19/coronavirus.R

@@ -1,0 +1,100 @@
+#Thiết lập đường dẫn vào file
+library(rstudioapi) 
+setwd(dirname(getActiveDocumentContext()$path))  
+
+
+library(dplyr)
+library(tidyr)
+library(plotly)
+library(coronavirus)
+
+
+#data source
+# -----https://covid19.who.int/info?openIndex=2
+#----- https://github.com/RamiKrispin/coronavirus/tree/master/csv
+
+
+#df <- coronavirus
+df <- read.csv("coronavirus.csv", header = TRUE, sep = ",")
+
+data("coronavirus")
+
+total_cases <- df %>% 
+  group_by(type) %>%
+  summarise(cases = sum(cases)) %>%
+  mutate(type = factor(type, levels = c("confirmed", "death", "recovery")))
+
+total_cases
+
+head(df)
+
+
+summary_df <- df %>% 
+  filter(type == "confirmed") %>%
+  group_by(country) %>%
+  summarise(total_cases = sum(cases)) %>%
+  arrange(-total_cases)
+
+
+summary_df %>% head(20) 
+
+df %>% 
+  filter(date == max(date)) %>%
+  select(country, type, cases) %>%
+  group_by(country, type) %>%
+  summarise(total_cases = sum(cases)) %>%
+  pivot_wider(names_from = type,
+              values_from = total_cases) %>%
+  arrange(-confirmed)
+
+
+df %>% 
+  group_by(type, date) %>%
+  summarise(total_cases = sum(cases)) %>%
+  pivot_wider(names_from = type, values_from = total_cases) %>%
+  arrange(date) %>%
+  mutate(active = confirmed - death - recovery) %>%
+  mutate(active_total = cumsum(active),
+         recovered_total = cumsum(recovery),
+         death_total = cumsum(death)) %>%
+  plot_ly(x = ~ date,
+          y = ~ active_total,
+          name = 'Active', 
+          fillcolor = '#1f77b4',
+          type = 'scatter',
+          mode = 'none', 
+          stackgroup = 'one') %>%
+  add_trace(y = ~ death_total, 
+            name = "Death",
+            fillcolor = '#E41317') %>%
+  add_trace(y = ~recovered_total, 
+            name = 'Recovered', 
+            fillcolor = 'forestgreen') %>%
+  layout(title = "Distribution of Covid19 Cases Worldwide",
+         legend = list(x = 0.1, y = 0.9),
+         yaxis = list(title = "Number of Cases"),
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Kiem-Dinh-Cong-Trinh/main.R

@@ -47,7 +47,9 @@ ggplot(missing, aes(x = reorder(Variable, -value),y = value)) +
 ###
 # 
 df1 <- aggregate(df$NPV, by=list(Discipline = df$Discipline), FUN=sum)%>%filter(x>0)
+
 df2 <- aggregate(df$NPV_HighRisk, by=list(Discipline = df$Discipline), FUN=sum)%>%filter(x>0)
+
 df3 <- aggregate(df$NPV_Option, by=list(Discipline = df$Discipline), FUN=sum)%>%filter(x>0)
 write.csv(df1, file = "df1.csv")
 

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/baocao.Rmd
update

---
title: "Report"
author: "Nam Le"
date: "1/22/2022"
#output: word_document
#output: html_document
output: slidy_presentation

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chương 1

Đây là nội dung của chương 1, tóm tắt số liệu tổng hợp của gói dữ liệu có tên là cars.

```{r cars}
summary(cars)
```

## Chương 2

Vẽ biểu đồ liên quan đến biến áp suất

```{r pressure, echo=FALSE}
plot(pressure)
```

Ví dụ

## Chương 3

Tổng hợp dữ liệu của phần đã học và vẽ biểu đồ liên quan.

```{r ECHO = FALSE, message= FALSE}
library(ggplot2)
library(dplyr)

# Load dataset from github
data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/1_OneNum.csv", header=TRUE)
data <- data.frame(data)
summary(data)
```

```{r ECHO = FALSE, message= FALSE}
source("ggplot-density.r")
a1
```

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Data/dieutra-hogiadinh-hcm/nhom1.Rmd
update

---
title: "Phân Tích Dữ Liệu"
subtitle: "Bài Tập Nhóm - Tuần 3"
author: "Nhom 1"
date: "2/11/2022"
output: html_document
---

## Giới Thiệu

Đây là gói dữ liệu khảo sát hơn 20000 hộ dân thành phố HCMC. Dữ liệu được khảo sát trong giai đoạn lập dự án khả thi xây dựng hệ thống Metro cho thành phố hồ chí minh.

Xem phần đầu của dữ liệu

```{r echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
library(readxl)# gói phụ kiện để đọc dữ liệu từ bảng tính excel
library(DT)
df=data.frame(read_excel("dieutra-hogiadinh-hcm.xlsx",sheet="ketquadieutra",skip = 0)) # 
p<-datatable(head(df),rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

Xem phần đuôi của dữ liệu

```{r echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
datatable(tail(df),rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
```

Xem cấu trúc của dữ liệu

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
str(head(df))
```

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
library(dplyr)
glimpse(head(df))
```

File gốc

```{r, echo=TRUE, include = TRUE,results = 'hide',message=FALSE,warning=FALSE}
source("dieutra-hogiadinh-hcm.r")
```

Tả phải chuyển đổi dạng dữ liệu của từng biến

```{r, echo=TRUE, include = FALSE,results = 'hide',message=FALSE,warning=FALSE}
df$Q1a <- as.integer(df$Q1a) #chọn ra một số biến và định nghĩa lại
library(purrr) #
df <- df %>%
  purrr::modify_at(c(2:10, 13:14, 16:21,25:30,44,46,50:51,54:55,58:59,62:63,66:67,70:71,74:75,78,79,82:83,86:87,89:90,92:229), factor)
df$Q4<- as.character(df$Q4)
df$Q6b <- as.character(df$Q6b)
#df$Q9 <- as.integer(df$Q9) #chọn ra một số biến và định nghĩa lại
df$Q11<- as.character(df$Q11)
df$Q18a8<- as.numeric(df$Q18a8)
df$Q18b8<- as.numeric(df$Q18b8)
df$Q18a9<- as.numeric(df$Q18a9)
df$Q18b9<- as.numeric(df$Q18b9)
df$Q18a10<- as.numeric(df$Q18a10)
df$Q18b10<- as.numeric(df$Q18b10)
df$Q44b <- as.character(df$Q44b)
```

Xem lại dữ liệu xem có sự thay đổi gì không?

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
glimpse(head(df))
```

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
df01 <- df%>%
  select(Q3,Q4,Q7,Q8a,Q8b,Q8c,Q8d,Q8e,Q9,Q10,Q11)
head(df01)
summary(df01)
```

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
table(df01$Q8a) # Cách tính thứ 1 sử dụng trực tiếp hàm table trong R
df01 %>%
count("Q8a")   # cách tính thứ 2 sử dụng dplyr với count
count(df01, "Q8a")
```

Vẽ đồ thị thanh

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
ggplot(df01 %>%
         count("Q8a"), aes(x=Q8a, y=freq))+
  geom_bar(stat="identity") 
```

Trang điểm thêm cho đồ thị

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
df01 %>%
  count("Q8a") %>%
  ggplot(aes(x=Q8a, y=freq))+ #khai bao trục hoành x và trục tung
  geom_bar(stat="identity", fill="red", alpha =0.5) + # khai báo biểu đồ dạng barchart, loại biểu đồ frequenty, tô đỏ và làm mờ vùng tô đỏ
  labs(title ="Trình trạng nhà ở", # khai báo tên của biểu đồ
       subtitle = "Chủ sở hữu", # khai báo tên phụ của biểu đồ
       x = "Loại chủ sở hữu", # khai báo tên trục hoành
       y = "Số lượng") # khai báo tên trục tung
```

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
df01 %>%
  count("Q8a") %>%
  ggplot(aes(x=as.factor(Q8a), y=freq))+ # cách vẽ thứ 2 trong đó biến các giá trị 1, 2, 3, 4 thành dạng factor (category) thay vì là số, vì bản chất giá trị này không phải là số.
  geom_bar(stat="identity", fill="red", alpha =0.5) +
  labs(title ="Trình trạng nhà ở",
       subtitle = "Chủ sở hữu",
       x = "Loại chủ sở hữu",
       y = "Số lượng")+
  scale_x_discrete(labels = xticks) + # điền tên các giá trị tương ứng ở trục hoành
  geom_text(aes(label=freq), position=position_dodge(width=0.9), vjust=-0.25, color ="blue") # thêm trên đầu mỗi thanh các giá trị tương ứng với nhóm

```

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
tyle <- 1000000
df01%>%
  filter(!is.na(Q9), Q9 != 0, Q9 <10000000)%>%
  select(Q9)%>%
  ggplot(aes(x=Q9/tyle))+
  geom_density()+
  geom_histogram(aes(y = ..density..),colour = 1, fill = "white", alpha=0.2)

```

Vẽ đồ thị động

```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
tyle <- 1000000
p<-df01%>%
  filter(!is.na(Q9), Q9 != 0, Q9 <10000000)%>%
  select(Q9)%>%
  ggplot(aes(x=Q9/tyle))+
  geom_density()+
  geom_histogram(aes(y = ..density..),colour = 1, fill = "white", alpha=0.2)
library(plotly)

ggplotly(p)
```

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/01-gioi-thieu.Rmd
update

---
title     : "Phân Tích và Tính Toán  với Dữ Liệu Đô Thị <br> Urban Computation"
subtitle  : "Giới Thiệu Chung"
author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
date      : "4-tháng 8-2023"
output    :
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Giới Thiệu Bản Thân

<div class='left' style='float:left;width:50%'>
- Tiểu Sử
  - Ngành nghiên cứu chính về phân tích rủi do bằng phương pháp định lượng và quản lý cơ sở hạ tầng xây dựng và các hệ thống kĩ thuật. 
  - 4 năm sống ở Nhật (làm Tiến Sĩ tại Đại Học Kyoto và 1  năm làm nghiên cứu sau TS tại ĐH Osaka, Nhật Bản)
  - <2 năm ở Thái Lan (học Thạc Sĩ tại AIT từ năm 2002-2004)
  - 6+ năm sống ở Thụy Sĩ (2010-2016, giảng dạy và nghiên cứu tại học viện kĩ thuật liên bang Thụy Sĩ, ETH Zurich)
  - 6 năm làm tư vấn tại Manila, Philippines (2016-2021) cho 2 công ty (GHD và Arcadis)
  - Từ cuối năm 2021, sống ở thủ đô New Delhi, Ấn Độ và thường xuyên về Việt Nam tham gia giảng dạy.

- Các môn thể thao yêu thích
  - Chạy bộ đường dài `r emo::ji("running")``r emo::ji("running")``r emo::ji("running_man")`
  - Lặn biển `r emo::ji("swimming_man")`
  - và một số môn khác ^_^ `r emo::ji("smile")``r emo::ji("smiling_imp")`

</div>
<div class='right' style='float:right;width:50%'>

```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
source("../code/bio.R")
p
```
</div>

##  Giới Thiệu Bản Thân

<div class='left' style='float:left;width:50%'>

```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
source("../code/bio.R")
q
```

</div>

<div class='right' style='float:right;width:50%'>

```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
source("../code/bio.R")
ggplotly(w)
```
</div>

## Thành Viên của Lớp

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center',collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
library(DT)
library(tidyverse)
library(readxl)
library(lubridate)
df=read_excel("../data/hocvien.xlsx",sheet="Sheet1",skip = 0)
df$`Ngày Sinh` <- dmy(df$`Ngày Sinh`)
p <- datatable(df,rownames = FALSE, filter = 'top',
          options = list(pageLength = 10,autoWidth = TRUE)
          )
p
```

##  Thành Viên của Lớp

<div class='left' style='float:left;width:50%'>

```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
source("../code/hocvien.R")
boxplot(df$namsinh,
    #    main = "Mean ozone in parts per billion at Roosevelt Island",
        xlab = "Năm Sinh",
        ylab = "",
        col = "orange",
        border = "brown",
        horizontal = TRUE,
        notch = TRUE
)
```

</div>

<div class='right' style='float:right;width:50%'>

```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
source("../code/hocvien.R")
ggplotly(p3)
```
</div>

## Chương Trình Học và Thời Khóa Biểu

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center',collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
source("../code/schedule.R")
p
```

## Mong Muốn và Yêu Cầu

```{r, echo=FALSE, out.width="60%", fig.cap="", fig.align = 'center'}
library(ggplot2)

# Create test data.
data <- data.frame(
  category=c("Học Trên Lớp", "Thảo Luận và Trình Bày", "Bài Thi"),
 count=c(40, 30, 30)
)

# Compute percentages
data$fraction <- data$count / sum(data$count)

# Compute the cumulative percentages (top of each rectangle)
data$ymax <- cumsum(data$fraction)

# Compute the bottom of each rectangle
data$ymin <- c(0, head(data$ymax, n=-1))

# Compute label position
data$labelPosition <- (data$ymax + data$ymin) / 2

# Compute a good label
data$label <- paste0(data$category, "\n value: ", data$count)

# Make the plot
ggplot(data, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=category)) +
  geom_rect() +
  geom_label( x=3.5, aes(y=labelPosition, label=label), size=6) +
  scale_fill_brewer(palette=4) +
  coord_polar(theta="y") +
  xlim(c(2, 4)) +
  theme_void() +
  theme(legend.position = "none")
```

## Khoa Học về Dữ Liệu (1) - Sơ đồ Venn

```{r, echo=FALSE, out.width="40%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/khoahocdulieu01.png")
```

## Khoa Học về Dữ Liệu (2) - Khoa Học Dữ Liệu vs Phân Tích Dữ Liệu

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/khoahocdulieu02.png")
```

## Khoa Học về Dữ Liệu (3) - Khoa Học Dữ Liệu vs Phân Tích Dữ Liệu

<div class='left' style='float:left;width:30%'>
- Một vài phương pháp tính toán xác xuất thống kê cơ bản
- Dữ liệu và Dữ Liệu Sạch
- Cơ bản về quản lý dữ liệu
- Giới thiệu sơ qua về phần mềm R
- Học cơ bản về nền tảng phân tích KNIME và PowerBI
- Các công cụ phân tích khác
- Cơ bản về trực quan hóa dữ liệu
- Các kết luận rút ra từ phân tích dữ liệu
</div>

<div class='left' style='float:left;width:60%'>
```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/Data-Engineer-vs-Data-Scientist-vs-Data-Analyst.png")
```
</div>

## Các Bước Trong Phân Tích Dữ Liệu

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-science-explore.png")
```

## Vậy Học Thì Có Gì Là Bổ Ích?

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/cartoon01.png")
```

</div>

<div class='right' style='float:right;width:48%'>

**Bạn không cần phải trở thành một nhà nghiên cứu dữ liệu hay phân tích dữ liệu toàn thời gian, nhưng bạn có thể hiểu và làm công việc liên quan đến phân tích dữ liệu hiệu quả.**

- Thu thập dữ liệu và ghi dữ liệu một cách bài bản và hợp lý
- Sự khác biệt cơ bạn của các hệ thống quản lý dữ liệu
- Tự động hóa và sao lại qui trình công việc (workflow)
- Nâng cao năng xuất và tính hiệu quả trong công việc
- Chuyển đổi (số)

</div>

## Tài Liệu Tham Khảo

<div class='left' style='float:left;width:48%'>
- Sách và Hệ Thống
  - [R for Data Science](https://r4ds.had.co.nz/index.html)
  - [Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html)
  - [Github](https://github.com/)
  - [R Statistical Computing](https://www.r-project.org/) and  [Rstudio](https://www.rstudio.com/)
  - [R Graph Gallery](https://www.r-graph-gallery.com/)
  - [Plotly R Open Source Graphing Library](https://plotly.com/r/)
  - [Bookdown](https://bookdown.org/)s
  <!-- - [Stat545](https://stat545.com/index.html) -->
- Các gói thư Viện R
  - [Tidyverse](https://www.tidyverse.org/), [Dplyr](https://dplyr.tidyverse.org/), [Ggplot2](https://ggplot2.tidyverse.org/), [DT](https://rstudio.github.io/DT/),  [Rattle](https://rattle.togaware.com/)

- Trang Web của KNIME
  - https://www.knime.com/knimepress

</div>

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/01_tidyverse_data_science.png")
```

</div>

## Giới Thiệu Sơ Qua về Sự Phát Triển của R

<div class='left' style='float:left;width:48%'>
<div class='left' style='float:left;width:48%'>
[Robert Gentleman](https://computationalbiomed.hms.harvard.edu/people/robert-gentleman)
```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
url <- "https://computationalbiomed.hms.harvard.edu/wp-content/uploads/2022/02/Robert-Gentleman-200x240-1.png"
knitr::include_graphics(url)
#knitr::include_graphics("../Visual/data-atomic.png")
```
Trường Y Khoa - Đại Học Harvard

</div>

<div class='left' style='float:left;width:48%'>
[Ross Ihaka](https://www.stat.auckland.ac.nz/~ihaka/)
```{r, echo=FALSE, out.width="65%", fig.cap="", fig.align = 'center'}
url <- "https://www.stat.auckland.ac.nz/~ihaka/images/ross-ihaka.jpg"
knitr::include_graphics(url)
#knitr::include_graphics("../Visual/data-atomic.png")
```
Giáo Sư về Xác Xuất và Thống Kê - Đại Học Auckland
</div>

- 1991 Triển Khai Ngôn ngữ S
- 1993 Giới Thiệu R
- 1995 Xuất bản miễn phí R là phần mềm miễn phí và mã nguồn mở

</div>

<div class='left' style='float:left;width:48%'>

Ngôn ngữ S được phát triển tại phòng Lab BELL Lab từ năm 1976

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
url <- "https://www.telecomreview.com/images/stories/2020/03/Nokia_Bell_Labs_achieves_world_record_in_fiber_optics-article.jpg"
knitr::include_graphics(url)
#knitr::include_graphics("../Visual/data-atomic.png")
```

- 9 giải Nobels 
- Thiết kế ra chương trình tối ưu AMPL Optimizer
- Thiết kế ra chương trình C

</div>

## Giới Thiệu qua về R (1)

<div class='left' style='float:left;width:30%'>

[An Introduction to R](https://intro2r.com/) by Alex Douglas, Deon Roos, Francesca Mancini, Ana Couto & David Lusseau

```{r, echo=FALSE, out.width="65%", fig.cap="", fig.align = 'center'}
url <- "https://intro2r.com/images/cover_v3.png"
knitr::include_graphics(url)
#knitr
```
</div>

<div class='left' style='float:left;width:65%'>
[Phân Tích Dữ Liệu với R](https://nguyenvantuan.info/publication/#jp-carousel-1489) - Giáo Sư [Nguyễn Văn Tuấn](https://nguyenvantuan.info/), University of Technology Sydney, Australia.

<div class='left' style='float:left;width:48%'>
```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center'}
url <- "https://nguyenvantuan830970966.files.wordpress.com/2020/10/50879-86994c_3b7445736eab4e9d8863ca616c088736mv2.jpg?w=400&h=&zoom=2"
knitr::include_graphics(url)
#knitr
```
</div>

<div class='left' style='float:left;width:48%'>

<iframe width="560" height="315" src="https://www.youtube.com/embed/nUrR0t45gx4" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

</div>

</div>

## Giới Thiệu qua về R  (2)

<div class='left' style='float:left;width:25%'>
- Thư Mực Làm việc - Working directory
- Tên File và cấu trúc - File name and structure
- Tài liệu dự án 
- Các loại dữ liệu
- Kết cấu của dữ liệu 
</div>

<div class='left' style='float:left;width:75%'>

```{r, echo=FALSE, out.width="95%", fig.cap="", fig.align = 'center', warning=FALSE}
library(knitr)
knitr::include_graphics("assets/img/rstudio.png")
```

</div>

## Kết Cấu Dữ Liệu (1)

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="45%", fig.cap="", fig.align = 'center'}
url <- "https://intro2r.com/images/scal_vec.png"
knitr::include_graphics(url)
#knitr
```

```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center'}
url <- "https://intro2r.com/images/mat_array.png"
knitr::include_graphics(url)
#knitr
```

</div>

<div class='left' style='float:left;width:48%'>
Matrix
```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
my_mat <- matrix(1:16, nrow = 4, byrow = TRUE)
my_mat
```

Array
```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
my_array <- array(1:16, dim = c(2, 4, 2))
my_array
```
</div>

## Kết Cấu Dữ Liệu (2)

<div class='left' style='float:left;width:45%'>
List
```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
list_1 <- list(c("black", "yellow", "orange"),
               c(TRUE, TRUE, FALSE, TRUE, FALSE, FALSE),
               matrix(1:6, nrow = 3))
list_1
```

</div>

<div class='left' style='float:left;width:45%'>
Data Frame
```{r, echo=TRUE, include = TRUE,message=FALSE,warning=FALSE}
p.height <- c(180, 155, 160, 167, 181)
p.weight <- c(65, 50, 52, 58, 70)
p.names <- c("Joanna", "Charlotte", "Helen", "Karen", "Amy")

dataf <- data.frame(height = p.height, weight = p.weight, names = p.names)
dataf
```

</div>

## Quá Trình Xử Lý Dữ Liệu (Data Wrangling)

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
url <- "https://favtutor.com/resources/images/uploads/DATA_WRANGLING.jpg"
knitr::include_graphics(url)
#knitr
```

## Nhập Dữ Liệu vào Môi Trường R

<div class='left' style='float:left;width:45%'>
Common data files

- text/flat files
- cvs/excel files
- các bảng từ databases

[Nhập Dữ Liệu vào Môi Trường R](http://www.sthda.com/english/wiki/importing-data-into-r)

```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center'}
url <- "http://www.sthda.com/sthda/RDoc/images/importing-data-into-r.png"
knitr::include_graphics(url)
#knitr
```

</div>

<div class='left' style='float:left;width:45%'>
R Packages and function

```r
# Read tab separated values
read.delim(file.choose())
# Read comma (",") separated values
read.csv(file.choose())
# Read semicolon (";") separated values
read.csv2(file.choose())
```

```r
library("readr")
# Read tab separated values
read_tsv(file.choose())
# Read comma (",") separated values
read_csv(file.choose())
# Read semicolon (";") separated values
read_csv2(file.choose())
```

```r
#reading sheet in an excel file
library(readxl)
df=read_excel("../Data/introduction.xlsx",sheet="bio-time",skip = 0)
```

</div>

## Xuất/Ghi Dữ Liệu từ R

<div class='left' style='float:left;width:45%'>
Các files thông thường

- text/flat files
- cvs/excel files
- các bảng từ databases

[Xuất/Ghi Dữ Liệu từ R](http://www.sthda.com/english/wiki/exporting-data-from-r)

```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center'}
url <- "http://www.sthda.com/sthda/RDoc/images/exporting-data-from-r.png"
knitr::include_graphics(url)
#knitr
```

</div>

<div class='left' style='float:left;width:45%'>
R Packages and function

```r
# Loading mtcars data
data("mtcars")
# Write data to txt file: tab separated values
# sep = "\t"
write.table(mtcars, file = "mtcars.txt", sep = "\t",
            row.names = TRUE, col.names = NA)
# Write data to csv files:  
# decimal point = "." and value separators = comma (",")
write.csv(mtcars, file = "mtcars.csv")
# Write data to csv files:
# decimal point = comma (",") and value separators = semicolon (";")
write.csv2(mtcars, file = "mtcars.csv")
```

```r
# Loading mtcars data
data("mtcars")
library("readr")
# Writing mtcars data to a tsv file
write_tsv(mtcars, path = "mtcars.txt")
# Writing mtcars data to a csv file
write_csv(mtcars, path = "mtcars.csv")
```

```r
library("xlsx")
# Write the first data set in a new workbook
write.xlsx(USArrests, file = "myworkbook.xlsx",
      sheetName = "USA-ARRESTS", append = FALSE)
# Add a second data set in a new worksheet
write.xlsx(mtcars, file = "myworkbook.xlsx",
           sheetName="MTCARS", append=TRUE)
```

</div>

## Giới Thiệu qua về KNIME

<div class='left' style='float:left;width:25%'>
KNIME là một phần mềm và thường được các nhà phát triển gọi là hệ thống phân tích KNIME, do công ty KNIME AG có trụ sở ở thành phố Zurich, Thụy Sĩ phát triển.

KNIME có cấu trúc và cách tiến hành phân tích dữ liệu theo dạng Workflow và theo khái niệm Kéo và Thả (Drag and Drop) nên rất trực quan đối với người dùng không có kiến thức về lập trình.

KNIME có mã nguồn mở và miễn phí đối với người dùng cho bản tiêu chuẩn. KNIME đang bắt đầu được các doanh nghiệp lớn quan tâm và sử dụng, bên cạnh các phần mềm như PowerBI và Tableau.

</div>

<div class='left' style='float:left;width:75%'>

```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center', warning=FALSE}
library(knitr)
knitr::include_graphics("assets/img/Knime.jpg")
```

</div>

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/02-vidu.Rmd
update

---
title     : "Phân Tích và Tính Toán  với Dữ Liệu Đô Thị <br> Urban Computation"
subtitle  : "Ví Dụ"
author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
date      : "4-tháng 8-2023"
output    :
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Các Bước Trong Phân Tích Dữ Liệu

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-science-explore.png")
```

## Ví Dụ 1: Biến Động Giá Nhà Theo Thời Gian

> "Chỉ Số Giá các loại nhà được tổng hợp cho từng khu vực và theo từng năm. Giá nhà biến động theo thời gian do ảnh hưởng của các yêu tố như chính sách, di chuyển dân cư, và tình hình phát triển kinh tế xã hội. Việc thiết lập cơ sở dữ liệu để lưu trữ lại chỉ số biến động giá nhà và các biến đi kèm với nó cần được thực hiện có hệ thống. Ở các nước phát triển, hệ thống cơ sở dữ liệu này được chỉa sẻ trực tiếp từ trang web của các tổ chức chính phủ."
>

```{r, echo=FALSE, out.width="40%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/NHPI_plot.png")
```

https://tradingeconomics.com/canada/housing-index

https://www150.statcan.gc.ca/n1/pub/62f0014m/62f0014m2022003-eng.htm

## Ví Dụ 2: Vệ Sinh và An Toàn trong Cộng Đồng Dân Cư (WASH)

> "Các tổ chức tài chính như Ngân Hàng Thế Giới, Ngân Hàng Châu Á, Quĩ IFC, trước khi ra các quyết định liên quan đến đầu tư cơ sở hạ tầng để phục vụ phát triển an sinh xã hội. Họ thường yêu cầu lập các báo cáo nghiên cứu tiền khả thi và khả thi. Các báo cáo này gồm nhiều cấu phần và một trong những cấu phần là liên quan đến thực trạng của môi trường sống. Ví dụ để xây dựng một nhà máy sản xuất nước ở một số xã ở vùng ven đô thị và vùng nông thôn của Việt Nam, IFC và các tổ chức tài chính nước ngoài thường yêu cầu thực hiện WASH, tức là điều tra và đánh giá về vệ sinh và an toàn của các hộ dân sinh sống trong khu vực ảnh hưởng của dự án."
>

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/graph_a8_yes_perc.png")
```

</div>

<div class='right' style='float:right;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/graph_a7q5.png")
```

</div>

## Ví Dụ 3: Đánh Giá và Kiểm Định Công Trình ở Qui Mô Chiến Lược

> "Trước khi ra quyết định đầu tư liên quan đến mua bán và xát nhập công ty hay các công trình và hệ thống kĩ thuật. Chủ đầu tư thường thuê đơn vị tư vấn độc lập tiến hành khảo sát để đánh giá và kiểm định công trình ở mức độ chiến lược, việc đánh giá này thường được thực hiện trong một thời gian từ 2 đến 4 tuần, tùy vào mức độ qui mô của công trình. Để thực hiện được công việc này một cách bài bản, cần phát triển hệ thống ghi chép dữ liệu chặt chẽ ngay từ đầu."
>

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/cooling.png")
```

</div>

<div class='right' style='float:right;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/tdd01.png")
```

</div>

## Ví Dụ 4: Bảo Hiểm Nhân Thọ

> "Khi bạn là một nhà quản lý và có một số đội ở dưới thì lúc đó bạn cần nắm bắt nhanh nhất có thể thông tin và báo cáo của các tổ trưởng các đội, và thậm chí là thông tin và báo cáo của từng người trong các tổ đội đó. Đây là một ví dụ cụ thể về việc tổng hợp thông tin liên quan đến bán bảo hiểm nhân thọ của một Công Ty Bảo Hiểm quốc tế, có trụ sở tại Việt Nam."
>

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/knime-bao-hiem.png")
```

</div>

<div class='right' style='float:right;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/knime-bao-hiem01.png")
```

</div>

## Ví Dụ 5: Quản Lý Dự Án.

> "Khi tham gia vào một dự án, các nhà quản lý thường quan tâm đến tiến độ thực hiện dự án, tiến độ dự án không những về thời gian mà còn về khối lượng, hay đơn giản gọi là **tiến độ Giải Ngân**. Tiến độ này liên quan chặt chẽ tới thời gian và khối lượng công việc hoàn thành. Để nắm bắt được tiến độ này chính xác, bạn cần phải sử dụng các kiến thức liên quan đến EVM (Earned Value Management), mà để làm được việc này theo từng tuần, thậm chí là từng ngày thì bạn lại cần đến số liệu thu thập được từ từng cấu phần của dự án cho đến thời điểm hiện tại."
>

```{r, echo=FALSE, out.width="50%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/evm.png")
```

## Ví Dụ 6: Thiết Kế Biển Bản Cuộc Họp Như Thế Nào Cho Tối Ưu Nhất.

> "Việc viết biên bản cuộc họp là rất mất thời gian và đôi khi là mang nhiều tính rủi do liên quan đến câu chữ. Trong một tổ chức hay 1 dự án, có rất nhiều cuộc họp diễn ra và đôi khi bạn sẽ nhận thấy là các điểm nội dung của rất nhiều cuộc họp là giống nhau, và lặp đi lặp lại. Vậy chẳng nhẽ, mỗi lần như vậy, bạn cứ phải viết đi viết lại??? Bạn có muốn các biển bản cuộc họp được ghi chép một cách hệ thống để tiện cho công tác truy cứu và quản lý không?."
>

```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/mom.png")
```

## Ví Dụ 7: Bản Đồ Hóa Tình Hình Dịch Bệnh Covid-19.

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/covid-19.png")
```

## Các Bước Trong Phân Tích Dữ Liệu

```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-science-explore.png")
```

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/03-cac-nguyen-ly-co-ban.Rmd
update

---
title     : "Các Nguyên Lý Cơ Bản Trong Phân Tích Dữ Liệu"
subtitle  : "Lý Thuyết Về Dữ Liệu"
author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
date      : "4-tháng 8-2023"
output    : 
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH."
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

## Giới Thiệu

> Khoa học dữ liệu là khoa học và việc thiết kế của (1) liên tục tạo ra một câu hỏi để điều tra một giả định với dữ liệu, (2) thiết lập và kết nối câu hỏi đó với việc thu thập các dữ liệu liên quan và ứng dụng các phương pháp phù hợp, và (3) liên kết và ra quyết định dựa vào các kiến thức đã có hoặc kiến thức mới có được từ việc phân tích dữ liệu.

>
> `r tufte::quote_footer('--- Stephanie C. Hicks and Roger D. Peng, 2019')`

> Phân tích dữ liệu là một quá trình điều tra tìm hiểm để lấy ra các kiến thức, thông tin, và các nội dung của hiện thực.

>
> `r tufte::quote_footer('--- Grolemund and Wickham, 2014')`

- [50 năm ngành khoa học dữ liệu](http://courses.csail.mit.edu/18.337/2015/docs/50YearsDataScience.pdf)

- [Khoa học dữ liệu: Một rạp xiếc có 3 vòng hay là một cái lều lớn?](https://arxiv.org/ftp/arxiv/papers/1712/1712.07349.pdf)

- [Các Yếu Tố Cơ Bản và Nguyên Lý của Phân Tích Dữ Liệu](https://arxiv.org/pdf/1903.07639v1.pdf)

- [Khoa Học Dữ Liệu Giờ Đây Đã Khác Biệt](http://veekaybee.github.io/2019/02/13/data-science-is-different/)

## Giới Thiệu - Các Yếu Tố của Phân Tích Dữ Liệu

> Các Yếu Tố của phân tích dữ liệu là các Thành Phần cơ bản được nhà phân tích dùng như: Đoạn mã, ghi chú cho đoạn mã, trực quan hóa dữ liệu, các dòng viết diễn tả, các bản thống kê, các mô hình thống kê hay các thuật toán dùng để tính toán.

>
> `r tufte::quote_footer('--- Breiman, 2001')`

```{r echo=FALSE, message=FALSE,,warning=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read_excel("../data/introduction.xlsx",sheet="elementsofdata")
library(knitr)
library(kableExtra)
data %>%
  kbl() %>%
#  kable_classic(full_width = F, html_font = "Cambria")
  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
#kable(data, booktabs = TRUE)
```

## Các Nguyên Lý của Phân Tích Dữ Liệu - Công Việc Khớp Dữ Liệu

<div class='left' style='float:left;width:48%'>

 - Các dữ liệu có tính kết nối và tính khớp nhau cao sẽ giúp nhà phân tích xem xét các giả định và các vấn đề tốt hơn là việc xem xét các dữ liệu dời dạc và không khớp.

- Công việc khớp dữ liệu (đôi khi được biết đến là việc ghi hay liên kết dữ liệu, định danh...) là một tác vụ liên quan đến xác định, khớp nhau, kết nối các bảng ghi có cùng chung các thuộc tính từ các bảng dữ liệu khác nhau trong một hệ thống cơ sở dữ liệu.

</div>
<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/datamatching.png")
```
</div>

## Các Nguyên Lý của Phân Tích Dữ Liệu - Tính Toàn Diện

<div class='left' style='float:left;width:48%'>

- Một phân tích được cho là toàn diện nếu các câu hỏi cụ thể hay các giả định được đề cập sử dụng nhiều thành phần bổ sung.

- Việc kết hợp nhiều thành phần có thể đưa ra một bức tranh toàn diện liên quan đến chứng cứ trong dữ liệu hơn là việc sử dụng chỉ một thành phần.

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/dataexhausive.png")
```

</div>
<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/dataexhausive02.png")
```

</div>

## Các Nguyên Lý của Phân Tích Dữ Liệu - Tính Hoài Nghi

- Một phân tích có tính hoài nghi nếu nhiều giả định hoặc nhiều giải thích của một hiện tượng được quan sát là được xem xét khi sử dụng cùng nguồn dữ liệu và đưa ra được tính kiên định trong dữ liệu với các giải thích khác. Trái ngược lại, các phân tích mà không tính đến lời giải thích khác thì được coi là không có tính hoài nghi.

```{r, echo=FALSE, out.width="50%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-skeptical01.png")
```

## Các Nguyên Lý của Phân Tích Dữ Liệu - Bậc 2

- Một phân tích được coi có tính chất bậc 2 nếu nó bao gồm các thành phần khác mà không trực tiếp đề cập đến giải định/câu hỏi ban đầu, nhưng lại có thể đưa/chỉ ra được bối cảnh quan trọng của việc phân tích.

```{r, echo=FALSE, out.width="40%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-2ndorder.png")
```

## Các Nguyên Lý của Phân Tích Dữ Liệu - Sự Minh Bạch

<div class='left' style='float:left;width:48%'>

- Các phân tích có tính minh bạch thể hiện một yếu cố hay nhiều yêu tố thống kê lại/ hoặc trực quan hóa các dữ liệu có tính ảnh hưởng đến việc giải thích như thế nào mà các hiện tượng hay quá trình tạo ra dữ liệu kết nối với kết quả đầu ra hay các kết luận.

- Trong khi toàn bộ việc phân tích có thể là phức tạp và liên quan đến nhiều bước tính toán rất dài, các phân tích có tính minh bạch lấy ra một hoặc vài thành tố có thể giải thích được tính biến đổi nhiều nhất hay ảnh hưởng nhất tới việc hiểu biết các kết quả hay các kết luận.

</div>
<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-transparency.png")
```

</div>

## Các Nguyên Lý của Phân Tích Dữ Liệu - Có Tính Sao Chép Lại

<div class='left' style='float:left;width:48%'>

- Một phân tích là có tính sao chép nếu một người nào đó không phải là người phân tích ban đầu có thể sử dụng phân tích đó dễ dàng và tạo ra kết quả như nhau, giống với kết quả mà người phân tích ban đầu tạo ra.

- Vấn đề quan trọng của việc sao chép là khả năng có thể dùng được của cầu phần phân tích đối với người khác.

- Tính sao chép là quan trọng trong ngữ cảnh cùng làm việc theo Luồng công việc (Workflow).

</div>
<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-reproducible.png")
```

</div>

<!-- ## Các Nguyên Lý của Phân Tích Dữ Liệu - Tính Nguyên Tử -->

<!-- <div class='left' style='float:left;width:48%'> -->

<!-- - the data analyst first chooses a principle to investigate a hypothes is -->
<!-- or scientific question (Stage A).  -->

<!-- - the analyst alternates between Stages B and C until the analyst exits the atomic unit, either by choosing to end the line of investigation or choosing to invoke a new principle. -->

<!-- </div> -->
<!-- <div class='left' style='float:left;width:48%'> -->

<!-- ```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'} -->
<!-- knitr::include_graphics("assets/img/data-atomic.png") -->
<!-- ``` -->

<!-- </div> -->

<!-- ## Hiểu Rõ Dữ Liệu của Bạn (1) -->

<!-- **[Nền Tảng về Phân Tích Dữ Liệu: Hiểu Dữ Liệu Của Bạn. Phải Thật Sự, Thật Sự Hiểu Nó](https://towardsdatascience.com/data-science-foundations-know-your-data-really-really-know-it-a6bb97eb991c).** -->

<!-- > Bạn đồng nghiệp trong công ty, mặc dù họ không có chuyên môn về dữ liệu, nhưng họ là chuyên gia về lĩnh vực chuyên môn của họ. Nên nếu bạn là một nhà phân tích dữ liệu còn non trẻ, bạn không nên  -->

<!-- Your colleagues in business, despite not being data people, aren’t idiots of any sort (hopefully). They’re domain experts that know tons about the business. Don’t be the junior analyst presenting a brilliant new finding to the CEO that gets asked "so did you take out the 3 special high volume contract accounts we have?”, doesn’t end well. -->

<!-- ```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'} -->
<!-- url <- "https://miro.medium.com/max/875/1*MmmBLiC7GACdZWwZFykoqQ.png" -->
<!-- knitr::include_graphics(url) -->
<!-- #knitr::include_graphics("../Visual/data-atomic.png") -->
<!-- ``` -->

## Hiểu Dữ Liệu của Bạn

<div class='left' style='float:left;width:48%'>

- Hiểu về cấu trúc dữ liệu
- Hiểu về việc ghi dữ liệu
- Hiểu về các thói quen trong kinh doanh
- Hiểu dữ liệu từ đâu tới, nó được hình thành như thế nào
- Hiểu về các quyết định liên quan đến thu thập

<span style="color: red;"> Bạn không bao giờ có được bộ dữ liệu hoàn hảo hay hiểu được rõ ràng về toàn bộ dữ liệu bạn hiện có. Bởi vì, chỉ khi bạn đã nắm bắt và hiểu rõ về một điều gì đó, thì sẽ có những thuộc tính mới được thêm vào, thay đổi về tiêu chuẩn, hay hệ thông bị lỗi thời. Do đó, bạn nên **Tiếp Tục Học**, ghi lại nhưng điều cần thiết, để lại dấu tích để bạn có thể quay lại với công việc mà không mất thêm quá nhiều thời gian.</span>

</div>
<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
url <- "https://miro.medium.com/max/865/1*X1n4WeN_cWohvxRkeVCPMQ.png"
knitr::include_graphics(url)
#knitr::include_graphics("../Visual/data-atomic.png")
```

</div>

## Khuynh Hướng Sai của Dữ Liệu

<div class='left' style='float:left;width:48%'>

[Caroline Criado-Perez On Data Bias And 'Invisible Women'](https://www.npr.org/2019/03/17/704209639/caroline-criado-perez-on-data-bias-and-invisible-women)

<iframe src="https://www.npr.org/player/embed/704209639/704209640" width="90%" height="180" frameborder="0" scrolling="no" title="NPR embedded audio player"></iframe>

<div class='left' style='float:left;width:48%'>
```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
url <- "https://i.gr-assets.com/images/S/compressed.photo.goodreads.com/books/1617113740i/41104077._UY473_SS473_.jpg"
knitr::include_graphics(url)
```
</div>
<div class='left' style='float:left;width:48%'>
```{r, echo=FALSE, out.width="68%", fig.cap="", fig.align = 'center'}
url <- "https://api.time.com/wp-content/uploads/2016/08/weapons-of-math-destruction.jpeg"
knitr::include_graphics(url)
```
</div>

</div>
<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url <- "https://productiveclub.com/wp-content/uploads/2020/07/sampling-bias.jpeg"
knitr::include_graphics(url)
```
</div>

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/04-tidydata - Copy.Rmd
update

---
title     : "Các Nguyên Lý Cơ Bản Trong Phân Tích Dữ Liệu"
subtitle  : "Làm Sạch và Hợp Thức Hóa Dữ Liệu, Dữ Liệu Gọn Gàng"
author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
date      : "4-tháng 8-2023"
#date      : "`r format(Sys.time(), '%d %B, %Y')`"
output    : 
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

## Data Cleansing and Validation (1)

> Data cleansing or data cleaning is the process of detecting and correcting (or removing) corrupt or inaccurate records from a record set, table, or database and refers to identifying incomplete, incorrect, inaccurate or irrelevant parts of the data and then replacing, modifying, or deleting the dirty or coarse data. Data cleansing may be performed interactively with data wrangling tools, or as batch processing through scripting
>
> `r tufte::quote_footer('--- https://en.wikipedia.org/wiki/Data_cleansing')`

<span style="color: red;"> Data scientists spend **60% of their time on cleaning** and organizing data. Collecting data sets comes second at 19% of their time, meaning data scientists spend around **80% of their time on preparing and managing data for analysis**.</span>

> Data Validation is an activity verifying whether or not a combination of values is a member of a set of acceptable combinations
>
> `r tufte::quote_footer('--- [Zio et al. 2016](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)')`

## Data Cleansing and Validation (2)

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/data-validationprocess.png")
```

Source: [Methodology for Data Valication 1.0](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)

</div>

<div class='right' style='float:right;width:48%'>

- Validation Level 0: consistency with the expected IT structural requirements

- Validation Level 1: consistency within the data set
- Validation Level 2:  consistency with other data sets within the same domain and within the same data source
- Validation Level 3:  consistency within the same domain between different data sources
- Validation Level 4:  consistency between separate domains in the same data provider 
- Validation Level 5: consistency with data of other data providers.

</div>

## Tidy Data - Structure

<div class='left' style='float:left;width:48%'>

**Untidy**

Typical representation

```{r echo=FALSE, message=FALSE,warning=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read_excel("../data/tidydata.xlsx",sheet="treatment01")
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '')
data %>%
  kbl() %>%
#  kable_classic(full_width = F, html_font = "Cambria")
  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
#kable(data, booktabs = TRUE)
```

Another typical representation

```{r echo=FALSE, message=FALSE,warning=FALSE}
# library(ggplot2)
data=read_excel("../data/tidydata.xlsx",sheet="treatment02")
data %>%
  kbl() %>%
#  kable_classic(full_width = F, html_font = "Cambria")
  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
#kable(data, booktabs = TRUE)
```

- What are the <span style="color: red;">**variable**</span>, <span style="color: red;">**observation**</span>, <span style="color: red;">**value**</span>?

</div>

<div class='right' style='float:right;width:48%'>
> Happy families are all alike; every unhappy family is unhappy in its own way
>
> `r tufte::quote_footer('--- Leo Tolstoy')`

Like families, tidy datasets are all alike but every messy dataset is messy in its own way. Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).

```{r, echo=FALSE, out.width="60%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/tidydata.png")
```

</div>

## Tidy Data - Definition

<div class='left' style='float:left;width:48%'>

A dataset is a collection of <span style="color: red;">**values**</span>, usually either numbers (if quantitative) or strings (if qualitative). Values are organised in two ways. Every value belongs to a <span style="color: red;">**variable**</span> and an <span style="color: red;">**observation**</span>. A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units. An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes.

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/tidydata_2.jpg")
```

source: [Openscapes](https://www.openscapes.org/blog/2020/10/12/tidy-data/)
</div>

<div class='right' style='float:right;width:48%'>

**Tidy data is a standard way of mapping the meaning of a dataset to its structure. A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types.** In tidy data:

  - Each variable forms a column.
  - Each observation forms a row.
  - Each type of observational unit forms a table.

```{r echo=FALSE, message=FALSE,warning=FALSE}
# library(ggplot2)
data=read_excel("../data/tidydata.xlsx",sheet="treatment03")
data %>%
  kbl() %>%
#  kable_classic(full_width = F, html_font = "Cambria")
  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
#kable(data, booktabs = TRUE)
```

</div>

## Tidy Data - Tydy Our Room with Marie Kondo's Way (KONMARI)

<div class='left' style='float:left;width:48%'>

```{r, echo=FALSE, out.width="60%", fig.cap="", fig.align = 'center'}
url <- "https://kenh14cdn.com/thumb_w/620/2019/1/30/artboard-1-copy-3-1548865803939988101634.jpg"
knitr::include_graphics(url)
#knitr::include_graphics("../Visual/data-atomic.png")
```

[Phương pháp Marie Kondo: Nghệ thuật dọn dẹp nhà cửa](https://www.btaskee.com/kinh-nghiem-hay/don-dep-nha-voi-phuong-phap-marie-kondo/)

</div>

<div class='right' style='float:right;width:48%'>

Folding a perfect underwear drawer

<iframe width="560" height="315" src="https://www.youtube.com/embed/tglp9eWQEhY" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

Organize the world: Design Your Life to Spark Joy

<iframe width="560" height="315" src="https://www.youtube.com/embed/rAC4sprfqgU" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

</div>

## Tidying Messy Data (1)

<div class='left' style='float:left;width:48%'>
- Column headers are values, not variable names.

- Multiple variables are stored in one column.

- Variables are stored in both rows and columns.

- Multiple types of observational units are stored in the same table.

- A single observational unit is stored in multiple tables
</div>

<div class='left' style='float:left;width:48%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://www.openscapes.org/img/blog/tidydata/tidydata_3.jpg"
knitr::include_graphics(url)
```
source: [Openscapes](https://www.openscapes.org/blog/2020/10/12/tidy-data/)

</div>

## Column headers are values, not variable names (1)

<div class='left' style='float:left;width:48%'>

Messy Data

```{r echo=FALSE, message=FALSE,warning=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read.csv("../data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
data <-data%>%select(c(2:8))
#colnames(data) <- c('religion','<$10k','$10-20k','$20-30k','$30-40k','$40-50k','$50-75k')
library(DT)
p<-datatable(data,rownames = FALSE)
p
#library(knitr)
#library(kableExtra)
#data %>% 
#  kbl() %>%
#  kable_classic(full_width = F, html_font = "Cambria")
#  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
#kable(data, booktabs = TRUE)
```

</div>

<div class='right' style='float:left;width:15%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
knitr::include_graphics(url)
```

- Variables

  - religion
  - income
  - frequency

</div>

<div class='right' style='float:left;width:30%'>

Tidy Data

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../code/pew-tidying.R")
library(DT)
p<-datatable(molten.df,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left', targets = c(1,2)))
            ,autoWidth = TRUE)
             )
p
```

</div>

## Column headers are values, not variable names (2)

Billboard Ranking

```{r echo=FALSE, message=FALSE,warning=FALSE}
df=read.csv("../data/tidy-data/data/billboard.csv", sep=",", header=TRUE)
p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Column headers are values, not variable names (3)

Billboard Ranking

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
p<-datatable(clean,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left', targets = c(1,2,3,4,5,6)))
            ,autoWidth = TRUE)
             )
p
```

## Multiple variables stored in one column (1)

WHO data on turboculis (TB)

```{r echo=FALSE, message=FALSE,warning=FALSE}
df=read.csv("../data/tidy-data/data/tb.csv", sep=",", header=TRUE)
p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Multiple variables stored in one column (2)

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/tb.r")
library(DT)
p<-datatable(raw,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Multiple variables stored in one column (3)

<div class='right' style='float:left;width:45%'>

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/tb.r")
library(DT)
p<-datatable(clean1,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```

</div>

<div class='right' style='float:left;width:10%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
knitr::include_graphics(url)
```
</div>

<div class='right' style='float:left;width:45%'>
```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/tb.r")
library(DT)
p<-datatable(clean,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```
</div>

## Variables are stored in both rows and columns (1)

Global Historical Climatology Network for one weather station (MX17004)

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/weather.r")
library(DT)
p<-datatable(raw,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Variables are stored in both rows and columns (2)

<div class='right' style='float:left;width:45%'>

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/weather.r")
library(DT)
p<-datatable(clean1,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```

</div>

<div class='right' style='float:left;width:10%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
knitr::include_graphics(url)
```
</div>

<div class='right' style='float:left;width:45%'>
```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/weather.r")
library(DT)
p<-datatable(clean2,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```
</div>

## Multiple types in one table (1)

Billboard Ranking

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
p<-datatable(clean,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left', targets = c(1,2,3,4,5,6)))
            ,autoWidth = TRUE)
             )
p
```

## Multiple types in one table (2)

<div class='right' style='float:left;width:45%'>

Song Table

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
library(DT)
p<-datatable(song,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```

</div>

<div class='right' style='float:left;width:10%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://www.svgrepo.com/show/3193/arrows-couple.svg"
knitr::include_graphics(url)
```
</div>

<div class='right' style='float:left;width:45%'>

Ranking Table

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
library(DT)
p<-datatable(rank,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```
</div>

## Examples

**Energy Audit**

This project was conducted in 2018 when I was with GHD. I was the PM for a consultancy project of about 1 million USD for 6 months. The project includes following scope
- Reliability check of all major equipment and systems of 20+ large pump stations in Metro Manila
- Energy Audit
- Conceptual design to upgrade the system

[https://namkyodai.github.io/post/2019-09-15-excel-combine/](https://namkyodai.github.io/post/2019-09-15-excel-combine/)

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/04-tidydata.Rmd
update

--- 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/04-tidydata - Copy.Rmd
@@
 title     : "Các Nguyên Lý Cơ Bản Trong Phân Tích Dữ Liệu"
-subtitle  : "Làm Sạch và Hợp Thức Hóa Dữ Liệu, Dữ Liệu Gọn Gàng"
+subtitle  : "Làm Sạch và Xác Thực Dữ Liệu, Dữ Liệu Gọn Gàng"
 author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
@@
 
-## Data Cleansing and Validation (1)
+## Làm Sạch và Xác Thực Dữ Liệu (1)
 
-> Data cleansing or data cleaning is the process of detecting and correcting (or removing) corrupt or inaccurate records from a record set, table, or database and refers to identifying incomplete, incorrect, inaccurate or irrelevant parts of the data and then replacing, modifying, or deleting the dirty or coarse data.
-Data cleansing may be performed interactively with data wrangling tools, or as batch processing through scripting
+> Làm sạch dữ liệu là một quá trình phát hiện, sửa, loại bỏ, thay đổi các dòng dữ liệu không chính xác/ bị lỗi từ các bảng ghi, các cơ sở dữ liệu.
+Công việc làm sạch có thể được tiến hành bởi các công cụ chuyên dụng, hoặc có thể được làm theo lô thông qua dòng lệnh.
+
 >
@@
 
-<span style="color: red;"> Data scientists spend **60% of their time on cleaning** and organizing data.
-Collecting data sets comes second at 19% of their time, meaning data scientists spend around **80% of their time on preparing and managing data for analysis**.</span>
+<span style="color: red;"> Các nhà khoa học dữ liệu sử dụng **60% thời gian của họ để làm sạch ** và thiết lập dữ liệu.
+Thu thập dữ liệu chiếm khoảng 19% thời gian, điều này có nghĩa là khoảng **80% thời gian của họ là dành cho việc chuẩn bị và quản lý dữ liệu**.</span>
 
-> Data Validation is an activity verifying whether or not a combination of values is a member of a set of acceptable combinations
+> Xác thực dữ liệu là một hoạt động nhằm xác thực một tập hợp các giá trị là một thành viên của các tập hợp được chấp nhận.
 >
@@
 
-## Data Cleansing and Validation (2)
+## Làm Sạch và Xác Thực Dữ Liệu (2)
 
@@
 
-Source: [Methodology for Data Valication 1.0](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)
+Source: [Phương Pháp Xác Thực Dữ Liệu 1.0](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)
 
@@
 
-- Validation Level 0: consistency with the expected IT structural requirements
-
-- Validation Level 1: consistency within the data set
-- Validation Level 2:  consistency with other data sets within the same domain and within the same data source
-- Validation Level 3:  consistency within the same domain between different data sources
-- Validation Level 4:  consistency between separate domains in the same data provider 
-- Validation Level 5: consistency with data of other data providers.
+- Xác Thực Cấp Độ 0: Tính nhất quán với các yêu cầu về cấu trúc IT
+- Xác Thực Cấp Độ 1: Tính nhất quán trong chính gói dữ liệu
+- Xác Thực Cấp Độ 2: Tính nhất quán với các gói dữ liệu trong cùng một miền và trong cùng một nguồn dữ liệu
+- Xác Thực Cấp Độ 3:  Tính nhất quán trong cùng một miền với các nguồn dữ liệu khác nhau
+- Xác Thực Cấp Độ 4:  Tính nhất quán giữa các miền tách biệt trong cùng một nhà cung cấp
+- Xác Thực Cấp Độ 5: Tính nhất quán với dữ liệu của các nhà cung cấp.
 
@@
 
-## Tidy Data - Structure
+## Dữ Liệu Gọn Gàng - Cấu Trúc
 
@@
 
-**Untidy**
+**Không Gọn Gàng**
 
-Typical representation
+Thể hiện điển hình
 
@@
 
-Another typical representation
+Thể Hiện Điển Hình (Tiếp)
 
@@
 
-- What are the <span style="color: red;">**variable**</span>, <span style="color: red;">**observation**</span>, <span style="color: red;">**value**</span>?
+- Cái gì là <span style="color: red;">**Biến**</span>, <span style="color: red;">**Quan sát**</span>, <span style="color: red;">**Giá trị**</span>?
 
@@
 <div class='right' style='float:right;width:48%'>
-> Happy families are all alike; every unhappy family is unhappy in its own way
+> Các gia đình hạnh phúc thì đều giống nhau, các gia đình không hạnh phúc thì không hạnh phúc theo từng cách riêng của họ
+
 >
@@
 
-Like families, tidy datasets are all alike but every messy dataset is messy in its own way.
-Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).
+Như là các gia đình, các dữ liệu gọn gàng thì đều giống nhau, nhưng các dữ liệu lộn xộn, không gọn gàng thì lộn xộn theo từng cách riêng của nó và không giống nhau.
+Các gói dữ liệu gọn gàng cung cấp một cách tiêu chuẩn để kết nối cấu trúc của một tập dữ liệu (theo sơ đồ vật lý) và với nghữ nghĩa của nó (nghĩa).
 
@@
 
-## Tidy Data - Definition
+## Dữ Liệu Gọn Gàng - Định Nghĩa
 
@@
 
-A dataset is a collection of <span style="color: red;">**values**</span>, usually either numbers (if quantitative) or strings (if qualitative).
-Values are organised in two ways.
-Every value belongs to a <span style="color: red;">**variable**</span> and an <span style="color: red;">**observation**</span>.
-A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units.
-An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes.
+Một gói dữ liệu là một tập hợp các giá trị, thường ở dạng số hay dạng chữ.
+Các giá trị này được thiết lập theo 2 cách.
+Toàn bộ các giá trị thuộc về một **Biến** và một **Quan Sát**.
+Một biến chứa đựng trong nó toàn bộ các giá trị để đo lường cho cùng một thuộc tính (như chiều cao, nhiệt độ, thời gian) trên các đơn vị.
+Một Quan Sát chứa toàn bộ các giá trị được đo lường trên cùng một đơn vị (như một người, một ngày, một chủng tộng) trên các thuộc tính.
 
@@
 
-source: [Openscapes](https://www.openscapes.org/blog/2020/10/12/tidy-data/)
+Nguồn: [Openscapes](https://www.openscapes.org/blog/2020/10/12/tidy-data/)
 </div>
@@
 
-**Tidy data is a standard way of mapping the meaning of a dataset to its structure.
-A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types.** In tidy data:
+** Dữ liệu gọn gàng là một cách tiêu chuẩn của công việc chuyển thể (bản đồ hóa) ý nghĩa của một gói dữ liệu tới cấu trúc của nó.
+Một gói dữ liệu là lộn xộn hay gọn gàng phụ thuộc vào việc các dòng, các cột và các bảng của chúng có tương tích với các quan sát, các biến, và loại dữ liệu như thế nào**.
+Trong một dữ liệu gọn gàng thì:
 
-  - Each variable forms a column.
-  - Each observation forms a row.
-  - Each type of observational unit forms a table.
+  - Mỗi một biến cấu thành một cột.
+  - Mỗi một quan sát cấu thành một hàng.
+  - Mỗi một loại đơn vị quan sát cấu thành một bảng.
 
@@
 
-## Tidy Data - Tydy Our Room with Marie Kondo's Way (KONMARI)
+## Dữ Liệu Gọn Gàng - Dọn Dẹp Phòng Của Bạn theo Cách của Chuyên Gia Marie Kondo (KONMARI)
 
@@
 
-Folding a perfect underwear drawer
+Gấp Đồ Lót Trong Tủ Quần Áo
 
@@
 
-Organize the world: Design Your Life to Spark Joy
+Thiết Kế Cuộc Sống Để Luôn Vui Vẻ
 
@@
 
-## Tidying Messy Data (1)
+## Dọn Dẹp Dữ Liệu Lộn Xộn (1)
 
 <div class='left' style='float:left;width:48%'>
-- Column headers are values, not variable names.
 
-- Multiple variables are stored in one column.
+- Tên các cột là giá trị, không phải là Biến
+- Nhiều biến được ghi trong một cột
 
-- Variables are stored in both rows and columns.
+- Các biến được lưu ở cả hàng và cột
 
-- Multiple types of observational units are stored in the same table.
+- Các loại dữ liệu của nhiều đối tượng quan sát được ghi trong cùng một bảng
 
-- A single observational unit is stored in multiple tables
+- Một đối tượng quan sát được ghi ở nhiều bảng
+
 </div>
@@
 ```{r echo=FALSE, message=FALSE,warning=FALSE}
-library(readxl)
-library(dplyr)
-# library(ggplot2)
-data=read.csv("../data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
-data <-data%>%select(c(2:8))
-#colnames(data) <- c('religion','<$10k','$10-20k','$20-30k','$30-40k','$40-50k','$50-75k')
-library(DT)
-p<-datatable(data,rownames = FALSE)
-p
-#library(knitr)
-#library(kableExtra)
-#data %>% 
-#  kbl() %>%
-#  kable_classic(full_width = F, html_font = "Cambria")
-#  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
-#kable(data, booktabs = TRUE)
-```
-
+source("../Code/pew-tidying.R")
+#library(dplyr)
+#library(reshape2)
+#df=read.csv("../Data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
 </div>
-
-<div class='right' style='float:left;width:15%'>
-
-```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
-url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
-knitr::include_graphics(url)
-```
-
-- Variables
-
-  - religion
-  - income
-  - frequency
-
-</div>
-
-<div class='right' style='float:left;width:30%'>
-
-Tidy Data
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../code/pew-tidying.R")
-library(DT)
-p<-datatable(molten.df,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left', targets = c(1,2)))
-            ,autoWidth = TRUE)
-             )
-p
-```
-
-</div>
-
-## Column headers are values, not variable names (2)
-
-Billboard Ranking
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-df=read.csv("../data/tidy-data/data/billboard.csv", sep=",", header=TRUE)
-p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'),
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
-             )
-p
-```
-
-## Column headers are values, not variable names (3)
-
-Billboard Ranking
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/billboard.r")
-p<-datatable(clean,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left', targets = c(1,2,3,4,5,6)))
-            ,autoWidth = TRUE)
-             )
-p
-```
-
-## Multiple variables stored in one column (1)
-
-WHO data on turboculis (TB)
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-df=read.csv("../data/tidy-data/data/tb.csv", sep=",", header=TRUE)
-p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'),
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
-             )
-p
-```
-
-## Multiple variables stored in one column (2)
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/tb.r")
-library(DT)
-p<-datatable(raw,rownames = FALSE,extensions = c('FixedColumns'),
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
-             )
-p
-```
-
-## Multiple variables stored in one column (3)
-
-<div class='right' style='float:left;width:45%'>
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/tb.r")
-library(DT)
-p<-datatable(clean1,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE)
-             )
-p
-```
-
-</div>
-
-<div class='right' style='float:left;width:10%'>
-```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
-url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
-knitr::include_graphics(url)
-```
-</div>
-
-<div class='right' style='float:left;width:45%'>
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/tb.r")
-library(DT)
-p<-datatable(clean,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE)
-             )
-p
-```
-</div>
-
-## Variables are stored in both rows and columns (1)
-
-Global Historical Climatology Network for one weather station (MX17004)
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/weather.r")
-library(DT)
-p<-datatable(raw,rownames = FALSE,extensions = c('FixedColumns'),
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
-             )
-p
-```
-
-## Variables are stored in both rows and columns (2)
-
-<div class='right' style='float:left;width:45%'>
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/weather.r")
-library(DT)
-p<-datatable(clean1,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE)
-             )
-p
-```
-
-</div>
-
-<div class='right' style='float:left;width:10%'>
-```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
-url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
-knitr::include_graphics(url)
-```
-</div>
-
-<div class='right' style='float:left;width:45%'>
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/weather.r")
-library(DT)
-p<-datatable(clean2,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE)
-             )
-p
-```
-</div>
-
-## Multiple types in one table (1)
-
-Billboard Ranking
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/billboard.r")
-p<-datatable(clean,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left', targets = c(1,2,3,4,5,6)))
-            ,autoWidth = TRUE)
-             )
-p
-```
-
-## Multiple types in one table (2)
-
-<div class='right' style='float:left;width:45%'>
-
-Song Table
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/billboard.r")
-library(DT)
-p<-datatable(song,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE)
-             )
-p
-```
-
-</div>
-
-<div class='right' style='float:left;width:10%'>
-```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
-url = "https://www.svgrepo.com/show/3193/arrows-couple.svg"
-knitr::include_graphics(url)
-```
-</div>
-
-<div class='right' style='float:left;width:45%'>
-
-Ranking Table
-
-```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../data/tidy-data/data/billboard.r")
-library(DT)
-p<-datatable(rank,rownames = FALSE,
-             options = list(pageLength = 10,
-            columnDefs = list(list(className = 'dt-left'))
-            ,autoWidth = TRUE)
-             )
-p
-```
-</div>
-
-## Examples
-
-**Energy Audit**
-
-This project was conducted in 2018 when I was with GHD. I was the PM for a consultancy project of about 1 million USD for 6 months. The project includes following scope
-- Reliability check of all major equipment and systems of 20+ large pump stations in Metro Manila
-- Energy Audit
-- Conceptual design to upgrade the system
-
-[https://namkyodai.github.io/post/2019-09-15-excel-combine/](https://namkyodai.github.io/post/2019-09-15-excel-combine/)

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week2-01-rdms.Rmd
update

--- 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/01-gioi-thieu.Rmd
@@
 ---
-title     : "Phân Tích và Tính Toán  với Dữ Liệu Đô Thị <br> Urban Computation"
-subtitle  : "Giới Thiệu Chung"
-author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
-date      : "4-tháng 8-2023"
+title     : "Fundementals of Data Analytics"
+subtitle  : "An Introduction"
+author    : "[Nam LE](https://namkyodai.github.io)"
+date      : "8-Jan-2021"
 output    : 
@@
     highlight: pygments
-    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH"
+    footer: "Copyright of ASQ Vietnam Co. Ltd - Prepared for course at ISCM, UEH"
     includes:
@@
 
-## Giới Thiệu Bản Thân
+## About Me
 
-<div class='left' style='float:left;width:50%'>
-- Tiểu Sử
-  - Ngành nghiên cứu chính về phân tích rủi do bằng phương pháp định lượng và quản lý cơ sở hạ tầng xây dựng và các hệ thống kĩ thuật. 
-  - 4 năm sống ở Nhật (làm Tiến Sĩ tại Đại Học Kyoto và 1  năm làm nghiên cứu sau TS tại ĐH Osaka, Nhật Bản)
-  - <2 năm ở Thái Lan (học Thạc Sĩ tại AIT từ năm 2002-2004)
-  - 6+ năm sống ở Thụy Sĩ (2010-2016, giảng dạy và nghiên cứu tại học viện kĩ thuật liên bang Thụy Sĩ, ETH Zurich)
-  - 6 năm làm tư vấn tại Manila, Philippines (2016-2021) cho 2 công ty (GHD và Arcadis)
-  - Từ cuối năm 2021, sống ở thủ đô New Delhi, Ấn Độ và thường xuyên về Việt Nam tham gia giảng dạy.
+<div class='left' style='float:left;width:48%'>
+- Bio
+  - My main major is quantitative risk analysis and asset management for buildings and engineering systems. Data analytics work is one of my interest.
+  - 4 years lived in Japan (Doctoral Study at Kyoto University and postdoc at Osaka University) 
+  - <2 years in Thailand (Master Study 2002-2004) 
+  - 6 years in Switzerland (2010-2016, ETH Zurich) 
+  - 6 years in the Philippines (2016-2021) 
+  - Currently living in several locations Vietnam, Philippines, and India.
 
-- Các môn thể thao yêu thích
-  - Chạy bộ đường dài `r emo::ji("running")``r emo::ji("running")``r emo::ji("running_man")`
-  - Lặn biển `r emo::ji("swimming_man")`
-  - và một số môn khác ^_^ `r emo::ji("smile")``r emo::ji("smiling_imp")`
+- Sports
+  - Long distance running `r emo::ji("running")``r emo::ji("running")``r emo::ji("running_man")`
+  - Diving `r emo::ji("swimming_man")`
+  - and some others ^_^ `r emo::ji("smile")``r emo::ji("smiling_imp")`
 
 </div>
-<div class='right' style='float:right;width:50%'>
+<div class='right' style='float:right;width:48%'>
 
@@
 
-##  Giới Thiệu Bản Thân
+### About Me
 
-<div class='left' style='float:left;width:50%'>
+<div class='left' style='float:left;width:48%'>
 
@@
 </div>
-
-<div class='right' style='float:right;width:50%'>
+<div class='right' style='float:right;width:48%'>
 
@@
 
-## Thành Viên của Lớp
+## About YOU
 
@@
 library(lubridate)
-df=read_excel("../data/hocvien.xlsx",sheet="Sheet1",skip = 0)
-df$`Ngày Sinh` <- dmy(df$`Ngày Sinh`)
+df=read_excel("../data/students.xlsx",sheet="student",skip = 0)
+df$`Ngày sinh` <- as.Date(df$`Ngày sinh`,format = "%d/%m/%y")
 p <- datatable(df,rownames = FALSE, filter = 'top',
@@
 
-##  Thành Viên của Lớp
-
+## Voting for our Class Monitor
 <div class='left' style='float:left;width:50%'>
-
-```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
-source("../code/hocvien.R")
-boxplot(df$namsinh,
-    #    main = "Mean ozone in parts per billion at Roosevelt Island",
-        xlab = "Năm Sinh",
-        ylab = "",
-        col = "orange",
-        border = "brown",
-        horizontal = TRUE,
-        notch = TRUE
-)
+```{r, echo=FALSE, out.width="95%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/classmember-01.png")
 ```
@@
 
-<div class='right' style='float:right;width:50%'>
+<div class='left' style='float:left;width:50%'>
 
-```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
-source("../code/hocvien.R")
-ggplotly(p3)
+```{r, echo=FALSE, out.width="95%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/classmember-02.png")
+```
+
+</div>
+
+### Random voting
+```r
+# Selecting a Monitor for our Class.
+library(purrr)
+# reading data file
+library(readxl) 
+library(dplyr)
+library(ggChernoff)
+df=read_excel("../Data/students.xlsx",sheet="student",skip = 0)
+
+l <- length(df$Stt)
+
+df1 <- replicate(sample(1:l, 1, replace = FALSE), n = 10000)
+df1 <- data.frame(df1)
+df2 <- data.frame(table(df1))
+which.max(df2$Freq)
+
+df4 <- df %>%
+  select(c(1,2,3,5,7,13,14)) %>%
+  mutate(freq = df2$Freq)
+
+names(df4) <- c('ID', 'lastname', 'firstname', 'age', 'sex', 'field', 'fullname', 'freq')
+
+df4 <- data.frame(df4)
+
+library(ggplot2)
+
+ggplot(df4) +
+  aes(age, freq, fill = field) +
+  geom_chernoff()+
+  labs(y = "Số phiếu bầu", x ="Độ Tuổi")+ 
+  geom_text(aes(label=paste(lastname,firstname)), position=position_dodge(width=1), vjust=2, cex =2.5)
+
+ggplot(df4) +
+  aes(age, freq, fill = sex) +
+  geom_chernoff()+
+  labs(y = "Số phiếu bầu", x ="Độ Tuổi")+ 
+  geom_text(aes(label=paste(lastname,firstname)), position=position_dodge(width=1), vjust=2, cex =2.5)
+
+summary(df4$age)
+
+cat(" Chúc mừng bạn", paste(df$'Họ'[which.max(df2$Freq)]), paste(df$'Tên'[which.max(df2$Freq)]))
+```
+
+## About the Course (1) - the Venn Diagram
+
+```{r, echo=FALSE, out.width="40%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/datascience-venn-01.png")
+```
+
+## About the Course (2) - Data Science vs Data Analytics
+
+```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/datascience-venn-02.png")
+```
+
+## About the Course (3) - Data Science vs Data Analytics
+
+<div class='left' style='float:left;width:30%'>
+- Some basic statistical methods and computing
+- Data and Tidy data
+- Basic data management system
+- Basic programming with R
+- Data analytics tools
+- Data visualization
+- Data storytelling
+</div>
+
+<div class='left' style='float:left;width:60%'>
+```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/Data-Engineer-vs-Data-Scientist-vs-Data-Analyst.png")
 ```
@@
 
-## Chương Trình Học và Thời Khóa Biểu
+## What are the benefits?
+
+<div class='left' style='float:left;width:48%'>
+
+```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/before-after-trump-un.png")
+```
+
+```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/cartoon01.png")
+```
+
+</div>
+
+<div class='right' style='float:right;width:48%'>
+
+**You do not need to become a data scientist, but you can be knowledgeable and capable with data analytic works**
+
+- Collecting data and recording it properly
+- The difference and purpose of different database management systems
+- Automation and reproduction
+- Increase the level of efficiency and productivity
+- Transformation
+</div>
+
+## Course Program and Schedule
 
@@
 
-## Mong Muốn và Yêu Cầu
+## References and Reading Materials
+
+<div class='left' style='float:left;width:48%'>
+- Books & Platforms
+  - [R for Data Science](https://r4ds.had.co.nz/index.html)
+  - [Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html)
+  - [Github](https://github.com/)
+  - [R Statistical Computing](https://www.r-project.org/) and  [Rstudio](https://www.rstudio.com/)
+  - [R Graph Gallery](https://www.r-graph-gallery.com/)
+  - [Plotly R Open Source Graphing Library](https://plotly.com/r/)
+  - [Bookdown](https://bookdown.org/)s
+  <!-- - [Stat545](https://stat545.com/index.html) -->
+- R Packages
+  - [Tidyverse](https://www.tidyverse.org/), [Dplyr](https://dplyr.tidyverse.org/), [Ggplot2](https://ggplot2.tidyverse.org/), [DT](https://rstudio.github.io/DT/),  [Rattle](https://rattle.togaware.com/)
+
+- Repository and Data
+
+  - Google Drive: https://bit.ly/3qWe6Fh
+  - Github:       https://github.com/iscmueh/MS-DS001
+
+</div>
+
+<div class='left' style='float:left;width:48%'>
+
+```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/01_tidyverse_data_science.png")
+```
+
+</div>
+
+## Expectations and Requirements
 
@@
 data <- data.frame(
-  category=c("Học Trên Lớp", "Thảo Luận và Trình Bày", "Bài Thi"),
- count=c(40, 30, 30)
+  category=c("Class participation", "Assignment", "Exam"),
+ count=c(50, 20, 30)
 )
@@
 
-## Khoa Học về Dữ Liệu (1) - Sơ đồ Venn
-
-```{r, echo=FALSE, out.width="40%", fig.cap="", fig.align = 'center'}
-knitr::include_graphics("assets/img/khoahocdulieu01.png")
-```
-
-## Khoa Học về Dữ Liệu (2) - Khoa Học Dữ Liệu vs Phân Tích Dữ Liệu
-
-```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
-knitr::include_graphics("assets/img/khoahocdulieu02.png")
-```
-
-## Khoa Học về Dữ Liệu (3) - Khoa Học Dữ Liệu vs Phân Tích Dữ Liệu
-
-<div class='left' style='float:left;width:30%'>
-- Một vài phương pháp tính toán xác xuất thống kê cơ bản
-- Dữ liệu và Dữ Liệu Sạch
-- Cơ bản về quản lý dữ liệu
-- Giới thiệu sơ qua về phần mềm R
-- Học cơ bản về nền tảng phân tích KNIME và PowerBI
-- Các công cụ phân tích khác
-- Cơ bản về trực quan hóa dữ liệu
-- Các kết luận rút ra từ phân tích dữ liệu
-</div>
-
-<div class='left' style='float:left;width:60%'>
-```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
-knitr::include_graphics("assets/img/Data-Engineer-vs-Data-Scientist-vs-Data-Analyst.png")
-```
-</div>
-
-## Các Bước Trong Phân Tích Dữ Liệu
-
-```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
-knitr::include_graphics("assets/img/data-science-explore.png")
-```
-
-## Vậy Học Thì Có Gì Là Bổ Ích?
-
-<div class='left' style='float:left;width:48%'>
-
-```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
-knitr::include_graphics("assets/img/cartoon01.png")
-```
-
-</div>
-
-<div class='right' style='float:right;width:48%'>
-
-**Bạn không cần phải trở thành một nhà nghiên cứu dữ liệu hay phân tích dữ liệu toàn thời gian, nhưng bạn có thể hiểu và làm công việc liên quan đến phân tích dữ liệu hiệu quả.**
-
-- Thu thập dữ liệu và ghi dữ liệu một cách bài bản và hợp lý
-- Sự khác biệt cơ bạn của các hệ thống quản lý dữ liệu
-- Tự động hóa và sao lại qui trình công việc (workflow)
-- Nâng cao năng xuất và tính hiệu quả trong công việc
-- Chuyển đổi (số)
-
-</div>
-
-## Tài Liệu Tham Khảo
-
-<div class='left' style='float:left;width:48%'>
-- Sách và Hệ Thống
-  - [R for Data Science](https://r4ds.had.co.nz/index.html)
-  - [Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html)
-  - [Github](https://github.com/)
-  - [R Statistical Computing](https://www.r-project.org/) and  [Rstudio](https://www.rstudio.com/)
-  - [R Graph Gallery](https://www.r-graph-gallery.com/)
-  - [Plotly R Open Source Graphing Library](https://plotly.com/r/)
-  - [Bookdown](https://bookdown.org/)s
-  <!-- - [Stat545](https://stat545.com/index.html) -->
-- Các gói thư Viện R
-  - [Tidyverse](https://www.tidyverse.org/), [Dplyr](https://dplyr.tidyverse.org/), [Ggplot2](https://ggplot2.tidyverse.org/), [DT](https://rstudio.github.io/DT/),  [Rattle](https://rattle.togaware.com/)
-
-- Trang Web của KNIME
-  - https://www.knime.com/knimepress
-
-</div>
-
-<div class='left' style='float:left;width:48%'>
-
-```{r, echo=FALSE, out.width="100%", fig.cap="", fig.align = 'center'}
-knitr::include_graphics("assets/img/01_tidyverse_data_science.png")
-```
-
-</div>
-
-## Giới Thiệu Sơ Qua về Sự Phát Triển của R
+## A Brief on R Statistical Computing Program
 
@@
 ```{r, echo=FALSE, out.width="80%", fig.cap="", fig.align = 'center'}
-url <- "https://computationalbiomed.hms.harvard.edu/wp-content/uploads/2022/02/Robert-Gentleman-200x240-1.png"
+url <- "https://computationalbiomed.hms.harvard.edu/sites/g/files/mcu1016/files/profile-photos/Robert-Gentleman-200x240.png"
 knitr::include_graphics(url)
@@
 ```
-Trường Y Khoa - Đại Học Harvard
+Harvard Medical School
 
@@
 ```
-Giáo Sư về Xác Xuất và Thống Kê - Đại Học Auckland
+Professor of Statistics at the University of Auckland
 </div>
 
-- 1991 Triển Khai Ngôn ngữ S
-- 1993 Giới Thiệu R
-- 1995 Xuất bản miễn phí R là phần mềm miễn phí và mã nguồn mở
+- 1991 S implementation
+- 1993 Published R Code
+- 1995 Made R to be open source and FREE
 
@@
 
-Ngôn ngữ S được phát triển tại phòng Lab BELL Lab từ năm 1976
+S language was developed in BELL Lab since 1976
 
@@
 
-- 9 giải Nobels 
-- Thiết kế ra chương trình tối ưu AMPL Optimizer
-- Thiết kế ra chương trình C
+- 9 Nobels Recipients
+- AMPL Optimizer
+- C Programing Language
 
@@
 
-## Giới Thiệu qua về R (1)
+## An Introduction to R (1)
 
@@
 
-## Giới Thiệu qua về R  (2)
+## An Introduction to R (2)
 
 <div class='left' style='float:left;width:25%'>
-- Thư Mực Làm việc - Working directory
-- Tên File và cấu trúc - File name and structure
-- Tài liệu dự án 
-- Các loại dữ liệu
-- Kết cấu của dữ liệu 
+- Working directory
+- File name and structure
+- File names
+- Project documentation
+- Data types
+- Data structures
 </div>
@@
 
-## Kết Cấu Dữ Liệu (1)
+## Data Structures (1)
 
@@
 
-## Kết Cấu Dữ Liệu (2)
+## Data Structures (2)
 
@@
 
-## Quá Trình Xử Lý Dữ Liệu (Data Wrangling)
+## Data Wrangling
 
@@
 
-## Nhập Dữ Liệu vào Môi Trường R
+## Data Importing
 
@@
 - cvs/excel files
-- các bảng từ databases
+- tables from databases
 
-[Nhập Dữ Liệu vào Môi Trường R](http://www.sthda.com/english/wiki/importing-data-into-r)
+[Importing data into R](http://www.sthda.com/english/wiki/importing-data-into-r)
 
@@
 
-## Xuất/Ghi Dữ Liệu từ R
+## Data Exporting
 
 <div class='left' style='float:left;width:45%'>
-Các files thông thường
+Common data files
 
@@
 - cvs/excel files
-- các bảng từ databases
+- tables from databases
 
-[Xuất/Ghi Dữ Liệu từ R](http://www.sthda.com/english/wiki/exporting-data-from-r)
+[Exporting data from R](http://www.sthda.com/english/wiki/exporting-data-from-r)
 
@@
 </div>
-
-## Giới Thiệu qua về KNIME
-
-<div class='left' style='float:left;width:25%'>
-KNIME là một phần mềm và thường được các nhà phát triển gọi là hệ thống phân tích KNIME, do công ty KNIME AG có trụ sở ở thành phố Zurich, Thụy Sĩ phát triển.
-
-KNIME có cấu trúc và cách tiến hành phân tích dữ liệu theo dạng Workflow và theo khái niệm Kéo và Thả (Drag and Drop) nên rất trực quan đối với người dùng không có kiến thức về lập trình.
-
-KNIME có mã nguồn mở và miễn phí đối với người dùng cho bản tiêu chuẩn. KNIME đang bắt đầu được các doanh nghiệp lớn quan tâm và sử dụng, bên cạnh các phần mềm như PowerBI và Tableau.
-
-</div>
-
-<div class='left' style='float:left;width:75%'>
-
-```{r, echo=FALSE, out.width="70%", fig.cap="", fig.align = 'center', warning=FALSE}
-library(knitr)
-knitr::include_graphics("assets/img/Knime.jpg")
-```
-
-</div>

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week2-02-dplyr.Rmd
update

---
title     : "Nền Tảng Cần Thiết cho việc Phân Tích Dữ Liệu với R"
subtitle  : "Thay đổi dữ liệu với gói phụ kiện Dplyr"
author    : "[Lê Thanh Nam](https://namkyodai.github.io)"
date      : "Ngày 15, tháng 1, năm 2022"
output    :
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản quyền thuộc công ty TNHH ASQ Vietnam - Cùng phối hợp với Đại Học UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Dữ liệu thực hành

- Gói dữ liệu về dự án thẩm định công trình ()

- Gói dữ liệu iris có sẵn trên nền tảng R

- [Gói dữ liệu về giá nhà của thành phố boston](https://www.cs.toronto.edu/~delve/data/boston/bostonDetail.html)

- Gói dữ liệu về tình hình covid trên toàn thế giới

Gói dữ liệu được tải từ link sau

## Gói phụ kiện Dplyr

## Các động từ quan trọng của gói phụ kiện

- **select()**:     chọn các biến dựa vào tên của biến
- **filter()**:     lọc các trường hợp dựa vào giá trị của biến
- **arrange()**:    sắp xếp hay thay đổi thứ tự các hàng
- **mutate()**:     thêm các cột với giá trị là phương trình phụ thuộc (hoặc không phụ thuộc) vào các biến đã có.
- **summarise()**:    giản lược nhiều giá trị thành một giá trị tổng hợp

## Chọn các biến - sử dụng select()

## lọc và sắp sếp - sử dụng filter() và arrange()

### sắp sếp các quan sát

### lọc có điều kiện

### lọc và sắp sếp

## thêm các cột (biến) mới vào dữ liệu

### tính toán

### tính toán

### kết hợp với select(), mutate(), filter(), và arrange()

<!-- Aggregating data -->

## Các cách tính tổng hợp

### Động từ đếm count()

### Đếm theo XXX

### Đếm them YYY

### Thêm cột và đếm

## Tóm tắt, Gộp và tách dùng động từ summarize(), groupby() và ungroup()

### Tóm tắt 1

### Tóm tắt 2

### Tóm tắt 3

## the top_ verb

### selecting a county from each region

### finding the highest-income state in each region

### using summarize, top_n, và count together

<!-- Selecting and Transforming Data -->

## Chọn và Thay Đổi Dữ Liệu

### Chọn dữ liêu

#### Chọn cột

#### Chọn helpers

<!-- rename verb -->
### Thay đổi tên động từ

<!-- rename a column after count -->
#### thay đổi tên một cột sau khi đếm

<!-- rename a column as part of a select -->
#### thay đổi tên một cột như là một phần của lệch select

<!-- transmute verb -->
### Chuyển đổi động từ

<!-- Choosing among verb -->
#### Chọn giữa các động tư

<!-- using transmute -->
#### sử dụng động từ transmute

<!-- matching verbs to their definition -->
#### Tìm sự tương ứng các động từ

<!-- chosing among the four verb -->
#### chọn giữa 4 động từ

## Thực hành với gói dữ liệu AAA

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week4-01-modeling-methods.Rmd
update

---
title     : "Cơ Bản về Xữ Lý và Phân Tích Dữ Liệu"
subtitle  : "Các Phương Pháp Lập Mô Hình"
author    : "[TS Lê Thanh Nam](https://namkyodai.github.io/vi)"
date      : "5-Tháng 2-2022"
#date      : "`r format(Sys.time(), '%d %B, %Y')`"
output    : 
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Copyright of ASQ Vietnam Co. Ltd - Prepared for course at ISCM, UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

## Giới Thiệu về cách nhiệm vụ của công tác mô hình hóa

- Phân loại (Classification): Quyết định nếu thông tin thuộc một nhóm này hay nhóm khác
- Ghi điểm (Scoring): Tiên lượng/tính toán một giá trị số (ví dụ như giá cả, lượng mưa, hay đơn giản là xác xuất)
- Xếp hạng (ranking): Học để xắp xếp thứ tự các đối tượng theo sở thích
- Xếp nhóm (Clustering): Nhóm các đối tượng theo tính chất giống nhau
- Tìm mối liên hệ (finding relations): Tìm mối tương quan hay nguyên nhân gây ảnh hưởng thấy trong dữ liệu
- Mô tả đặc điểm (Characterization): Vẽ biểu đồ và tạo ra báo cáo từ dữ liệu.

## Lựa Chọn và Đánh Giá các Mô Hình

Làm thế nào để lựa chọn một mô hình phù hợp để giải quyết một bài toán kinh doanh/kĩ thuật cụ thể

Làm thế nào để đánh giá chất lượng và tính hiệu quả của các mô hình

- Bản đồ hóa các vấn đề liên quan đến công việc tới các nhiệm vụ của học máy
- Đánh giá chất lượng mô hình
- Kiểm định lại mô hình

### Bản đồ hóa

```{r, echo=FALSE, out.width="60%", fig.cap="", fig.align = 'center'}
knitr::include_graphics("assets/img/modelschema.png")
```

#### Phân loại (Classification)

```{r attr.source="style='display:inline-block;'", collapse=TRUE, echo = FALSE, message=FALSE,warning=FALSE}
library(readxl)
data=read_excel("../Data/modelling.xlsx",sheet="classification-methods")
library(DT)
p<-datatable(data,options = list(dom = 't'),rownames = FALSE)
p
```

Các ví dụ

- Phân tích và tiên đoán hành vi của khách hàng
- Phân loại tài liệu
- Lọc thư rác
- Phân loại ảnh
- Phân loại thông tin trang web
- Tiên đoán tỷ lệ bấm vào quảng cáo
- Phân loại sản phẩm
- Phân loại Malware
- Phân tích tính nhạy cảm của hình ảnh
- Phỏng đoán lỗi
- ETC

#### Ghi điểm (Scoring)

- Phương pháp hồi qui tuyến tính
- Phương pháp hồi qui nhị thức

#### Làm việc khi không biết đến các đích cụ thể

Các phương pháp đã được đề cập đòi hỏi chúng ta phải có một bộ dữ liệu học về các tình huống mà chúng ta đã biết trước kết quả. Nhưng trong thực tế thì chúng ta chưa chắc đã biết chắc chắn các biến đầu ra. Thay vào đó, chúng ta muốn quan tâm tới các hình thái học (patterns) và các mối quan hệ trong dữ liệu để từ đó chúng ta có thể hiểu được phần nào các đối tượng cần quan tâm (khách hàng hay một vấn đề nào đó trong kinh doanh).

Các tính huống này thì liên quan đến một dạng tiếp cận trong phân tích dữ liệu gọi là "Học mà không có hướng dẫn/giám sát". Với trường hợp này, chúng ta thường dùng các phương pháp nhóm (clustering).

- Phân nhóm K-means (K-means clustering)
- Thuật toán Apriori để tìm ra các qui luật liên kết
- Hàng xóm gần nhất

## Các Phương Pháp Ghi Nhớ (Memorization Methods)

Tiên đoán/ước lượng dựa chủ yếu vào thông số thống kê cơ bản:

- Các bảng tra cứu (lookup tables)
- Các phương pháp hàng xóm gần nhất (nearest neighbors)
- Phương pháp phân loại Naive Bayes
- Phương pháp cây quyết định (decision trees)

## Các phương pháp hàng xóm gần nhất (nearest neighbors)

## Phương pháp cây quyết định (decision trees)

Cây quyết định là một loại mô hình đơn giản dùng để tiên đoán theo từng mẩu (hay còn gọi là cành của cây). Dù đơn giản nhưng nó khá thú vị bởi vì [**GIẢ THUYẾT KHÔNG**](https://vi.wikipedia.org/wiki/Gi%E1%BA%A3_thuy%E1%BA%BFt_kh%C3%B4ng]) (Null Hypothesis) mà chúng ta cố gắng để làm tốt hơn thì thường là một hằng số cho cả gói dữ liệu, do đó chúng ta có thể xem cây quyết định như là một quá trình chia gói dữ liệu học thành những gói nhỏ và dùng hằng số đã được ghi nhớ cho từng gói dữ liệu nhỏ đó.

Phương pháp này thường được gọi là CART (Classification and Regression Trees) và được dùng để tiên lượng nhanh chóng cho cả các biến định lượng và biến định tính.

Ví dụ: **Ước tính lượng mưa hàng ngày.**

## Mô Hình Hồi Qui Tuyến Tính và Hồi Qui Nhị Thức
Các phương pháp lựa chọn các hàm đơn giản với cấu trúc cộng dồn (addictive structure)
- Hồi qui tuyến tính
- Hồi qui nhị thức

Các phương pháp này không những chỉ dùng để tiên lượng mà còn dùng để cung cấp thêm các thông tin liên quan đến mối quan hệ giữa các biến đầu vào và biến đầu ra.

## Các Phương Pháp không Giám Sát (unsupervized)

Các phương pháp trong học máy liên quan đến việc xếp nhóm (cluster) và liên quan đến tính liên kết (association).

Các phương pháp này không nhằm tiên đoán chính xác kết quả, mà chúng thường để dùng để tìm ra/phát hiện các mối quan hệ và cấu trúc ẩn trong dữ liệu.

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/2ndcars/2ndcars-dplyr.r

@@ -1,0 +1,207 @@
+###################
+#     BƯỚC 1: Khai báo các gói phụ kiện (package) R cần thiết để phân tích
+
+
+# -----------------------------------------------------------------
+library(dplyr) # gói phụ kiện để tương tác và thay đổi dữ liệu
+
+
+###################
+
+#     BƯỚC 2: Đọc dữ liệu vào R
+
+#toyota <- read.csv("ToyotaCorolla.csv")
+
+df <- read.csv("https://www.biz.uiowa.edu/faculty/jledolter/DataMining/ToyotaCorolla.csv")
+
+
+
+df <- data.frame(df) # Chuyển đổi dữ liệu thành dạng khung
+
+
+###################
+#   BƯỚC 3: Thực hành với gói phụ kiện Dplyr
+
+#---Các bạn có thể đọc thêm hướng dẫn từ link **https://dplyr.tidyverse.org/**, để dễ nhớ bằng minh họa, chúng ta có thể dùng bảng CheetSheet 
+# https://raw.githubusercontent.com/rstudio/cheatsheets/master/pngs/thumbnails/data-transformation-cheatsheet-thumbs.png
+
+#https://www.rstudio.com/resources/cheatsheets/
+
+
+# >>>>>>>>>       glimse() sử dụng hàm glimse để xem tổng hợp vắn tắt về dữ liệu
+glimpse(df)
+
+#>>>>>>>>>>       select()
+# chọn các cột Price, Age, KM
+
+#cách 1
+df %>%
+  select(Price, Age, KM,FuelType,Automatic)
+#cách 2
+df %>%
+  select(c(1,2,3,4,7))
+#cách 3
+df %>%
+  select(c(1:4,7))
+
+# cách 4
+
+# không chọn một hay nhiều biến trong toàn bộ các biến
+
+df %>%
+  select(-c(5,6,8,9,10))
+
+# không chọn một hay nhiều biến trong toàn bộ các biến
+
+# thay đổi tên của biến (tên của cột)
+
+
+rename(df, gia=Price)
+
+
+df01 %>%
+  rename(item = CAPEXassets)
+
+# or
+rename(df01, item = CAPEXassets)
+
+
+df01 %>%
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/2ndcars/main.R

@@ -1,0 +1,117 @@
+## first we read in the data
+toyota <- read.csv("https://www.biz.uiowa.edu/faculty/jledolter/DataMining/ToyotaCorolla.csv")
+#toyota <- read.csv("ToyotaCorolla.csv")
+toyota[1:3,]
+summary(toyota)
+hist(toyota$Price)
+## next we create indicator variables for the categorical variable
+## FuelType with its three nominal outcomes: CNG, Diesel, and Petrol
+v1=rep(1,length(toyota$FuelType))
+v2=rep(0,length(toyota$FuelType))
+toyota$FuelType1=ifelse(toyota$FuelType=="CNG",v1,v2) #return value of CNG to v1, otherwise 0
+toyota$FuelType2=ifelse(toyota$FuelType=="Diesel",v1,v2) #return value of Diesel to v1, otherwise 0
+auto=toyota[-4] #ignore column 4 (Fueltype)
+auto[1:3,]
+
+plot.new()
+#par(mfrow=c(1,1))
+par(mar=c(4,4,1,1)+0.1,mfrow=c(4,2),bg="white",cex = 0.7, cex.main = 1)
+
+plot(Price~Age,data=auto)
+plot(Price~KM,data=auto)
+plot(Price~HP,data=auto)
+plot(Price~MetColor,data=auto)
+plot(Price~Automatic,data=auto)
+plot(Price~CC,data=auto)
+plot(Price~Doors,data=auto)
+plot(Price~Weight,data=auto)
+
+dev.copy(png,'toyota_xyplot.png',width = 500, height = 800)
+dev.off()
+
+
+## regression on all data
+m1=lm(Price~.,data=auto)
+summary(m1)
+
+set.seed(1)
+## fixing the seed value for the random selection guarantees the 
+## same results in repeated runs
+n=length(auto$Price)
+n1=1000
+n2=n-n1
+train=sample(1:n,n1) # generate random n1 integer number of data among the size from 1 to n.
+
+## regression on training set
+m1=lm(Price~.,data=auto[train,])
+summary(m1)
+pred=predict(m1,newdat=auto[-train,]) #adding a set of n2 number into regression model
+obs=auto$Price[-train]
+diff=obs-pred
+percdiff=abs(diff)/obs
+me=mean(diff)
+rmse=sqrt(sum(diff**2)/n2)
+mape=100*(mean(percdiff))
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/EVM/main.R

@@ -1,0 +1,83 @@
+#Coded by Nam Le (namlt@prontonmail.ch)
+library(scales)
+library(reshape2)
+library(ggplot2)
+
+
+
+#----------DATA-----------------
+data <- read.csv(file="Scurve.csv", header=TRUE, sep=",")
+k=data.frame(data)
+
+# -----------------------------This needs to be modified whenever you plot
+
+plot.new()
+par(mar=c(4, 4, 4, 4), bg="ivory")
+week= length(data[!is.na(data[,3]),3]) #the lastest week that EV value is available. Or you can simply put the number of week in instead of using length(data[!is.na(data[,3]),3]). Here 3 is the column.
+scalefactor=100
+trucy=100
+ticktime= length(data[!is.na(data[,1]),1])   #Set the Project Week at which you want to limit the X-axis
+varrange=c(-20,50)
+
+#Posion of texbox (PV, EV, and VAR in the graph)
+htextbox=c(week-5,week)
+vtextbox=20
+
+# --------------------------------
+
+
+# ----------------------------
+#Plot the baseline
+
+plot(data$BS01[1:ticktime],pch=4,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="blue",type="o",lwd=1,lty=1, main =paste("S-curve of week", week))
+axis(2, ylim=c(0,trucy),col="darkblue",las=1)
+mtext(expression(paste("Cummulative percentage (%)")),side=2,line=2.2, adj = 0.5 )
+axis(1,pretty(range(data$Week),40))
+mtext(expression(paste("Project Week")),side=1,col="black",line=2.2)  
+box()
+
+#plot the EV curve
+par(new=TRUE)
+plot(data$EV[1:ticktime],pch=18,axes=FALSE,ylim=c(0,trucy),ylab="", xlab="",col="red",type="o",lwd=1,lty=2)
+
+#adding stick at the actual week
+abline(v=week, col="darkviolet",lty=5)
+
+par(new=TRUE)
+#library(gplots)
+library(astro) #this package is for astronomy but it has some fantastic function for plotting. Here I use the textbox function
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/bio.R

@@ -1,0 +1,36 @@
+library(ggplot2)
+library(plotly)
+# read excel file
+library(readxl) 
+library(reshape)
+library(reshape2)
+df=read_excel("../data/introduction.xlsx",sheet="bio-time-vn",skip = 0)
+df<-melt(df,id="activity")
+# Grouped
+p<-ggplot(df, aes(fill=activity, y=value, x=variable)) + 
+  geom_bar(position="dodge", stat="identity")+
+  labs(title = "Số Giờ Trung Bình Hàng Tuần cho chuyện ấy", x = "Thời Kì",y = "Số Giờ",fill="Chuyện ấy")+
+  theme(axis.text.x=element_text(angle=0, hjust=0.5))+ 
+  geom_text(aes(label=value), position=position_dodge(width=0.9), vjust=-0.25)
+
+q<-ggplot(df, aes(fill=variable, y=value, x=activity)) + 
+  geom_bar(position="dodge", stat="identity")+
+  labs(title = "Số Giờ Trung Bình Hàng Tuần cho chuyện ấy", x = "Chuyện ấy",y = "Số Giờ",fill="Thời Kì")+
+  theme(axis.text.x=element_text(angle=25, hjust=0.5))+ 
+  geom_text(aes(label=value), position=position_dodge(width=0.9), vjust=-0.25)
+
+
+
+w<-ggplot(df, aes(fill=variable, y=value, x=reorder(activity,value))) + 
+  geom_bar(position="dodge", stat="identity")+
+  coord_flip() +
+  labs(title = "Số Giờ Trung Bình Hàng Tuần cho chuyện ấy", x = "Chuyện ấy",y = "Số Giờ",fill="Thời Kì")+
+  theme(axis.text.x=element_text(angle=0, hjust=0.5))
+ # geom_text(aes(label=value), position=position_dodge(width=1), vjust=0.25)
+
+
+
+
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/cleaning.R

@@ -1,0 +1,7 @@
+# Làm sạch dữ liệu, sử dụng thư viện Janitor để viết lại tên cho các cột. Đồng thời đặt lại ngày và tháng theo định dạng chuẩn thông qua việc sử dụng hàm ymd trong gói lubricate
+
+thedata <- janitor::clean_names(thedata)
+thedata <- thedata %>%
+  mutate(ref_date = ymd(ref_date, truncated = 2)) 
+head(thedata)
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/f_monthyear.R

@@ -1,0 +1,18 @@
+
+#phương trình tính % thay đổi theo tháng
+get_mom_stats <- function(df) {
+  df %>%
+    arrange(ref_date) %>%
+    mutate(mom_val = lag(value), 
+           mom_pct = ((value / lag(value, n = 1)) - 1) * 100,
+           mom_chg = (value - lag(value, n = 1)))
+}
+
+#phương trình tính % thay đổi theo năm
+get_yoy_stats <- function(df) {
+  df %>%
+    arrange(ref_date) %>%
+    mutate(yoy_val = lag(value, n = 12),
+           yoy_pct = ((value / lag(value, n = 12)) - 1) * 100,
+           yoy_chg = (value - lag(value, n = 12)))
+}

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/filtering.R

@@ -1,0 +1,11 @@
+# Chỉ chọn ra 2 vùng là BC - Bristish Columbia và Canada
+
+startdate <- as.Date("2007-01-01") #vì số liệu từ năm 1981 nên nó quá nhiều, nên chọn số liệu từ ngăm 2007 cho tới nay, lưu ý đây là số liệu được cập nhật theo tháng và được xuất bản định kì trên trang web của chính phủ canada.
+
+# Chọn BC và Canada, đồng thời chỉ chọn riêng 4 cột cần quan tâm.
+thedata_BC_Can <- thedata %>%
+  filter(ref_date >= startdate) %>%
+  filter(geo %in% c("British Columbia", "Canada"), 
+         new_housing_price_indexes == "Total (house and land)") %>%
+  select(ref_date, geo, new_housing_price_indexes, value)
+head(thedata_BC_Can)

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/import.R

@@ -1,0 +1,13 @@
+#Nhập Dữ Liệu
+
+table_id <- "18-10-0205-01" # --> chọn bảng --> xem trên trang web https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1810020501
+thedata <- get_cansim(table_id) #sử dụng hàm get_cansim để chọn bảng
+get_cansim_table_overview(table_id)
+ls.str(thedata) #xem dữ liệu
+
+#xem tổng hợp dữ liêu
+thedata %>%
+  group_by(GEO, GeoUID) %>%
+  tally()
+
+tail(thedata) #xem đuôi của dữ liệu

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/main.R

@@ -1,0 +1,45 @@
+#Load vào RAM máy tính các thư viện cần thiết
+library(tidyverse)
+library(lubridate)  # chứa các chương trình xử lý về ngày tháng
+library(scales)     # bổ trợ cho thư viện  {ggplot2}
+library(glue)       # dùng để gắn kết các dữ liệu string với nhau
+#
+# utilities
+library(cansim)     # đây là thư viện chứa gói dữ liệu về giá nhà ở Canada
+library(janitor)    # dùng để làm sạch tên cột
+library(knitr)      # dùng để xuất bản báo cáo, có chứa các hàm để hiển thị bảng như `kable()`
+library(kableExtra) # - dùng để định dạng và trang trí bảng 
+library(flextable)  # dùng để định dạng và trang trí bảng
+
+
+#Tải vào các phương trình định sẵn liên quan đến chuyển đổi ngày tháng và vẽ đồ thị theo 1 định dạng chuẩn.
+
+source("f_monthyear.R")
+source("bida_chart_theme.R")
+# CHÚNG TA BẮT ĐẦU TỪ ĐÂY
+
+source("import.R")            # BƯỚC 1  - Nhập Dữ Liệu
+source("cleaning.R")          # BƯỚC 2  - Làm Sạch 
+source("filtering.R")         # BƯỚC 3  - Sàng Lọc
+source("table_visualize.R")   # BƯỚC 4  - Hiển THị Bảng
+source("table_communicate.R") # BƯỚC 5  - Thông Tin Bảng
+source("text.R")              # BƯỚC 6  - Chữ
+source("visualize.R")         # BƯỚC 7  - Trực Quan Hóa Bằng Đồ Thị
+
+
+
+
+
+
+
+
+
+
+#text
+
+
+
+
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/table_communicate.R

@@ -1,0 +1,46 @@
+#Truyền đạt thông tin từ các bảng
+#trong môi trường R, có rất nhiều cách để trình bày bảng, có thể xuất ra các định dạng khác nhau và có thể trang điểm cho các bảng một cách rất chuyên nghiệp.
+
+
+# Thể hiện  table với định dạng {flextable}   
+NHPI_table4 %>%
+  flextable() %>% 
+  #  add_footer(glue::glue("Source: Statistics Canada, Table {table_id}"))
+  add_footer_row(values = glue::glue("Source: Statistics Canada, Table {table_id}"),
+                 colwidths = 14)
+
+print(NHPI_table4)
+
+#table: monthly summary
+
+# Tạo ra bảng tổng hợp
+tbl_month_BC <-
+  thedata_BC_Can %>%
+  filter(geo == "British Columbia") %>%
+  arrange(ref_date) %>%
+  # calculate percent change stats --> tính toán phần trăm thay đổi theo tháng
+  get_mom_stats() %>%
+  get_yoy_stats() %>%
+  # pull year and month
+  mutate(year = year(ref_date),
+         month = month(ref_date, label = TRUE)) %>%
+  # select relevant columns, rename as necessary
+  select(year, month, value, 
+         "from previous month" = mom_chg, 
+         "from same month, previous year" = yoy_chg) %>%
+  arrange(desc(year), desc(month)) %>%
+  # just print rows 1 to 13 --> chỉ in ra từ hạng 1 đến 13
+  slice(1:13)
+
+print(tbl_month_BC)
+
+# Trình bày bảng với {kableExtra} formatting  
+tbl_month_BC %>%
+  kable(caption = "NHPI, British Columbia", digits = 1) %>%
+  kable_styling(bootstrap_options = "striped") %>%
+  row_spec(0, bold = T, font_size = 14) %>%
+  row_spec(1, bold = T) %>%
+  add_header_above(c(" " = 3, "index point change" = 2), font_size = 14)
+
+print(tbl_month_BC)
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/table_visualize.R

@@ -1,0 +1,60 @@
+#Tìm hiểu về dữ liệu
+## Thiết lập các bảng pivot
+
+
+#chỉ chọn ra vùng BC sau đó thêm cột năm (year), cột tháng,sau đó chọn ra 3 cột là năm, tháng và giá trị.
+
+#sử dụng pivot_wider để coi theo tháng
+NHPI_table <- thedata_BC_Can %>%
+  filter(geo == "British Columbia") %>%
+  mutate(year = year(ref_date),
+         month = month(ref_date, label = TRUE)) %>%
+  select(year, month, value)%>%
+  pivot_wider(names_from = month, values_from = value) %>%
+  arrange(desc(year))
+
+# display the table
+print(head(NHPI_table))
+
+
+
+# Làm thế nào để thêm vào tổng hợp theo năm
+# Julie's genius solution --> giải pháp của Julie, là sử dụng chính bảng pivot lúc trước và tính giá trị trung bình các cột, trừ cột thứ 1 (là cột năm)
+NHPI_table2 <- NHPI_table %>%
+  mutate(annual_avg = rowMeans(NHPI_table[-1], na.rm = TRUE))
+print(head(NHPI_table2))
+
+
+
+# Stephanie's genius solution --> Cách của Stephanie
+# starts with the raw data table --> Sử dụng luôn dữ liệu ban đầu chứ không phải từ bảng pivot
+
+NHPI_table3 <- thedata_BC_Can %>%
+  filter(geo == "British Columbia") %>%
+  mutate(year = year(ref_date),
+         month = month(ref_date, label = TRUE)) %>%
+  select(year, month, value) %>%
+  group_by(year) %>%  # đến đây các bạn có thể thấy là Stephanie đã sử dụng hàm group_by
+  mutate(annual_avg = mean(value, na.rm = TRUE)) %>%  # sau đó cô ấy thêm 1 cột và tính trung bình của cột value
+  pivot_wider(names_from = month, values_from = value) %>%
+  arrange(desc(year))
+
+print(head(NHPI_table3))
+
+# hoặc là tính thủ công như thế này
+NHPI_table4 <- NHPI_table %>%
+  rowwise() %>% 
+  mutate(annual_avg = mean(c(Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec), na.rm = TRUE))
+
+print(head(NHPI_table4))
+
+
+#---------------
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/text.R

@@ -1,0 +1,28 @@
+#lấy các giá trị tháng gần nhất của vùng B.C 
+
+# lọc 
+
+thedata_BC <- thedata_BC_Can %>%
+  filter(geo == "British Columbia") %>%
+  arrange(ref_date) %>%
+  # calculate percent change stats --> Tính toán % thay đổi theo tháng, năm
+  get_mom_stats() %>%
+  get_yoy_stats()
+
+# determine most recent month --> Xác định tháng gần nhất
+
+latest_ref_date <- max(thedata_BC$ref_date)
+
+this_month <- month(latest_ref_date, label = TRUE)
+this_year <- year(latest_ref_date)
+
+this_month_nhpi <- thedata_BC %>%
+  filter(ref_date == latest_ref_date) %>%
+  pull(value)
+this_month_mom <- thedata_BC %>%
+  filter(ref_date == latest_ref_date) %>%
+  pull(mom_chg)
+this_month_yoy <- thedata_BC %>%
+  filter(ref_date == latest_ref_date) %>%
+  pull(yoy_chg)
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/canadahousing/visualize.R

@@ -1,0 +1,62 @@
+# Vẽ Đồ Thị!
+# Cơ Bản
+ggplot(thedata_BC_Can, aes(x=ref_date, y=value, group=geo)) + 
+  geom_line()
+# các bạn đã thấy chưa, R cung cấp cho ta 1 dạng đồ thị cơ bản và cũng khá là bắt mắt. Và chúng ta có thể thêm thắt và làm đẹp cho nó vì khi nhìn vào đồ thị này, chúng ta để ý là có 2 đường thể hiện, 1 đường cho Canada và 1 cho B.C, nhưng chúng ta không biết đường nào thuộc về vùng nào.
+
+#
+# trang điểm
+dataplot <- ggplot(thedata_BC_Can, aes(x=ref_date, y=value, colour=geo)) + 
+  geom_line(size=1.5) 
+
+plot(dataplot) 
+
+
+dataplot2 <- dataplot +
+  scale_x_date(date_breaks = "2 years", labels = year) +
+  scale_y_continuous(labels = comma, limits = c(80, 125)) +
+  scale_colour_manual(name=NULL,
+                      breaks=c("Canada", "British Columbia"),
+                      labels=c("Canada", "British Columbia"), 
+                      values=c("#325A80", "#CCB550")) +
+  bida_chart_theme
+#  theme_bw() +
+#  theme(
+#    panel.border = element_rect(colour="white"),
+#    plot.title = element_text(face="bold"),
+#    legend.position=c(1,0), 
+#    legend.justification=c(1,0),
+#    legend.title = element_text(size=12),
+#    legend.text = element_text(size=11),
+#    axis.line = element_line(colour="black"),
+#    axis.title = element_text(size=12),
+#    axis.text = element_text(size=12)
+#  )
+#
+plot(dataplot2)
+
+
+# experiments with ggplot2's new subtitle and caption options
+NHPI_title <- as.character("New Housing Price Index, Canada & B.C.")
+NHPI_subtitle <- as.character("December 2016 = 100")
+NHPI_caption <- as.character("Source: Statistics Canada, CANSIM table 18-10-0205-01")
+# add titles / X-Y axis labels
+NHPI_plot <- dataplot2 +
+  labs(title = NHPI_title,
+       subtitle = NHPI_subtitle,
+       caption = NHPI_caption, 
+       x = NULL, y = "NHPI (Dec. 2016 = 100)")
+
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/coronavirus-vietnam.R

@@ -1,0 +1,3 @@
+# https://data.opendevelopmentmekong.net/vi/dataset/coronavirus-covid-19-cases-in-vietnam
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/coronavirus.R

@@ -1,0 +1,95 @@
+library(dplyr)
+library(tidyr)
+library(plotly)
+library(coronavirus)
+
+
+#data source
+# -----https://covid19.who.int/info?openIndex=2
+#----- https://github.com/RamiKrispin/coronavirus/tree/master/csv
+
+
+#df <- coronavirus
+df <- read.csv("../data/coronavirus/coronavirus.csv", header = TRUE, sep = ",")
+
+data("coronavirus")
+
+total_cases <- df %>% 
+  group_by(type) %>%
+  summarise(cases = sum(cases)) %>%
+  mutate(type = factor(type, levels = c("confirmed", "death", "recovery")))
+
+total_cases
+
+head(df)
+
+
+summary_df <- df %>% 
+  filter(type == "confirmed") %>%
+  group_by(country) %>%
+  summarise(total_cases = sum(cases)) %>%
+  arrange(-total_cases)
+
+
+summary_df %>% head(20) 
+
+df %>% 
+  filter(date == max(date)) %>%
+  select(country, type, cases) %>%
+  group_by(country, type) %>%
+  summarise(total_cases = sum(cases)) %>%
+  pivot_wider(names_from = type,
+              values_from = total_cases) %>%
+  arrange(-confirmed)
+
+
+df %>% 
+  group_by(type, date) %>%
+  summarise(total_cases = sum(cases)) %>%
+  pivot_wider(names_from = type, values_from = total_cases) %>%
+  arrange(date) %>%
+  mutate(active = confirmed - death - recovery) %>%
+  mutate(active_total = cumsum(active),
+         recovered_total = cumsum(recovery),
+         death_total = cumsum(death)) %>%
+  plot_ly(x = ~ date,
+          y = ~ active_total,
+          name = 'Active', 
+          fillcolor = '#1f77b4',
+          type = 'scatter',
+          mode = 'none', 
+          stackgroup = 'one') %>%
+  add_trace(y = ~ death_total, 
+            name = "Death",
+            fillcolor = '#E41317') %>%
+  add_trace(y = ~recovered_total, 
+            name = 'Recovered', 
+            fillcolor = 'forestgreen') %>%
+  layout(title = "Distribution of Covid19 Cases Worldwide",
+         legend = list(x = 0.1, y = 0.9),
+         yaxis = list(title = "Number of Cases"),
+         xaxis = list(title = "Source: Johns Hopkins University Center for Systems Science and Engineering"))
+
+
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/hocvien.R

@@ -1,0 +1,56 @@
+library(ggplot2)
+library(plotly)
+# read excel file
+library(readxl) 
+library(reshape)
+library(reshape2)
+df=read_excel("../data/hocvien.xlsx",sheet="Sheet1",skip = 0)
+#df<-melt(df,id="activity")
+# Grouped
+df<-data.frame(df)
+
+colnames(df) <- c("Stt","hoten","ngaysinh","gioitinh","noisinh","truong","nganh","email","sdt","chucvu","donvi")
+
+
+df$ngaysinh<-as.Date(parse_date_time(df$ngaysinh,c('dmy')),format = "%d/%m/%y")
+
+format(as.Date("27/5/1987", format="%d/%m/%Y"),"%Y")
+
+year(df$ngaysinh)
+
+df <- df %>%
+  mutate(namsinh = year(ngaysinh))
+
+
+
+
+p1<-ggplot(df, aes(x = (2023-namsinh), y = gioitinh, fill = namsinh)) +
+  geom_boxplot() +
+  theme(legend.position = "top")
+
+p2<-ggplot(df, aes(x = (2023-namsinh))) +
+  geom_histogram()+
+  theme(legend.position = "top")
+
+library(hrbrthemes)
+p3 <- df %>%
+  ggplot( aes(x=(2023-namsinh))) +
+  geom_histogram( binwidth=3, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
+  labs(
+    x ="Độ Tuổi",
+    y ="Số Lượng"
+  )
+  theme_ipsum()
+
+
+a=df %>%
+  group_by(truong) %>%
+  tally()
+
+b=df %>%
+  group_by(nganh) %>%
+  tally()
+
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/monitor-selection.r

@@ -1,0 +1,52 @@
+# Selecting a Monitor for our Class.
+library(purrr)
+# reading data file
+library(readxl) 
+library(dplyr)
+df=read_excel("../Data/students.xlsx",sheet="student",skip = 0)
+
+l <- length(df$Stt) #Data Frame (df)
+
+df1 <- replicate(sample(1:l, 1, replace = FALSE), n = 10000)
+df1 <- data.frame(df1)
+df2 <- data.frame(table(df1))
+which.max(df2$Freq)
+
+
+df4 <- df %>%
+  select(c(1,2,3,5,7,13,14)) %>%
+  mutate(freq = df2$Freq) 
+
+names(df4) <- c('ID', 'lastname', 'firstname', 'age', 'sex', 'field', 'fullname', 'freq')
+
+df4 <- data.frame(df4)
+
+library(ggChernoff)
+library(ggplot2)
+
+ggplot(df4) +
+  aes(age, freq, fill = field) +
+  geom_chernoff()+
+  labs(y = "Số phiếu bầu", x ="Độ Tuổi")+ 
+  geom_text(aes(label=paste(lastname,firstname)), position=position_dodge(width=1), vjust=2, cex =2.5)
+
+
+ggplot(df4) +
+  aes(age, freq, fill = sex) +
+  geom_chernoff()+
+  labs(y = "Số phiếu bầu", x ="Độ Tuổi")+ 
+  geom_text(aes(label=paste(lastname,firstname)), position=position_dodge(width=1), vjust=2, cex =2.5)
+
+summary(df4$age)
+
+cat(" Chúc mừng bạn", paste(df$'Họ'[which.max(df2$Freq)]), paste(df$'Tên'[which.max(df2$Freq)]))
+
+
+
+p.height <- c(180, 155, "aaa", 167, 181)
+p.weight <- c(65, 50, 52, 58, 70)
+p.names <- c("Joanna", "Charlotte", "Helen", "Karen", "Amy")
+
+dataf <- data.frame(height = p.height, weight = p.weight, names = p.names)
+dataf
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/outliers-removal.r

@@ -1,0 +1,26 @@
+# Removing outliers method 1
+
+data("warpbreaks")
+boxplot(warpbreaks)$out
+
+library(ggstatsplot)
+ggbetweenstats(warpbreaks,
+               wool, breaks, outlier.tagging = TRUE)
+
+Q <- quantile(warpbreaks$breaks, probs=c(.25, .75), na.rm = FALSE)
+iqr <- IQR(warpbreaks$breaks)
+
+up <-  Q[2]+1.5*iqr # Upper Range
+low<- Q[1]-1.5*iqr # Lower Range
+eliminated<- subset(warpbreaks, warpbreaks$breaks > (Q[1] - 1.5*iqr) & warpbreaks$breaks < (Q[2]+1.5*iqr))
+
+ggbetweenstats(eliminated, wool, breaks, outlier.tagging = TRUE)
+
+
+# removing outliner method 2
+
+data("warpbreaks")
+boxplot(warpbreaks$breaks, plot=FALSE)$out
+outliers <- boxplot(warpbreaks$breaks, plot=FALSE)$out
+x<-warpbreaks
+x<- x[-which(x$breaks %in% outliers),]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/pdfmanipulation.r

@@ -1,0 +1,10 @@
+library(pdftools)
+#pdf_combine(c(
+#"1.pdf",
+#"2.pdf"
+#), 
+#output = "output.pdf")
+
+#extract some pages
+#pdf_subset('1.pdf',pages = 1:1, output = "output.pdf")
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/pew-tidying.R

@@ -1,0 +1,7 @@
+library(dplyr)
+library(reshape2)
+df=read.csv("../Data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
+df <-data.frame(df)
+molten.df <- melt(df, id = c("X","religion"))
+colnames(molten.df)<- c('no','religion','income','freq')
+molten.df <- molten.df%>%select(c(2,3,4))

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/r-introduction.r

@@ -1,0 +1,340 @@
+# A brief to R
+
+# Working directory
+getwd()
+
+# Getting Start
+
+# Print out Hello World
+
+a <- "Hello World!"
+print("Hello World")
+print(a)
+cat(a)
+b <- "Welcome to ISCM"
+print("Welcome to ISCM")
+print(b)
+cat(b)
+cat(a, b)
+print(paste(a, b))
+
+a=2
+b=3
+c = a+b
+c
+
+
+library(tidyverse)
+library(dplyr)
+library(reshape)
+library(reshape2)
+library(readxl)
+library(xlsx)
+
+
+
+
+
+
+
+
+# Package in R
+#install package (using CRAN, or using Dev.Tool)
+#---------->install.packages()
+#loading package
+#---------->library()
+
+# let play some funs with Emoji package
+#https://cran.r-project.org/web/packages/emoji/index.html
+
+library(emoji)
+
+arrow("up-down")
+
+arrow(c("up", "up", "down", "down", "left", "right", "left", "right"))
+
+
+times <- as.POSIXct("2021-09-17 14:33:21 PDT") + seq(1:30) * 3500
+clock(times)
+
+string <- paste(c(letters[1:4], emoji_name[1:6]), collapse = " ")
+emoji_count(string)
+emoji_count(emoji_name[1:6])
+
+#https://github.com/hadley/emo
+
+run = "running"
+replicate(3,emo::ji(run))
+
+# Cong Tru Nhan Chia
+
+library(car)
+
+
+
+
+# R is a calculator
+3+4
+
+3-4
+
+3*4
+
+3/4
+
+# Rounding
+
+1/6
+
+round(1/6,2)
+n = 1:5
+round(1/6,n)
+
+#power and log
+
+3**4
+3^4
+
+2^3
+
+2**3
+
+2*3
+
+log(3)
+
+d <- log(3)
+
+exp(3)
+
+e = exp(3)
+
+f = d+e +pi
+
+
+
+
+pi
+
+
+
+
+
+# --> there are some predetermine function name in R and we should avoid to use them.
+
+
+# Object in R --> R is a object oriented program.
+
+object1 <- 3
+object2 <-4
+
+x <- object1+object2
+x
+
+
+# Objects can be numeric or character
+
+
+char_obj1 <- "Chao"
+char_obj2 <- "ISCM!"
+
+char_obj3 <- char_obj1 + char_obj2
+
+
+
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/schedule.R

@@ -1,0 +1,16 @@
+library(DT)
+library(tidyverse)
+library(readxl) 
+library(lubridate)
+#df=read_excel("../data/introduction.xlsx",sheet="course-program",skip = 0)
+df=read_excel("../Data/introduction.xlsx",sheet="course-program",skip = 0)
+df$Ngày <- as.Date(df$Ngày,format = "%d/%m/%y")
+p <- datatable(df,rownames = FALSE, filter = 'top',
+          options = list(pageLength = 10,
+            columnDefs = list(list(className = 'dt-left', targets = c(1,2,4)),list(className = 'dt-center', targets = c(0,3,5)))
+            ,autoWidth = TRUE)
+          )
+
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/student-knowledge-efficiency.R

@@ -1,0 +1,15 @@
+library(ggplot2)
+N<-10000 # Total number of of students in the class
+R <- seq(1:10) # Level of efficiency
+
+
+BE <-sample(c(1:6),size=N,replace = TRUE)
+AF <-sample(c(7:9),size=N,replace = TRUE)
+
+df <- data.frame(BE,AF)
+
+p <- ggplot(df, aes(x=BE)) + 
+  geom_bar()
+p
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/test.r

@@ -1,0 +1,9 @@
+2+2
+
+
+my_mat <- matrix(1:16, nrow = 4, byrow = TRUE)
+my_mat
+
+my_array <- array(1:16, dim = c(2, 4, 2))
+my_array
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/tidydata.r

@@ -1,0 +1,8 @@
+library(readxl)
+library(dplyr)
+# library(ggplot2)
+df1=read_excel("../data/tidydata.xlsx",sheet="treatment01")
+
+df2=read_excel("../data/tidydata.xlsx",sheet="treatment02")
+
+df3=read_excel("../data/tidydata.xlsx",sheet="treatment03")

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/venndiagram.R

@@ -1,0 +1,10 @@
+library(ggVennDiagram)
+library(ggplot2)
+x <- list(A = 1:5, B = 2:7, C = 5:10)
+
+
+v<-ggVennDiagram(x,category.names = c("Database Management","Statistics","Modelling"), label = "none")+ 
+  theme(legend.position = "none")+
+  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")
+
+v

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/visualization/ggplot2-esquisse.r

@@ -1,0 +1,3 @@
+library(ggplot2)
+library(esquisse)
+esquisse::esquisser()

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/AlumniContributions/main.R

@@ -1,0 +1,220 @@
+library(lattice)
+library(ggplot2)
+
+#Using multiple plot function when using with ggplot
+#source("https://raw.githubusercontent.com/namkyodai/BusinessAnalytics/master/genericfunctions/multiplot.R")
+
+
+#----1. Data
+don <- read.csv("contribution.csv")
+#or read directly from the web
+#don <- read.csv("https://www.biz.uiowa.edu/faculty/jledolter/DataMining/contribution.csv")
+#or
+
+don[1:5,] #display the first 5 data rows
+
+table(don$Class.Year) #display total numbers of data points for each batch of year
+a=barchart(table(don$Class.Year),horizontal=FALSE,xlab="Class Year",col="black")
+p=ggplot(data.frame(table(don$Class.Year)), aes(x=Var1, y=Freq))+labs(y="Freq", x="Class Year") + geom_bar(stat="identity",width=0.8,color="blue",fill="steelblue")+geom_text(aes(label=Freq), vjust=-0.3, size=3.5)
+plot.new()
+par(mar=c(4.5,4.3,1,1)+0.1,mfrow=c(1,2),bg="white")
+library(gridExtra) #this package allows to plot multiple graphs in the same plot despite the difference in plotting engines (e.g. ggplot or barchart)
+grid.arrange(a, p, ncol = 2) #display the two plot a and p
+
+dev.copy(png,'alumni_classyear_bar.png',width = 1200, height = 500)
+dev.off()
+
+
+don$TGiving=don$FY00Giving+don$FY01Giving+don$FY02Giving+don$FY03Giving+don$FY04Giving #total giving
+mean(don$TGiving)
+
+sd(don$TGiving)
+quantile(don$TGiving,probs=seq(0,1,0.05)) #quantile from 0 to 1 with a step of 5%
+quantile(don$TGiving,probs=seq(0.95,1,0.01)) #quantilte from 0.95 to 1 with a step of 1%
+
+
+#---------------------
+plot.new()
+par(mar=c(4.5,4.3,1,1)+0.1,mfrow=c(2,2))
+hist(don$TGiving,main=NULL,xlab="Total Contribution") #histograph with outliners
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/Births/birth_statistic01.R

@@ -1,0 +1,240 @@
+#Graphs are saved in GITHUB
+#https://raw.githubusercontent.com/namkyodai/BusinessAnalytics/master/Births/XXX.png
+
+library(lattice)
+# library(nutshell.bbdb)
+# library(nutshell.audioscrobbler)
+# library(nutshell)
+#data(births2006.smpl)
+#write.csv(births2006.smpl, file = "births2006.csv")
+
+df <-read.csv("births2006.csv",header = TRUE, sep = ",")
+
+df <- data.frame(df)
+
+
+#https://rdrr.io/cran/nutshell/man/births2006.smpl.html
+#R In a Nutshell: A Desktop Quick Reference (Adler, 2009).
+
+##
+
+# DOB_MM
+# Month of date of birth/ Tháng sinh
+#
+# DOB_WK
+# Day of week of birth/Ngày trong tuần sinh
+#
+# MAGER
+# Mother's age / tuổi của mẹ
+#
+# TBO_REC
+# Total birth order / Thứ tự sinh
+#
+# WTGAIN
+# Weight gain by mother / Cân nặng của Mẹ
+#
+# SEX
+# a factor with levels F M, representing the sex of the child / Giới tính của trẻ
+#
+# APGAR5
+# APGAR score / Chỉ số APGAR, đây là chỉ số kiểm tra nhanh trong vòng 1 đến 5 phút khi trẻ chào đời. CHỉ số càng cao thì càng tốts
+#https://vi.wikipedia.org/wiki/Ch%E1%BB%89_s%E1%BB%91_Apgar
+
+#
+# DMEDUC
+# Mother's education level / trình độ học vấn phụ sản
+
+# UPREVIS
+# Number of prenatal visits / số buổi phụ sản đi gặp bác sĩ để tư vấn khám trước khi sinh
+#
+# ESTGEST
+# Estimated weeks of gestation / ước tính số tuần thai kỳ
+#
+# DMETH_REC
+# Delivery Method / phương pháp sinh
+#
+# DPLURAL
+# "Plural Births;" levels include 1 Single, 2 Twin, 3 Triplet, 4 Quadruplet, and 5 Quintuplet or higher / số tẻ được sinh ra (đơn, sinh đôi, sinh ba)
+#
+# DBWT
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/Births/birth_statistic02.r

@@ -1,0 +1,164 @@
+#Graphs are saved in GITHUB
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+df <-read.csv("births2006.csv",header = TRUE, sep = ",")
+df <-data.frame(df)
+
+glimpse(df)
+
+df %>%
+  select(DOB_WK,DMETH_REC,WTGAIN,DBWT)
+
+df %>%
+  filter(DMETH_REC =="Vaginal"|DMETH_REC =="C-section")
+
+
+df %>%
+  filter(DMETH_REC %in% c("Vaginal","C-section"))
+
+df %>%
+  arrange(desc(DBWT))
+
+df %>%
+  mutate(cotmoi01 = DBWT*DOB_WK,cotmoi02 = "abc" )
+
+
+df %>%
+  select(DOB_WK,DMETH_REC,WTGAIN,DBWT) %>%
+  filter(DMETH_REC == "Vaginal") %>%
+  filter(DOB_WK %in% c(2,3,4,5,6)) %>%
+  filter(DOB_WK >= 3)
+  
+
+df %>%
+  select(DOB_WK,DMETH_REC,WTGAIN,DBWT) %>%
+  filter(DMETH_REC == "Vaginal", DOB_WK %in% c(2,3,4,5,6)) %>%
+  arrange(desc(DBWT))
+
+
+
+
+#select()
+#filter()
+#arrange()
+#mutate() 
+
+
+
+
+
+
+
+# So luong tre so sinh theo ngay trong tuan
+df.dow <- df %>%
+  count(DOB_WK, sort = FALSE)
+
+# ve bieu do barchart
+# https://www.r-graph-gallery.com/barplot.html  --> 
+# http://www.sthda.com/english/wiki/ggplot2-axis-scales-and-transformations  --> tỷ lệ hiển thị ở các trục
+#https://ggplot2.tidyverse.org/reference/labs.html  --> Đặt tên các biến trên đồ thị
+ggplot(df.dow, aes(x=DOB_WK, y=n)) + 
+  geom_bar(stat = "identity",color='red',fill='darkolivegreen1') +
+  labs(
+    title = "Số lượng trẻ sinh ra theo ngày0 trong tuần",
+    y = "Số trẻ sơ sinh",
+    x = "Ngày trong tuần"
+  )
+
+
+ggplot(df %>%
+         count(DOB_WK, sort = FALSE), 
+       aes(x=DOB_WK, y=n)) + 
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/Births/week2-assignment-solution.r

@@ -1,0 +1,321 @@
+#000000000000000000000000000000000000000000000000000000000
+
+
+# A. Sử dụng thư viện dplyr của R để trả lời các câu hỏi sau liên quan đến gói dữ liệu Births2006
+
+#000000000000000000000000000000000000000000000000000000000
+
+
+# 1. Có bao nhiêu trẻ sơ sinh là Nam, bao nhiêu là Nữ?
+# 2. Có bao nhiêu phụ sản mà chúng ta không biết trình độ học vấn?
+# 3. Có bao nhiêu phụ sản sinh ít hơn 3 bé sơ sinh và sử dụng phương pháp đẻ tự nhiên?
+# 4. Có bao nhiêu trẻ sinh ra có chỉ số APGAR5 nhỏ hơn 3 và có cân nặng lớn hơn 2.5 kg? Trong số trẻ này thì có bao nhiêu trẻ sinh bằng phương pháp tự nhiên, bao nhiêu sinh ra bằng phương pháp cắt.
+# 5. Tháng nào trong năm có số lượng trẻ sơn sinh nhiều nhất?
+# 6. Câu hỏi phụ (không nhất thiết phải trả lời), các bạn có thẻ vẽ biểu đồ để thể hiện dữ liệu cho 5 câu hỏi trên không? --> sử dụng biểu đồ barchart, histogram sử dụng phụ kiện ggplot2
+# 7. Tìm mối tương quan giữa cân nặng của trẻ sơ sinh với các biến còn lại? Các biến nào các bạn sẽ chọn để dùng phân tích? Các bạn có kết luận gì không sau khi chạy phân tích với hồi qui tuyến tính chuẩn?
+
+
+library(dplyr)
+
+df <-read.csv("births2006.csv",header = TRUE, sep = ",")
+df <- data.frame(df)
+
+
+glimpse(df)
+
+# 1. Có bao nhiêu trẻ sơ sinh là Nam, bao nhiêu là Nữ?
+
+summary(df %>% select(SEX))
+
+#cách 1
+df %>%
+  select(SEX) %>%
+  count(SEX)
+
+
+#cách 2
+
+df %>%
+  count(SEX)
+
+
+# 2. Có bao nhiêu phụ sản mà chúng ta không biết trình độ học vấn
+
+# cách 1
+df %>%
+  filter(DMEDUC == "NULL")%>%
+  count(DMEDUC)
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/FuelEfficiency/main (namlt's conflicted copy 2019-12-15).R

@@ -1,0 +1,100 @@
+## first we read in the data
+FuelEff <- read.csv("https://www.biz.uiowa.edu/faculty/jledolter/DataMining/FuelEfficiency.csv")
+#FuelEff <- read.csv("FuelEfficiency.csv")
+
+FuelEff
+
+#GPM-->Gallon per 100 miles
+#MPG--> miles per Gallons
+#WT --> weigt of the car
+#DIS --> Cubic displacement
+#NC --> No. of cylinders
+#HP --> Hourse power
+#ACC --> Acceleration
+# ET --> engine type
+
+#----------------------------
+plot.new()
+# par(mfrow=c(1,1))
+par(mar=c(4,4,1,1)+0.1,mfrow=c(3,3),bg="white",cex = 1, cex.main = 1)
+plot(GPM~MPG,data=FuelEff)
+plot(GPM~WT,data=FuelEff)
+plot(GPM~DIS,data=FuelEff)
+plot(GPM~NC,data=FuelEff)
+plot(GPM~HP,data=FuelEff)
+plot(GPM~ACC,data=FuelEff)
+plot(GPM~ET,data=FuelEff)
+
+dev.copy(png,'fueleff_xyplot.png',width = 800, height = 800)
+dev.off()
+
+
+FuelEff=FuelEff[-1] #ignore the MPG column
+FuelEff
+
+## regression on all data
+m1=lm(GPM~.,data=FuelEff)
+summary(m1)
+
+cor(FuelEff)
+
+## best subset regression in R
+library(leaps)
+X=FuelEff[,2:7]
+y=FuelEff[,1]
+
+#use the regsubsets function from package leaps to compute regression of the subsets
+#https://www.rdocumentation.org/packages/leaps/versions/2.1-1/topics/regsubsets
+
+
+out=summary(regsubsets(X,y,nbest=2,nvmax=ncol(X)))
+tab=cbind(out$which,out$rsq,out$adjr2,out$cp)
+tab
+
+m2=lm(GPM~WT,data=FuelEff)
+summary(m2)
+
+## cross-validation (leave one out) for the model on all six regressors
+n=length(FuelEff$GPM)
+diff=dim(n)
+percdiff=dim(n)
+for (k in 1:n) {
+  train1=c(1:n)
+  train=train1[train1!=k]
+  ## the R expression "train1[train1!=k]" picks from train1 those
+  ## elements that are different from k and stores those elements in the
+  ## object train.
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/FuelEfficiency/main.R

@@ -1,0 +1,119 @@
+## first we read in the data
+#FuelEff <- read.csv("https://www.biz.uiowa.edu/faculty/jledolter/DataMining/FuelEfficiency.csv")
+df <- read.csv("FuelEfficiency.csv")
+
+df <-data.frame(df)
+
+library(dplyr)
+
+glimpse(df)
+
+#GPM--> Gallon per 100 miles
+#MPG--> miles per Gallons
+#WT --> weiggt of the car
+#DIS --> Cubic displacement
+#NC --> No. of cylinders
+#HP --> Hourse power
+#ACC --> Acceleration
+# ET --> engine type
+
+
+#----------------------------
+plot.new()
+#par(mfrow=c(1,1))
+par(mar=c(4,4,1,1)+0.1,mfrow=c(3,3),bg="white",cex = 1, cex.main = 1)
+
+plot(GPM~MPG,data=df)
+
+plot(GPM~WT,data=df)
+
+plot(GPM~DIS,data=df)
+
+plot(GPM~NC,data=df)
+
+plot(GPM~HP,data=df)
+
+
+plot(GPM~ACC,data=df)
+plot(GPM~ET,data=df)
+
+dev.copy(png,'fueleff_xyplot.png',width = 800, height = 800)
+dev.off()
+
+df=df[-1] #ignore the MPG column
+df
+
+## regression on all data
+m1=lm(GPM~.,data=df)
+summary(m1)
+
+cor(df)
+
+## best subset regres0sion in R
+library(leaps)
+X=df[,2:7]
+y=df[,1]
+
+#use the regsubsets function from package leaps to compute regression of the subsets
+#https://www.rdocumentation.org/packages/leaps/versions/2.1-1/topics/regsubsets
+
+
+out=summary(regsubsets(X,y,nbest=2,nvmax=ncol(X)))
+tab=cbind(out$which,out$rsq,out$adjr2,out$cp)
+tab
+
+
+
+m2=lm(GPM~WT,data=df)
+summary(m2)
+
+m3=lm(GPM~WT+HP,data=df)
+m3
+
+summary(m3)
+
+
+## cross-validation (leave one out) for the model on all six regressors
+n=length(df$GPM)
+diff=dim(n)
+percdiff=dim(n)
+for (k in 1:n) {
+  train1=c(1:n)
+  train=train1[train1!=k]
+  ## the R expression "train1[train1!=k]" picks from train1 those
+  ## elements that are different from k and stores those elements in the
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/dplyr-general.r

@@ -1,0 +1,212 @@
+###################
+#     BƯỚC 1: Khai báo các gói phụ kiện (package) R cần thiết để phân tích
+
+# -----------------------------------------------------------------
+library(dplyr) # gói phụ kiện để tương tác và thay đổi dữ liệu
+
+###################
+#   BƯỚC 2: Thực hành với gói phụ kiện Dplyr
+
+#---Các bạn có thể đọc thêm hướng dẫn từ link **https://dplyr.tidyverse.org/**, để dễ nhớ bằng minh họa, chúng ta có thể dùng bảng CheetSheet 
+# https://raw.githubusercontent.com/rstudio/cheatsheets/master/pngs/thumbnails/data-transformation-cheatsheet-thumbs.png
+
+#https://www.rstudio.com/resources/cheatsheets/
+
+# Data
+
+
+# >>>>>>>>>       glimse() sử dụng hàm glimse để xem tổng hợp vắn tắt về dữ liệu
+iris <- data.frame(iris)
+
+car
+
+
+glimpse(iris)
+
+#>>>>>>>>>>       select()
+
+
+
+#cách 1
+df %>%
+  select(CAPEXassets, Facilities, FloorLevel, Area, Tenant, WorkType, Disciplines,Quantity,  Unit, States, Risk, InteType,IntePec, UnitCost, LifeSpan, Prob, InterYear, NPVCapexP2Php)
+#cách 2
+df %>%
+  select(c(2,3,4,5,6,7,8,9,11,12,13,14,15,17,18,19,20,21,22,26,27))
+#cách 3
+df %>%
+  select(c(2:9,11:15,17:22,26,27))
+
+# cách 4
+
+df %>%
+  select(-c(No,YearBuilt,CodeComply,TenantInterYear,IncludedPhase1,NPVCAPEXP1USD,NPVCapexP2Php,NPVCapexUSD,ReturntoAs.built,ReasonsforIntervention,Scopes,PictureReferenceCode,Remarks))
+
+
+
+
+
+# chọn các cột CAPEXassets, Facilities, FloorLevel, Area, Tenant, WorkType, Disciplines,Quantity,  Unit, States, Risk, NPVCapexP2Php
+
+#đặt tên mới là df01
+df01 <- df%>%
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/dplyr-hanjinshipyard.r

@@ -1,0 +1,212 @@
+###################
+#     BƯỚC 1: Khai báo các gói phụ kiện (package) R cần thiết để phân tích
+
+
+# -----------------------------------------------------------------
+library(dplyr) # gói phụ kiện để tương tác và thay đổi dữ liệu
+library(readxl)# gói phụ kiện để đọc dữ liệu từ bảng tính excel
+
+###################
+
+#     BƯỚC 2: Đọc dữ liệu vào R
+
+df=read_excel("../../data/tdd/hanjinshipyard.xlsx",sheet="CAPEX",skip = 2) # Tải dữ liệu vào R, đọc bảng tính tab có tên CAPEX, bỏ qua 2 dòng đầu tiên
+
+df <- data.frame(df) # Chuyển đổi dữ liệu thành dạng khung
+
+
+###################
+#   BƯỚC 3: Thực hành với gói phụ kiện Dplyr
+
+#---Các bạn có thể đọc thêm hướng dẫn từ link **https://dplyr.tidyverse.org/**, để dễ nhớ bằng minh họa, chúng ta có thể dùng bảng CheetSheet 
+# https://raw.githubusercontent.com/rstudio/cheatsheets/master/pngs/thumbnails/data-transformation-cheatsheet-thumbs.png
+
+#https://www.rstudio.com/resources/cheatsheets/
+
+
+# >>>>>>>>>       glimse() sử dụng hàm glimse để xem tổng hợp vắn tắt về dữ liệu
+glimpse(df)
+
+#>>>>>>>>>>       select()
+# chọn các cột CAPEXassets, Facilities, FloorLevel, Area, Tenant, WorkType, Disciplines,Quantity,  Unit, States, Risk, IntePec, UnitCost, LifeSpan, Prob, InterYear, NPVCapexP2Php
+
+#cách 1
+df %>%
+  select(CAPEXassets, Facilities, FloorLevel, Area, Tenant, WorkType, Disciplines,Quantity,  Unit, States, Risk, InteType,IntePec, UnitCost, LifeSpan, Prob, InterYear, NPVCapexP2Php)
+#cách 2
+df %>%
+  select(c(2,3,4,5,6,7,8,9,11,12,13,14,15,17,18,19,20,21,22,26,27))
+#cách 3
+df %>%
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/dplyr-iris.R

@@ -1,0 +1,99 @@
+## Người soạn bài giảng: Lê Thanh Nam (namlt@asq.vn), https://namkyodai.github.io
+
+###################
+#     BƯỚC 1: Khai báo các gói phụ kiện (package) R cần thiết để phân tích
+
+
+# -----------------------------------------------------------------
+library(dplyr) # gói phụ kiện để tương tác và thay đổi dữ liệu
+
+###################
+
+#     BƯỚC 2: Đọc dữ liệu vào R
+
+df <- iris
+
+# để biết có bao nhiêu gói dữ liệu có sẵn trên R, các bạn có thể dùng lệnh
+
+# https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html
+
+df <- data.frame(df) # Chuyển đổi dữ liệu thành dạng khung
+
+
+###################
+#   BƯỚC 3: Thực hành với gói phụ kiện Dplyr
+
+#---Các bạn có thể đọc thêm hướng dẫn từ link **https://dplyr.tidyverse.org/**, để dễ nhớ bằng minh họa, chúng ta có thể dùng bảng CheetSheet 
+# https://raw.githubusercontent.com/rstudio/cheatsheets/master/pngs/thumbnails/data-transformation-cheatsheet-thumbs.png
+
+
+# >>>>>>>>>       glimse() sử dụng hàm glimse để xem tổng hợp vắn tắt về dữ liệu
+glimpse(df)
+
+# giải thích dữ liệu
+# iris là một loài hoa, tên tiếng Việt là hoa Chi Diên Vĩ và có họ hàng với hoa Lay Ơn
+# https://eva.vn/cay-canh-vuon/hoa-dien-vi-y-nghia-nguon-goc-cong-dung-va-cach-trong-c283a474719.html
+# -------5 cột, tương ứng với 5 biến
+#--------
+
+
+
+#>>>>>>>>>>       select()
+# chọn các cột CAPEXassets, Facilities, FloorLevel, Area, Tenant, WorkType, Disciplines,Quantity,  Unit, States, Risk, IntePec, UnitCost, LifeSpan, Prob, InterYear, NPVCapexP2Php
+
+#cách 1
+df %>%
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/orangejuice/main.R

@@ -1,0 +1,137 @@
+## Install packages from CRAN; use any USA mirror
+library(lattice)
+#oj <- read.csv("https://www.biz.uiowa.edu/faculty/jledolter/DataMining/oj.csv")
+oj <- read.csv("oj.csv")
+oj$store <- factor(oj$store) #change numberic value of store into categorical data
+oj[1:2,]
+t1=tapply(oj$logmove,oj$brand,FUN=mean,na.rm=TRUE) #calculate the mean of each brand using logmove value.
+t1
+t2=tapply(oj$logmove,INDEX=list(oj$brand,oj$week),FUN=mean,na.rm=TRUE) #calculate the mean of logmove value based on index lists per week.
+t2
+#plot each graph as time serieas data per week.
+plot.new()
+par(mar=c(4.5,4.3,1,1)+0.1,mfrow=c(2,2),bg="white",cex = 1, cex.main = 0.6)
+plot(t2[1,],type= "l",xlab="week",ylab="dominicks",ylim=c(7,12),cex.axis = 1,las = 1)
+plot(t2[2,],type= "l",xlab="week",ylab="minute.maid",ylim=c(7,12),cex.axis = 1,las = 1)
+plot(t2[3,],type= "l",xlab="week",ylab="tropicana",ylim=c(7,12),cex.axis = 1,las = 1)
+dev.copy(png,'oj_weekmean01.png',width = 1600, height = 600)
+dev.off()
+#-------------------------------
+#now we combine the three above graphs into one single graphs for ease of comparison
+logmove=c(t2[1,],t2[2,],t2[3,])
+week1=c(40:160)
+week=c(week1,week1,week1)
+brand1=rep(1,121)
+brand2=rep(2,121)
+brand3=rep(3,121)
+brand=c(brand1,brand2,brand3)
+plot.new()
+xyplot(logmove~week|factor(brand),type= "l",layout=c(1,3),col="black")
+dev.copy(png,'oj_weekmean02.png',width = 1000, height = 600)
+dev.off()
+#-----------------------------
+plot.new()
+par(mfrow=c(1,1))
+#par(mar=c(4.5,4.3,1,1)+0.1,mfrow=c(2,2),bg="white",cex = 1, cex.main = 0.6)
+boxplot(logmove~brand,data=oj) # compare logmove of 3 branch using boxplot
+dev.copy(png,'oj_logmovebrandboxplot.png',width = 800, height = 600)
+dev.off()
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week2/week2-assignment-solution.r

@@ -1,0 +1,321 @@
+#000000000000000000000000000000000000000000000000000000000
+
+
+# A. Sử dụng thư viện dplyr của R để trả lời các câu hỏi sau liên quan đến gói dữ liệu Births2006
+
+#000000000000000000000000000000000000000000000000000000000
+
+
+# 1. Có bao nhiêu trẻ sơ sinh là Nam, bao nhiêu là Nữ?
+# 2. Có bao nhiêu phụ sản mà chúng ta không biết trình độ học vấn?
+# 3. Có bao nhiêu phụ sản sinh ít hơn 3 bé sơ sinh và sử dụng phương pháp đẻ tự nhiên?
+# 4. Có bao nhiêu trẻ sinh ra có chỉ số APGAR5 nhỏ hơn 3 và có cân nặng lớn hơn 2.5 kg? Trong số trẻ này thì có bao nhiêu trẻ sinh bằng phương pháp tự nhiên, bao nhiêu sinh ra bằng phương pháp cắt.
+# 5. Tháng nào trong năm có số lượng trẻ sơn sinh nhiều nhất?
+# 6. Câu hỏi phụ (không nhất thiết phải trả lời), các bạn có thẻ vẽ biểu đồ để thể hiện dữ liệu cho 5 câu hỏi trên không? --> sử dụng biểu đồ barchart, histogram sử dụng phụ kiện ggplot2
+# 7. Tìm mối tương quan giữa cân nặng của trẻ sơ sinh với các biến còn lại? Các biến nào các bạn sẽ chọn để dùng phân tích? Các bạn có kết luận gì không sau khi chạy phân tích với hồi qui tuyến tính chuẩn?
+
+
+library(dplyr)
+
+df <-read.csv("births2006.csv",header = TRUE, sep = ",")
+df <- data.frame(df)
+
+
+glimpse(df)
+
+# 1. Có bao nhiêu trẻ sơ sinh là Nam, bao nhiêu là Nữ?
+
+summary(df %>% select(SEX))
+
+#cách 1
+df %>%
+  select(SEX) %>%
+  count(SEX)
+
+
+#cách 2
+
+df %>%
+  count(SEX)
+
+
+# 2. Có bao nhiêu phụ sản mà chúng ta không biết trình độ học vấn
+
+# cách 1
+df %>%
+  filter(DMEDUC == "NULL")%>%
+  count(DMEDUC)
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/NOxEmissions/main.R

@@ -1,0 +1,92 @@
+#set working directory for the file
+path <- rstudioapi::getActiveDocumentContext()$path
+Encoding(path) <- "UTF-8"
+setwd(dirname(path))
+getwd()
+#-------------------------------------
+
+library(locfit)
+
+## first we read in the data
+ethanol <- read.csv("ethanol.csv")
+
+ethanol[1:3,]
+
+## density histogram
+hist(ethanol$NOx,freq=FALSE)
+## smoothed density histogram
+fit <- locfit(~lp(NOx),data=ethanol)
+plot(fit)
+
+## experiment with the parameters of locfit
+fit <- locfit(~lp(NOx,deg=1),data=ethanol)
+plot(fit)
+fit <- locfit(~lp(NOx,nn=0.7,deg=1),data=ethanol)
+plot(fit)
+fit <- locfit(~lp(NOx,nn=0.5,deg=1),data=ethanol)
+plot(fit)
+
+fit <- locfit(~lp(NOx,deg=2),data=ethanol)
+plot(fit)
+fit <- locfit(~lp(NOx,nn=0.7,deg=2),data=ethanol)
+plot(fit)
+fit <- locfit(~lp(NOx,nn=0.5,deg=2),data=ethanol)
+plot(fit)
+
+fit <- locfit(~lp(NOx,deg=3),data=ethanol)
+plot(fit)
+fit <- locfit(~lp(NOx,nn=0.7,deg=3),data=ethanol)
+plot(fit)
+fit <- locfit(~lp(NOx,nn=0.5,deg=3),data=ethanol)
+plot(fit)
+
+## standard regression of NOx on the equivalence ratio
+plot(NOx~EquivRatio,data=ethanol)
+fitreg=lm(NOx~EquivRatio,data=ethanol)
+plot(NOx~EquivRatio,data=ethanol)
+abline(fitreg)
+
+## local polynomial regression of NOx on the equivalence ratio
+## fit with a 50% nearest neighbor bandwidth.
+fit <- locfit(NOx~lp(EquivRatio,nn=0.5),data=ethanol)
+plot(fit)
+
+## experiment with the parameters of locfit
+fit <- locfit(NOx~lp(EquivRatio,nn=0.2),data=ethanol)
+plot(fit)
+fit <- locfit(NOx~lp(EquivRatio,nn=0.8),data=ethanol)
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/ggplot-boxplot-Viet.r

@@ -1,0 +1,328 @@
+# Load ggplot2
+library(ggplot2)
+
+# The mtcars dataset is natively available
+head(mtcars)
+glimpse(mtcars)
+# A really basic boxplot.
+
+# Biểu đồ hộp cho 1 biến
+
+ggplot(mtcars, aes(x="", y=mpg)) + #ggplot(cơ sở dữ liệu, aes: thiết lập biến cho trục tung và trục hoành: trục x = rỗng, trục y = cột mpg) +
+  geom_boxplot() # goem_boxplot() hàm mặc định chạy hiển thị biểu đồ boxplot, tuy nhiên trục x do không gán giá trị nào nên sẽ hiển thị dấu X
+
+
+
+
+ggplot(mtcars, aes(x="", y=mpg)) + #ggplot(cơ sở dữ liệu, aes: thiết lập biến cho trục tung và trục hoành: trục x = rỗng, trục y = cột mpg) +
+  geom_boxplot() + # goem_boxplot() hàm mặc định chạy hiển thị biểu đồ boxplot
+  xlab("") #đặt tên trục x là rỗng
+
+
+ggplot(mtcars, aes(x="", y=mpg)) + #ggplot(cơ sở dữ liệu, aes: thiết lập biến cho trục tung và trục hoành: trục x = rỗng, trục y = cột mpg) +
+  geom_boxplot(alpha=0.1) + # goem_boxplot() chạy hiển thị biểu đồ boxplot, tuy nhiên làm mờ toàn bộ biểu đồ với giá trị 0.1
+  xlab("") #đặt tên trục x là rỗng
+
+
+ggplot(mtcars, aes(x="", y=mpg)) + #ggplot(cơ sở dữ liệu, aes: thiết lập biến cho trục tung và trục hoành: trục x = rỗng, trục y = cột mpg)
+  geom_boxplot(fill="slateblue", alpha=0.2) + # goem_boxplot() chạy hiển thị biểu đồ boxplot, đổ màu xanh và làm mờ toàn bộ biểu đồ với giá trị 0.2
+  xlab("")+ #đặt tên trục x là rỗng
+  labs(title="Biểu đồ hộp cho biến mpg", subtitle="ví dụ về thêm màu") #đặt tiêu đề cho toàn bộ biểu đồ và tiêu đề phụ
+
+#thêm một lớp cho biểu đồ
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/ggplot-boxplot.r

@@ -1,0 +1,293 @@
+# Load ggplot2
+library(ggplot2)
+
+# The mtcars dataset is natively available
+# head(mtcars)
+
+# A really basic boxplot.
+
+# Biểu đồ hộp cho 1 biến
+
+mtcars
+
+ggplot(mtcars, aes(x="", y=mpg)) +
+  geom_boxplot()
+
+
+ggplot(mtcars, aes(x="", y=mpg)) +
+  geom_boxplot() +
+  xlab("")
+
+
+ggplot(mtcars, aes(x="", y=mpg)) +
+  geom_boxplot(alpha=0.2, colour = "red", fill = "blue") +
+  xlab("")
+
+
+ggplot(mtcars, aes(x="", y=mpg)) +
+  geom_boxplot(fill="midnightblue", alpha=0.3, colour = "mediumseagreen") +
+  xlab("")+
+  labs(title="Biểu đồ hộp cho biến mpg",
+       subtitle="ví dụ về thêm màu",
+       y = "Quãng đường trên 1 lít xăng")
+
+
+#thêm một lớp cho biểu đồ
+ggplot(mtcars, aes(x="", y=mpg)) +
+  geom_boxplot(fill="slateblue", alpha=0.2) +
+  xlab("")+
+  ylab("Quãng đường trên 1 lít xăng")
+
+# Cùng một biểu đồ nhưng thêm biến
+
+#http://www.sthda.com/english/wiki/ggplot2-point-shapes
+ggplot(mtcars, aes(x=as.factor(cyl), y=mpg)) +
+  geom_boxplot(fill="red", alpha=0.2)+
+  xlab("Dung tích xi lanh")+
+  ylab("Quãng đường trên 1 lít xăng")+
+  stat_summary(fun=mean, geom="point", shape=20, size=4, color="red", fill="red")+
+  theme(legend.position="none") +
+  scale_fill_brewer(palette="Set1")
+
+
+
+ggplot(mtcars, aes(x=as.factor(cyl), y=mpg, fill = as.factor(cyl))) +
+  geom_boxplot(alpha=0.7) +
+  xlab("Dung tích xi lanh")+
+  ylab("Quãng đường trên 1 lít xăng")+
+  stat_summary(fun=mean, geom="point", shape=20, size=4, color="red", fill="red")+
+  theme(legend.position="none") +
+  scale_fill_brewer(palette="Set1")
+
+
+
+
+ggplot(mtcars, aes(x=as.factor(cyl), y=mpg, fill = as.factor(cyl))) +
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/ggplot-density.r

@@ -1,0 +1,88 @@
+library(ggplot2)
+library(dplyr)
+
+# Load dataset from github
+data <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/1_OneNum.csv", header=TRUE)
+data <- data.frame(data)
+#view(data)
+# Make the histogram
+
+
+
+
+data %>%
+  # filter( price<300 ) %>%
+  ggplot( aes(x=price)) +
+  geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)
+
+
+data %>%
+  # filter( price<300 ) %>%
+  ggplot( aes(x=price)) +
+  geom_boxplot()
+
+quantile(data$price,probs=seq(0,1,0.05)) #quantile from 0 to 1 with a step of 5%
+quantile(data$price,probs=seq(0.95,1,0.01)) #quantilte from 0.95 to 1 with a step of 1%
+
+
+data %>%
+  filter( price<300 ) %>%
+  ggplot( aes(x=price)) +
+  geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)
+
+
+# thay đổi một chút cho đẹp hơn :)
+
+p<-data %>%
+  filter( price<300 ) %>%
+  ggplot( aes(x=price)) +
+  geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)
+
+p+labs(title = "Biểu đồ mật độ",
+       y = "Mật độ",
+       x = "Giá (đô la mỹ)"
+
+)
+# biểu đồ tần số
+
+q<-data %>%
+  filter( price<300 ) %>%
+  ggplot( aes(x=price))+
+  geom_histogram(fill="red",alpha=0.4) #fill="#69b3a2", color="#e9ecef", alpha=0.8
+
+
+q
+# Giờ ta có thể vẻ thêm một chút cho đẹp
+
+# vẽ density plot vào
+
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/ggplot-scatterplot.r

@@ -1,0 +1,41 @@
+library(ggplot2)
+
+# The iris dataset is provided natively by R
+#head(iris)
+
+# basic scatterplot
+
+ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width)) +
+   geom_point()
+
+
+p<-ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width)) 
+
+p + geom_point()
+
+p + geom_point()+
+  annotate("rect", xmin = 5.5, xmax = 7, ymin = 2, ymax = 3.5, alpha = .2) + # tô thêm màu thế nào cho đẹp 
+  annotate("segment", x = 5.5, xend = 7, y = 2.5, yend = 3.3)+
+  annotate("pointrange", x = 6, y = 3, ymin = 2, ymax = 4, colour = "red", size = 1.5)
+
+# let play with some emoji
+
+#https://cran.r-project.org/web/packages/emojifont/vignettes/emojifont.html
+library(emojifont)
+p + geom_point()+
+  annotate("rect", xmin = 5.5, xmax = 7, ymin = 2, ymax = 3.5, alpha = .2) + # tô thêm màu thế nào cho đẹp 
+  annotate("segment", x = 5.5, xend = 7, y = 2.5, yend = 3.3)+
+  annotate("pointrange", x = 6, y = 3, ymin = 2, ymax = 4, colour = "red", size = 1.5)+
+  geom_emoji('smiley', x=7, y=4, size=20)
+
+
+p + geom_point()+
+   annotate("segment", x = 5.5, xend = 7, y = 2.5, yend = 3.3)+
+  annotate("rect", xmin = 5.5, xmax = 7, ymin = 2, ymax = 3.5, alpha = .2) +
+  labs(title= "Biểu đồ scatterplot",
+       x ="Chiều rộng loài Sepal",
+       y = "chiều dài loài Sepal")
+  
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/lm-prediction.r

@@ -1,0 +1,42 @@
+# Values of height
+x <- c(151, 174, 138, 186, 128, 136, 179, 163, 152, 131)
+
+# Values of weight.
+y <- c(63, 81, 56, 91, 47, 57, 76, 72, 62, 48)
+
+df <- data.frame(x,y)
+
+relation <- lm(y~x)
+
+summary(relation)
+
+# Tiên đón cân năng của 1 người có chiều cao là a
+
+a <- data.frame(x = 170)
+
+result <-  predict(relation,a)
+
+result
+
+# Tiên đón cân nặng của một nhóm người
+
+b <- data.frame(x= c(144,165,180))
+
+result <-  predict(relation,b)
+
+result
+
+
+# Vẽ biểu đồ
+
+library(ggplot2)
+
+p <-df %>% 
+  ggplot(aes(x,y))+
+  geom_point()+
+  geom_smooth(method='lm', formula= y~x, se = FALSE)
+p
+
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/test.r

@@ -1,0 +1,6 @@
+library(plotly)
+fig <- plot_ly(x = ~rnorm(50), type = "histogram")
+
+fig
+
+x = ~rnorm(50)

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week3/week2-assignment-solution.r

@@ -1,0 +1,327 @@
+#000000000000000000000000000000000000000000000000000000000
+
+
+# A. Sử dụng thư viện dplyr của R để trả lời các câu hỏi sau liên quan đến gói dữ liệu Births2006
+
+#000000000000000000000000000000000000000000000000000000000
+
+
+# 1. Có bao nhiêu trẻ sơ sinh là Nam, bao nhiêu là Nữ?
+# 2. Có bao nhiêu phụ sản mà chúng ta không biết trình độ học vấn?
+# 3. Có bao nhiêu phụ sản sinh ít hơn 3 bé sơ sinh và sử dụng phương pháp đẻ tự nhiên?
+# 4. Có bao nhiêu trẻ sinh ra có chỉ số APGAR5 nhỏ hơn 3 và có cân nặng lớn hơn 2.5 kg? Trong số trẻ này thì có bao nhiêu trẻ sinh bằng phương pháp tự nhiên, bao nhiêu sinh ra bằng phương pháp cắt.
+# 5. Tháng nào trong năm có số lượng trẻ sơn sinh nhiều nhất?
+# 6. Câu hỏi phụ (không nhất thiết phải trả lời), các bạn có thẻ vẽ biểu đồ để thể hiện dữ liệu cho 5 câu hỏi trên không? --> sử dụng biểu đồ barchart, histogram sử dụng phụ kiện ggplot2
+# 7. Tìm mối tương quan giữa cân nặng của trẻ sơ sinh với các biến còn lại? Các biến nào các bạn sẽ chọn để dùng phân tích? Các bạn có kết luận gì không sau khi chạy phân tích với hồi qui tuyến tính chuẩn?
+
+
+library(dplyr)
+
+df <-read.csv("../../Code/week2/Births/births2006.csv",header = TRUE, sep = ",")
+df <- data.frame(df)
+
+
+glimpse(df)
+
+# 1. Có bao nhiêu trẻ sơ sinh là Nam, bao nhiêu là Nữ?
+
+summary(df %>% select(SEX))
+
+#cách 1
+df %>%
+  select(SEX) %>%
+  count(SEX)
+
+
+#cách 2
+
+df %>%
+  count(SEX)
+
+
+# 2. Có bao nhiêu phụ sản mà chúng ta không biết trình độ học vấn
+
+# cách 1
+df %>%
+  filter(DMEDUC == "NULL")%>%
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/FlightDelays.r

@@ -1,0 +1,101 @@
+library(car) ## needed to recode variables
+set.seed(1)
+
+## read and print the data
+del <- read.csv("FlightDelays.csv")
+del[1:3,]
+
+## define hours of departure
+del$sched=factor(floor(del$schedtime/100))
+table(del$sched)
+table(del$carrier)
+table(del$dest)
+table(del$origin)
+table(del$weather)
+table(del$dayweek)
+table(del$daymonth)
+table(del$delay)
+del$delay=recode(del$delay,"'delayed'=1;else=0")
+del$delay=as.numeric(levels(del$delay)[del$delay])
+table(del$delay)
+## Delay: 1=Monday; 2=Tuesday; 3=Wednesday; 4=Thursday; 
+## 5=Friday; 6=Saturday; 7=Sunday
+## 7=Sunday and 1=Monday coded as 1
+del$dayweek=recode(del$dayweek,"c(1,7)=1;else=0")
+table(del$dayweek)
+## omit unused variables
+del=del[,c(-1,-3,-5,-6,-7,-11,-12)]
+del[1:3,]
+n=length(del$delay)
+n
+n1=floor(n*(0.6))
+n1
+n2=n-n1
+n2
+train=sample(1:n,n1)
+
+## estimation of the logistic regression model
+## explanatory variables: carrier, destination, origin, weather, day of week 
+## (weekday/weekend), scheduled hour of departure
+## create design matrix; indicators for categorical variables (factors)
+Xdel <- model.matrix(delay~.,data=del)[,-1] 
+Xdel[1:3,]
+xtrain <- Xdel[train,]
+xnew <- Xdel[-train,]
+ytrain <- del$delay[train]
+ynew <- del$delay[-train]
+m1=glm(delay~.,family=binomial,data=data.frame(delay=ytrain,xtrain))
+summary(m1)
+
+## prediction: predicted default probabilities for cases in test set
+ptest <- predict(m1,newdata=data.frame(xnew),type="response")
+data.frame(ynew,ptest)[1:10,]
+## first column in list represents the case number of the test element
+plot(ynew~ptest)
+
+## coding as 1 if probability 0.5 or larger
+gg1=floor(ptest+0.5)	## floor function; see help command
+ttt=table(ynew,gg1)
+ttt
+error=(ttt[1,2]+ttt[2,1])/n2
+error
+
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/OldFaithful/main.R

@@ -1,0 +1,114 @@
+library(locfit)  
+library(ggplot2)
+library(dplyr)
+## first we read in the data
+## first we read in the data
+
+df <- read.csv("OldFaithful.csv")
+
+df <- data.frame(df)
+
+df[1:10,]
+
+## density histograms and smoothed density histograms
+## time of eruption
+plot.new()
+#par(mfrow=c(1,1))
+par(mar=c(4,4,1,1)+0.1,mfrow=c(1,2),bg="white",cex = 1, cex.main = 1)
+
+hist(df$TimeEruption,freq=FALSE)
+
+
+#use locfit https://www.rdocumentation.org/packages/locfit/versions/19980714-2/topics/locfit
+fit1 <- locfit(~lp(TimeEruption),data=df)
+
+#lp function https://www.rdocumentation.org/packages/lpSolve/versions/5.6.13.1/topics/lp
+#lp is a local polynomial model term for Locfit models.
+#https://www.rdocumentation.org/packages/locfit/versions/1.5-9.1/topics/lp
+plot(fit1)
+#dev.copy(png,'oldfaithful_timeeruption01.png',width = 800, height = 400)
+#dev.off()
+
+plot.new()
+
+df %>%
+  ggplot(aes(TimeEruption))+
+  geom_histogram()
+
+df %>%
+  ggplot(aes(TimeEruption))+
+  geom_density()+
+  geom_histogram(aes(y = ..density..),
+                 colour = 1, fill = "white", alpha =0.4)
+  
+
+plot.new()
+par(mar=c(4,4,1,1)+0.1,mfrow=c(1,2),bg="white",cex = 1, cex.main = 1)
+## waiting time to next eruption
+hist(df$TimeWaiting,freq=FALSE)
+
+fit2 <- locfit(~lp(TimeWaiting),data=df)
+plot(fit2)
+#dev.copy(png,'oldfaithful_TimeWaiting01.png',width = 800, height = 400)
+#dev.off()
+
+#------------------------------
+## experiment with different smoothing constants
+fit3 <- locfit(~lp(TimeWaiting,nn=0.9,deg=2),data=df) #nn is the nearest neighbour component, and deg is the degree of polynomial. default value of nn is 0.6 and deg is 2.
+plot(fit3)
+fit4 <- locfit(~lp(TimeWaiting,nn=0.3,deg=2),data=df)
+plot(fit4)
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/dealthpenalty.r

@@ -1,0 +1,35 @@
+## analyzing individual observations
+df <- data.frame(read.csv("DeathPenalty.csv"))
+df[1:4,]
+df[359:362,]
+m1=glm(Death~VRace+Agg,family=binomial,data=df)
+m1
+summary(m1)
+## calculating logits
+exp(m1$coef[2])
+exp(m1$coef[3])
+## plotting probability of getting death penalty as a function of aggravation
+## separately for black (in black) and white (in red) victim
+fitBlack=dim(501)
+fitWhite=dim(501)
+ag=dim(501)
+for (i in 1:501) {
+  ag[i]=(99+i)/100
+  fitBlack[i]=exp(m1$coef[1]+ag[i]*m1$coef[3])/(1+exp(m1$coef[1]+ag[i]*m1$coef[
+    3]))
+  fitWhite[i]=exp(m1$coef[1]+m1$coef[2]+ag[i]*m1$coef[3])/(1+exp(m1$coef[1]+m1$
+                                                                   coef[2]+ag[i]*m1$coef[3]))
+}
+plot(fitBlack~ag,type="l",col="black",ylab="Prob[Death]",xlab="Aggravation",ylim=c(0,1),main="red line for white victim; black line for black victim")
+points(fitWhite~ag,type="l",col="red")
+## analyzing summarized data
+dpenother <- read.csv("DeathPenaltyOther.csv")
+dpenother
+m1=glm(Death~VRace+Agg,family=binomial,weights=Freq,data=dpenother)
+m1
+summary(m1)
+exp(m1$coef[2])
+exp(m1$coef[3])
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/mushrooms.r

@@ -1,0 +1,54 @@
+library(rpart,quietly = TRUE)
+library(caret,quietly = TRUE)
+library(rpart.plot,quietly = TRUE)
+library(rattle)
+
+
+mushrooms <- read.csv ("mushrooms.csv")
+str(mushrooms)
+
+
+mushrooms <- mushrooms%>%mutate_if(is.character, as.factor)
+str(mushrooms)
+
+
+nrow(mushrooms) - sum(complete.cases(mushrooms))
+
+mushrooms$veil.type <- NULL
+table(mushrooms$class,mushrooms$odor)
+
+
+number.perfect.splits <- apply(X=mushrooms[-1], MARGIN = 2, FUN = function(col){
+  t <- table(mushrooms$class,col)
+  sum(t == 0)
+})
+
+set.seed(12345)
+train <- sample(1:nrow(mushrooms),size = ceiling(0.80*nrow(mushrooms)),replace = FALSE)
+# training set
+mushrooms_train <- mushrooms[train,]
+# test set
+mushrooms_test <- mushrooms[-train,]
+
+# penalty matrix
+penalty.matrix <- matrix(c(0,1,10,0), byrow=TRUE, nrow=2)
+
+# building the classification tree with rpart
+tree <- rpart(class~.,
+              data=mushrooms_train,
+              parms = list(loss = penalty.matrix),
+              method = "class")
+
+# Visualize the decision tree with rpart.plot
+rpart.plot(tree, nn=TRUE)
+
+
+#https://www.guru99.com/r-decision-trees.html
+accuracy_tune <- function(fit) {
+  predict_unseen <- predict(fit, data_test, type = 'class')
+  table_mat <- table(data_test$survived, predict_unseen)
+  accuracy_Test <- sum(diag(table_mat)) / sum(table_mat)
+  accuracy_Test
+}
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/radiant.r

@@ -1,0 +1,4 @@
+library(radiant)
+radiant()
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/titanic.r

@@ -1,0 +1,39 @@
+set.seed(678)
+path <- 'https://raw.githubusercontent.com/guru99-edu/R-Programming/master/titanic_data.csv'
+titanic <-read.csv(path)
+head(titanic)
+tail(titanic)
+shuffle_index <- sample(1:nrow(titanic))
+head(shuffle_index)
+titanic <- titanic[shuffle_index, ]
+head(titanic)
+library(dplyr)
+# Drop variables
+clean_titanic <- titanic %>%
+  select(-c(home.dest, cabin, name, x, ticket)) %>% 
+  #Convert to factor level
+  mutate(pclass = factor(pclass, levels = c(1, 2, 3), labels = c('Upper', 'Middle', 'Lower')),
+         survived = factor(survived, levels = c(0, 1), labels = c('No', 'Yes'))) %>%
+  na.omit()
+glimpse(clean_titanic)
+
+create_train_test <- function(data, size = 0.8, train = TRUE) {
+  n_row = nrow(data)
+  total_row = size * n_row
+  train_sample < - 1: total_row
+  if (train == TRUE) {
+    return (data[train_sample, ])
+  } else {
+    return (data[-train_sample, ])
+  }
+}
+
+
+data_train <- create_train_test(clean_titanic, 0.8, train = TRUE)
+
+
+data_test <- create_train_test(clean_titanic, 0.8, train = FALSE)
+dim(data_train)
+
+
+

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/weather01-optimal.r

@@ -1,0 +1,43 @@
+# Đọc dữ liệu thời tiết từ file excel
+library(rstudioapi) # thư viện dùng để đặt đường dẫn tới file gốc
+setwd(dirname(getActiveDocumentContext()$path))       # Đặt thư mục làm việc trùng với thư mục của file
+library(readxl) 
+library(dplyr)
+library(ggplot2)
+library(rpart)
+library(rpart.plot)
+
+thoi.tiet <-data.frame(read_excel("weather.xlsx",sheet="01",skip = 0)) # đọc dữ liệu từ file excel
+
+thoi.tiet$temp.d <- cut(thoi.tiet$temp, breaks = c(-Inf,20,26,Inf),
+                        labels = c("cool","mild","hot"),include.lowest = F)
+
+thoi.tiet$humi.d <- cut(thoi.tiet$humi, breaks = c(-Inf,80,Inf),
+                        labels = c("normal","high"),include.lowest = F)
+
+thoi.tiet$wind.d <- cut(thoi.tiet$wind, breaks = c(-Inf,20,Inf),
+                        labels = c("weak","strong"),include.lowest = F)
+
+# Biến các biến rời rạc thành dạng factor
+thoi.tiet <- thoi.tiet%>%mutate_if(is.character, as.factor)
+
+# Sử dụng thư viện vtree để vẽ thống kê cơ bản.
+library(vtree) #https://www.infoworld.com/article/3573577/how-to-count-by-groups-in-r.html
+vtree(thoi.tiet, c("outlook", "temp.d","humi.d", "play"),horiz = FALSE)
+
+
+## bắt đầu tính giá trị hàm số entropy
+control <- rpart.control(minsplit = 2,
+                         minbucket = 1,
+                         maxdepth = 4,
+                         cp = 0.01)
+#https://www.rdocumentation.org/packages/rpart/versions/4.1-15/topics/rpart.control
+
+# building the classification tree with rpart
+tree <- rpart(play~ outlook + temp.d + humi.d + wind.d,
+              data=thoi.tiet,
+              parms=list(split='information'),
+              control = control,
+              method = "class")
+tree
+rpart.plot(tree, nn=TRUE)

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/weather01.r

@@ -1,0 +1,207 @@
+library(rstudioapi) # thư viện dùng để đặt đường dẫn tới file gốc
+setwd(dirname(getActiveDocumentContext()$path))  
+# Đọc dữ liệu thời tiết từ file excel
+
+library(readxl)
+library(tidyverse)
+library(dplyr)
+library(ggplot2)
+library(rpart)
+library(rpart.plot)
+
+
+thoi.tiet <-data.frame(read_excel("weather.xlsx",sheet="01",skip = 0)) # đọc dữ liệu từ file excel
+
+thoi.tiet$temp.d <- cut(thoi.tiet$temp, breaks = c(-Inf,20,26,Inf),
+                        labels = c("cool","mild","hot"),include.lowest = F)
+
+thoi.tiet$humi.d <- cut(thoi.tiet$humi, breaks = c(-Inf,80,Inf),
+                        labels = c("normal","high"),include.lowest = F)
+
+thoi.tiet$wind.d <- cut(thoi.tiet$wind, breaks = c(-Inf,20,Inf),
+                        labels = c("weak","strong"),include.lowest = F)
+
+# Biến các biến rời rạc thành dạng factor
+## Cách 1
+
+
+thoi.tiet$outlook <- factor(thoi.tiet$outlook, levels = c("sunny","overcast","rain"))
+thoi.tiet$temp.d <- factor(thoi.tiet$temp.d, levels = c("hot","mild","cool"))
+thoi.tiet$humi.d <- factor(thoi.tiet$humi.d, levels = c("high","normal"))
+thoi.tiet$wind.d <- factor(thoi.tiet$wind.d, levels = c("strong","weak"))
+thoi.tiet$play <- factor(thoi.tiet$play, levels = c("yes","no"))
+
+## Cách 2
+#thoi.tiet <- thoi.tiet%>%mutate_if(is.character, as.factor)
+#str(thoi.tiet)
+
+
+3*3*2*2
+
+write.csv(thoi.tiet,"weather01-factor.csv", row.names = FALSE)
+
+
+names(thoi.tiet) #kiểm tra tên biến
+
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week4/weather02.r

@@ -1,0 +1,903 @@
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(rstudioapi) # thư viện dùng để đặt đường dẫn tới file gốc
+setwd(dirname(getActiveDocumentContext()$path))
+
+
+#### PHẦN 1: Giới thiệu về gói dữ liệu liên quan đến lượng mưa đo được ở một trạm khí tượng.
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(readxl)
+
+#weather <-data.frame(read_excel("weather.xlsx",sheet="02", skip = 0)) # đọc dữ liệu từ file excel
+weather <- data.frame(read.csv("weather.csv",sep=";",stringsAsFactors=FALSE))
+str(weather)
+library(dplyr)
+glimpse(weather)
+
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+#### PHẦN 2: Chuẩn bị dữ liệu (Wrangling)
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+
+dim(weather) # xem tổng quan hàng và cột của gói dữ liệu
+
+names(weather) # xem tên của các biến (Cột)
+
+head(weather) # xem 10 dòng đầu tiên của gói dữ liệu
+
+str(weather) # xem dạng của từng biến
+
+
+# Kiểm tra xem gói dữ liệu có bao nhiêu mục bị bỏ trống
+sum(is.na(weather))
+
+# xem số dòng của gói dữ liệu
+nrow(weather)
+sum(complete.cases(weather))
+
+nrow(weather) == sum(complete.cases(weather))
+
+
+#############         TẠO FACTORS
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Code/week5/housingprice-beijin.r

@@ -1,0 +1,381 @@
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(rstudioapi) # thư viện dùng để đặt đường dẫn tới file gốc
+setwd(dirname(getActiveDocumentContext()$path))
+library(tidyverse) #
+library(dplyr)
+library(psych)
+library(lubridate) # sử dụng để chuyển đổi biến thời gian
+library(xts) # một thư viện khác để sử lý biến thời gian
+library(tseries) # gói này dùng để sử dụng để phân tích dữ liệu theo thời gian
+library(forecast) # dự đoán tương lại cho các biến thười gian thông qua các mô hình
+
+#Visualisation  --> Các gói thư viện dùng để vẽ đồ thị và trang trí đồ thị
+library(corrplot) # đồ thị vẽ tương quan giữa các biến
+library(plotly) # đồ thị động
+library(viridis) # dùng để tô màu cho bản đồ
+library(ggmap) # sử dụng để lấy bản đồ từ google
+library(knitr) # gói này dùng để tạo bảng đẹp --> trong trường hợp cần dùng
+library(dygraphs) # sử dụng để vẽ đồ thị theo giời gian --> nhưng cũng k cần thiết vì thư viện ggplot cũng có rồi
+library(ggthemes) # tạo theme cho các đồ thị
+
+#### PHẦN 1: Giới thiệu về gói dữ liệu liên quan đến giá nhà tại thành phố Bắc Kinh
+#https://www.kaggle.com/eraw0x/house-prices-in-beijing-eda-arima
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+beijin <- data.frame(read.csv("housingprice-beijin.csv",sep=","))
+beijin <- beijin%>% mutate(floor = str_trim(str_extract(floor,"( .*)"), side = "both"))
+
+beijin <- beijin %>%
+            select(-url, -id, -Cid)
+
+glimpse(beijin)
+summary(beijin)
+
+# 1. Xử lý dữ liệu trống
+
+# Tìm hiểu về các biến có dữ liệu trống
+
+library(psych)
+library(lubridate)
[…]

### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Data/dieutra-hogiadinh-hcm/dieutra-hogiadinh-hcm.r

@@ -1,0 +1,358 @@
+###################
+#     BƯỚC 1: Khai báo các gói phụ kiện (package) R cần thiết để phân tích
+library(rstudioapi) # thư viện dùng để đặt đường dẫn tới file gốc
+setwd(dirname(getActiveDocumentContext()$path))
+
+# -----------------------------------------------------------------
+library(dplyr) # gói phụ kiện để tương tác và thay đổi dữ liệu
+library(readxl)# gói phụ kiện để đọc dữ liệu từ bảng tính excel
+library(ggplot2)
+library(plyr)
+
+###################
+
+#     BƯỚC 2: Đọc dữ liệu vào R
+
+df=data.frame(read_excel("dieutra-hogiadinh-hcm.xlsx",sheet="ketquadieutra",skip = 1)) #
+
+head(df)
+
+library(DT)
+datatable(head(df))
+datatable(tail(df))
+
+str(df)
+glimpse(df)
+summary(df)
+
+df$Q1a <- as.integer(df$Q1a) #chọn ra một số biến và định nghĩa lại
+
+library(purrr) #
+
+df <- df %>%
+  purrr::modify_at(c(2:10, 13:14, 16:21,25:30,44,46,50:51,54:55,58:59,62:63,66:67,70:71,74:75,78,79,82:83,86:87,89:90,92:229), factor)
+
+glimpse(df)
+
+df$Q4<- as.character(df$Q4)
+df$Q6b <- as.character(df$Q6b)
+#df$Q9 <- as.integer(df$Q9) #chọn ra một số biến và định nghĩa lại
+df$Q11<- as.character(df$Q11)
+df$Q18a8<- as.numeric(df$Q18a8)
+df$Q18b8<- as.numeric(df$Q18b8)
+df$Q18a9<- as.numeric(df$Q18a9)
+df$Q18b9<- as.numeric(df$Q18b9)
+df$Q18a10<- as.numeric(df$Q18a10)
+df$Q18b10<- as.numeric(df$Q18b10)
+df$Q44b <- as.character(df$Q44b)
+str(df)
+
+#df<-df%>%
+#  mutate_if(is.numeric,as.factor) #Biến đổi toàn bộ giá trị numeric thành factor
+
+# check if there is any strange
+str(df)
+glimpse(df)
+# Tại sao các cột cần được mã hóa.
+
+# ********************************************************************************
+#
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/01-gioi-thieu.Rmd
update

@@
 author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
-date      : "4-tháng 8-2023"
+date      : "tháng 8-2023"
 output    :

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/03-cac-nguyen-ly-co-ban.Rmd
update

@@
 author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
-date      : "4-tháng 8-2023"
+date      : "tháng 8-2023"
 output    : 

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/04-tidydata.Rmd
update

@@
 author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
-date      : "4-tháng 8-2023"
+date      : "tháng 8-2023"
 #date      : "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-## Column headers are values, not variable names (1)
+## Tên Các Cột Là Giá Trị, Không Phải là Tên Biến (1)
 
-<div class='left' style='float:left;width:48%'>
+```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
+knitr::include_graphics("assets/img/pew-raw.png")
+```
 
-Messy Data
+<!-- ## Column headers are values, not variable names (2) -->
+
+<!-- Billboard Ranking -->
+
+<!-- ```{r echo=FALSE, message=FALSE,warning=FALSE} -->
+<!-- df=read.csv("../data/tidy-data/data/billboard.csv", sep=",", header=TRUE) -->
+<!-- library(DT) -->
+<!-- p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'), -->
+<!--              options = list(pageLength = 10, -->
+<!--             columnDefs = list(list(className = 'dt-left')) -->
+<!--             ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE) -->
+<!--              ) -->
+<!-- p -->
+<!-- ``` -->
+
+## Tên Các Cột Là Giá Trị, Không Phải là Tên Biến (2)
+
+Bảng xếp hạng âm nhạc Billboard
 
 ```{r echo=FALSE, message=FALSE,warning=FALSE}
-source("../Code/pew-tidying.R")
-#library(dplyr)
-#library(reshape2)
-#df=read.csv("../Data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
-</div>
+library(DT)
+source("../data/tidy-data/data/billboard.r")
+p<-datatable(clean,rownames = FALSE,
+             options = list(pageLength = 10,
+            columnDefs = list(list(className = 'dt-left', targets = c(1:6)))
+            ,autoWidth = TRUE)
+             )
+p
+```

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/06-Data-Wrangling-Truc-Quan-Hoa/ggplot2-practice.r

@@ -1,0 +1,412 @@
+# Load ggplot2
+library(ggplot2)
+library(dplyr)
+
+
+# The mtcars dataset is natively available
+# head(mtcars)
+
+# A really basic boxplot.
+
+# 
+
+#### PART4 : Visualization with ggplot2
+#--------------------------------------------------------------
+
+# This libaray was created by Hadley Wickham based on graphical method developed by Leland Wilkison. This defines a standard grammar used to create graphs. Since 2014, ggplots is the most downloadable libaries in R.
+# 
+# GGplot2 works based on the concept of Layering. Each part of the graph is visualized using a specific layer such as layer for coordinate, layer for size, layer for lable, layer for specific type of graph.. 
+# 
+# Ok then how can we practice, now if we want to visualize the relationship between low and high temperature, change color according to season and choose size, how can we do.
+# 
+#  Steps in using ggplot2
+# 
+# A graph is created using ggplot() function, and right after this, we can add layer. 
+# 
+# data <- data.frame(data1)
+# 
+# 
+# ggplot(data = dataframe, aes(x = var1, y= var2, colour = var3, size = var4))
+# 
+# geom_xxx()   ---> this layer is added after ggplot() and used to create type of graph such as boxplot, histogram, etc. Note that if we do not specify anything inside the bracket, R will create a default graph, but if we want to decorate, we can declare inside the bracket what attributes we want for our graph. When we add attribute, new layer will be added on top of the previous declared attribute. that is how it works.
+# 
+# ggplot(data, coordinate)+
+#   layer1+
+#   layer2+
+#   layer3
+
+
+
+
+
+mtcars <-data.frame(mtcars)
+
+
+
+ggplot(mtcars, aes(x="", y=mpg))+
+  geom_boxplot()
+
+
+ggplot(mtcars, aes(x="", y=mpg)) +
+  geom_boxplot() +
+  xlab("")+
+  ylab("Mile per gallon")
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/06-Data-Wrangling-Truc-Quan-Hoa/weather01.r

@@ -1,0 +1,175 @@
+library(rstudioapi) #
+setwd(dirname(getActiveDocumentContext()$path))
+#
+
+library(readxl)
+library(tidyverse)
+library(dplyr)
+library(ggplot2)
+library(rpart)
+library(rpart.plot)
+
+
+weather <-data.frame(read_excel("weather.xlsx",sheet="01",skip = 0)) #
+
+weather$temp.d <- cut(weather$temp, breaks = c(-Inf,20,26,Inf),
+                        labels = c("cool","mild","hot"),include.lowest = F)
+
+weather$humi.d <- cut(weather$humi, breaks = c(-Inf,80,Inf),
+                        labels = c("normal","high"),include.lowest = F)
+
+weather$wind.d <- cut(weather$wind, breaks = c(-Inf,20,Inf),
+                        labels = c("weak","strong"),include.lowest = F)
+
+weather$outlook <- factor(weather$outlook, levels = c("sunny","overcast","rain"))
+weather$temp.d <- factor(weather$temp.d, levels = c("hot","mild","cool"))
+weather$humi.d <- factor(weather$humi.d, levels = c("high","normal"))
+weather$wind.d <- factor(weather$wind.d, levels = c("strong","weak"))
+weather$play <- factor(weather$play, levels = c("yes","no"))
+
+3*3*2*2
+
+write.csv(weather,"weather01-factor.csv", row.names = FALSE)
+
+
+names(weather) #
+
+number.weather <- length(unique(weather$outlook))*length(unique(weather$temp.d))*length(unique(weather$wind.d))*length(unique(weather$play)) #
+number.weather
+table(weather$outlook)
+table(weather$temp.d)
+table(weather$humi.d)
+table(weather$wind.d)
+
+library(vtree) #https://www.infoworld.com/article/3573577/how-to-count-by-groups-in-r.html
+
+vtree(weather, "outlook", palette = 3, sortfill = TRUE)
+
+vtree(weather, "temp.d", palette = 3, sortfill = TRUE)
+
+vtree(weather, "humi.d", palette = 3, sortfill = TRUE)
+vtree(weather, "wind.d", palette = 3, sortfill = TRUE)
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/06-Data-Wrangling-Truc-Quan-Hoa/weather02.r

@@ -1,0 +1,743 @@
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(rstudioapi) # We load this libary in order to use the below command to set the working directory to the file --> path
+setwd(dirname(getActiveDocumentContext()$path))
+
+
+#### PART 1: Introduction to data on weather recorded in on station
+
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(readxl)
+
+#weather <-data.frame(read_excel("weather.xlsx",sheet="02", skip = 0)) # if your data is recorded in excel file
+
+#read from csv file
+weather <- data.frame(read.csv("weather.csv",sep=";",stringsAsFactors=FALSE))
+str(weather)
+library(dplyr)
+glimpse(weather)
+
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+#### PART 2: Data Wrangling
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+
+dim(weather) # Explore the dimension of data
+
+names(weather) # see name of variables (columns)
+
+head(weather) # view the first 10 rows of the data
+
+str(weather) # see the type of each variable
+
+
+# Check missing data
+sum(is.na(weather))
+
+# check number of observations
+nrow(weather)
+sum(complete.cases(weather))
+
+nrow(weather) == sum(complete.cases(weather))
+
+
+#############        CREATE FACTORS
+# Variables can be convertted into Factors type by simply using as.factor() function. THis function is a function of factor() functional family. When using this function, R will arrange automatically based on alphabetic order if the data is STRING type.
+
+class(weather$season) # check the class type for season variables
+summary(weather$season)
+
+weather$season <- factor(weather$season, levels = c("Spring","Summer","Autumn","Winter"))
+
+class(weather$season) # now we have this variable as factor types instead of character type.
+summary(weather$season)
+
+#similarly we us as.factor() to convert data type for other columns.
+weather$day <- as.factor(weather$day)
+weather$month <- as.factor(weather$month)
+weather$dir.wind <- as.factor(weather$dir.wind)
+
+#we check our data again
+str(weather)
+
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week5-examples/2-EDA.r

@@ -1,0 +1,77 @@
+source("1-loadingdata-in-R.r")
+head(Census.Data)
+library(dplyr)
+library(ggplot2)
+
+str(Census.Data)
+summary(Census.Data)
+
+#EDA
+
+boxplot(Census.Data[,2:5])
+
+library(vioplot)
+vioplot(Census.Data$Unemployed, Census.Data$Qualification, Census.Data$White_British, Census.Data$Low_Occupancy, ylim=c(0,100), col = "dodgerblue", rectCol="dodgerblue3", colMed="dodgerblue4")
+
+vioplot(Census.Data$Unemployed, Census.Data$Qualification, Census.Data$White_British, Census.Data$Low_Occupancy, ylim=c(0,100), col = "dodgerblue", rectCol="dodgerblue3", colMed="dodgerblue4", names=c("Unemployed", "Qualifications", "White British", "Occupancy"))
+
+
+#boxplot
+library(reshape)
+library(reshape2)
+Census.Data.Vi <- melt(Census.Data,id.vars ='OA',measure.vars =c('White_British','Low_Occupancy','Unemployed','Qualification'))
+head(Census.Data.Vi)
+
+
+ggplot(Census.Data.Vi , aes(x= variable, y=value, color = variable))+
+  geom_boxplot() +
+  stat_summary(fun=mean, geom="point", shape=20, size=1, color="red", fill="red")+
+  theme(axis.text.x = element_text(angle=0, vjust=0.5,hjust=0.5))+
+  theme(legend.position = "none")
+
+
+
+
+p <- ggplot(Census.Data, aes(Unemployed,Qualification))
+
+p + geom_point(aes(colour = White_British, size = Low_Occupancy))
+
+#correlallation
+
+
+
+# Runs a Pearson's correlation
+cor(Census.Data$Unemployed, Census.Data$Qualification)
+
+data1 <- Census.Data[,2:5]
+
+cor(data1)
+round(cor(data1),2)
+
+corr <- cor(data1)
+
+
+# creates a qplot from our corr matrix
+qplot(x=Var1, y=Var2, data=melt(corr), fill=value, geom="tile") +
+  scale_fill_gradient2(limits=c(-1, 1))
+
+library(corrplot)
+
+# creates a lower triangular corrplot from our corr matrix
+corrplot(corr, type="lower", tl.col="black", tl.srt=45)
+
+# runs a regressions model (y ~ x, data frame)
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week5-examples/4-mapping-pointdata.r

@@ -1,0 +1,44 @@
+source("3-mapping-in-R.r")
+# load the house prices csv file
+#https://dataviz.spatial.ly/worksheets/camden/
+houses <- read.csv("Camden/tables/CamdenHouseSales15.csv")
+
+# we only need a few columns for this practical
+
+houses <- houses[,c(1,2,8,9)]
+# 2D scatter plot
+plot(houses$oseast1m, houses$osnrth1m)
+
+# create a House.Points SpatialPointsDataFrame
+House.Points <-SpatialPointsDataFrame(houses[,3:4], houses, proj4string = CRS("+init=EPSG:27700"))
+
+# This plots a blank base map, we have set the transparency of the borders to 0.4
+tm_shape(OA.Census) + tm_borders(alpha=.4)
+
+# creates a coloured dot map
+tm_shape(OA.Census) + tm_borders(alpha=.4)+
+  tm_shape(House.Points) + tm_dots(col = "Price", palette = "Reds", style = "quantile")
+
+# creates a coloured dot map
+
+tm_shape(OA.Census) + tm_borders(alpha=.4) +
+  tm_shape(House.Points) + tm_dots(col = "Price", scale = 1.5, palette = "Reds", style = "quantile", title = "Price Paid (£)")
+
+# creates a coloured dot map
+tm_shape(OA.Census) + tm_borders(alpha=.4) +
+  tm_shape(House.Points) + tm_dots(col = "Price", scale = 1.5, palette = "Purples", style = "quantile", title = "Price Paid (£)")  +
+  tm_compass() +
+  tm_layout(legend.text.size = 1.1, legend.title.size = 1.4, frame = FALSE)
+
+# creates a proportional symbol map
+tm_shape(OA.Census) + tm_borders(alpha=.4) +
+  tm_shape(House.Points) + tm_bubbles(size = "Price", col = "Price", palette = "Blues", style = "quantile", legend.size.show = FALSE, title.col = "Price Paid (£)") +
+  tm_layout(legend.text.size = 1.1, legend.title.size = 1.4, frame = FALSE)
+
+
+# creates a proportional symbol map
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week5-examples/5-GISandR.r

@@ -1,0 +1,87 @@
+source("4-mapping-pointdata.r")
+#proj4string(OA.Census) <- CRS("+init=EPSG:27700")
+#proj4string(House.Points) <- CRS("+init=EPSG:27700")
+
+# Declaire points in polygon 
+pip <- over(House.Points, OA.Census)
+
+#This is the function of the libarary spover package, used to connect spatial points
+
+# include pip into the dataset
+House.Points@data <- cbind(House.Points@data, pip)
+
+#View(House.Points@data)
+
+# plot the relationshiop between price and unemployment
+plot(log(House.Points@data$Price), House.Points@data$Unemployed)
+
+
+# because there is many house to be sold in one place (polygon), we now care about the average housing for each polygon
+
+OA <- aggregate(House.Points@data$Price, by = list(House.Points@data$OA11CD), mean)
+
+# rename it
+names(OA) <- c("OA11CD", "Price")
+
+# include this into OA.Cencus
+OA.Census@data <- merge(OA.Census@data, OA, by = "OA11CD", all.x = TRUE)
+
+
+head(OA.Census@data)
+
+### plot on the map
+tm_shape(OA.Census) + 
+  tm_fill(col = "Price", style = "quantile", title = "Mean House Price (£)")
+
+## regression model
+
+model <- lm(OA.Census@data$Price ~ OA.Census@data$Unemployed)
+summary(model)
+
+###### Buffer.
+
+
+# a buffer zone of 200 m diameter 
+house_buffers <- gBuffer(House.Points, width = 200, byid = TRUE)
+
+# plot
+tm_shape(OA.Census) + tm_borders() +
+  tm_shape(house_buffers) + tm_borders(col = "blue") +
+  tm_shape(House.Points) + tm_dots(col = "red") 
+
+
+# ----> unite them
+union.buffers <- gUnaryUnion(house_buffers)
+
+# map
+tm_shape(OA.Census) + 
+  tm_borders() +
+  tm_shape(union.buffers) + 
+  tm_fill(col = "blue", alpha = .4) + 
+  tm_borders(col = "blue") +
+  tm_shape(House.Points)+
+  tm_dots(col = "red") 
+
+###Interractive map
+#tmap_mode("view")
+
+tm_shape(House.Points) + 
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week5-examples/6-representing-density.r

@@ -1,0 +1,89 @@
+source("5-GISandR.r")
+library(sp)
+library(raster)
+library(adehabitatHR)
+library(tmap)
+
+
+#### 
+
+# 
+tm_shape(OA.Census) + 
+  tm_fill("Qualification", palette = "Reds", style = "quantile", title = "% Qualification") + tm_borders(alpha=.4) + 
+  tm_shape(House.Points) +   
+  tm_bubbles(size = "Price", col = "Price", palette = "Blues", style = "quantile", legend.size.show = FALSE, title.col = "Price Paid (£)", border.col = "black", border.lwd = 0.1, border.alpha = 0.1) +  
+  tm_layout(legend.text.size = 0.8, legend.title.size = 1.1, frame = FALSE)
+
+## how many points in a polygon
+
+library(GISTools)
+
+house.count <- poly.counts(House.Points, OA.Census) 
+
+choropleth(OA.Census,house.count/poly.areas(OA.Census))
+
+head(house.count)
+
+names(house.count)
+
+### calculating area
+
+## change to sf
+
+library(sf)
+
+OA.Census.sf <- st_as_sf(OA.Census)
+House.Points.sf <- st_as_sf(House.Points)
+
+poly.areas(OA.Census) #  sp
+st_area(OA.Census.sf) # sf
+
+### KDE
+kde.output <- kernelUD(House.Points, h="href", grid = 1000)
+plot(kde.output)
+
+## Raster KDE
+
+kde <- raster(kde.output)
+# sets projection to British National Grid
+projection(kde) <- CRS("+init=EPSG:27700")
+
+# density map and raster, ud is the parameter of kde
+tm_shape(kde) + tm_raster("ud")
+
+bounding_box <- bbox(Output.Areas) # zoom in an area using box
+tm_shape(kde, bbox = bounding_box) + 
+  tm_raster("ud") #plotting the density
+
+# add raster kde on targettted area
+masked_kde <- mask(kde, Output.Areas)
+
+# mapping kde, then add polygon layer
+tm_shape(masked_kde, bbox = bounding_box) + 
+  tm_raster("ud", style = "quantile", n = 100, legend.show = FALSE, palette = "YlGnBu") +
+  tm_shape(Output.Areas) + tm_borders(alpha=.3, col = "white") +
+  tm_layout(frame = FALSE)
+
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week5-examples/practice.r

@@ -1,0 +1,700 @@
+x2 <- seq(0,2*pi,len=100)
+
+y2 <- sin(x2)
+
+y4 <- cos(x2)
+
+
+plot(y2)
+plot(y4)
+
+par(mfrow = c(1,2))
+plot(y2,y4)
+
+polygon(y2,y4,col='lightgreen')
+
+plot(y2,y4, asp=1, type='n')
+polygon(y2,y4,col='lightgreen')
+#install.packages("GISTools", depend = T)
+
+library(GISTools)
+
+
+data(georgia)
+
+class(georgia)
+
+georgia@data
+
+
+appling <- georgia.polys[[1]]
+
+
+View(georgia@data)
+
+plot.new()
+plot(appling, asp=1, type='n', xlab="Easting", ylab="Northing")
+
+polygon(appling, density=10, angle=145)
+
+View(appling[[1]])
+colours()
+
+polygon(appling, col="blue")
+polygon(appling, col=rgb(0,0.5,0.7,0.4))
+
+# set the plot extent
+plot(appling, asp=1, type='n', xlab="Easting", ylab="Northing")
+
+
+# plot the points
+points(x = runif(500,126,132)*10000,
+       y = runif(500,103,108)*10000, pch=16, col='red')
+# plot the polygon with a transparency factor
+
+
+polygon(appling, col=rgb(0,0.5,0.7,0.4))
+
+
+
+
+plot(c(-1.5,1.5),c(-1.5,1.5),asp=1, type='n')
+# plot the green/blue rectangle
+rect(-0.5,-0.5,0.5,0.5, border=NA, col=rgb(0,0.5,0.5,0.7))
+# then the second one
+rect(0,0,1,1, col=rgb(1,0.5,0.5,0.7))
+
+
+data(meuse.grid)
+
+mat = SpatialPixelsDataFrame(points = meuse.grid[c("x", "y")],
+                             data = meuse.grid)
+image(mat, "dist")
+
+
+
+library(RColorBrewer)
+
+greenpal <- brewer.pal(7,'Greens')
+
+image(mat, "dist", col=greenpal)
+library(ggplot2)
+theme_bw()
+theme_dark()
+library(gridExtra)
+
+
+
+appling
+
+appling <- data.frame(appling)
+
+
+colnames(appling) <- c("X", "Y")
+
+#ggplot2
+
+
+p1 <- qplot(X, Y, data = appling, geom = "polygon", asp = 1,
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/week6-examples/house-distance-centroid.r

@@ -1,0 +1,180 @@
+#We will calculate the following
+
+# Diversity index
+
+# shortest distance of one POI layers to other POI layer
+
+# shortest distance to nearby road
+
+#loading in library
+library(dplyr)
+library(ggplot2)
+library(reshape2)
+library(rgdal)
+library(sf)
+library(rgeos)
+library(tmap)
+library(RColorBrewer)
+
+
+
+
+#reading the shapefile
+
+OA.Census <- st_read("Camden/Census_OA_Shapefile.shp")
+plot(OA.Census)
+House.Points <- st_read("Camden/Camden_house_sales.shp")
+plot(House.Points)
+
+
+
+#calculate centroid points
+
+cent_sf <- st_centroid(OA.Census)
+cent_sf
+plot(cent_sf)
+
+tm_shape(OA.Census) +
+  tm_fill("olivedrab4") +
+  tm_borders("grey", lwd = 1) +
+  # the points layer
+  tm_shape(cent_sf) +
+  tm_bubbles("Lw_Occp", title.size = "% Occupancy", col = "gold")
+
+## now we want to estimate the distance from every house to the Centroid --> pretending that the POIs is like Metro, Bus Stop or House White :)
+
+## we can do this separately
+
+House.Points
+
+plot(House.Points)
+
+nrow(House.Points)
+nrow(cent_sf)
+
+house.cordinate <-data.frame(st_coordinates(House.Points))
+cent.cordinate <-data.frame(st_coordinates(cent_sf))
+
+head(house.cordinate)
+head(cent.cordinate)
+
+coordinates(house.cordinate) <- ~X + Y
+
+
+str(house.cordinate)
+class(house.cordinate)
+
+proj4string(house.cordinate)
+
+proj4string(house.cordinate) <- CRS("+init=epsg:27700")
+
+
+
+class(house.cordinate)
+house.cordinate <- spTransform(house.cordinate, CRS("+init=epsg:4326"))
+head(house.cordinate@coords)
+head(house.cordinate@coords)
+
+coordinates(cent.cordinate) <- ~X + Y
+str(cent.cordinate)
+proj4string(cent.cordinate)
+proj4string(cent.cordinate) <- CRS("+init=epsg:27700")
+class(cent.cordinate)
+cent.cordinate <- spTransform(cent.cordinate, CRS("+init=epsg:4326"))
+
+
+
+
+
+head(cent.cordinate@coords)
+
+
+
+plot.new()
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Gia-Nha-o-Singapore/eda-visualization.r

@@ -1,0 +1,517 @@
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(rstudioapi) # setting working directory
+setwd(dirname(getActiveDocumentContext()$path)) 
+library(tidyverse)
+library(dplyr)
+library(ggplot2)
+library(scales)
+library(vtree)
+
+df <- data.frame(read.csv("sg-housing-resale-data.csv",sep=",",skip = "0"))
+
+### EDA AND VISUALIZATION
+
+### USING DPLYR
+
+
+df%>%
+  filter(year> 2015, flat_type =="2 ROOM")
+
+
+df%>%
+  filter(year> 2015, flat_type %in% c("2 ROOM","5 ROOM"))
+
+
+df%>%
+  filter(year> 2015, flat_type %in% c("2 ROOM","5 ROOM"))%>%
+  arrange(desc(resale_price))
+
+df%>%
+  group_by(year)%>%
+  summarise(
+    resale_price = mean(resale_price)
+  )
+
+
+
+# 
+df%>%
+  group_by(year)%>%
+  summarise(
+    resale_price = mean(resale_price)
+  )%>%
+  ggplot( aes(x=year, y=resale_price)) +
+  geom_line(color="darkorchid4")+
+  scale_y_continuous(labels = comma)
+# 
+
+
+ggplot(df , aes(x= as.factor(year), y=resale_price, color = as.factor(year)))+
+  geom_boxplot() +
+  stat_summary(fun=mean, geom="point", shape=20, size=1, color="red", fill="red")+
+  scale_y_continuous(breaks = round(seq(0, max(df$resale_price), by = 50000),1))+
+  theme(axis.text.x = element_text(angle=45, vjust=1,hjust=1))+
+  theme(legend.position = "none")
+
+
+
+ggplot(df , aes(x= as.factor(year), y=price_m2, color = as.factor(year)))+
+  geom_boxplot() +
+  stat_summary(fun=mean, geom="point", shape=20, size=1, color="red", fill="red")+
+  scale_y_continuous(breaks = round(seq(0, 15000, by = 500),1))+
+  theme(axis.text.x = element_text(angle=45, vjust=1,hjust=1))+
+  theme(legend.position = "none")+
+  labs(
+         x= "Year",
+         y = "Price/m2"
+       )
+
+
+#we notice that for prediction data prior to 2010 might be not relevant.
+df1 <- df %>%
+  filter(year >= 2010)
+
+# df1 %>% 
[…]

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Gia-Nha-o-Singapore/postgresql-connection.R

@@ -1,0 +1,31 @@
+require(DBI)
+require(RPostgres)
+require(RPostgreSQL)
+
+#con<-dbConnect(dbDriver("PostgreSQL"), dbname="hcmc_population2019", host="localhost", port=5432, user="postgres",password="St029477")
+
+
+con<- dbConnect(RPostgres::Postgres(),user="postgres",password="St029477",host="localhost",port=5432,dbname="sghouseprice")
+
+dbListTables(con)
+# Select
+
+df1 <- dbGetQuery(con, "SELECT * FROM communes")
+library(dplyr)
+
+print(df1)
+
+glimpse(df1)
+
+
+
+### below code is for another example (for testing only)
+
+
+df2 <- dbGetQuery(con, "select id as id, geom, ST_Area(geom,false)/1000000 as area, tongdans01/(ST_Area(geom,false)/1000000) as density
+from communes")
+
+
+
+df3 <- dbGetQuery(con, "select id as id, geom, ST_Area(geom,false)/1000000 as area, tongdans01/(ST_Area(geom,false)/1000000) as density
+from communes")

### 2023-08-03 2023-NHA-TRANG-Data-Analytics-for-Managers/Gia-Nha-o-Singapore/sg-house-price-prediction.r

@@ -1,0 +1,680 @@
+#--------------------------------------------------------------
+#--------------------------------------------------------------
+library(rstudioapi) # setting working directory
+setwd(dirname(getActiveDocumentContext()$path)) 
+
+
+#loading required libraries
+library(tidyverse)
+
+
+library(dplyr)
+library(ggplot2)
+
+
+
+library(lubridate)
+library(scales)
+library(vtree)
+library(plotly)
+library(hrbrthemes)
+library(gridExtra)
+library(ggthemes)
+library(hydroTSM)
+
+
+df <- data.frame(read.csv("csv/merged_file.csv",sep=",")) #reading data from csv file. Refer to other code for compilation of data.
+
+View(df)
+
+df[2]
+
+
+
+df<-df[-1] #delete the first column
+glimpse(df)
+str(df)
+
+head(df) #see first 10 observations
+tail(df) # see last 10 observations
+
+
+#dealing with data, month, and year
+df$month<- paste(df$month,"-01",sep="") #adding -01 in so the system can regconize as data. To be use later.
+
+df$month <- as.Date(df$month) #converting to data
+
+
+###################33
+
+######################
+df <- df%>%
+  mutate(date = month,.before = month)%>% #define new column name date
+  mutate(year = year(date), .after = date)%>% # define new column name year
+  mutate(month=month(month))# extract only value of month
+str(df)
+
+
+df$year <- as.integer(df$year) # convert year to integer
+df$month <- as.integer(df$month) # convert month to integer
+head(df)
+
+summary(df) # summary
+
+x1 <- map_df(df, function(x){sum(is.na(x))}) # function to check if there are any NULL and how many NULL for each variables.
+x1
+
+
+#counting the number of towns
+df%>%
+  count(town)
+# or
+table(df$town)
+
+
+df %>%
+  tally()
+
+# counting the number of flat by type
+df%>%
+  count(flat_type)
+
+# we notice that there are MULTI GENERATION", "MULTI-GENERATION", which they shall be the same. Thus, we use the following code to recode the wrongly duplicated one
+df<-df%>%
+  mutate_at("flat_type", str_replace, "MULTI GENERATION", "MULTI-GENERATION")
+
+
[…]
